# Multiple (Linear) Regression

```{r packages}
#| include: false
library("tidyverse")
theme_set(theme_bw())
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **write** the multiple regression model in matrix form and **derive** the
least squares estimator $\hat\beta = (\mathbf{X}^\top \mathbf{X})^{-1}\mathbf{X}^\top y$,
- **encode** a categorical explanatory variable with dummy variables and
**interpret** the resulting coefficients relative to the reference level,
- **interpret** a coefficient and a coefficient on a logged response or
explanatory variable,
- **construct** t-tests and confidence intervals for coefficients, and
**distinguish** a confidence interval for the mean response at $x_0$ from a
prediction interval for a new observation, and
- **conduct** an F-test comparing two nested models and **explain** the
F-statistic in terms of the residual sums of squares of the reduced and full
models.
:::

Recall our non-linear regression model:
$$Y = f(X) + \epsilon$$

where $f$ is an unknown function and $\epsilon$ is a random error term with mean 0 and variance $\sigma^2$. 
We now consider $X$ a vector of $p$ explanatory variables, 
and we can use a multiple linear regression model to approximate $f$.

## Multiple explanatory variables

Meadowfoam is grown for the oil in its seed, and how much light the plants get
affects how many flowers they produce. In one experiment, plots of meadowfoam
were raised under a range of light intensities, with the light begun either
early or late in the growing cycle, and the average number of flowers per
plant recorded for each plot (`Sleuth3::case0901`). Two explanatory variables
were varied on purpose, so any model of flower production has to account for
both of them at once.

```{r case0901}
d <- Sleuth3::case0901 |>
    mutate(Time = case_when(
        Time == 1 ~ "Late",
        Time == 2 ~ "Early"
    ))

g <- ggplot(d, aes(x = Intensity, y = Flowers, color = Time, shape = Time)) +
    geom_point() +
    labs(title = "Sleuth3::case0901")

g
```

Flower production falls as intensity rises, and at nearly every intensity the
plots whose light began early out-produced those whose light began late. A
model containing intensity alone would have to charge that vertical gap to
error, and a model containing timing alone would have to charge the downward
trend to error. Multiple regression holds both.


## Model

The multiple (linear) regression model is

$$Y_i = \beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip} + \epsilon_i$$

where $\epsilon_i$ is a random error term with $E[\epsilon_i] = 0$ and 
$Var[\epsilon_i] = \sigma^2$.

Alternatively the model can be written

$$Y = \mathbf{X}\beta + \epsilon$$

where

- $Y = (Y_1, \ldots, Y_n)^\top$ is the vector of responses,
- $\mathbf{X}$ is the $n \times (p+1)$ model matrix with rows
$$X_i = (1, X_{i1}, \ldots, X_{ip})$$
- $\beta = (\beta_0, \ldots, \beta_p)^\top$ is the vector of regression coefficients, and
- $\epsilon = (\epsilon_1, \ldots, \epsilon_n)^\top$ is the vector of random errors.

The error assumptions are now $E[\epsilon] = 0$ and $Var[\epsilon] = \sigma^2 I$, where $I$ is the $n \times n$ identity matrix.

Throughout we will assume the $\mathbf{X}$ matrix is of full rank, i.e. $\mathrm{rank}(\mathbf{X}) = p+1$,
but later we will learn about methods such as *ridge regression*, *lasso* and 
*principal component regression* that can be used when this assumption is violated.

## Categorical explanatory variables

If you have categorical explanatory variables, you can include them in the model using *dummy variables*.
Suppose you have a categorical variable $Z$ with $k$ levels. You can create $k-1$ dummy variables to represent the levels of $Z$. 

For example, if $Z$ has levels A, B, and C, let's choose C as the 
*reference level* and create two dummy variables for A and B:

$$D_A = \mathrm{I}(Z=A) = \left\{ \begin{array}{ll} 1 & \text{if $Z$ = A} \\
0 & \text{otherwise} \end{array} \right.$$
and 

$$D_B = \mathrm{I}(Z=B) = \left\{ \begin{array}{ll} 1 & \text{if $Z$ = B} \\
0 & \text{otherwise} \end{array} \right.$$

Then, you can include these dummy variables in the regression model as follows:
$$Y_i = \beta_0 + \beta_1 D_{iA} + \beta_2 D_{iB} + \epsilon_i.$$



## Interpretation

The interpretation of the parameters is as follows:

- $\beta_0$ is the expected response when all explanatory variables are zero and
- $\beta_j$ is the expected change in $Y$ for a one-unit increase in $X_j$ when all other explanatory variables are held constant.

When using polynomial terms, e.g. quadratic, or interactions, 
the interpretation of the coefficients becomes more complex. 


::: {.callout-note collapse="true"}
### Interpretations using logarithms
A common transformation is to take the logarithm of the response variable, the explanatory variable, or both. The interpretations of the parameters change accordingly:

- If we take the logarithm of the response variable, then $e^{\beta_j}$ represents the multiplicative change in the median response for a one-unit increase in the explanatory variable *when all other explanatory variables are held constant*.
- If we take the logarithm of the explanatory variable, then $\beta_j\log(2)$ represents the expected change in the response for a doubling of the explanatory variable *when all other explanatory variables are held constant*.
- If we take the logarithm of both the response and explanatory variables, then $2^{\beta_j}$ represents the multiplicative change in the median response for a doubling of the explanatory variable *when all other explanatory variables are held constant*.
:::

::: {.callout-note collapse="true"}
### Beyond this course
Another tool that helps interpretation of models is the use of contrasts.
Contrasts are linear combinations of the regression coefficients that allow us
to test specific hypotheses about the relationships between the explanatory
variables and the response variable.
:::

## Estimation

The vector $\beta = (\beta_0, \beta_1, \ldots, \beta_p)^\top$ is estimated using 
the method of least squares:

$$\begin{array}{rl}\hat{\beta}
&= \arg\min_{\beta} \sum_{i=1}^n (y_i - \beta_0 - \beta_1 X_{i1} - \cdots - \beta_p X_{ip})^2\\
&= \arg\min_{\beta} (y-\mathbf{X}\beta)^\top (y-\mathbf{X}\beta) \\
&= \left(\mathbf{X}^\top \mathbf{X}\right)^{-1} \mathbf{X}^\top y
\end{array}$$

These are commonly referred to as the *ordinary least squares* (OLS) estimators. 

The residual sum of squares (RSS) are

$$RSS = \left(y-\mathbf{X}\hat\beta\right)^\top \left(y-\mathbf{X}\hat\beta\right)$$

We can use this to estimate the error variance $\sigma^2$ using

$$\hat{\sigma}^2 = \frac{RSS}{n-(p+1)}$$

where $n-(p+1)$ is the degrees of freedom, 
since we have estimated $(p+1)$ coefficients ($\beta_0$ through $\beta_p$).

```{r fit-model}
m <- lm(Flowers ~ Intensity + Time, data = d) # Additive model
```

Visualization of residuals:
```{r visualize-residuals}
shared_jitter <- position_jitter(width = 0.5, height = 0, seed = 20260914)

fits <- tibble(fitted = fitted(m), observed = fitted(m) + residuals(m))

ggplot(fits, aes(x = fitted, y = observed)) +
    geom_point(position = shared_jitter) +
    geom_segment(
        aes(xend = fitted, yend = fitted),
        position = shared_jitter,
        color = "red") +
    geom_abline(slope = 1, intercept = 0, color = "blue") +
    labs(
        title = "Fitted vs Observed",
        x = "Predicted value", 
        y = "Observed value")


```

Parameter estimates:
```{r parameter-estimates}
coef(m)
summary(m)$sigma
```


## Inference

For inference, we need to assume normality of the errors, 
i.e. $$\epsilon_i \stackrel{ind}{\sim} N(0, \sigma^2).$$
This results in the model
$$Y_i \stackrel{ind}{\sim} N(\beta_0 + \beta_1 X_{i1}+\cdots+\beta_pX_{ip}, \sigma^2)$$
or, equivalently,

$$Y \sim N_n(\mathbf{X}\beta, \sigma^2 I)$$
where $N_n$ is the $n$-dimensional multivariate normal distribution.

With the normality assumption, we can write the likelihood function:

$$\begin{array}{rl}
L(\beta, \sigma^2) &= \prod_{i=1}^n \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{1}{2\sigma^2}[Y_i - \beta_0 - \beta_1 X_{i1} - \cdots - \beta_pX_{ip}]^2\right) \\
&= (2\pi\sigma^2)^{-n/2} \exp\left(-\frac{1}{2\sigma^2} [y-\mathbf{X}\beta]^\top [y-\mathbf{X}\beta] \right)
\end{array}$$

the estimates for $\beta$ that maximize this likelihood are the same as those obtained from least squares.
These are referred to as the *maximum likelihood estimators* (MLEs) of $\beta$.

### t-tests

For testing hypotheses about the regression coefficients, we use t-tests. The test statistic for $\beta_j$ is

$$t = \frac{\hat{\beta}_j - \beta_j^0}{\text{SE}(\hat{\beta}_j)}$$

which, under the null hypothesis $H_0: \beta_j = \beta_j^0$,
follows a t-distribution with $n-(p+1)$ degrees of freedom.

```{r t-tests}
summary(m)$coefficients
```

### Confidence intervals

A $(1-\alpha)100\%$ confidence interval for $\beta_j$ is given by

$$\hat{\beta}_j \pm t_{n-(p+1), 1-\alpha/2} \cdot \text{SE}\left(\hat{\beta}_j\right)$$

where $t_{n-(p+1), 1-\alpha/2}$ is the $(1-\alpha/2)$ quantile of the t-distribution with $n-(p+1)$ degrees of freedom.

The standard error of $\hat{\beta}_j$ is given by
$$\text{SE}\left(\hat{\beta}_j\right) = \sqrt{\hat{\sigma}^2 \left(\mathbf{X}^\top \mathbf{X}\right)^{-1}_{jj}}$$
where $\left(\mathbf{X}^\top \mathbf{X}\right)^{-1}_{jj}$ is the $j$-th diagonal element of the 
inverse of the matrix $\mathbf{X}^\top \mathbf{X}$.

```{r confidence-intervals}
confint(m, level = 0.95)
```

::: {.callout-note collapse="true"}
### Beyond this course
Contrasts also utilize t-tools, e.g. t-tests and confidence intervals.
:::

### Confidence intervals at $x$

A $(1-\alpha)100\%$ confidence interval for the **mean response** at a specific value $x_0$ is given by

$$\hat{f}(x_0) \pm t_{n-(p+1), 1-\alpha/2} \cdot \text{SE}\left(\hat{f}(x_0)\right)$$

where $x_0$ is now a vector of feature values,
$\hat{f}(x_0) = x_0^\top\hat\beta = \hat{\beta}_0 + \hat{\beta}_1 x_{01} + \cdots + \hat\beta_p x_{0p}$,
and $\text{SE}(\hat{f}(x_0))$ is the *standard error of the predicted mean response*.
This is the uncertainty in the line at a given point $x_0$.

```{r confidence-intervals-at-x}
nd <- unique(d |> select(Intensity, Time))

cx <- bind_cols(nd, 
  predict(
    m, 
    newdata = unique(d |> select(Intensity, Time)), 
    interval = "confidence", 
    level = 0.95)) |>
  rename(
    Flowers = fit
  )

cx
```

```{r confidence-intervals-at-x-plot}
g +
    geom_ribbon(data = cx, aes(ymin = lwr, ymax = upr, fill = Time), color = NA, alpha = 0.2) +
    geom_line(data = cx, aes(linetype = Time))
```

## Prediction

A $(1-\alpha)100\%$ prediction interval for a new observation at a specific value $x_0$ is given by
$$\hat{Y}_0 \pm t_{n-(p+1), 1-\alpha/2} \cdot \text{SE}_{\text{pred}}(\hat{Y}_0)$$

where $\hat{Y}_0 = \hat{f}(x_0) = x_0^\top\hat\beta$ and $\text{SE}_{\text{pred}}(\hat{Y}_0)$ is the *standard error of the predicted value*,
which accounts for both the uncertainty in the mean response and the variability of individual observations.
The key here is that the prediction interval is wider than the confidence interval for the mean response, reflecting the additional uncertainty in predicting a single new observation due to the residual error $\epsilon$.

```{r prediction-intervals-at-x}
px <- bind_cols(nd, 
  predict(
    m, 
    newdata = unique(d |> select(Intensity, Time)), 
    interval = "prediction",  # THIS CHANGED
    level = 0.95)) |>
  rename(
    Flowers = fit
  )

px 
```

```{r prediction-intervals-at-x-plot}
g +
    geom_ribbon(data = px, aes(ymin = lwr, ymax = upr, fill = Time), color = NA, alpha = 0.2) +
    geom_line(data = px, aes(linetype = Time))
```

## F-tests

To evaluate the statistical evidence for a group of explanatory variables,
e.g. a single categorical variable with $k+1$ levels, we can use an F-test.
The null hypothesis is that all coefficients for the group of variables are equal to zero, i.e
$$H_0: \beta_{j_1} = \beta_{j_2} = \cdots = \beta_{j_k} = 0$$
where $j_1, j_2, \ldots, j_k$ are the indices of the coefficients
for the group of variables. The alternative hypothesis is that at least one of the coefficients is not equal to zero. The F-statistic is calculated as
$$F_{k, n-(p+1)} 
= \frac{(RSS_r - RSS_f)/k}{RSS_f/(n-[p+1])}
= \frac{(RSS_r - RSS_f)/k}{\hat\sigma^2_{f}}$$

where 

- $RSS_r$ is the residual sum of squares for the reduced model (without the group of variables), 
- $RSS_f$ is the residual sum of squares for the full model (with the group of variables),
- $k$ is the difference in the number of coefficients between the two models, and 
- $n-(p+1)$ is the degrees of freedom for the full model. 

Under the null hypothesis, 
the F-statistic follows an F-distribution with $k$ and $n-(p+1)$ degrees of freedom.
A $p$-value can be calculated from the F-distribution to assess the statistical significance of the group of variables.

```{r f-tests}
# Default
anova(m)
drop1(m, test="F")
tail(capture.output(summary(m)), n = 2)[1]

# Manual (for nested models)
m0 <- lm(Flowers ~ 1, data = d) # Intercept only model
anova(m0, m)
```


## Model Assessment

### R-squared

The coefficient of determination, $R^2$, measures the proportion of the variance in the response variable that is predictable from the explanatory variable(s). It is defined as
$$R^2 = 1 - \frac{RSS}{TSS}$$

where $RSS$ is the residual sum of squares and $TSS$ is the total sum of squares, i.e.

$$TSS = \sum_{i=1}^n (y_i - \bar{y})^2.$$

Thus, $R^2$ can be interpreted as the proportion of the total variability in the response variable that is explained by the model.

```{r r-squared}
summary(m)$r.squared
```

## Meadowfoam flower production

Collecting the pieces above into one analysis: does light intensity change the
number of flowers a meadowfoam plant produces, and does the timing of the
light matter once intensity is accounted for?

The experiment varied intensity over `r length(unique(d$Intensity))` levels
and timing over `r length(unique(d$Time))`, with
`r nrow(d) / (length(unique(d$Intensity)) * length(unique(d$Time)))` plots at
each combination and `r nrow(d)` plots in all, so the design is balanced and
neither explanatory variable can stand in for the other. Before fitting the
model, look at the data with a separate least squares line through each timing
group:

```{r flowers-plot}
g + geom_smooth(method = "lm", formula = y ~ x)
```

Both groups trend downward at a similar rate, with the early group's line
sitting above the late group's across the whole range of intensity. The two
slopes are close enough that a model forcing them to be equal — one slope on
intensity, plus a constant vertical shift for timing — looks like a reasonable
description of these data.

That model is the additive fit `m` from above, `Flowers ~ Intensity + Time`:

```{r flowers-summary}
summary(m)
```

Read the coefficients in order.

- $\hat\beta_0 = `r round(coef(m)[1], 1)`$ is the expected number of flowers
at zero intensity for the reference timing, which R takes to be `Early`
because it orders the levels alphabetically. Zero light lies outside the
observed intensity range of `r min(d$Intensity)` to `r max(d$Intensity)`
μmol/m²/sec, so the intercept anchors the fitted lines rather than predicting
a realizable condition.
- $\hat\beta_1 = `r round(coef(m)[2], 4)`$ says that each additional
μmol/m²/sec of light is associated with `r round(-coef(m)[2], 4)` fewer
flowers, holding timing fixed. Over the `r diff(sort(unique(d$Intensity))[1:2])`
units separating consecutive intensities in the design, that is
`r round(-diff(sort(unique(d$Intensity))[1:2]) * coef(m)[2], 1)` fewer flowers.
- $\hat\beta_2 = `r round(coef(m)[3], 1)`$ is the coefficient on the dummy
variable for `Late`, so plots whose light began late produced that many fewer
flowers than plots at the *same* intensity whose light began early. Its 95%
confidence interval, $(`r round(confint(m)[3, 1], 1)`,
`r round(confint(m)[3, 2], 1)`)$, excludes zero.

The F-test against the intercept-only model gives
$F = `r round(summary(m)$fstatistic[1], 1)`$ on
$`r summary(m)$fstatistic[2]`$ and $`r summary(m)$fstatistic[3]`$ degrees of
freedom, and the two explanatory variables together account for
$R^2 = `r round(100 * summary(m)$r.squared)`\%$ of the variability in flower
counts.

Both explanatory variables earn their place: flower production declines with
intensity, and at any fixed intensity beginning the light early is worth about
`r round(-coef(m)[3], 0)` additional flowers per plant. The one assumption the
data have not been asked about is the parallel-lines restriction the additive
model imposes — that the slope on intensity is the same for both timings.

## Conclusion

Multiple linear regression writes the mean response as a linear combination of
$p$ explanatory variables,
$Y_i = \beta_0 + \beta_1X_{i1} + \cdots + \beta_pX_{ip} + \epsilon_i$, or in
matrix form $Y = \mathbf{X}\beta + \epsilon$. Least squares gives
$\hat\beta = (\mathbf{X}^\top\mathbf{X})^{-1}\mathbf{X}^\top y$ whenever
$\mathbf{X}$ has full rank, the normal-error likelihood makes that same vector
the maximum likelihood estimator, and the error variance is estimated by
$\hat\sigma^2 = RSS/[n - (p+1)]$, one degree of freedom spent per estimated
coefficient.

A categorical explanatory variable with $k$ levels enters through $k-1$ dummy
variables defined against a reference level: the intercept absorbs the
reference level, and each dummy's coefficient is a difference from it. Every
coefficient in the model, dummy or not, now carries the qualifier *holding the
other explanatory variables fixed*, which is what separates it from the
coefficient the same variable would get in a regression by itself.

Inference divides by the question. A t-test and a confidence interval speak
about one coefficient; a confidence interval for the mean response at $x_0$
speaks about where the fitted surface is; a prediction interval speaks about
where a new observation will fall and is wider by that observation's own error
variance. Testing several coefficients at once — every dummy from one
categorical variable, say — requires the F-test, which compares the residual
sums of squares of a reduced and a full model and asks whether the drop is
larger than the extra parameters can explain by chance.

Next lecture keeps this model and the estimator derived for it, but stops
requiring the columns of $\mathbf{X}$ to be the raw explanatory variables.
Polynomials and interactions are features built from those variables, and
because the model stays linear in $\beta$, everything above carries over
unchanged.
