# Multiple (Linear) Regression

```{r packages}
#| include: false
library("tidyverse")
theme_set(theme_bw())
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **write** the multiple regression model in matrix form and **derive** the
least squares estimator $\hat\beta = (X^\top X)^{-1}X^\top y$,
- **encode** a categorical explanatory variable with dummy variables and
**interpret** the resulting coefficients relative to the baseline level,
- **interpret** a coefficient, a contrast, and a coefficient on a logged
response or explanatory variable,
- **construct** t-tests and confidence intervals for coefficients and contrasts,
and **distinguish** a confidence interval for the mean response at $x_0$ from a
prediction interval for a new observation, and
- **conduct** an F-test comparing two nested models and **explain** the
F-statistic in terms of the residual sums of squares of the reduced and full
models.
:::

Recall our non-linear regression model:
$$Y = f(X) + \epsilon$$

where $f$ is an unknown function and $\epsilon$ is a random error term with mean 0 and variance $\sigma^2$. 
We now consider $X$ a vector of $p$ explanatory variables, 
and we can use a multiple linear regression model to approximate $f$.

## Example

```{r case0901}
d <- Sleuth3::case0901 |>
    mutate(Time = case_when(
        Time == 1 ~ "Late",
        Time == 2 ~ "Early"
    ))

g <- ggplot(d, aes(x = Intensity, y = Flowers, color = Time, shape = Time)) +
    geom_point() +
    labs(title = "Sleuth3::case0901")

g
```


## Model

The multiple (linear) regression model is

$$Y_i = \beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip} + \epsilon_i$$

where $\epsilon_i$ is a random error term with $E[\epsilon_i] = 0$ and 
$Var[\epsilon_i] = \sigma^2$.

Alternatively the model can be written

$$Y = X\beta + \epsilon$$

where

- $Y = (Y_1, \ldots, Y_n)^\top$ is the vector of responses,
- $X$ is the $n \times (p+1)$ model matrix with rows
$$X_i = (1, X_{i1}, \ldots, X_{ip})$$
- $\beta = (\beta_0, \ldots, \beta_p)^\top$ is the vector of regression coefficients, and
- $\epsilon = (\epsilon_1, \ldots, \epsilon_n)^\top$ is the vector of random errors.

The error assumptions are now $E[\epsilon] = 0$ and $Var[\epsilon] = \sigma^2 I$, where $I$ is the $n \times n$ identity matrix.

Throughout we will assume the $X$ matrix is of full rank, i.e. $rank(X) = p+1$,
but later we will learn about methods such as *ridge regression*, *lasso* and 
*principal component regression* that can be used when this assumption is violated.

## Categorical explanatory variables

If you have categorical explanation variables, you can include them in the model using *dummy variables*.
Suppse you have a categorical variable $Z$ with $k$ levels. You can create $k-1$ dummy variables to represent the levels of $Z$. 

For example, if $Z$ has levels "A", "B", and "C", 
you can create two dummy variables:

- $D_1 = 1$ if $Z = $"A", 0 otherwise
- $D_2 = 1$ if $Z = $"B", 0 otherwise

Then, you can include these dummy variables in the regression model as follows:
$$Y_i = \beta_0 + \beta_1 D_{i1} + \beta_2 D_{i2} + \epsilon_i.$$



## Interpretation

The interpretation of the parameters is as follows:

- $\beta_0$ is the expected response when all explanatory variables are zero and
- $\beta_j$ is the expected change in $Y$ for a one-unit increase in $X_j$ when all other explanatory variables are held constant.

When using polynomial terms, e.g. quadratic, or interactions, 
the interpretation of the coefficients becomes more complex. 


::: {.callout-note collapse="true"}
### Interpretations using logarithms
A common transformation is to take the logarithm of the response variable, the explanatory variable, or both. The interpretations of the parameters change accordingly:

- If we take the logarithm of the response variable, then $e^{\beta_j}$ represents the multiplicative change in the median response for a one-unit increase in the explanatory variable *when all other explanatory variables are held constant*.
- If we take the logarithm of the explanatory variable, then $\beta_1\log(2)$ represents the expected change in the response for a doubling of the explanatory variable *when all other explanatory variables are held constant*.
- If we take the logarithm of both the response and explanatory variables, then $2^\beta_1$ represents the multiplicative change in the median response for a doubling of the explanatory variable *when all other explanatory variables are held constant*.
:::

::: {.callout-note collapse="true"}
### Contrasts
Another tool that helps interpretation of models is the use of contrasts. 
Contrasts are linear combinations of the regression coefficients that allow us 
to test specific hypotheses about the relationships between the explanatory 
variables and the response variable.
:::

## Estimation

The vector $\beta = (\beta_0, \beta_1, \ldots, \beta_p)^T$ is estimated using 
the method of least squares:

$$\begin{array}{rl}\hat{\beta}
&= \arg\min_{\beta} \sum_{i=1}^n (y_i - \beta_0 - \beta_1 X_{i1} - \cdots - \beta_p X_{ip})^2\\
&= \arg\min_{\beta} (y-X\beta)^\top (y-X\beta) \\
&= \left(X^\top X\right)^{-1} X^\top y
\end{array}$$

This is commonly referred to as the *ordinary least squares* (OLS) estimators. 

The residual sum of squares (RSS) are

$$RSS = \left(y-X\hat\beta\right)^\top \left(y-X\hat\beta\right)$$

We can use this to estimate the error variance $\sigma^2$ using

$$\hat{\sigma}^2 = \frac{RSS}{n-(p+1)}$$

where $n-(p+1)$ is the degrees of freedom, 
since we have estimated $(p+1)$ coefficients ($\beta_0$ through $\beta_p$).

```{r fit-model}
m <- lm(Flowers ~ Intensity + Time, data = d) # Additive model
```

Visualization of residuals:
```{r visualize-residuals}
shared_jitter <- position_jitter(width = 0.5, height = 0, seed = 20260914)
ggplot(m, aes(x = .fitted, y = .fitted + .resid)) +
    geom_point(position = shared_jitter) +
    geom_segment(
        aes(xend = .fitted, yend = .fitted), 
        position = shared_jitter,
        color = "red") +
    geom_abline(slope = 1, intercept = 0, color = "blue") +
    labs(
        title = "Fitted vs Observed",
        x = "Predicted value", 
        y = "Observed value")


```

Parameter estimates:
```{r parameter-estimates}
coef(m)
summary(m)$sigma
```


## Inference

For inference, we need to assume normality of the errors, 
i.e. $$\epsilon_i \stackrel{ind}{\sim} N(0, \sigma^2).$$
This results in the model
$$Y_i \stackrel{ind}{\sim} N(\beta_0 + \beta_1 X_{i1}+\cdots+\beta_pX_{ip}, \sigma^2)$$
or, equivalently,

$$Y \sim N_n(X\beta, \sigma^2\mathrm{I})$$
where $N_n$ is the $n$-dimensional multivariate normal distribution.

With the normality assumption, we can write the likelihood function:

$$\begin{array}{rl}
L(\beta, \sigma^2) &= \prod_{i=1}^n \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{1}{2\sigma^2}[Y_i - \beta_0 - \beta_1 X_{i1} - \cdots - \beta_pX_{ip}]^2\right) \\
&= (2\pi\sigma^2)^{-n/2} \exp\left(-\frac{1}{2\sigma^2} [y-X\beta]^\top [y-X\beta] \right)
\end{array}$$

the estimates for $\beta$ that maximize this likelihood are the same as those obtained from least squares.
These are referred to as the *maximum likelihood estimators* (MLEs) of $\beta$.

### t-tests

For testing hypotheses about the regression coefficients, we use t-tests. The test statistic for $\beta_j$ is

$$t = \frac{\hat{\beta}_j - \beta_j^0}{\text{SE}(\hat{\beta}_j)}$$

which, under the null hypothesis $H_0: \beta_j = \beta_j^0$,
follows a t-distribution with $n-(p+1)$ degrees of freedom.

```{r t-tests}
summary(m)$coefficients
```

::: {.callout-note collapse="true"}
### Contrasts
Contrasts also utilize t-tools, e.g. t-tests and confidence intervals.
:::


### Confidence intervals

A $(1-\alpha)100\%$ confidence interval for $\beta_j$ is given by

$$\hat{\beta}_j \pm t_{n-(p+1), 1-\alpha/2} \cdot \text{SE}\left(\hat{\beta}_j\right)$$

where $t_{n-2, 1-\alpha/2}$ is the $(1-\alpha/2)$ quantile of the t-distribution with $n-2$ degrees of freedom.

The standard error of $\hat{\beta}_j$ is given by
$$\text{SE}\left(\hat{\beta}_j\right) = \sqrt{\hat{\sigma}^2 \left(X^\top X\right)^{-1}_{jj}}$$
where $\left(X^\top X\right)^{-1}_{jj}$ is the $j$-th diagonal element of the 
inverse of the matrix $X^\top X$.

```{r confidence-intervals}
confint(m, level = 0.95)
```

::: {.callout-note collapse="true"}
### Contrasts
Contrasts also utilize t-tools, e.g. t-tests and confidence intervals.
:::

### Confidence intervals at $x$

A $(1-\alpha)100\%$ confidence interval for the **mean response** at a specific value $x_0$ is given by

$$\hat{f}(x_0) \pm t_{n-(p+1), 1-\alpha/2} \cdot \text{SE}\left(\hat{f}(x_0)\right)$$

where $\hat{f}(x_0) = \hat{\beta}_0 + \hat{\beta}_1 x_0$ and $\text{SE}(\hat{f}(x_0))$ is the *standard error of the predicted mean response*.
This is the uncertainty in the line at a given point $x_0$.

```{r confidence-intervals-at-x}
nd <- unique(d |> select(Intensity, Time))

cx <- bind_cols(nd, 
  predict(
    m, 
    newdata = unique(d |> select(Intensity, Time)), 
    interval = "confidence", 
    level = 0.95)) |>
  rename(
    Flowers = fit
  )

cx
```

```{r confidence-intervals-at-x-plot}
g + 
    geom_ribbon(data = cx, aes(ymin = lwr, ymax = upr), color = NA, alpha = 0.2) +
    geom_line(data = cx)
```

## Prediction

A $(1-\alpha)100\%$ prediction interval for a new observation at a specific value $x_0$ is given by
$$\hat{Y}_0 \pm t_{n-(p+1), 1-\alpha/2} \cdot \text{SE}_{\text{pred}}(\hat{Y}_0)$$

where $\hat{Y}_0 = \hat{f}(x_0) = \hat{\beta}_0 + \hat{\beta}_1 x_0$ and $\text{SE}_{\text{pred}}(\hat{Y}_0)$ is the *standard error of the predicted value*,
which accounts for both the uncertainty in the mean response and the variability of individual observations.
The key here is that the prediction interval is wider than the confidence interval for the mean response, reflecting the additional uncertainty in predicting a single new observation due to the residual error $\epsilon$.

```{r prediction-intervals-at-x}
px <- bind_cols(nd, 
  predict(
    m, 
    newdata = unique(d |> select(Intensity, Time)), 
    interval = "prediction",  # THIS CHANGED
    level = 0.95)) |>
  rename(
    Flowers = fit
  )

px 
```

```{r prediction-intervals-at-x-plot}
g + 
    geom_ribbon(data = px, aes(ymin = lwr, ymax = upr), color = NA, alpha = 0.2) +
    geom_line(data = px)
```

## F-tests

To evaluate the statistical evidence for a group of explanatory variables,
e.g. a single categorical variable with $k$ levels, we can use an F-test.
The null hypothesis is that all coefficients for the group of variables are equal to zero, i.e
$$H_0: \beta_{j_1} = \beta_{j_2} = \cdots = \beta_{j_k} = 0$$
where $j_1, j_2, \ldots, j_k$ are the indices of the coefficients
for the group of variables. The alternative hypothesis is that at least one of the coefficients is not equal to zero. The F-statistic is calculated as
$$F_{k, n-(p+1)} 
= \frac{(RSS_r - RSS_f)/k}{RSS_f/(n-[p+1])}
= \frac{(RSS_r - RSS_f)/k}{\hat\sigma^2_{f}}$$

where 

- $RSS_r$ is the residual sum of squares for the reduced model (without the group of variables), 
- $RSS_f$ is the residual sum of squares for the full model (with the group of variables),
- $k$ is the difference in the number of coefficients between the two models, and 
- $n-(p+1)$ is the degrees of freedom for the full model. 

Under the null hypothesis, 
the F-statistic follows an F-distribution with $k$ and $n-(p+1)$ degrees of freedom.
A $p$-value can be calculated from the F-distribution to assess the statistical significance of the group of variables.

```{r f-tests}
# Default
anova(m)
drop1(m, test="F")
tail(capture.output(summary(m)), n = 2)[1]

# Manual (for nested models)
m0 <- lm(Flowers ~ 1, data = d) # Intercept only model
anova(m0, m)
```


## Model Assessment

### R-squared

The coefficient of determination, $R^2$, measures the proportion of the variance in the dependent variable that is predictable from the independent variable(s). It is defined as
$$R^2 = 1 - \frac{RSS}{TSS}$$

where $RSS$ is the residual sum of squares and $TSS$ is the total sum of squares, i.e.

$$TSS = \sum_{i=1}^n (y_i - \bar{y})^2.$$

Thus, $R^2$ can be interpreted as the proportion of the total variability in the response variable that is explained by the model.

```{r r-squared}
summary(m)$r.squared
```

## Examples

### Flowers

Default plot:
```{r flowers-plot}
g + geom_smooth(method = "lm") # lines are not parallel 
```

Model summary:
```{r flowers-summary}
summary(m)
```
