# Feature Engineering

```{r packages}
#| include: false
library("tidyverse")
theme_set(theme_bw())
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **explain** why a model built from polynomials and interactions is still a
*linear* model,
- **construct** these features and **write** the corresponding `lm()` formula,
- **interpret** a coefficient in each case, and explain why "a one-unit increase,
holding all else constant" can stop making sense, and
- **use** an F-test to decide how much curvature, or whether an interaction, the
data actually support.
:::

## Linear in the parameters

Multiple linear regression looks restrictive: a straight line in each feature.
It is not. Consider

$$Y_i = \beta_0 + \beta_1 h_1(X_i) + \cdots + \beta_p h_p(X_i) + \epsilon_i$$

where $h_1, \ldots, h_p$ are known, fixed functions of the features $X_i$,
called *basis functions*.

This model is linear in $\beta$, which is all least squares ever required.
Every result from last lecture therefore carries over **unchanged**:
$\hat\beta = (X^\top X)^{-1}X^\top y$, the t-tests, the F-tests, the intervals.
Only the columns of the design matrix change.

Choosing the $h_j$ is **feature engineering**, and it is where most of the
modeling judgment in a regression analysis lives.


## Polynomials

Use the powers of a single feature as the basis:

$$Y_i = \beta_0 + \beta_1 X_i + \beta_2 X_i^2 + \cdots + \beta_d X_i^d + \epsilon_i.$$

The fit is now a curve, and no single coefficient is "the effect of $X$." That
effect is the derivative

$$\frac{\partial}{\partial x} E[Y|X = x] = \beta_1 + 2\beta_2 x + \cdots + d\beta_d x^{d-1},$$

which **depends on where you are**. Reporting $\hat\beta_1$ by itself is
meaningless.

```{r polynomials}
#| eval: false
lm(y ~ poly(x, 3))               # orthogonal polynomial basis
lm(y ~ x + I(x^2) + I(x^3))      # raw powers
lm(y ~ poly(x, 3, raw = TRUE))   # also raw powers
```

Two cautions:

- **Hierarchy**: if $X^2$ is in the model, keep $X$. 
- **Extrapolation**: high-degree polynomials behave badly outside the range of the data.

::: {.callout-note collapse="true"}
### Why `poly()` instead of raw powers?
Over a range like $x \in [0,10]$, the columns $x$, $x^2$, and $x^3$ are
enormously correlated — they all increase together. That is collinearity, and it
inflates standard errors and destabilizes $\hat\beta$ numerically.

`poly()` builds an *orthogonal* basis spanning the same space, so $\hat{y}$,
the RSS, and $R^2$ are **identical**. What changes is that the coefficients are
now uncorrelated, so their t-tests can be read one at a time — a convenient way
to ask "is the cubic term needed?"
:::




### Example: Galileo's falling bodies

Galileo released a ball from a ramp at seven different heights and measured how
far it travelled horizontally. Physics says that relationship should be curved,
so this is a natural place to ask *how much* curvature the data support.

Before fitting anything, look at the relationship itself:

::: {.callout-note collapse="true"}
### Code for the plots below
```{r galileo-plot-setup}
#| output: false
m1 <- lm(Distance ~ Height,          data = Sleuth3::case1001)
m2 <- lm(Distance ~ poly(Height, 2), data = Sleuth3::case1001)
m3 <- lm(Distance ~ poly(Height, 3), data = Sleuth3::case1001)

model_levels    <- c("linear", "quadratic", "cubic")
model_colors    <- c(linear = "#E69F00", quadratic = "#56B4E9", cubic = "#009E73")
model_linetypes <- c(linear = "solid",   quadratic = "dashed",  cubic = "dotted")

galileo_grid <- tibble(Height = seq(100, 1000, length.out = 200))

galileo_pred <- bind_rows(
  galileo_grid |> mutate(Distance = predict(m1, galileo_grid), Model = "linear"),
  galileo_grid |> mutate(Distance = predict(m2, galileo_grid), Model = "quadratic"),
  galileo_grid |> mutate(Distance = predict(m3, galileo_grid), Model = "cubic")
)

# A shared base layer, so the four views below differ only in which lines
# (and legend rows) are layered on top of it. The color/linetype scales are
# attached separately below, only on the plots that actually map them --
# attaching an unused manual scale to the data-only plot triggers a spurious
# "no shared levels" warning from ggplot2.
galileo_base <- ggplot(Sleuth3::case1001, aes(x = Height, y = Distance)) +
  geom_point() +
  coord_cartesian(xlim = range(galileo_grid$Height),
                   ylim = range(Sleuth3::case1001$Distance, galileo_pred$Distance)) +
  labs(title = "Sleuth3::case1001", x = "Release height", y = "Horizontal distance") +
  # Legend drawn inside the panel, bottom right, so the panel itself is the
  # same size on every tab -- an outside legend would grow the plot's total
  # width as rows are added, shifting the panel between tabs.
  theme(legend.position = "inside",
        legend.position.inside = c(0.98, 0.02),
        legend.justification = c(1, 0),
        legend.background = element_rect(fill = alpha("white", 0.8), color = "grey70"))

galileo_data_plot <- galileo_base

galileo_linear_plot <- galileo_base +
  geom_line(
    data = galileo_pred |>
      filter(Model == "linear") |>
      mutate(Model = factor(Model, levels = "linear")),
    aes(color = Model, linetype = Model)
  ) +
  scale_color_manual(values = model_colors) +
  scale_linetype_manual(values = model_linetypes)

galileo_quadratic_plot <- galileo_base +
  geom_line(
    data = galileo_pred |>
      filter(Model %in% c("linear", "quadratic")) |>
      mutate(Model = factor(Model, levels = c("linear", "quadratic"))),
    aes(color = Model, linetype = Model)
  ) +
  scale_color_manual(values = model_colors) +
  scale_linetype_manual(values = model_linetypes)

galileo_cubic_plot <- galileo_base +
  geom_line(
    data = galileo_pred |> mutate(Model = factor(Model, levels = model_levels)),
    aes(color = Model, linetype = Model)
  ) +
  scale_color_manual(values = model_colors) +
  scale_linetype_manual(values = model_linetypes)
```
:::

::: {.panel-tabset}
#### Data
```{r galileo-data-plot}
#| echo: false
galileo_data_plot
```

#### + Linear
```{r galileo-linear-plot}
#| echo: false
galileo_linear_plot
```

#### + Quadratic
```{r galileo-quadratic-plot}
#| echo: false
galileo_quadratic_plot
```

#### + Cubic
```{r galileo-cubic-plot}
#| echo: false
galileo_cubic_plot
```
:::

Click through the tabs: a line is added, and a row joins the legend, each
time. Even by eye, the cubic barely bends past the quadratic — the question is
whether that bend is worth the extra parameter.

Now fit the three models explicitly and compare them with a nested F-test:

```{r galileo-models}
m1 <- lm(Distance ~ Height,          data = Sleuth3::case1001)
m2 <- lm(Distance ~ poly(Height, 2), data = Sleuth3::case1001)
m3 <- lm(Distance ~ poly(Height, 3), data = Sleuth3::case1001)

galileo_anova <- anova(m1, m2, m3)
galileo_anova
```

Because each model is nested in the next, `anova()` reports a *sequential*
F-test on each row, comparing that model to the one above it. With three
models compared at once, every row's F-test uses the residual degrees of
freedom of the **largest** model, `r df.residual(m3)`, in the denominator —
not each row's own `Res.Df`. Row 2 tests $H_0: \beta_2 = 0$ and gives
$F = `r round(galileo_anova$F[2], 2)`$ on $`r galileo_anova$Df[2]`$ and
$`r df.residual(m3)`$ degrees of freedom — the quadratic term is clearly
needed. Row 3 tests $H_0: \beta_3 = 0$ and gives
$F = `r round(galileo_anova$F[3], 2)`$, so the cubic term improves the fit as
well. The residual sum of squares falls from
$`r round(galileo_anova$RSS[1])`$ to $`r round(galileo_anova$RSS[2])`$ to
$`r round(galileo_anova$RSS[3])`$.

Be careful about what that second F-test has established. There are
$n = `r nrow(Sleuth3::case1001)`$ observations, and the cubic model spends
$`r length(coef(m3))`$ parameters on them, leaving
$`r df.residual(m3)`$ degrees of freedom — the cubic curve nearly interpolates
the data. A significant
F-test says the cubic fits the *training* data better than the quadratic. It
does not say the cubic will predict a new release height better. Physics in a
vacuum argues for a quadratic, but physics with air resistance can make a
cubic a reasonable approximation too — statistical significance alone should
not settle which curve to extrapolate with. We return to this distinction next
lecture.


## Interactions

Let the effect of one feature depend on another:

$$Y_i = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \beta_3 X_{i1}X_{i2} + \epsilon_i.$$

Collect the terms involving $X_{i1}$:

$$Y_i = \underbrace{(\beta_0 + \beta_2 X_{i2})}_{\text{intercept on } X_1} + \underbrace{(\beta_1 + \beta_3 X_{i2})}_{\text{slope on } X_1} X_{i1} + \epsilon_i.$$

The slope on $X_1$ is a *function of the other feature*. This is exactly where
"holding all else constant" breaks: you cannot move $X_1$ while holding
$X_1X_2$ fixed. And $\beta_1$ is now the slope on $X_1$ **when $X_2 = 0$**,
which may be a value $X_2$ never takes — centering usually makes the main
effects interpretable again.

```{r interactions}
#| eval: false
lm(y ~ x1 * x2)   # expands to x1 + x2 + x1:x2
```

Hierarchy applies here too: keep $X_1$ and $X_2$ whenever $X_1X_2$ is present,
whatever their individual p-values say.

::: {.callout-note collapse="true"}
### Interaction with a categorical feature
With $D = 1$ for group B and $0$ for group A,

$$Y_i = \beta_0 + \beta_1 X_i + \beta_2 D_i + \beta_3 X_i D_i + \epsilon_i$$

fits group A with intercept $\beta_0$ and slope $\beta_1$, and group B with
intercept $\beta_0 + \beta_2$ and slope $\beta_1 + \beta_3$. Without the
interaction, the groups are forced to share a slope (parallel lines). With it,
$H_0: \beta_3 = 0$ tests whether the slopes differ.
:::

### Example: meadowfoam, where the interaction is not needed

Twenty-four meadowfoam plots were grown at six light intensities, with the light
begun either early or late. Does the timing change the *slope* on intensity, or
only the intercept?

Before fitting anything, look at the relationship itself:

::: {.callout-note collapse="true"}
### Code for the plots below
```{r meadowfoam-plot-setup}
#| output: false
meadowfoam <- Sleuth3::case0901 |>
  mutate(Time = factor(Time, levels = 1:2, labels = c("Late", "Early")))

meadowfoam_single      <- lm(Flowers ~ Intensity,        data = meadowfoam)
meadowfoam_additive    <- lm(Flowers ~ Intensity + Time, data = meadowfoam)
meadowfoam_interaction <- lm(Flowers ~ Intensity * Time, data = meadowfoam)

meadowfoam_grid <- expand_grid(
  Intensity = seq(min(meadowfoam$Intensity), max(meadowfoam$Intensity), length.out = 100),
  Time = levels(meadowfoam$Time)
) |>
  mutate(Time = factor(Time, levels = levels(meadowfoam$Time)))

# A shared base layer, so the four views below differ only in which line(s)
# are layered on top of it -- the panel itself never changes size or shape.
meadowfoam_base <- ggplot(meadowfoam, aes(x = Intensity, y = Flowers)) +
  geom_point(aes(color = Time, shape = Time)) +
  coord_cartesian(xlim = range(meadowfoam$Intensity),
                   ylim = range(meadowfoam$Flowers)) +
  labs(title = "Sleuth3::case0901") +
  theme(legend.position = "inside",
        legend.position.inside = c(0.98, 0.98),
        legend.justification = c(1, 1),
        legend.background = element_rect(fill = alpha("white", 0.8), color = "grey70"))

meadowfoam_data_plot <- meadowfoam_base

meadowfoam_single_plot <- meadowfoam_base +
  geom_line(
    data = meadowfoam_grid |> mutate(Flowers = predict(meadowfoam_single, meadowfoam_grid)),
    linewidth = 1
  )

meadowfoam_additive_plot <- meadowfoam_base +
  geom_line(
    data = meadowfoam_grid |> mutate(Flowers = predict(meadowfoam_additive, meadowfoam_grid)),
    aes(color = Time, linetype = Time)
  )

meadowfoam_interaction_plot <- meadowfoam_base +
  geom_line(
    data = meadowfoam_grid |> mutate(Flowers = predict(meadowfoam_interaction, meadowfoam_grid)),
    aes(color = Time, linetype = Time)
  )
```
:::

::: {.panel-tabset}
#### Data
```{r meadowfoam-data-plot}
#| echo: false
meadowfoam_data_plot
```

#### + Single line
```{r meadowfoam-single-plot}
#| echo: false
meadowfoam_single_plot
```

#### + Additive
```{r meadowfoam-additive-plot}
#| echo: false
meadowfoam_additive_plot
```

#### + Interaction
```{r meadowfoam-interaction-plot}
#| echo: false
meadowfoam_interaction_plot
```
:::

A single line, ignoring timing, is pulled between the two groups and fits
neither well. The additive model gives each timing its own intercept but
forces a common slope. The interaction model frees the slopes entirely — click
through and notice how little that changes the fit here: the two slopes were
already nearly identical without it.

Now fit the interaction model explicitly and test whether it earns its extra
parameter:

```{r meadowfoam-anova}
meadowfoam_anova <- anova(meadowfoam_interaction)
meadowfoam_anova
```

The interaction gives
$F = `r round(meadowfoam_anova["Intensity:Time", "F value"], 3)`$ on
$`r meadowfoam_anova["Intensity:Time", "Df"]`$ and
$`r meadowfoam_anova["Residuals", "Df"]`$ degrees of freedom, so
$p = `r round(meadowfoam_anova["Intensity:Time", "Pr(>F)"], 2)`$. The two
slopes are about as close as noisy data allow, and the test agrees: no
evidence that the slope on intensity depends on timing. Drop the interaction
and report the additive model, in which timing shifts the intercept by a
constant amount and "the effect of intensity" is a single number again.

Dropping the *interaction* is fine. Dropping a main effect would not be.

### Example: alcohol metabolism, where the interaction matters

Thirty-two subjects had their first-pass metabolism of alcohol measured along
with the activity of gastric alcohol dehydrogenase, the enzyme that breaks
alcohol down in the stomach.

Before fitting anything, look at the relationship itself:

::: {.callout-note collapse="true"}
### Code for the plots below
```{r metabolism-plot-setup}
#| output: false
noteworthy <- Sleuth3::case1101 |> filter(Subject %in% c(17, 31, 32))

metabolism_single      <- lm(Metabol ~ Gastric,        data = Sleuth3::case1101)
metabolism_additive    <- lm(Metabol ~ Gastric + Sex,  data = Sleuth3::case1101)
metabolism_interaction <- lm(Metabol ~ Gastric * Sex,  data = Sleuth3::case1101)

metabolism_grid <- expand_grid(
  Gastric = seq(min(Sleuth3::case1101$Gastric), max(Sleuth3::case1101$Gastric), length.out = 100),
  Sex = levels(Sleuth3::case1101$Sex)
) |>
  mutate(Sex = factor(Sex, levels = levels(Sleuth3::case1101$Sex)))

# A shared base layer, so the four views below differ only in which line(s)
# are layered on top of it -- the panel itself never changes size or shape.
metabolism_base <- ggplot(Sleuth3::case1101, aes(x = Gastric, y = Metabol)) +
  geom_point(aes(color = Sex, shape = Sex)) +
  geom_text(data = noteworthy, aes(label = Subject),
            hjust = -0.6, size = 3, show.legend = FALSE) +
  coord_cartesian(xlim = c(min(Sleuth3::case1101$Gastric), 5.6),
                   ylim = range(Sleuth3::case1101$Metabol)) +
  labs(title = "Sleuth3::case1101",
       x = "Gastric alcohol dehydrogenase activity",
       y = "First-pass metabolism") +
  theme(legend.position = "inside",
        legend.position.inside = c(0.02, 0.98),
        legend.justification = c(0, 1),
        legend.background = element_rect(fill = alpha("white", 0.8), color = "grey70"))

metabolism_data_plot <- metabolism_base

metabolism_single_plot <- metabolism_base +
  geom_line(
    data = metabolism_grid |> mutate(Metabol = predict(metabolism_single, metabolism_grid)),
    linewidth = 1
  )

metabolism_additive_plot <- metabolism_base +
  geom_line(
    data = metabolism_grid |> mutate(Metabol = predict(metabolism_additive, metabolism_grid)),
    aes(color = Sex, linetype = Sex)
  )

metabolism_interaction_plot <- metabolism_base +
  geom_line(
    data = metabolism_grid |> mutate(Metabol = predict(metabolism_interaction, metabolism_grid)),
    aes(color = Sex, linetype = Sex)
  )
```
:::

::: {.panel-tabset}
#### Data
```{r metabolism-data-plot}
#| echo: false
metabolism_data_plot
```

#### + Single line
```{r metabolism-single-plot}
#| echo: false
metabolism_single_plot
```

#### + Additive
```{r metabolism-additive-plot}
#| echo: false
metabolism_additive_plot
```

#### + Interaction
```{r metabolism-interaction-plot}
#| echo: false
metabolism_interaction_plot
```
:::

A single line, ignoring sex, cuts across both groups. The additive model gives
each sex its own intercept but forces a common slope. The interaction model
frees the slopes — and here, unlike meadowfoam, that visibly matters: the
lines are far from parallel.

Now fit the interaction model explicitly and test whether it earns its extra
parameter:

```{r metabolism-fit}
metabolism_anova <- anova(metabolism_interaction)
metabolism_anova
coef(metabolism_interaction)
```

Here the interaction gives
$F = `r round(metabolism_anova["Gastric:Sex", "F value"], 2)`$ on
$`r metabolism_anova["Gastric:Sex", "Df"]`$ and
$`r metabolism_anova["Residuals", "Df"]`$ degrees of freedom, so
$p = `r round(metabolism_anova["Gastric:Sex", "Pr(>F)"], 3)`$, and the lines
are visibly non-parallel. The fitted slope is
$`r round(coef(metabolism_interaction)["Gastric"], 2)`$ for females and
$`r round(coef(metabolism_interaction)["Gastric"], 2)` + `r round(coef(metabolism_interaction)["Gastric:SexMale"], 2)` = `r round(coef(metabolism_interaction)["Gastric"] + coef(metabolism_interaction)["Gastric:SexMale"], 2)`$
for males — nearly three times steeper.

In this model there is no such thing as "the effect of gastric activity." There
are two of them, and you have to say which sex you mean.

::: {.callout-note collapse="true"}
### A preview of a problem
Three observations are labeled by subject number in each plot above. Subjects 31 and
32 have gastric activity well beyond any other man's, and subject 17 is the most
extreme of the women. These are high-leverage points: they sit far from the bulk
of the feature values, so they do a disproportionate share of the work in
estimating the two slopes.

Notice that they are not all the same kind of unusual. Subject 31 sits close to
the male line, while subject 32 is far above it. Next lecture we return to these
three by number and separate leverage — an unusual *feature* value — from
outliers, an unusual *response*.
:::

## What we have not fixed

Polynomials and interactions both stay inside the linear model: we pick the
basis functions $h_k$ by hand and least squares does the rest. Next lecture we
let the data take over more of that choice — step functions and K-nearest
neighbors — and then take stock of what can go wrong.
