# Feature Engineering

```{r packages}
#| include: false
library("tidyverse")
theme_set(theme_bw())
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **explain** why a model built from polynomials and interactions is still a
*linear* model,
- **construct** these features and **write** the corresponding `lm()` formula,
- **interpret** a coefficient in each case, and explain why "a one-unit increase,
holding all else constant" can stop making sense, and
- **use** an F-test to decide how much curvature, or whether an interaction, the
data actually support.
:::

## Linear in the parameters

Multiple linear regression looks restrictive: a straight line in each feature.
It is not. Consider

$$Y_i = \beta_0 + \beta_1 h_1(X_i) + \cdots + \beta_p h_K(X_i) + \epsilon_i$$

where $h_1, \ldots, h_K$ are known, fixed functions of the features $X_i$,
called *basis functions*.

This model is linear in $\beta$, which is all least squares ever required.
Every result from last lecture therefore carries over **unchanged**:
$\hat\beta = (X^\top X)^{-1}X^\top y$, the t-tests, the F-tests, the intervals.
Only the columns of the design matrix change.

Choosing the $h_k$ is **feature engineering**, and it is where most of the
modeling judgment in a regression analysis lives.


## Polynomials

Use the powers of a single feature as the basis:

$$Y_i = \beta_0 + \beta_1 X_i + \beta_2 X_i^2 + \cdots + \beta_d X_i^d + \epsilon_i.$$

The fit is now a curve, and no single coefficient is "the effect of $X$." That
effect is the derivative

$$\frac{\partial}{\partial x} E[Y|X = x] = \beta_1 + 2\beta_2 x + \cdots + d\beta_d x^{d-1},$$

which **depends on where you are**. Reporting $\hat\beta_1$ by itself is
meaningless.

```{r polynomials}
#| eval: false
lm(y ~ poly(x, 3))               # orthogonal polynomial basis
lm(y ~ x + I(x^2) + I(x^3))      # raw powers
lm(y ~ poly(x, 3, raw = TRUE))   # also raw powers
```

Two cautions. 

- **Hierarchy**: if $X^2$ is in the model, keep $X$. 
- **Extrapolation**: high-degree polynomials behave badly outside the range of the data.

::: {.callout-note collapse="true"}
### Why `poly()` instead of raw powers?
Over a range like $x \in [0,10]$, the columns $x$, $x^2$, and $x^3$ are
enormously correlated — they all increase together. That is collinearity, and it
inflates standard errors and destabilizes $\hat\beta$ numerically.

`poly()` builds an *orthogonal* basis spanning the same space, so $\hat{y}$,
the RSS, and $R^2$ are **identical**. What changes is that the coefficients are
now uncorrelated, so their t-tests can be read one at a time — a convenient way
to ask "is the cubic term needed?"
:::




### Example: Galileo's falling bodies

Galileo released a ball from a ramp at seven different heights and measured how
far it travelled horizontally. Physics says that relationship should be curved,
so this is a natural place to ask *how much* curvature the data support.

```{r galileo-models}
m1 <- lm(Distance ~ Height,          data = Sleuth3::case1001)
m2 <- lm(Distance ~ poly(Height, 2), data = Sleuth3::case1001)
m3 <- lm(Distance ~ poly(Height, 3), data = Sleuth3::case1001)

anova(m1, m2, m3)
```

Because each model is nested in the next, `anova()` reports a *sequential*
F-test on each row, comparing that model to the one above it. Row 2 tests
$H_0: \beta_2 = 0$ and gives $F = 306.33$ on $1$ and $4$ degrees of freedom — the
quadratic term is clearly needed. Row 3 tests $H_0: \beta_3 = 0$ and gives
$F = 43.26$, so the cubic term improves the fit as well. The residual sum of
squares falls from $5671$ to $744$ to $48$.


::: {.callout-note collapse="true"}
### Code for the figure below
```{r galileo-plot}
galileo_grid <- tibble(Height = seq(100, 1000, length.out = 200))

galileo_pred <- bind_rows(
  galileo_grid |> mutate(Distance = predict(m1, galileo_grid), Model = "linear"),
  galileo_grid |> mutate(Distance = predict(m2, galileo_grid), Model = "quadratic"),
  galileo_grid |> mutate(Distance = predict(m3, galileo_grid), Model = "cubic")
) |>
  mutate(Model = factor(Model, levels = c("linear", "quadratic", "cubic")))

galileo_plot <- ggplot(Sleuth3::case1001, aes(x = Height, y = Distance)) +
  geom_point() +
  geom_line(data = galileo_pred, aes(color = Model, linetype = Model)) +
  labs(title = "Sleuth3::case1001",
       x = "Release height", y = "Horizontal distance")
```
:::

```{r galileo-figure}
#| echo: false
galileo_plot
```

Be careful about what that second F-test has established. There are $n = 7$
observations, and the cubic model spends $4$ parameters on them, leaving $3$
degrees of freedom — the cubic curve nearly interpolates the data. A significant
F-test says the cubic fits the *training* data better than the quadratic. It
does not say the cubic will predict a new release height better, and the physics
argues for a quadratic. We return to this distinction next lecture.


## Interactions

Let the effect of one feature depend on another:

$$Y_i = \beta_0 + \beta_1 X_{i1} + \beta_2 X_{i2} + \beta_3 X_{i1}X_{i2} + \epsilon_i.$$

Collect the terms involving $X_{i1}$:

$$Y_i = \underbrace{(\beta_0 + \beta_2 X_{i2})}_{\text{intercept on } X_1} + \underbrace{(\beta_1 + \beta_3 X_{i2})}_{\text{slope on } X_1} X_{i1} + \epsilon_i.$$

The slope on $X_1$ is a *function of the other feature*. This is exactly where
"holding all else constant" breaks: you cannot move $X_1$ while holding
$X_1X_2$ fixed. And $\beta_1$ is now the slope on $X_1$ **when $X_2 = 0$**,
which may be a value $X_2$ never takes — centering usually makes the main
effects interpretable again.

```{r interactions}
#| eval: false
lm(y ~ x1 * x2)   # expands to x1 + x2 + x1:x2
```

Hierarchy applies here too: keep $X_1$ and $X_2$ whenever $X_1X_2$ is present,
whatever their individual p-values say.

::: {.callout-note collapse="true"}
### Interaction with a categorical feature
With $D = 1$ for group B and $0$ for group A,

$$Y_i = \beta_0 + \beta_1 X_i + \beta_2 D_i + \beta_3 X_i D_i + \epsilon_i$$

fits group A with intercept $\beta_0$ and slope $\beta_1$, and group B with
intercept $\beta_0 + \beta_2$ and slope $\beta_1 + \beta_3$. Without the
interaction, the groups are forced to share a slope (parallel lines). With it,
$H_0: \beta_3 = 0$ tests whether the slopes differ.
:::

### Example: meadowfoam, where the interaction is not needed

Twenty-four meadowfoam plots were grown at six light intensities, with the light
begun either early or late. Does the timing change the *slope* on intensity, or
only the intercept?

```{r meadowfoam-data}
meadowfoam <- Sleuth3::case0901 |>
  mutate(Time = factor(Time, levels = 1:2, labels = c("Late", "Early")))
```

::: {.callout-note collapse="true"}
### Code for the figure below
```{r meadowfoam-plot}
#| output: false
meadowfoam_plot <- ggplot(meadowfoam,
                          aes(x = Intensity, y = Flowers,
                              color = Time, shape = Time, linetype = Time)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x) +
  labs(title = "Sleuth3::case0901")
```
:::

```{r meadowfoam-figure}
#| echo: false
meadowfoam_plot
```

```{r meadowfoam-anova}
anova(lm(Flowers ~ Intensity * Time, data = meadowfoam))
```

The interaction gives $F = 0.013$ on $1$ and $20$ degrees of freedom, so
$p = 0.91$. The separately fitted lines are about as close to parallel as noisy
data allow, and the test agrees: no evidence that the slope on intensity depends
on timing. Drop the interaction and report the additive model, in which timing
shifts the intercept by a constant amount and "the effect of intensity" is a
single number again.

Dropping the *interaction* is fine. Dropping a main effect would not be.

### Example: alcohol metabolism, where the interaction matters

Thirty-two subjects had their first-pass metabolism of alcohol measured along
with the activity of gastric alcohol dehydrogenase, the enzyme that breaks
alcohol down in the stomach.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r metabolism-plot}
#| output: false
metabolism_plot <- ggplot(Sleuth3::case1101,
                          aes(x = Gastric, y = Metabol,
                              color = Sex, shape = Sex, linetype = Sex)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, formula = y ~ x) +
  labs(title = "Sleuth3::case1101",
       x = "Gastric alcohol dehydrogenase activity",
       y = "First-pass metabolism")
```
:::

```{r metabolism-figure}
#| echo: false
metabolism_plot
```

```{r metabolism-fit}
metabolism <- lm(Metabol ~ Gastric * Sex, data = Sleuth3::case1101)

anova(metabolism)
coef(metabolism)
```

Here the interaction gives $F = 7.26$ on $1$ and $28$ degrees of freedom, so
$p = 0.012$, and the lines are visibly non-parallel. The fitted slope is $0.84$
for females and $0.84 + 1.51 = 2.34$ for males — nearly three times steeper.

In this model there is no such thing as "the effect of gastric activity." There
are two of them, and you have to say which sex you mean.

::: {.callout-note collapse="true"}
### A preview of a problem
The two male subjects at the far right have gastric activity well beyond anyone
else's. Those are high-leverage points, and they are doing a large share of the
work in estimating the male slope. Leverage appears in the list of potential
problems next lecture; this is what it looks like in practice.
:::

## What we have not fixed

Polynomials and interactions both stay inside the linear model: we pick the
basis functions $h_k$ by hand and least squares does the rest. Next lecture we
let the data take over more of that choice — step functions and K-nearest
neighbors — and then take stock of what can go wrong.
