# Flexibility and Its Costs

```{r packages}
#| include: false
library("tidyverse")
theme_set(theme_bw())
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **construct** a step function model and **explain** why it is the
dummy-variable machinery applied to a quantitative feature,
- **describe** K-nearest neighbors regression and **identify** $K$ as its
flexibility parameter,
- **compare** a parametric fit against a nonparametric one and **state** the
conditions under which each is preferable,
- **diagnose** the potential problems of ISLR2 Section 3.3.3, and
- **evaluate** added flexibility in terms of the bias-variance tradeoff.
:::

## Step functions

Last lecture we used polynomials, which impose one global shape on the whole
range of $X$. A step function gives up smoothness instead. Choose cut points
$c_1 < \cdots < c_K$, define indicators
$C_k(X) = I(c_k \le X < c_{k+1})$, and fit

$$Y_i = \beta_0 + \beta_1 C_1(X_i) + \cdots + \beta_K C_K(X_i) + \epsilon_i$$

dropping the first bin as the baseline. This is **exactly** the dummy-variable
machinery from multiple regression: we have turned a quantitative feature into a
categorical one. The fitted value in a bin is the sample mean of $y$ in that
bin, and $\beta_k$ is that bin's mean minus the baseline bin's mean.

```{r step-functions}
#| eval: false
lm(y ~ cut(x, breaks = c(0, 2, 4, 6, 8, 10)))
```

::: {.callout-note collapse="true"}
### Code for the figure below
```{r flexibility-comparison}
#| output: false
set.seed(20260916)

sim <- tibble(x = runif(100, 0, 10)) |>
  mutate(y = 2 + 0.6 * x + 1.8 * sin(x / 1.4) + rnorm(100, 0, 0.9))

cuts     <- seq(0, 10, length.out = 6)
sim_grid <- tibble(x = seq(0, 10, length.out = 400))

fit_grid <- function(model, label, binned = FALSE) {
  sim_grid |>
    mutate(yhat = predict(model, sim_grid),
           fit  = label,
           bin  = if (binned) as.character(cut(x, cuts, include.lowest = TRUE))
                  else "all")
}

sim_pred <- bind_rows(
  fit_grid(lm(y ~ x,                                   data = sim), "Linear in x"),
  fit_grid(lm(y ~ poly(x, 5),                          data = sim), "Polynomial (degree 5)"),
  fit_grid(lm(y ~ cut(x, cuts, include.lowest = TRUE), data = sim), "Step function (5 bins)",
           binned = TRUE)
)

flexibility_plot <- ggplot(sim, aes(x = x, y = y)) +
  geom_point(alpha = 0.35) +
  geom_line(data = sim_pred, aes(y = yhat, group = interaction(fit, bin)),
            color = "blue", linewidth = 1) +
  facet_wrap(~ fit)
```
:::

```{r flexibility-comparison-figure}
#| echo: false
#| fig-width: 9
#| fig-height: 3.2
flexibility_plot
```

The costs are visible on the right: the fit jumps at the cut points and ignores
any trend within a bin. And *we* chose those cut points — nothing in the data
told us where to put them.

::: {.callout-important collapse="true"}
### Where this is going
That last complaint is the whole idea behind **regression trees** (Chapter 8).
A tree is a step function whose cut points are *learned from the data*, applied
recursively and in several features at once — so a tree gets interactions for
free, without our specifying them. A **random forest** averages many such step
functions to cut their variance.
:::

## K-nearest neighbors regression

A step function lets the data choose the *value* taken in each bin, but we still
chose the bins. K-nearest neighbors abandons fixed bins altogether and lets the
neighborhood move with the point being predicted.

Fix a value $K$ and a target point $x_0$, and let $\mathcal{N}_0$ be the set of
the $K$ training observations whose feature values are closest to $x_0$. Then

$$\hat{f}(x_0) = \frac{1}{K} \sum_{i \in \mathcal{N}_0} y_i.$$

That is the whole method: average the responses of the nearest neighbors. There
is no $\beta$, nothing is fit by least squares, and no functional form is
assumed anywhere. KNN is a **nonparametric** method, unlike every model we have
seen so far.

```{r knn-function}
knn_reg <- function(x0, x, y, K) {
  sapply(x0, function(z) mean(y[order(abs(x - z))[1:K]]))
}
```

$K$ is the flexibility dial, and it runs backwards:

- $K = 1$ makes $\hat{f}$ interpolate the training data — zero training error
and enormous variance.
- $K = n$ makes $\hat{f}$ the overall mean $\bar{y}$, a flat line — maximum bias
and minimum variance.

So it is $1/K$ that plays the role of flexibility.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r knn-comparison}
#| output: false
knn_pred <- map_df(c(1, 10, 50), function(K) {
  sim_grid |>
    mutate(yhat = knn_reg(x, sim$x, sim$y, K),
           K    = paste("K =", K))
}) |>
  mutate(K = factor(K, levels = paste("K =", c(1, 10, 50))))

knn_plot <- ggplot(sim, aes(x = x, y = y)) +
  geom_point(alpha = 0.35) +
  geom_line(data = knn_pred, aes(y = yhat), color = "blue", linewidth = 1) +
  facet_wrap(~ K)
```
:::

```{r knn-comparison-figure}
#| echo: false
#| fig-width: 9
#| fig-height: 3.2
knn_plot
```

### When does KNN beat linear regression?

When the true $f$ is far from linear, KNN wins, and it does so without our
having to guess the right polynomial degree or the right cut points.

When $f$ really is close to linear, least squares wins, because it estimates
that line using every observation at once rather than only $K$ of them.

The decisive issue is the number of features. As $p$ grows, the $K$ "nearest"
neighbors of $x_0$ stop being anywhere near it — in high dimensions everything
is far from everything — so averaging their responses estimates $f$ somewhere
other than at $x_0$, which is bias. This is the **curse of dimensionality**, and
it is severe: ISLR2 shows KNN losing to linear regression at $p = 20$ and
$n = 100$ *even when the true relationship is not linear*.

A parametric model that is merely approximately right will usually beat a
nonparametric one when $p$ is large relative to $n$.

## Potential problems

Last lecture's feature engineering is the *cure* for the first of these and the
*cause* of the last two.

1. **Non-linearity.** Structure in a plot of residuals against fitted values
means the model is missing systematic signal — the motivation for polynomials,
step functions, and KNN alike.
2. **Correlated errors.** Common with data over time or space. Standard errors
come out too small, so you are more confident than you have earned.
3. **Non-constant variance.** A funnel in the residual plot; often handled by
transforming the response, $\log Y$ or $\sqrt{Y}$.
4. **Outliers.** An unusual *response*. Inflates $\hat\sigma^2$ and depresses
$R^2$ without necessarily moving the line.
5. **High-leverage points.** An unusual *feature* value. These do move the line,
and polynomials make it worse — an extreme $x$ raised to the fifth power is an
extraordinarily influential row.
6. **Collinearity.** Features carrying nearly the same information. The data
cannot separate their effects, so $\text{SE}(\hat\beta_j)$ inflates and the
t-test loses power: a coefficient can matter and still look "insignificant."
Polynomials and interactions create collinearity by construction.

::: {.callout-note collapse="true"}
### Detecting leverage and collinearity
**Leverage** is $h_i$, the $i$th diagonal element of the hat matrix
$H = X(X^\top X)^{-1}X^\top$. It averages $(p+1)/n$, so values several times
that deserve a look. The dangerous points have high leverage *and* a large
studentized residual, say beyond $\pm 3$.

**Collinearity** is measured by the variance inflation factor

$$\text{VIF}(\hat\beta_j) = \frac{1}{1 - R^2_{X_j | X_{-j}}}$$

where $R^2_{X_j|X_{-j}}$ comes from regressing $X_j$ on all other features. VIF
is 1 under no collinearity; past 5 or 10 is commonly flagged. Remedies: drop one
feature, combine them, center them, or — for polynomials — use `poly()`.
:::

## The cost of flexibility

Every basis function you add, and every reduction in $K$, is another turn of the
flexibility dial from Chapter 2. The training RSS can only fall as you turn it;
the **test** MSE need not, because you are trading bias for variance.

So these methods hand us more candidate models than we can judge by eye, and
training error cannot referee among them. Honest test-error estimates come from
resampling (Chapter 5), and automatic ways to choose which features to keep come
from selection and regularization (Chapter 6).
