# Flexibility and Its Costs

```{r packages}
#| include: false
library("tidyverse")
theme_set(theme_bw())
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **construct** a step function model and **explain** why it is the
dummy-variable machinery applied to a quantitative feature,
- **describe** K-nearest neighbors regression and **identify** $K$ as its
flexibility parameter,
- **compare** a parametric fit against a nonparametric one and **state** the
conditions under which each is preferable,
- **diagnose** the potential problems of ISLR2 Section 3.3.3 from a residual
plot, leverage values, and variance inflation factors,
- **distinguish** a high-leverage point from an outlier, and **use** Cook's
distance to judge which observations actually influence the fit, and
- **evaluate** added flexibility in terms of the bias-variance tradeoff.
:::

## Step functions

Last lecture we used polynomials, which impose one global shape on the whole
range of $X$. A step function gives up smoothness instead. Choose cut points
$c_1 < \cdots < c_K$, define indicators
$C_k(X) = I(c_k \le X < c_{k+1})$, and fit

$$Y_i = \beta_0 + \beta_1 C_1(X_i) + \cdots + \beta_K C_K(X_i) + \epsilon_i$$

dropping the first bin as the baseline. This is **exactly** the dummy-variable
machinery from multiple regression: we have turned a quantitative feature into a
categorical one. The fitted value in a bin is the sample mean of $y$ in that
bin, and $\beta_k$ is that bin's mean minus the baseline bin's mean.

```{r step-functions}
#| eval: false
lm(y ~ cut(x, breaks = c(0, 2, 4, 6, 8, 10)))
```

::: {.callout-note collapse="true"}
### Code for the figure below
```{r flexibility-comparison}
#| output: false
set.seed(20260916)

sim <- tibble(x = runif(100, 0, 10)) |>
  mutate(y = 2 + 0.6 * x + 1.8 * sin(x / 1.4) + rnorm(100, 0, 0.9))

cuts     <- seq(0, 10, length.out = 6)
sim_grid <- tibble(x = seq(0, 10, length.out = 400))

fit_grid <- function(model, label, binned = FALSE) {
  sim_grid |>
    mutate(yhat = predict(model, sim_grid),
           fit  = label,
           bin  = if (binned) as.character(cut(x, cuts, include.lowest = TRUE))
                  else "all")
}

sim_pred <- bind_rows(
  fit_grid(lm(y ~ x,                                   data = sim), "Linear in x"),
  fit_grid(lm(y ~ poly(x, 5),                          data = sim), "Polynomial (degree 5)"),
  fit_grid(lm(y ~ cut(x, cuts, include.lowest = TRUE), data = sim), "Step function (5 bins)",
           binned = TRUE)
)

flexibility_plot <- ggplot(sim, aes(x = x, y = y)) +
  geom_point(alpha = 0.35) +
  geom_line(data = sim_pred, aes(y = yhat, group = interaction(fit, bin)),
            color = "blue", linewidth = 1) +
  facet_wrap(~ fit)
```
:::

```{r flexibility-comparison-figure}
#| echo: false
#| fig-width: 9
#| fig-height: 3.2
flexibility_plot
```

The costs are visible on the right: the fit jumps at the cut points and ignores
any trend within a bin. And *we* chose those cut points — nothing in the data
told us where to put them.

::: {.callout-important collapse="true"}
### Where this is going
That last complaint is the whole idea behind **regression trees** (Chapter 8).
A tree is a step function whose cut points are *learned from the data*, applied
recursively and in several features at once — so a tree gets interactions for
free, without our specifying them. A **random forest** averages many such step
functions to cut their variance.
:::

## K-nearest neighbors regression

A step function lets the data choose the *value* taken in each bin, but we still
chose the bins. K-nearest neighbors abandons fixed bins altogether and lets the
neighborhood move with the point being predicted.

Fix a value $K$ and a target point $x_0$, and let $\mathcal{N}_0$ be the set of
the $K$ training observations whose feature values are closest to $x_0$. Then

$$\hat{f}(x_0) = \frac{1}{K} \sum_{i \in \mathcal{N}_0} y_i.$$

That is the whole method: average the responses of the nearest neighbors. There
is no $\beta$, nothing is fit by least squares, and no functional form is
assumed anywhere. KNN is a **nonparametric** method, unlike every model we have
seen so far.

```{r knn-function}
knn_reg <- function(x0, x, y, K) {
  sapply(x0, function(z) mean(y[order(abs(x - z))[1:K]]))
}
```

$K$ is the flexibility dial, and it runs backwards:

- $K = 1$ makes $\hat{f}$ interpolate the training data — zero training error
and enormous variance.
- $K = n$ makes $\hat{f}$ the overall mean $\bar{y}$, a flat line — maximum bias
and minimum variance.

So it is $1/K$ that plays the role of flexibility.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r knn-comparison}
#| output: false
knn_pred <- map_df(c(1, 10, 50), function(K) {
  sim_grid |>
    mutate(yhat = knn_reg(x, sim$x, sim$y, K),
           K    = paste("K =", K))
}) |>
  mutate(K = factor(K, levels = paste("K =", c(1, 10, 50))))

knn_plot <- ggplot(sim, aes(x = x, y = y)) +
  geom_point(alpha = 0.35) +
  geom_line(data = knn_pred, aes(y = yhat), color = "blue", linewidth = 1) +
  facet_wrap(~ K)
```
:::

```{r knn-comparison-figure}
#| echo: false
#| fig-width: 9
#| fig-height: 3.2
knn_plot
```

### When does KNN beat linear regression?

When the true $f$ is far from linear, KNN wins, and it does so without our
having to guess the right polynomial degree or the right cut points.

When $f$ really is close to linear, least squares wins, because it estimates
that line using every observation at once rather than only $K$ of them.

The decisive issue is the number of features. As $p$ grows, the $K$ "nearest"
neighbors of $x_0$ stop being anywhere near it — in high dimensions everything
is far from everything — so averaging their responses estimates $f$ somewhere
other than at $x_0$, which is bias. This is the **curse of dimensionality**, and
it is severe: ISLR2 shows KNN losing to linear regression at $p = 20$ and
$n = 100$ *even when the true relationship is not linear*.

A parametric model that is merely approximately right will usually beat a
nonparametric one when $p$ is large relative to $n$.

## Potential problems

Last lecture's feature engineering is the *cure* for the first of these and the
*cause* of the last two.

1. **Non-linearity.** Structure in a plot of residuals against fitted values
means the model is missing systematic signal — the motivation for polynomials,
step functions, and KNN alike.
2. **Correlated errors.** Common with data over time or space. Standard errors
come out too small, so you are more confident than you have earned.
3. **Non-constant variance.** A funnel in the residual plot; often handled by
transforming the response, $\log Y$ or $\sqrt{Y}$.
4. **Outliers.** An unusual *response*. Inflates $\hat\sigma^2$ and depresses
$R^2$ without necessarily moving the line.
5. **High-leverage points.** An unusual *feature* value. These do move the line,
and polynomials make it worse — an extreme $x$ raised to the fifth power is an
extraordinarily influential row.
6. **Collinearity.** Features carrying nearly the same information. The data
cannot separate their effects, so $\text{SE}(\hat\beta_j)$ inflates and the
t-test loses power: a coefficient can matter and still look "insignificant."
Polynomials and interactions create collinearity by construction.

::: {.callout-note collapse="true"}
### Detecting leverage and collinearity
**Leverage** is $h_i$, the $i$th diagonal element of the hat matrix
$H = X(X^\top X)^{-1}X^\top$. It averages $(p+1)/n$, so values several times
that deserve a look. Leverage depends only on the features — you can compute it
before seeing a single response.

**Cook's distance** combines leverage with the size of the residual into a
single number measuring how much the fitted values would move if an observation
were deleted. Values near or above $1$ warrant attention. In R, `hatvalues()`,
`rstudent()`, and `cooks.distance()` compute all three.

**Collinearity** is measured by the variance inflation factor

$$\text{VIF}(\hat\beta_j) = \frac{1}{1 - R^2_{X_j | X_{-j}}}$$

where $R^2_{X_j|X_{-j}}$ comes from regressing $X_j$ on all other features. VIF
is 1 under no collinearity; past 5 or 10 is commonly flagged. Remedies: drop one
feature, combine them, center them, or — for polynomials — use `poly()`.
:::

Every one of these except correlated errors shows up in data we have already
used this semester. Number 2 is the exception only because none of our examples
are collected over time; the CAPM demonstration, with weekly returns ordered by
date, is exactly where you would worry about it.

### Reading residuals: curvature and changing spread

A plot of residuals against fitted values diagnoses problems 1 and 3 at a
glance. Both of our regressions from last lecture have something to confess.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r residual-diagnostics}
#| output: false
galileo_lin <- lm(Distance ~ Height,          data = Sleuth3::case1001)
metabol_fit <- lm(Metabol  ~ Gastric * Sex,   data = Sleuth3::case1101)

residual_data <- bind_rows(
  tibble(fitted = fitted(galileo_lin), resid = residuals(galileo_lin),
         panel  = "Galileo, linear fit: curvature"),
  tibble(fitted = fitted(metabol_fit), resid = residuals(metabol_fit),
         panel  = "Metabolism: non-constant variance")
)

residual_plot <- ggplot(residual_data, aes(x = fitted, y = resid)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_point() +
  facet_wrap(~ panel, scales = "free") +
  labs(x = "Fitted value", y = "Residual")
```
:::

```{r residual-diagnostics-figure}
#| echo: false
#| fig-width: 9
#| fig-height: 3.3
residual_plot
```

On the left, the residuals from the *linear* fit to Galileo's data run negative,
positive, then negative again — the inverted U that says a straight line is the
wrong shape. That is problem 1, and it is precisely what the quadratic term
fixed last lecture.

On the right, the residuals from the alcohol metabolism model fan out: near zero
for small fitted values and spanning several units for large ones. That is
problem 3. The constant-variance assumption behind every standard error in that
output is not holding, and a $\log$ or $\sqrt{\cdot}$ transformation of the
response would be the usual response.

### Leverage and outliers are different things

Last lecture we noticed two men at the far right of the metabolism plot. Now we
can be precise about them. With $p + 1 = 4$ coefficients and $n = 32$
observations, average leverage is $4/32 = 0.125$.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r leverage-diagnostics}
#| output: false
average_leverage <- length(coef(metabol_fit)) / nrow(Sleuth3::case1101)

influence_data <- Sleuth3::case1101 |>
  mutate(leverage    = hatvalues(metabol_fit),
         studentized = rstudent(metabol_fit))

influence_plot <- ggplot(influence_data, aes(x = leverage, y = studentized)) +
  geom_hline(yintercept = c(-3, 3),            linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 2 * average_leverage, linetype = "dashed", color = "grey40") +
  geom_point() +
  geom_text(aes(label = ifelse(leverage > 2 * average_leverage | abs(studentized) > 3,
                               paste("Subject", Subject), "")),
            hjust = -0.15, size = 3) +
  expand_limits(x = 0.65) +
  labs(x = "Leverage", y = "Studentized residual")
```
:::

```{r leverage-diagnostics-figure}
#| echo: false
influence_plot
```

Three subjects sit outside the dashed guides, and they are outside for different
reasons. **Cook's distance** settles which of them actually matter.

::: {.callout-note collapse="true"}
### Code for the table below
```{r influence-table}
#| output: false
influence_table <- tibble(
  Subject     = Sleuth3::case1101$Subject,
  Studentized = rstudent(metabol_fit),
  Leverage    = hatvalues(metabol_fit),
  CooksD      = cooks.distance(metabol_fit)) |>
  filter(Subject %in% c(17, 31, 32)) |>
  knitr::kable(digits = 3,
               col.names = c("Subject", "Studentized residual", "Leverage", "Cook's D"))
```
:::

```{r influence-table-display}
#| echo: false
influence_table
```

Cook's distance answers a direct question: if we deleted this one observation
and refit, how much would all the fitted values move? A large value means the
model depends meaningfully on that single point.

It rises with both of the columns beside it. A point has to disagree with the
model *and* have room to pull the line toward itself before it can do real
damage — and having room is exactly what leverage measures. Either ingredient
alone is harmless; the combination is not.

The table makes the distinction concrete. Subject 17 has leverage $0.393$, more
than three times the average of $4/32 = 0.125$, and yet Cook's distance is
$0.006$. It sits far from the other feature values and agrees completely with
the trend, so deleting it would change essentially nothing. High leverage by
itself is not a problem.

Subjects 31 and 32 both land near or above $1$, the conventional cutoff, and
they get there by opposite routes. Subject 32 has only moderate leverage but a
studentized residual of $5.12$ — it disagrees violently. Subject 31 has the
largest leverage in the data with an unremarkable residual of $-1.91$ — it
barely disagrees, but it has so much room that a small disagreement is enough.

Deleting either one would visibly move the male line, which is worth knowing
before reporting that men metabolize alcohol at three times the female rate.

### Collinearity you create yourself

Last lecture's `poly()` callout claimed that raw powers are dangerously
correlated. Galileo's data lets us put a number on it.

```{r vif-example}
vif <- function(X) {
  sapply(seq_along(X), function(j)
    1 / (1 - summary(lm(X[[j]] ~ ., data = X[-j]))$r.squared))
}

Height     <- Sleuth3::case1001$Height
raw_powers <- data.frame(Height = Height, Height2 = Height^2, Height3 = Height^3)

round(vif(raw_powers), 1)
round(vif(as.data.frame(poly(Height, 3))), 3)
```

The raw powers carry variance inflation factors of roughly $172$, $960$, and
$356$ — the standard errors on those coefficients are inflated by factors of
$\sqrt{960} \approx 31$. The orthogonal basis from `poly()` gives exactly $1$,
$1$, and $1$.

Both parameterizations produce **identical** fitted values, residuals, and
$R^2$. Nothing about the quality of the fit changed. What changed is whether the
individual coefficients can be interpreted and tested one at a time — and this
is collinearity we manufactured ourselves by choosing a basis, not a defect in
Galileo's measurements.

### Why not just look at the data?

Every problem in this section could have been caught by eye. Seven observations
of Galileo's, thirty-two subjects, one or two features — plot the data and you
can see the curvature, see the two men stranded out on the right, see the spread
widening.

That is about to stop being true. From here on the course adds features faster
than you can plot them and models whose fitted surface cannot be drawn at all.
There is no scatterplot of a random forest on twenty features, and no way to
squint at observation 4,812 of 50,000 to decide whether it is dragging the fit.

What survives the transition is the numbers. Leverage, studentized residuals,
Cook's distance, and variance inflation factors are defined for any number of
features, and they can be sorted, thresholded, and scanned without a human
looking at anything. So can the diagnostic plots we just used: residuals against
fitted values, or leverage against studentized residual, stay two-dimensional no
matter how large $p$ gets, because they plot *model output* rather than the data
itself.

The habit worth building now, while you can still check the answer by looking,
is to compute the diagnostic anyway and learn what its values mean. Later the
number is all you will have.

## The cost of flexibility

Every basis function you add, and every reduction in $K$, is another turn of the
flexibility dial from Chapter 2. The training RSS can only fall as you turn it;
the **test** MSE need not, because you are trading bias for variance.

So these methods hand us more candidate models than we can judge by eye, and
training error cannot referee among them. Honest test-error estimates come from
resampling (Chapter 5), and automatic ways to choose which features to keep come
from selection and regularization (Chapter 6).
