# Cross-Validation

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
set.seed(20261008)

# class::knn() is called with :: rather than attached.
class_colors <- c(No = "#0072B2", Yes = "#D55E00")

# Fixed decimals with thousands separators, and percentages, for prose.
fmt <- function(x, digits = 2) {
  formatC(x, format = "f", digits = digits, big.mark = ",")
}
pct <- function(x, digits = 1) paste0(fmt(100 * x, digits), "%")
# A number in LaTeX scientific notation for use inside $...$.
sci <- function(x, digits = 1) {
  e <- floor(log10(abs(x)))
  sprintf("%s \\times 10^{%d}", formatC(x / 10^e, format = "f", digits = digits), e)
}
# "a", "a and b", "a, b, and c".
and_list <- function(x) {
  if (length(x) == 1) x else if (length(x) == 2) paste(x, collapse = " and ") else
    paste0(paste(x[-length(x)], collapse = ", "), ", and ", x[length(x)])
}
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **explain** why training error cannot choose a model's flexibility, and
**identify** the two weaknesses of the validation set approach: its estimate
varies from one random split to the next, and it overestimates the test error
of a model fit to all $n$ observations,
- **define** the leave-one-out and $k$-fold cross-validation estimates of test
MSE and test error rate, and **implement** each as a loop that fits on all but
one fold and predicts the held-out fold,
- **derive** the least squares shortcut
$\text{CV}_{(n)} = \frac{1}{n}\sum_{i=1}^n\left[\left(y_i - \hat y_i\right)/\left(1 - h_i\right)\right]^2$
and **interpret** the role of the leverage $h_i$ in it,
- **compare** the validation set approach, leave-one-out, and $k$-fold
cross-validation by the bias and variance of their test-error estimates and by
computation, and **justify** the usual choice $k = 5$ or $10$,
- **use** cross-validation to choose a flexibility setting, such as a
polynomial degree or the number of neighbors $K$ in KNN, and
- **identify** the data-dependent steps of a fit (feature screening,
standardization, tuning) that must be repeated inside every fold, and
**explain** why performing them once on all the data makes the
cross-validation estimate optimistic.
:::

## Test error estimation

### Horsepower and fuel efficiency

**Question.** How flexible should a regression of a car's fuel efficiency on
its engine horsepower be?

**Data.** `ISLR2::Auto` records `r nrow(Auto)` cars. The response is `mpg`,
miles per gallon, and the explanatory variable is `horsepower`. The figure
plots one against the other.

::: {.callout-note collapse="true"}
### Code
```{r auto-base}
#| output: false
auto_base <- ggplot(Auto, aes(x = horsepower, y = mpg)) +
  geom_point(alpha = 0.4, size = 1) +
  coord_cartesian(ylim = range(Auto$mpg)) +
  labs(x = "Horsepower", y = "Miles per gallon", title = "ISLR2::Auto")
```
:::

```{r auto-data-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4
auto_base
```

Fuel efficiency falls as horsepower rises, steeply at first and then more
slowly.

**Model.** To follow that curvature, the candidate models are polynomial
regressions of degree $d$,

$$\texttt{mpg}_i = \beta_0 + \beta_1\,\texttt{horsepower}_i + \beta_2\,\texttt{horsepower}_i^2 + \cdots + \beta_d\,\texttt{horsepower}_i^d + \epsilon_i, \qquad \epsilon_i \stackrel{ind}{\sim} N\left(0, \sigma^2\right),$$

where

- $\texttt{mpg}_i$ and $\texttt{horsepower}_i$ are the fuel efficiency and
horsepower of car $i = 1, \ldots, n$,
- $d$ is the **degree** of the polynomial, the flexibility setting to be
chosen,
- $\beta_0, \ldots, \beta_d$ are the coefficients and $\epsilon_i$ the error,
with variance $\sigma^2$, and
- the features are the $d$ powers of horsepower, so row $i$ of the model
matrix $\mathbf{X}$ is
$x_i = \left(1, \texttt{horsepower}_i, \ldots, \texttt{horsepower}_i^d\right)$.

As in the feature engineering lecture, `poly(horsepower, d)` fits the same
model with an orthogonal basis of the same column space, so its fitted values
are identical.

::: {.callout-note collapse="true"}
### Code
```{r auto-curves}
#| output: false
degrees <- 1:10
show_degrees <- c(1, 2, 10)
degree_labels <- paste("Degree", show_degrees)
degree_colors <- setNames(c("#0072B2", "#D55E00", "#009E73"), degree_labels)
degree_linetypes <- setNames(c("solid", "dashed", "dotdash"), degree_labels)

hp_grid <- tibble(horsepower = seq(min(Auto$horsepower), max(Auto$horsepower),
                                   length.out = 400))
degree_curves <- map(show_degrees, \(d) {
  fit <- lm(mpg ~ poly(horsepower, d), data = Auto)
  mutate(hp_grid, degree = paste("Degree", d), mpg = predict(fit, hp_grid))
}) |>
  list_rbind() |>
  mutate(degree = factor(degree, levels = degree_labels))

add_curves <- function(through) {
  auto_base +
    geom_line(data = filter(degree_curves, as.integer(degree) <= through),
              aes(color = degree, linetype = degree), linewidth = 1) +
    scale_color_manual(values = degree_colors) +
    scale_linetype_manual(values = degree_linetypes) +
    labs(color = "Polynomial", linetype = "Polynomial")
}
auto_curve_plots <- map(seq_along(show_degrees), add_curves)
```
:::

The tabs add least squares fits of degree `r and_list(show_degrees)` to the
plot above.

::: {.panel-tabset}
#### Degree 1
```{r auto-degree1-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4
auto_curve_plots[[1]]
```

#### + Degree 2
```{r auto-degree2-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4
auto_curve_plots[[2]]
```

#### + Degree 10
```{r auto-degree10-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4
auto_curve_plots[[3]]
```
:::

The straight line misses the curvature: it underpredicts at both ends
and overpredicts in the middle. The quadratic follows the curve. The degree-10
polynomial follows it too, with extra wiggles at both ends of the range of
horsepower, where cars are few.

### Training MSE

The function below fits a degree-$d$ polynomial to a data set; applied to all
`r nrow(Auto)` cars it gives the training MSE of each degree.

```{r fit-degree}
fit_degree <- function(d, data) lm(mpg ~ poly(horsepower, d), data = data)
train_mse <- map_dbl(degrees, \(d) mean(residuals(fit_degree(d, Auto))^2))
```

::: {.callout-note collapse="true"}
### Code
```{r train-mse-plot}
#| output: false
train_mse_plot <- ggplot(tibble(degree = degrees, mse = train_mse),
                         aes(x = degree, y = mse)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = degrees) +
  labs(x = "Polynomial degree d", y = "Training MSE", title = "ISLR2::Auto")
```
:::

```{r train-mse-figure}
#| echo: false
#| fig-width: 6
#| fig-height: 3.5
train_mse_plot
```

The training MSE falls from `r fmt(train_mse[1])` at degree 1 to
`r fmt(train_mse[2])` at degree 2 and keeps falling, to
`r fmt(train_mse[10])` at degree 10. It cannot do otherwise: the degree-$d$
model is the degree-$(d+1)$ model with $\beta_{d+1} = 0$, so least squares
over the larger model can only match or lower the residual sum of squares.
Training error rewards flexibility whether or not the flexibility captures
anything that will appear in new data, so it cannot choose $d$.

### Held-out data

The regression lecture defined the test MSE as the average squared error on
observations that were not used to estimate $\hat f$. With a large, separate
test set, choosing $d$ would be easy: fit each degree to the training data and
keep the degree with the smallest test MSE. A separate test set is rarely
available. Collecting one costs as much as collecting the training data, and
setting aside part of the data already in hand leaves less to fit with.

**Resampling methods** estimate test error from the training data alone, by
fitting the model to part of the data and measuring its error on the part
left out. This lecture develops three, in increasing order of reuse: the
validation set approach, leave-one-out cross-validation, and $k$-fold
cross-validation.

## Validation set approach

The figure below shows the cars split at random into two halves of
`r nrow(Auto) / 2`, a training set (blue circles) and a validation set (orange
triangles).

::: {.callout-note collapse="true"}
### Code
```{r validation-splits}
#| output: false
n_auto <- nrow(Auto)
n_splits <- 100
splits <- map(seq_len(n_splits), \(r) sample(n_auto, n_auto / 2))
one_split <- splits[[1]]

set_colors <- c(Training = "#0072B2", Validation = "#E69F00")
split_plot <- Auto |>
  mutate(set = if_else(row_number() %in% one_split, "Training", "Validation")) |>
  ggplot(aes(x = horsepower, y = mpg, color = set, shape = set)) +
  geom_point(alpha = 0.7, size = 1.3) +
  scale_color_manual(values = set_colors) +
  scale_shape_manual(values = c(Training = 16, Validation = 17)) +
  labs(x = "Horsepower", y = "Miles per gallon", color = NULL, shape = NULL,
       title = "ISLR2::Auto, one random split")
```
:::

```{r validation-split-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4
split_plot
```

Both halves cover the range of horsepower, and either half shows the same
curved decline as the full data.

### Algorithm

The **validation set approach** estimates test MSE in three steps:

1. Split the $n$ observations at random into a training set and a validation
(hold-out) set $\mathcal{V}$.
2. Fit the model to the training set.
3. Predict every observation in the validation set and average the squared
errors,

$$\text{MSE}_{\mathcal{V}} = \frac{1}{n_{\mathcal{V}}}\sum_{i \in \mathcal{V}}\left[y_i - \hat f(x_i)\right]^2,$$

where

- $\mathcal{V}$ is the set of indices of the validation observations and
$n_{\mathcal{V}}$ their number, and
- $\hat f$ is the model fit to the training set only.

Every validation observation is new to $\hat f$, so $\text{MSE}_{\mathcal{V}}$
is a test MSE in the sense of the regression lecture. In R, with `train_id`
the row numbers of the training set:

```{r validation-mse}
validation_mse <- function(d, train_id, data = Auto) {
  fit <- fit_degree(d, data[train_id, ])
  valid <- data[-train_id, ]
  mean((valid$mpg - predict(fit, valid))^2)
}
one_split_mse <- map_dbl(degrees, validation_mse, train_id = one_split)
```

### Split variability

The split above is one of `r n_splits` random splits into halves. The first
tab plots its validation MSE against degree, the second adds the curves of all
`r n_splits` splits (grey, with the first in black), and the third counts the
degree with the smallest validation MSE in each split.

::: {.callout-note collapse="true"}
### Code
```{r validation-curves}
#| output: false
val_curves <- expand_grid(split = seq_len(n_splits), degree = degrees) |>
  mutate(mse = map2_dbl(degree, split, \(d, r) validation_mse(d, splits[[r]])))
val_best <- val_curves |>
  group_by(split) |>
  slice_min(mse, n = 1, with_ties = FALSE) |>
  ungroup()
val_choice_counts <- count(val_best, degree)
val_mse_d2 <- filter(val_curves, degree == 2)$mse
# Within each split, the spread of the estimates over degrees 2 through 10.
within_split_range <- val_curves |>
  filter(degree >= 2) |>
  group_by(split) |>
  summarize(range = max(mse) - min(mse), .groups = "drop")

mse_limits <- c(14, 30)
n_off_plot <- n_distinct(filter(val_curves, mse > mse_limits[2])$split)
degree_axis <- list(scale_x_continuous(breaks = degrees),
                    coord_cartesian(ylim = mse_limits))
one_split_plot <- ggplot(tibble(degree = degrees, mse = one_split_mse),
                         aes(x = degree, y = mse)) +
  geom_line() +
  geom_point() +
  degree_axis +
  labs(x = "Polynomial degree d", y = "Validation MSE",
       title = "One random split into halves")
all_splits_plot <- ggplot(val_curves, aes(x = degree, y = mse, group = split)) +
  geom_line(color = "grey70", linewidth = 0.3) +
  geom_line(data = filter(val_curves, split == 1), color = "black") +
  geom_point(data = filter(val_curves, split == 1), color = "black") +
  degree_axis +
  labs(x = "Polynomial degree d", y = "Validation MSE",
       title = paste(n_splits, "random splits into halves"))
choice_plot <- ggplot(val_best, aes(x = factor(degree, levels = degrees))) +
  geom_bar() +
  scale_x_discrete(drop = FALSE) +
  labs(x = "Degree with the smallest validation MSE", y = "Number of splits",
       title = paste(n_splits, "random splits into halves"))
```
:::

::: {.panel-tabset}
#### One split
```{r one-split-figure}
#| echo: false
#| fig-width: 6
#| fig-height: 3.5
one_split_plot
```

#### `r n_splits` splits
```{r all-splits-figure}
#| echo: false
#| fig-width: 6
#| fig-height: 3.5
all_splits_plot
```

#### Chosen degree
```{r val-choice-figure}
#| echo: false
#| fig-width: 6
#| fig-height: 3.5
choice_plot
```
:::

Every curve drops sharply from degree 1 to degree 2, so every split agrees
that a quadratic beats a line. Beyond that the splits disagree. The validation
MSE of the quadratic ranges from `r fmt(min(val_mse_d2))` to
`r fmt(max(val_mse_d2))` across splits, while within a split the estimates
for degrees 2 through 10 span a median of
`r fmt(median(within_split_range$range))`. Of the `r n_splits` curves,
`r n_off_plot` leave the top of the plot at high degree, where a polynomial
fit to half the cars extrapolates badly to a validation car outside their
range of horsepower. The degree with the smallest validation MSE is
`r and_list(as.character(sort(val_choice_counts$degree)))` in different splits;
the most common choice, degree
`r val_choice_counts$degree[which.max(val_choice_counts$n)]`, wins only
`r max(val_choice_counts$n)` of the `r n_splits` splits.

This is the first weakness of the validation set approach: the estimate, and
the model it selects, depend on which observations happen to land in which
half.

### Smaller training sets

The second weakness is that each model is fit to only part of the data.

::: {.callout-note collapse="true"}
### Code
```{r learning-curve}
#| output: false
train_sizes <- c(60, 80, 120, 160, n_auto / 2, 240, 280, 320, 360)
learning_degrees <- c(2, 7)
n_learning_reps <- 400
learning <- expand_grid(m = train_sizes, degree = learning_degrees,
                        rep = seq_len(n_learning_reps)) |>
  mutate(mse = map2_dbl(m, degree, \(m, d) validation_mse(d, sample(n_auto, m))))
learning_summary <- learning |>
  group_by(m, degree) |>
  summarize(median_mse = median(mse), mean_mse = mean(mse), .groups = "drop") |>
  mutate(degree = factor(paste("Degree", degree),
                         levels = paste("Degree", learning_degrees)))
learning_of <- function(m_value, d, col = "median_mse") {
  learning_summary[[col]][learning_summary$m == m_value &
                            learning_summary$degree == paste("Degree", d)]
}

learning_plot <- ggplot(learning_summary,
                        aes(x = m, y = median_mse, color = degree,
                            linetype = degree, shape = degree)) +
  geom_vline(xintercept = n_auto / 2, color = "grey60") +
  geom_line() +
  geom_point(size = 2) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Training set size m (vertical line: half of the cars)",
       y = "Median validation MSE", color = NULL, linetype = NULL,
       shape = NULL, title = paste(n_learning_reps, "random splits at each m"))
```
:::

To see what that costs, the cars were split at random into a training set of
$m$ cars and a validation set of the remaining $n - m$, for $m$ from
`r min(train_sizes)` to `r max(train_sizes)`, `r n_learning_reps` times at
each $m$. Because the validation cars are new to each fit, the typical
validation MSE at a given $m$ estimates the test MSE of a model fit to $m$
cars. The figure plots the median over splits for degrees
`r and_list(learning_degrees)`.

```{r learning-curve-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.8
learning_plot
```

Test MSE generally falls as the training set grows, and it falls more steeply
for the more flexible model. The degree-7 polynomial has a median validation MSE of
`r fmt(learning_of(min(train_sizes), 7))` when fit to `r min(train_sizes)`
cars, against `r fmt(learning_of(n_auto / 2, 7))` when fit to
`r n_auto / 2`; the quadratic goes from
`r fmt(learning_of(min(train_sizes), 2))` to
`r fmt(learning_of(n_auto / 2, 2))`. (The median is plotted because the mean
is dominated by a few catastrophic extrapolations: at
$m = `r min(train_sizes)`$ the mean validation MSE of degree 7 is
`r fmt(learning_of(min(train_sizes), 7, "mean_mse"), 0)`.) The model that will be used is
fit to all `r n_auto` cars, so a validation set estimate computed from fits to
half of them estimates the error of a worse model. The validation set
approach tends to **overestimate** the test error of the final fit, and more
so for flexible models and small $n$.

## Leave-one-out cross-validation

The validation set approach leaves out half the data. Leave-one-out
cross-validation leaves out one observation at a time. The figure below holds
out the car with the largest horsepower, a
`r str_to_title(as.character(Auto$name[which.max(Auto$horsepower)]))` with
`r max(Auto$horsepower)` horsepower (black triangle), and draws the fit to
all `r n_auto` cars (dashed) and the fit to the other `r n_auto - 1`
(solid), for degree 2 in the first tab and degree 10 in the second.

::: {.callout-note collapse="true"}
### Code
```{r held-out-car}
#| output: false
i_out <- which.max(Auto$horsepower)
held_out_summary <- map(c(2, 10), \(d) {
  fit_all <- fit_degree(d, Auto)
  fit_without <- fit_degree(d, Auto[-i_out, ])
  tibble(degree = d,
         residual = Auto$mpg[i_out] - fitted(fit_all)[[i_out]],
         loo_error = Auto$mpg[i_out] - predict(fit_without, Auto[i_out, ]),
         leverage = hatvalues(fit_all)[[i_out]])
}) |>
  list_rbind()
held_of <- function(d, col) held_out_summary[[col]][held_out_summary$degree == d]

fit_labels <- c("All cars", "Car held out")
held_out_plot <- function(d) {
  curves <- bind_rows(
    mutate(hp_grid, fit = fit_labels[1],
           mpg = predict(fit_degree(d, Auto), hp_grid)),
    mutate(hp_grid, fit = fit_labels[2],
           mpg = predict(fit_degree(d, Auto[-i_out, ]), hp_grid))) |>
    mutate(fit = factor(fit, levels = fit_labels))
  ggplot(Auto, aes(x = horsepower, y = mpg)) +
    geom_point(data = Auto[-i_out, ], color = "grey60", alpha = 0.5, size = 1) +
    geom_line(data = curves, aes(color = fit, linetype = fit), linewidth = 0.9) +
    geom_point(data = Auto[i_out, ], shape = 17, size = 3.5) +
    scale_color_manual(values = c("#0072B2", "#D55E00")) +
    scale_linetype_manual(values = c("dashed", "solid")) +
    coord_cartesian(ylim = range(Auto$mpg)) +
    labs(x = "Horsepower", y = "Miles per gallon", color = "Fit to",
         linetype = "Fit to", title = paste("Degree", d))
}
held_out_plots <- map(c(2, 10), held_out_plot)
```
:::

::: {.panel-tabset}
#### Degree 2
```{r held-out-d2-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4
held_out_plots[[1]]
```

#### Degree 10
```{r held-out-d10-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4
held_out_plots[[2]]
```
:::

For the quadratic the two fits nearly coincide, and the held-out car is
predicted about as well without it as with it. For degree 10 the fit to all
cars bends to pass near the held-out car, and without it the curve swings far
below. The squared error of the held-out prediction measures how well the
model predicts a car it has not seen, which the residual from the fit to all
cars does not.

### LOOCV estimate

**Leave-one-out cross-validation** (LOOCV) repeats that for every
observation:

$$\text{CV}_{(n)} = \frac{1}{n}\sum_{i=1}^n\left[y_i - \hat f^{(-i)}(x_i)\right]^2,$$

where

- $\hat f^{(-i)}$ is the model fit to the $n - 1$ observations other than
observation $i$, and
- $\hat f^{(-i)}(x_i)$ is its prediction for the held-out observation.

Each fit uses $n - 1$ observations, nearly all of the data, and there is no
random split: every observation is held out exactly once. Written as a loop,
LOOCV costs $n$ fits per model.

```{r loocv-loop}
loocv_mse <- function(d, data = Auto) {
  sq_error <- numeric(nrow(data))
  for (i in seq_len(nrow(data))) {
    fit <- fit_degree(d, data[-i, ])
    sq_error[i] <- (data$mpg[i] - predict(fit, data[i, ]))^2
  }
  mean(sq_error)
}
loocv_loop <- map_dbl(degrees, loocv_mse)
```

### Least squares shortcut

For a model fit by least squares, the $n$ fits are unnecessary. Write the
least squares fit to all $n$ observations as $\hat y = \mathbf{H}y$, with

$$\mathbf{H} = \mathbf{X}\left(\mathbf{X}^\top\mathbf{X}\right)^{-1}\mathbf{X}^\top, \qquad \hat y_i = \sum_{j=1}^n h_{ij}\,y_j, \qquad h_i = h_{ii} = x_i\left(\mathbf{X}^\top\mathbf{X}\right)^{-1}x_i^\top,$$

where

- $\mathbf{H}$ is the hat matrix and $h_{ij}$ its $(i, j)$ entry,
- $\hat y_i = x_i\hat\beta$ is the fitted value of observation $i$ from the fit
to all $n$ observations, and
- $h_i$ is the **leverage** of observation $i$, as in the flexibility lecture,
with $0 \le h_i \le 1$.

Let $\hat\beta^{(-i)}$ be the least squares estimate without observation $i$,
so $\hat f^{(-i)}(x_i) = x_i\hat\beta^{(-i)}$, and assume $\mathbf{X}$ without
row $i$ still has full column rank so that $\hat\beta^{(-i)}$ is unique.

**Step 1: a full-data fit that reproduces $\hat\beta^{(-i)}$.** Define a
response vector $z$ equal to $y$ except in entry $i$, which is replaced by the
held-out prediction: $z_j = y_j$ for $j \ne i$ and
$z_i = x_i\hat\beta^{(-i)}$. For any $\beta$,

$$\sum_{j=1}^n\left(z_j - x_j\beta\right)^2 = \sum_{j \ne i}\left(y_j - x_j\beta\right)^2 + \left(x_i\hat\beta^{(-i)} - x_i\beta\right)^2.$$

The first sum is minimized at $\beta = \hat\beta^{(-i)}$, by definition, and
the second term is zero there, its smallest possible value. So
$\hat\beta^{(-i)}$ minimizes the left side: it is the least squares estimate
for response $z$ with all $n$ rows,
$\hat\beta^{(-i)} = \left(\mathbf{X}^\top\mathbf{X}\right)^{-1}\mathbf{X}^\top z$.

**Step 2: apply the hat matrix.** Multiplying by $x_i$,
$x_i\hat\beta^{(-i)}$ is entry $i$ of $\mathbf{H}z$:

$$\hat f^{(-i)}(x_i) = \sum_{j=1}^n h_{ij}\,z_j = \sum_{j=1}^n h_{ij}\,y_j - h_i\,y_i + h_i\,\hat f^{(-i)}(x_i) = \hat y_i - h_i\,y_i + h_i\,\hat f^{(-i)}(x_i).$$

**Step 3: solve.** Collecting the $\hat f^{(-i)}(x_i)$ terms,
$\left(1 - h_i\right)\hat f^{(-i)}(x_i) = \hat y_i - h_i y_i$. The
full-rank assumption on $\mathbf{X}$ without row $i$ makes $h_i < 1$, so
dividing by $1 - h_i$,

$$y_i - \hat f^{(-i)}(x_i) = \frac{\left(1 - h_i\right)y_i - \hat y_i + h_i y_i}{1 - h_i} = \frac{y_i - \hat y_i}{1 - h_i}.$$

Squaring and averaging over $i$,

$$\text{CV}_{(n)} = \frac{1}{n}\sum_{i=1}^n\left(\frac{y_i - \hat y_i}{1 - h_i}\right)^2.$$

The held-out error is the ordinary residual inflated by $1/(1 - h_i)$. With
$h_i = 0$ for every $i$ this would be the training MSE. The inflation is
small for a typical observation and large for a high-leverage one, whose
fitted value the full-data fit pulls toward its own response. The formula
needs one fit, its residuals, and `hatvalues()`. It holds for any model fit
by least squares on fixed features: polynomials, interactions, step functions
with fixed cut points. It does not hold for logistic regression or KNN, where
LOOCV takes $n$ fits.

For the held-out car in the figure, the quadratic gives leverage
$h_i = `r fmt(held_of(2, "leverage"), 4)`$, residual
$y_i - \hat y_i = `r fmt(held_of(2, "residual"), 3)`$, and signed held-out
error $y_i - \hat f^{(-i)}(x_i) = `r fmt(held_of(2, "loo_error"))`$; the
degree-10 polynomial gives leverage
$h_i = `r fmt(held_of(10, "leverage"), 4)`$, residual
`r fmt(held_of(10, "residual"), 3)`, and signed held-out error
`r fmt(held_of(10, "loo_error"))`, which equals
$`r fmt(held_of(10, "residual"), 3)`/(1 - `r fmt(held_of(10, "leverage"), 4)`)$
up to rounding.

The shortcut for every degree:

```{r loocv-shortcut}
loocv_shortcut <- function(d, data = Auto) {
  fit <- fit_degree(d, data)
  mean(((data$mpg - fitted(fit)) / (1 - hatvalues(fit)))^2)
}
loocv_fast <- map_dbl(degrees, loocv_shortcut)
```

::: {.callout-note collapse="true"}
### Code
```{r loocv-table}
#| output: false
loocv_table <- tibble(degree = degrees,
                      loop = fmt(loocv_loop, 4),
                      shortcut = fmt(loocv_fast, 4)) |>
  knitr::kable(col.names = c("Degree $d$",
                             paste0("Loop (", n_auto, " fits)"),
                             "Shortcut (1 fit)"),
               align = "ccc")
max_loocv_diff <- max(abs(loocv_loop - loocv_fast))
```
:::

```{r loocv-table-display}
#| echo: false
loocv_table
```

The two columns agree to the digits shown; the largest absolute difference
is $`r sci(max_loocv_diff)`$, rounding error. The loop needed
`r format(n_auto * length(degrees), big.mark = ",")` least squares fits and
the shortcut `r length(degrees)`.

### Degree selection

The figure below plots the LOOCV estimate and the training MSE against
degree.

::: {.callout-note collapse="true"}
### Code
```{r loocv-curve}
#| output: false
loocv_best <- degrees[which.min(loocv_fast)]
error_labels <- c("Training MSE", "LOOCV")
loocv_curve_data <- bind_rows(
  tibble(degree = degrees, mse = train_mse, estimate = error_labels[1]),
  tibble(degree = degrees, mse = loocv_fast, estimate = error_labels[2])) |>
  mutate(estimate = factor(estimate, levels = error_labels))
loocv_curve_plot <- ggplot(loocv_curve_data,
                           aes(x = degree, y = mse, color = estimate,
                               linetype = estimate, shape = estimate)) +
  geom_line() +
  geom_point(size = 2) +
  scale_x_continuous(breaks = degrees) +
  scale_color_manual(values = c("grey40", "#D55E00")) +
  labs(x = "Polynomial degree d", y = "MSE", color = NULL, linetype = NULL,
       shape = NULL, title = "ISLR2::Auto")
```
:::

```{r loocv-curve-figure}
#| echo: false
#| fig-width: 6.5
#| fig-height: 3.5
loocv_curve_plot
```

The LOOCV estimate drops from `r fmt(loocv_fast[1])` at degree 1 to
`r fmt(loocv_fast[2])` at degree 2 and then barely moves. Its minimum,
`r fmt(min(loocv_fast))`, is at degree `r loocv_best`, only
`r fmt(loocv_fast[2] - min(loocv_fast))` below the quadratic. Unlike the
training MSE, it does not keep falling: at degree 10 it is
`r fmt(loocv_fast[10])`, above its minimum, while the training MSE is at its
lowest.

## k-fold cross-validation

::: {.callout-note collapse="true"}
### Code
```{r auto-folds}
#| output: false
n_folds <- 10
fold_auto <- sample(rep(seq_len(n_folds), length.out = n_auto))
fold_sizes <- table(fold_auto)

fold_roles <- c("Training", "Validation")
fold_plot <- map(seq_len(n_folds), \(l) {
  mutate(Auto, panel = paste("Fold", l, "held out"),
         role = if_else(fold_auto == l, fold_roles[2], fold_roles[1]))
}) |>
  list_rbind() |>
  mutate(panel = factor(panel, levels = paste("Fold", seq_len(n_folds), "held out")),
         role = factor(role, levels = fold_roles)) |>
  arrange(role) |>
  ggplot(aes(x = horsepower, y = mpg, color = role, shape = role, size = role)) +
  geom_point(alpha = 0.8) +
  facet_wrap(~ panel, nrow = 2) +
  scale_color_manual(values = c("grey70", "black")) +
  scale_shape_manual(values = c(16, 17)) +
  scale_size_manual(values = c(0.5, 1.3)) +
  labs(x = "Horsepower", y = "Miles per gallon", color = NULL, shape = NULL,
       size = NULL, title = "ISLR2::Auto in 10 folds") +
  theme(legend.position = "bottom")
```
:::

The figure below splits the `r n_auto` cars at random into `r n_folds` groups
of about equal size. Each panel marks one group as black triangles and the
rest as grey circles.

```{r auto-folds-figure}
#| echo: false
#| fig-width: 9
#| fig-height: 4.8
fold_plot
```

The groups hold `r and_list(as.character(sort(unique(as.vector(fold_sizes)))))`
cars, and every car appears as a triangle in exactly one panel. Each group is
scattered across the whole range of horsepower.

### k-fold estimate

**$k$-fold cross-validation** divides the observations at random into $k$
groups, or **folds**, $\mathcal{F}_1, \ldots, \mathcal{F}_k$, of about equal
size. For each fold $\ell = 1, \ldots, k$ it fits the model to the other $k - 1$
folds and computes the MSE on fold $\ell$,

$$\text{MSE}_\ell = \frac{1}{n_\ell}\sum_{i \in \mathcal{F}_\ell}\left[y_i - \hat f^{(-\ell)}(x_i)\right]^2, \qquad \text{CV}_{(k)} = \sum_{\ell=1}^k \frac{n_\ell}{n}\,\text{MSE}_\ell,$$

where

- $k$ is the **number of folds**, written lowercase as in the name $k$-fold
cross-validation; it is unrelated to $K$, the number of neighbors in KNN (or
of classes in the generative classifier lectures),
- $\mathcal{F}_\ell$ is the set of indices of the observations in fold $\ell$
and $n_\ell$ their number,
- $\hat f^{(-\ell)}$ is the model fit to the observations outside fold
$\ell$, and
- $\text{CV}_{(k)}$ is the average of the fold MSEs weighted by fold size,
which is $\frac{1}{k}\sum_\ell \text{MSE}_\ell$ when every fold has $n/k$
observations, and is the average squared held-out error over all $n$
observations.

Every observation is held out exactly once and used for fitting $k - 1$
times. LOOCV is the special case $k = n$, with one observation per fold.

```{r kfold-function}
kfold_mse <- function(d, fold, data = Auto) {
  sq_error <- numeric(nrow(data))
  for (l in unique(fold)) {
    held_out <- fold == l
    fit <- fit_degree(d, data[!held_out, ])
    sq_error[held_out] <- (data$mpg[held_out] - predict(fit, data[held_out, ]))^2
  }
  mean(sq_error)
}
kfold_10 <- map_dbl(degrees, kfold_mse, fold = fold_auto)
```

With the folds in the figure, the 10-fold estimate is
`r fmt(kfold_10[1])` at degree 1 and `r fmt(kfold_10[2])` at degree 2, and is
smallest, `r fmt(min(kfold_10))`, at degree `r degrees[which.min(kfold_10)]`.

### Partition variability

::: {.callout-note collapse="true"}
### Code
```{r kfold-partitions}
#| output: false
n_partitions <- n_splits
partition_k <- c(5, 10)
kfold_curves <- expand_grid(k = partition_k, partition = seq_len(n_partitions)) |>
  mutate(fold = map(k, \(k) sample(rep(seq_len(k), length.out = n_auto))),
         mse = map(fold, \(f) map_dbl(degrees, kfold_mse, fold = f)),
         degree = list(degrees)) |>
  select(-fold) |>
  unnest(c(degree, mse))

method_labels <- c("Validation set", "5-fold", "10-fold")
resample_curves <- bind_rows(
  transmute(val_curves, method = method_labels[1], rep = split, degree, mse),
  transmute(kfold_curves, method = paste0(k, "-fold"), rep = partition, degree, mse)) |>
  mutate(method = factor(method, levels = method_labels))
resample_summary <- resample_curves |>
  group_by(method) |>
  summarize(sd_d2 = sd(mse[degree == 2]), .groups = "drop")
summary_of <- function(m, col) resample_summary[[col]][resample_summary$method == m]
chosen_by <- resample_curves |>
  group_by(method, rep) |>
  slice_min(mse, n = 1, with_ties = FALSE) |>
  ungroup()
chosen_degrees <- function(m) sort(unique(chosen_by$degree[chosen_by$method == m]))

loocv_line <- tibble(degree = degrees, mse = loocv_fast)
resample_plot <- function(m) {
  ggplot(filter(resample_curves, method == m), aes(x = degree, y = mse)) +
    geom_line(aes(group = rep), color = "grey65", linewidth = 0.3) +
    geom_line(data = loocv_line, color = "black", linetype = "dashed",
              linewidth = 1) +
    degree_axis +
    labs(x = "Polynomial degree d", y = "Estimated test MSE",
         title = paste0(m, ", ", n_partitions,
                        " random partitions; LOOCV dashed"))
}
resample_plots <- map(method_labels, resample_plot)
```
:::

The `r n_folds`-fold estimate above comes from one random partition of the
cars. To
see whether it changes from one partition to another, and by how much, the
tabs below repeat
the validation set curves of `r n_splits` random splits and add the curves of
`r n_partitions` random partitions into `r partition_k[1]` folds and into
`r partition_k[2]` folds, each over the LOOCV curve (black, dashed), on the
same axes.

::: {.panel-tabset}
#### Validation set
```{r resample-val-figure}
#| echo: false
#| fig-width: 6.5
#| fig-height: 3.6
resample_plots[[1]]
```

#### 5-fold
```{r resample-5fold-figure}
#| echo: false
#| fig-width: 6.5
#| fig-height: 3.6
resample_plots[[2]]
```

#### 10-fold
```{r resample-10fold-figure}
#| echo: false
#| fig-width: 6.5
#| fig-height: 3.6
resample_plots[[3]]
```
:::

The band of curves narrows from tab to tab. At degree 2 the standard
deviation of the estimate across random partitions is
`r fmt(summary_of("Validation set", "sd_d2"))` for the validation set,
`r fmt(summary_of("5-fold", "sd_d2"), 3)` for 5-fold, and
`r fmt(summary_of("10-fold", "sd_d2"), 3)` for 10-fold cross-validation. The
5-fold and 10-fold curves hug the LOOCV curve, while the validation set
curves scatter widely on both sides of it: every observation is held out once
and the held-out errors are averaged over all $n$ of them, so the estimate
depends far less on which observations share a fold. The degree with the smallest estimate
is still not fixed: across partitions 10-fold cross-validation chooses degree
`r and_list(as.character(chosen_degrees("10-fold")))`, because the curves are
nearly flat from degree 2 on.

### Fold count

Any $k$ from 2 to $n$ gives a cross-validation estimate, and the choice
affects its bias, its variance, and its computation.

::: {.callout-note collapse="true"}
### Code
```{r k-simulation}
#| output: false
n_sample <- 50
n_sims <- 2000
k_values <- c(2, 5, 10, n_sample)
k_labels <- c("2-fold", "5-fold", "10-fold", paste0("LOOCV (k = ", n_sample, ")"))
k_sims <- map(seq_len(n_sims), \(r) {
  id <- sample(n_auto, n_sample)
  cars <- Auto[id, ]
  others <- Auto[-id, ]
  test_mse <- mean((others$mpg - predict(fit_degree(2, cars), others))^2)
  cv <- map_dbl(k_values, \(k) {
    if (k == n_sample) return(loocv_shortcut(2, cars))
    kfold_mse(2, sample(rep(seq_len(k), length.out = n_sample)), cars)
  })
  tibble(sim = r, method = k_labels, estimate = cv, test_mse = test_mse)
}) |>
  list_rbind() |>
  mutate(method = factor(method, levels = k_labels))
k_summary <- k_sims |>
  group_by(method) |>
  summarize(mean_estimate = mean(estimate), sd_estimate = sd(estimate),
            bias = mean(estimate - test_mse),
            se_bias = sd(estimate - test_mse) / sqrt(n()), .groups = "drop")
mean_test_mse <- mean(k_sims$test_mse[k_sims$method == k_labels[1]])
k_of <- function(m, col) k_summary[[col]][k_summary$method == m]

# A few 2-fold estimates are extreme: a quadratic fit to half of a small
# sample can extrapolate badly to a held-out car.
two_fold <- k_sims$estimate[k_sims$method == k_labels[1]]
extreme <- two_fold > 5 * mean_test_mse
n_extreme <- sum(extreme)
max_extreme <- max(two_fold)
sd_two_fold_trimmed <- sd(two_fold[!extreme])

k_plot <- ggplot(k_sims, aes(x = method, y = estimate)) +
  geom_boxplot(outlier.size = 0.6, outlier.alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3,
               fill = "#E69F00") +
  geom_hline(yintercept = mean_test_mse, linetype = "dashed") +
  coord_cartesian(ylim = quantile(k_sims$estimate, c(0.005, 0.995))) +
  labs(x = NULL, y = "Cross-validation estimate of test MSE",
       title = paste0(format(n_sims, big.mark = ","), " samples of ", n_sample,
                      " cars; quadratic; diamond = mean; dashed = mean test MSE"),
       caption = "Vertical axis clipped to the middle 99% of the estimates")
```
:::

To see how $k$ affects the bias and variance of the estimate, the
`r n_auto` cars were treated as a population. Each of `r format(n_sims, big.mark = ",")` times, a
sample of `r n_sample` cars was drawn without replacement and a quadratic was
fit to it; the actual test MSE of that fit is its average squared error on the
`r n_auto - n_sample` cars not drawn. The figure compares the cross-validation
estimates from the same `r n_sample` cars, with $k$ equal to
`r and_list(k_values)` (LOOCV), against it.

```{r k-simulation-figure}
#| echo: false
#| fig-width: 7.5
#| fig-height: 4
k_plot
```

The average actual test MSE of the fits is `r fmt(mean_test_mse)`. The
estimates average `r fmt(k_of("2-fold", "mean_estimate"))` for 2-fold,
`r fmt(k_of("5-fold", "mean_estimate"))` for 5-fold,
`r fmt(k_of("10-fold", "mean_estimate"))` for 10-fold, and
`r fmt(k_of(k_labels[4], "mean_estimate"))` for LOOCV. 2-fold
cross-validation, fitting to `r n_sample / 2` cars, overestimates by
`r fmt(k_of(k_labels[1], "bias"))`, with a simulation standard error of
`r fmt(k_of(k_labels[1], "se_bias"))`; by $k = `r k_values[2]`$ the
overestimate is small, `r fmt(k_of(k_labels[2], "bias"))`, with a simulation
standard error of `r fmt(k_of(k_labels[2], "se_bias"))`. The standard
deviations of the estimates are `r fmt(k_of("2-fold", "sd_estimate"))`,
`r fmt(k_of("5-fold", "sd_estimate"))`,
`r fmt(k_of("10-fold", "sd_estimate"))`, and
`r fmt(k_of(k_labels[4], "sd_estimate"))`, so 10-fold and LOOCV vary about
equally. The 2-fold standard deviation is dominated by `r n_extreme` of the
`r format(n_sims, big.mark = ",")` samples, whose estimates exceed five times
the mean test MSE (the largest is `r fmt(max_extreme)`), where a quadratic fit
to `r n_sample / 2` cars extrapolates badly to a held-out car; without them it
is `r fmt(sd_two_fold_trimmed)`. The figure's vertical axis is clipped, so
those estimates are not shown.

Three considerations explain these results and trade off in choosing $k$.

- **Computation.** $k$-fold cross-validation needs $k$ fits per model, LOOCV
$n$. For least squares the shortcut makes LOOCV cost one fit; for most other
methods it is $n$ fits, and $k = 5$ or $10$ is a large saving.
- **Bias.** Each fit uses $n(k-1)/k$ observations: for all `r n_auto` cars,
`r n_auto / 2` with $k = 2$, about `r round(n_auto * 4 / 5)` with $k = 5$,
about `r round(n_auto * 9 / 10)` with $k = 10$, and `r n_auto - 1` with
LOOCV. By the learning curve above, a model fit to fewer observations has
larger test error, so $\text{CV}_{(k)}$ overestimates the test error of the
model fit to all $n$, by more as $k$ shrinks. LOOCV is nearly unbiased.
- **Variance.** $\text{CV}_{(k)}$ is an average of $k$ fold MSEs. If each has
variance $\tau^2$ and each pair has correlation $\rho$,

$$Var\left[\frac{1}{k}\sum_{\ell=1}^k\text{MSE}_\ell\right] = \frac{1}{k^2}\sum_{\ell=1}^k\sum_{\ell'=1}^k Cov\left[\text{MSE}_\ell, \text{MSE}_{\ell'}\right] = \tau^2\left(\rho + \frac{1-\rho}{k}\right),$$

where $\ell'$ is a second fold index. Averaging more folds shrinks the
$(1-\rho)/k$ term but not $\rho\tau^2$. In LOOCV any two fits share $n - 2$
of their $n - 1$ observations, so their held-out errors are strongly
correlated and averaging $n$ of them reduces the variance less than the
number $n$ suggests; with larger $k$ each fold MSE is also computed from fewer
observations, raising $\tau^2$. Which effect dominates depends on the
method and the data.

The standard deviations in the simulation combine two sources: which
`r n_sample` cars were drawn, shared by all four estimates, and, for $k$-fold,
the random partition. The partition alone is the cleaner comparison. With all
`r n_auto` cars held fixed, the degree-2 estimate varied across random
partitions with standard deviation `r fmt(summary_of("5-fold", "sd_d2"), 3)`
for 5-fold and `r fmt(summary_of("10-fold", "sd_d2"), 3)` for 10-fold
cross-validation, and LOOCV, with no partition, does not vary at all.

For a quadratic fit to `r n_sample` cars, 5-fold and 10-fold
cross-validation had small bias and variance close to LOOCV's, and for a
method without the least squares shortcut they need 5 or 10 fits rather than
$n$. That combination is why $k = 5$ or $10$ is the usual choice.

## Classification

### Credit card default data

::: {.callout-note collapse="true"}
### Code
```{r default-split}
#| output: false
# The QDA lecture's split, recreated with its seed; the generator's state is
# restored afterward so this lecture's later draws are unaffected.
rng_state <- .Random.seed
set.seed(20261001)
train_id <- sample(nrow(Default), nrow(Default) / 2)
assign(".Random.seed", rng_state, envir = globalenv())
train <- Default[train_id, ]
test  <- Default[-train_id, ]

# The QDA lecture's rule of thumb: the smallest odd integer above sqrt(n).
odd_above_sqrt <- function(n) {
  k <- floor(sqrt(n))
  if (k %% 2 == 0) k + 1 else k + 2
}
K_rule <- odd_above_sqrt(nrow(train))

dollar_labels <- function(v) {
  ifelse(is.na(v), NA,
         paste0("$", format(v, big.mark = ",", scientific = FALSE, trim = TRUE)))
}
default_plot <- ggplot(arrange(train, default),
                       aes(x = balance, y = income, color = default,
                           shape = default)) +
  geom_point(alpha = 0.4, size = 0.9) +
  scale_color_manual(values = class_colors) +
  scale_shape_manual(values = c(No = 16, Yes = 17)) +
  scale_x_continuous(labels = dollar_labels) +
  scale_y_continuous(labels = dollar_labels) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  labs(x = "Credit card balance ($)", y = "Income ($)", color = "Default",
       shape = "Default", title = "ISLR2::Default, training half")
```
:::

The QDA and naive Bayes lecture classified `ISLR2::Default` credit card
customers as defaulters or not from `balance` and `income`, fitting every
classifier to a random half of the
`r format(nrow(Default), big.mark = ",")` customers and evaluating it on the
other half. Its KNN classifier used $K = `r K_rule`$ neighbors, the smallest
odd integer above $\sqrt{n}$ for the training half, a rule of thumb. The
figure below plots the same training half, recreated with the same seed.

```{r default-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
default_plot
```

The training half holds `r format(nrow(train), big.mark = ",")` customers, of
whom `r sum(train$default == "Yes")` (`r pct(mean(train$default == "Yes"), 2)`)
defaulted. Defaulters sit at high balances at every income, mixed among
non-defaulters with similar balances.

### Cross-validation error rate

Cross-validation for a classifier replaces the squared error with a
classification loss. With the misclassification indicator,

$$\text{Err}_\ell = \frac{1}{n_\ell}\sum_{i \in \mathcal{F}_\ell}\mathrm{I}\left(y_i \ne \hat y_i^{(-\ell)}\right), \qquad \text{CV}_{(k)} = \sum_{\ell=1}^k\frac{n_\ell}{n}\,\text{Err}_\ell,$$

where

- $\hat y_i^{(-\ell)}$ is the class predicted for observation $i$ by the
classifier fit to the observations outside fold $\ell$, and
- $\text{Err}_\ell$ is the test error rate of that classifier on fold $\ell$.

Any loss can take the place of the indicator. Log loss, from the
classification lecture, scores the predicted probabilities rather than the
classes: with $y_i$ coded $1$ for the positive class and $0$ otherwise, it
replaces $\mathrm{I}\left(y_i \ne \hat y_i^{(-\ell)}\right)$ with
$-\left\{y_i\log\hat p^{(-\ell)}(x_i) + (1 - y_i)\log\left[1 - \hat p^{(-\ell)}(x_i)\right]\right\}$,
where $\hat p^{(-\ell)}(x_i)$ is the predicted probability of the positive
class. A KNN vote share of exactly $0$ or $1$ on a misclassified observation
makes it infinite, so log loss suits classifiers whose probabilities stay
away from $0$ and $1$, such as logistic regression.

For KNN, the classifier fit to the training folds includes the
standardization of the features: their means and standard deviations are
computed from the training folds, as they were computed from the training half
in the QDA lecture.

```{r knn-cv-function}
features <- c("balance", "income")
knn_classify <- function(fit_data, new_data, K) {
  center <- colMeans(fit_data[features])
  spread <- map_dbl(fit_data[features], sd)
  class::knn(scale(fit_data[features], center, spread),
             scale(new_data[features], center, spread),
             fit_data$default, k = K)
}
knn_cv_error <- function(K, fold, data = train) {
  wrong <- logical(nrow(data))
  for (l in unique(fold)) {
    held_out <- fold == l
    predicted <- knn_classify(data[!held_out, ], data[held_out, ], K)
    wrong[held_out] <- predicted != data$default[held_out]
  }
  mean(wrong)
}
```

### Neighbor count

::: {.callout-note collapse="true"}
### Code
```{r knn-cv}
#| output: false
fold_default <- sample(rep(seq_len(n_folds), length.out = nrow(train)))
K_grid <- seq(1, 201, by = 2)
knn_errors <- tibble(K = K_grid) |>
  mutate(Training = map_dbl(K, \(K) mean(knn_classify(train, train, K) != train$default)),
         `10-fold CV` = map_dbl(K, knn_cv_error, fold = fold_default),
         `Test half` = map_dbl(K, \(K) mean(knn_classify(train, test, K) != test$default)))
K_cv <- knn_errors$K[which.min(knn_errors$`10-fold CV`)]
K_test <- knn_errors$K[which.min(knn_errors$`Test half`)]
share_test_below <- mean(knn_errors$`Test half` < knn_errors$`10-fold CV`)
knn_of <- function(K_value, col) knn_errors[[col]][knn_errors$K == K_value]
test_below_cv_min <- knn_of(K_cv, "Test half") < min(knn_errors$`10-fold CV`)

knn_long <- knn_errors |>
  pivot_longer(-K, names_to = "error", values_to = "rate") |>
  mutate(error = factor(error, levels = c("Training", "10-fold CV", "Test half")))
knn_plot <- ggplot(knn_long, aes(x = 1 / K, y = rate, color = error,
                                 linetype = error, shape = error)) +
  geom_vline(xintercept = 1 / c(K_cv, K_rule), color = "grey70") +
  geom_line() +
  geom_point(data = filter(knn_long, K %in% c(1, 3, 5, 11, 21, 51, 101, 201)),
             size = 2) +
  annotate("text", x = 1 / c(K_cv, K_rule), y = max(knn_long$rate), hjust = -0.1,
           label = c(paste("K =", K_cv, "(CV)"), paste("K =", K_rule, "(rule)")),
           size = 3) +
  scale_x_log10(breaks = 1 / c(1, 3, 10, 30, 100, 200),
                labels = c("1", "1/3", "1/10", "1/30", "1/100", "1/200")) +
  scale_y_continuous(labels = \(v) paste0(100 * v, "%")) +
  scale_color_manual(values = c("grey40", "#D55E00", "#0072B2")) +
  labs(x = "Flexibility 1/K (log scale)", y = "Error rate", color = NULL,
       linetype = NULL, shape = NULL, title = "KNN on ISLR2::Default")
```
:::

The training half was divided into `r n_folds` folds, and the
`r n_folds`-fold cross-validation error rate was computed for every odd $K$
from `r min(K_grid)` to `r max(K_grid)`. The figure plots it against $1/K$,
the flexibility, with the training error rate and the error rate on the test
half for comparison.

```{r knn-cv-figure}
#| echo: false
#| fig-width: 7.5
#| fig-height: 4
knn_plot
```

The training error rate falls to `r pct(knn_of(1, "Training"), 2)` at $K = 1$,
where every training customer is its own nearest neighbor. The
cross-validation error rate is U-shaped instead: `r pct(knn_of(1, "10-fold CV"), 2)`
at $K = 1$, a minimum of `r pct(min(knn_errors$"10-fold CV"), 2)` at
$K = `r K_cv`$, and `r pct(knn_of(max(K_grid), "10-fold CV"), 2)` at
$K = `r max(K_grid)`$, where
KNN classifies every customer as a non-defaulter and its error rate equals the
proportion of defaulters. The rule of thumb's $K = `r K_rule`$ has a
cross-validation error rate of `r pct(knn_of(K_rule, "10-fold CV"), 2)`.

The test half, which the cross-validation never touched, gives the same
shape: its error rate is `r pct(knn_of(K_cv, "Test half"), 2)` at
$K = `r K_cv`$ and `r pct(knn_of(K_rule, "Test half"), 2)` at $K = `r K_rule`$,
and is smallest, `r pct(min(knn_errors$"Test half"), 2)`, at $K = `r K_test`$.
The test curve lies below the cross-validation curve
`r if (share_test_below == 1) "at every value" else paste("at", pct(share_test_below, 0), "of the values")`
of $K$, in part because defaulters
are rarer in the test half, `r pct(mean(test$default == "Yes"), 2)` against
`r pct(mean(train$default == "Yes"), 2)`, which also sets the level of each
curve at large $K$.

## Data-dependent steps

The data sets in this section have thousands of features and only dozens of
observations, and none of the features carries any information about the
response. They are simulated rather than real because the point concerns the
data-generating process: with no information in the features, the true test
error rate of any classifier is known, so a cross-validation estimate can be
checked against it.

### Noise feature screening

::: {.callout-note collapse="true"}
### Code
```{r noise-data}
#| output: false
n_noise <- 50
p_noise <- 5000
q_keep <- 20
K_noise <- 5
# Two classes of equal size and no information in the features.
true_error <- 1 / 2
simulate_noise <- function() {
  list(X = matrix(rnorm(n_noise * p_noise), n_noise, p_noise),
       y = factor(rep(c("A", "B"), each = n_noise / 2)))
}
screen <- function(X, y, q) {
  order(abs(cor(X, as.numeric(y))), decreasing = TRUE)[seq_len(q)]
}

noise_example <- simulate_noise()
top_two <- screen(noise_example$X, noise_example$y, 2)
top_cor <- abs(cor(noise_example$X[, top_two], as.numeric(noise_example$y)))
# Which class sits higher (second feature) and further right (first feature).
class_means <- sapply(top_two, \(j) tapply(noise_example$X[, j], noise_example$y, mean))
upper_class <- rownames(class_means)[which.max(class_means[, 2])]
right_class <- rownames(class_means)[which.max(class_means[, 1])]
other_class <- function(cl) setdiff(levels(noise_example$y), cl)
noise_plot <- tibble(first = noise_example$X[, top_two[1]],
                     second = noise_example$X[, top_two[2]],
                     class = noise_example$y) |>
  ggplot(aes(x = first, y = second, color = class, shape = class)) +
  geom_point(size = 2.2) +
  scale_color_manual(values = c(A = "#0072B2", B = "#D55E00")) +
  labs(x = paste("Feature", top_two[1]), y = paste("Feature", top_two[2]),
       color = "Class", shape = "Class",
       title = paste("The two of", format(p_noise, big.mark = ","),
                     "noise features most correlated with the class"))
```
:::

Each simulated data set has $n = `r n_noise`$ observations in two classes of
`r n_noise / 2` and $p = `r format(p_noise, big.mark = "{,}")`$ features, all
independent standard normal draws unrelated to the class, so the true test
error rate of any classifier is `r pct(true_error, 0)`. A common first step
with so many features is **screening**: keep the `r q_keep` features most
correlated with the response, then fit a classifier to those. The figure plots
the two features of one data set with the largest absolute correlation with
the class.

```{r noise-figure}
#| echo: false
#| fig-width: 6
#| fig-height: 4.2
noise_plot
```

The classes look partly separated: class `r upper_class` tends to sit higher
than class `r other_class(upper_class)`, and class `r right_class` further
right than class `r other_class(right_class)`. The absolute correlations of
these two features with the class are `r fmt(top_cor[1])` and
`r fmt(top_cor[2])`. Among
`r format(p_noise, big.mark = ",")` features unrelated to the class, some are
correlated with it in a sample of `r n_noise` by chance, and screening finds
exactly those.

### Within-fold screening

The function below estimates the error rate of "screen the `r q_keep` most
correlated features, then classify by KNN with $K = `r K_noise`$" by
`r n_folds`-fold cross-validation, in one of two ways. With `inside = FALSE`
it screens once, on all $n$ observations, and cross-validates only the KNN step. With
`inside = TRUE` it screens again within each fold, using only the training
folds.

```{r screen-cv}
screen_cv_error <- function(X, y, inside,
                            fold = sample(rep(seq_len(n_folds), length.out = nrow(X)))) {
  keep_all <- screen(X, y, q_keep)
  wrong <- logical(nrow(X))
  for (l in seq_len(n_folds)) {
    held_out <- fold == l
    keep <- if (inside) screen(X[!held_out, ], y[!held_out], q_keep) else keep_all
    predicted <- class::knn(X[!held_out, keep], X[held_out, keep, drop = FALSE],
                            y[!held_out], k = K_noise)
    wrong[held_out] <- predicted != y[held_out]
  }
  mean(wrong)
}
```

::: {.callout-note collapse="true"}
### Code
```{r screen-simulation}
#| output: false
n_noise_sims <- 200
screen_labels <- c("Screen once, then cross-validate",
                   "Screen inside each fold")
screen_sims <- map(seq_len(n_noise_sims), \(r) {
  d <- simulate_noise()
  tibble(approach = screen_labels,
         error = c(screen_cv_error(d$X, d$y, inside = FALSE),
                   screen_cv_error(d$X, d$y, inside = TRUE)))
}) |>
  list_rbind() |>
  mutate(approach = factor(approach, levels = screen_labels))
screen_mean <- function(a) mean(screen_sims$error[screen_sims$approach == a])

# Screening inside each fold on further data sets, three ways: folds drawn as
# above; stratified folds, each holding equally many of each class; and class
# labels drawn independently rather than fixed at n / 2 per class.
n_check_sims <- 1000
stratified_fold <- function(y) {
  fold <- integer(length(y))
  for (cl in levels(y)) {
    fold[y == cl] <- sample(rep(seq_len(n_folds), length.out = sum(y == cl)))
  }
  fold
}
check_labels <- c("random", "stratified", "independent")
inside_checks <- map(seq_len(n_check_sims), \(r) {
  d <- simulate_noise()
  y_independent <- factor(sample(levels(d$y), n_noise, replace = TRUE),
                          levels = levels(d$y))
  tibble(folds = check_labels,
         error = c(screen_cv_error(d$X, d$y, inside = TRUE),
                   screen_cv_error(d$X, d$y, inside = TRUE,
                                   fold = stratified_fold(d$y)),
                   screen_cv_error(d$X, y_independent, inside = TRUE)))
}) |>
  list_rbind() |>
  group_by(folds) |>
  summarize(mean = mean(error), se = sd(error) / sqrt(n()), .groups = "drop")
check_of <- function(f, col = "mean") inside_checks[[col]][inside_checks$folds == f]

screen_plot <- ggplot(screen_sims, aes(x = error)) +
  geom_histogram(binwidth = 0.04, boundary = 0, fill = "grey60", color = "white") +
  geom_vline(xintercept = true_error, linetype = "dashed") +
  facet_wrap(~ approach, ncol = 1) +
  scale_x_continuous(labels = \(v) paste0(100 * v, "%"), limits = c(0, 1)) +
  labs(x = "10-fold cross-validation error rate", y = "Simulated data sets",
       title = paste0(n_noise_sims, " simulated data sets; true error rate ",
                      pct(true_error, 0), " (dashed)"))
```
:::

```{r screen-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
screen_plot
```

Across `r n_noise_sims` simulated data sets, screening once and then
cross-validating gives an average estimated error rate of
`r pct(screen_mean(screen_labels[1]))`, against a true error rate of
`r pct(true_error, 0)`. Screening inside each fold gives
`r pct(screen_mean(screen_labels[2]))`. The first approach selected the
features using the responses of the very observations it later held out, so
those observations were not new to the classifier.

Screening is part of the fit. So is every other step that uses the data to
build a classifier or a regression, and a cross-validation estimate is honest
only if all of those steps are repeated, inside every fold, using only the
training folds.

The second estimate sits slightly above `r pct(true_error, 0)`. To see why,
screening inside each fold was cross-validated on
`r format(n_check_sims, big.mark = ",")` further simulated data sets in three
ways. With folds drawn at random as
above, the average estimated error rate is `r pct(check_of("random"))`. With
stratified folds, each holding equally many observations of each class, it is
`r pct(check_of("stratified"))`. With each class label drawn independently,
`r and_list(levels(noise_example$y))` equally likely, so that the class counts
vary from one data set to the next, it is `r pct(check_of("independent"))`.
Each of these averages has a simulation standard error of at most
`r fmt(100 * max(inside_checks$se))` percentage points.

Only the first combines fixed class counts with folds that ignore the class.
Each data set has exactly `r n_noise / 2` observations per class, so a
held-out fold with more of one class leaves the training folds with fewer of
it. A classifier with no information leans toward the class that is more
common in its training folds, which is the less common class in the held-out
fold. Stratified folds remove the imbalance between held-out and training
folds, and independent labels remove the fixed counts that tie them together,
and both estimates are within simulation error of `r pct(true_error, 0)`.

### Standardization and tuning

The same rule covers every data-dependent step.

- **Standardization.** `knn_classify()` computes the feature means and
standard deviations from the training folds and applies them to the held-out
fold. Standardizing the whole data set first lets the held-out observations
influence the scale, the same kind of leak as screening.
- **Tuning.** Choosing $K$ by cross-validation is itself a data-dependent
step. The smallest of `r length(K_grid)` cross-validation error rates,
`r pct(min(knn_errors$"10-fold CV"), 2)`, is the minimum of noisy estimates,
so over repeated data sets it tends to fall below the test error rate of the
classifier it chooses. An honest estimate of the error of "choose $K$ by
`r n_folds`-fold cross-validation, then fit" needs data not used in the
choosing: here the test half, where $K = `r K_cv`$ has error rate
`r pct(knn_of(K_cv, "Test half"), 2)`. On this one split that is
`r if (test_below_cv_min) "below" else "above"` the cross-validation minimum,
`r if (test_below_cv_min) "the opposite of the tendency, because the test curve lies below the cross-validation curve" else "as the tendency predicts"`.

::: {.callout-note collapse="true"}
### Beyond this course
- **One-standard-error rule.** When the cross-validation curve is nearly flat
near its minimum, as for the polynomial degree of the cars, choose the
simplest model whose estimate is within one standard error of the minimum,
the standard error computed from the $k$ fold MSEs.
- **Repeated and stratified $k$-fold.** Averaging $k$-fold estimates over
several random partitions reduces their dependence on the partition;
stratifying the folds keeps the class proportions the same in every fold,
which matters when one class is rare, as defaulters are.
- **Nested cross-validation.** An outer cross-validation loop around the
whole procedure "choose the tuning parameter by inner cross-validation, then
fit" estimates the test error of that procedure without a separate test set.
:::

## Conclusion

Training error falls as flexibility increases, so it cannot choose a model's
flexibility. Resampling methods estimate test error by fitting to part of the
data and measuring error on the part held out.

The validation set approach holds out one random part. Its estimate varies
with the split, and because each model is fit to only part of the data it
tends to overestimate the test error of the model fit to all $n$. Leave-one-out
cross-validation holds out each observation in turn and averages the $n$
squared held-out errors. It involves no random split and is nearly unbiased.
For least squares it needs a single fit, since the held-out error of
observation $i$ is its residual divided by $1 - h_i$.

$k$-fold cross-validation divides the data into $k$ folds, fits $k$ times, and
averages the held-out errors. Choosing $k$ trades bias, variance, and
computation. Smaller $k$ means smaller training sets and so more upward bias;
larger $k$ averages more, but more strongly correlated, fold errors, so its
variance falls more slowly than the number of folds suggests; and each
additional fold is another fit. $k = 5$ or $10$ is the usual choice because
each fit uses 80% or 90% of the data, which keeps the bias modest, while a
method without the least squares shortcut needs only 5 or 10 fits rather than
$n$. For classification the squared
error becomes the misclassification indicator or another loss such as log
loss, and the cross-validation curve chooses a flexibility setting, such as
the number of neighbors in KNN, as it chooses a polynomial degree.

A cross-validation estimate is honest only when every data-dependent step,
including feature screening, standardization, and tuning, is repeated inside
each fold using only the training folds. Selecting features or tuning
parameters on all the data first lets the held-out observations shape the
fit, and the estimate becomes optimistic.

The next lecture turns from estimating test error to estimating the
variability of an estimate: the bootstrap resamples the observations with
replacement to approximate a statistic's sampling distribution.
