# Bootstrap

```{r packages}
#| include: false
library("tidyverse")
library("Sleuth3")
theme_set(theme_bw())
# The date this file was created. Every random draw below (the jitter, the
# bootstrap samples, and the simulated experiments) follows from this seed.
set.seed(20261008)

treatment_colors <- c("Unseeded" = "#0072B2", "Seeded" = "#D55E00")
treatment_shapes <- c("Unseeded" = 16, "Seeded" = 17)
interval_colors  <- c("Percentile" = "#009E73",
                      "Normal approximation" = "#CC79A7")
interval_lines   <- c("Percentile" = "dashed",
                      "Normal approximation" = "dotdash")

# Rounds and adds thousands separators for inline numbers.
fmt <- function(x, digits = 0) {
  format(round(x, digits), big.mark = ",", nsmall = digits, scientific = FALSE,
         trim = TRUE)
}
# Axis labels such as 1,000 rather than 1e+03.
comma_labels <- function(v) {
  ifelse(is.na(v), NA, format(v, big.mark = ",", scientific = FALSE,
                              trim = TRUE, drop0trailing = TRUE))
}
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **explain** the bootstrap principle, that drawing $n$ observations with
replacement from the sample imitates drawing new samples from the population,
and **compute** a bootstrap standard error of a statistic with the bootstrap
algorithm,
- **compare** a bootstrap standard error with a formula-based standard error
when one exists, such as the standard error of a regression coefficient from
`lm()`, and **explain** why a bootstrap standard error is itself an estimate
whose accuracy depends on the statistic and the sample size,
- **construct** a percentile interval and a normal-approximation interval from
a bootstrap distribution, and **explain** why the two differ when the
bootstrap distribution is skewed or not centered at the estimate,
- **derive** the probability $1 - (1 - 1/n)^n \approx 1 - 1/e \approx 0.632$
that a given observation appears in a bootstrap sample, and **explain** why it
makes the bootstrap a poor estimator of test error, and
- **identify** settings in which bootstrap standard errors and intervals are
unreliable: small samples, heavy-tailed or skewed populations summarized by a
mean-type statistic, extreme statistics such as the maximum, and dependent
data.
:::

## Cloud seeding

### Data

**Question.** Does seeding clouds with silver iodide increase rainfall, and by
what factor?

**Data.** `Sleuth3::case0301` records the volume of rainfall, in acre-feet, in
a target area on `r nrow(case0301)` days suitable for cloud seeding. In a
randomized experiment, the clouds were seeded on
`r sum(case0301$Treatment == "Seeded")` of the days and left unseeded on the
other `r sum(case0301$Treatment == "Unseeded")`. Each day is one observation.

The table gives each treatment's number of days and the mean, geometric mean,
median, standard deviation, minimum, and maximum of its rainfall, where the
geometric mean is the exponential of the mean of log rainfall. The figure
plots each day's rainfall by treatment, with segments at each treatment's
mean (solid), geometric mean (dotted), and median (dashed); the second tab
draws rainfall on a log scale.

::: {.callout-note collapse="true"}
### Code
```{r cloud-data}
#| output: false
clouds <- case0301 |>
  mutate(Treatment = factor(Treatment, levels = c("Unseeded", "Seeded")),
         seeded = as.integer(Treatment == "Seeded"),
         # Horizontal jitter, computed once so that every figure below puts
         # each day in the same place.
         x_jit = as.integer(Treatment) + runif(n(), -0.15, 0.15))
n <- nrow(clouds)

cloud_summary <- clouds |>
  group_by(Treatment) |>
  summarize(days = n(), mean = mean(Rainfall),
            geo_mean = exp(mean(log(Rainfall))), median = median(Rainfall),
            sd = sd(Rainfall), min = min(Rainfall), max = max(Rainfall),
            sd_log = sd(log(Rainfall)), .groups = "drop")
summary_of <- function(trt, col) cloud_summary[[col]][cloud_summary$Treatment == trt]

cloud_table <- cloud_summary |>
  transmute(Treatment, days,
            across(c(mean, geo_mean, median, sd, min, max), \(v) fmt(v, 1))) |>
  knitr::kable(col.names = c("Treatment", "Days", "Mean", "Geometric mean",
                             "Median", "SD", "Min", "Max"),
               align = "lccccccc")

center_types <- c("Mean", "Geometric mean", "Median")
center_lines <- cloud_summary |>
  select(Treatment, Mean = mean, `Geometric mean` = geo_mean,
         Median = median) |>
  pivot_longer(-Treatment, names_to = "summary", values_to = "rainfall") |>
  mutate(x = as.integer(Treatment),
         summary = factor(summary, levels = center_types))

eda_plot <- ggplot(clouds, aes(x = x_jit, y = Rainfall)) +
  geom_point(aes(color = Treatment, shape = Treatment), size = 2) +
  geom_segment(data = center_lines,
               aes(x = x - 0.3, xend = x + 0.3, y = rainfall, yend = rainfall,
                   linetype = summary)) +
  scale_color_manual(values = treatment_colors) +
  scale_shape_manual(values = treatment_shapes) +
  scale_linetype_manual(values = c("Mean" = "solid",
                                   "Geometric mean" = "dotted",
                                   "Median" = "dashed")) +
  scale_x_continuous(breaks = 1:2, labels = levels(clouds$Treatment)) +
  labs(x = NULL, y = "Rainfall (acre-feet)", linetype = NULL,
       title = "Sleuth3::case0301, rainfall by treatment")

eda_plot_log <- eda_plot +
  scale_y_log10(labels = comma_labels) +
  labs(y = "Rainfall (acre-feet, log scale)")
```
:::

```{r cloud-table}
#| echo: false
cloud_table
```

::: {.panel-tabset}
#### Acre-feet
```{r cloud-eda-plot}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4
eda_plot
```

#### Log scale
```{r cloud-eda-plot-log}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4
eda_plot_log
```
:::

Both groups are strongly right-skewed: a few days with very large rainfall
pull each mean far above its median, and the largest seeded day,
`r fmt(summary_of("Seeded", "max"), 1)` acre-feet, is more than
`r floor(summary_of("Seeded", "max") / summary_of("Seeded", "median"))` times
the seeded median. On the log scale the two groups are roughly symmetric with
similar spread (standard deviations of log rainfall
`r fmt(summary_of("Unseeded", "sd_log"), 2)` unseeded and
`r fmt(summary_of("Seeded", "sd_log"), 2)` seeded), and on that axis each
geometric mean lies close to its median. Seeded days tend to have more
rainfall than unseeded days, though the two groups overlap heavily.

### Multiplicative effect

The log-scale tab suggests modeling log rainfall as normal with a mean that
depends on treatment,

$$\log Y_i = \beta_0 + \beta_1 D_i + \epsilon_i, \qquad \epsilon_i \stackrel{ind}{\sim} N\left(0, \sigma^2\right),$$

where

- $Y_i$ is the rainfall on day $i$, in acre-feet,
- $D_i = \mathrm{I}(\text{day } i \text{ seeded})$ is the indicator of
seeding,
- $\beta_0$ is the mean log rainfall on unseeded days,
- $\beta_1$ is the difference in mean log rainfall, seeded minus unseeded,
and
- $\sigma^2$ is the variance of log rainfall about its treatment mean.

Adding $\beta_1$ to log rainfall multiplies rainfall by $e^{\beta_1}$, so the
effect of seeding is the factor

$$\theta = e^{\beta_1}, \qquad \hat\theta = e^{\hat\beta_1} = \frac{\tilde y_S}{\tilde y_U},$$

where

- $\theta$ is the factor by which seeding multiplies median rainfall, since
under the model the median of $Y_i$ is $e^{\beta_0 + \beta_1 D_i}$,
- $\hat\beta_1 = \overline{\log y}_S - \overline{\log y}_U$ is the least
squares estimate of $\beta_1$, the difference between the sample means of log
rainfall on seeded and unseeded days, and
- $\tilde y_S = \exp\left(\overline{\log y}_S\right)$ and
$\tilde y_U = \exp\left(\overline{\log y}_U\right)$ are the sample geometric
means of rainfall on seeded and unseeded days, so that $\hat\theta$ is their
ratio.

Throughout this chapter $\theta$ is the quantity whose standard error is
wanted and $\hat\theta$ its estimate; for the cloud data,
$\theta = e^{\beta_1}$.

```{r cloud-lm}
cloud_lm <- lm(log(Rainfall) ~ seeded, data = clouds)
summary(cloud_lm)$coefficients
theta_hat <- exp(coef(cloud_lm)[["seeded"]])
theta_hat
```

```{r cloud-lm-values}
#| include: false
b_lm  <- coef(cloud_lm)
se_lm <- summary(cloud_lm)$coefficients[, "Std. Error"]
ci_lm <- confint(cloud_lm)["seeded", ]
```

The estimated mean log rainfall on unseeded days is
$\hat\beta_0 = `r fmt(b_lm[["(Intercept)"]], 2)`$, a geometric mean of
$e^{\hat\beta_0} = `r fmt(exp(b_lm[["(Intercept)"]]), 1)`$ acre-feet. Seeded
days averaged $\hat\beta_1 = `r fmt(b_lm[["seeded"]], 2)`$ more log rainfall,
with standard error `r fmt(se_lm[["seeded"]], 2)`, so seeding multiplied the
geometric mean rainfall by $\hat\theta = `r fmt(theta_hat, 2)`$. `lm()`
reports a standard error for $\hat\beta_1$ but none for $\hat\theta$, the
number that answers the question.

## Standard error

### Sampling distribution

Repeating the experiment on `r n` new days would give different rainfall and
a different $\hat\theta$. Before the data are collected, $\hat\theta$ is a
random variable, and its distribution over repeated samples from the
population is its **sampling distribution**. The **standard error** of
$\hat\theta$ is the standard deviation of that distribution,

$$\text{SE}(\hat\theta) = \sqrt{Var\left[\hat\theta\right]}.$$

Simulating the sampling distribution would mean running the experiment again
and again, which requires access to the population.

### Formulas

For some statistics, probability theory gives the standard error as a
function of population quantities that the sample can estimate. The sample
mean of $n$ independent observations with standard deviation $\sigma$ has
$\text{SE}(\bar y) = \sigma/\sqrt n$, estimated by $s/\sqrt n$, and `lm()`
reports the standard error of each regression coefficient. The estimate
$\hat\theta = e^{\hat\beta_1}$ is a nonlinear function of a coefficient, and
no column of the `lm()` output gives its standard error. A median, a
quantile, and the AUC of a fitted classifier have no simple standard-error
formula either, and a statistic computed by an algorithm rather than a
formula often has none at all.

## Bootstrap

### Resampling

The first tab below repeats the `r n` days on the log scale. Each later tab
draws `r n` days *with replacement* from those `r n`, so that a day can be
drawn more than once or not at all. Filled points are days that were drawn,
larger the more times they were drawn, open grey points are days that were
not drawn, and the solid segments mark each treatment's geometric mean
rainfall in that sample. Each title gives the sample's ratio of geometric
means.

::: {.callout-note collapse="true"}
### Code
```{r resample-plots}
#| output: false
# Draw three illustrative bootstrap samples and record how many times each
# day was drawn.
illustrative_counts <- map(1:3, \(b) tabulate(sample(n, replace = TRUE),
                                              nbins = n))
# One size scale for every tab, so a point's size means the same thing in each.
max_count <- max(unlist(illustrative_counts))

resample_plot <- function(counts, title) {
  d <- clouds |> mutate(count = counts)
  drawn   <- filter(d, count > 0)
  undrawn <- filter(d, count == 0)
  group_geo_means <- drawn |>
    group_by(Treatment) |>
    summarize(rainfall = exp(weighted.mean(log(Rainfall), count)),
              .groups = "drop") |>
    mutate(x = as.integer(Treatment))
  ratio <- group_geo_means$rainfall[2] / group_geo_means$rainfall[1]

  ggplot(drawn, aes(x = x_jit, y = Rainfall)) +
    geom_point(data = undrawn, color = "grey60",
               shape = ifelse(undrawn$Treatment == "Seeded", 2, 1), size = 2) +
    geom_point(aes(color = Treatment, shape = Treatment, size = count)) +
    geom_segment(data = group_geo_means,
                 aes(x = x - 0.3, xend = x + 0.3, y = rainfall,
                     yend = rainfall)) +
    scale_color_manual(values = treatment_colors) +
    scale_shape_manual(values = treatment_shapes) +
    scale_size_area(max_size = 5, breaks = 1:max_count,
                    limits = c(0, max_count)) +
    scale_x_continuous(breaks = 1:2, labels = levels(clouds$Treatment),
                       limits = c(0.6, 2.4)) +
    scale_y_log10(labels = comma_labels) +
    labs(x = NULL, y = "Rainfall (acre-feet, log scale)",
         size = "Times drawn",
         title = sprintf("%s: ratio of geometric means = %.2f", title, ratio))
}

original_plot <- resample_plot(rep(1, n), "Original sample")
resample_plots <- imap(illustrative_counts,
                       \(cnt, b) resample_plot(cnt, paste("Bootstrap sample", b)))
n_distinct_illustrative <- map_int(illustrative_counts, \(cnt) sum(cnt > 0))
```
:::

::: {.panel-tabset}
#### Original sample
```{r resample-original}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4
original_plot
```

#### Bootstrap sample 1
```{r resample-1}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4
resample_plots[[1]]
```

#### Bootstrap sample 2
```{r resample-2}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4
resample_plots[[2]]
```

#### Bootstrap sample 3
```{r resample-3}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4
resample_plots[[3]]
```
:::

The three resamples contain
`r n_distinct_illustrative[1]`, `r n_distinct_illustrative[2]`, and
`r n_distinct_illustrative[3]` of the `r n` days, each with some days
repeated, and their ratios of geometric means differ from one another and
from $\hat\theta = `r fmt(theta_hat, 2)`$.

This is the **bootstrap**: it treats the sample as if it were the population.
The population of all suitable days is unavailable, but the `r n` observed
days are, and drawing `r n` days with replacement from them imitates running
the experiment again. Each resample is a **bootstrap sample**, and how much
the statistic varies across bootstrap samples estimates how much $\hat\theta$
varies across experiments.

### Algorithm

Write the data set of $n$ observations as $Z$ (here `r n` days, each a
rainfall and a treatment). The bootstrap standard error of a statistic
$\hat\theta$ computed from $Z$ is found in three steps.

1. For $b = 1, \ldots, B$, draw a bootstrap sample $Z^{*b}$ of $n$
observations from $Z$ with replacement, each draw choosing each observation
with probability $1/n$.
2. Compute the statistic on each bootstrap sample, exactly as $\hat\theta$
was computed on $Z$, giving $\hat\theta^{*1}, \ldots, \hat\theta^{*B}$.
3. Take their standard deviation,

$$\text{SE}_B(\hat\theta) = \sqrt{\frac{1}{B-1}\sum_{b=1}^B \left(\hat\theta^{*b} - \bar\theta^{*}\right)^2}, \qquad \bar\theta^{*} = \frac{1}{B}\sum_{b=1}^B \hat\theta^{*b},$$

where

- $B$ is the number of bootstrap samples, chosen by the analyst (a few
thousand is typical),
- $\hat\theta^{*b}$ is the statistic computed on bootstrap sample $b$,
- $\bar\theta^{*}$ is the mean of the $B$ bootstrap statistics, and
- $\text{SE}_B(\hat\theta)$ is the **bootstrap standard error** of
$\hat\theta$.

The $B$ values $\hat\theta^{*b}$ form the **bootstrap distribution** of
$\hat\theta$. In R, step 1 is `sample(n, replace = TRUE)`, which returns the
row numbers of one bootstrap sample. It draws the `r n` days without regard
to treatment, so the numbers of seeded and unseeded days vary from one
bootstrap sample to the next. Step 2 refits the regression and
exponentiates its slope; the loop also keeps the slope $\hat\beta_1^{*b}$
itself, used in the next subsection.

```{r bootstrap-loop}
B <- 2000
boot <- tibble(b = 1:B) |>
  mutate(
    i     = map(b, \(b) sample(n, replace = TRUE)),                 # step 1
    slope = map_dbl(i, \(i) coef(lm(log(Rainfall) ~ seeded,
                                    data = clouds[i, ]))[["seeded"]]),
    theta = exp(slope)                                              # step 2
  )
se_boot_theta <- sd(boot$theta)                                     # step 3
se_boot_theta
```

The histogram shows the `r fmt(B)` values of $\hat\theta^{*b}$, with a solid
line at $\hat\theta$.

::: {.callout-note collapse="true"}
### Code
```{r boot-hist}
#| output: false
boot_hist_plot <- ggplot(boot, aes(x = theta)) +
  geom_histogram(bins = 50, fill = "grey75", color = "grey45") +
  geom_vline(xintercept = theta_hat, linewidth = 0.9) +
  labs(x = expression("Bootstrap ratio of geometric means, " *
                        hat(theta)^"*b"),
       y = "Bootstrap samples",
       title = "Bootstrap distribution of the ratio of geometric means")
boot_theta_range <- range(boot$theta)
```
:::

```{r boot-hist-figure}
#| echo: false
#| fig-width: 6.5
#| fig-height: 3.8
boot_hist_plot
```

The bootstrap distribution is right-skewed, running from
`r fmt(boot_theta_range[1], 2)` to `r fmt(boot_theta_range[2], 2)`. Its
standard deviation is the bootstrap standard error,
$\text{SE}_B(\hat\theta) = `r fmt(se_boot_theta, 2)`$.

### Regression slope

Where a formula exists, the bootstrap should agree with it. The slope
$\hat\beta_1$ has the standard error `lm()` reported, and the loop above
recorded $\hat\beta_1^{*b}$ on each bootstrap sample. The histogram shows
those `r fmt(B)` bootstrap slopes on the density scale, with the normal
density centered at $\hat\beta_1$ with the `lm()` standard error drawn as a
dashed curve.

::: {.callout-note collapse="true"}
### Code
```{r slope-check}
#| output: false
se_boot_slope <- sd(boot$slope)
slope_grid <- tibble(slope = seq(min(boot$slope), max(boot$slope),
                                 length.out = 400)) |>
  mutate(density = dnorm(slope, b_lm[["seeded"]], se_lm[["seeded"]]))

slope_plot <- ggplot(boot, aes(x = slope)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = "grey75", color = "grey45") +
  geom_line(data = slope_grid, aes(y = density), linetype = "dashed",
            linewidth = 0.9) +
  geom_vline(xintercept = b_lm[["seeded"]], linewidth = 0.9) +
  labs(x = expression("Bootstrap slope, " * hat(beta)[1]^"*b" *
                        " (difference in mean log rainfall)"),
       y = "Density",
       title = "Bootstrap slopes and the normal density from lm()")
```
:::

```{r slope-plot}
#| echo: false
#| fig-width: 6.5
#| fig-height: 3.8
slope_plot
```

The bootstrap standard error of the slope is
$\text{SE}_B(\hat\beta_1) = `r fmt(se_boot_slope, 3)`$, against
`r fmt(se_lm[["seeded"]], 3)` from `lm()`, and the histogram follows the
normal curve. Each $\hat\theta^{*b}$ is $e^{\hat\beta_1^{*b}}$, so the
right skew of the $\hat\theta^{*b}$ histogram is this bell shape passed
through the exponential function.

### Known population

Whether the bootstrap distribution of $\hat\theta$ imitates its sampling
distribution can only be checked where the sampling distribution is known,
so this subsection simulates experiments from a known population: the
log-scale regression model with $\beta_0$, $\beta_1$, and $\sigma$ set to
the cloud data's estimates
$\hat\beta_0 = `r fmt(b_lm[["(Intercept)"]], 2)`$,
$\hat\beta_1 = `r fmt(b_lm[["seeded"]], 2)`$, and
$\hat\sigma = `r fmt(sigma(cloud_lm), 2)`$. For this population
$\theta = e^{\beta_1} = `r fmt(theta_hat, 2)`$.

The simulation runs the experiment, `r sum(clouds$seeded == 0)` unseeded and
`r sum(clouds$seeded == 1)` seeded days, many times to obtain the sampling
distribution of $\hat\theta$. It then runs one more experiment, treats it as
the only sample available, and bootstraps it `r fmt(B)` times, resampling all
its days together as for the cloud data. The first tab shows the sampling
distribution, with a solid line at $\theta$; the second shows the bootstrap
distribution from the one experiment, with a solid line at that experiment's
$\hat\theta$. Both use the same horizontal axis, cut at the 99.9th percentile
of the simulated values.

::: {.callout-note collapse="true"}
### Code
```{r known-population}
#| output: false
pop <- list(beta0 = b_lm[["(Intercept)"]], beta1 = b_lm[["seeded"]],
            sigma = sigma(cloud_lm))
theta_pop <- exp(pop$beta1)

# R simulated experiments with m days per treatment, as an m-row matrix of
# unseeded log rainfall and an m-row matrix of seeded log rainfall, one column
# per experiment.
draw_experiments <- function(m, R) {
  list(U = matrix(rnorm(m * R, pop$beta0, pop$sigma), m, R),
       S = matrix(rnorm(m * R, pop$beta0 + pop$beta1, pop$sigma), m, R))
}
# Every statistic in this chapter, for each column (experiment): the slope
# (difference in mean log rainfall, the least squares slope), theta-hat, and
# the ratio of mean rainfall (used under Limitations).
experiment_stats <- function(x) {
  slope <- colMeans(x$S) - colMeans(x$U)
  tibble(slope = slope, theta = exp(slope),
         ratio = colMeans(exp(x$S)) / colMeans(exp(x$U)))
}
# B bootstrap samples of one data set (rainfall y, seeding indicator s), all at
# once: column b of idx holds the row numbers of bootstrap sample b, drawn from
# all the days together as in the cloud bootstrap.
bootstrap_stats <- function(y, s, B) {
  N <- length(y)
  idx <- matrix(sample.int(N, N * B, replace = TRUE), N, B)
  Y <- matrix(y[idx], N, B)
  S <- matrix(s[idx], N, B)
  n_S <- colSums(S)
  slope <- colSums(log(Y) * S) / n_S - colSums(log(Y) * (1 - S)) / (N - n_S)
  tibble(slope = slope, theta = exp(slope),
         ratio = (colSums(Y * S) / n_S) / (colSums(Y * (1 - S)) / (N - n_S)))
}

# Statistics for R experiments, simulated in batches of 10,000 to limit memory.
simulate_stats <- function(m, R, batch = 10000) {
  map(seq_len(R / batch), \(k) experiment_stats(draw_experiments(m, batch))) |>
    list_rbind()
}

# The ratio of means has a heavy-tailed sampling distribution, so its standard
# deviation needs many experiments to pin down: 10,000 can be off by over 10%.
n_experiments <- 100000
m_cloud <- sum(clouds$seeded == 1)   # 26 days per treatment
sampling <- simulate_stats(m_cloud, n_experiments)
se_true <- map_dbl(sampling, sd)
# se_true is itself a simulation estimate. Its Monte Carlo standard error, from
# the spread of the standard deviations of 20 equal batches of experiments, is
# reported relative to se_true.
se_true_mcse <- map_dbl(sampling, \(v) {
  sd(tapply(v, rep(1:20, each = n_experiments / 20), sd)) / sqrt(20) / sd(v)
})

one_experiment <- draw_experiments(m_cloud, 1)
one_y <- exp(c(one_experiment$U, one_experiment$S))
one_s <- rep(0:1, each = m_cloud)
theta_hat_one <- experiment_stats(one_experiment)$theta
one_boot      <- bootstrap_stats(one_y, one_s, B)$theta
se_boot_one   <- sd(one_boot)

# A shared axis from 0 to the 99.9th percentile of the simulated values; the
# few experiments beyond it would otherwise squeeze the histograms flat.
x_limits <- c(0, quantile(c(sampling$theta, one_boot), 0.999))
known_hist <- function(values, line_at, title) {
  ggplot(tibble(value = values), aes(x = value)) +
    geom_histogram(aes(y = after_stat(density)), boundary = 0,
                   binwidth = diff(x_limits) / 60,
                   fill = "grey75", color = "grey45") +
    geom_vline(xintercept = line_at, linewidth = 0.9) +
    coord_cartesian(xlim = x_limits) +
    labs(x = "Ratio of geometric means", y = "Density", title = title)
}
sampling_plot <- known_hist(
  sampling$theta, theta_pop,
  sprintf("Sampling distribution: %s experiments, SE = %.2f",
          fmt(n_experiments), se_true[["theta"]]))
one_boot_plot <- known_hist(
  one_boot, theta_hat_one,
  sprintf("Bootstrap distribution: one experiment, SE_B = %.2f", se_boot_one))
```
:::

::: {.panel-tabset}
#### Sampling distribution
```{r sampling-plot}
#| echo: false
#| fig-width: 6.5
#| fig-height: 3.8
sampling_plot
```

#### Bootstrap distribution
```{r one-boot-plot}
#| echo: false
#| fig-width: 6.5
#| fig-height: 3.8
one_boot_plot
```
:::

Both distributions are right-skewed. Over `r fmt(n_experiments)`
experiments, $\hat\theta$ has standard error `r fmt(se_true[["theta"]], 2)`
(a simulation estimate, accurate to about
$\pm$`r fmt(200 * se_true_mcse[["theta"]])`%, two Monte Carlo standard
errors). The bootstrap distribution is
centered at that experiment's $\hat\theta = `r fmt(theta_hat_one, 2)`$, not
at $\theta = `r fmt(theta_pop, 2)`$: it estimates the spread and shape of the
sampling distribution, not the location of $\theta$. Its standard deviation,
$\text{SE}_B(\hat\theta) = `r fmt(se_boot_one, 2)`$, is
`r fmt(se_boot_one / se_true[["theta"]], 2)` times the true standard error,
and that experiment's $\hat\theta$ is
`r fmt(theta_hat_one / theta_pop, 2)` times $\theta$. The next subsection
shows that these two ratios tend to move together.

### Accuracy

```{r accuracy-count}
#| include: false
n_check <- 500   # simulated experiments bootstrapped
B_check <- 1000  # bootstrap samples for each
```

A bootstrap standard error is itself an estimate, computed from one sample,
and a different sample gives a different value. To see how far it typically
lands from the truth, the simulation bootstraps `r n_check` further
experiments of `r m_cloud` days per treatment, `r fmt(B_check)` times
each, and divides each experiment's $\text{SE}_B$ by the true standard error.
The first tab shows these ratios for the slope $\hat\beta_1$ and for
$\hat\theta$; the second plots, for each experiment, the ratio for
$\hat\theta$ against how far that experiment's $\hat\theta$ fell from
$\theta$.

::: {.callout-note collapse="true"}
### Code
```{r accuracy}
#| output: false
# Bootstrap n_check simulated experiments with m days per treatment. Each row
# holds one experiment's estimates, its bootstrap SEs, and its 95% percentile
# and normal-approximation intervals for theta.
check_experiments <- function(m, n_check, B_check) {
  map(1:n_check, \(r) {
    x <- draw_experiments(m, 1)
    est <- experiment_stats(x)
    bs <- bootstrap_stats(exp(c(x$U, x$S)), rep(0:1, each = m), B_check)
    tibble(m = m, theta_hat = est$theta,
           se_b_slope = sd(bs$slope), se_b_theta = sd(bs$theta),
           se_b_ratio = sd(bs$ratio),
           perc_lo = quantile(bs$theta, 0.025),
           perc_hi = quantile(bs$theta, 0.975),
           norm_lo = est$theta - qnorm(0.975) * sd(bs$theta),
           norm_hi = est$theta + qnorm(0.975) * sd(bs$theta))
  }) |> list_rbind()
}
checks <- check_experiments(m_cloud, n_check, B_check) |>
  mutate(rel_slope = se_b_slope / se_true[["slope"]],
         rel_theta = se_b_theta / se_true[["theta"]])

# Typical relative error: the median of |SE_B / SE - 1| over experiments.
typical_error <- c(slope = median(abs(checks$rel_slope - 1)),
                   theta = median(abs(checks$rel_theta - 1)))
median_rel    <- c(slope = median(checks$rel_slope),
                   theta = median(checks$rel_theta))
slope_within  <- mean(abs(checks$rel_slope - 1) <= 0.10)

accuracy_labels <- c(rel_slope = "Slope, difference in mean log rainfall",
                     rel_theta = "Ratio of geometric means")
accuracy_long <- checks |>
  select(rel_slope, rel_theta) |>
  pivot_longer(everything(), names_to = "stat", values_to = "relative") |>
  mutate(stat = factor(accuracy_labels[stat], levels = accuracy_labels))
accuracy_plot <- ggplot(accuracy_long, aes(x = relative)) +
  geom_histogram(bins = 40, fill = "grey75", color = "grey45") +
  geom_vline(xintercept = 1, linewidth = 0.9) +
  scale_x_log10() +
  facet_wrap(~ stat, ncol = 1) +
  labs(x = "Bootstrap SE divided by true SE (log scale)",
       y = "Simulated experiments",
       title = sprintf("%d simulated experiments, %d days per treatment",
                       n_check, m_cloud))

accuracy_scatter <- ggplot(checks, aes(x = theta_hat / theta_pop,
                                       y = rel_theta)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = expression("Estimate divided by truth, " * hat(theta) / theta *
                        " (log scale)"),
       y = "Bootstrap SE divided by true SE (log scale)",
       title = "Bootstrap SE of the ratio of geometric means")
scatter_cor <- cor(log(checks$theta_hat), log(checks$rel_theta))

coverage <- checks |>
  summarize(perc = mean(perc_lo < theta_pop & theta_pop < perc_hi),
            perc_above = mean(perc_hi < theta_pop),
            perc_below = mean(perc_lo > theta_pop),
            norm = mean(norm_lo < theta_pop & theta_pop < norm_hi),
            norm_above = mean(norm_hi < theta_pop),
            norm_below = mean(norm_lo > theta_pop),
            norm_negative = mean(norm_lo < 0))
```
:::

::: {.panel-tabset}
#### SE accuracy
```{r accuracy-plot}
#| echo: false
#| fig-width: 6.5
#| fig-height: 5
accuracy_plot
```

#### SE and estimate
```{r accuracy-scatter}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4
accuracy_scatter
```
:::

For the slope, $\text{SE}_B$ is typically within
`r fmt(100 * typical_error[["slope"]])`% of the true standard error (the
median of $|\text{SE}_B/\text{SE} - 1|$ over the experiments), and within
10% in `r fmt(100 * slope_within)`% of them. For $\hat\theta$ the median
ratio is `r fmt(median_rel[["theta"]], 2)`, so $\text{SE}_B$ is not
systematically too small or too large, but its typical error is
`r fmt(100 * typical_error[["theta"]])`%.

The second tab shows where that error comes from: experiments whose
$\hat\theta$ overshoots $\theta$ also overstate its standard error
(correlation `r fmt(scatter_cor, 2)` on the log scales). Each bootstrap
statistic is
$\hat\theta^{*b} = \hat\theta\, e^{\hat\beta_1^{*b} - \hat\beta_1}$, so

$$\text{SE}_B(\hat\theta) = \hat\theta \times \text{SD}_B\left(e^{\hat\beta_1^{*b} - \hat\beta_1}\right),$$

where

- $\text{SD}_B(\cdot)$ is the standard deviation over the $B$ bootstrap
samples, and
- the second factor depends only on the spread of the bootstrap slopes, which
the bootstrap estimates as well as it estimates
$\text{SE}(\hat\beta_1)$.

The first factor is $\hat\theta$ itself. The standard error of $\hat\theta$
is proportional to $\theta$, which is known only through $\hat\theta$, so no
estimate of $\text{SE}(\hat\theta)$, bootstrap or formula, can be much more
accurate than $\hat\theta$. For the cloud data,
$\text{SE}_B(\hat\theta) = `r fmt(se_boot_theta, 2)`$ is best read as
"about `r fmt(se_boot_theta, 1)`, give or take
`r fmt(100 * typical_error[["theta"]])`%."

## Confidence intervals

A standard error describes the spread of a statistic; a confidence interval
uses the whole bootstrap distribution. The first tab shows the bootstrap
distributions of the slope, $\hat\beta_1^{*b}$, and the ratio of geometric
means, $\hat\theta^{*b}$, each with a solid line at the original estimate.
The second tab adds each statistic's 95% percentile interval as dashed lines,
and the third adds its 95% normal-approximation interval as dot-dashed lines;
both intervals are defined below the figure.

::: {.callout-note collapse="true"}
### Code
```{r intervals}
#| output: false
alpha <- 0.05
z <- qnorm(1 - alpha / 2)
stat_labels <- c(slope = "Slope, difference in mean log rainfall",
                 theta = "Ratio of geometric means")
estimates <- c(slope = b_lm[["seeded"]], theta = theta_hat)
se_boot   <- c(slope = se_boot_slope, theta = se_boot_theta)

intervals <- tibble(stat = names(estimates), estimate = estimates,
                    se = se_boot) |>
  mutate(perc_lo = map_dbl(stat, \(s) quantile(boot[[s]], alpha / 2)),
         perc_hi = map_dbl(stat, \(s) quantile(boot[[s]], 1 - alpha / 2)),
         norm_lo = estimate - z * se,
         norm_hi = estimate + z * se,
         # The t interval for the slope from lm(), and its exponential for
         # theta.
         lm_lo = if_else(stat == "slope", ci_lm[[1]], exp(ci_lm[[1]])),
         lm_hi = if_else(stat == "slope", ci_lm[[2]], exp(ci_lm[[2]])))
interval_of <- function(s, col) intervals[[col]][intervals$stat == s]

boot_long <- boot |>
  select(slope, theta) |>
  pivot_longer(everything(), names_to = "stat") |>
  mutate(stat = factor(stat_labels[stat], levels = stat_labels))
estimate_lines <- tibble(stat = factor(stat_labels[names(estimates)],
                                       levels = stat_labels),
                         value = estimates)
interval_lines_df <- intervals |>
  pivot_longer(c(perc_lo, perc_hi, norm_lo, norm_hi),
               names_to = c("method", "end"), names_sep = "_") |>
  mutate(method = if_else(method == "perc", "Percentile",
                          "Normal approximation"),
         method = factor(method, levels = names(interval_colors)),
         stat = factor(stat_labels[stat], levels = stat_labels))

interval_base <- ggplot(boot_long, aes(x = value)) +
  geom_histogram(bins = 50, fill = "grey75", color = "grey45") +
  geom_vline(data = estimate_lines, aes(xintercept = value),
             linewidth = 0.9) +
  facet_wrap(~ stat, scales = "free") +
  labs(x = "Bootstrap statistic", y = "Bootstrap samples")
interval_layer <- function(methods) {
  list(
    geom_vline(data = filter(interval_lines_df, method %in% methods),
               aes(xintercept = value, color = method, linetype = method),
               linewidth = 0.9, key_glyph = "path"),
    scale_color_manual(values = interval_colors, drop = FALSE),
    scale_linetype_manual(values = interval_lines, drop = FALSE),
    labs(color = "95% interval", linetype = "95% interval"),
    theme(legend.position = "bottom")
  )
}
interval_plot_0 <- interval_base
interval_plot_1 <- interval_base + interval_layer("Percentile")
interval_plot_2 <- interval_base + interval_layer(names(interval_colors))

interval_text <- function(lo, hi) sprintf("(%s, %s)", fmt(lo, 2), fmt(hi, 2))
interval_table <- intervals |>
  mutate(Statistic = stat_labels[stat],
         Estimate = fmt(estimate, 2),
         `Bootstrap SE` = fmt(se, 2),
         Percentile = interval_text(perc_lo, perc_hi),
         `Normal approximation` = interval_text(norm_lo, norm_hi),
         `From lm()` = interval_text(lm_lo, lm_hi)) |>
  select(Statistic, Estimate, `Bootstrap SE`, Percentile,
         `Normal approximation`, `From lm()`) |>
  knitr::kable(align = "lccccc")
```
:::

::: {.panel-tabset}
#### Bootstrap distributions
```{r interval-plot-0}
#| echo: false
#| fig-width: 8
#| fig-height: 3.8
interval_plot_0
```

#### + Percentile
```{r interval-plot-1}
#| echo: false
#| fig-width: 8
#| fig-height: 4.2
interval_plot_1
```

#### + Normal approximation
```{r interval-plot-2}
#| echo: false
#| fig-width: 8
#| fig-height: 4.2
interval_plot_2
```
:::

### Percentile interval

If some monotone transformation of $\hat\theta$ has an approximately normal
sampling distribution, as $\log\hat\theta = \hat\beta_1$ does here, the
middle $1 - \alpha$ of the bootstrap distribution gives a range of plausible
values of $\theta$, without the transformation having to be known. The $(1 - \alpha)100\%$ **percentile interval** is

$$\left(\hat\theta^{*}_{(\alpha/2)},\ \hat\theta^{*}_{(1-\alpha/2)}\right),$$

where $\hat\theta^{*}_{(q)}$ is the $q$ quantile of the $B$ bootstrap
statistics $\hat\theta^{*1}, \ldots, \hat\theta^{*B}$. In R it is
`quantile(boot$theta, c(0.025, 0.975))` for a 95% interval.

For the ratio of geometric means the 95% percentile interval is
(`r fmt(interval_of("theta", "perc_lo"), 2)`,
`r fmt(interval_of("theta", "perc_hi"), 2)`).

::: {.callout-note collapse="true"}
### For example,
A percentile interval moves with a monotone transformation of the statistic.
Since $\hat\theta^{*b} = e^{\hat\beta_1^{*b}}$, the quantiles of the
$\hat\theta^{*b}$ are the exponentials of the quantiles of the
$\hat\beta_1^{*b}$. The 95% percentile interval for $\beta_1$ is
(`r fmt(interval_of("slope", "perc_lo"), 3)`,
`r fmt(interval_of("slope", "perc_hi"), 3)`), and its exponential,
(`r fmt(exp(interval_of("slope", "perc_lo")), 2)`,
`r fmt(exp(interval_of("slope", "perc_hi")), 2)`), is the percentile interval
for $\theta$ above. A normal-approximation interval does not move this way:
exponentiating the one for $\beta_1$ does not give the one for $\theta$.
:::

### Normal-approximation interval

If the sampling distribution of $\hat\theta$ is approximately normal, the
usual form of interval applies with the bootstrap standard error,

$$\hat\theta \pm z_{1-\alpha/2}\,\text{SE}_B(\hat\theta),$$

where $z_{1-\alpha/2}$ is the $(1 - \alpha/2)$ quantile of the standard normal
distribution, `r fmt(z, 2)` for a 95% interval.

For the ratio of geometric means the 95% normal-approximation interval is
(`r fmt(interval_of("theta", "norm_lo"), 2)`,
`r fmt(interval_of("theta", "norm_hi"), 2)`).

### Comparison

The table collects both intervals for both statistics. Its last column is
the $t$ interval from `lm()` for $\beta_1$ and, for $\theta$, the
exponential of that interval.

```{r interval-table}
#| echo: false
interval_table
```

```{r slope-interval-gaps}
#| include: false
# Largest distance between a bootstrap interval's endpoint and the t
# interval's corresponding endpoint, for the slope.
slope_gap <- max(abs(
  c(interval_of("slope", "perc_lo"), interval_of("slope", "norm_lo"),
    interval_of("slope", "perc_hi"), interval_of("slope", "norm_hi")) -
  c(ci_lm[[1]], ci_lm[[1]], ci_lm[[2]], ci_lm[[2]])))
```

For the slope the three intervals nearly coincide: each endpoint differs
from the $t$ interval's by at most `r fmt(slope_gap, 2)`, because the
bootstrap distribution of $\hat\beta_1^{*b}$ is close to normal and centered
at $\hat\beta_1$. For the ratio of geometric means the two bootstrap
intervals differ. The normal-approximation interval is symmetric about
$\hat\theta$, and its lower limit,
`r fmt(interval_of("theta", "norm_lo"), 2)`, lies
`r if (interval_of("theta", "norm_lo") < min(boot$theta)) "below every bootstrap value" else "in the thin lower tail of the bootstrap distribution"`.
The percentile interval follows the skew: its upper limit is
`r fmt(interval_of("theta", "perc_hi") - theta_hat, 2)` above $\hat\theta$
and its lower limit `r fmt(theta_hat - interval_of("theta", "perc_lo"), 2)`
below it, like the exponentiated $t$ interval
(`r fmt(interval_of("theta", "lm_lo"), 2)`,
`r fmt(interval_of("theta", "lm_hi"), 2)`).

The two intervals agree when the bootstrap distribution is approximately
normal and centered at $\hat\theta$. Skewness separates them, and so does
bias, a bootstrap distribution centered away from $\hat\theta$, which shifts
the percentile interval but not the normal-approximation interval.

In the `r n_check` simulated experiments of the Accuracy subsection, the 95%
percentile interval contained $\theta$ in `r fmt(100 * coverage$perc, 1)`% of
them, missing low (entirely below $\theta$) in
`r fmt(100 * coverage$perc_above, 1)`% and high in
`r fmt(100 * coverage$perc_below, 1)`%. The normal-approximation interval
contained $\theta$ in `r fmt(100 * coverage$norm, 1)`%, missing low in
`r fmt(100 * coverage$norm_above, 1)`% and high in
`r fmt(100 * coverage$norm_below, 1)`%, and its lower limit was negative, a
value no ratio of positive quantities can take, in
`r fmt(100 * coverage$norm_negative, 1)`%. Both cover $\theta$ at close to the
nominal rate, but only the percentile interval stays within the possible
values of $\theta$ and splits its misses between the two sides.

## Sample overlap

The bootstrap samples in the Resampling tabs each left out some of the `r n`
days. The bar chart counts, over all `r fmt(B)` bootstrap samples, the number
of distinct days each one contains.

::: {.callout-note collapse="true"}
### Code
```{r distinct-days}
#| output: false
boot <- boot |> mutate(distinct = map_int(i, n_distinct))
distinct_mean <- mean(boot$distinct)
distinct_plot <- ggplot(boot, aes(x = distinct)) +
  geom_bar(fill = "grey75", color = "grey45") +
  labs(x = sprintf("Distinct days in a bootstrap sample of %d", n),
       y = "Bootstrap samples",
       title = "Distinct observations in the cloud bootstrap samples")
```
:::

```{r distinct-plot}
#| echo: false
#| fig-width: 6.5
#| fig-height: 3.5
distinct_plot
```

The bootstrap samples contain between `r min(boot$distinct)` and
`r max(boot$distinct)` distinct days, `r fmt(distinct_mean, 1)` on average, or
`r fmt(100 * distinct_mean / n, 1)`% of the `r n`.

### Inclusion probability

Each of the $n$ draws that make up a bootstrap sample picks observation $j$
with probability $1/n$, independently of the other draws. So

$$P\left(\text{observation } j \text{ not chosen on one draw}\right) = 1 - \frac{1}{n}, \qquad P\left(\text{observation } j \notin Z^{*b}\right) = \left(1 - \frac{1}{n}\right)^n,$$

and the probability that observation $j$ appears in the bootstrap sample is

$$P\left(\text{observation } j \in Z^{*b}\right) = 1 - \left(1 - \frac{1}{n}\right)^n \longrightarrow 1 - e^{-1} \approx 0.632 \quad \text{as } n \to \infty,$$

using $\lim_{n \to \infty} (1 + x/n)^n = e^x$ with $x = -1$. The number of
distinct observations in $Z^{*b}$ is
$\sum_{j=1}^n \mathrm{I}\left(\text{observation } j \in Z^{*b}\right)$, so by
linearity of expectation

$$E\left[\sum_{j=1}^n \mathrm{I}\left(\text{observation } j \in Z^{*b}\right)\right] = n\left[1 - \left(1 - \frac{1}{n}\right)^n\right].$$

```{r inclusion-theory}
#| include: false
p_incl <- function(n) 1 - (1 - 1 / n)^n
```

For the `r n` cloud seeding days, $1 - (1 - 1/`r n`)^{`r n`} = `r fmt(p_incl(n), 3)`$,
so a bootstrap sample is expected to contain `r fmt(n * p_incl(n), 1)`
distinct days, against the `r fmt(distinct_mean, 1)` averaged over the
`r fmt(B)` samples above. The figure plots the inclusion probability against
$n$, with a dashed line at $1 - e^{-1}$ and a point at $n = `r n`$.

::: {.callout-note collapse="true"}
### Code
```{r inclusion-curve}
#| output: false
inclusion_plot <- tibble(n = 1:100) |>
  mutate(p = p_incl(n)) |>
  ggplot(aes(x = n, y = p)) +
  geom_hline(yintercept = 1 - exp(-1), linetype = "dashed") +
  geom_line() +
  geom_point(data = tibble(n = nrow(clouds), p = p_incl(nrow(clouds))),
             size = 3, shape = 21, fill = "white") +
  labs(x = "Sample size n",
       y = "P(observation j in bootstrap sample)",
       title = "Inclusion probability 1 - (1 - 1/n)^n")
```
:::

```{r inclusion-plot}
#| echo: false
#| fig-width: 6.5
#| fig-height: 3.5
inclusion_plot
```

The inclusion probability is `r fmt(p_incl(2), 2)` at $n = 2$ and falls
quickly toward $1 - e^{-1} = `r fmt(1 - exp(-1), 3)`$; by $n = 20$ it is
`r fmt(p_incl(20), 3)`.

### Test error

A bootstrap sample could serve as a training set: fit the model to $Z^{*b}$,
predict the original $n$ observations, and average the error over $b$. About
`r fmt(100 * (1 - exp(-1)), 1)`% of the observations being predicted were
also used to fit the model, some of them several times, so this estimate
mixes training error into test error and is too optimistic, increasingly so
for flexible models that fit their training data closely. Cross-validation,
the subject of the previous lecture, avoids the overlap: each fold's held-out
observations are not in the data the model is fit to.

## Limitations

The bootstrap replaces the population with the sample, so it is only as good
as that replacement.

- **Small samples.** A small sample may miss the population's tails, and the
bootstrap cannot reproduce what the sample does not contain. Even for a
well-behaved statistic, $\text{SE}_B$ is an estimate with sampling error, as
the Accuracy subsection showed at `r m_cloud` days per treatment, and that
error shrinks only as $n$ grows, as the first tab below shows for
$\hat\theta$.
- **Heavy-tailed or skewed populations with a mean-type statistic.** A mean
of a heavy right tail is driven by its few largest values. The first tab
compares $\text{SE}_B$ with the true standard error for $\hat\theta$ and for
the ratio of mean rainfall, $\bar y_S/\bar y_U$, where $\bar y_S$ and
$\bar y_U$ are the sample mean rainfall on seeded and unseeded days, over
simulated experiments of three sizes.
- **Extreme statistics.** The bootstrap maximum can never exceed the sample
maximum, and it equals the sample maximum whenever that observation is drawn,
with probability $1 - (1 - 1/n)^n$. The second tab shows the bootstrap
distribution of the maximum rainfall on a seeded day, from the `r fmt(B)`
cloud bootstrap samples.
- **Dependent data.** Resampling individual observations assumes they are
independent. Temperatures on consecutive days or repeated measurements on
the same subject are correlated, and resampling them one at a time breaks that
correlation, which usually makes $\text{SE}_B$ too small.

::: {.callout-note collapse="true"}
### Code
```{r limitations}
#| output: false
# The ratio of mean rainfall, and its bootstrap SE, on the cloud data.
ratio_of_means <- function(d) {
  mean(d$Rainfall[d$seeded == 1]) / mean(d$Rainfall[d$seeded == 0])
}
ratio_hat <- ratio_of_means(clouds)
boot <- boot |> mutate(ratio = map_dbl(i, \(i) ratio_of_means(clouds[i, ])))
se_boot_ratio <- sd(boot$ratio)

# True SEs and bootstrap SEs at three numbers of days per treatment. The
# 26-day experiments are the ones already bootstrapped in the Accuracy
# subsection; the larger sizes reuse the same functions.
sizes <- c(m_cloud, 100, 400)
size_true <- bind_rows(
  tibble(m = m_cloud, theta = se_true[["theta"]], ratio = se_true[["ratio"]]),
  map(sizes[-1], \(m) {
    st <- simulate_stats(m, n_experiments)
    tibble(m = m, theta = sd(st$theta), ratio = sd(st$ratio),
           ratio_mcse = sd(tapply(st$ratio,
                                  rep(1:20, each = n_experiments / 20),
                                  sd)) / sqrt(20) / sd(st$ratio))
  }) |> list_rbind())
size_checks <- bind_rows(checks,
                         map(sizes[-1], \(m) check_experiments(m, n_check,
                                                               B_check))) |>
  left_join(size_true, by = "m") |>
  transmute(m,
            theta = se_b_theta / theta,
            ratio = se_b_ratio / ratio)
size_summary <- size_checks |>
  pivot_longer(c(theta, ratio), names_to = "stat", values_to = "relative") |>
  group_by(stat, m) |>
  summarize(within = mean(abs(relative - 1) <= 0.25),
            median = median(relative), .groups = "drop")
size_of <- function(st, mm, col) {
  size_summary[[col]][size_summary$stat == st & size_summary$m == mm]
}

limit_labels <- c(theta = "Ratio of geometric means",
                  ratio = "Ratio of means")
size_plot <- size_checks |>
  pivot_longer(c(theta, ratio), names_to = "stat", values_to = "relative") |>
  mutate(stat = factor(limit_labels[stat], levels = limit_labels)) |>
  ggplot(aes(x = factor(m), y = relative)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 0.75, ymax = 1.25,
           alpha = 0.15) +
  geom_hline(yintercept = 1, linewidth = 0.9) +
  geom_boxplot(outlier.size = 0.8) +
  scale_y_log10() +
  facet_wrap(~ stat) +
  labs(x = "Days per treatment",
       y = "Bootstrap SE divided by true SE (log scale)",
       title = sprintf("%d simulated experiments per size; shaded band: within 25%%",
                       n_check))

seeded_max <- max(clouds$Rainfall[clouds$seeded == 1])
boot <- boot |>
  mutate(max_seeded = map_dbl(i, \(i) {
    max(clouds$Rainfall[i][clouds$seeded[i] == 1])
  }))
share_at_max <- mean(boot$max_seeded == seeded_max)
max_plot <- ggplot(boot, aes(x = max_seeded)) +
  geom_histogram(bins = 60, fill = "grey75", color = "grey45") +
  labs(x = "Bootstrap maximum seeded-day rainfall (acre-feet)",
       y = "Bootstrap samples",
       title = "Bootstrap distribution of the maximum")
```
:::

::: {.panel-tabset}
#### SE accuracy by size
```{r size-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 4.2
size_plot
```

#### Maximum
```{r max-plot}
#| echo: false
#| fig-width: 6.5
#| fig-height: 3.8
max_plot
```
:::

Under the known population, log rainfall is normal with a common $\sigma$
in both treatments, so the mean rainfall on a day with seeding indicator $D$
is $e^{\beta_0 + \beta_1 D + \sigma^2/2}$, and the ratio of means
$\mu_S/\mu_U$ also equals $\theta$, where $\mu_S$ and $\mu_U$ are the
population mean rainfall on seeded and unseeded days. The two statistics
estimate the same quantity, but at `r m_cloud` days per treatment the ratio
of means $\bar y_S/\bar y_U$ has true standard error
`r fmt(se_true[["ratio"]], 2)` (accurate to about
$\pm$`r fmt(200 * se_true_mcse[["ratio"]])`%), against
`r fmt(se_true[["theta"]], 2)` for $\hat\theta$.

The table gives, for each statistic and size, the percentage of the
`r n_check` simulated experiments whose $\text{SE}_B$ fell within 25% of the
true standard error. With `r n_check` experiments each percentage is accurate
to about $\pm$`r fmt(200 * sqrt(0.25 / n_check))` percentage points or
better (two binomial standard errors).

```{r size-table}
#| echo: false
size_summary |>
  mutate(Statistic = limit_labels[stat],
         within = paste0(fmt(100 * within), "%")) |>
  select(Statistic, m, within) |>
  pivot_wider(names_from = m, values_from = within) |>
  knitr::kable(col.names = c("Statistic",
                             paste(sizes, "days per treatment")),
               align = "lccc")
```

For the ratio of geometric means the percentage climbs quickly with $n$, as
its $\text{SE}_B$ inherits the shrinking error of $\hat\theta$. For the ratio
of means it is lower at every size and reaches only
`r fmt(100 * size_of("ratio", sizes[3], "within"))`% at `r sizes[3]` days per
treatment. The median ratio of $\text{SE}_B$ to the truth for the ratio of
means is `r fmt(size_of("ratio", sizes[1], "median"), 2)`,
`r fmt(size_of("ratio", sizes[2], "median"), 2)`, and
`r fmt(size_of("ratio", sizes[3], "median"), 2)` at the three sizes (each
true standard error is from `r fmt(n_experiments)` experiments, accurate to
about $\pm$`r fmt(200 * max(se_true_mcse[["ratio"]], size_true$ratio_mcse, na.rm = TRUE))`%
or better), so its $\text{SE}_B$ is typically too small.

A likely reason is that a sample seldom holds the population's rare, very
wet days in their population proportion: it usually holds too few, giving an
$\text{SE}_B$ that is too small, and occasionally holds several, giving one
far too large, and the bootstrap resamples only the days the sample
contains. On the log scale those days are not extreme. The two bootstrap
standard errors therefore err for different reasons: that of the ratio of
geometric means is noisy because $\text{SE}(\hat\theta)$ is proportional to
$\theta$, which is known only through $\hat\theta$, while that of the ratio
of means is also biased low by the heavy right tail. On the cloud data
the ratio of means is `r fmt(ratio_hat, 2)` with
$\text{SE}_B = `r fmt(se_boot_ratio, 2)`$, likely well below its true
standard error.

In the cloud bootstrap samples, `r fmt(100 * share_at_max, 1)`% of the
bootstrap maxima equal the sample maximum of `r fmt(seeded_max, 1)`
acre-feet, against $1 - (1 - 1/`r n`)^{`r n`} = `r fmt(100 * p_incl(n), 1)`$%
expected. Most of the bootstrap distribution is a single spike at the sample
maximum, so it carries no information about how far the population maximum
lies above the sample's.

::: {.callout-note collapse="true"}
### Beyond this course
- The **delta method** gives an approximate formula standard error for a
smooth function of estimates from a first-order Taylor expansion; for
$\hat\theta = e^{\hat\beta_1}$ it is $\hat\theta\,\text{SE}(\hat\beta_1)$.
- **Out-of-bag error** repairs the bootstrap's test-error estimate by
predicting each observation only from bootstrap samples that left it out;
it reappears with bagging and random forests (ISLR2 Chapter 8).
- The **parametric bootstrap** resamples from a fitted model rather than from
the data, much as the known-population simulation above drew experiments from
the fitted log-scale regression.
- **Bias-corrected and accelerated (BCa)** and **bootstrap-$t$** intervals
improve on the percentile interval's coverage in small or skewed samples.
- The **block bootstrap** resamples blocks of consecutive observations to
keep the dependence in a time series, and **stratified resampling** draws
within each group (here, `r m_cloud` days from each treatment) to match a
design with fixed group sizes.
:::

## Conclusion

The standard error of a statistic is the standard deviation of its sampling
distribution. Simple formulas give it for a few statistics, such as a mean
or a regression coefficient, and `lm()` reports it for each coefficient but
not for a function of the coefficients, such as an exponentiated
coefficient. A median, a quantile, an AUC, and many statistics computed by
an algorithm have no simple formula. The bootstrap estimates the standard
error by treating the sample as the population: draw $B$ bootstrap samples
of $n$ observations with replacement, compute the statistic on each, and
take the standard deviation of the $B$ results. Where a formula exists, such
as for a regression coefficient, the two estimate the same standard error
and can be checked against each other; where no simple formula is
available, the bootstrap is computed in exactly the same way. A bootstrap standard error is itself an estimate, and for a
statistic whose standard error grows with the parameter, such as a ratio, it
is no more accurate than the estimate itself.

The bootstrap distribution gives two confidence intervals. The percentile
interval takes the $\alpha/2$ and $1 - \alpha/2$ quantiles of the bootstrap
statistics, follows any skew, and moves with a monotone transformation of the
statistic; the normal-approximation interval is
$\hat\theta \pm z_{1-\alpha/2}\,\text{SE}_B(\hat\theta)$ and is symmetric.
They agree when the bootstrap distribution is approximately normal and
centered at $\hat\theta$, and skewness or bias separates them.

A bootstrap sample contains each observation with probability
$1 - (1 - 1/n)^n$, which approaches $1 - 1/e \approx 0.632$, so a model fit
to a bootstrap sample and evaluated on the original data is evaluated mostly
on its own training observations; cross-validation, not the bootstrap,
estimates test error. Bootstrap standard errors and intervals are unreliable
for small samples, for heavy-tailed or skewed populations summarized by a
mean-type statistic, for extreme statistics such as the maximum, and for
dependent data resampled one observation at a time.

The next lecture begins regularization (ISLR2 Sections 6.1-6.2): selecting a
subset of the features, and shrinking the least squares coefficient estimates
toward zero with ridge regression and the lasso, with cross-validation
choosing how much.
