# Simple Logistic Regression

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
set.seed(20260919)
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **explain** why a linear regression model is a poor model for a binary
response,
- **write** the simple logistic regression model in both its probability and
log-odds forms, and **define** the odds of an event,
- **derive** the Bernoulli log-likelihood for $\beta_0$ and $\beta_1$ and
**explain** why, unlike least squares, it has no closed-form maximizer,
- **fit** a simple logistic regression with `glm()`, **predict** the
probability of the event at a given feature value, and **construct** a
confidence interval for that probability by transforming a log-odds interval
through the logistic function,
- **interpret** $\hat\beta_0$ as a log-odds and $\hat\beta_1$ as a change in
log-odds, with $e^{\hat\beta_1}$ an odds ratio, and
- **test** $H_0: \beta_1 = 0$ with a $z$-statistic and **construct** a
confidence interval for $\beta_1$ and for the odds ratio $e^{\beta_1}$.
:::

## Binary response

Every model we have *fit* so far has had a quantitative response. Now take the
classification setup from the classification lecture, with $C = 2$ classes, and
code the response numerically,

$$Y_i = \begin{cases} 1 & \text{the event of interest occurs for observation } i \\ 0 & \text{otherwise,}\end{cases}$$

where

- $Y_i$ is the response for observation $i$, and
- "the event of interest" is whichever of the two classes we choose to call a
success.

The coding is a choice, and it is not innocent: the model below describes the
class coded $1$, so swapping the labels flips the sign of every coefficient.

The expected value of a $0/1$ response is a probability,

$$E[Y_i|X_i] = 1 \cdot P(Y_i = 1|X_i) + 0 \cdot P(Y_i = 0|X_i) = P(Y_i = 1|X_i) \equiv p(X_i),$$

where

- $X_i$ is the feature for observation $i$ (a scalar throughout this lecture),
and
- $p(X_i) = P(Y_i = 1|X_i)$ is the conditional probability that the event
occurs.

So modeling the mean of a binary response *is* modeling a probability, and any
model for $p(X)$ has to respect $0 \le p(X) \le 1$ for every $X$.

### Credit card default data

**Question.** Does a customer's credit card balance tell us whether that
customer will default?

**Data.** `ISLR2::Default` holds `r format(nrow(Default), big.mark = ",")`
credit card customers. The data are simulated rather than observed, and we use
them because they are the canonical demonstration of this method. Today we use
two of its four variables:

- `default`: did the customer default? A factor with levels
`r paste(levels(Default$default), collapse = " and ")`. This is the response.
- `balance`: the average balance remaining on the card after the monthly
payment, in dollars. This is the feature.

Defaulting is the rarer and more consequential outcome, so it is the event of
interest: $Y_i = 1$ when customer $i$ defaults. In this sample
`r round(100 * mean(Default$default == "Yes"), 2)`% of customers default.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r default-eda}
#| output: false
Default <- Default |>
  mutate(default01 = as.numeric(default == "Yes"),
         default_jit = default01 + runif(n(), -0.03, 0.03))

default_plot <- ggplot(Default, aes(x = balance, y = default_jit)) +
  geom_point(alpha = 0.08, size = 0.7) +
  scale_y_continuous(breaks = c(0, 1), labels = c("No (0)", "Yes (1)")) +
  labs(x = "Credit card balance ($)", y = "Default",
       title = "ISLR2::Default")
```
:::

```{r default-eda-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_plot
```

The points are jittered vertically, since every response is exactly $0$ or
exactly $1$ and thousands of customers would otherwise stack on top of one
another.

No customer with a balance under
\$`r round(min(Default$balance[Default$default == "Yes"]))` defaulted, and
defaults become steadily more common as the balance grows, while above \$2,000
the non-defaulters thin out. Exact
proportions cannot be read off jittered points, but the shape can: whatever
curve describes $p(\text{balance})$ sits near $0$ on the left, rises through
the middle, and has to stay below $1$ on the right.

The obvious next step is to fit the model we already have, by least squares,
to this $0/1$ response — expand the note below to see why that does not work.

::: {.callout-note collapse="true"}
### Linear regression fails

Applying the linear regression model of the earlier lectures to this $0/1$
response gives

$$Y_i = \beta_0 + \beta_1 X_i + \epsilon_i.$$

Since $E[Y_i|X_i] = p(X_i)$, this model says the probability of default is a
straight line in balance. Least squares will happily fit that line to the
default data, so the fastest way to see what goes wrong is to draw it:

```{r default-lm}
#| output: false
default_lm <- lm(default01 ~ balance, data = Default)

lm_plot <- ggplot(Default, aes(x = balance, y = default_jit)) +
  geom_point(alpha = 0.08, size = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_abline(intercept = coef(default_lm)[1], slope = coef(default_lm)[2],
              color = "#0072B2", linewidth = 1) +
  coord_cartesian(xlim = range(Default$balance),
                  ylim = c(min(fitted(default_lm)) - 0.02, 1.03)) +
  scale_y_continuous(breaks = c(-0.05, 0, 0.5, 1)) +
  labs(x = "Credit card balance ($)", y = "Default (0/1)",
       title = "ISLR2::Default")

zero_crossing <- -coef(default_lm)[1] / coef(default_lm)[2]
percent_negative <- 100 * mean(fitted(default_lm) < 0)
```

```{r default-lm-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
lm_plot
```

The fitted line drops below the dashed zero line on the left. It crosses zero
at a balance of \$`r round(zero_crossing)`, so every customer carrying less
than that — `r round(percent_negative)`% of the sample — is assigned a
*negative* estimated probability of default. A negative probability is not a
poor estimate; it is not a probability.

Three objections:

1. **Fitted values leave $[0,1]$.** A line is unbounded, and $p(X)$ is not.
Truncating at $0$ and $1$ after the fact fixes the printed number without
fixing the model that produced it.

2. **A constant additive effect cannot be right.** The fitted slope says every
extra dollar of balance raises the line's value by the same
$`r format(signif(coef(default_lm)[2], 3), scientific = FALSE)`$ regardless of
where we start. Wherever the line's value reaches $0.99$, only $0.01$ of room
is left, and just
`r round(0.01 / coef(default_lm)[2])` more dollars of balance pushes it past
$1$ — a "probability" of more than 100%. A probability has to flatten as it
approaches either boundary, and a line never flattens.

3. **The error assumptions fail by construction.** $Y_i$ is Bernoulli, not
normal, so $\epsilon_i = Y_i - p(X_i)$ takes exactly two values, and
$$Var[Y_i|X_i] = p(X_i)\left[1 - p(X_i)\right]$$
depends on $X_i$ through $p(X_i)$. Constant variance is impossible unless
$p(X)$ is constant. Every standard error, t-test, and interval from `lm()`
rests on assumptions the response itself contradicts.

What we need is a model that is flexible in the middle and bends toward $0$
and $1$ at the ends — the shape the data already showed us.
:::

## Logistic regression model

The **simple logistic regression** model is

$$p(X) = P(Y = 1|X) = \frac{e^{\beta_0 + \beta_1 X}}{1 + e^{\beta_0 + \beta_1 X}},$$

where

- $p(X)$ is the probability the event occurs at feature value $X$,
- $\beta_0$ is the intercept on the log-odds scale (defined below), and
- $\beta_1$ is the slope on the log-odds scale.

Write $\eta = \beta_0 + \beta_1 X$ for the **linear predictor**. The function

$$\text{logistic}(\eta) = \frac{e^{\eta}}{1 + e^{\eta}} = \left(1 + e^{-\eta}\right)^{-1}$$

is the **logistic function**, available in R as `plogis()` — the second form
follows from dividing numerator and denominator of the first by $e^\eta$, and
the two are used interchangeably depending on which is easier to work with in
a given derivation. The logistic function maps the whole real line into
$(0,1)$: $e^{\eta} > 0$ makes $\text{logistic}(\eta) > 0$, and
$e^\eta < 1 + e^\eta$ makes $\text{logistic}(\eta) < 1$. As
$\eta \to -\infty$, $\text{logistic}(\eta) \to 0$; as $\eta \to \infty$,
$\text{logistic}(\eta) \to 1$. The boundary problem raised above is gone no
matter what values $\beta_0$, $\beta_1$, and $X$ take.

::: {.callout-note collapse="true"}
### Beyond this course
The logistic function is not the only function that maps the real line onto
$(0,1)$. Replacing it with the standard normal CDF gives **probit
regression**, and the whole family — logistic, probit, Poisson regression, and
others — travels under the name **generalized linear models**; the inverse
map, which takes a probability back to the linear predictor, is called the
*link function*. `glm()` is named for the family, not for logistic regression
specifically.
:::

Plotting the curve at a few parameter values shows what $\beta_0$ and $\beta_1$
each control:

::: {.callout-note collapse="true"}
### Code for the figure below
```{r logistic-shapes}
#| output: false
shape_data <- tibble(b0 = c(0, 0, 0, -4), b1 = c(1, 2, -1, 1)) |>
  mutate(curve = sprintf("beta[0] == %d * ',' ~ beta[1] == %d", b0, b1)) |>
  mutate(curve = fct_inorder(curve)) |>
  expand_grid(x = seq(-8, 8, length.out = 400)) |>
  mutate(p = plogis(b0 + b1 * x))

logistic_shape_plot <- ggplot(shape_data,
                              aes(x = x, y = p,
                                  color = curve, linetype = curve)) +
  geom_hline(yintercept = c(0, 1), linetype = "dashed", color = "grey60") +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7"),
                     labels = scales::parse_format()) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted", "dotdash"),
                        labels = scales::parse_format()) +
  labs(x = "X", y = "p(X)", color = NULL, linetype = NULL)
```
:::

```{r logistic-shapes-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
logistic_shape_plot
```

Every curve stays strictly between the two dashed boundaries. Comparing the
first two, doubling $\beta_1$ makes the curve rise more steeply without moving
where it crosses $0.5$; the third shows that a negative $\beta_1$ reverses the
direction; the fourth shows that changing $\beta_0$ shifts the curve
horizontally without changing its shape. $\beta_1 = 0$ would give a horizontal
line at $\text{logistic}(\beta_0)$: no relationship between $X$ and the
probability.

### Odds and logit

The **odds** of an event with probability $p$ are

$$\text{odds} = \frac{p}{1 - p},$$

the ratio of the probability the event happens to the probability it does not.
Odds run over $(0, \infty)$: $p = 0.5$ gives odds of $1$, and $p = 0.8$ gives
odds of $4$, quoted as "4 to 1."

Apply this to the logistic model. Since

$$1 - p(X) = 1 - \frac{e^{\beta_0 + \beta_1 X}}{1 + e^{\beta_0 + \beta_1 X}} = \frac{1}{1 + e^{\beta_0 + \beta_1 X}},$$

the odds are

$$\frac{p(X)}{1 - p(X)} = e^{\beta_0 + \beta_1 X}.$$

Taking logarithms,

$$\log\left(\frac{p(X)}{1 - p(X)}\right) = \beta_0 + \beta_1 X.$$

The left side is the **logit**, or **log-odds**, of $p(X)$. It is the inverse
of the logistic function, and it carries $(0,1)$ onto the whole real line.

This is the payoff of the logistic form. The model is *linear* — in the
parameters and in the feature — once the response is expressed on the log-odds
scale, so the entire linear-model vocabulary transfers over. Only the scale on
which linearity holds has changed.

### Interpreting coefficients

Setting $X = 0$ in the log-odds form gives $\beta_0$ as the log-odds of the
event when $X = 0$, so $e^{\beta_0}$ is the odds and
$\text{logistic}(\beta_0) = e^{\beta_0}/(1 + e^{\beta_0})$ the probability, at
$X = 0$.

Now increase the feature by one unit from an arbitrary $x$:

$$\log\left(\frac{p(x+1)}{1-p(x+1)}\right) - \log\left(\frac{p(x)}{1-p(x)}\right) = \left[\beta_0 + \beta_1 (x+1)\right] - \left[\beta_0 + \beta_1 x\right] = \beta_1.$$

So $\beta_1$ is the change in the log-odds of the event per one-unit increase
in $X$. Exponentiating turns the difference into a ratio,

$$\frac{p(x+1)/[1-p(x+1)]}{p(x)/[1-p(x)]} = e^{\beta_1},$$

so $e^{\beta_1}$ is the **odds ratio**: the multiplicative factor by which the
odds of the event change per one-unit increase in $X$. An odds ratio of $1$
($\beta_1 = 0$) means the odds do not change.

What no coefficient reports is a constant change in the probability, because
there isn't one. Differentiating $p(x) = \text{logistic}(\beta_0 + \beta_1 x)$,

$$\frac{d\, p(x)}{dx} = \beta_1\, p(x)\left[1 - p(x)\right],$$

which depends on $x$ through $p(x)$. It is largest in magnitude,
$|\beta_1|/4$, where $p(x) = 0.5$, and it vanishes as $p(x)$ approaches $0$ or
$1$ — the flattening a bounded probability needs and a line could never
provide. A single number is no more "the effect of $X$ on $p$" here than a
single coefficient was the effect of $X$ in a polynomial regression.

### Maximum likelihood estimation

Under the model, the observations are independent Bernoulli trials whose
success probability depends on the feature,

$$Y_i \stackrel{ind}{\sim} \text{Bernoulli}\left(p(x_i)\right), \qquad i = 1,\ldots,n,$$

so the probability mass function of a single observation is
$P(Y_i = y_i|x_i) = p(x_i)^{y_i}\left[1 - p(x_i)\right]^{1-y_i}$ for
$y_i \in \{0,1\}$. The likelihood is the product over the sample,

$$L(\beta_0, \beta_1) = \prod_{i=1}^n p(x_i)^{y_i} \left[1 - p(x_i)\right]^{1 - y_i},$$

where

- $(x_i, y_i)$ is the observed feature and response for observation $i$, and
- $p(x_i) = e^{\beta_0 + \beta_1 x_i}/(1 + e^{\beta_0 + \beta_1 x_i})$ depends
on the parameters being estimated.

There is no $\sigma^2$ anywhere: once $p(x_i)$ is specified, the variance
$p(x_i)[1-p(x_i)]$ is determined, so the model has exactly two unknown
parameters.

Take logarithms, and use the odds identity
$p(x_i)/[1 - p(x_i)] = e^{\beta_0 + \beta_1 x_i}$ to simplify:

$$\begin{array}{rl}
\ell(\beta_0, \beta_1) &= \displaystyle\sum_{i=1}^n \left\{ y_i \log p(x_i) + (1 - y_i)\log\left[1 - p(x_i)\right] \right\} \\[2mm]
&= \displaystyle\sum_{i=1}^n \left\{ y_i \log\left(\frac{p(x_i)}{1 - p(x_i)}\right) + \log\left[1 - p(x_i)\right] \right\} \\[2mm]
&= \displaystyle\sum_{i=1}^n \left\{ y_i (\beta_0 + \beta_1 x_i) - \log\left(1 + e^{\beta_0 + \beta_1 x_i}\right) \right\}.
\end{array}$$

The **maximum likelihood estimators** $\hat\beta_0$ and $\hat\beta_1$ are the
values maximizing $\ell$. Differentiating,

$$\frac{\partial \ell}{\partial \beta_0} = \sum_{i=1}^n \left[y_i - p(x_i)\right], \qquad \frac{\partial \ell}{\partial \beta_1} = \sum_{i=1}^n x_i\left[y_i - p(x_i)\right],$$

and setting both to zero gives the **score equations**

$$\sum_{i=1}^n \left[y_i - p(x_i)\right] = 0, \qquad \sum_{i=1}^n x_i\left[y_i - p(x_i)\right] = 0.$$

These say the residuals $y_i - p(x_i)$ sum to zero and are orthogonal to the
feature — the same two conditions the least squares normal equations impose.
The difference is what happens next. In least squares, $p(x_i)$ would be
replaced by $\beta_0 + \beta_1 x_i$, the equations would be *linear* in
$\beta_0$ and $\beta_1$, and solving them would deliver the closed-form least
squares estimators of the simple linear regression lecture. Here $p(x_i)$ is
a nonlinear function of $\beta_0$ and $\beta_1$, so the equations are
nonlinear and no rearrangement produces a closed form.

The estimates are therefore found numerically. The log-likelihood is strictly
concave whenever the feature is not constant, so when a maximum exists it is
unique, and `glm()` locates it with an iterative algorithm.

::: {.callout-note collapse="true"}
### Beyond this course
The algorithm `glm()` uses is **iteratively reweighted least squares** (IRLS),
which is Newton-Raphson on $\ell$ in disguise: each iteration solves a weighted
least squares problem whose weights $p(x_i)[1 - p(x_i)]$ are recomputed from
the current estimate. Convergence usually takes a handful of iterations, which
is why `glm()` returns nearly as fast as `lm()` despite having no formula to
evaluate.
:::

::: {.callout-note collapse="true"}
### Log-likelihood and log loss
Divide the negative log-likelihood by $n$:

$$-\frac{1}{n}\ell(\beta_0,\beta_1) = -\frac{1}{n}\sum_{i=1}^n \left\{ y_i \log p(x_i) + (1 - y_i)\log\left[1 - p(x_i)\right]\right\}.$$

This is exactly the **log loss** defined in the classification lecture,
specialized to $C = 2$ classes and evaluated on the training data. Maximizing
the likelihood and minimizing training log loss are the same optimization
problem.
:::

### Fitting

```{r glm-syntax}
#| eval: false
glm(default ~ balance, data = Default, family = "binomial")
```

`family = "binomial"` is what makes this logistic regression; without it,
`glm()` fits a linear model. Two details govern what the coefficients mean:

- With a factor response, R models the probability of the **second** level.
Here `levels(Default$default)` is
`r paste(levels(Default$default), collapse = ", ")`, so `glm()` models
$P(\texttt{default} = \texttt{Yes})$, which is the event we chose.
- The response may also be supplied as a $0/1$ numeric vector, with $1$ as the
event.

::: {.callout-note collapse="true"}
### Other ways to write this model
`glm()` only needs to know which outcome counts as $Y_i = 1$; it does not
care how that information is encoded.

`default`'s second level is already `"Yes"`, since factor levels default to
alphabetical order — but relying on alphabetical order to pick the event is
fragile. `relevel()` sets the reference level explicitly, moving the named
level to the front so every other level (here, the one level left) is the
modeled event regardless of how the factor happened to be built:

```{r glm-relevel}
#| eval: false
Default$default <- relevel(Default$default, ref = "No") # make the reference explicit
```

The response can also be recoded into a $0/1$ numeric column or a
`TRUE`/`FALSE` logical column, or compared to the event of interest right
inside the formula. All four of the calls below fit identically:

```{r glm-equivalent-forms}
#| eval: false
Default$default01 <- as.numeric(Default$default == "Yes") # 0/1 numeric
Default$defaultTF <- Default$default == "Yes"              # TRUE/FALSE logical

glm(default ~ balance, data = Default, family = "binomial")           # factor
glm(default01 ~ balance, data = Default, family = "binomial")         # 0/1 numeric
glm(defaultTF ~ balance, data = Default, family = "binomial")         # logical
glm(default == "Yes" ~ balance, data = Default, family = "binomial")  # comparison
```

::: {.callout-note collapse="true"}
### Code for the table below
```{r glm-equivalent-forms-check}
#| output: false
default_prep <- Default |>
  mutate(default = relevel(default, ref = "No"),
         default01 = as.numeric(default == "Yes"),
         defaultTF = default == "Yes")

m_factor     <- glm(default ~ balance, data = default_prep, family = "binomial")
m_numeric    <- glm(default01 ~ balance, data = default_prep, family = "binomial")
m_logical    <- glm(defaultTF ~ balance, data = default_prep, family = "binomial")
m_comparison <- glm(default == "Yes" ~ balance, data = default_prep, family = "binomial")

equivalent_forms_table <- rbind(
  factor     = coef(m_factor),
  "0/1 numeric" = coef(m_numeric),
  "TRUE/FALSE logical" = coef(m_logical),
  comparison = coef(m_comparison)
)

stopifnot(isTRUE(all.equal(coef(m_factor), coef(m_numeric))),
          isTRUE(all.equal(coef(m_factor), coef(m_logical))),
          isTRUE(all.equal(coef(m_factor), coef(m_comparison))))
```
:::

```{r glm-equivalent-forms-table}
#| echo: false
knitr::kable(equivalent_forms_table, digits = 6)
```

The coefficients agree exactly. A logical vector is coerced to $0/1$ with
`TRUE` as $1$, which is exactly what `default01` does by hand, and
`default == "Yes"` builds that same logical vector inline without storing it
as a column at all. Of the four, the comparison form is often the easiest to
read: the formula states the event being modeled directly, rather than
leaving it implicit in a factor's level order or a column name.

::: {.callout-note collapse="true"}
### Beyond this course
The comparison form generalizes beyond a two-category response. With a
response that has more than two categories, `glm(y == "target_category" ~ x,
family = "binomial")` fits a binary logistic regression for that one category
against every other category combined, without constructing a separate $0/1$
or logical column first.
:::
:::

Fitting it to the credit card data,

```{r default-glm}
default_glm <- glm(default ~ balance, data = Default, family = "binomial")
coef(default_glm)
```

```{r default-fit-plot-setup}
#| include: false
fit_bin_width <- 200
```

Putting the fitted curve back on the data checks the shape against what the
plot suggested. Alongside it, the observed proportion defaulting within
\$`r fit_bin_width`-wide balance bins gives a model-free estimate of $p(X)$ to
compare against; point size shows how many customers are in each bin, since
the right-most bins hold very few.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r default-fit-plot}
#| output: false
balance_grid <- tibble(balance = seq(0, max(Default$balance), length.out = 400))
balance_grid$p_hat <- predict(default_glm, newdata = balance_grid, type = "response")

binned <- Default |>
  mutate(bin = cut(balance, breaks = seq(0, 2800, by = fit_bin_width),
                   include.lowest = TRUE)) |>
  group_by(bin) |>
  summarize(balance = mean(balance), p = mean(default01), n = n(), .groups = "drop")

fit_plot <- ggplot(Default, aes(x = balance, y = default_jit)) +
  geom_point(alpha = 0.08, size = 0.7) +
  geom_point(data = binned, aes(y = p, size = n), shape = 1, color = "#D55E00") +
  geom_line(data = balance_grid, aes(y = p_hat),
            color = "#0072B2", linewidth = 1) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_size_area(max_size = 5) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       size = "Customers\nin bin", title = "ISLR2::Default")
```
:::

```{r default-fit-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
fit_plot
```

The fitted curve stays inside $[0,1]$ over the full range of balances, sits
near zero out to roughly \$1,000, and passes through the binned proportions
where those bins contain enough customers to be worth reading.

### Coefficient inference

The maximum likelihood estimators are approximately normal in large samples,
which gives the test statistic

$$z = \frac{\hat\beta_j - \beta_j^0}{\text{SE}(\hat\beta_j)},$$

where

- $\beta_j^0$ is the value of $\beta_j$ under the null hypothesis, almost
always $0$, and
- $\text{SE}(\hat\beta_j)$ is the standard error of $\hat\beta_j$, computed by
`glm()` from the curvature of the log-likelihood at its maximum.

Under $H_0: \beta_j = \beta_j^0$, $z$ is approximately standard normal. The
statistic is the same ratio as the t-statistic of the linear regression
lectures; the reference distribution is normal rather than $t$ because no
separate error variance is estimated and because the normality is asymptotic
rather than exact.

`glm()` reports this test for every coefficient at once:

```{r default-summary}
summary(default_glm)
```

The coefficient table has the same layout as `lm()`'s, with one visible
change: `z value` and `Pr(>|z|)` in place of `t value` and `Pr(>|t|)`, for the
reason just given.

::: {.callout-note collapse="true"}
### Beyond this course
The two deviance lines are the likelihood's answer to RSS and TSS. For a
binary response, the **residual deviance** is $-2\ell$ evaluated at
$\hat\beta_0, \hat\beta_1$, and the **null deviance** is $-2\ell$ for the
intercept-only model. Differences between deviances of nested models play the
role that differences in RSS play for an F-test, and `AIC` penalizes the
residual deviance by twice the number of estimated parameters.
:::

An approximate $(1-\alpha)100\%$ confidence interval for $\beta_j$ follows:

$$\hat\beta_j \pm z_{1-\alpha/2}\,\text{SE}(\hat\beta_j),$$

with $z_{1-\alpha/2}$ the $(1-\alpha/2)$ quantile of the standard normal
distribution. In R, `confint.default()` on a fitted `glm` returns exactly this
interval for $\beta_j$. Exponentiating both endpoints gives an interval for
the odds ratio $e^{\beta_j}$, because $e^x$ is increasing and therefore
preserves the order of the endpoints.

::: {.callout-note collapse="true"}
### Beyond this course
Plain `confint()` on a `glm` object returns a different, *profile likelihood*
interval instead, built from the log-likelihood directly rather than from
$\hat\beta_j \pm z_{1-\alpha/2}\text{SE}(\hat\beta_j)$, so it will not match
`confint.default()`'s numbers.
:::

For `balance`,

```{r default-inference}
z_balance <- summary(default_glm)$coefficients["balance", "z value"]
ci_beta1  <- confint.default(default_glm)["balance", ]

round(z_balance, 2)  # z-statistic for beta_1
round(ci_beta1,  5)  # 95% CI for beta_1, per dollar
```

$z = `r round(z_balance, 2)`$ is far beyond any conventional critical value,
so the data are overwhelmingly inconsistent with $\beta_1 = 0$. The confidence
interval for $\beta_1$ runs from $`r signif(ci_beta1[1], 4)`$ to
$`r signif(ci_beta1[2], 4)`$ log-odds per dollar, and exponentiating both
endpoints gives an interval for the odds ratio per dollar, entirely above $1$,
the odds ratio corresponding to no association.

### Prediction

The fitted model returns a probability at any feature value,

$$\hat p(x) = \frac{e^{\hat\beta_0 + \hat\beta_1 x}}{1 + e^{\hat\beta_0 + \hat\beta_1 x}}.$$

In R, `predict()` on a `glm` object returns the linear predictor
$\hat\beta_0 + \hat\beta_1 x$ by default; `type = "response"` applies
$\text{logistic}(\cdot)$ and returns $\hat p(x)$.

Evaluating $\hat p(x)$ at a few balances spread across the range of the data
shows how the three scales — log-odds, odds, and probability — move against
one another:

::: {.callout-note collapse="true"}
### Code for the table below
```{r default-predictions}
#| output: false
pred_balances <- c(500, 1000, 1500, 2000, 2500)

pred_data <- tibble(balance = pred_balances) |>
  mutate(log_odds = predict(default_glm, newdata = tibble(balance = pred_balances)),
         odds     = exp(log_odds),
         p_hat    = predict(default_glm, newdata = tibble(balance = pred_balances),
                            type = "response"))

pred_table <- pred_data |>
  mutate(log_odds = sprintf("%.2f", log_odds),
         odds     = formatC(odds,  format = "g", digits = 3),
         p_hat    = formatC(p_hat, format = "g", digits = 3)) |>
  knitr::kable(col.names = c("Balance ($)", "Log-odds", "Odds",
                             "Estimated P(default)"),
               align = "cccc")

p_at <- function(x) pred_data$p_hat[pred_data$balance == x]
```
:::

```{r default-predictions-table}
#| echo: false
pred_table
```

The log-odds column increases by the same amount for every \$500 step, by
construction, and the odds column is multiplied by the same factor. The
probability column does neither: the step from \$1,500 to \$2,000 raises the
estimated probability by `r round(p_at(2000) - p_at(1500), 3)`, while the
identical \$500 step from \$500 to \$1,000 raises it by only
`r round(p_at(1000) - p_at(500), 3)`, because down there the curve is still
pressed against $0$.

$\hat p(x)$ is itself an estimate, with uncertainty inherited from
$\hat\beta_0$ and $\hat\beta_1$. Building a confidence interval for $p(x)$
climbs the same three scales as prediction did above: log-odds, then odds,
then probability.

`predict()` with `type = "link"` and `se.fit = TRUE` returns $\hat\eta(x)$ and
$\text{SE}(\hat\eta(x))$, giving a Wald interval for the log-odds at $x$, built
the same way as the interval for $\beta_1$ above:

$$\hat\eta(x) \pm z_{1-\alpha/2}\,\text{SE}(\hat\eta(x)).$$

Exponentiating both endpoints — the same increasing transform that turned the
interval for $\beta_1$ into an interval for the odds ratio $e^{\beta_1}$ —
carries this to an interval for the **odds** at $x$,
$\text{odds}(x) = p(x)/[1-p(x)]$. A second increasing transform,
$p = \text{odds}/(1+\text{odds})$ — the same map that turns $e^\eta$ into
$\text{logistic}(\eta) = e^\eta/(1+e^\eta)$ earlier in this section — carries
the odds interval to an interval for $p(x)$ itself.

```{r default-ci-walkthrough-setup}
x0 <- 1500
```

At a balance of \$`r format(x0, big.mark = ",")`:

```{r default-ci-walkthrough}
#| output: false
link_x0    <- predict(default_glm, newdata = tibble(balance = x0),
                      type = "link", se.fit = TRUE)
eta_hat_x0 <- link_x0$fit
eta_ci_x0  <- eta_hat_x0 + c(-1, 1) * qnorm(0.975) * link_x0$se.fit
odds_ci_x0 <- exp(eta_ci_x0)
p_ci_x0    <- odds_ci_x0 / (1 + odds_ci_x0)

ci_walkthrough <- tibble(
  Scale    = c("Log-odds", "Odds", "P(default)"),
  Estimate = c(eta_hat_x0, exp(eta_hat_x0), plogis(eta_hat_x0)),
  Lower    = c(eta_ci_x0[1], odds_ci_x0[1], p_ci_x0[1]),
  Upper    = c(eta_ci_x0[2], odds_ci_x0[2], p_ci_x0[2])
) |>
  mutate(across(c(Estimate, Lower, Upper), \(x) sprintf("%.4f", x))) |>
  knitr::kable(col.names = c("Scale", "Estimate", "2.5%", "97.5%"),
               align = "lccc")
```

```{r default-ci-walkthrough-table}
#| echo: false
ci_walkthrough
```

The composition of two increasing transforms is increasing, so this two-step
route lands on the same interval as applying $\text{logistic}(\cdot)$ directly
to the $\eta(x)$ endpoints in one step — the shortcut used to build the whole
band below. This is a *pointwise* interval: each vertical slice is a separate
95% interval for $p(x)$ at that one balance, not a region covering the whole
curve at once.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r default-confidence-band}
#| output: false
band_grid <- tibble(balance = seq(0, max(Default$balance), length.out = 400))

band_link <- predict(default_glm, newdata = band_grid,
                     type = "link", se.fit = TRUE)

band_grid <- band_grid |>
  mutate(eta   = band_link$fit,
         se    = band_link$se.fit,
         p_hat = plogis(eta),
         lower = plogis(eta - qnorm(0.975) * se),
         upper = plogis(eta + qnorm(0.975) * se))

confidence_band_plot <- ggplot(band_grid, aes(x = balance, y = p_hat)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey70", alpha = 0.6) +
  geom_line(color = "#0072B2", linewidth = 1) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       title = "ISLR2::Default")

peak_balance <- band_grid$balance[which.max(band_grid$upper - band_grid$lower)]
```
:::

```{r default-confidence-band-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
confidence_band_plot
```

The gray band is squeezed to a hairline at both ends, where the curve is
pinned near $0$ or near $1$ regardless of the exact coefficients, and is
widest through the steep middle, peaking near a balance of
\$`r format(round(peak_balance), big.mark = ",")`. This is not because the
coefficients are estimated less precisely there — the standard error of
$\hat\eta(x)$ is actually
*smallest* near the middle of the balance range and largest at the two ends —
but because a fixed-width interval on the log-odds scale converts to a wider
probability-scale interval exactly where the logistic curve is steepest, and
to a narrower one where it is nearly flat against $0$ or $1$. This is also
why the band is not symmetric around the curve the way a linear regression's
confidence interval for the mean response would be.

## Credit card default probability

**Question.** Does a customer's credit card balance predict whether they
default, and by how much?

**Data.** `ISLR2::Default` again: `r format(nrow(Default), big.mark = ",")`
credit card customers, `default` as the response
(`r round(100 * mean(Default$default == "Yes"), 2)`% default), and `balance`
as the feature.

```{r default-recap-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_plot
```

Defaults are rare at low balances and steadily more common as balance climbs,
the shape a probability model needs to reproduce.

### Fitting

Fitting $\texttt{default} \sim \texttt{balance}$ by logistic regression,

```{r default-recap-fit}
coef(default_glm)
```

```{r default-recap-summary}
summary(default_glm)
```

Both coefficients are far from zero relative to their standard errors, exactly
as the $z$-test above found.

### Estimated odds ratios

::: {.callout-note collapse="true"}
### Code for the values quoted below
```{r default-coefficients}
#| output: false
b         <- coef(default_glm)
or_dollar <- unname(exp(b["balance"]))
or_100    <- unname(exp(100 * b["balance"]))
ci_or_100 <- exp(100 * ci_beta1)
```
:::

The estimated log-odds of default are

$$\log\left(\frac{\hat p(X)}{1 - \hat p(X)}\right) = `r round(b[1], 2)` + `r signif(b[2], 4)` \times \texttt{balance}.$$

The intercept $\hat\beta_0 = `r round(b[1], 2)`$ is the estimated log-odds of
default for a customer carrying no balance, which corresponds to estimated odds
of $e^{`r round(b[1], 2)`} = `r format(exp(b[1]), digits = 3, scientific = FALSE)`$
and an estimated probability of
$`r format(plogis(b[1]), digits = 3, scientific = FALSE)`$ — odds and
probability agree to three significant digits here because $p/(1-p) \approx p$ whenever $p$ is
small. Unlike many intercepts, this one describes customers who actually exist:
`r sum(Default$balance == 0)` customers in this sample carry a zero balance,
and `r sum(Default$balance == 0 & Default$default == "Yes")` of them defaulted.

The slope $\hat\beta_1 = `r signif(b[2], 4)`$ is positive, so every additional
dollar of balance raises the log-odds of default by
$`r signif(b[2], 4)`$ — equivalently, multiplies the odds of default by
$e^{\hat\beta_1} = `r round(or_dollar, 5)`$.

A dollar is too small a unit to read comfortably, and nothing forces us to use
it: `balance` can be rescaled by any positive constant without changing the
model, since $\beta_1 X = (100\beta_1)(X/100)$ — measuring balance in hundreds
of dollars instead of dollars multiplies the slope by $100$ and leaves the
fitted probabilities untouched. Scaling up this way, an additional \$100 of
balance multiplies the odds of default by

$$e^{100\hat\beta_1} = `r round(or_100, 2)`,$$

a `r round(100 * (or_100 - 1))`% increase in the **odds** of default per \$100.
It is not a `r round(100 * (or_100 - 1))`% increase in the probability of
default; the change in probability depends on the balance you started from.
\$100 is simply the increment a person managing a credit line can picture
without doing arithmetic first.

The same rescaling applies to the confidence interval: from the coefficient
inference above, multiplying the endpoints of the interval for $\beta_1$ by
$100$ and exponentiating gives $(`r round(ci_or_100[1], 2)`,
`r round(ci_or_100[2], 2)`)$ on the per-\$100 odds-ratio scale, every value of
it well above $1$.

### Fitted probability

```{r default-recap-fit-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
fit_plot
```

The fitted curve stays inside $[0,1]$ everywhere, and tracks the binned
observed proportions closely wherever a bin holds enough customers to estimate
one.

## Conclusion

Simple logistic regression models a binary response's probability as the
logistic function of a linear predictor,
$p(X) = \text{logistic}(\beta_0 + \beta_1 X)$, equivalently a statement that
the *log-odds* of the event are linear in $X$. Estimation is by maximum
likelihood rather than least squares because a Bernoulli response has a
variance determined by its own mean, which breaks the least squares error
assumptions by construction; and because $p(x_i)$ is a nonlinear function of
$\beta_0$ and $\beta_1$, the score equations have no closed-form solution, so
`glm()` finds $\hat\beta_0$ and $\hat\beta_1$ numerically.

Coefficients are read on the log-odds scale — $\beta_0$ as a baseline log-odds,
$\beta_1$ as a change in log-odds per unit of $X$ — and exponentiated into odds
ratios for a multiplicative statement about the odds. Neither describes a
constant change in probability, because the logistic curve has no constant
slope. The same increasing-transform idea that turns a log-odds interval into
an odds-ratio interval also turns a confidence interval for the linear
predictor at a point into a confidence interval for the predicted probability
there, one increasing map — exponentiate, then odds-to-probability — at a
time.

The next lecture keeps this model and adds features: multiple logistic
regression asks what each feature contributes once the others are held fixed,
the same question multiple linear regression asked of a quantitative
response.
