# Simple Logistic Regression

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **explain** why a linear regression model is a poor model for a binary
response,
- **write** the simple logistic regression model in both its probability and
log-odds forms, and **define** the odds of an event,
- **derive** the Bernoulli log-likelihood for $\beta_0$ and $\beta_1$ and
**explain** why, unlike least squares, it has no closed-form maximizer,
- **fit** a simple logistic regression with `glm()` and **predict** the
probability of the event at a given feature value,
- **interpret** $\hat\beta_1$ as a change in log-odds and $e^{\hat\beta_1}$ as
an odds ratio, and **contrast** both with the additive interpretation of the
simple linear regression slope, and
- **test** $H_0: \beta_1 = 0$ with a $z$-statistic and **construct** a
confidence interval for $\beta_1$ and for the odds ratio $e^{\beta_1}$.
:::

## A binary response

Every model we have *fit* so far has had a quantitative response. Now take the
classification setup from the classification lecture, with $C = 2$ classes, and
code the response numerically,

$$Y_i = \begin{cases} 1 & \text{the event of interest occurs for observation } i \\ 0 & \text{otherwise,}\end{cases}$$

where

- $Y_i$ is the response for observation $i$, and
- "the event of interest" is whichever of the two classes we choose to call a
success.

The coding is a choice, and it is not innocent: the model below describes the
class coded $1$, so swapping the labels flips the sign of every coefficient.

The expected value of a $0/1$ response is a probability,

$$E[Y_i|X_i] = 1 \cdot P(Y_i = 1|X_i) + 0 \cdot P(Y_i = 0|X_i) = P(Y_i = 1|X_i) \equiv p(X_i),$$

where

- $X_i$ is the feature for observation $i$ (a scalar throughout this lecture),
and
- $p(X_i) = P(Y_i = 1|X_i)$ is the conditional probability that the event
occurs.

So modeling the mean of a binary response *is* modeling a probability, and any
model for $p(X)$ has to respect $0 \le p(X) \le 1$ for every $X$.

## Example: credit card default

**Question.** Does a customer's credit card balance tell us whether that
customer will default?

**Data.** `ISLR2::Default` holds `r format(nrow(Default), big.mark = ",")`
credit card customers. The data are simulated rather than observed, and we use
them because they are the canonical demonstration of this method. Today we use
two of its four variables:

- `default`: did the customer default? A factor with levels
`r paste(levels(Default$default), collapse = " and ")`. This is the response.
- `balance`: the average balance remaining on the card after the monthly
payment, in dollars. This is the feature.

Defaulting is the rarer and more consequential outcome, so it is the event of
interest: $Y_i = 1$ when customer $i$ defaults. In this sample
`r round(100 * mean(Default$default == "Yes"), 2)`% of customers default.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r default-eda}
#| output: false
Default <- Default |>
  mutate(default01 = as.numeric(default == "Yes"))

default_plot <- ggplot(Default, aes(x = balance, y = default01)) +
  geom_jitter(height = 0.03, width = 0, alpha = 0.08, size = 0.7) +
  scale_y_continuous(breaks = c(0, 1), labels = c("No (0)", "Yes (1)")) +
  labs(x = "Credit card balance ($)", y = "Default",
       title = "ISLR2::Default")
```
:::

```{r default-eda-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_plot
```

The points are jittered vertically, since every response is exactly $0$ or
exactly $1$ and thousands of customers would otherwise stack on top of one
another.

No customer with a balance under
\$`r round(min(Default$balance[Default$default == "Yes"]))` defaulted, and
defaults become steadily more common as the balance grows, while above \$2,000
the non-defaulters thin out. Exact
proportions cannot be read off jittered points, but the shape can: whatever
curve describes $p(\text{balance})$ sits near $0$ on the left, rises through
the middle, and has to stay below $1$ on the right.

## Why not linear regression?

The obvious thing to try is the model we already have,

$$Y_i = \beta_0 + \beta_1 X_i + \epsilon_i,$$

fit by least squares to the $0/1$ response. Since $E[Y_i|X_i] = p(X_i)$, this
model says the probability of default is a straight line in balance.

### Example: fitting a line to default

Least squares will happily fit that line to the default data, so the fastest
way to see what goes wrong is to draw it:

::: {.callout-note collapse="true"}
### Code for the figure below
```{r default-lm}
#| output: false
default_lm <- lm(default01 ~ balance, data = Default)

lm_plot <- ggplot(Default, aes(x = balance, y = default01)) +
  geom_jitter(height = 0.03, width = 0, alpha = 0.08, size = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  geom_abline(intercept = coef(default_lm)[1], slope = coef(default_lm)[2],
              color = "#0072B2", linewidth = 1) +
  coord_cartesian(xlim = range(Default$balance),
                  ylim = c(min(fitted(default_lm)) - 0.02, 1.03)) +
  scale_y_continuous(breaks = c(-0.05, 0, 0.5, 1)) +
  labs(x = "Credit card balance ($)", y = "Default (0/1)",
       title = "ISLR2::Default")

zero_crossing <- -coef(default_lm)[1] / coef(default_lm)[2]
percent_negative <- 100 * mean(fitted(default_lm) < 0)
```
:::

```{r default-lm-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
lm_plot
```

The fitted line drops below the dashed zero line on the left. It crosses zero
at a balance of \$`r round(zero_crossing)`, so every customer carrying less
than that — `r round(percent_negative)`% of the sample — is assigned a
*negative* estimated probability of default. A negative probability is not a
poor estimate; it is not a probability.

### Three objections

1. **Fitted values leave $[0,1]$.** A line is unbounded, and $p(X)$ is not.
Truncating at $0$ and $1$ after the fact fixes the printed number without
fixing the model that produced it.

2. **A constant additive effect cannot be right.** The fitted slope says every
extra dollar of balance raises the line's value by the same
$`r format(signif(coef(default_lm)[2], 3), scientific = FALSE)`$ regardless of
where we start. Wherever the line's value reaches $0.99$, only $0.01$ of room
is left, and just
`r round(0.01 / coef(default_lm)[2])` more dollars of balance pushes it past
$1$ — a "probability" of more than 100%. A probability has to flatten as it
approaches either boundary, and a line never flattens.

3. **The error assumptions fail by construction.** $Y_i$ is Bernoulli, not
normal, so $\epsilon_i = Y_i - p(X_i)$ takes exactly two values, and
$$Var[Y_i|X_i] = p(X_i)\left[1 - p(X_i)\right]$$
depends on $X_i$ through $p(X_i)$. Constant variance is impossible unless
$p(X)$ is constant. Every standard error, t-test, and interval from `lm()`
rests on assumptions the response itself contradicts.

What we need is a model that is flexible in the middle and bends toward $0$ and
$1$ at the ends — the shape the data already showed us.

## The logistic model

The **simple logistic regression** model is

$$p(X) = P(Y = 1|X) = \frac{e^{\beta_0 + \beta_1 X}}{1 + e^{\beta_0 + \beta_1 X}},$$

where

- $p(X)$ is the probability the event occurs at feature value $X$,
- $\beta_0$ is the intercept on the log-odds scale (defined below), and
- $\beta_1$ is the slope on the log-odds scale.

Write $\eta = \beta_0 + \beta_1 X$ for the **linear predictor**. The function

$$\text{logistic}(\eta) = \frac{e^{\eta}}{1 + e^{\eta}}$$

is the **logistic function**, available in R as `plogis()`. It maps the whole
real line into $(0,1)$: $e^{\eta} > 0$ makes $\text{logistic}(\eta) > 0$, and
$e^\eta < 1 + e^\eta$ makes $\text{logistic}(\eta) < 1$. As
$\eta \to -\infty$, $\text{logistic}(\eta) \to 0$; as $\eta \to \infty$,
$\text{logistic}(\eta) \to 1$. Objection 1 is gone no matter what values
$\beta_0$, $\beta_1$, and $X$ take.

Plotting the curve at a few parameter values shows what $\beta_0$ and $\beta_1$
each control:

::: {.callout-note collapse="true"}
### Code for the figure below
```{r logistic-shapes}
#| output: false
shape_data <- tibble(b0 = c(0, 0, 0, -4), b1 = c(1, 2, -1, 1)) |>
  mutate(curve = sprintf("beta[0] == %d * ',' ~ beta[1] == %d", b0, b1)) |>
  mutate(curve = fct_inorder(curve)) |>
  expand_grid(x = seq(-8, 8, length.out = 400)) |>
  mutate(p = plogis(b0 + b1 * x))

logistic_shape_plot <- ggplot(shape_data,
                              aes(x = x, y = p,
                                  color = curve, linetype = curve)) +
  geom_hline(yintercept = c(0, 1), linetype = "dashed", color = "grey60") +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7"),
                     labels = scales::parse_format()) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted", "dotdash"),
                        labels = scales::parse_format()) +
  labs(x = "X", y = "p(X)", color = NULL, linetype = NULL)
```
:::

```{r logistic-shapes-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
logistic_shape_plot
```

Every curve stays strictly between the two dashed boundaries. Comparing the
first two, doubling $\beta_1$ makes the curve rise more steeply without moving
where it crosses $0.5$; the third shows that a negative $\beta_1$ reverses the
direction; the fourth shows that changing $\beta_0$ shifts the curve
horizontally without changing its shape. $\beta_1 = 0$ would give a horizontal
line at $\text{logistic}(\beta_0)$: no relationship between $X$ and the
probability.

### Odds and the logit

The **odds** of an event with probability $p$ are

$$\text{odds} = \frac{p}{1 - p},$$

the ratio of the probability the event happens to the probability it does not.
Odds run over $(0, \infty)$: $p = 0.5$ gives odds of $1$, and $p = 0.8$ gives
odds of $4$, quoted as "4 to 1."

Apply this to the logistic model. Since

$$1 - p(X) = 1 - \frac{e^{\beta_0 + \beta_1 X}}{1 + e^{\beta_0 + \beta_1 X}} = \frac{1}{1 + e^{\beta_0 + \beta_1 X}},$$

the odds are

$$\frac{p(X)}{1 - p(X)} = e^{\beta_0 + \beta_1 X}.$$

Taking logarithms,

$$\log\left(\frac{p(X)}{1 - p(X)}\right) = \beta_0 + \beta_1 X.$$

The left side is the **logit**, or **log-odds**, of $p(X)$. It is the inverse
of the logistic function, and it carries $(0,1)$ onto the whole real line.

This is the payoff of the logistic form. The model is *linear* — in the
parameters and in the feature — once the response is expressed on the log-odds
scale, so the entire linear-model vocabulary transfers over. Only the scale on
which linearity holds has changed.

## Estimation by maximum likelihood

Under the model, the observations are independent Bernoulli trials whose
success probability depends on the feature,

$$Y_i \stackrel{ind}{\sim} \text{Bernoulli}\left(p(x_i)\right), \qquad i = 1,\ldots,n,$$

so the probability mass function of a single observation is
$P(Y_i = y_i|x_i) = p(x_i)^{y_i}\left[1 - p(x_i)\right]^{1-y_i}$ for
$y_i \in \{0,1\}$. The likelihood is the product over the sample,

$$L(\beta_0, \beta_1) = \prod_{i=1}^n p(x_i)^{y_i} \left[1 - p(x_i)\right]^{1 - y_i},$$

where

- $(x_i, y_i)$ is the observed feature and response for observation $i$, and
- $p(x_i) = e^{\beta_0 + \beta_1 x_i}/(1 + e^{\beta_0 + \beta_1 x_i})$ depends
on the parameters being estimated.

There is no $\sigma^2$ anywhere: once $p(x_i)$ is specified, the variance
$p(x_i)[1-p(x_i)]$ is determined, so the model has exactly two unknown
parameters.

Take logarithms, and use the odds identity from the previous section to
simplify:

$$\begin{array}{rl}
\ell(\beta_0, \beta_1) &= \displaystyle\sum_{i=1}^n \left\{ y_i \log p(x_i) + (1 - y_i)\log\left[1 - p(x_i)\right] \right\} \\[2mm]
&= \displaystyle\sum_{i=1}^n \left\{ y_i \log\left(\frac{p(x_i)}{1 - p(x_i)}\right) + \log\left[1 - p(x_i)\right] \right\} \\[2mm]
&= \displaystyle\sum_{i=1}^n \left\{ y_i (\beta_0 + \beta_1 x_i) - \log\left(1 + e^{\beta_0 + \beta_1 x_i}\right) \right\}.
\end{array}$$

The **maximum likelihood estimators** $\hat\beta_0$ and $\hat\beta_1$ are the
values maximizing $\ell$. Differentiating,

$$\frac{\partial \ell}{\partial \beta_0} = \sum_{i=1}^n \left[y_i - p(x_i)\right], \qquad \frac{\partial \ell}{\partial \beta_1} = \sum_{i=1}^n x_i\left[y_i - p(x_i)\right],$$

and setting both to zero gives the **score equations**

$$\sum_{i=1}^n \left[y_i - p(x_i)\right] = 0, \qquad \sum_{i=1}^n x_i\left[y_i - p(x_i)\right] = 0.$$

These say the residuals $y_i - p(x_i)$ sum to zero and are orthogonal to the
feature — the same two conditions the least squares normal equations impose.
The difference is what happens next. In least squares, $p(x_i)$ would be
replaced by $\beta_0 + \beta_1 x_i$, the equations would be *linear* in
$\beta_0$ and $\beta_1$, and solving them would deliver the closed-form least
squares estimators of the simple linear regression lecture. Here $p(x_i)$ is
a nonlinear function of $\beta_0$ and $\beta_1$, so the equations are
nonlinear and no rearrangement produces a closed form.

The estimates are therefore found numerically. The log-likelihood is strictly
concave whenever the feature is not constant, so when a maximum exists it is
unique, and `glm()` locates it with an iterative algorithm.

::: {.callout-note collapse="true"}
### Beyond this course
The algorithm `glm()` uses is **iteratively reweighted least squares** (IRLS),
which is Newton-Raphson on $\ell$ in disguise: each iteration solves a weighted
least squares problem whose weights $p(x_i)[1 - p(x_i)]$ are recomputed from
the current estimate. Convergence usually takes a handful of iterations, which
is why `glm()` returns nearly as fast as `lm()` despite having no formula to
evaluate.

The logit is also not the only function that maps $(0,1)$ onto the real line.
Replacing it with the standard normal quantile function gives **probit
regression**, and the whole family — logistic, probit, Poisson regression, and
others — travels under the name **generalized linear models**, with the
transformation called the *link function*. `glm()` is named for the family, not
for logistic regression specifically.
:::

::: {.callout-note collapse="true"}
### The log-likelihood is the log loss
Divide the negative log-likelihood by $n$:

$$-\frac{1}{n}\ell(\beta_0,\beta_1) = -\frac{1}{n}\sum_{i=1}^n \left\{ y_i \log p(x_i) + (1 - y_i)\log\left[1 - p(x_i)\right]\right\}.$$

This is exactly the **log loss** defined in the classification lecture,
specialized to $C = 2$ classes and evaluated on the training data. Maximizing
the likelihood and minimizing training log loss are the same optimization
problem.
:::

## Fitting a logistic regression in R

```{r glm-syntax}
#| eval: false
glm(default ~ balance, data = Default, family = "binomial")
```

`family = "binomial"` is what makes this logistic regression; without it,
`glm()` fits a linear model. Two details govern what the coefficients mean:

- With a factor response, R models the probability of the **second** level.
Here `levels(Default$default)` is
`r paste(levels(Default$default), collapse = ", ")`, so `glm()` models
$P(\texttt{default} = \texttt{Yes})$, which is the event we chose.
- The response may also be supplied as a $0/1$ numeric vector, with $1$ as the
event.

::: {.callout-note collapse="true"}
### Other ways to write this model
`glm()` only needs to know which outcome counts as $Y_i = 1$; it does not
care how that information is encoded.

`default`'s second level is already `"Yes"`, since factor levels default to
alphabetical order — but relying on alphabetical order to pick the event is
fragile. `relevel()` sets the reference level explicitly, moving the named
level to the front so every other level (here, the one level left) is the
modeled event regardless of how the factor happened to be built:

```{r glm-relevel}
#| eval: false
Default$default <- relevel(Default$default, ref = "No") # make the reference explicit
```

The response can also be recoded into a $0/1$ numeric column or a
`TRUE`/`FALSE` logical column, or compared to the event of interest right
inside the formula. All four of the calls below fit identically:

```{r glm-equivalent-forms}
#| eval: false
Default$default01 <- as.numeric(Default$default == "Yes") # 0/1 numeric
Default$defaultTF <- Default$default == "Yes"              # TRUE/FALSE logical

glm(default ~ balance, data = Default, family = "binomial")           # factor
glm(default01 ~ balance, data = Default, family = "binomial")         # 0/1 numeric
glm(defaultTF ~ balance, data = Default, family = "binomial")         # logical
glm(default == "Yes" ~ balance, data = Default, family = "binomial")  # comparison
```

::: {.callout-note collapse="true"}
### Code for the table below
```{r glm-equivalent-forms-check}
#| output: false
default_prep <- Default |>
  mutate(default = relevel(default, ref = "No"),
         default01 = as.numeric(default == "Yes"),
         defaultTF = default == "Yes")

m_factor     <- glm(default ~ balance, data = default_prep, family = "binomial")
m_numeric    <- glm(default01 ~ balance, data = default_prep, family = "binomial")
m_logical    <- glm(defaultTF ~ balance, data = default_prep, family = "binomial")
m_comparison <- glm(default == "Yes" ~ balance, data = default_prep, family = "binomial")

equivalent_forms_table <- rbind(
  factor     = coef(m_factor),
  "0/1 numeric" = coef(m_numeric),
  "TRUE/FALSE logical" = coef(m_logical),
  comparison = coef(m_comparison)
)

stopifnot(isTRUE(all.equal(coef(m_factor), coef(m_numeric))),
          isTRUE(all.equal(coef(m_factor), coef(m_logical))),
          isTRUE(all.equal(coef(m_factor), coef(m_comparison))))
```
:::

```{r glm-equivalent-forms-table}
#| echo: false
knitr::kable(equivalent_forms_table, digits = 6)
```

The coefficients agree exactly. A logical vector is coerced to $0/1$ with
`TRUE` as $1$, which is exactly what `default01` does by hand, and
`default == "Yes"` builds that same logical vector inline without storing it
as a column at all. Of the four, the comparison form is often the easiest to
read: the formula states the event being modeled directly, rather than
leaving it implicit in a factor's level order or a column name.

::: {.callout-note collapse="true"}
### Beyond this course
The comparison form generalizes beyond a two-category response. With a
response that has more than two categories, `glm(y == "target_category" ~ x,
family = "binomial")` fits a binary logistic regression for that one category
against every other category combined, without constructing a separate $0/1$
or logical column first.
:::
:::

### Example: fitting the logistic model

```{r default-glm}
default_glm <- glm(default ~ balance, data = Default, family = "binomial")
summary(default_glm)
```

The coefficient table has the same layout as `lm()`'s, with one visible change:
`z value` and `Pr(>|z|)` in place of `t value` and `Pr(>|t|)`, for the reason
given in the inference section below.

::: {.callout-note collapse="true"}
### Beyond this course
The two deviance lines are the likelihood's answer to RSS and TSS. For a
binary response, the **residual deviance** is $-2\ell$ evaluated at
$\hat\beta_0, \hat\beta_1$, and the **null deviance** is $-2\ell$ for the
intercept-only model. Differences between deviances of nested models play the
role that differences in RSS play for an F-test, and `AIC` penalizes the
residual deviance by twice the number of estimated parameters.
:::

Putting the fitted curve back on the data checks the shape against what the
plot suggested. Alongside it, the observed proportion defaulting within
\$200-wide balance bins gives a model-free estimate of $p(X)$ to compare
against; point size shows how many customers are in each bin, since the
right-most bins hold very few.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r default-fit-plot}
#| output: false
balance_grid <- tibble(balance = seq(0, max(Default$balance), length.out = 400))
balance_grid$p_hat <- predict(default_glm, newdata = balance_grid, type = "response")

binned <- Default |>
  mutate(bin = cut(balance, breaks = seq(0, 2800, by = 200), include.lowest = TRUE)) |>
  group_by(bin) |>
  summarize(balance = mean(balance), p = mean(default01), n = n(), .groups = "drop")

fit_plot <- ggplot(Default, aes(x = balance, y = default01)) +
  geom_jitter(height = 0.03, width = 0, alpha = 0.08, size = 0.7) +
  geom_point(data = binned, aes(y = p, size = n), shape = 1, color = "#D55E00") +
  geom_line(data = balance_grid, aes(y = p_hat),
            color = "#0072B2", linewidth = 1) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_size_area(max_size = 5) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       size = "Customers\nin bin", title = "ISLR2::Default")
```
:::

```{r default-fit-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
fit_plot
```

The fitted curve stays inside $[0,1]$ over the full range of balances, sits
near zero out to roughly \$1,000, and passes through the binned proportions
where those bins contain enough customers to be worth reading.

## Interpreting the coefficients

Start from the log-odds form and increase the feature by one unit:

$$\log\left(\frac{p(x+1)}{1-p(x+1)}\right) - \log\left(\frac{p(x)}{1-p(x)}\right) = \left[\beta_0 + \beta_1 (x+1)\right] - \left[\beta_0 + \beta_1 x\right] = \beta_1.$$

So $\beta_1$ is the change in the log-odds of the event per one-unit increase
in $X$. Exponentiating turns the difference into a ratio,

$$\frac{p(x+1)/[1-p(x+1)]}{p(x)/[1-p(x)]} = e^{\beta_1},$$

so $e^{\beta_1}$ is the **odds ratio**: the multiplicative factor by which the
odds of the event change per one-unit increase in $X$. An odds ratio of $1$
($\beta_1 = 0$) means the odds do not change.

Setting $X = 0$ in the log-odds form gives $\beta_0$ as the log-odds of the
event when $X = 0$, so $e^{\beta_0}$ is the odds and
$\text{logistic}(\beta_0) = e^{\beta_0}/(1 + e^{\beta_0})$ the probability, at $X = 0$.

### Contrast with simple linear regression

In simple linear regression, $\beta_1 = E[Y|X = x+1] - E[Y|X = x]$: a one-unit
increase in $X$ *adds* $\beta_1$ to the mean response, the same amount at every
$x$. That is the interpretation from the simple linear regression lecture, and
it does not survive here.

| Quantity | Simple linear regression | Simple logistic regression |
| -------- | ------------------------ | -------------------------- |
| $\beta_0$ | mean response at $X = 0$ | log-odds of the event at $X = 0$ |
| $\beta_1$ | additive change in the mean response per unit of $X$ | additive change in the **log-odds** per unit of $X$ |
| $e^{\beta_1}$ | — | multiplicative change in the **odds** per unit of $X$ |

The pattern — additive on one scale, multiplicative on another — is the same
one that appeared when the *response* was logged in the simple linear
regression lecture, where $e^{\beta_1}$ became a multiplicative change in the
median response. The
difference is that here the transformation is not something we chose to apply
to the data; it is built into the model.

What no coefficient reports is a constant change in the probability, because
there isn't one. Differentiating $p(x) = \text{logistic}(\beta_0 + \beta_1 x)$,

$$\frac{d\, p(x)}{dx} = \beta_1\, p(x)\left[1 - p(x)\right],$$

which depends on $x$ through $p(x)$. It is largest in magnitude,
$|\beta_1|/4$, where $p(x) = 0.5$, and it vanishes as $p(x)$ approaches $0$ or
$1$ — the flattening
that objection 2 demanded. A single number is no more "the effect of $X$ on
$p$" here than a single coefficient was the effect of $X$ in a polynomial
regression.

### Example: interpreting the default coefficients

```{r default-coefficients}
b         <- coef(default_glm)
or_dollar <- unname(exp(b["balance"]))
or_100    <- unname(exp(100 * b["balance"]))

round(c("odds ratio per $1" = or_dollar, "odds ratio per $100" = or_100), 4)
```

The estimated log-odds of default are

$$\log\left(\frac{\hat p(X)}{1 - \hat p(X)}\right) = `r round(b[1], 2)` + `r signif(b[2], 4)` \times \texttt{balance}.$$

The slope $\hat\beta_1 = `r signif(b[2], 4)`$ is positive, so every additional
dollar of balance raises the log-odds of default by
$`r signif(b[2], 4)`$ — equivalently, multiplies the odds of default by
$e^{\hat\beta_1} = `r round(or_dollar, 5)`$.

A dollar is too small a unit to read comfortably, and nothing forces us to use
it. Scaling up by $100$, an additional \$100 of balance multiplies the odds of
default by

$$e^{100\hat\beta_1} = `r round(or_100, 2)`,$$

a `r round(100 * (or_100 - 1))`% increase in the **odds** of default per \$100.
It is not a `r round(100 * (or_100 - 1))`% increase in the probability of
default; the change in probability depends on the balance you started from.

The intercept $\hat\beta_0 = `r round(b[1], 2)`$ is the estimated log-odds of
default for a customer carrying no balance, which corresponds to estimated odds
of $e^{`r round(b[1], 2)`} = `r format(exp(b[1]), digits = 3, scientific = FALSE)`$
and an estimated probability of
$`r format(plogis(b[1]), digits = 3, scientific = FALSE)`$ — odds and
probability agree to three significant digits here because $p/(1-p) \approx p$ whenever $p$ is
small. Unlike many intercepts, this one describes customers who actually exist:
`r sum(Default$balance == 0)` customers in this sample carry a zero balance,
and `r sum(Default$balance == 0 & Default$default == "Yes")` of them defaulted.

## Inference for $\beta_1$

The maximum likelihood estimators are approximately normal in large samples,
which gives the test statistic

$$z = \frac{\hat\beta_1 - \beta_1^0}{\text{SE}(\hat\beta_1)},$$

where

- $\beta_1^0$ is the value of $\beta_1$ under the null hypothesis, almost
always $0$, and
- $\text{SE}(\hat\beta_1)$ is the standard error of $\hat\beta_1$, computed by
`glm()` from the curvature of the log-likelihood at its maximum.

Under $H_0: \beta_1 = \beta_1^0$, $z$ is approximately standard normal. The
statistic is the same ratio as the t-statistic of the linear regression
lectures; the reference distribution is normal rather than $t$ because no
separate error variance is estimated and because the normality is asymptotic
rather than exact. This is
why the `summary()` output above says `z value` and `Pr(>|z|)`.

An approximate $(1-\alpha)100\%$ confidence interval for $\beta_1$ follows:

$$\hat\beta_1 \pm z_{1-\alpha/2}\,\text{SE}(\hat\beta_1),$$

with $z_{1-\alpha/2}$ the $(1-\alpha/2)$ quantile of the standard normal
distribution. Exponentiating both endpoints gives an interval for the odds
ratio $e^{\beta_1}$, because $e^x$ is increasing and therefore preserves the
order of the endpoints.

### Example: testing and estimating the default slope

```{r default-inference}
z_balance <- summary(default_glm)$coefficients["balance", "z value"]
ci_beta1  <- confint.default(default_glm)["balance", ]
ci_or_100 <- exp(100 * ci_beta1)

round(ci_beta1,  5)   # 95% CI for beta_1, per dollar
round(ci_or_100, 2)   # 95% CI for the odds ratio, per $100
```

`confint.default()` is the interval derived above,
$\hat\beta_1 \pm z_{1-\alpha/2}\text{SE}(\hat\beta_1)$. Plain `confint()`
on a `glm` object returns a *profile likelihood* interval instead, which is
built from the log-likelihood directly and will not match this formula.

The `summary()` output above reports $z = `r round(z_balance, 2)`$ for
`balance`, far beyond any conventional critical value, so the data are
overwhelmingly inconsistent with $\beta_1 = 0$.

The confidence interval for $\beta_1$ runs from $`r signif(ci_beta1[1], 4)`$ to
$`r signif(ci_beta1[2], 4)`$ log-odds per dollar. Multiplying both endpoints by
$100$ and exponentiating puts it on the odds-ratio scale per \$100:
$(`r round(ci_or_100[1], 2)`, `r round(ci_or_100[2], 2)`)$. Every value in that
interval is well above $1$, the odds ratio corresponding to no association.

## Prediction

The fitted model returns a probability at any feature value,

$$\hat p(x) = \frac{e^{\hat\beta_0 + \hat\beta_1 x}}{1 + e^{\hat\beta_0 + \hat\beta_1 x}}.$$

In R, `predict()` on a `glm` object returns the linear predictor
$\hat\beta_0 + \hat\beta_1 x$ by default; `type = "response"` applies
$\text{logistic}(\cdot)$ and returns $\hat p(x)$.

### Example: predicting default probabilities

Evaluating $\hat p(x)$ at a few balances spread across the range of the data
shows how the three scales — log-odds, odds, and probability — move against
one another:

::: {.callout-note collapse="true"}
### Code for the table below
```{r default-predictions}
#| output: false
pred_balances <- c(500, 1000, 1500, 2000, 2500)

pred_data <- tibble(balance = pred_balances) |>
  mutate(log_odds = predict(default_glm, newdata = tibble(balance = pred_balances)),
         odds     = exp(log_odds),
         p_hat    = predict(default_glm, newdata = tibble(balance = pred_balances),
                            type = "response"))

pred_table <- pred_data |>
  mutate(log_odds = sprintf("%.2f", log_odds),
         odds     = formatC(odds,  format = "g", digits = 3),
         p_hat    = formatC(p_hat, format = "g", digits = 3)) |>
  knitr::kable(col.names = c("Balance ($)", "Log-odds", "Odds",
                             "Estimated P(default)"),
               align = "cccc")

p_at <- function(x) pred_data$p_hat[pred_data$balance == x]
```
:::

```{r default-predictions-table}
#| echo: false
pred_table
```

The log-odds column increases by the same amount for every \$500 step, by
construction, and the odds column is multiplied by the same factor. The
probability column does neither: the step from \$1,500 to \$2,000 raises the
estimated probability by `r round(p_at(2000) - p_at(1500), 3)`, while the
identical \$500 step from \$500 to \$1,000 raises it by only
`r round(p_at(1000) - p_at(500), 3)`, because down there the curve is still
pressed against $0$.

## Conclusion

Credit card balance is strongly associated with default. Each additional \$100
of balance multiplies the estimated odds of default by
`r round(or_100, 2)` (95% CI `r round(ci_or_100[1], 2)` to
`r round(ci_or_100[2], 2)`), and the estimated probability of default rises
from `r round(100 * p_at(500), 2)`% at a \$500 balance to
`r round(100 * p_at(2000))`% at \$2,000.

The logistic model delivered this while a linear regression could not, because
it constrains $p(X)$ to $[0,1]$, lets the effect of a dollar flatten as the
probability approaches either boundary, and is estimated under a Bernoulli
likelihood that matches how a binary response actually varies.

What it has not done is use the rest of the data. `ISLR2::Default` also records
whether each customer is a student and what their income is, and answering
"does balance matter *after accounting for* those?" needs more than one
feature — multiple logistic regression.
