# Multiple Logistic Regression

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
set.seed(20260919)

fmt_p <- function(p) {
  e <- floor(log10(p))
  sprintf("%.1f \\times 10^{%d}", p / 10^e, e)
}
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **write** the multiple logistic regression model in its probability and
log-odds forms using the linear predictor $X_i\beta$, and **encode** a
two-level categorical feature as an indicator against a reference level,
- **interpret** $\beta_j$ as a change in log-odds and $e^{\beta_j}$ as an odds
ratio *holding every other feature fixed*, and **distinguish** that from the
interpretation of the same feature's coefficient fit alone,
- **fit** an additive multiple logistic regression with `glm()`, **predict**
$\hat p(x)$ for a specified combination of feature values, and **construct** a
confidence interval for that probability by transforming a log-odds interval
to the odds scale and then to the probability scale,
- **test** $H_0: \beta_j = 0$ with a $z$-statistic and **interpret** a
confidence interval for the odds ratio $e^{\beta_j}$ as a statement about
feature $j$ after accounting for the other features,
- **conduct** a drop-in-deviance test comparing two nested logistic regression
models and **explain** its parallel with the F-test comparing two nested linear
regression models, and
- **diagnose** a coefficient whose sign reverses between a one-feature and a
multi-feature model as confounding, and **explain** the mechanism from
descriptive summaries of the features.
:::

## Multiple features

The previous lecture modeled the probability of default from a single feature,
credit card balance. `ISLR2::Default` records two more variables on the same
`r format(nrow(Default), big.mark = ",")` customers:

- `student`: is the customer a student? A factor with levels
`r paste(levels(Default$student), collapse = " and ")`.
- `income`: the customer's income, in dollars. This one is set aside for this
lecture; everything below uses `balance` and `student`.

Having more than one feature raises a question that a one-feature model cannot
answer: what does each feature tell us that the others do not? This lecture
answers it for an **additive** model, in which each feature contributes its own
term to the linear predictor.

### Categorical features

A categorical feature enters a logistic regression exactly as it enters a
multiple linear regression: through dummy variables against a reference level.
A feature with $L$ levels contributes $L-1$ indicators, and the level left
without an indicator is the reference.

`student` has two levels, so it contributes one indicator. Taking
`r levels(Default$student)[1]` as the reference level,

$$D_i = \mathrm{I}(\texttt{student}_i = \texttt{Yes}) = \begin{cases} 1 & \text{customer } i \text{ is a student} \\ 0 & \text{otherwise,}\end{cases}$$

where

- $D_i$ is the indicator for observation $i$, and
- $\mathrm{I}(\cdot)$ is $1$ when its argument is true and $0$ otherwise.

R builds this column automatically from a factor, using the first level as the
reference, and names the coefficient after the level being indicated —
`studentYes` below.

### Student status and default

**Question.** Are students more likely to default than non-students?

**Data.** The same `r format(nrow(Default), big.mark = ",")` customers as in the
previous lecture, now using `student` as the feature and `default` as the
response, with $Y_i = 1$ when customer $i$ defaults.

::: {.callout-note collapse="true"}
### Code for the table below
```{r student-eda}
#| output: false
Default <- Default |>
  mutate(default01 = as.numeric(default == "Yes"),
         default_jit = default01 + runif(n(), -0.03, 0.03))

student_rates <- Default |>
  group_by(student) |>
  summarize(n = n(),
            defaults = sum(default01),
            rate = mean(default01),
            .groups = "drop")

student_rate_table <- student_rates |>
  mutate(n = format(n, big.mark = ","),
         rate = sprintf("%.2f%%", 100 * rate)) |>
  knitr::kable(col.names = c("Student", "Customers", "Defaults",
                             "Percent defaulting"),
               align = "cccc")

rate_of <- function(s) student_rates$rate[student_rates$student == s]
```
:::

```{r student-eda-table}
#| echo: false
student_rate_table
```

Students default at `r sprintf("%.2f", 100 * rate_of("Yes"))`% versus
`r sprintf("%.2f", 100 * rate_of("No"))`% for non-students, so the student
default rate is `r sprintf("%.2f", rate_of("Yes") / rate_of("No"))` times the
non-student rate. That is a ratio of *rates*, not of odds; the two happen to
be close here only because both rates are small.

Fitting the simple logistic regression with $D$ as the feature turns that
comparison into a coefficient:

```{r student-glm}
student_glm <- glm(default ~ student, data = Default, family = "binomial")
coef(student_glm)
```

::: {.callout-note collapse="true"}
### Code for the estimates quoted below
```{r student-glm-extract}
#| output: false
bs <- coef(student_glm)
or_student_alone <- unname(exp(bs["studentYes"]))
ci_student_alone <- exp(confint.default(student_glm)["studentYes", ])
```
:::

The estimated log-odds of default are

$$\log\left(\frac{\hat p(D)}{1 - \hat p(D)}\right) = `r sprintf("%.2f", bs[1])` `r sprintf("%+.3f", bs[2])` \, D.$$

With a single indicator feature, the two fitted values are the two observed
log-odds. The intercept $\hat\beta_0 = `r sprintf("%.2f", bs[1])`$ is the
log-odds of default among non-students, and $\hat\beta_0 + \hat\beta_1$ is the
log-odds among students. The coefficient
$\hat\beta_1 = `r sprintf("%.3f", bs[2])`$ is the difference between them, so
$e^{\hat\beta_1} = `r sprintf("%.2f", or_student_alone)`$ is the odds of
default for a student divided by the odds for a non-student. The confidence
interval for that odds ratio runs from
`r sprintf("%.2f", ci_student_alone[1])` to
`r sprintf("%.2f", ci_student_alone[2])`, entirely above $1$.

**Conclusion.** Taken by itself, student status is associated with a
`r round(100 * (or_student_alone - 1))`% increase in the odds of default.

### Balance and student together

The previous lecture found balance strongly associated with default; this
lecture has now found student status associated with default as well. Before
putting both in one model, start with the individual customers rather than a
summary of them: the same jittered scatter of default against balance from
the previous lecture, now split into two panels by student status.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r student-default-raw}
#| output: false
raw_plot <- ggplot(Default, aes(x = balance, y = default_jit)) +
  geom_point(alpha = 0.08, size = 0.7) +
  facet_wrap(~ student, labeller = label_both) +
  scale_y_continuous(breaks = c(0, 1), labels = c("No (0)", "Yes (1)")) +
  labs(x = "Credit card balance ($)", y = "Default",
       title = "ISLR2::Default")
```
:::

```{r student-default-raw-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
raw_plot
```

The student panel's points sit further right on average, a first hint of a
balance difference between the two groups, but with
`r format(sum(Default$student == "Yes"), big.mark = ",")` students against
`r format(sum(Default$student == "No"), big.mark = ",")` non-students plotted
at the same jitter and transparency, that shift is hard to read precisely by
eye. Two descriptive summaries make it precise.

The first compares the distribution of balance for students and non-students.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r balance-by-student}
#| output: false
balance_means <- Default |>
  group_by(student) |>
  summarize(mean_balance = mean(balance), .groups = "drop")

balance_density_plot <- ggplot(Default,
                               aes(x = balance,
                                   color = student, linetype = student)) +
  geom_density(linewidth = 0.9) +
  geom_vline(data = balance_means,
             aes(xintercept = mean_balance,
                 color = student, linetype = student),
             linewidth = 0.5, show.legend = FALSE) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Credit card balance ($)", y = "Density",
       color = "Student", linetype = "Student",
       title = "ISLR2::Default")

mean_balance_of <- function(s) {
  balance_means$mean_balance[balance_means$student == s]
}
```
:::

```{r balance-by-student-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
balance_density_plot
```

The two densities are shifted relative to one another, with the vertical lines
marking the group means: students average
\$`r format(round(mean_balance_of("Yes")), big.mark = ",")` versus
\$`r format(round(mean_balance_of("No")), big.mark = ",")` for non-students, a
difference of
\$`r format(round(mean_balance_of("Yes") - mean_balance_of("No")), big.mark = ",")`.
Students carry more balance.

```{r binned-rates-setup}
#| include: false
bin_width <- 400
min_bin_n <- 100
```

The second summary asks what happens to the default rate within a narrow range
of balance. Splitting balance into \$`r bin_width`-wide bins and computing the
observed default rate separately for students and non-students within each bin
gives a model-free comparison of two customers who carry similar balances.
Bins holding too few customers of a given student status to estimate a rate
are dropped.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r binned-rates}
#| output: false
binned_rates <- Default |>
  mutate(bin = cut(balance, breaks = seq(0, 2800, by = bin_width),
                   include.lowest = TRUE)) |>
  group_by(student, bin) |>
  summarize(balance = mean(balance), rate = mean(default01), n = n(),
            .groups = "drop") |>
  filter(n >= min_bin_n)

binned_plot <- ggplot(binned_rates,
                      aes(x = balance, y = rate,
                          color = student, shape = student,
                          linetype = student)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Mean balance within bin ($)",
       y = "Observed proportion defaulting",
       color = "Student", shape = "Student", linetype = "Student",
       title = paste0("ISLR2::Default, $", bin_width, "-wide balance bins"))
```
:::

```{r binned-rates-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
binned_plot
```

Within every bin shown, the student curve sits at or below the non-student
curve. Comparing customers who carry similar balances reverses the direction of
the association that the aggregate rates showed. The curves stop near
\$`r format(round(max(binned_rates$balance)), big.mark = ",")` because every
bin beyond that holds fewer than `r min_bin_n` customers in at least one of the
two groups.

Neither summary used a model. Together they say that students carry higher
balances, that higher balances default more, and that at a fixed balance
students default less; a model with both features lets us separate those
statements.

## Multiple logistic regression model

Let $X_{i1}, \ldots, X_{ip}$ be the $p$ features recorded on observation $i$.
The **multiple logistic regression** model is

$$p(X_i) = P(Y_i = 1|X_i) = \frac{e^{\beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}}}{1 + e^{\beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}}},$$

where

- $p(X_i)$ is the probability the event occurs for observation $i$,
- $X_{ij}$ is the value of feature $j$ for observation $i$, and
- $\beta_0, \beta_1, \ldots, \beta_p$ are the regression coefficients.

The same algebra as in the previous lecture — form $1 - p(X_i)$, divide, take
logarithms — puts the model in its log-odds form,

$$\log\left(\frac{p(X)}{1 - p(X)}\right) = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p.$$

The right side is exactly the linear predictor of the multiple linear
regression lecture. Using that lecture's model matrix notation together with
the linear predictor $\eta$ of the previous lecture,

- $X_i = (1, X_{i1}, \ldots, X_{ip})$ is the $i$th row of the $n \times (p+1)$
model matrix $X$,
- $\beta = (\beta_0, \ldots, \beta_p)^\top$ is the vector of coefficients, and
- $\eta = X\beta$ is the $n$-vector of linear predictors, with $i$th element
$\eta_i = X_i\beta$,

the model is

$$p(X_i) = \text{logistic}(X_i\beta) = \frac{e^{X_i\beta}}{1 + e^{X_i\beta}}.$$

So multiple logistic regression is multiple linear regression's linear
predictor passed through the logistic function. Everything the linear predictor
can express — several quantitative features, dummy variables for categorical
features, polynomial terms, step functions — is available here unchanged,
because none of it touches the logistic function wrapped around it.

The model above is **additive**: each feature contributes its own term
$\beta_j X_{ij}$ to $\eta_i$, and no term involves two features at once.

### Maximum likelihood estimation

The observations are still independent Bernoulli trials,
$Y_i \stackrel{ind}{\sim} \text{Bernoulli}(p(x_i))$, so the log-likelihood
derived in the previous lecture carries over with $\beta_0 + \beta_1 x_i$
replaced by $x_i\beta$:

$$\ell(\beta) = \ell(\beta_0, \beta_1, \ldots, \beta_p) = \sum_{i=1}^n \left\{ y_i \, x_i\beta - \log\left(1 + e^{x_i\beta}\right)\right\},$$

where

- $x_i = (1, x_{i1}, \ldots, x_{ip})$ is the observed feature row for
observation $i$, and
- $y_i \in \{0, 1\}$ is the observed response.

Nothing structural changed. There are now $p+1$ parameters rather than $2$,
still no $\sigma^2$, and still no closed-form maximizer, because $p(x_i)$
remains a nonlinear function of the coefficients. The **maximum likelihood
estimator** $\hat\beta$ is the maximizer of $\ell$, and `glm()` finds it
numerically by the same iterative algorithm as before.

Differentiating with respect to $\beta_j$ and setting the result to zero gives
one score equation per coefficient,

$$\sum_{i=1}^n x_{ij}\left[y_i - p(x_i)\right] = 0, \qquad j = 0, 1, \ldots, p,$$

with $x_{i0} = 1$. As in the one-feature case, these say the residuals
$y_i - p(x_i)$ sum to zero and are orthogonal to every feature — the same
conditions the least squares normal equations impose, though here the
equations are nonlinear in $\beta$.

### Interpreting coefficients

Fix the values of all features other than $X_j$ and increase $X_j$ by one unit.
On the log-odds scale, every term except the $j$th is identical on both sides,
so the difference is

$$\left[\beta_0 + \cdots + \beta_j(x_j + 1) + \cdots + \beta_p x_p\right] - \left[\beta_0 + \cdots + \beta_j x_j + \cdots + \beta_p x_p\right] = \beta_j.$$

So $\beta_j$ is the change in the log-odds of the event per one-unit increase
in $X_j$ **holding every other feature fixed**, and exponentiating,
$e^{\beta_j}$ is the multiplicative change in the odds of the event per
one-unit increase in $X_j$ **holding every other feature fixed**.

The qualifier is the whole difference between this lecture and the last one,
and it is the same qualifier that attaches to a multiple linear regression
coefficient. The cancellation above is what licenses it: the comparison
$\beta_j$ describes is between two observations that differ in $X_j$ and agree
on everything else in the model.

Two consequences follow.

1. **$\beta_j$ from a multi-feature fit and from a one-feature fit answer
different questions.** Fit alone, a feature's coefficient describes the
comparison between observations that differ in that feature and in whatever
else happens to vary with it. In an additive model, it describes the
comparison holding the other included features fixed. The two can differ in
magnitude, and they can differ in sign.

2. **"Holding fixed" covers only features in the model.** A variable not in
$X$ is not held fixed by including other features, so a coefficient never
describes a comparison that adjusts for something the model never saw.

For a dummy variable — a feature $X_j$ taking only the values $0$ and $1$ — a
"one-unit increase" is the change from the reference level to the indicated
level, so $e^{\beta_j}$ is the odds of the event at that level divided by the
odds at the reference level, again holding the other features fixed.

The intercept $\beta_0$ is the log-odds of the event when every feature is $0$
— for a dummy variable, when the observation is at the reference level.

### Fitting

```{r additive-syntax}
#| eval: false
glm(default ~ balance + student, data = Default, family = "binomial")
```

The `+` makes the model additive: `balance` and `student` each contribute one
term to the linear predictor. Terms joined by `*` or `:` would add products of
features, which the additive model excludes.

Fitting it to the credit card data,

```{r additive-glm}
additive_glm <- glm(default ~ balance + student, data = Default,
                    family = "binomial")
coef(additive_glm)
```

`r length(coef(additive_glm))` coefficients are estimated for
$p = `r length(coef(additive_glm)) - 1`$ features, and the third is named
`studentYes`: R built the indicator $D$ from the factor and took
`r levels(Default$student)[1]` as the reference level, as promised above.
Writing $X_1 = \texttt{balance}$ and $X_2 = D$, the three estimates are
$\hat\beta_0$, $\hat\beta_1$, and $\hat\beta_2$.

### Coefficient inference

Inference proceeds coefficient by coefficient with the same machinery as the
previous lecture. Maximum likelihood estimators are approximately normal in
large samples, so

$$z = \frac{\hat\beta_j - \beta_j^0}{\text{SE}(\hat\beta_j)}$$

is approximately standard normal under $H_0: \beta_j = \beta_j^0$, where

- $\beta_j^0$ is the null value, almost always $0$, and
- $\text{SE}(\hat\beta_j)$ is the standard error computed by `glm()` from the
curvature of the log-likelihood at $\hat\beta$.

An approximate $(1-\alpha)100\%$ confidence interval is
$\hat\beta_j \pm z_{1-\alpha/2}\text{SE}(\hat\beta_j)$, and exponentiating both
endpoints gives an interval for the odds ratio $e^{\beta_j}$.

Every one of these statements is about $\beta_j$ with the other features in the
model held fixed, so the test of $H_0: \beta_j = 0$ asks whether feature $j$ is
associated with the response *after accounting for* the other features — not
whether it is associated with the response at all.

`glm()` reports the test for every coefficient at once:

```{r additive-summary}
summary(additive_glm)
```

::: {.callout-note collapse="true"}
### Code for the estimates quoted below
```{r additive-extract}
#| output: false
ba <- coef(additive_glm)
or_balance_100 <- unname(exp(100 * ba["balance"]))
or_student_adj <- unname(exp(ba["studentYes"]))
ci_adj <- confint.default(additive_glm)
ci_or_student_adj <- exp(ci_adj["studentYes", ])
ci_or_balance_100 <- exp(100 * ci_adj["balance", ])
z_student_adj <- summary(additive_glm)$coefficients["studentYes", "z value"]
p_student_adj <- summary(additive_glm)$coefficients["studentYes", "Pr(>|z|)"]
```
:::

Collecting the two tests and the two odds ratios on interpretable scales:

::: {.callout-note collapse="true"}
### Code for the table below
```{r additive-inference}
#| output: false
inference_table <- tibble(
  coefficient = c("balance (per $100)", "student (Yes vs. No)"),
  estimate    = c(100 * ba["balance"], ba["studentYes"]),
  z           = summary(additive_glm)$coefficients[c("balance", "studentYes"),
                                                   "z value"],
  odds_ratio  = c(or_balance_100, or_student_adj),
  lower       = c(ci_or_balance_100[1], ci_or_student_adj[1]),
  upper       = c(ci_or_balance_100[2], ci_or_student_adj[2])
) |>
  mutate(across(c(estimate, z, odds_ratio, lower, upper),
                \(x) sprintf("%.3f", x))) |>
  knitr::kable(col.names = c("Coefficient", "Estimate (log-odds)", "z",
                             "Odds ratio", "95% CI lower", "95% CI upper"),
               align = "lccccc")
```
:::

```{r additive-inference-table}
#| echo: false
inference_table
```

Both $z$-statistics are far from zero. The interval for the balance odds ratio
per \$100 lies entirely above $1$, and the interval for the student odds ratio
lies entirely below $1$, running from
`r sprintf("%.2f", ci_or_student_adj[1])` to
`r sprintf("%.2f", ci_or_student_adj[2])`.

### Drop-in-deviance test

A $z$-statistic tests one coefficient at a time. Asking whether a *group* of
coefficients is all zero at once — every dummy variable of a five-level
categorical feature, say — requires comparing two models rather than reading
one row of a coefficient table.

The multiple linear regression lecture answered that question with an F-test:
fit a **reduced** model without the group and a **full** model with it, and
compare their residual sums of squares. Logistic regression has no residual
sum of squares, but it has the quantity that plays the same role.

For a binary response, the **residual deviance** of a fitted logistic
regression is

$$D = -2\,\ell(\hat\beta),$$

where

- $\ell$ is the Bernoulli log-likelihood above and
- $\hat\beta$ is the maximum likelihood estimate.

`glm()` prints it as `Residual deviance` in the summary above, alongside
`Null deviance`, which is $-2\ell$ for the intercept-only model. Smaller
deviance means a larger maximized likelihood. Because a reduced model's
coefficients are a special case of the full model's (the dropped ones fixed
at zero), the full model's maximized likelihood is at least as large, so
adding features cannot increase the deviance — exactly as adding features
cannot increase RSS in a linear model.

The **drop-in-deviance** statistic compares a reduced model, nested in a full
model with $k$ more parameters, by

$$G^2 = D_r - D_f,$$

where

- $D_r$ is the residual deviance of the reduced model, and
- $D_f$ is the residual deviance of the full model.

Under $H_0$ that the reduced model is adequate — that all $k$ of the dropped
coefficients are zero — $G^2$ is approximately $\chi^2$ with $k$ degrees of
freedom in large samples, and the $p$-value is the upper tail probability
$P(\chi^2_k > G^2)$.

Two differences from the F-test are worth naming. The F-statistic divides the
drop in RSS *per degree of freedom* by $\hat\sigma^2_f$, because a linear
model has an error variance to estimate; a logistic regression has none, so
$G^2$ needs no such divisor and carries $k$ in its reference distribution
instead. And the F-distribution is exact under normal errors, while the
$\chi^2$ reference distribution here is a large-sample approximation.

In R, `anova()` on two fitted models with `test = "Chisq"` performs the test.
Applying it to this lecture's two models — reduced `default ~ balance`, full
`default ~ balance + student` — tests $H_0: \beta_2 = 0$, that student status
adds nothing once balance is in the model:

```{r deviance-test}
balance_glm <- glm(default ~ balance, data = Default, family = "binomial")
anova(balance_glm, additive_glm, test = "Chisq")
```

The same numbers assembled from the deviances directly:

```{r deviance-manual}
G2    <- deviance(balance_glm) - deviance(additive_glm)
k     <- length(coef(additive_glm)) - length(coef(balance_glm))
p_dev <- pchisq(G2, df = k, lower.tail = FALSE)

data.frame(G2 = G2, k = k, p_value = p_dev)
```

Adding `student` to the balance-only model drops the deviance by
$G^2 = `r sprintf("%.2f", G2)`$ on $k = `r k`$ degree of freedom, giving
$p = `r fmt_p(p_dev)`$. The data are strongly inconsistent with a model in
which student status contributes nothing beyond balance.

The $z$-test from the coefficient table tests the same $H_0$, since the two
models differ by one coefficient. It gives
$z = `r sprintf("%.2f", z_student_adj)`$, so $z^2 = `r sprintf("%.2f", z_student_adj^2)`$
against $G^2 = `r sprintf("%.2f", G2)`$, and $p = `r fmt_p(p_student_adj)`$
against $p = `r fmt_p(p_dev)`$. The two agree on the conclusion without
agreeing to the digit, and there is no reason they should: they are different
statistics, each approximately $\chi^2_1$ under the null rather than exactly
equal to one another.

When the two models differ by more than one coefficient, no single $z$-test
answers the question and the drop-in-deviance test is the tool that does.
Comparing the intercept-only model to the additive model, for instance, tests
$H_0: \beta_1 = \beta_2 = 0$ with $k = 2$, and the two deviances needed are the
`Null deviance` and `Residual deviance` lines already printed in the summary
above.

::: {.callout-note collapse="true"}
### Beyond this course
$G^2$ is a **likelihood-ratio statistic**: because
$D_r - D_f = -2\left[\ell(\hat\beta_r) - \ell(\hat\beta_f)\right] = -2\log\left(L(\hat\beta_r)/L(\hat\beta_f)\right)$,
the test compares the maximized likelihoods of the two models. Wilks' theorem
supplies the $\chi^2_k$ limiting distribution. The per-coefficient $z$-test is a
**Wald** test instead, built from $\hat\beta_j$ and its standard error rather
than from the likelihood at two parameter values, which is why the two statistics
differ in finite samples while sharing a null distribution in the limit.
:::

### Prediction

The fitted model returns a probability for any combination of feature values,

$$\hat p(x) = \frac{e^{x\hat\beta}}{1 + e^{x\hat\beta}},$$

where $x = (1, x_1, \ldots, x_p)$ is a row of feature values. In R,
`predict()` with `type = "response"` applies the logistic function; the
`newdata` argument takes a data frame with one column per feature, and a
categorical feature must be supplied at one of its observed levels.

Evaluating $\hat p(x)$ for a student and a non-student at three balances gives
six predictions, and the odds ratio between the two customers at each balance:

::: {.callout-note collapse="true"}
### Code for the table below
```{r additive-predictions}
#| output: false
pred_grid <- expand_grid(balance = c(1000, 1500, 2000),
                         student = factor(c("No", "Yes"),
                                          levels = levels(Default$student)))
pred_grid$p_hat <- predict(additive_glm, newdata = pred_grid,
                           type = "response")

pred_table <- pred_grid |>
  pivot_wider(names_from = student, values_from = p_hat,
              names_prefix = "student_") |>
  mutate(odds_ratio = (student_Yes / (1 - student_Yes)) /
                      (student_No  / (1 - student_No))) |>
  mutate(across(c(student_No, student_Yes), \(x) sprintf("%.4f", x)),
         odds_ratio = sprintf("%.3f", odds_ratio),
         balance = format(balance, big.mark = ",")) |>
  knitr::kable(col.names = c("Balance ($)", "Non-student", "Student",
                             "Odds ratio"),
               align = "cccc")

p_at <- function(b, s) {
  pred_grid$p_hat[pred_grid$balance == b & pred_grid$student == s]
}
```
:::

```{r additive-predictions-table}
#| echo: false
pred_table
```

At a balance of \$1,500 the estimated probability of default is
`r sprintf("%.1f", 100 * p_at(1500, "No"))`% for a non-student and
`r sprintf("%.1f", 100 * p_at(1500, "Yes"))`% for a student. At \$2,000 the two
estimates are `r sprintf("%.1f", 100 * p_at(2000, "No"))`% and
`r sprintf("%.1f", 100 * p_at(2000, "Yes"))`%.

The final column is constant down the table at `r round(or_student_adj, 3)`,
which is $e^{\hat\beta_2}$: the model forces the student-versus-non-student
odds ratio to be the same at every balance. The difference in *probabilities*
is not constant — `r sprintf("%.3f", p_at(1000, "No") - p_at(1000, "Yes"))` at
\$1,000 against `r sprintf("%.3f", p_at(2000, "No") - p_at(2000, "Yes"))` at
\$2,000 — for the same reason a logistic curve's slope varies with $x$.

Predicting across the whole range of balance for each student status shows what
a constant odds ratio looks like on the probability scale, with the binned
observed rates from the exploratory plot overlaid as a check.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r additive-curves}
#| output: false
curve_grid <- expand_grid(
  balance = seq(0, max(Default$balance), length.out = 400),
  student = factor(c("No", "Yes"), levels = levels(Default$student)))
curve_link <- predict(additive_glm, newdata = curve_grid,
                      type = "link", se.fit = TRUE)
curve_grid <- curve_grid |>
  mutate(p_hat = plogis(curve_link$fit),
         # lower/upper are not plotted below; reused for the confidence
         # ribbon in the "Balance confounds student status" worked example
         lower = plogis(curve_link$fit - qnorm(0.975) * curve_link$se.fit),
         upper = plogis(curve_link$fit + qnorm(0.975) * curve_link$se.fit))

curve_gap <- curve_grid |>
  select(balance, student, p_hat) |>
  pivot_wider(names_from = student, values_from = p_hat) |>
  mutate(gap = No - Yes)

max_gap          <- max(curve_gap$gap)
max_gap_balance  <- curve_gap$balance[which.max(curve_gap$gap)]

curve_plot <- ggplot(curve_grid,
                     aes(x = balance, y = p_hat,
                         color = student, linetype = student)) +
  geom_line(linewidth = 0.9) +
  geom_point(data = binned_rates,
             aes(y = rate, shape = student), size = 2.5,
             inherit.aes = TRUE) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Student", linetype = "Student", shape = "Student",
       title = "Fitted additive model with binned observed rates")
```
:::

```{r additive-curves-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
curve_plot
```

The two curves have the same shape, with the student curve shifted to the
right, and both sets of binned points track their own curve. On the log-odds
scale the curves are parallel lines separated by $\hat\beta_2$ at every
balance; on the probability scale they never cross, and the gap between them
closes at both ends. The gap is widest near a balance of
\$`r format(round(max_gap_balance), big.mark = ",")`, where it reaches
`r sprintf("%.3f", max_gap)` in estimated probability — not at the midpoint of
the balance range, but where both curves are passing through the steep part of
their climb.

A confidence interval for $p(x)$ goes through the same three scales as
the one-feature case, with $x$ now a full row of feature values.
`predict()` with `type = "link"` and `se.fit = TRUE` returns $\hat\eta(x) =
x\hat\beta$ and $\text{SE}(\hat\eta(x))$, giving a Wald interval for the
log-odds at $x$; exponentiating both endpoints carries it to an interval for
the **odds** at $x$; and applying $p = \text{odds}/(1+\text{odds})$ to both
endpoints of that carries it to an interval for $p(x)$ itself.

```{r additive-ci-walkthrough-setup}
x0 <- tibble(balance = 1500,
            student = factor("No", levels = levels(Default$student)))
```

For a non-student at a balance of \$`r format(x0$balance, big.mark = ",")`:

```{r additive-ci-walkthrough}
#| output: false
link_x0    <- predict(additive_glm, newdata = x0, type = "link", se.fit = TRUE)
eta_hat_x0 <- link_x0$fit
eta_ci_x0  <- eta_hat_x0 + c(-1, 1) * qnorm(0.975) * link_x0$se.fit
odds_ci_x0 <- exp(eta_ci_x0)
p_ci_x0    <- odds_ci_x0 / (1 + odds_ci_x0)

ci_walkthrough <- tibble(
  Scale    = c("Log-odds", "Odds", "P(default)"),
  Estimate = c(eta_hat_x0, exp(eta_hat_x0), plogis(eta_hat_x0)),
  Lower    = c(eta_ci_x0[1], odds_ci_x0[1], p_ci_x0[1]),
  Upper    = c(eta_ci_x0[2], odds_ci_x0[2], p_ci_x0[2])
) |>
  mutate(across(c(Estimate, Lower, Upper), \(x) sprintf("%.4f", x))) |>
  knitr::kable(col.names = c("Scale", "Estimate", "2.5%", "97.5%"),
               align = "lccc")
```

```{r additive-ci-walkthrough-table}
#| echo: false
ci_walkthrough
```

Nothing about the three-scale route changed by adding a second feature: $x$
now has three entries instead of two, but $\hat\eta(x)$, its standard error,
and the two increasing transforms that follow are exactly what the one-feature
lecture used.

## Balance confounds student status

**Question.** Does being a student change a customer's risk of default, or
does that comparison change once their balances are put on equal footing?

**Data.** The same `r format(nrow(Default), big.mark = ",")` customers as
throughout this lecture, `default` as the response, with `balance` and
`student` both as features.

The binned default rates from the exploratory analysis are worth having in
view again:

```{r student-recap-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
binned_plot
```

Within every bin the student curve sat at or below the non-student curve,
while a student's *balance* averaged higher than a non-student's — the two
descriptive facts a model with both features has to reconcile.

### Fitting both models

Both fits are already in hand: `student_glm`, with `student` alone, and
`additive_glm`, with `balance` and `student` together.

::: {.callout-note collapse="true"}
### Code for the table below
```{r student-vs-additive-table}
#| output: false
fits_comparison <- tibble(
  term = names(ba),
  "Student alone" = c(bs["(Intercept)"], NA, bs["studentYes"]),
  "Balance + student" = ba
) |>
  mutate(across(c("Student alone", "Balance + student"),
                \(x) ifelse(is.na(x), "", sprintf("%.4f", x)))) |>
  knitr::kable(col.names = c("Coefficient", "Student alone",
                             "Balance + student"),
               align = "lcc")
```
:::

```{r student-vs-additive-table-display}
#| echo: false
fits_comparison
```

The two intercepts are not comparable: each is the log-odds for a non-student
at whatever balance is $0$ in *that* model, and the student-alone model never
holds balance fixed at all.

### Fitted probabilities

Each model's fitted probabilities, against the same binned observed rates
used above:

::: {.callout-note collapse="true"}
### Code for the plots below
```{r default-prob-setup}
#| output: false
student_newdata <- tibble(student = factor(c("No", "Yes"),
                                           levels = levels(Default$student)))
student_link    <- predict(student_glm, newdata = student_newdata,
                           type = "link", se.fit = TRUE)
student_ci <- student_newdata |>
  mutate(eta   = student_link$fit,
         se    = student_link$se.fit,
         p_hat = plogis(eta),
         lower = plogis(eta - qnorm(0.975) * se),
         upper = plogis(eta + qnorm(0.975) * se))

student_band <- curve_grid |>
  select(balance, student) |>
  left_join(student_ci |> select(student, p_hat, lower, upper), by = "student")

default_prob_base <- ggplot(mapping = aes(x = balance, color = student,
                                          linetype = student)) +
  geom_point(data = binned_rates, aes(y = rate, shape = student, size = n)) +
  coord_cartesian(xlim = c(0, max(Default$balance)), ylim = c(0, 1)) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  scale_size_area(max_size = 5) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Student", linetype = "Student", shape = "Student",
       size = "Customers\nin bin", title = "ISLR2::Default")

default_prob_data <- default_prob_base

default_prob_student_alone <- default_prob_base +
  geom_ribbon(data = student_band, aes(ymin = lower, ymax = upper, fill = student),
             color = NA, alpha = 0.2) +
  geom_line(data = student_band, aes(y = p_hat), linewidth = 0.9) +
  scale_fill_manual(values = c("#0072B2", "#D55E00"), guide = "none")

default_prob_balance_fixed <- default_prob_base +
  geom_ribbon(data = curve_grid, aes(ymin = lower, ymax = upper, fill = student),
             color = NA, alpha = 0.2) +
  geom_line(data = curve_grid, aes(y = p_hat), linewidth = 0.9) +
  scale_fill_manual(values = c("#0072B2", "#D55E00"), guide = "none")
```
:::

::: {.panel-tabset}
#### Data
```{r default-prob-data-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
default_prob_data
```

#### Student alone
```{r default-prob-student-alone-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
default_prob_student_alone
```

#### Balance held fixed
```{r default-prob-balance-fixed-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
default_prob_balance_fixed
```
:::

Fit alone, `student` cannot respond to balance at all — its tab shows two flat
lines, and the "Yes" line sits *above* the "No" line, matching the higher
aggregate default rate students showed at the start of this lecture. Once
`balance` is in the model, its tab shows two curves that both climb with
balance, and now the "Yes" curve sits *below* the "No" curve at every balance
on the page. The direction of the student effect flips depending on whether
balance is held fixed.

### Coefficient reversal

Putting the pieces together algebraically, the fitted log-odds are

$$\log\left(\frac{\hat p(X)}{1 - \hat p(X)}\right) = `r sprintf("%.2f", ba[1])` `r sprintf("%+.6f", ba[2])` \times \texttt{balance} `r sprintf("%+.3f", ba[3])` \times D.$$

The intercept $\hat\beta_0 = `r sprintf("%.2f", ba[1])`$ is the estimated
log-odds of default for a non-student carrying no balance — both features at
zero — which corresponds to an estimated probability of
`r format(plogis(ba[1]), digits = 2, scientific = FALSE)`.

The other two coefficients are easier to read as odds ratios, with `balance`
rescaled to \$100 increments as in the previous lecture:

```{r additive-odds-ratios}
round(c("balance, per $100" = or_balance_100,
        "student vs. non-student" = or_student_adj), 3)
```

Holding student status fixed, an additional \$100 of balance multiplies the
estimated odds of default by `r sprintf("%.2f", or_balance_100)`.

Holding balance fixed, a student's estimated odds of default are
`r sprintf("%.2f", or_student_adj)` times a non-student's — a
`r round(100 * (1 - or_student_adj))`% *reduction* in the odds. Fit alone,
student status multiplied the odds by
`r sprintf("%.2f", or_student_alone)`, a
`r round(100 * (or_student_alone - 1))`% increase. The sign of the `student`
coefficient reversed, from $`r sprintf("%.3f", bs[2])`$ to
$`r sprintf("%.3f", ba[3])`$, and nothing about the data changed between the
two fits — only which features were held fixed when reading the coefficient.

A marginal association that reverses direction once a third variable is held
fixed, at every level of that third variable, is an instance of **Simpson's
paradox**.

::: {.callout-note collapse="true"}
### Code for the table below
```{r comparison-table}
#| output: false
comparison_table <- tibble(
  Model = c("`default ~ student`", "`default ~ balance + student`"),
  Compares = c("a student against a non-student, balances as they happen to be",
               "a student against a non-student **at the same balance**"),
  OR = sprintf("%.2f", c(or_student_alone, or_student_adj))
) |>
  knitr::kable(col.names = c("Model", "What the `student` coefficient compares",
                              "Estimated odds ratio"),
               align = "lll")
```
:::

```{r comparison-table-display}
#| echo: false
comparison_table
```

The two descriptive summaries from the exploratory analysis supply the
mechanism. Students carry
\$`r format(round(mean_balance_of("Yes") - mean_balance_of("No")), big.mark = ",")`
more balance on average, and balance raises the odds of default sharply. A
comparison of students to non-students that does not hold balance fixed is
therefore partly a comparison of higher balances to lower balances. Once
balance is in the model and held fixed, that path is closed, and what remains
is the within-balance comparison the binned plot already showed: at a given
balance, students default less.

This is **confounding**: a feature left out of the model is associated both
with the feature being interpreted and with the response, so the included
feature's coefficient reports the two associations combined rather than its
own. In the one-feature model, `student` stood in for balance, and the
absorbed part was large enough and of the opposite sign to flip the total.

::: {.callout-note collapse="true"}
### Choosing which coefficient to report
Neither is wrong; they answer different questions, and the question decides.

A credit card company deciding what a *new applicant's* student status implies
about risk, knowing nothing else about them, wants the one-feature answer:
among the customers we see, students default more. A company setting a credit
limit for a customer whose balance it already knows wants the additive answer:
at that balance, a student is the safer of the two.

What is not defensible is reporting one number and describing it with the
other's interpretation.
:::

## Conclusion

Multiple logistic regression passes multiple linear regression's linear
predictor $X\beta$ through the logistic function, estimated by the same
Bernoulli maximum likelihood as the one-feature case. Each $\beta_j$ is a
change in log-odds, and $e^{\beta_j}$ an odds ratio, *holding every other
feature in the model fixed* — the qualifier that distinguishes a coefficient
here from the same feature's coefficient fit alone, and the reason the two can
disagree in magnitude or even in sign.

Testing several coefficients at once needs more than a $z$-statistic per
coefficient: the drop-in-deviance test compares a reduced and a full model's
residual deviances the way an F-test compares two models' residual sums of
squares, with a $\chi^2$ reference distribution in place of an $F$.

A coefficient's sign can reverse between a model that omits a feature and one
that includes it whenever the omitted feature is associated with both the
included feature and the response — confounding, and when the association
runs in the opposite direction at every level of the third variable, an
instance of Simpson's paradox. The question being asked decides which
comparison — and which coefficient — is the right one to report.

The additive model forces every feature's odds ratio to be the same regardless
of the other features' values; relaxing that constraint requires interactions
between features.
