# Multiple Logistic Regression

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **write** the multiple logistic regression model in its probability and
log-odds forms using the linear predictor $X_i\beta$,
- **encode** a two-level categorical feature as an indicator and **interpret**
its coefficient as an odds ratio relative to the reference level,
- **interpret** $\beta_j$ as a change in log-odds and $e^{\beta_j}$ as an odds
ratio *holding every other feature fixed*, and **distinguish** that from the
interpretation of the same feature's coefficient fit alone,
- **fit** an additive multiple logistic regression with `glm()` and **predict**
$\hat p(x)$ for a specified combination of feature values,
- **diagnose** a coefficient whose sign reverses between a one-feature and a
multi-feature model as confounding, and **explain** the mechanism from
descriptive summaries of the features, and
- **test** $H_0: \beta_j = 0$ with a $z$-statistic and **interpret** a
confidence interval for the odds ratio $e^{\beta_j}$ as a statement about
feature $j$ after accounting for the other features.
:::

## More than one feature

The previous lecture modeled the probability of default from a single feature,
credit card balance. `ISLR2::Default` records two more variables on the same
`r format(nrow(Default), big.mark = ",")` customers:

- `student`: is the customer a student? A factor with levels
`r paste(levels(Default$student), collapse = " and ")`.
- `income`: the customer's income, in dollars. This one is set aside for this
lecture; everything below uses `balance` and `student`.

Having more than one feature raises a question that a one-feature model cannot
answer: what does each feature tell us that the others do not? This lecture
answers it for an **additive** model, in which each feature contributes its own
term to the linear predictor.

## Categorical features

A categorical feature enters a logistic regression exactly as it enters a
multiple linear regression: through dummy variables against a reference level.
A feature with $k$ levels contributes $k-1$ indicators, and the level left
without an indicator is the reference.

`student` has two levels, so it contributes one indicator. Taking
`r levels(Default$student)[1]` as the reference level,

$$D_i = \mathrm{I}(\texttt{student}_i = \texttt{Yes}) = \begin{cases} 1 & \text{customer } i \text{ is a student} \\ 0 & \text{otherwise,}\end{cases}$$

where

- $D_i$ is the indicator for observation $i$, and
- $\mathrm{I}(\cdot)$ is $1$ when its argument is true and $0$ otherwise.

R builds this column automatically from a factor, using the first level as the
reference, and names the coefficient after the level being indicated —
`studentYes` below.

## Example: student status and default

**Question.** Are students more likely to default than non-students?

**Data.** The same `r format(nrow(Default), big.mark = ",")` customers as in the
previous lecture, now using `student` as the feature and `default` as the
response, with $Y_i = 1$ when customer $i$ defaults.

::: {.callout-note collapse="true"}
### Code for the table below
```{r student-eda}
#| output: false
Default <- Default |>
  mutate(default01 = as.numeric(default == "Yes"))

student_rates <- Default |>
  group_by(student) |>
  summarize(n = n(),
            defaults = sum(default01),
            rate = mean(default01),
            .groups = "drop")

student_rate_table <- student_rates |>
  mutate(n = format(n, big.mark = ","),
         rate = sprintf("%.2f%%", 100 * rate)) |>
  knitr::kable(col.names = c("Student", "Customers", "Defaults",
                             "Percent defaulting"),
               align = "cccc")

rate_of <- function(s) student_rates$rate[student_rates$student == s]
```
:::

```{r student-eda-table}
#| echo: false
student_rate_table
```

Students default at `r sprintf("%.2f", 100 * rate_of("Yes"))`% versus
`r sprintf("%.2f", 100 * rate_of("No"))`% for non-students, so the student
default rate is `r sprintf("%.2f", rate_of("Yes") / rate_of("No"))` times the
non-student rate. That is a ratio of *rates*, not of odds; the two are
different quantities, and they happen to be close here only because both rates
are small.

Fitting the simple logistic regression with $D$ as the feature turns that
comparison into a coefficient:

```{r student-glm}
student_glm <- glm(default ~ student, data = Default, family = "binomial")
summary(student_glm)$coefficients
```

::: {.callout-note collapse="true"}
### Code for the estimates quoted below
```{r student-glm-extract}
#| output: false
bs <- coef(student_glm)
or_student_alone <- unname(exp(bs["studentYes"]))
ci_student_alone <- exp(confint.default(student_glm)["studentYes", ])
```
:::

The estimated log-odds of default are

$$\log\left(\frac{\hat p(D)}{1 - \hat p(D)}\right) = `r sprintf("%.2f", bs[1])` `r sprintf("%+.3f", bs[2])` \, D.$$

With a single indicator feature, the two fitted values are the two observed
log-odds: $\hat\beta_0$ is the log-odds of default among non-students, and
$\hat\beta_0 + \hat\beta_1$ is the log-odds among students. The coefficient
$\hat\beta_1 = `r sprintf("%.3f", bs[2])`$ is the difference between them, so
$e^{\hat\beta_1} = `r sprintf("%.2f", or_student_alone)`$ is the odds of
default for a student divided by the odds for a non-student. The confidence
interval for that odds ratio runs from
`r sprintf("%.2f", ci_student_alone[1])` to
`r sprintf("%.2f", ci_student_alone[2])`, entirely above $1$.

**Conclusion.** Taken by itself, student status is associated with a
`r round(100 * (or_student_alone - 1))`% increase in the odds of default.

## Two features at once

The previous lecture found balance strongly associated with default; this
lecture has now found student status associated with default as well. Before
putting both in one model, start with the individual customers rather than a
summary of them: the same jittered scatter of default against balance from
the previous lecture, now split into two panels by student status.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r student-default-raw}
#| output: false
raw_plot <- ggplot(Default, aes(x = balance, y = default01)) +
  geom_jitter(height = 0.03, width = 0, alpha = 0.08, size = 0.7) +
  facet_wrap(~ student, labeller = label_both) +
  scale_y_continuous(breaks = c(0, 1), labels = c("No (0)", "Yes (1)")) +
  labs(x = "Credit card balance ($)", y = "Default",
       title = "ISLR2::Default")
```
:::

```{r student-default-raw-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
raw_plot
```

The student panel's points sit further right on average, a first hint of a
balance difference between the two groups, but with
`r format(sum(Default$student == "Yes"), big.mark = ",")` students against
`r format(sum(Default$student == "No"), big.mark = ",")` non-students plotted
at the same jitter and transparency, that shift is hard to read precisely by
eye. Two descriptive summaries make it precise.

The first compares the distribution of balance for students and non-students.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r balance-by-student}
#| output: false
balance_means <- Default |>
  group_by(student) |>
  summarize(mean_balance = mean(balance), .groups = "drop")

balance_density_plot <- ggplot(Default,
                               aes(x = balance,
                                   color = student, linetype = student)) +
  geom_density(linewidth = 0.9) +
  geom_vline(data = balance_means,
             aes(xintercept = mean_balance,
                 color = student, linetype = student),
             linewidth = 0.5, show.legend = FALSE) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Credit card balance ($)", y = "Density",
       color = "Student", linetype = "Student",
       title = "ISLR2::Default")

mean_balance_of <- function(s) {
  balance_means$mean_balance[balance_means$student == s]
}
```
:::

```{r balance-by-student-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
balance_density_plot
```

The two densities are shifted relative to one another, with the vertical lines
marking the group means: students average
\$`r format(round(mean_balance_of("Yes")), big.mark = ",")` versus
\$`r format(round(mean_balance_of("No")), big.mark = ",")` for non-students, a
difference of
\$`r format(round(mean_balance_of("Yes") - mean_balance_of("No")), big.mark = ",")`.
Students carry more balance.

The second summary asks what happens to the default rate within a narrow range
of balance. Splitting balance into \$400-wide bins and computing the observed
default rate separately for students and non-students within each bin gives a
model-free comparison of two customers who carry similar balances. Bins holding
too few customers of a given student status to estimate a rate are dropped.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r binned-rates}
#| output: false
min_bin_n <- 100

binned_rates <- Default |>
  mutate(bin = cut(balance, breaks = seq(0, 2800, by = 400),
                   include.lowest = TRUE)) |>
  group_by(student, bin) |>
  summarize(balance = mean(balance), rate = mean(default01), n = n(),
            .groups = "drop") |>
  filter(n >= min_bin_n)

binned_plot <- ggplot(binned_rates,
                      aes(x = balance, y = rate,
                          color = student, shape = student,
                          linetype = student)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Mean balance within bin ($)",
       y = "Observed proportion defaulting",
       color = "Student", shape = "Student", linetype = "Student",
       title = "ISLR2::Default, $400-wide balance bins")
```
:::

```{r binned-rates-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
binned_plot
```

Within every bin shown, the student curve sits at or below the non-student
curve. Comparing customers who carry similar balances reverses the direction of
the association that the aggregate rates showed. The curves stop near
\$`r format(round(max(binned_rates$balance)), big.mark = ",")` because every
bin beyond that holds fewer than `r min_bin_n` customers in at least one of the
two groups.

Neither summary used a model. Together they say that students carry higher
balances, that higher balances default more, and that at a fixed balance
students default less; a model with both features lets us separate those
statements.

## The multiple logistic regression model

Let $X_{i1}, \ldots, X_{ip}$ be the $p$ features recorded on observation $i$.
The **multiple logistic regression** model is

$$p(X_i) = P(Y_i = 1|X_i) = \frac{e^{\beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}}}{1 + e^{\beta_0 + \beta_1 X_{i1} + \cdots + \beta_p X_{ip}}},$$

where

- $p(X_i)$ is the probability the event occurs for observation $i$,
- $X_{ij}$ is the value of feature $j$ for observation $i$, and
- $\beta_0, \beta_1, \ldots, \beta_p$ are the regression coefficients.

The same algebra as in the previous lecture — form $1 - p(X_i)$, divide, take
logarithms — puts the model in its log-odds form,

$$\log\left(\frac{p(X)}{1 - p(X)}\right) = \beta_0 + \beta_1 X_1 + \cdots + \beta_p X_p.$$

The right side is exactly the linear predictor of the multiple linear
regression lecture. Using that lecture's notation,

- $X_i = (1, X_{i1}, \ldots, X_{ip})$ is the $i$th row of the $n \times (p+1)$
model matrix $X$,
- $\beta = (\beta_0, \ldots, \beta_p)^\top$ is the vector of coefficients, and
- $\eta = X\beta$ is the $n$-vector of linear predictors, with $i$th element
$\eta_i = X_i\beta$,

the model is

$$p(X_i) = \text{logistic}(X_i\beta) = \frac{e^{X_i\beta}}{1 + e^{X_i\beta}}.$$

So multiple logistic regression is multiple linear regression's linear
predictor passed through the logistic function. Everything the linear predictor
can express — several quantitative features, dummy variables for categorical
features, polynomial terms, step functions — is available here unchanged,
because none of it touches the logistic function wrapped around it.

The model above is **additive**: each feature contributes its own term
$\beta_j X_{ij}$ to $\eta_i$, and no term involves two features at once.

## Estimation

The observations are still independent Bernoulli trials,
$Y_i \stackrel{ind}{\sim} \text{Bernoulli}(p(x_i))$, so the log-likelihood
derived in the previous lecture carries over with $\beta_0 + \beta_1 x_i$
replaced by $x_i\beta$:

$$\ell(\beta) = \ell(\beta_0, \beta_1, \ldots, \beta_p) = \sum_{i=1}^n \left\{ y_i \, x_i\beta - \log\left(1 + e^{x_i\beta}\right)\right\},$$

where

- $x_i = (1, x_{i1}, \ldots, x_{ip})$ is the observed feature row for
observation $i$, and
- $y_i \in \{0, 1\}$ is the observed response.

Nothing structural changed. There are now $p+1$ parameters rather than $2$,
still no $\sigma^2$, and still no closed-form maximizer, because $p(x_i)$
remains a nonlinear function of the coefficients. The **maximum likelihood
estimator** $\hat\beta$ is the maximizer of $\ell$, and `glm()` finds it
numerically by the same iterative algorithm as before.

Differentiating with respect to $\beta_j$ and setting the result to zero gives
one score equation per coefficient,

$$\sum_{i=1}^n x_{ij}\left[y_i - p(x_i)\right] = 0, \qquad j = 0, 1, \ldots, p,$$

with $x_{i0} = 1$. As in the one-feature case, these say the residuals
$y_i - p(x_i)$ sum to zero and are orthogonal to every feature — the same
conditions the least squares normal equations impose, though here the
equations are nonlinear in $\beta$.

## Interpreting the coefficients

Fix the values of all features other than $X_j$ and increase $X_j$ by one unit.
On the log-odds scale, every term except the $j$th is identical on both sides,
so the difference is

$$\left[\beta_0 + \cdots + \beta_j(x_j + 1) + \cdots + \beta_p x_p\right] - \left[\beta_0 + \cdots + \beta_j x_j + \cdots + \beta_p x_p\right] = \beta_j.$$

So $\beta_j$ is the change in the log-odds of the event per one-unit increase
in $X_j$ **holding every other feature fixed**, and exponentiating,
$e^{\beta_j}$ is the multiplicative change in the odds of the event per
one-unit increase in $X_j$ **holding every other feature fixed**.

The qualifier is the whole difference between this lecture and the last one,
and it is the same qualifier that attaches to a multiple linear regression
coefficient. The cancellation above is what licenses it: the comparison
$\beta_j$ describes is between two observations that differ in $X_j$ and agree
on everything else in the model.

Two consequences follow.

1. **$\beta_j$ from a multi-feature fit and from a one-feature fit answer
different questions.** Fit alone, a feature's coefficient describes the
comparison between observations that differ in that feature and in whatever
else happens to vary with it. In an additive model, it describes the
comparison holding the other included features fixed. The two can differ in
magnitude, and they can differ in sign.

2. **"Holding fixed" covers only features in the model.** A variable not in
$X$ is not held fixed by including other features, so a coefficient never
describes a comparison that adjusts for something the model never saw.

For a dummy variable — a feature $X_j$ taking only the values $0$ and $1$ — a
"one-unit increase" is the change from the reference level to the indicated
level, so $e^{\beta_j}$ is the odds of the event at that level divided by the
odds at the reference level, again holding the other features fixed.

The intercept $\beta_0$ is the log-odds of the event when every feature is $0$
— for a dummy variable, when the observation is at the reference level.

## Fitting an additive model in R

```{r additive-syntax}
#| eval: false
glm(default ~ balance + student, data = Default, family = "binomial")
```

The `+` makes the model additive: `balance` and `student` each contribute one
term to the linear predictor. Terms joined by `*` or `:` would add products of
features, which the additive model excludes.

### Example: fitting the additive model

```{r additive-glm}
additive_glm <- glm(default ~ balance + student, data = Default,
                    family = "binomial")
summary(additive_glm)
```

::: {.callout-note collapse="true"}
### Code for the estimates quoted below
```{r additive-extract}
#| output: false
ba <- coef(additive_glm)
or_balance_100 <- unname(exp(100 * ba["balance"]))
or_student_adj <- unname(exp(ba["studentYes"]))
ci_adj <- confint.default(additive_glm)
ci_or_student_adj <- exp(ci_adj["studentYes", ])
ci_or_balance_100 <- exp(100 * ci_adj["balance", ])
```
:::

The fitted log-odds are

$$\log\left(\frac{\hat p(X)}{1 - \hat p(X)}\right) = `r sprintf("%.2f", ba[1])` `r sprintf("%+.6f", ba[2])` \times \texttt{balance} `r sprintf("%+.3f", ba[3])` \times D,$$

with $X_1 = \texttt{balance}$ the quantitative feature and $X_2 = D$ the
student indicator, so $p = 2$.

The `studentYes` coefficient is $`r sprintf("%.3f", ba[3])`$. Fit alone it was
$`r sprintf("%.3f", bs[2])`$. The sign reversed.

## Confounding: a coefficient that changes sign

Nothing about the data changed between the one-feature and additive fits —
only which features were held fixed when reading the `student` coefficient.
The reversal has to be explained, not just reported.

### Example: interpreting the additive coefficients

```{r additive-odds-ratios}
round(c("balance, per $100" = or_balance_100,
        "student vs. non-student" = or_student_adj), 3)
```

Holding student status fixed, an additional \$100 of balance multiplies the
estimated odds of default by `r sprintf("%.2f", or_balance_100)`.

Holding balance fixed, a student's estimated odds of default are
`r sprintf("%.2f", or_student_adj)` times a non-student's — a
`r round(100 * (1 - or_student_adj))`% *reduction* in the odds. Fit alone,
student status multiplied the odds by
`r sprintf("%.2f", or_student_alone)`, a
`r round(100 * (or_student_alone - 1))`% increase.

The two estimates are not in conflict. They are estimates of different
comparisons.

::: {.callout-note collapse="true"}
### Code for the table below
```{r comparison-table}
#| output: false
comparison_table <- tibble(
  Model = c("`default ~ student`", "`default ~ balance + student`"),
  Compares = c("a student against a non-student, balances as they happen to be",
               "a student against a non-student **at the same balance**"),
  OR = sprintf("%.2f", c(or_student_alone, or_student_adj))
) |>
  knitr::kable(col.names = c("Model", "What the `student` coefficient compares",
                              "Estimated odds ratio"),
               align = "lll")
```
:::

```{r comparison-table-display}
#| echo: false
comparison_table
```

The two descriptive summaries above supply the mechanism. Students carry
\$`r format(round(mean_balance_of("Yes") - mean_balance_of("No")), big.mark = ",")`
more balance on average, and balance raises the odds of default sharply. A
comparison of students to non-students that does not hold balance fixed is
therefore partly a comparison of higher balances to lower balances. Once
balance is in the model and held fixed, that path is closed, and what remains
is the within-balance comparison the binned plot already showed: at a given
balance, students default less.

This is **confounding**: a feature left out of the model is associated both
with the feature being interpreted and with the response, so the included
feature's coefficient reports the two associations combined rather than its
own. In the one-feature model, `student` stood in for balance, and the
absorbed part was large enough and of the opposite sign to flip the total.

::: {.callout-note collapse="true"}
### Which coefficient is the right one to report?
Neither is wrong; they answer different questions, and the question decides.

A credit card company deciding what a *new applicant's* student status implies
about risk, knowing nothing else about them, wants the one-feature answer:
among the customers we see, students default more. A company setting a credit
limit for a customer whose balance it already knows wants the additive answer:
at that balance, a student is the safer of the two.

What is not defensible is reporting one number and describing it with the
other's interpretation.
:::

## Inference

Inference proceeds coefficient by coefficient with the same machinery as the
previous lecture. Maximum likelihood estimators are approximately normal in
large samples, so

$$z = \frac{\hat\beta_j - \beta_j^0}{\text{SE}(\hat\beta_j)}$$

is approximately standard normal under $H_0: \beta_j = \beta_j^0$, where

- $\beta_j^0$ is the null value, almost always $0$, and
- $\text{SE}(\hat\beta_j)$ is the standard error computed by `glm()` from the
curvature of the log-likelihood at $\hat\beta$.

An approximate $(1-\alpha)100\%$ confidence interval is
$\hat\beta_j \pm z_{1-\alpha/2}\text{SE}(\hat\beta_j)$, and exponentiating both
endpoints gives an interval for the odds ratio $e^{\beta_j}$.

Every one of these statements is about $\beta_j$ with the other features in the
model held fixed, so the test of $H_0: \beta_j = 0$ asks whether feature $j$ is
associated with the response *after accounting for* the other features — not
whether it is associated with the response at all.

### Example: testing the additive coefficients

::: {.callout-note collapse="true"}
### Code for the table below
```{r additive-inference}
#| output: false
inference_table <- tibble(
  coefficient = c("balance (per $100)", "student (Yes vs. No)"),
  estimate    = c(100 * ba["balance"], ba["studentYes"]),
  z           = summary(additive_glm)$coefficients[c("balance", "studentYes"),
                                                   "z value"],
  odds_ratio  = c(or_balance_100, or_student_adj),
  lower       = c(ci_or_balance_100[1], ci_or_student_adj[1]),
  upper       = c(ci_or_balance_100[2], ci_or_student_adj[2])
) |>
  mutate(across(c(estimate, z, odds_ratio, lower, upper),
                \(x) sprintf("%.3f", x))) |>
  knitr::kable(col.names = c("Coefficient", "Estimate (log-odds)", "z",
                             "Odds ratio", "95% CI lower", "95% CI upper"),
               align = "lccccc")
```
:::

```{r additive-inference-table}
#| echo: false
inference_table
```

Both $z$-statistics are far from zero. The interval for the balance odds ratio
per \$100 lies entirely above $1$, and the interval for the student odds ratio
lies entirely below $1$, running from
`r sprintf("%.2f", ci_or_student_adj[1])` to
`r sprintf("%.2f", ci_or_student_adj[2])`.

The one-feature fit produced an interval for the student odds ratio entirely
above $1$ (`r sprintf("%.2f", ci_student_alone[1])` to
`r sprintf("%.2f", ci_student_alone[2])`), and the additive fit produces one
entirely below $1$. The two intervals do not overlap, and that is not a
contradiction: they are intervals for different quantities.

## Prediction

The fitted model returns a probability for any combination of feature values,

$$\hat p(x) = \frac{e^{x\hat\beta}}{1 + e^{x\hat\beta}},$$

where $x = (1, x_1, \ldots, x_p)$ is a row of feature values. In R,
`predict()` with `type = "response"` applies the logistic function; the
`newdata` argument takes a data frame with one column per feature, and a
categorical feature must be supplied at one of its observed levels.

### Example: comparing a student and a non-student at the same balance

::: {.callout-note collapse="true"}
### Code for the table below
```{r additive-predictions}
#| output: false
pred_grid <- expand_grid(balance = c(1000, 1500, 2000),
                         student = factor(c("No", "Yes"),
                                          levels = levels(Default$student)))
pred_grid$p_hat <- predict(additive_glm, newdata = pred_grid,
                           type = "response")

pred_table <- pred_grid |>
  pivot_wider(names_from = student, values_from = p_hat,
              names_prefix = "student_") |>
  mutate(odds_ratio = (student_Yes / (1 - student_Yes)) /
                      (student_No  / (1 - student_No))) |>
  mutate(across(c(student_No, student_Yes), \(x) sprintf("%.3f", x)),
         odds_ratio = sprintf("%.3f", odds_ratio),
         balance = format(balance, big.mark = ",")) |>
  knitr::kable(col.names = c("Balance ($)", "Non-student", "Student",
                             "Odds ratio"),
               align = "cccc")

p_at <- function(b, s) {
  pred_grid$p_hat[pred_grid$balance == b & pred_grid$student == s]
}
```
:::

```{r additive-predictions-table}
#| echo: false
pred_table
```

At a balance of \$1,500 the estimated probability of default is
`r sprintf("%.1f", 100 * p_at(1500, "No"))`% for a non-student and
`r sprintf("%.1f", 100 * p_at(1500, "Yes"))`% for a student. At \$2,000 the two
estimates are `r sprintf("%.1f", 100 * p_at(2000, "No"))`% and
`r sprintf("%.1f", 100 * p_at(2000, "Yes"))`%.

The final column is constant down the table at `r round(or_student_adj, 3)`,
which is $e^{\hat\beta_2}$: the model forces the student-versus-non-student
odds ratio to be the same at every balance. The difference in *probabilities*
is not constant — `r sprintf("%.3f", p_at(1000, "No") - p_at(1000, "Yes"))` at
\$1,000 against `r sprintf("%.3f", p_at(2000, "No") - p_at(2000, "Yes"))` at
\$2,000 — for the same reason a logistic curve's slope varies with $x$.

Plotting $\hat p$ across the range of balance for each student status shows
what a constant odds ratio looks like on the probability scale.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r additive-curves}
#| output: false
curve_grid <- expand_grid(
  balance = seq(0, max(Default$balance), length.out = 400),
  student = factor(c("No", "Yes"), levels = levels(Default$student)))
curve_grid$p_hat <- predict(additive_glm, newdata = curve_grid,
                            type = "response")

curve_gap <- curve_grid |>
  pivot_wider(names_from = student, values_from = p_hat) |>
  mutate(gap = No - Yes)

max_gap          <- max(curve_gap$gap)
max_gap_balance  <- curve_gap$balance[which.max(curve_gap$gap)]

curve_plot <- ggplot(curve_grid,
                     aes(x = balance, y = p_hat,
                         color = student, linetype = student)) +
  geom_line(linewidth = 0.9) +
  geom_point(data = binned_rates,
             aes(y = rate, shape = student), size = 2.5,
             inherit.aes = TRUE) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Student", linetype = "Student", shape = "Student",
       title = "Fitted additive model with binned observed rates")
```
:::

```{r additive-curves-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
curve_plot
```

The two curves have the same shape, with the student curve shifted to the
right. On the log-odds scale they are parallel lines separated by
$\hat\beta_2$ at every balance; on the probability scale they never cross, and
the gap between them closes at both ends. The gap is widest near a balance of
\$`r format(round(max_gap_balance), big.mark = ",")`, where it reaches
`r sprintf("%.3f", max_gap)` in estimated probability — not at the midpoint of
the balance range, but where both curves are passing through the steep part of
their climb. The binned observed rates from the exploratory plot are overlaid,
and both sets of points track their own curve.

## Conclusion

Student status looked like a risk factor on its own, multiplying the odds of
default by `r sprintf("%.2f", or_student_alone)`, and looked like a protective
factor once balance was held fixed, multiplying the odds by
`r sprintf("%.2f", or_student_adj)`. The reversal is confounding: students
carry
\$`r format(round(mean_balance_of("Yes") - mean_balance_of("No")), big.mark = ",")`
more balance on average, so the one-feature coefficient was reporting balance's
association with default alongside student status's own.

The model that delivered this is the linear predictor from multiple linear
regression, passed through the logistic function, estimated by the same
Bernoulli maximum likelihood as the one-feature case, with each $\beta_j$
interpreted as a change in log-odds holding the other features fixed.

The additive model constrains the student-versus-non-student odds ratio to be
identical at every balance; relaxing that constraint requires interactions
between features.
