# Flexible Logistic Regression

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
set.seed(20260919)
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **explain** how interactions, polynomials, and step functions all enter
logistic regression as basis-function columns of the linear predictor, and
**interpret** a coefficient on any such column as a change in log-odds,
- **distinguish** categorical-categorical, continuous-categorical, and
continuous-continuous interactions, and **explain** why a coefficient's raw
magnitude does not by itself indicate whether an interaction matters,
- **write** a logistic regression model with an interaction between a
quantitative feature and an indicator, **derive** the two group-specific
log-odds slopes $\beta_1$ and $\beta_1 + \beta_3$, and **interpret**
$e^{\beta_1}$ and $e^{\beta_1 + \beta_3}$ as group-specific odds ratios,
- **compare** an additive fit against an interaction fit on the
linear-predictor scale, where the additive model forces two *parallel* lines,
and **translate** that comparison onto the probability scale and the decision
boundary,
- **test** a single interaction coefficient with a $z$-statistic and several
at once with a drop-in-deviance test, **recognizing** that a plausible
interaction often turns out not to be, and
- **describe**, from a picture, how a fitted curve changes as a polynomial's
degree or a step function's number of bins increases.
:::

Additive logistic regression gives each feature its own term in the linear
predictor and assumes nothing else. Three ways to relax that assumption carry
over from linear regression unchanged: **interactions**, which let one
feature's effect depend on another feature's value; **polynomials**, which let
a single feature's effect curve instead of running straight; and **step
functions**, which let a single feature's effect jump between bins instead of
following any smooth curve at all.

Each of these is a choice of **basis function** $h_j(X)$ from the feature
engineering lecture — a squared term, a product of two features, an indicator
for a range of $X$ — entering the linear predictor as its own column,

$$\eta = \beta_0 + \beta_1 h_1(X) + \beta_2 h_2(X) + \cdots.$$

Logistic regression changes only what happens to $\eta$ next:
$p(X) = \text{logistic}(\eta)$ in place of $E[Y \mid X] = \eta$, and
$\hat\beta$ maximizes the Bernoulli log-likelihood in place of minimizing a sum
of squared residuals. Every coefficient is still read the way earlier lectures
read it — $\beta_j$ a change in log-odds per unit of $h_j(X)$ with every other
column held fixed, $e^{\beta_j}$ the corresponding odds ratio — whatever $h_j$
turns out to be. That "held fixed" is worth flagging now: it breaks down
whenever two columns move together, which is exactly what happens for $X$ and
$X^2$, or for $X$ and $XD$, so it will need qualifying again once those columns
appear below. What changes from one choice of $h_j$ to the next is only what a
"one-unit change" of it means.

This lecture works through interactions first, since they raise the most that
is genuinely new, then polynomials and step functions more briefly, since
transferring them needs nothing beyond what this introduction already said.

## Interactions

### Additive assumption

The previous lecture's model was **additive**: each feature contributed its own
term $\beta_j X_{ij}$ to the linear predictor, and no term involved two
features at once. That structure is what licensed the interpretation "a
one-unit increase in $X_j$ multiplies the odds by $e^{\beta_j}$, holding every
other feature fixed" — the same $e^{\beta_j}$ at every value of every other
feature.

That is an assumption, not a convenience. The fitted curves of the previous
lecture showed it directly: two curves of identical shape, separated by a
constant $\hat\beta_2$ on the log-odds scale at every balance.

There is no reason a real process has to work that way. Whether it does is a
question about the data, and it can be asked and answered with a model that
does not impose it. What that model looks like depends on whether the two
interacting features are categorical or quantitative.

### Categorical-categorical interactions

When both features are categorical, an interaction adds one indicator for
every combination of their non-reference levels beyond what the two main
effects already supply. With indicators $D_1$ and $D_2$ for two binary
features,

$$\eta = \beta_0 + \beta_1 D_1 + \beta_2 D_2 + \beta_3 D_1 D_2,$$

and the product $D_1 D_2$ is $1$ only when an observation is at the
non-reference level of both features, so $\beta_3$ is the *extra* log-odds a
case gets from being at both non-reference levels at once, beyond what
$\beta_1$ and $\beta_2$ already give it separately.

A quantitative feature becomes categorical the same way it did in the feature
engineering lecture: pick a cutoff and take an indicator. Splitting `balance`
at \$1,500 turns the running `Default` example into a categorical-categorical
case, `balance_cat * student`:

```{r cat-cat-fit}
Default <- Default |>
  mutate(balance_cat = factor(ifelse(balance > 1500, "High", "Low"),
                              levels = c("Low", "High")))
balance_cat_glm <- glm(default ~ balance_cat * student, data = Default,
                       family = "binomial")
summary(balance_cat_glm)$coefficients
```

```{r cat-cat-extract}
#| include: false
bc_b <- coef(balance_cat_glm)
bc_z <- summary(balance_cat_glm)$coefficients["balance_catHigh:studentYes", "z value"]
```

The reference cell is a non-student with balance at or below \$1,500, at
log-odds $\hat\beta_0 = `r sprintf("%.3f", bc_b[1])`$. Crossing \$1,500 shifts
that by $\hat\beta_1 = `r sprintf("%.3f", bc_b[2])`$, being a student shifts it
by $\hat\beta_2 = `r sprintf("%.3f", bc_b[3])`$, and being *both* a
high-balance customer *and* a student adds a further
$\hat\beta_3 = `r sprintf("%.3f", bc_b[4])`$ beyond the two separately — small
here, and one the coefficient table above cannot distinguish from zero
($z = `r sprintf("%.2f", bc_z)`$). Both features are already $0/1$, so
$\hat\beta_3$ carries no units to worry about and comparing its size to
$\hat\beta_1$ or $\hat\beta_2$ directly is fair; that comparison stops being
fair once a quantitative feature is involved, as the continuous-continuous
subsection below shows.

Four coefficients describe four cells, so the whole fit can be drawn: one
point per cell, at the log-odds of that cell's default rate, with the two
balance categories in separate panels and the student status on the
horizontal axis.

::: {.callout-note collapse="true"}
### Code
```{r cat-cat-setup}
#| output: false
# default01 and default_jit are used by figures throughout this lecture, so
# both columns are added here, at the first figure that needs either. The
# jitter is drawn once into a column rather than left to geom_jitter(), which
# would re-randomize every time a tab is drawn and make the same points land
# in different places from one tab to the next.
Default <- Default |>
  mutate(default01 = as.numeric(default == "Yes"),
         default_jit = default01 + runif(n(), -0.03, 0.03))

cat_add <- glm(default ~ balance_cat + student, data = Default,
               family = "binomial")

cat_cells <- expand_grid(
  balance_cat = factor(c("Low", "High"), levels = c("Low", "High")),
  student     = factor(c("No", "Yes"), levels = levels(Default$student)))

cat_obs <- Default |>
  group_by(balance_cat, student) |>
  summarize(rate = mean(default01), .groups = "drop") |>
  mutate(eta = log(rate / (1 - rate)))

cat_fit <- bind_rows(
  cat_cells |> mutate(Model = "Additive",
                      eta = predict(cat_add, newdata = cat_cells)),
  cat_cells |> mutate(Model = "Interaction",
                      eta = predict(balance_cat_glm, newdata = cat_cells)))

# Each panel gets its own half-log-odds window, both of the same width: the
# two balance categories are about four log-odds apart, so a common axis
# would flatten every student difference in the figure to nothing, while
# equal window widths keep a segment's slope comparable across panels.
cat_window <- cat_obs |>
  group_by(balance_cat) |>
  summarize(mid = mean(range(eta)), .groups = "drop") |>
  reframe(balance_cat = rep(balance_cat, 2), eta = c(mid - 0.25, mid + 0.25)) |>
  mutate(student = factor("No", levels = levels(Default$student)))

cat_labels <- c(Low = "Balance up to $1,500", High = "Balance above $1,500")

# The same shared-base idea used throughout this lecture: one base layer
# carrying the observed cells, with each tab adding only its own fit.
cat_base <- ggplot(mapping = aes(x = student, y = eta,
                                 color = student, shape = student)) +
  geom_blank(data = cat_window) +
  geom_point(data = cat_obs, size = 3) +
  facet_wrap(~ balance_cat, scales = "free_y",
             labeller = labeller(balance_cat = cat_labels)) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Student", y = "Log-odds of default",
       color = "Student", shape = "Student", title = "ISLR2::Default")

cat_data <- cat_base

cat_additive <- cat_base +
  geom_line(data = filter(cat_fit, Model == "Additive"), aes(group = 1),
            color = "grey40", linetype = 2, linewidth = 0.8)

cat_interaction <- cat_base +
  geom_line(data = filter(cat_fit, Model == "Interaction"), aes(group = 1),
            color = "grey40", linetype = 2, linewidth = 0.8)
```
:::

::: {.panel-tabset}
#### Data
```{r cat-cat-data-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
cat_data
```

#### + Additive
```{r cat-cat-additive-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
cat_additive
```

#### + Interaction
```{r cat-cat-interaction-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
cat_interaction
```
:::

```{r cat-cat-gaps}
#| include: false
cat_gap <- cat_obs |>
  group_by(balance_cat) |>
  summarize(gap = eta[student == "Yes"] - eta[student == "No"],
            .groups = "drop")
```

The points are the four observed cell log-odds; each dashed segment connects
the two log-odds a model fits within one balance category. The two panels
carry separate vertical axes of the same width, since the two balance
categories are about four log-odds apart and a shared axis would flatten every
student difference in the figure; a segment's steepness still means the same
thing in both. The additive
model's segments fall at the same slope in both panels, since one coefficient
$\hat\beta_2 = `r sprintf("%.3f", coef(cat_add)[3])`$ has to serve as the
student effect at both balance levels, and neither segment passes through both
of its panel's points. The interaction model's segments pass exactly through
all four points — four coefficients for four cells — with slopes differing by
$\hat\beta_3$: the observed student gap is
$`r sprintf("%.3f", cat_gap$gap[cat_gap$balance_cat == "Low"])`$ log-odds below
\$1,500 and $`r sprintf("%.3f", cat_gap$gap[cat_gap$balance_cat == "High"])`$
above it. That difference in slope is what the $z$-statistic above finds
indistinguishable from sampling error.

The same fit looks different on the probability scale, where the four points
and segments are bounded to $[0, 1]$ instead of spread across four log-odds.

::: {.callout-note collapse="true"}
### Code
```{r cat-cat-prob-setup}
#| output: false
cat_fit_p <- cat_fit |> mutate(p_hat = plogis(eta))

cat_prob_base <- ggplot(mapping = aes(x = student, y = p_hat,
                                      color = student, shape = student)) +
  geom_point(data = cat_obs |> rename(p_hat = rate), size = 3) +
  facet_wrap(~ balance_cat, labeller = labeller(balance_cat = cat_labels)) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Student", y = "P(default = Yes)",
       color = "Student", shape = "Student", title = "ISLR2::Default")

cat_prob_data <- cat_prob_base

cat_prob_additive <- cat_prob_base +
  geom_line(data = filter(cat_fit_p, Model == "Additive"), aes(group = 1),
            color = "grey40", linetype = 2, linewidth = 0.8)

cat_prob_interaction <- cat_prob_base +
  geom_line(data = filter(cat_fit_p, Model == "Interaction"), aes(group = 1),
            color = "grey40", linetype = 2, linewidth = 0.8)
```
:::

::: {.panel-tabset}
#### Data
```{r cat-cat-prob-data-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
cat_prob_data
```

#### + Additive
```{r cat-cat-prob-additive-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
cat_prob_additive
```

#### + Interaction
```{r cat-cat-prob-interaction-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
cat_prob_interaction
```
:::

Both balance categories now share one axis, since probabilities need no
per-panel window to stay comparable. The roughly four-log-odds gap between
balance categories compresses into two clusters, one near $0$ and one near
$1$; the additive segments still share a slope and the interaction segments
still pass through all four points, but the difference between the two models
is harder to see here than on the log-odds scale above.

### Continuous-categorical interactions

The combination easiest to visualize pairs one quantitative feature $X$ with
one indicator $D$:

$$\eta = \beta_0 + \beta_1 X + \beta_2 D + \beta_3 XD.$$

Collecting the terms involving $X$ shows what each coefficient does:

$$\log\left(\frac{p(X,D)}{1 - p(X,D)}\right) = \underbrace{(\beta_0 + \beta_2 D)}_{\text{log-odds intercept}} + \underbrace{(\beta_1 + \beta_3 D)}_{\text{log-odds slope on } X} X.$$

Setting $X = 0$ and $D = 0$ leaves $\beta_0$, the log-odds of the event in the
reference group at $X = 0$, so $e^{\beta_0}$ are the odds and
$\text{logistic}(\beta_0)$ the probability there. Adding $\beta_2$ shifts that
reference value to the indicated group, still at $X = 0$.

The slope on $X$ is a function of $D$, which is the whole point of the
interaction. Writing out the two levels:

- **Reference level ($D = 0$):** log-odds $= \beta_0 + \beta_1 X$. A one-unit
increase in $X$ adds $\beta_1$ to the log-odds and multiplies the odds by
$e^{\beta_1}$.
- **Indicated level ($D = 1$):** log-odds
$= (\beta_0 + \beta_2) + (\beta_1 + \beta_3) X$. A one-unit increase in $X$
adds $\beta_1 + \beta_3$ to the log-odds and multiplies the odds by
$e^{\beta_1 + \beta_3}$.

So there is no single odds ratio for $X$. There are two, one per group, and
their ratio is

$$\frac{e^{\beta_1 + \beta_3}}{e^{\beta_1}} = e^{\beta_3}.$$

$e^{\beta_3}$ is a ratio of odds ratios: the factor by which one group's
per-unit odds ratio for $X$ differs from the other's. $\beta_3 = 0$ makes that
factor $1$ and returns the additive model.

One caution carries over from the feature engineering lecture: $\beta_1$ is the
log-odds slope **when $D = 0$**, not an average across the two groups. A second
is that the $X = 0$ at which $\beta_0$ and $\beta_2$ are defined may be a value
$X$ never takes. Hierarchy applies here as it did in linear regression: keep
$X$ and $D$ in the model whenever $XD$ is present.

```{r interaction-syntax}
#| eval: false
glm(default ~ balance * student, data = Default, family = "binomial")
```

The `*` expands to `balance + student + balance:student`, so the fitted model
contains both main effects and the product, satisfying hierarchy
automatically. Writing `balance:student` alone would drop the main effects,
violating hierarchy: R responds by giving each group its own balance slope
while forcing both to share a single intercept.

Fitting both models on `ISLR2::Default`, with `balance` as $X$ and the student
indicator as $D$, gives the pair the rest of this section compares:

```{r cont-cat-fits}
default_additive    <- glm(default ~ balance + student,
                           data = Default, family = "binomial")
default_interaction <- glm(default ~ balance * student,
                           data = Default, family = "binomial")
coef(default_interaction)
```

```{r cont-cat-extract}
#| include: false
bda <- coef(default_additive)
bdi <- coef(default_interaction)
```

The intercept $\hat\beta_0 = `r sprintf("%.2f", bdi[1])`$ is the log-odds of
default for a non-student carrying no balance at all, a value far outside
anything the data contain. From that reference, $\hat\beta_2$ moves a student
with no balance by $`r sprintf("%.3f", bdi[3])`$ log-odds. The balance slope is
$\hat\beta_1 = `r signif(bdi[2], 4)`$ for non-students and
$\hat\beta_1 + \hat\beta_3 = `r signif(bdi[2] + bdi[4], 4)`$ for students, so
the two per-\$100 odds ratios are
$`r sprintf("%.3f", exp(100 * bdi[2]))`$ and
$`r sprintf("%.3f", exp(100 * (bdi[2] + bdi[4])))`$.

Written as functions of $X$, the fitted log-odds for the two groups are

$$\hat\eta(X, 0) = \hat\beta_0 + \hat\beta_1 X, \qquad \hat\eta(X, 1) = (\hat\beta_0 + \hat\beta_2) + (\hat\beta_1 + \hat\beta_3) X.$$

Under the additive model $\beta_3 = 0$, both lines have slope
$\hat\beta_1$: they are **parallel**, separated by exactly $\hat\beta_2$ at
every $X$. Under the interaction model the slopes differ by $\hat\beta_3$, so
the lines are not parallel and, extended far enough, must cross. Parallel
versus non-parallel on the linear-predictor scale is the visual signature that
separates the two models.

In R, `predict()` on a `glm` object returns both scales:

```{r predict-scales}
#| eval: false
predict(fit, newdata = grid)                    # eta-hat, the linear predictor
predict(fit, newdata = grid, type = "response") # p-hat, the probability
```

Plotting $\hat\eta$ against balance for the two `Default` fits above draws
those lines.

::: {.callout-note collapse="true"}
### Code
```{r default-link-setup}
#| output: false
default_grid <- expand_grid(
  balance = seq(0, max(Default$balance), length.out = 300),
  student = factor(c("No", "Yes"), levels = levels(Default$student)))

default_link <- bind_rows(
  default_grid |> mutate(Model = "Additive",
                         eta = predict(default_additive, newdata = default_grid)),
  default_grid |> mutate(Model = "Interaction",
                         eta = predict(default_interaction, newdata = default_grid)))

# A shared base layer, so the two tabs below differ only in which model's
# lines are drawn on it -- the panel itself never changes size or shape.
default_link_base <- ggplot(mapping = aes(x = balance, y = eta,
                                          color = student, linetype = student)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  coord_cartesian(xlim = range(default_grid$balance),
                  ylim = range(default_link$eta)) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Credit card balance ($)", y = expression(hat(eta)),
       color = "Student", linetype = "Student", title = "ISLR2::Default")

default_link_additive <- default_link_base +
  geom_line(data = filter(default_link, Model == "Additive"), linewidth = 0.9)

default_link_interaction <- default_link_base +
  geom_line(data = filter(default_link, Model == "Interaction"), linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Additive
```{r default-link-additive-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_link_additive
```

#### Interaction
```{r default-link-interaction-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_link_interaction
```
:::

The additive tab's two lines are exactly parallel, both with slope
$`r signif(bda[2], 4)`$ log-odds per dollar and separated by
$\hat\beta_2 = `r sprintf("%.3f", bda[3])`$ at every balance. The interaction
tab's two lines have slopes $`r signif(bdi[2], 4)`$ and
$`r signif(bdi[2] + bdi[4], 4)`$, a difference too small to see: switching
between the tabs barely moves the picture.

The probability scale is what a decision actually uses, and it is a distorted
view of the same two lines. Passing each line through
$\text{logistic}(\cdot)$ compresses it toward $0$ and $1$, so two parallel
lines become two S-curves of identical shape that never cross, while two
non-parallel lines become two S-curves of different steepness that can cross
within the range of the data. Building the same comparison up from the raw
data puts both fits on that scale.

::: {.callout-note collapse="true"}
### Code
```{r default-prob-setup}
#| output: false
default_prob <- bind_rows(
  default_grid |>
    mutate(Model = "Additive",
           p_hat = predict(default_additive, newdata = default_grid,
                           type = "response")),
  default_grid |>
    mutate(Model = "Interaction",
           p_hat = predict(default_interaction, newdata = default_grid,
                           type = "response")))

# The same shared-base idea, with the raw data as the layer every tab keeps.
# The legend is kept on every tab (with more visible swatches than the
# points themselves) rather than suppressed on this one layer, so the panel
# is the same width in all three tabs -- an unequal panel width across tabs
# looks, to a reader clicking between them, like the data changed.
default_prob_base <- ggplot(mapping = aes(x = balance, y = p_hat,
                                          color = student, linetype = student)) +
  geom_point(data = Default,
             aes(y = default_jit, shape = student),
             alpha = 0.08, size = 0.7) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  coord_cartesian(xlim = range(default_grid$balance), ylim = c(-0.05, 1.05)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Student", linetype = "Student", shape = "Student",
       title = "ISLR2::Default")

default_prob_data <- default_prob_base

default_prob_additive <- default_prob_base +
  geom_line(data = filter(default_prob, Model == "Additive"), linewidth = 0.9)

default_prob_interaction <- default_prob_base +
  geom_line(data = filter(default_prob, Model == "Interaction"), linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Data
```{r default-prob-data-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_prob_data
```

#### + Additive
```{r default-prob-additive-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_prob_additive
```

#### + Interaction
```{r default-prob-interaction-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_prob_interaction
```
:::

The data tab plots each customer's outcome, jittered vertically, against
balance, with students and non-students distinguished by color and shape. The
two model tabs are nearly the same picture: two S-curves of all but identical
shape, the student curve shifted right, whether or not the interaction is in
the model.

### Continuous-continuous interactions

Two quantitative features multiply into a single new column, $X_1 X_2$:

$$\eta = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 X_2.$$

The interpretation follows the same pattern as before: fixing $X_2$, a
one-unit increase in $X_1$ changes the log-odds by $\beta_1 + \beta_3 X_2$ — a
slope that itself depends on $X_2$, rather than two group-specific slopes.

A coefficient's raw size says nothing about whether that dependence matters,
because it depends on the units the two features happen to be measured in.
Fitting `balance * income` on `ISLR2::Default`:

```{r cont-cont-fit}
balance_income_glm <- glm(default ~ balance * income, data = Default,
                          family = "binomial")
summary(balance_income_glm)$coefficients
```

```{r cont-cont-extract}
#| include: false
bi_b <- coef(balance_income_glm)
bi_z <- summary(balance_income_glm)$coefficients["balance:income", "z value"]
```

The estimated interaction, $\hat\beta_3 \approx `r signif(bi_b[4], 3)`$, looks
negligible beside $\hat\beta_1 \approx `r signif(bi_b[2], 3)`$ — but `income`
runs into the tens of thousands of dollars while `balance` runs into the
thousands, and rescaling `income` to thousands of dollars changes that
comparison completely without changing the fitted model at all:

```{r cont-cont-rescale}
Default_k <- Default |> mutate(income_k = income / 1000)
balance_income_k_glm <- glm(default ~ balance * income_k, data = Default_k,
                            family = "binomial")
rbind(dollars = coef(balance_income_glm),
      thousands = coef(balance_income_k_glm))
```

The interaction coefficient moved by a factor of $1{,}000$, the same factor
`income` was rescaled by, while the $z$-statistic
($`r sprintf("%.3f", bi_z)`$ either way) and every fitted probability are
unchanged to well beyond the precision this table shows. A coefficient's
*size* is an artifact of the units it was
estimated in; whether an interaction matters is a question for the $z$-test or
the drop-in-deviance test below, never for comparing $\hat\beta_3$ against
$\hat\beta_1$ by eye.

Neither feature has levels to split the picture by, so the fit is drawn by
splitting one of them: customers are grouped into income quartiles, and each
group's fitted curve is drawn at that group's median income against the full
range of balance. Both models are fit on `income` as recorded; the quartiles
only decide which four curves to draw.

::: {.callout-note collapse="true"}
### Code
```{r cont-cont-link-setup}
#| output: false
Default <- Default |>
  mutate(income_q = cut(income, breaks = quantile(income, 0:4 / 4),
                        include.lowest = TRUE,
                        labels = c("Q1 (lowest)", "Q2", "Q3", "Q4 (highest)")))

balance_income_add <- glm(default ~ balance + income, data = Default,
                          family = "binomial")

# One curve per income quartile, each evaluated at that quartile's median
# income -- the fit is a function of income itself, so a group needs a single
# income value to be drawn at.
income_median <- Default |>
  group_by(income_q) |>
  summarize(income = median(income), .groups = "drop")

income_grid <- expand_grid(
  balance  = seq(0, max(Default$balance), length.out = 300),
  income_q = income_median$income_q) |>
  left_join(income_median, by = "income_q")

income_link <- bind_rows(
  income_grid |> mutate(Model = "Additive",
                        eta = predict(balance_income_add, newdata = income_grid)),
  income_grid |> mutate(Model = "Interaction",
                        eta = predict(balance_income_glm, newdata = income_grid)))

# Four groups, so shape and linetype cannot simply duplicate a two-level color
# scale; viridis keeps the four distinguishable without color. No Data tab on
# this scale, the same way the continuous-categorical case has none -- raw
# 0/1 outcomes have no finite log-odds to plot.
income_link_base <- ggplot(mapping = aes(x = balance, y = eta,
                                         color = income_q, linetype = income_q)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  coord_cartesian(xlim = range(income_grid$balance),
                  ylim = range(income_link$eta)) +
  scale_color_viridis_d(end = 0.8) +
  labs(x = "Credit card balance ($)", y = expression(hat(eta)),
       color = "Income\nquartile", linetype = "Income\nquartile",
       title = "ISLR2::Default")

income_link_additive <- income_link_base +
  geom_line(data = filter(income_link, Model == "Additive"), linewidth = 0.9)

income_link_interaction <- income_link_base +
  geom_line(data = filter(income_link, Model == "Interaction"), linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Additive
```{r cont-cont-link-additive-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
income_link_additive
```

#### Interaction
```{r cont-cont-link-interaction-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
income_link_interaction
```
:::

Under the additive model the four lines are exactly parallel, each shifted
from the next by a fixed amount of log-odds. Under the interaction model the
slope on balance itself grows with income, so the four lines fan out instead
of staying parallel — the same parallel-versus-not signature the
continuous-categorical case showed, now with four lines instead of two.

The probability scale is what a decision actually uses, and it is a distorted
view of the same four lines: passing each one through $\text{logistic}(\cdot)$
compresses it toward $0$ and $1$.

::: {.callout-note collapse="true"}
### Code
```{r cont-cont-prob-setup}
#| output: false
income_prob <- bind_rows(
  income_grid |> mutate(Model = "Additive",
                        p_hat = predict(balance_income_add,
                                        newdata = income_grid,
                                        type = "response")),
  income_grid |> mutate(Model = "Interaction",
                        p_hat = predict(balance_income_glm,
                                        newdata = income_grid,
                                        type = "response")))

# Observed default rates in $400-wide balance bins within each income
# quartile, dropped where a bin holds fewer than 50 customers.
income_binned <- Default |>
  mutate(bin = cut(balance, breaks = seq(0, 2800, by = 400),
                   include.lowest = TRUE)) |>
  group_by(income_q, bin) |>
  summarize(balance = mean(balance), rate = mean(default01), n = n(),
            .groups = "drop") |>
  filter(n >= 50)

# Four groups, so shape and linetype cannot simply duplicate a two-level
# color scale; viridis keeps the four distinguishable without color.
income_base <- ggplot(mapping = aes(x = balance, y = p_hat,
                                    color = income_q, linetype = income_q)) +
  geom_point(data = income_binned,
             aes(y = rate, size = n, shape = income_q)) +
  coord_cartesian(xlim = range(income_grid$balance), ylim = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_color_viridis_d(end = 0.8) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_size_area(max_size = 4) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Income\nquartile", linetype = "Income\nquartile",
       shape = "Income\nquartile", size = "Customers\nin bin",
       title = "ISLR2::Default")

income_data <- income_base

income_additive <- income_base +
  geom_line(data = filter(income_prob, Model == "Additive"), linewidth = 0.9)

income_interaction <- income_base +
  geom_line(data = filter(income_prob, Model == "Interaction"), linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Data
```{r cont-cont-data-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
income_data
```

#### + Additive
```{r cont-cont-additive-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
income_additive
```

#### + Interaction
```{r cont-cont-interaction-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
income_interaction
```
:::

```{r cont-cont-spread}
#| include: false
income_at_2000 <- income_prob |> filter(abs(balance - 2000) ==
                                        min(abs(balance - 2000)))
income_spread <- income_at_2000 |>
  group_by(Model) |>
  summarize(spread = max(p_hat) - min(p_hat), .groups = "drop")
```

The additive tab draws four curves of identical shape, each shifted from the
next by a fixed amount of log-odds, so no pair of them ever crosses. The
interaction tab lets the balance slope grow with income, and its four curves
pull apart at high balances and close up at low ones: at a balance of \$2,000
the fitted probabilities span
`r sprintf("%.2f", income_spread$spread[income_spread$Model == "Additive"])`
under the additive model and
`r sprintf("%.2f", income_spread$spread[income_spread$Model == "Interaction"])`
under the interaction. Switching between the tabs changes the picture very
little, which is the same verdict the $z$-statistic above reaches.

### Fitted model geometry

Besides the shape of each group's fitted curve, shown above for `balance` and
`student`, a fitted interaction differs from a fitted additive model in one
more visible way: the feature value at which each group's predicted class
flips.

A fitted probability is often turned into a predicted class by comparing it to
a threshold, most commonly $0.5$. The **decision boundary** is the set of
feature values at which the model estimates the two classes as equally likely,

$$\hat p(x) = 0.5 \iff \frac{\hat p(x)}{1 - \hat p(x)} = 1 \iff \hat\eta(x) = 0,$$

where $\hat p(x)$ is the estimated probability of the event at feature values
$x$ and $\hat\eta(x) = x\hat\beta$ is the estimated linear predictor. Because
the logistic function is strictly increasing, $\hat p(x) > 0.5$ exactly when
$\hat\eta(x) > 0$, so the boundary is where the *linear* predictor crosses
zero — for one quantitative feature and one indicator, a single threshold
value of $X$ computed separately within each level of $D$:

$$x^{*}_{D=0} = -\frac{\hat\beta_0}{\hat\beta_1}, \qquad x^{*}_{D=1} = -\frac{\hat\beta_0 + \hat\beta_2}{\hat\beta_1 + \hat\beta_3}.$$

Dropping the interaction sets $\beta_3 = 0$, so the two thresholds share a
denominator, $x^{*}_{D=1} = x^{*}_{D=0} - \hat\beta_2/\hat\beta_1$: the
additive model still gives the two groups different thresholds, but what it
cannot give them is different *steepness*, since both groups cross the
boundary at the common rate $\hat\beta_1$ per unit of $X$.

### Interaction tests

A single interaction coefficient is tested the way any other coefficient is,
with the Wald $z$-statistic from the simple logistic regression lecture:
$H_0: \beta_3 = 0$ against $z = \hat\beta_3/\text{SE}(\hat\beta_3)$, printed by
`summary()` alongside every other coefficient.

A categorical feature with more than two levels contributes more than one
interaction coefficient — one per non-reference level — and no single
$z$-test addresses whether the feature interacts with $X$ *at all*. That
question compares two models with the drop-in-deviance test of the multiple
logistic regression lecture: a reduced model without any of the interaction
terms against a full model with all of them,
$G^2 = D_r - D_f$ approximately $\chi^2_k$ where

- $D_r$ is the deviance of the reduced model,
- $D_f$ is the deviance of the full model, and
- $k$ is the number of interaction terms dropped.

The drop-in-deviance test is the one that always applies: it asks the same
question for $k = 1$ as for $k = 10$, where the $z$-test only ever addresses
one coefficient at a time. Both tests apply to the `balance`-by-`student`
interaction, which contributes exactly one coefficient:

```{r default-interaction-z}
summary(default_interaction)$coefficients["balance:studentYes", ]
```

```{r default-interaction-deviance}
anova(default_additive, default_interaction, test = "Chisq")
```

```{r default-interaction-test-extract}
#| include: false
dz   <- summary(default_interaction)$coefficients["balance:studentYes", ]
ddev <- anova(default_additive, default_interaction, test = "Chisq")
dG2  <- ddev$Deviance[2]
dp   <- ddev[["Pr(>Chi)"]][2]
```

Both are testing $H_0: \beta_3 = 0$. The $z$-statistic is
$`r sprintf("%.3f", dz["z value"])`$ with $p = `r sprintf("%.3f", dz["Pr(>|z|)"])`$;
dropping the interaction raises the deviance by
$G^2 = `r sprintf("%.3f", dG2)`$ on $`r ddev$Df[2]`$ degree of freedom, giving
$p = `r sprintf("%.3f", dp)`$. With $k = 1$ the two are different statistics
for the same hypothesis, each with an approximate $\chi^2_1$ null
distribution: the Wald statistic reads the curvature of the log-likelihood at
$\hat\beta$, while the deviance compares the log-likelihood itself at the two
fits, which is why $z^2 = `r sprintf("%.3f", dz["z value"]^2)`$ and $G^2$ come
out close without being equal. Only the deviance test survives $k > 1$.

The price example later in this lecture is exactly that multi-level case: a
store's identity has more than two levels, so testing whether price interacts
with store at all takes a drop-in-deviance test, not a single $z$.

## Polynomials

A polynomial is a basis function like any other: $\beta_1 X + \beta_2 X^2$ adds
one more column, $X^2$, to the linear predictor, and `poly(x, 2)` builds it in
R exactly as it did for linear regression. It combines with an interaction the
same way a linear term did — put it on both sides of `*` — giving each group
its own quadratic shape rather than its own straight line.

```{r poly-fits}
poly_linear_additive <- glm(default ~ balance + student,
                            data = Default, family = "binomial")
poly_additive        <- glm(default ~ poly(balance, 2) + student,
                            data = Default, family = "binomial")
poly_interaction     <- glm(default ~ poly(balance, 2) * student,
                            data = Default, family = "binomial")
```

The first fit is the linear additive model again, kept as the reduced model
for the test below. Its curves, and the linear interaction model's, were
plotted in the Interactions section above, so the tabs here add only the
quadratic fits to the same raw data.

::: {.callout-note collapse="true"}
### Code
```{r poly-prob-setup}
#| output: false
poly_grid <- expand_grid(
  balance = seq(0, max(Default$balance), length.out = 300),
  student = factor(c("No", "Yes"), levels = levels(Default$student)))

poly_probs <- bind_rows(
  poly_grid |> mutate(Model = "Additive (quadratic)",
                      p_hat = predict(poly_additive, newdata = poly_grid,
                                      type = "response")),
  poly_grid |> mutate(Model = "Interaction (quadratic)",
                      p_hat = predict(poly_interaction, newdata = poly_grid,
                                      type = "response"))
) |> mutate(Model = fct_inorder(Model))

# The same shared-base idea used throughout this chapter: one base layer
# carrying the raw (jittered) data, with each tab adding only its own line.
poly_base <- ggplot(mapping = aes(x = balance, y = p_hat,
                                  color = student, linetype = student)) +
  geom_point(data = Default, aes(y = default_jit, shape = student),
             alpha = 0.08, size = 0.7) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  coord_cartesian(xlim = range(poly_grid$balance), ylim = c(-0.05, 1.05)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Student", linetype = "Student", shape = "Student",
       title = "ISLR2::Default")

poly_data     <- poly_base
poly_quad_add <- poly_base +
  geom_line(data = filter(poly_probs, Model == "Additive (quadratic)"), linewidth = 0.9)
poly_quad_int <- poly_base +
  geom_line(data = filter(poly_probs, Model == "Interaction (quadratic)"), linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Data
```{r poly-data-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
poly_data
```

#### + Additive (quadratic)
```{r poly-quad-add-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
poly_quad_add
```

#### + Interaction (quadratic)
```{r poly-quad-int-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
poly_quad_int
```
:::

```{r poly-quad-extract}
#| include: false
poly_quad_z <- summary(poly_additive)$coefficients["poly(balance, 2)2", "z value"]
```

The two quadratic tabs are all but identical to each other, and to the linear
fits of the Interactions section: the quadratic term itself is not significant
($z = `r sprintf("%.2f", poly_quad_z)`$), so these data give no evidence that
balance's effect on the log-odds curves at all. Adding curvature does not
manufacture an interaction where the data do not support one, any more than
adding an interaction would manufacture curvature that is not there.

The quadratic term can equally be tested by comparing the linear-only and
quadratic additive fits:

```{r poly-deviance}
anova(poly_linear_additive, poly_additive, test = "Chisq")
```

```{r poly-deviance-extract}
#| include: false
poly_dev <- anova(poly_linear_additive, poly_additive, test = "Chisq")
poly_G2  <- poly_dev$Deviance[2]
poly_p   <- poly_dev[["Pr(>Chi)"]][2]
```

Dropping the quadratic column costs $G^2 = `r sprintf("%.3f", poly_G2)`$ of
deviance on $`r poly_dev$Df[2]`$ degree of freedom
($p = `r sprintf("%.3f", poly_p)`$), matching
$z^2 = `r sprintf("%.3f", poly_quad_z^2)`$ to three decimals — one added
parameter, so the two tests carry the same information. The reason to reach
for this one anyway
is that it would still be a single test if the polynomial were cubic or
quartic, where the $z$-column would offer one statistic per added degree and
no verdict on the polynomial as a whole.

## Step functions

A step function is likewise just a set of indicator columns,
$C_j(X) = \mathrm{I}(c_j \le X < c_{j+1})$ for $j = 1, \ldots, p$, built from
cut points $-\infty = c_0 < c_1 < \cdots < c_p < c_{p+1} = \infty$ exactly as
in the flexibility lecture, with $[c_0, c_1)$ the reference bin and `cut(x,
breaks)` in place of `poly(x, degree)`. Coefficients are read the same way:
$\beta_j$ is the log-odds difference between bin $j$ and the reference bin, so
$e^{\beta_j}$ is the odds ratio between those two bins — where in linear
regression $\beta_j$ was a difference between bin means.

More bins mean a more flexible fit, at the same bias-variance cost flexibility
always carries: with enough bins the step function tracks the training data
closely, at the price of a noisier estimate in any bin with few observations.

::: {.callout-note collapse="true"}
### Code
```{r step-setup}
#| output: false
step_grid <- expand_grid(
  balance = seq(0, max(Default$balance), length.out = 400),
  student = factor(c("No", "Yes"), levels = levels(Default$student)))

# The same $400-wide-bin, by-student grouping used to check the additive fit
# in the previous lecture, so the observed data look the same way here as
# they did there. default01 was added to Default in the Interactions section
# above.
balance_bin_width <- 400
balance_binned <- Default |>
  mutate(bin = cut(balance, breaks = seq(0, 2800, by = balance_bin_width),
                   include.lowest = TRUE)) |>
  group_by(student, bin) |>
  summarize(balance = mean(balance), rate = mean(default01), n = n(),
            .groups = "drop") |>
  filter(n >= 100)

step_base <- ggplot(mapping = aes(x = balance)) +
  geom_point(data = balance_binned,
             aes(y = rate, color = student, shape = student, size = n)) +
  coord_cartesian(xlim = range(step_grid$balance), ylim = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  scale_size_area(max_size = 5) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Student", shape = "Student", size = "Customers\nin bin",
       title = "ISLR2::Default")

step_curve <- function(breaks) {
  fit <- glm(default ~ cut(balance, breaks = breaks, include.lowest = TRUE) *
               student,
             data = Default, family = "binomial")
  step_grid |> mutate(p_hat = predict(fit, newdata = step_grid,
                                      type = "response"))
}

step_line <- function(breaks) {
  list(geom_line(data = step_curve(breaks),
                 aes(y = p_hat, color = student, linetype = student),
                 linewidth = 0.9),
       labs(linetype = "Student"))
}

# Explicit, unequal-width breaks rather than cut(balance, breaks = k): every
# bin now has to hold both outcomes within *each* student group, not just
# overall, and equal-width bins leave the lowest bin with no student defaults
# at all once k reaches 4 (an empty cell is a preview of the separation the
# next lecture covers). Predicting from a re-fit cut() on new data is also
# fragile unless the same breaks are passed to both the fit and the grid.
step_data <- step_base
step_2bin <- step_base + step_line(c(0, 1350, 2700))
step_4bin <- step_base + step_line(c(0, 1100, 1500, 1900, 2700))
step_8bin <- step_base +
  step_line(c(0, 1100, 1300, 1500, 1700, 1900, 2100, 2300, 2700))
```
:::

Each fit below uses `cut(balance, breaks) * student`, giving the two groups
separate staircases over the same bins; the observed points are the default
rate in the same \$400-wide bins the previous lecture used, split by student.

::: {.panel-tabset}
#### Data
```{r step-data-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
step_data
```

#### + 2 bins
```{r step-2bin-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
step_2bin
```

#### + 4 bins
```{r step-4bin-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
step_4bin
```

#### + 8 bins
```{r step-8bin-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
step_8bin
```
:::

```{r step-extract}
#| include: false
step_top <- Default |>
  filter(balance > 2300) |>
  count(student)
```

Two bins give each group one flat stretch and one jump, barely distinguishing
low balances from high; each doubling adds detail the staircases can track,
until eight bins trace nearly the same rise the smooth logistic curve of the
earlier lectures did. Interacting the bins with `student` gives each group its
own height in every bin, so the two staircases run together at low balances,
where almost nobody defaults, and separate above roughly \$1,500, the student
staircase sitting below the non-student one over most of that range.

The price of that detail is visible in the eight-bin tab, where the staircases
stop rising monotonically, cross each other, and jump around: only
`r step_top$n[step_top$student == "No"]` non-students and
`r step_top$n[step_top$student == "Yes"]` students carry a balance above
\$2,300, so the top bin's two estimates are each built on a handful of
customers. Every added bin is another pair of indicator columns to estimate —
`cut(balance, breaks) * student` is a categorical-categorical interaction with
one parameter per bin-by-group cell — and the data have to be spread across
all of them.

## Student-specific balance effects

**Question.** The previous lecture fit `default ~ balance + student`
additively, which forces the student-versus-non-student odds ratio to be
identical at every balance and the balance odds ratio to be identical for
students and non-students. Is that additive assumption right, or does balance
act differently on the two groups?

**Data.** The same `r format(nrow(Default), big.mark = ",")` simulated credit
card customers as in the previous two lectures, with `balance` as the
quantitative feature $X$, the student indicator
$D = \mathrm{I}(\texttt{student} = \texttt{Yes})$, and $Y_i = 1$ when customer
$i$ defaults.

### Exploratory analysis

::: {.callout-note collapse="true"}
### Code
```{r default-eda}
#| output: false
default_eda_plot <- ggplot(Default, aes(x = balance, y = default_jit,
                                        color = student, shape = student)) +
  geom_point(alpha = 0.08, size = 0.7) +
  scale_y_continuous(breaks = c(0, 1), labels = c("No (0)", "Yes (1)")) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Credit card balance ($)", y = "Default",
       color = "Student", shape = "Student", title = "ISLR2::Default")
```
:::

```{r default-eda-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_eda_plot
```

Both groups show the same rough shape — defaults rare at low balances,
steadily more common as balance grows — with the student points sitting a
little to the right, a first hint of the confounding the previous lecture
already worked through. Whether the *rate of increase* also differs by group
is what an interaction asks and the raw scatter alone cannot answer.

### Fitting both models

```{r default-fits}
default_additive    <- glm(default ~ balance + student,
                           data = Default, family = "binomial")
default_interaction <- glm(default ~ balance * student,
                           data = Default, family = "binomial")
coef(default_interaction)
```

::: {.callout-note collapse="true"}
### Code
```{r default-coef-setup}
#| output: false
bda <- coef(default_additive)
bdi <- coef(default_interaction)

default_coef_table <- tibble(
  term        = names(bdi),
  additive    = c(bda, NA),
  interaction = bdi
) |>
  mutate(across(c(additive, interaction),
                \(x) ifelse(is.na(x), "", sprintf("%.6f", x)))) |>
  knitr::kable(col.names = c("Coefficient", "Additive fit",
                             "Interaction fit"),
               align = "lcc")
```
:::

```{r default-coef-table}
#| echo: false
default_coef_table
```

The two fits agree closely on the intercept and on the `balance` slope.
Whether the interaction $\hat\beta_3 = `r sprintf("%.6f", bdi[4])`$ is
negligible is a question for the $z$-test below, not for its size relative to
the `balance` main effect: a ratio of two point estimates says nothing about
how precisely $\hat\beta_3$ was estimated.

These are the two fits drawn in the Interactions section above. On the
linear-predictor scale, the additive model's lines are exactly parallel, both
with slope $`r signif(bda[2], 4)`$ log-odds per dollar and separated by
$\hat\beta_2 = `r sprintf("%.3f", bda[3])`$ at every balance, while the
interaction model's lines have slopes $`r signif(bdi[2], 4)`$ and
$`r signif(bdi[2] + bdi[4], 4)`$ — a difference too small to see. On the
probability scale the two are equally hard to tell apart: two S-curves of all
but identical shape, the student curve shifted right, whether or not the
interaction is in the model.

### Interaction test

```{r default-interaction-summary}
summary(default_interaction)$coefficients
```

::: {.callout-note collapse="true"}
### Code
```{r default-inference}
#| output: false
dcoef   <- summary(default_interaction)$coefficients
z_d     <- dcoef["balance:studentYes", "z value"]
p_d     <- dcoef["balance:studentYes", "Pr(>|z|)"]
ci_d    <- confint.default(default_interaction)["balance:studentYes", ]

or_d <- tibble(
  student = c("No (reference)", "Yes"),
  slope   = c(bdi[2], bdi[2] + bdi[4]),
  or100   = exp(100 * c(bdi[2], bdi[2] + bdi[4]))
)

or_d_table <- or_d |>
  mutate(slope = sprintf("%.6f", slope),
         or100 = sprintf("%.3f", or100)) |>
  knitr::kable(col.names = c("Student", "Slope (log-odds per \\$1)",
                             "Odds ratio per \\$100"),
               align = "lcc")

```
:::

```{r default-or-table}
#| echo: false
or_d_table
```

The two per-\$100 odds ratios differ by
`r sprintf("%.2f", abs(diff(or_d$or100)))`, which is the difference the
interaction was added to capture. The coefficient table above already gives
the answer: $z = `r sprintf("%.2f", z_d)`$ with a $p$-value of
$`r sprintf("%.3f", p_d)`$, and the 95% confidence interval for $\beta_3$
runs from $`r sprintf("%.5f", ci_d[1])`$ to $`r sprintf("%.5f", ci_d[2])`$,
comfortably containing $0$.

::: {.callout-note collapse="true"}
### Beyond this course
The **Akaike information criterion**, $\text{AIC} = -2\ell(\hat\beta) + 2k$ for
a model with $k$ estimated parameters, scores a fit by its log-likelihood with
a penalty for the number of parameters it used; lower is better. It gives a
second, test-free reading of whether a term earns its place.

```{r default-aic-fit}
#| output: false
aic_d_table <- tibble(
  Model = c("`default ~ balance + student`", "`default ~ balance * student`"),
  k     = c(length(bda), length(bdi)),
  AIC   = sprintf("%.2f", c(AIC(default_additive), AIC(default_interaction)))
) |>
  knitr::kable(col.names = c("Model", "Parameters", "AIC"), align = "lcc")
```

```{r default-aic-table}
#| echo: false
aic_d_table
```

Adding the interaction makes AIC *worse* here: the extra parameter buys less
log-likelihood than it costs.
:::

**Conclusion.** Balance's effect on the odds of default does not depend on
student status in these data. The $z$-test gives no reason to prefer the more
complicated model, and the two fitted models are visually indistinguishable on
both scales, so the additive model of the previous lecture is the one to
report. The decision boundaries move correspondingly little: the balance at
which $\hat p = 0.5$ shifts from
\$`r format(round(-bda[1] / bda[2]), big.mark = ",")` to
\$`r format(round(-bdi[1] / bdi[2]), big.mark = ",")` for non-students and from
\$`r format(round(-(bda[1] + bda[3]) / bda[2]), big.mark = ",")` to
\$`r format(round(-(bdi[1] + bdi[3]) / (bdi[2] + bdi[4])), big.mark = ",")` for
students.

Not every plausible interaction is a real one. Fitting it was still the right
move: the additive assumption was checked rather than assumed.

## Store-specific price effects

**Question.** Citrus Hill and Minute Maid are competing orange juice brands,
sold at several different stores. If a store raises the price of Citrus Hill,
fewer customers should buy it — but does price act the same way in every
store, or does the size of its effect vary from one store to the next?

**Data.** `ISLR2::OJ` records `r format(nrow(OJ), big.mark = ",")` individual
orange juice purchases. These are observed purchases, not simulated ones.
Three of its `r ncol(OJ)` variables are used here:

- `Purchase`: which brand the customer bought, a factor with levels
`r paste(levels(OJ$Purchase), collapse = " and ")` (Citrus Hill and Minute
Maid). This is the response.
- `SalePriceCH`: the price actually paid for Citrus Hill, in dollars, ranging
from \$`r sprintf("%.2f", min(OJ$SalePriceCH))` to
\$`r sprintf("%.2f", max(OJ$SalePriceCH))`. This is the quantitative feature
$X$.
- `StoreID`: which of `r length(unique(OJ$StoreID))` stores the purchase
happened at, with `r paste(sort(unique(OJ$StoreID)), collapse = ", ")` as the
store identifiers. This is a categorical feature with more than two levels —
exactly the case the drop-in-deviance test above was built for.

Buying Citrus Hill is the event of interest, so $Y_i = 1$ when purchase $i$ is
a Citrus Hill purchase. `glm()` models the probability of a factor's *second*
level, and `r levels(OJ$Purchase)[2]` is second, so the event has to be set
explicitly with `relevel()`:

```{r oj-relevel}
OJ <- OJ |> mutate(Purchase = relevel(Purchase, ref = "MM"))
levels(OJ$Purchase)
```

With `MM` as the reference level, every model below estimates
$P(\texttt{Purchase} = \texttt{CH})$.

### Exploratory analysis

Before fitting anything, plot each purchase against the price paid for
Citrus Hill, split by store:

::: {.callout-note collapse="true"}
### Code
```{r oj-eda}
#| output: false
OJ <- OJ |> mutate(ch01 = as.numeric(Purchase == "CH"))

oj_rates <- OJ |>
  group_by(StoreID, SalePriceCH) |>
  summarize(n = n(), rate = mean(ch01), .groups = "drop")

oj_eda_plot <- ggplot(OJ, aes(x = SalePriceCH, y = ch01)) +
  geom_jitter(height = 0.05, width = 0.008, alpha = 0.15, size = 0.8) +
  geom_point(data = oj_rates, aes(y = rate, size = n),
             shape = 1, color = "#D55E00") +
  facet_wrap(~ StoreID, labeller = label_both) +
  scale_y_continuous(breaks = c(0, 0.5, 1),
                     labels = c("MM (0)", "0.5", "CH (1)")) +
  scale_size_area(max_size = 5) +
  labs(x = "Citrus Hill sale price ($)", y = "Purchase",
       size = "Purchases\nat that price", title = "ISLR2::OJ")

store_cors <- OJ |>
  group_by(StoreID) |>
  summarize(cor = cor(SalePriceCH, ch01), .groups = "drop")
```
:::

```{r oj-eda-figure}
#| echo: false
#| fig-width: 7.5
#| fig-height: 4.5
oj_eda_plot
```

The open circles are the observed proportion of Citrus Hill purchases at each
distinct price, sized by how many purchases were made at that price. Even
before any model is fit, the raw correlation between price and purchase
already changes sign from store to store: it runs from
`r sprintf("%.2f", min(store_cors$cor))` at store
`r store_cors$StoreID[which.min(store_cors$cor)]` (higher prices, fewer Citrus
Hill purchases, as expected) to
`r sprintf("%.2f", max(store_cors$cor))` at store
`r store_cors$StoreID[which.max(store_cors$cor)]` (the opposite).

### Fitting both models

```{r oj-fits}
oj_additive    <- glm(Purchase ~ SalePriceCH + factor(StoreID),
                      data = OJ, family = "binomial")
oj_interaction <- glm(Purchase ~ SalePriceCH * factor(StoreID),
                      data = OJ, family = "binomial")
coef(oj_interaction)
```

::: {.callout-note collapse="true"}
### Code
```{r oj-coef-setup}
#| output: false
boa <- coef(oj_additive)
boi <- coef(oj_interaction)

oj_coef_table <- tibble(
  term        = names(boi),
  additive    = c(boa, rep(NA, length(boi) - length(boa))),
  interaction = boi
) |>
  mutate(across(c(additive, interaction),
                \(x) ifelse(is.na(x), "", sprintf("%.3f", x)))) |>
  knitr::kable(col.names = c("Coefficient", "Additive fit",
                             "Interaction fit"),
               align = "lcc")
```
:::

```{r oj-coef-table}
#| echo: false
oj_coef_table
```

The intercept is the log-odds of a Citrus Hill purchase at store
`r sort(unique(OJ$StoreID))[1]` (the reference store, chosen because it is
first in numeric order) at a Citrus Hill price of \$0, an extrapolation well
below the \$`r sprintf("%.2f", min(OJ$SalePriceCH))` minimum in the data. The
additive fit puts a single `SalePriceCH` coefficient of
$`r sprintf("%.3f", boa[2])`$ on every store. The interaction fit instead gives
each store its own `SalePriceCH` coefficient, `r length(unique(OJ$StoreID))`
in all.

### Fitted log-odds

Plotting $\hat\eta$ against price for each store shows what those separate
coefficients mean for the fitted lines:

::: {.callout-note collapse="true"}
### Code
```{r oj-link-setup}
#| output: false
# Each store's own observed price range, not the pooled range -- stores 3 and
# 4 were never sampled below $1.79, so a pooled grid would draw their fitted
# lines across prices those stores never charged.
oj_grid <- OJ |>
  group_by(StoreID) |>
  reframe(SalePriceCH = seq(min(SalePriceCH), max(SalePriceCH), length.out = 100))

oj_link <- bind_rows(
  oj_grid |> mutate(Model = "Additive",
                    eta = predict(oj_additive, newdata = oj_grid)),
  oj_grid |> mutate(Model = "Interaction",
                    eta = predict(oj_interaction, newdata = oj_grid)))

store_colors <- c("#E69F00", "#56B4E9", "#009E73", "#0072B2", "#D55E00")

# A shared base layer, so the two tabs below differ only in which model's
# lines are drawn on it.
oj_link_base <- ggplot(mapping = aes(x = SalePriceCH, y = eta,
                                     color = factor(StoreID),
                                     linetype = factor(StoreID))) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  coord_cartesian(xlim = range(oj_grid$SalePriceCH),
                  ylim = range(oj_link$eta)) +
  scale_color_manual(values = store_colors) +
  labs(x = "Citrus Hill sale price ($)", y = expression(hat(eta)),
       color = "Store", linetype = "Store", title = "ISLR2::OJ")

oj_link_additive <- oj_link_base +
  geom_line(data = filter(oj_link, Model == "Additive"), linewidth = 0.9)

oj_link_interaction <- oj_link_base +
  geom_line(data = filter(oj_link, Model == "Interaction"), linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Additive
```{r oj-link-additive-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
oj_link_additive
```

#### Interaction
```{r oj-link-interaction-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
oj_link_interaction
```
:::

The additive tab shows `r length(unique(OJ$StoreID))` parallel lines, all
sloping down at $`r sprintf("%.3f", boa[2])`$ log-odds per dollar. The
interaction tab lets
every store's line take its own slope, and some of them now slope *up*.
Switching between these two tabs changes the picture substantially, which is
what the `Default` example did not do.

### Fitted probabilities

The same two model fits, now on the scale a purchase decision actually uses,
with the observed proportions overlaid:

::: {.callout-note collapse="true"}
### Code
```{r oj-prob-setup}
#| output: false
oj_prob <- bind_rows(
  oj_grid |> mutate(Model = "Additive",
                    p_hat = predict(oj_additive, newdata = oj_grid,
                                    type = "response")),
  oj_grid |> mutate(Model = "Interaction",
                    p_hat = predict(oj_interaction, newdata = oj_grid,
                                    type = "response")))

# The same shared-base idea, with the observed proportions as the layer every
# tab keeps.
oj_prob_base <- ggplot(mapping = aes(x = SalePriceCH, y = p_hat,
                                     color = factor(StoreID),
                                     linetype = factor(StoreID))) +
  geom_hline(yintercept = 0.5, color = "grey60", linewidth = 0.3) +
  geom_point(data = oj_rates, aes(y = rate, size = n, shape = factor(StoreID)),
             alpha = 0.8) +
  coord_cartesian(xlim = range(oj_grid$SalePriceCH), ylim = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_size_area(max_size = 5) +
  scale_color_manual(values = store_colors) +
  labs(x = "Citrus Hill sale price ($)", y = "P(Purchase = CH)",
       color = "Store", linetype = "Store", shape = "Store",
       size = "Purchases\nat that price", title = "ISLR2::OJ")

oj_prob_data <- oj_prob_base

oj_prob_additive <- oj_prob_base +
  geom_line(data = filter(oj_prob, Model == "Additive"), linewidth = 0.9)

oj_prob_interaction <- oj_prob_base +
  geom_line(data = filter(oj_prob, Model == "Interaction"), linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Data
```{r oj-prob-data-plot}
#| echo: false
#| fig-width: 7.5
#| fig-height: 3.2
oj_prob_data
```

#### + Additive
```{r oj-prob-additive-plot}
#| echo: false
#| fig-width: 7.5
#| fig-height: 3.2
oj_prob_additive
```

#### + Interaction
```{r oj-prob-interaction-plot}
#| echo: false
#| fig-width: 7.5
#| fig-height: 3.2
oj_prob_interaction
```
:::

The data tab shows the observed proportion of Citrus Hill purchases at each
price, by store. The additive tab fits them with `r length(unique(OJ$StoreID))`
curves of identical shape that never cross, all drifting down with price. The
interaction tab lets
each store's curve rise or fall at its own rate, and several now cross within
the observed price range.

### Interpreting price slopes

Reading each store's slope off the interaction fit, and converting it to an
odds ratio, makes the split concrete:

::: {.callout-note collapse="true"}
### Code
```{r oj-slope-setup}
#| output: false
store_ids <- sort(unique(OJ$StoreID))
store_slope <- function(s) {
  if (s == store_ids[1]) return(unname(boi["SalePriceCH"]))
  term <- paste0("SalePriceCH:factor(StoreID)", s)
  unname(boi["SalePriceCH"] + boi[term])
}
store_slopes <- sapply(store_ids, store_slope)

oj_slope_table <- tibble(
  store = store_ids,
  slope = store_slopes,
  or10  = exp(0.10 * store_slopes)
) |>
  mutate(slope = sprintf("%.3f", slope),
         or10  = sprintf("%.3f", or10)) |>
  knitr::kable(col.names = c("Store", "Slope (log-odds per \\$1)",
                             "Odds ratio per \\$0.10"),
               align = "ccc")

n_negative <- sum(store_slopes < 0)
n_positive <- sum(store_slopes > 0)
```
:::

```{r oj-slope-table}
#| echo: false
oj_slope_table
```

The fitted slopes run downward, the direction a price increase is expected to
push demand, at `r n_negative` of the `r length(store_ids)` stores; the other
`r n_positive` run backwards, the same \$0.10 *raising* the estimated odds.
None of these five point estimates is individually far enough from zero to
rule out sampling error on its own — the drop-in-deviance test below is what
supports "not all five slopes are the same," not any one store's estimate by
itself. What these data cannot say is why the slopes that do run backwards do
so — `SalePriceCH` records what was paid, not what else varied
alongside it across stores and weeks, so the association it carries need not
be the effect of price by itself.

### Break-even prices

The decision boundary here is the Citrus Hill price at which the model puts
the two brands at 50/50, computed separately for each store.

::: {.callout-note collapse="true"}
### Code
```{r oj-boundaries}
#| output: false
store_intercept <- function(s) {
  if (s == store_ids[1]) return(unname(boi["(Intercept)"]))
  term <- paste0("factor(StoreID)", s)
  unname(boi["(Intercept)"] + boi[term])
}
store_intercepts <- sapply(store_ids, store_intercept)
store_boundaries <- -store_intercepts / store_slopes

# Each store's own observed price range, not the pooled range -- store 3's
# break-even price, for instance, is inside the range OJ$SalePriceCH covers
# overall but above the highest price ever charged at store 3 specifically.
store_ranges <- OJ |>
  group_by(StoreID) |>
  summarize(lo = min(SalePriceCH), hi = max(SalePriceCH), .groups = "drop")

oj_boundary_table <- tibble(
  store = store_ids,
  price = store_boundaries
) |>
  left_join(store_ranges, by = c("store" = "StoreID")) |>
  mutate(observed = ifelse(price >= lo & price <= hi, "yes", "no"),
         price = sprintf("\\$%.2f", price)) |>
  select(store, price, observed) |>
  knitr::kable(col.names = c("Store", "Price where $\\hat p = 0.5$",
                             "In observed range?"),
               align = "ccc")

n_outside <- sum(store_boundaries < store_ranges$lo | store_boundaries > store_ranges$hi)
```
:::

```{r oj-boundary-table}
#| echo: false
oj_boundary_table
```

`r n_outside` of the `r length(store_ids)` break-even prices fall outside the
range of prices that store itself charged — a store whose own data never
approached 50/50 can only place its break-even point by extrapolating past
what was observed there, which is a property of the fitted line, not evidence
about prices that store never charged.

### Interaction test

```{r oj-interaction-summary}
oj_dev <- anova(oj_additive, oj_interaction, test = "Chisq")
oj_dev
```

```{r oj-inference}
#| include: false
oj_G2 <- oj_dev$Deviance[2]
oj_df <- oj_dev$Df[2]
oj_p  <- oj_dev[["Pr(>Chi)"]][2]
```

With `r length(store_ids)` stores, the interaction contributes
`r length(store_ids) - 1` coefficients, one per non-reference store, so no
single $z$-test asks whether price acts differently across stores *in
general* — this is the multi-coefficient case the "Interaction tests" section
set up. Dropping all `r oj_df` price-by-store interaction terms at
once would raise the deviance by $G^2 = `r sprintf("%.2f", oj_G2)`$ on
$`r oj_df`$ degrees of freedom, giving $p = `r signif(oj_p, 3)`$: price's
effect on the odds of a Citrus Hill purchase is not the same across stores.

**Conclusion.** The association between Citrus Hill's price and the chance a
customer buys it differs sharply from store to store, and the drop-in-deviance
test rejects a common price slope decisively ($p = `r signif(oj_p, 3)`$). Two
of the five stores show the expected pattern of higher prices costing Citrus
Hill sales; three run the opposite way. The fitted lines diverge visibly on
both scales, the interaction model is the one to report, and what remains
unresolved is *why* three stores run backwards — these data describe the
pattern, not its cause.

::: {.callout-note collapse="true"}
### Beyond this course
Giving each store its own fixed intercept and slope, as the interaction model
above does, spends one parameter per store and says nothing about a store that
was not in the sample. A **random effects** (or **mixed-effects**) logistic
regression instead treats the stores actually observed as a sample from a
population of stores, and estimates the *distribution* of intercepts and
slopes across that population rather than a separate fixed value for each one.

A **random intercept** model lets each store's baseline log-odds vary around a
shared mean, $b_{0,\text{store}} \sim N(0, \sigma_0^2)$, while every store
keeps the same price slope — the mixed-effects analogue of the additive model
above. A **random slope** model goes further and lets the price slope itself
vary by store, $b_{1,\text{store}} \sim N(0, \sigma_1^2)$, typically allowed to
correlate with $b_{0,\text{store}}$ — the analogue of the interaction model,
but estimated with two variances and a correlation instead of one fixed
coefficient per store, which matters far more once there are dozens of stores
instead of five. In R, `lme4::glmer()` fits both:

```{r glmer-syntax}
#| eval: false
glmer(Purchase ~ SalePriceCH + (1 | StoreID),
      family = "binomial")                          # random intercept
glmer(Purchase ~ SalePriceCH + (SalePriceCH | StoreID),
      family = "binomial")                          # random intercept and slope
```
:::

## Two verdicts, two tests

Both examples asked whether one feature's effect on the odds of the response
depends on the value of another feature. `Default` answered no: the fitted
slopes for students and non-students differed by almost nothing, the curves
were indistinguishable, and both the $z$-test and the drop-in-deviance test
agreed ($p = `r sprintf("%.3f", p_d)`$). `OJ` answered yes, and needed a
different tool to ask the question at all — `r length(store_ids)` stores means
testing the interaction is a drop-in-deviance test rather than a single $z$,
and that test rejected the common-slope assumption decisively
($p = `r signif(oj_p, 3)`$).

Nothing about either pair of features announced which answer was coming, or
which test the answer would take. The additive assumption is checkable, and
checking it is the only way to find out — which is a reason to fit the
interaction even when you expect the additive model to survive, and to know
which test a categorical feature's number of levels calls for before running
it.

## Conclusion

Interactions, polynomials, and step functions are all choices of basis
function entering the linear predictor as their own column, estimated by the
same Bernoulli maximum likelihood as every other logistic regression, with
every coefficient still read as a change in log-odds. An interaction lets one
feature's effect depend on another's: categorical-categorical, adding an
extra log-odds for the combination of two non-reference levels;
continuous-categorical, giving each group its own log-odds slope and so its
own odds ratio; continuous-continuous, letting one feature's slope vary
continuously with the other's value. A coefficient's raw magnitude never
settles whether an interaction matters, since it depends on arbitrary units —
only a $z$-test, for one coefficient, or a drop-in-deviance test, for several
at once, does.

The diagnostic for the continuous-categorical case is a picture on the
linear-predictor scale, where both fits are straight lines: if the interaction
fit's two lines are all but parallel, the additive model already captures each
group's log-odds slope; if they visibly cross or diverge, the slope changes
from one group to the next. Each group's decision boundary is where its own
fitted line crosses zero, so an interaction moves the two boundaries
independently rather than by a common shift.

Polynomials and step functions raise no new estimation or interpretation
issues once basis functions are in view — only the question of how much
flexibility the data can support, the same bias-variance tradeoff the
flexibility lecture raised for a quantitative response. Adding columns to a
binary-response model can also break the estimation itself in a way that least
squares never breaks, a failure that goes by the name separation.
