# Flexible Logistic Regression

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **write** a logistic regression model with an interaction between a
quantitative feature and an indicator, and **derive** the two group-specific
log-odds slopes $\beta_1$ and $\beta_1 + \beta_3$,
- **interpret** $e^{\beta_1}$ and $e^{\beta_1 + \beta_3}$ as group-specific
odds ratios, and **explain** why no single odds ratio describes the
quantitative feature once an interaction is present,
- **compare** an additive fit against an interaction fit on the
linear-predictor scale, where the additive model forces two *parallel* lines,
and **translate** that comparison onto the probability scale,
- **define** the decision boundary as the feature value at which
$\hat p(x) = 0.5$ and **compute** it separately within each level of a
categorical feature,
- **test** $H_0: \beta_3 = 0$ with a $z$-statistic and **evaluate** whether an
interaction is supported, **recognizing** that a plausible interaction often
turns out not to be, and
- **explain** why polynomial and step-function features transfer from linear
regression to logistic regression unchanged.
:::

## What the additive model assumes

The previous lecture's model was **additive**: each feature contributed its own
term $\beta_j X_{ij}$ to the linear predictor, and no term involved two
features at once. That structure is what licensed the interpretation "a
one-unit increase in $X_j$ multiplies the odds by $e^{\beta_j}$, holding every
other feature fixed" — the same $e^{\beta_j}$ at every value of every other
feature.

That is an assumption, not a convenience. The fitted curves of the previous
lecture showed it directly: two curves of identical shape, separated by a
constant $\hat\beta_2$ on the log-odds scale at every balance.

There is no reason a real process has to work that way. Whether it does is a
question about the data, and it can be asked and answered with a model that
does not impose it.

## The interaction model

The fix is the one from the feature engineering lecture, applied to the linear
predictor instead of to the mean response. With a quantitative feature $X$ and
an indicator $D$, that lecture wrote

$$Y_i = \beta_0 + \beta_1 X_i + \beta_2 D_i + \beta_3 X_i D_i + \epsilon_i,$$

giving group A (where $D = 0$) intercept $\beta_0$ and slope $\beta_1$, and
group B (where $D = 1$) intercept $\beta_0 + \beta_2$ and slope
$\beta_1 + \beta_3$. Logistic regression uses that same linear predictor and
passes it through the logistic function,

$$p(X, D) = P(Y = 1 | X, D) = \text{logistic}(\eta) = \frac{e^{\eta}}{1 + e^{\eta}}, \qquad \eta = \beta_0 + \beta_1 X + \beta_2 D + \beta_3 X D,$$

where

- $X$ is the quantitative feature,
- $D$ is the indicator for the non-reference level of the categorical feature,
- $\eta$ is the linear predictor, the log-odds of the event, and
- $\beta_3$ is the coefficient on the product $XD$, the **interaction**.

Equivalently, on the log-odds scale,

$$\log\left(\frac{p(X,D)}{1 - p(X,D)}\right) = \beta_0 + \beta_1 X + \beta_2 D + \beta_3 X D.$$

The product $XD$ is just another column of the model matrix — written
$\mathbf{X}$ here, to keep it distinct from the scalar feature $X$ — built from
two columns already in it. Estimation is unchanged: $\hat\beta$ still maximizes
the Bernoulli log-likelihood

$$\ell(\beta) = \sum_{i=1}^n \left\{ y_i \, x_i\beta - \log\left(1 + e^{x_i\beta}\right)\right\},$$

with $x_i$ the $i$th row of $\mathbf{X}$, so it still has no closed-form
maximizer and is still found numerically by `glm()`.

Hierarchy applies here as it did in linear regression: keep $X$ and $D$ in the
model whenever $XD$ is present.

## Interpreting the coefficients

Collect the terms involving $X$:

$$\log\left(\frac{p(X,D)}{1 - p(X,D)}\right) = \underbrace{(\beta_0 + \beta_2 D)}_{\text{log-odds intercept}} + \underbrace{(\beta_1 + \beta_3 D)}_{\text{log-odds slope on } X} X.$$

The slope on $X$ is a function of $D$, which is the whole point of the
interaction. Writing out the two levels:

- **Reference level ($D = 0$):** log-odds $= \beta_0 + \beta_1 X$. A one-unit
increase in $X$ adds $\beta_1$ to the log-odds and multiplies the odds by
$e^{\beta_1}$.
- **Indicated level ($D = 1$):** log-odds
$= (\beta_0 + \beta_2) + (\beta_1 + \beta_3) X$. A one-unit increase in $X$
adds $\beta_1 + \beta_3$ to the log-odds and multiplies the odds by
$e^{\beta_1 + \beta_3}$.

So there is no single odds ratio for $X$. There are two, one per group, and
their ratio is

$$\frac{e^{\beta_1 + \beta_3}}{e^{\beta_1}} = e^{\beta_3}.$$

$e^{\beta_3}$ is a ratio of odds ratios: the factor by which one group's
per-unit odds ratio for $X$ differs from the other's. $\beta_3 = 0$ makes that
factor $1$ and returns the additive model.

One caution carries over from the feature engineering lecture, and has a
partner here: $\beta_1$ is the log-odds slope **when $D = 0$**, not an average
across groups, and $\beta_2$ is the difference in log-odds between the groups
**at $X = 0$**, which may be a value $X$ never takes.

## Two scales, two pictures

The two models are easiest to tell apart on the scale where both are linear.
Written as functions of $X$, the fitted log-odds are two straight lines:

$$\hat\eta(X, 0) = \hat\beta_0 + \hat\beta_1 X, \qquad \hat\eta(X, 1) = (\hat\beta_0 + \hat\beta_2) + (\hat\beta_1 + \hat\beta_3) X.$$

Under the additive model $\hat\beta_3 = 0$, both lines have slope
$\hat\beta_1$: they are **parallel**, separated by exactly $\hat\beta_2$ at
every $X$. Under the interaction model the slopes differ by $\hat\beta_3$, so
the lines are not parallel and, extended far enough, must cross. Parallel
versus non-parallel on the linear-predictor scale is the visual signature that
separates the two models.

The probability scale is what a decision actually uses, and it is a distorted
view of the same two lines. Passing each line through
$\text{logistic}(\cdot)$ compresses it toward $0$ and $1$, so two parallel
lines become two S-curves of identical shape that never cross, while two
non-parallel lines become two S-curves of different steepness that can cross
within the range of the data.

In R, `predict()` on a `glm` object returns these two scales:

```{r predict-scales}
#| eval: false
predict(fit, newdata = grid)                    # eta-hat, the linear predictor
predict(fit, newdata = grid, type = "response") # p-hat, the probability
```

The examples below plot both for each model: the linear predictor first, to
diagnose whether the fitted slopes differ, and the probability scale after, to
see what that difference is worth.

## Decision boundaries

A fitted probability is often turned into a predicted class by comparing it to
a threshold, most commonly $0.5$. The **decision boundary** is the set of
feature values at which the model estimates the two classes as equally likely,

$$\hat p(x) = 0.5 \iff \frac{\hat p(x)}{1 - \hat p(x)} = 1 \iff \hat\eta(x) = 0,$$

where

- $\hat p(x)$ is the estimated probability of the event at feature values $x$,
and
- $\hat\eta(x) = x\hat\beta$ is the estimated linear predictor.

Because the logistic function is strictly increasing, $\hat p(x) > 0.5$ exactly
when $\hat\eta(x) > 0$. The boundary is therefore where the *linear* predictor
crosses zero, which is why it is a simple set — here, with one quantitative
feature, a single threshold value of that feature computed separately within
each level of the categorical feature.

Solving $\hat\eta = 0$ in the interaction model gives

$$x^{*}_{D=0} = -\frac{\hat\beta_0}{\hat\beta_1}, \qquad x^{*}_{D=1} = -\frac{\hat\beta_0 + \hat\beta_2}{\hat\beta_1 + \hat\beta_3},$$

where $x^{*}_{D=d}$ is the value of $X$ at which $\hat p = 0.5$ for group $d$.
Dropping the interaction sets $\hat\beta_3 = 0$, so the two thresholds share a
denominator:

$$x^{*}_{D=0} = -\frac{\hat\beta_0}{\hat\beta_1}, \qquad x^{*}_{D=1} = -\frac{\hat\beta_0 + \hat\beta_2}{\hat\beta_1} = x^{*}_{D=0} - \frac{\hat\beta_2}{\hat\beta_1}.$$

The additive model still gives the two groups different thresholds; what it
cannot give them is different *steepness*, since both groups cross the boundary
at the common rate $\hat\beta_1$ per unit of $X$.

## Fitting an interaction model in R

```{r interaction-syntax}
#| eval: false
glm(default ~ balance * student, data = Default, family = "binomial")
```

The `*` expands to `balance + student + balance:student`, so the fitted model
contains both main effects and the product, satisfying hierarchy
automatically. Writing `balance:student` alone would drop the main effects,
violating hierarchy: R responds by giving each group its own balance slope
while forcing both to share a single intercept.

## Example: does balance's effect on default depend on student status?

**Question.** The previous lecture fit `default ~ balance + student`
additively, which forces the student-versus-non-student odds ratio to be
identical at every balance and the balance odds ratio to be identical for
students and non-students. Is that additive assumption right, or does balance
act differently on the two groups?

**Data.** The same `r format(nrow(Default), big.mark = ",")` simulated credit
card customers as in the previous two lectures, with `balance` as the
quantitative feature $X$, the student indicator
$D = \mathrm{I}(\texttt{student} = \texttt{Yes})$, and $Y_i = 1$ when customer
$i$ defaults.

### Fitting both models

```{r default-fits}
default_additive    <- glm(default ~ balance + student,
                           data = Default, family = "binomial")
default_interaction <- glm(default ~ balance * student,
                           data = Default, family = "binomial")
summary(default_interaction)$coefficients
```

::: {.callout-note collapse="true"}
### Code for the table below
```{r default-coef-setup}
#| output: false
Default <- Default |>
  mutate(default01 = as.numeric(default == "Yes"))

bda <- coef(default_additive)
bdi <- coef(default_interaction)

default_coef_table <- tibble(
  term        = names(bdi),
  additive    = c(bda, NA),
  interaction = bdi
) |>
  mutate(across(c(additive, interaction),
                \(x) ifelse(is.na(x), "", sprintf("%.6f", x)))) |>
  knitr::kable(col.names = c("Coefficient", "Additive fit",
                             "Interaction fit"),
               align = "lcc")
```
:::

```{r default-coef-table}
#| echo: false
default_coef_table
```

The two fits agree closely on the intercept and on the `balance` slope, and the
estimated interaction is $\hat\beta_3 = `r signif(bdi[4], 3)`$ —
`r sprintf("%.0f", abs(bdi[2] / bdi[4]))` times smaller than the `balance`
main effect it modifies.

### Comparing the fits on the log-odds scale

Plotting $\hat\eta$ against balance for each model shows the two straight lines
each one produces.

::: {.callout-note collapse="true"}
### Code for the plots below
```{r default-link-setup}
#| output: false
default_grid <- expand_grid(
  balance = seq(0, max(Default$balance), length.out = 300),
  student = factor(c("No", "Yes"), levels = levels(Default$student)))

default_link <- bind_rows(
  default_grid |> mutate(Model = "Additive",
                         eta = predict(default_additive, newdata = default_grid)),
  default_grid |> mutate(Model = "Interaction",
                         eta = predict(default_interaction, newdata = default_grid)))

# A shared base layer, so the two tabs below differ only in which model's
# lines are drawn on it -- the panel itself never changes size or shape.
default_link_base <- ggplot(mapping = aes(x = balance, y = eta,
                                          color = student, linetype = student)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  coord_cartesian(xlim = range(default_grid$balance),
                  ylim = range(default_link$eta)) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Credit card balance ($)", y = expression(hat(eta)),
       color = "Student", linetype = "Student", title = "ISLR2::Default")

default_link_additive <- default_link_base +
  geom_line(data = filter(default_link, Model == "Additive"), linewidth = 0.9)

default_link_interaction <- default_link_base +
  geom_line(data = filter(default_link, Model == "Interaction"), linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Additive
```{r default-link-additive-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_link_additive
```

#### Interaction
```{r default-link-interaction-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_link_interaction
```
:::

The additive tab's two lines are exactly parallel, both with slope
$`r signif(bda[2], 4)`$ log-odds per dollar and separated by
$\hat\beta_2 = `r sprintf("%.3f", bda[3])`$ at every balance. The interaction
tab's two lines have slopes $`r signif(bdi[2], 4)`$ and
$`r signif(bdi[2] + bdi[4], 4)`$, a difference too small to see: switching
between the tabs barely moves the picture.

### Comparing the fits on the probability scale

Building the same comparison up from the raw data puts both fitted models on
the scale a lending decision would use.

::: {.callout-note collapse="true"}
### Code for the plots below
```{r default-prob-setup}
#| output: false
default_prob <- bind_rows(
  default_grid |>
    mutate(Model = "Additive",
           p_hat = predict(default_additive, newdata = default_grid,
                           type = "response")),
  default_grid |>
    mutate(Model = "Interaction",
           p_hat = predict(default_interaction, newdata = default_grid,
                           type = "response")))

# The jitter is computed once, here, and stored as a column rather than left
# to geom_jitter() -- geom_jitter() re-randomizes every time a plot is drawn,
# and each tab below is drawn separately, so leaving it to geom_jitter() would
# put the points in slightly different places from one tab to the next and
# make it look like the data had changed. Seeded so re-rendering the chapter
# does not also churn the committed figure for no reason.
set.seed(20260925)
default_jittered <- Default |>
  mutate(default01_jittered = default01 + runif(n(), -0.03, 0.03))

# The same shared-base idea, with the raw data as the layer every tab keeps.
# The legend is kept on every tab (with more visible swatches than the
# points themselves) rather than suppressed on this one layer, so the panel
# is the same width in all three tabs -- an unequal panel width across tabs
# is the same "the data looks different" problem the jitter fix above solves.
default_prob_base <- ggplot(mapping = aes(x = balance, y = p_hat,
                                          color = student, linetype = student)) +
  geom_point(data = default_jittered,
             aes(y = default01_jittered, shape = student),
             alpha = 0.08, size = 0.7) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  coord_cartesian(xlim = range(default_grid$balance), ylim = c(-0.05, 1.05)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Student", linetype = "Student", shape = "Student",
       title = "ISLR2::Default")

default_prob_data <- default_prob_base

default_prob_additive <- default_prob_base +
  geom_line(data = filter(default_prob, Model == "Additive"), linewidth = 0.9)

default_prob_interaction <- default_prob_base +
  geom_line(data = filter(default_prob, Model == "Interaction"), linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Data
```{r default-prob-data-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_prob_data
```

#### + Additive
```{r default-prob-additive-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_prob_additive
```

#### + Interaction
```{r default-prob-interaction-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_prob_interaction
```
:::

The data tab replots the previous lecture's jittered scatter in a single
panel, with students and non-students distinguished by color and shape. The
two model tabs are nearly the same picture: two S-curves of all but identical
shape, the student curve shifted right, whether or not
the interaction is in the model.

### Testing the interaction

::: {.callout-note collapse="true"}
### Code for the values quoted below
```{r default-inference}
#| output: false
dcoef   <- summary(default_interaction)$coefficients
z_d     <- dcoef["balance:studentYes", "z value"]
p_d     <- dcoef["balance:studentYes", "Pr(>|z|)"]
ci_d    <- confint.default(default_interaction)["balance:studentYes", ]

or_d <- tibble(
  student = c("No (reference)", "Yes"),
  slope   = c(bdi[2], bdi[2] + bdi[4]),
  or100   = exp(100 * c(bdi[2], bdi[2] + bdi[4]))
)

or_d_table <- or_d |>
  mutate(slope = sprintf("%.6f", slope),
         or100 = sprintf("%.3f", or100)) |>
  knitr::kable(col.names = c("Student", "Slope (log-odds per \\$1)",
                             "Odds ratio per \\$100"),
               align = "lcc")

aic_d_table <- tibble(
  Model = c("`default ~ balance + student`", "`default ~ balance * student`"),
  k     = c(length(bda), length(bdi)),
  AIC   = sprintf("%.2f", c(AIC(default_additive), AIC(default_interaction)))
) |>
  knitr::kable(col.names = c("Model", "Parameters", "AIC"), align = "lcc")
```
:::

```{r default-or-table}
#| echo: false
or_d_table
```

The two per-\$100 odds ratios differ by
`r sprintf("%.2f", abs(diff(or_d$or100)))`, which is the difference the
interaction was added to capture. The $z$-test asks whether that difference is
larger than sampling error can explain:
$z = `r sprintf("%.2f", z_d)`$ with a $p$-value of
$`r sprintf("%.3f", p_d)`$, and the 95% confidence interval for $\beta_3$
runs from $`r sprintf("%.5f", ci_d[1])`$ to $`r sprintf("%.5f", ci_d[2])`$,
comfortably containing $0$.

::: {.callout-note collapse="true"}
### Beyond this course
The **Akaike information criterion**, $\text{AIC} = -2\ell(\hat\beta) + 2k$ for
a model with $k$ estimated parameters, scores a fit by its log-likelihood with
a penalty for the number of parameters it used; lower is better. It gives a
second, test-free reading of whether a term earns its place.

```{r default-aic-table}
#| echo: false
aic_d_table
```

Adding the interaction makes AIC *worse* here: the extra parameter buys less
log-likelihood than it costs.
:::

**Conclusion.** Balance's effect on the odds of default does not depend on
student status in these data. The estimated interaction is tiny, the two
fitted models are visually indistinguishable on both scales, and the $z$-test
gives no reason to prefer the more complicated one. The additive model of the
previous lecture is the one to report. The decision boundaries move
correspondingly little: the balance at which $\hat p = 0.5$ shifts from
\$`r format(round(-bda[1] / bda[2]), big.mark = ",")` to
\$`r format(round(-bdi[1] / bdi[2]), big.mark = ",")` for non-students and from
\$`r format(round(-(bda[1] + bda[3]) / bda[2]), big.mark = ",")` to
\$`r format(round(-(bdi[1] + bdi[3]) / (bdi[2] + bdi[4])), big.mark = ",")` for
students.

Not every plausible interaction is a real one. Fitting it was still the right
move: the additive assumption was checked rather than assumed.

## Example: does a store's price effect depend on the store?

**Question.** Citrus Hill and Minute Maid are competing orange juice brands. If
a store raises the price of Citrus Hill, fewer customers should buy it — but
does price act the same way in every store, or does the size of its effect
depend on where the purchase happens?

**Data.** `ISLR2::OJ` records `r format(nrow(OJ), big.mark = ",")` individual
orange juice purchases. These are observed purchases, not simulated ones. Three
of its `r ncol(OJ)` variables are used here:

- `Purchase`: which brand the customer bought, a factor with levels
`r paste(levels(OJ$Purchase), collapse = " and ")` (Citrus Hill and Minute
Maid). This is the response.
- `SalePriceCH`: the price actually paid for Citrus Hill, in dollars, ranging
from \$`r sprintf("%.2f", min(OJ$SalePriceCH))` to
\$`r sprintf("%.2f", max(OJ$SalePriceCH))`. This is the quantitative feature
$X$.
- `Store7`: whether the purchase happened at store 7, a factor with
`r sum(OJ$Store7 == "No")` `No` and `r sum(OJ$Store7 == "Yes")` `Yes`
purchases. This supplies the indicator
$D = \mathrm{I}(\texttt{Store7} = \texttt{Yes})$.

Buying Citrus Hill is the event of interest, so $Y_i = 1$ when purchase $i$ is
a Citrus Hill purchase. `glm()` models the probability of a factor's *second*
level, and `r levels(OJ$Purchase)[2]` is second, so the event has to be set
explicitly with `relevel()`:

```{r oj-relevel}
OJ <- OJ |> mutate(Purchase = relevel(Purchase, ref = "MM"))
levels(OJ$Purchase)
```

With `MM` as the reference level, every model below estimates
$P(\texttt{Purchase} = \texttt{CH})$.

### Exploring the data

Before fitting anything, plot each purchase against the price paid for
Citrus Hill, split by whether the store was store 7:

::: {.callout-note collapse="true"}
### Code for the figure below
```{r oj-eda}
#| output: false
OJ <- OJ |> mutate(ch01 = as.numeric(Purchase == "CH"))

oj_rates <- OJ |>
  group_by(Store7, SalePriceCH) |>
  summarize(n = n(), rate = mean(ch01), .groups = "drop")

oj_eda_plot <- ggplot(OJ, aes(x = SalePriceCH, y = ch01)) +
  geom_jitter(height = 0.05, width = 0.008, alpha = 0.15, size = 0.8) +
  geom_point(data = oj_rates, aes(y = rate, size = n),
             shape = 1, color = "#D55E00") +
  facet_wrap(~ Store7, labeller = label_both) +
  scale_y_continuous(breaks = c(0, 0.5, 1),
                     labels = c("MM (0)", "0.5", "CH (1)")) +
  scale_size_area(max_size = 5) +
  labs(x = "Citrus Hill sale price ($)", y = "Purchase",
       size = "Purchases\nat that price", title = "ISLR2::OJ")

cheap_max     <- 1.70
expensive_min <- 1.90

oj_ends <- OJ |>
  filter(SalePriceCH <= cheap_max | SalePriceCH >= expensive_min) |>
  mutate(band = ifelse(SalePriceCH <= cheap_max, "cheap", "expensive")) |>
  group_by(Store7, band) |>
  summarize(n = n(), rate = mean(ch01), .groups = "drop")

end_of <- function(store, band, column) {
  oj_ends[[column]][oj_ends$Store7 == store & oj_ends$band == band]
}
```
:::

```{r oj-eda-figure}
#| echo: false
#| fig-width: 7.5
#| fig-height: 3.4
oj_eda_plot
```

The open circles are the observed proportion of Citrus Hill purchases at each
distinct price, sized by how many purchases were made at that price. Citrus
Hill takes `r sprintf("%.0f", 100 * mean(OJ$ch01[OJ$Store7 == "Yes"]))`% of
purchases at store 7 against
`r sprintf("%.0f", 100 * mean(OJ$ch01[OJ$Store7 == "No"]))`% elsewhere.

Within each panel the proportions move with price, and they move in opposite
directions. Comparing the two ends of the price range, at store 7 Citrus Hill's
share falls from
`r sprintf("%.0f", 100 * end_of("Yes", "cheap", "rate"))`% of the
`r end_of("Yes", "cheap", "n")` purchases priced at or below
\$`r sprintf("%.2f", cheap_max)` to
`r sprintf("%.0f", 100 * end_of("Yes", "expensive", "rate"))`% of the
`r end_of("Yes", "expensive", "n")` purchases at or above
\$`r sprintf("%.2f", expensive_min)`. Elsewhere it *rises* across the same two
bands, from `r sprintf("%.0f", 100 * end_of("No", "cheap", "rate"))`% of
`r end_of("No", "cheap", "n")` purchases to
`r sprintf("%.0f", 100 * end_of("No", "expensive", "rate"))`% of
`r end_of("No", "expensive", "n")`.

### Fitting both models

```{r oj-fits}
oj_additive    <- glm(Purchase ~ SalePriceCH + Store7,
                      data = OJ, family = "binomial")
oj_interaction <- glm(Purchase ~ SalePriceCH * Store7,
                      data = OJ, family = "binomial")
summary(oj_interaction)$coefficients
```

::: {.callout-note collapse="true"}
### Code for the table below
```{r oj-coef-setup}
#| output: false
boa <- coef(oj_additive)
boi <- coef(oj_interaction)

oj_coef_table <- tibble(
  term        = names(boi),
  additive    = c(boa, NA),
  interaction = boi
) |>
  mutate(across(c(additive, interaction),
                \(x) ifelse(is.na(x), "", sprintf("%.3f", x)))) |>
  knitr::kable(col.names = c("Coefficient", "Additive fit",
                             "Interaction fit"),
               align = "lcc")
```
:::

```{r oj-coef-table}
#| echo: false
oj_coef_table
```

The additive fit puts a single `SalePriceCH` coefficient of
$`r sprintf("%.3f", boa[2])`$ on both stores. The interaction fit replaces it
with $`r sprintf("%.3f", boi[2])`$ and
$`r sprintf("%.3f", boi[2] + boi[4])`$, which have opposite signs.

### Comparing the fits on the log-odds scale

Plotting $\hat\eta$ against price for each store shows what those two signs
mean for the fitted lines:

::: {.callout-note collapse="true"}
### Code for the plots below
```{r oj-link-setup}
#| output: false
oj_grid <- expand_grid(
  SalePriceCH = seq(min(OJ$SalePriceCH), max(OJ$SalePriceCH), length.out = 300),
  Store7 = factor(c("No", "Yes"), levels = levels(OJ$Store7)))

oj_link <- bind_rows(
  oj_grid |> mutate(Model = "Additive",
                    eta = predict(oj_additive, newdata = oj_grid)),
  oj_grid |> mutate(Model = "Interaction",
                    eta = predict(oj_interaction, newdata = oj_grid)))

# A shared base layer, so the two tabs below differ only in which model's
# lines are drawn on it.
oj_link_base <- ggplot(mapping = aes(x = SalePriceCH, y = eta,
                                     color = Store7, linetype = Store7)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  coord_cartesian(xlim = range(oj_grid$SalePriceCH),
                  ylim = range(oj_link$eta)) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Citrus Hill sale price ($)", y = expression(hat(eta)),
       color = "Store 7", linetype = "Store 7", title = "ISLR2::OJ")

oj_link_additive <- oj_link_base +
  geom_line(data = filter(oj_link, Model == "Additive"), linewidth = 0.9)

oj_link_interaction <- oj_link_base +
  geom_line(data = filter(oj_link, Model == "Interaction"), linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Additive
```{r oj-link-additive-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
oj_link_additive
```

#### Interaction
```{r oj-link-interaction-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
oj_link_interaction
```
:::

The additive tab shows the parallel lines the additive model requires, both
sloping down at $`r sprintf("%.3f", boa[2])`$ log-odds per dollar. The
interaction tab shows lines of opposite sign: the store 7 line falls steeply
while the other rises, and they cross at
\$`r sprintf("%.2f", -boi[3] / boi[4])`, inside the observed price range.
Switching between these two tabs changes the picture substantially, which is
what the `Default` example did not do.

### Comparing the fits on the probability scale

The same two model fits, now on the scale a purchase decision actually uses,
with the observed proportions overlaid:

::: {.callout-note collapse="true"}
### Code for the plots below
```{r oj-prob-setup}
#| output: false
oj_prob <- bind_rows(
  oj_grid |> mutate(Model = "Additive",
                    p_hat = predict(oj_additive, newdata = oj_grid,
                                    type = "response")),
  oj_grid |> mutate(Model = "Interaction",
                    p_hat = predict(oj_interaction, newdata = oj_grid,
                                    type = "response")))

# The same shared-base idea, with the observed proportions as the layer every
# tab keeps.
oj_prob_base <- ggplot(mapping = aes(x = SalePriceCH, y = p_hat,
                                     color = Store7, linetype = Store7)) +
  geom_hline(yintercept = 0.5, color = "grey60", linewidth = 0.3) +
  geom_point(data = oj_rates, aes(y = rate, size = n, shape = Store7),
             alpha = 0.8) +
  coord_cartesian(xlim = range(oj_grid$SalePriceCH), ylim = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_size_area(max_size = 5) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Citrus Hill sale price ($)", y = "P(Purchase = CH)",
       color = "Store 7", linetype = "Store 7", shape = "Store 7",
       size = "Purchases\nat that price", title = "ISLR2::OJ")

oj_prob_data <- oj_prob_base

oj_prob_additive <- oj_prob_base +
  geom_line(data = filter(oj_prob, Model == "Additive"), linewidth = 0.9)

oj_prob_interaction <- oj_prob_base +
  geom_line(data = filter(oj_prob, Model == "Interaction"), linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Data
```{r oj-prob-data-plot}
#| echo: false
#| fig-width: 7.5
#| fig-height: 3.2
oj_prob_data
```

#### + Additive
```{r oj-prob-additive-plot}
#| echo: false
#| fig-width: 7.5
#| fig-height: 3.2
oj_prob_additive
```

#### + Interaction
```{r oj-prob-interaction-plot}
#| echo: false
#| fig-width: 7.5
#| fig-height: 3.2
oj_prob_interaction
```
:::

The data tab shows the observed proportion of Citrus Hill purchases at each
price, by store. The additive tab fits them with two curves of identical shape
that never cross, both drifting down with price; the store 7 curve sits above
the other but is forced to decline at the same log-odds rate. The interaction
tab lets store 7's curve fall steeply while the other edges upward, and the two
meet near the top of the observed price range.

### Interpreting the two price slopes

Reading the two group-specific slopes off the interaction fit, and converting
each to an odds ratio, makes the reversal concrete:

::: {.callout-note collapse="true"}
### Code for the table below
```{r oj-slope-setup}
#| output: false
slope_no   <- unname(boi[2])
slope_yes  <- unname(boi[2] + boi[4])

oj_slope_table <- tibble(
  store = c("Not store 7 (reference)", "Store 7"),
  param = c("$\\beta_1$", "$\\beta_1 + \\beta_3$"),
  slope = c(slope_no, slope_yes),
  or10  = exp(0.10 * c(slope_no, slope_yes))
) |>
  mutate(slope = sprintf("%.3f", slope),
         or10  = sprintf("%.3f", or10)) |>
  knitr::kable(col.names = c("Store", "Slope parameter",
                             "Estimate (log-odds per \\$1)",
                             "Odds ratio per \\$0.10"),
               align = "llcc")

or10_no  <- exp(0.10 * slope_no)
or10_yes <- exp(0.10 * slope_yes)
```
:::

```{r oj-slope-table}
#| echo: false
oj_slope_table
```

At store 7, each additional \$0.10 on the Citrus Hill price multiplies the
estimated odds of a Citrus Hill purchase by
`r sprintf("%.3f", or10_yes)` — a
`r round(100 * (1 - or10_yes))`% drop in the odds, the direction a price
increase is expected to push demand. Everywhere else the same \$0.10 multiplies
the odds by `r sprintf("%.3f", or10_no)`, a
`r round(100 * (or10_no - 1))`% *increase*.

That second estimate runs backwards from what price alone would predict, and
the observed proportions in the exploratory plot show the same thing without
any model: outside store 7, Citrus Hill's share does not fall as its price
rises. What these data cannot say is why — `SalePriceCH` records what was
paid, not what else varied alongside it across stores and weeks, so the association it
carries need not be the effect of price by itself. The honest report is the
pattern and its limits, not a mechanism the data cannot supply.

### Break-even prices

The decision boundary here is the Citrus Hill price at which the model puts the
two brands at 50/50, computed separately for each store.

::: {.callout-note collapse="true"}
### Code for the table below
```{r oj-boundaries}
#| output: false
oj_boundary <- function(b, interaction = TRUE) {
  b3 <- if (interaction) b[4] else 0
  c(`Not store 7` = unname(-b[1] / b[2]),
    `Store 7`     = unname(-(b[1] + b[3]) / (b[2] + b3)))
}

oj_boundaries <- bind_rows(
  tibble(Model = "Additive",
         Store = c("Not store 7", "Store 7"),
         price = oj_boundary(boa, interaction = FALSE)),
  tibble(Model = "Interaction",
         Store = c("Not store 7", "Store 7"),
         price = oj_boundary(boi)))

oj_boundary_table <- oj_boundaries |>
  mutate(observed = ifelse(price >= min(OJ$SalePriceCH) &
                           price <= max(OJ$SalePriceCH), "yes", "no"),
         price = sprintf("\\$%.2f", price)) |>
  relocate(observed, .after = price) |>
  knitr::kable(col.names = c("Model", "Store", "Price where $\\hat p = 0.5$",
                             "In observed range?"),
               align = "llcc")

boundary_at <- function(m, s) {
  oj_boundaries$price[oj_boundaries$Model == m & oj_boundaries$Store == s]
}
```
:::

```{r oj-boundary-table}
#| echo: false
oj_boundary_table
```

Under the additive model, store 7's break-even price is
\$`r sprintf("%.2f", boundary_at("Additive", "Store 7"))`, well beyond the
\$`r sprintf("%.2f", max(OJ$SalePriceCH))` maximum any customer in the data
actually paid: a single shared slope can only explain store 7's high Citrus
Hill share by pushing its break-even point outside the observed prices
entirely. The interaction model brings it back to
\$`r sprintf("%.2f", boundary_at("Interaction", "Store 7"))`, inside the price
range, and moves the other stores' break-even point from
\$`r sprintf("%.2f", boundary_at("Additive", "Not store 7"))` to
\$`r sprintf("%.2f", boundary_at("Interaction", "Not store 7"))`. All four
thresholds differ, and only the interaction model places both of its own
within the prices customers actually paid.

### Testing the interaction

::: {.callout-note collapse="true"}
### Code for the values quoted below
```{r oj-inference}
#| output: false
ocoef <- summary(oj_interaction)$coefficients
z_o   <- ocoef["SalePriceCH:Store7Yes", "z value"]
p_o   <- ocoef["SalePriceCH:Store7Yes", "Pr(>|z|)"]
ci_o  <- confint.default(oj_interaction)["SalePriceCH:Store7Yes", ]

aic_o_table <- tibble(
  Model = c("`Purchase ~ SalePriceCH + Store7`",
            "`Purchase ~ SalePriceCH * Store7`"),
  k     = c(length(boa), length(boi)),
  AIC   = sprintf("%.2f", c(AIC(oj_additive), AIC(oj_interaction)))
) |>
  knitr::kable(col.names = c("Model", "Parameters", "AIC"), align = "lcc")
```
:::

The interaction coefficient is $\hat\beta_3 = `r sprintf("%.3f", boi[4])`$ with
$z = `r sprintf("%.2f", z_o)`$ and a $p$-value of $`r signif(p_o, 2)`$. The 95% confidence
interval for $\beta_3$ runs from $`r sprintf("%.2f", ci_o[1])`$ to
$`r sprintf("%.2f", ci_o[2])`$, entirely below $0$, so the interval for
$e^{\beta_3}$ lies entirely below $1$: store 7's per-dollar odds ratio is
firmly smaller than the other stores'.

::: {.callout-note collapse="true"}
### Beyond this course
The same AIC comparison used in the previous example, applied here:

```{r oj-aic-table}
#| echo: false
aic_o_table
```

The interaction lowers AIC by
`r sprintf("%.1f", AIC(oj_additive) - AIC(oj_interaction))`, so the extra
parameter earns more log-likelihood than it costs — the opposite of the verdict
in the `Default` example.
:::

**Conclusion.** The association between Citrus Hill's price and the chance a
customer buys it differs sharply between store 7 and the rest. At store 7 the
estimated odds fall `r round(100 * (1 - or10_yes))`% per \$0.10; elsewhere they
do not fall at all. The $z$-test rejects $H_0: \beta_3 = 0$ decisively
($z = `r sprintf("%.2f", z_o)`$), the two models' fitted lines diverge visibly
on both scales, and the interaction model is the one to report — while noting
that the non-store-7 slope points the wrong way for a price effect and that
these data cannot say why.

## Two examples, two verdicts

Both examples asked the same question with the same syntax, the same estimator,
and the same test. `Default` answered no: the fitted slopes differed by almost
nothing, the curves were indistinguishable, and the $p$-value was
$`r sprintf("%.3f", p_d)`$. `OJ` answered yes: the fitted slopes had opposite
signs and the $p$-value was $`r signif(p_o, 2)`$.

Nothing about either pair of features announced which answer was coming. The
additive assumption is checkable, and checking it is the only way to find out —
which is a reason to fit the interaction even when you expect the additive
model to survive.

## Polynomials and step functions carry over

Interactions are not the only flexibility available. Everything built in the
feature engineering and flexibility lectures for linear regression transfers to
logistic regression without modification, because those lectures worked on the
linear predictor and the linear predictor is unchanged here.

Recall the structure. The log-odds form is

$$\log\left(\frac{p(X_i)}{1 - p(X_i)}\right) = x_i\beta,$$

so anything that can be a column of the model matrix $\mathbf{X}$ is available:
a squared feature, a step-function indicator
$C_j(X) = \mathrm{I}(c_j \le X < c_{j+1})$ built from cut points
$c_1 < c_2 < \cdots$ exactly as in the flexibility lecture, a log-transformed
feature, a product of two features. The logistic function wrapped around
$x_i\beta$ never inspects what those columns contain.

In R this means the same formula syntax:

```{r flexible-syntax}
#| eval: false
glm(y ~ poly(x, 3),       data = d, family = "binomial")  # cubic in x
glm(y ~ cut(x, breaks),   data = d, family = "binomial")  # step function in x
glm(y ~ poly(x, 2) * sex, data = d, family = "binomial")  # separate quadratics by group
```

Three things do change, and all three are consequences of the logistic form
rather than of the features:

1. **The fitted curve is flexible on the log-odds scale**, and the logistic
function then bends it toward $0$ and $1$. A cubic in the log-odds is not a
cubic in $p(X)$.
2. **Coefficients are read as changes in log-odds and odds ratios**, exactly as
in the simple logistic regression lecture. For a step function, $\beta_j$ is
the log-odds difference between bin $j$ and the baseline bin, so
$e^{\beta_j}$ is the odds ratio between those two bins — where in linear
regression $\beta_j$ was a difference between bin means.
3. **Estimation is maximum likelihood, not least squares.** The extra columns
add parameters to the same Bernoulli log-likelihood, which `glm()` maximizes
numerically.

The costs are also unchanged. More flexible features fit the training data
better and estimate more parameters from the same $n$, which is the
bias-variance tradeoff of the flexibility lecture; a high-degree polynomial
extrapolates just as badly in the log-odds as it does in a mean response.

## Conclusion

An interaction lets a feature's effect on the log-odds depend on another
feature's value. With a quantitative feature and an indicator, it gives each
group its own log-odds slope — $\beta_1$ and $\beta_1 + \beta_3$ — and so its
own odds ratio for that feature, with $e^{\beta_3}$ the ratio between the two.
That is the same interaction structure used in linear regression, evaluated on
the log-odds scale instead of the mean-response scale, and fit by the same
maximum likelihood machinery as every other logistic regression.

The diagnostic is a picture on the linear-predictor scale: parallel lines mean
the additive model is adequate, non-parallel lines mean it is not.
`ISLR2::Default` produced lines that were parallel to the eye and an
interaction the $z$-test could not distinguish from zero. `ISLR2::OJ` produced
lines with opposite slopes, an interaction
$\hat\beta_3 = `r sprintf("%.2f", boi[4])`$ that the same test rejected
decisively, and two stores whose break-even prices sit
\$`r sprintf("%.2f", abs(diff(oj_boundaries$price[oj_boundaries$Model == "Interaction"])))`
apart.

Polynomials, step functions, and interactions all enter logistic regression as
columns of the same model matrix, so the model can be made as flexible as the
data support. Adding columns to a binary-response model can also break the
estimation itself in a way that least squares never breaks, a failure that goes
by the name separation.
