# Flexible Logistic Regression

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
set.seed(20260919)
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **explain** how interactions, polynomials, and step functions all enter
logistic regression as basis-function columns of the linear predictor, and
**interpret** a coefficient on any such column as a change in log-odds,
- **distinguish** categorical-categorical, continuous-categorical, and
continuous-continuous interactions, and **explain** why a coefficient's raw
magnitude does not by itself indicate whether an interaction matters,
- **write** a logistic regression model with an interaction between a
quantitative feature and an indicator, **derive** the two group-specific
log-odds slopes $\beta_1$ and $\beta_1 + \beta_3$, and **interpret**
$e^{\beta_1}$ and $e^{\beta_1 + \beta_3}$ as group-specific odds ratios,
- **compare** an additive fit against an interaction fit on the
linear-predictor scale, where the additive model forces two *parallel* lines,
and **translate** that comparison onto the probability scale and the decision
boundary,
- **test** a single interaction coefficient with a $z$-statistic and several
at once with a drop-in-deviance test, **recognizing** that a plausible
interaction often turns out not to be, and
- **describe** how raising a polynomial's degree adds curvature to the linear
predictor, and, from a picture, how a fitted curve changes as a step
function's number of bins increases.
:::

Additive logistic regression gives each feature its own term in the linear
predictor and assumes nothing else. Three ways to relax that assumption carry
over from linear regression unchanged: **interactions**, which let one
feature's effect depend on another feature's value; **polynomials**, which let
a single feature's effect curve instead of running straight; and **step
functions**, which let a single feature's effect jump between bins instead of
following any smooth curve at all.

Each of these is a choice of **basis function** $h_j(X)$ from the feature
engineering lecture — a squared term, a product of two features, an indicator
for a range of $X$ — entering the linear predictor as its own column,

$$\eta = \beta_0 + \beta_1 h_1(X) + \beta_2 h_2(X) + \cdots.$$

Logistic regression changes only what happens to $\eta$ next:
$p(X) = \text{logistic}(\eta)$ in place of $E[Y \mid X] = \eta$, and
$\hat\beta$ maximizes the Bernoulli log-likelihood in place of minimizing a sum
of squared residuals. Every coefficient is still read the way earlier lectures
read it — $\beta_j$ a change in log-odds per unit of $h_j(X)$ with every other
column held fixed, $e^{\beta_j}$ the corresponding odds ratio — whatever $h_j$
turns out to be. That "held fixed" is worth flagging now: it breaks down
whenever two columns move together, which is exactly what happens for $X$ and
$X^2$, or for $X$ and $XD$, so it will need qualifying again once those columns
appear below. What changes from one choice of $h_j$ to the next is only what a
"one-unit change" of it means.

This lecture works through interactions first, since they raise the most that
is genuinely new, then polynomials and step functions more briefly, since the
transfer is mostly mechanical.

## Interactions

### Additive assumption

The previous lecture's model was **additive**: each feature contributed its own
term $\beta_j X_{ij}$ to the linear predictor, and no term involved two
features at once. That structure is what licensed the interpretation "a
one-unit increase in $X_j$ multiplies the odds by $e^{\beta_j}$, holding every
other feature fixed" — the same $e^{\beta_j}$ at every value of every other
feature.

That is an assumption, not a convenience. The fitted curves of the previous
lecture showed it directly: two curves of identical shape, separated by a
constant $\hat\beta_2$ on the log-odds scale at every balance.

There is no reason a real process has to work that way. Whether it does is a
question about the data, and it can be asked and answered with a model that
does not impose it. What that model looks like depends on whether the two
interacting features are categorical or quantitative.

### Categorical-categorical interactions

When both features are categorical, an interaction adds one indicator for
every combination of their non-reference levels beyond what the two main
effects already supply. With indicators $D_1$ and $D_2$ for two binary
features,

$$\eta = \beta_0 + \beta_1 D_1 + \beta_2 D_2 + \beta_3 D_1 D_2,$$

and the product $D_1 D_2$ is $1$ only when an observation is at the
non-reference level of both features, so $\beta_3$ is the *extra* log-odds a
case gets from being at both non-reference levels at once, beyond what
$\beta_1$ and $\beta_2$ already give it separately.

A quantitative explanatory variable becomes categorical the same way it did in
the flexibility lecture: pick a cutoff and take an indicator.
Splitting `balance` at \$1,500 turns the running `Default` example into a
categorical-categorical case, `balance_cat * student`, with four
balance-by-student cells to describe. The additive model spends three
coefficients on them and forces a single student effect to serve at both
balance levels; the interaction model spends four, one per cell.

Both fits are drawn below against the four observed cells, on the log-odds
scale at left and the probability scale at right.

::: {.callout-note collapse="true"}
### Code
```{r cat-cat-setup}
#| output: false
# default01 and default_jit are used by figures throughout this lecture, so
# both columns are added here, at the first figure that needs either. The
# jitter is drawn once into a column rather than left to geom_jitter(), which
# would re-randomize every time a tab is drawn and make the same points land
# in different places from one tab to the next.
Default <- Default |>
  mutate(balance_cat = factor(ifelse(balance > 1500, "High", "Low"),
                              levels = c("Low", "High")),
         default01 = as.numeric(default == "Yes"),
         default_jit = default01 + runif(n(), -0.03, 0.03))

cat_add        <- glm(default ~ balance_cat + student, data = Default,
                      family = "binomial")
balance_cat_glm <- glm(default ~ balance_cat * student, data = Default,
                       family = "binomial")

cat_cells <- expand_grid(
  balance_cat = factor(c("Low", "High"), levels = c("Low", "High")),
  student     = factor(c("No", "Yes"), levels = levels(Default$student)))

cat_obs <- Default |>
  group_by(balance_cat, student) |>
  summarize(rate = mean(default01), .groups = "drop") |>
  mutate(eta = log(rate / (1 - rate)))

cat_fit <- bind_rows(
  cat_cells |> mutate(Model = "Additive",
                      eta = predict(cat_add, newdata = cat_cells)),
  cat_cells |> mutate(Model = "Interaction",
                      eta = predict(balance_cat_glm, newdata = cat_cells))) |>
  mutate(p_hat = plogis(eta))

cat_x <- scale_x_discrete(labels = c(Low = "Up to $1,500",
                                     High = "Above $1,500"))

# The same shared-base idea used throughout this lecture: one base layer
# carrying the observed cells, with each tab adding only its own fit. One
# line per student group joins that group's two balance categories, so a
# model's student effect is the vertical gap between the two lines.
cat_base <- ggplot(mapping = aes(x = balance_cat, color = student,
                                 shape = student, linetype = student,
                                 group = student)) +
  cat_x +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Credit card balance", color = "Student", shape = "Student",
       linetype = "Student", title = "ISLR2::Default")

cat_link_base <- cat_base +
  geom_point(data = cat_obs, aes(y = eta), size = 3) +
  coord_cartesian(ylim = range(c(cat_obs$eta, cat_fit$eta))) +
  labs(y = expression(hat(eta)))

cat_prob_base <- cat_base +
  geom_point(data = cat_obs, aes(y = rate), size = 3) +
  coord_cartesian(ylim = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  labs(y = "P(default = Yes)")

cat_link_data        <- cat_link_base
cat_link_additive    <- cat_link_base +
  geom_line(data = filter(cat_fit, Model == "Additive"), aes(y = eta),
            linewidth = 0.9)
cat_link_interaction <- cat_link_base +
  geom_line(data = filter(cat_fit, Model == "Interaction"), aes(y = eta),
            linewidth = 0.9)

cat_prob_data        <- cat_prob_base
cat_prob_additive    <- cat_prob_base +
  geom_line(data = filter(cat_fit, Model == "Additive"), aes(y = p_hat),
            linewidth = 0.9)
cat_prob_interaction <- cat_prob_base +
  geom_line(data = filter(cat_fit, Model == "Interaction"), aes(y = p_hat),
            linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Data

::: {layout-ncol=2}
```{r cat-cat-data-link-plot}
#| echo: false
#| fig-width: 4.5
#| fig-height: 3.4
cat_link_data
```

```{r cat-cat-data-prob-plot}
#| echo: false
#| fig-width: 4.5
#| fig-height: 3.4
cat_prob_data
```
:::

#### + Additive

::: {layout-ncol=2}
```{r cat-cat-additive-link-plot}
#| echo: false
#| fig-width: 4.5
#| fig-height: 3.4
cat_link_additive
```

```{r cat-cat-additive-prob-plot}
#| echo: false
#| fig-width: 4.5
#| fig-height: 3.4
cat_prob_additive
```
:::

#### + Interaction

::: {layout-ncol=2}
```{r cat-cat-interaction-link-plot}
#| echo: false
#| fig-width: 4.5
#| fig-height: 3.4
cat_link_interaction
```

```{r cat-cat-interaction-prob-plot}
#| echo: false
#| fig-width: 4.5
#| fig-height: 3.4
cat_prob_interaction
```
:::
:::

```{r cat-cat-gaps}
#| include: false
cat_gap <- cat_obs |>
  group_by(balance_cat) |>
  summarize(gap = eta[student == "Yes"] - eta[student == "No"],
            .groups = "drop")

cat_climb <- cat_obs |>
  group_by(student) |>
  summarize(climb = eta[balance_cat == "High"] - eta[balance_cat == "Low"],
            .groups = "drop")
```

On the log-odds scale the lines climb
$`r sprintf("%.1f", min(cat_climb$climb))`$–$`r sprintf("%.1f", max(cat_climb$climb))`$
log-odds from the low-balance cells to the high-balance ones, so the student
gap — the vertical distance between the two lines — is a small quantity
inside that climb. The
additive fit's two lines are exactly parallel: one coefficient
$\hat\beta_2 = `r sprintf("%.3f", coef(cat_add)[3])`$ serves as the student
gap at both balance levels, so neither line passes through both of its group's
observed points. The interaction fit's lines pass through all four points —
four coefficients for four cells — with the observed student gap
$`r sprintf("%.3f", cat_gap$gap[cat_gap$balance_cat == "Low"])`$ log-odds below
\$1,500 and $`r sprintf("%.3f", cat_gap$gap[cat_gap$balance_cat == "High"])`$
above it. On the probability scale the low-balance cells collapse together
near $0$, and only the high-balance cells separate visibly.

The same four cells read as fitted probabilities:

::: {.callout-note collapse="true"}
### Code
```{r cat-cat-prob-tables}
#| output: false
cat_cell_table <- function(model_name) {
  cat_fit |>
    filter(Model == model_name) |>
    mutate(p_hat = sprintf("%.4f", p_hat)) |>
    select(balance_cat, student, p_hat) |>
    pivot_wider(names_from = student, values_from = p_hat) |>
    mutate(balance_cat = c(Low = "Up to \\$1,500",
                           High = "Above \\$1,500")[as.character(balance_cat)]) |>
    knitr::kable(col.names = c("Balance", "Non-student", "Student"),
                 align = "lcc")
}

cat_add_table <- cat_cell_table("Additive")
cat_int_table <- cat_cell_table("Interaction")
```
:::

::: {.panel-tabset}
#### Additive
```{r cat-cat-add-table}
#| echo: false
cat_add_table
```

#### Interaction
```{r cat-cat-int-table}
#| echo: false
cat_int_table
```
:::

The interaction model's four probabilities are the four observed cell
proportions exactly, since it has one free coefficient per cell. The additive
model's four are not, since its three coefficients cannot reproduce four
arbitrary numbers.

The coefficients behind those probabilities:

```{r cat-cat-coefficients}
summary(balance_cat_glm)$coefficients
```

```{r cat-cat-extract}
#| include: false
bc_b <- coef(balance_cat_glm)
bc_z <- summary(balance_cat_glm)$coefficients["balance_catHigh:studentYes", "z value"]
```

The reference cell is a non-student with balance at or below \$1,500, at
log-odds $\hat\beta_0 = `r sprintf("%.3f", bc_b[1])`$. Crossing \$1,500 shifts
that by $\hat\beta_1 = `r sprintf("%.3f", bc_b[2])`$, being a student shifts it
by $\hat\beta_2 = `r sprintf("%.3f", bc_b[3])`$, and being *both* a
high-balance customer *and* a student adds a further
$\hat\beta_3 = `r sprintf("%.3f", bc_b[4])`$ beyond the two separately, which
the $z$-statistic $`r sprintf("%.2f", bc_z)`$ cannot distinguish from zero.
Both features are $0/1$ indicators, so $\hat\beta_3$ carries no units and
comparing its size to $\hat\beta_1$ or $\hat\beta_2$ is fair.

The same hypothesis $H_0: \beta_3 = 0$ can be tested by dropping the
interaction column and comparing deviances, using the drop-in-deviance test of
the previous lecture:

```{r cat-cat-deviance}
anova(cat_add, balance_cat_glm, test = "Chisq")
```

```{r cat-cat-deviance-extract}
#| include: false
cat_dev <- anova(cat_add, balance_cat_glm, test = "Chisq")
cat_G2  <- cat_dev$Deviance[2]
cat_p   <- cat_dev[["Pr(>Chi)"]][2]
```

Forcing one student effect on both balance levels costs
$G^2 = `r sprintf("%.3f", cat_G2)`$ of deviance on $`r cat_dev$Df[2]`$ degree
of freedom, $p = `r sprintf("%.3f", cat_p)`$. These data give no evidence that
the student effect differs between the two balance categories.

### Continuous-categorical interactions

The combination easiest to visualize pairs one quantitative feature $X$ with
one indicator $D$:

$$\eta = \beta_0 + \beta_1 X + \beta_2 D + \beta_3 XD.$$

Collecting the terms involving $X$ shows what each coefficient does:

$$\log\left(\frac{p(X,D)}{1 - p(X,D)}\right) = \underbrace{(\beta_0 + \beta_2 D)}_{\text{log-odds intercept}} + \underbrace{(\beta_1 + \beta_3 D)}_{\text{log-odds slope on } X} X.$$

Setting $X = 0$ and $D = 0$ leaves $\beta_0$, the log-odds of the event in the
reference group at $X = 0$, so $e^{\beta_0}$ are the odds and
$\text{logistic}(\beta_0)$ the probability there. Adding $\beta_2$ shifts that
reference value to the indicated group, still at $X = 0$.

The slope on $X$ is a function of $D$, which is the whole point of the
interaction. Writing out the two levels:

- **Reference level ($D = 0$):** log-odds $= \beta_0 + \beta_1 X$. A one-unit
increase in $X$ adds $\beta_1$ to the log-odds and multiplies the odds by
$e^{\beta_1}$.
- **Indicated level ($D = 1$):** log-odds
$= (\beta_0 + \beta_2) + (\beta_1 + \beta_3) X$. A one-unit increase in $X$
adds $\beta_1 + \beta_3$ to the log-odds and multiplies the odds by
$e^{\beta_1 + \beta_3}$.

So there is no single odds ratio for $X$. There are two, one per group, and
their ratio is

$$\frac{e^{\beta_1 + \beta_3}}{e^{\beta_1}} = e^{\beta_3}.$$

$e^{\beta_3}$ is a ratio of odds ratios: the factor by which one group's
per-unit odds ratio for $X$ differs from the other's. $\beta_3 = 0$ makes that
factor $1$ and returns the additive model.

One caution carries over from the feature engineering lecture: $\beta_1$ is the
log-odds slope **when $D = 0$**, not an average across the two groups. A second
is that the $X = 0$ at which $\beta_0$ and $\beta_2$ are defined may be a value
$X$ never takes. Hierarchy applies here as it did in linear regression: keep
$X$ and $D$ in the model whenever $XD$ is present.

```{r interaction-syntax}
#| eval: false
glm(default ~ balance * student, data = Default, family = "binomial")
```

The `*` expands to `balance + student + balance:student`, so the fitted model
contains both main effects and the product, satisfying hierarchy
automatically. Writing `balance:student` alone would drop the main effects,
violating hierarchy: R responds by giving each group its own balance slope
while forcing both to share a single intercept.

Written as functions of $X$, the fitted log-odds for the two groups are

$$\hat\eta(X, 0) = \hat\beta_0 + \hat\beta_1 X, \qquad \hat\eta(X, 1) = \left(\hat\beta_0 + \hat\beta_2\right) + \left(\hat\beta_1 + \hat\beta_3\right) X.$$

Under the additive model $\beta_3 = 0$, both lines have slope
$\hat\beta_1$: they are **parallel**, separated by exactly $\hat\beta_2$ at
every $X$. Under the interaction model the slopes differ by $\hat\beta_3$, so
the lines are not parallel and, extended far enough, must cross. Parallel
versus non-parallel on the linear-predictor scale is the visual signature that
separates the two models. Each group's **decision boundary** — the $X$ where
its own line crosses zero, since $\hat p = 0.5$ there — sits wherever
$\hat\eta(X, d) = 0$ solves for that group's $d$; under the additive model a
constant $\hat\beta_2$ shifts both groups' boundaries by the same amount,
while under the interaction model the differing slopes can move them by
different amounts entirely.

In R, `predict()` on a `glm` object returns both scales:

```{r predict-scales}
#| eval: false
predict(fit, newdata = grid)                    # eta-hat, the linear predictor
predict(fit, newdata = grid, type = "response") # p-hat, the probability
```

Fitting both models on `ISLR2::Default`, with `balance` as $X$ and the student
indicator as $D$, and plotting $\hat\eta$ against balance draws those lines.

::: {.callout-note collapse="true"}
### Code
```{r default-link-setup}
#| output: false
default_additive    <- glm(default ~ balance + student,
                           data = Default, family = "binomial")
default_interaction <- glm(default ~ balance * student,
                           data = Default, family = "binomial")

bda <- coef(default_additive)
bdi <- coef(default_interaction)

default_grid <- expand_grid(
  balance = seq(0, max(Default$balance), length.out = 300),
  student = factor(c("No", "Yes"), levels = levels(Default$student)))

default_link <- bind_rows(
  default_grid |> mutate(Model = "Additive",
                         eta = predict(default_additive, newdata = default_grid)),
  default_grid |> mutate(Model = "Interaction",
                         eta = predict(default_interaction, newdata = default_grid)))

# A shared base layer, so the two tabs below differ only in which model's
# lines are drawn on it -- the panel itself never changes size or shape.
default_link_base <- ggplot(mapping = aes(x = balance, y = eta,
                                          color = student, linetype = student)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  coord_cartesian(xlim = range(default_grid$balance),
                  ylim = range(default_link$eta)) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Credit card balance ($)", y = expression(hat(eta)),
       color = "Student", linetype = "Student", title = "ISLR2::Default")

default_link_additive <- default_link_base +
  geom_line(data = filter(default_link, Model == "Additive"), linewidth = 0.9)

default_link_interaction <- default_link_base +
  geom_line(data = filter(default_link, Model == "Interaction"), linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Additive
```{r default-link-additive-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_link_additive
```

#### Interaction
```{r default-link-interaction-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_link_interaction
```
:::

The additive tab's two lines are exactly parallel, both with slope
$`r signif(bda[2], 4)`$ log-odds per dollar and separated by
$\hat\beta_2 = `r sprintf("%.3f", bda[3])`$ at every balance. The interaction
tab's two lines have slopes $`r signif(bdi[2], 4)`$ and
$`r signif(bdi[2] + bdi[4], 4)`$, a difference too small to see: switching
between the tabs barely moves the picture.

The probability scale is what a decision actually uses, and it is a distorted
view of the same two lines. Passing each line through
$\text{logistic}(\cdot)$ compresses it toward $0$ and $1$, so two parallel
lines become two S-curves of identical shape that never cross, while two
non-parallel lines become two S-curves of different steepness that can cross
within the range of the data. Building the same comparison up from the raw
data puts both fits on that scale.

::: {.callout-note collapse="true"}
### Code
```{r default-prob-setup}
#| output: false
default_prob <- bind_rows(
  default_grid |>
    mutate(Model = "Additive",
           p_hat = predict(default_additive, newdata = default_grid,
                           type = "response")),
  default_grid |>
    mutate(Model = "Interaction",
           p_hat = predict(default_interaction, newdata = default_grid,
                           type = "response")))

# The same shared-base idea, with the raw data as the layer every tab keeps.
# The legend is kept on every tab (with more visible swatches than the
# points themselves) rather than suppressed on this one layer, so the panel
# is the same width in all three tabs -- an unequal panel width across tabs
# looks, to a reader clicking between them, like the data changed.
default_prob_base <- ggplot(mapping = aes(x = balance, y = p_hat,
                                          color = student, linetype = student)) +
  geom_point(data = Default,
             aes(y = default_jit, shape = student),
             alpha = 0.08, size = 0.7) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  coord_cartesian(xlim = range(default_grid$balance), ylim = c(-0.05, 1.05)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Student", linetype = "Student", shape = "Student",
       title = "ISLR2::Default")

default_prob_data <- default_prob_base

default_prob_additive <- default_prob_base +
  geom_line(data = filter(default_prob, Model == "Additive"), linewidth = 0.9)

default_prob_interaction <- default_prob_base +
  geom_line(data = filter(default_prob, Model == "Interaction"), linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Data
```{r default-prob-data-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_prob_data
```

#### + Additive
```{r default-prob-additive-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_prob_additive
```

#### + Interaction
```{r default-prob-interaction-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
default_prob_interaction
```
:::

The data tab plots each customer's outcome, jittered vertically, against
balance, with students and non-students distinguished by color and shape. The
two model tabs are nearly the same picture: two S-curves of all but identical
shape, the student curve shifted right, whether or not the interaction is in
the model.

The coefficients behind the interaction tab, with the $z$-test of
$H_0: \beta_3 = 0$ in its last row:

```{r default-interaction-z}
summary(default_interaction)$coefficients
```

```{r cont-cat-extract}
#| include: false
dz <- summary(default_interaction)$coefficients["balance:studentYes", ]
```

The intercept $\hat\beta_0 = `r sprintf("%.2f", bdi[1])`$ is the log-odds of
default for a non-student carrying no balance at all, a value far outside
anything the data contain. From that reference, $\hat\beta_2$ moves a student
with no balance by $`r sprintf("%.3f", bdi[3])`$ log-odds. The balance slope is
$\hat\beta_1 = `r signif(bdi[2], 4)`$ for non-students and
$\hat\beta_1 + \hat\beta_3 = `r signif(bdi[2] + bdi[4], 4)`$ for students, so
the two per-\$100 odds ratios are
$`r sprintf("%.3f", exp(100 * bdi[2]))`$ and
$`r sprintf("%.3f", exp(100 * (bdi[2] + bdi[4])))`$.

The same hypothesis can be tested by dropping the interaction column and
comparing deviances:

```{r default-interaction-deviance}
anova(default_additive, default_interaction, test = "Chisq")
```

```{r default-interaction-test-extract}
#| include: false
ddev <- anova(default_additive, default_interaction, test = "Chisq")
dG2  <- ddev$Deviance[2]
dp   <- ddev[["Pr(>Chi)"]][2]
```

The $z$-statistic is $`r sprintf("%.3f", dz["z value"])`$ with
$p = `r sprintf("%.3f", dz["Pr(>|z|)"])`$; dropping the interaction raises the
deviance by $G^2 = `r sprintf("%.3f", dG2)`$ on $`r ddev$Df[2]`$ degree of
freedom, giving $p = `r sprintf("%.3f", dp)`$. With one coefficient at stake
the two are different statistics for the same hypothesis, each with an
approximate $\chi^2_1$ null distribution: the Wald statistic reads the
curvature of the log-likelihood at $\hat\beta$, while the deviance compares the
log-likelihood itself at the two fits, which is why
$z^2 = `r sprintf("%.3f", dz["z value"]^2)`$ and $G^2$ come out close without
being equal. Neither gives evidence that balance acts differently on students
than on non-students.

Only the deviance test survives more than one coefficient. An indicator
contributes exactly one interaction column, but a categorical feature with
more than two levels contributes one per non-reference level, and no single
$z$-test asks whether that feature interacts with $X$ *at all*; the
drop-in-deviance test asks the same question for $k$ dropped columns as for
one, $G^2 = D_r - D_f$ approximately $\chi^2_k$ where

- $D_r$ is the deviance of the reduced model,
- $D_f$ is the deviance of the full model, and
- $k$ is the number of interaction columns dropped.

### Continuous-continuous interactions

Two quantitative features multiply into a single new column, $X_1 X_2$:

$$\eta = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_1 X_2.$$

Collecting the terms involving $X_1$ puts it in the same form the
continuous-categorical case took:

$$\log\left(\frac{p(X_1,X_2)}{1 - p(X_1,X_2)}\right) = \underbrace{(\beta_0 + \beta_2 X_2)}_{\text{log-odds intercept}} + \underbrace{(\beta_1 + \beta_3 X_2)}_{\text{log-odds slope on } X_1} X_1.$$

Both the intercept and the slope on $X_1$ are now functions of $X_2$, and
$X_2$ is quantitative, so instead of one intercept and one slope per group
there is a separate line for every value $X_2$ can take. Fixing $X_2$, a
one-unit increase in $X_1$ still changes the log-odds by
$\beta_1 + \beta_3 X_2$ and multiplies the odds by $e^{\beta_1 + \beta_3 X_2}$.

Taking `balance` as $X_1$ and `income` as $X_2$ on `ISLR2::Default`, neither
feature has levels to split the picture by, so the fit is drawn by splitting
one of them: customers are grouped into income quartiles, and each group's
fitted curve is drawn at that group's median income against the full range of
balance. Both models are fit on `income` as recorded; the quartiles only
decide which four curves to draw.

::: {.callout-note collapse="true"}
### Code
```{r cont-cont-link-setup}
#| output: false
Default <- Default |>
  mutate(income_q = cut(income, breaks = quantile(income, 0:4 / 4),
                        include.lowest = TRUE,
                        labels = c("Q1 (lowest)", "Q2", "Q3", "Q4 (highest)")))

balance_income_add <- glm(default ~ balance + income, data = Default,
                          family = "binomial")
balance_income_glm <- glm(default ~ balance * income, data = Default,
                          family = "binomial")

# One curve per income quartile, each evaluated at that quartile's median
# income -- the fit is a function of income itself, so a group needs a single
# income value to be drawn at.
income_median <- Default |>
  group_by(income_q) |>
  summarize(income = median(income), .groups = "drop")

income_grid <- expand_grid(
  balance  = seq(0, max(Default$balance), length.out = 300),
  income_q = income_median$income_q) |>
  left_join(income_median, by = "income_q")

income_link <- bind_rows(
  income_grid |> mutate(Model = "Additive",
                        eta = predict(balance_income_add, newdata = income_grid)),
  income_grid |> mutate(Model = "Interaction",
                        eta = predict(balance_income_glm, newdata = income_grid)))

# Four groups, so shape and linetype cannot simply duplicate a two-level color
# scale; viridis keeps the four distinguishable without color. No Data tab on
# this scale, the same way the continuous-categorical case has none -- raw
# 0/1 outcomes have no finite log-odds to plot.
income_link_base <- ggplot(mapping = aes(x = balance, y = eta,
                                         color = income_q, linetype = income_q)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  coord_cartesian(xlim = range(income_grid$balance),
                  ylim = range(income_link$eta)) +
  scale_color_viridis_d(end = 0.8) +
  labs(x = "Credit card balance ($)", y = expression(hat(eta)),
       color = "Income\nquartile", linetype = "Income\nquartile",
       title = "ISLR2::Default")

income_link_additive <- income_link_base +
  geom_line(data = filter(income_link, Model == "Additive"), linewidth = 0.9)

income_link_interaction <- income_link_base +
  geom_line(data = filter(income_link, Model == "Interaction"), linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Additive
```{r cont-cont-link-additive-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
income_link_additive
```

#### Interaction
```{r cont-cont-link-interaction-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
income_link_interaction
```
:::

Under the additive model the four lines are exactly parallel, each shifted
from the next by a fixed amount of log-odds. Under the interaction model the
slope on balance itself grows with income, so the four lines fan out instead
of staying parallel — the same parallel-versus-not signature the
continuous-categorical case showed, now with four lines instead of two.

The probability scale is what a decision actually uses, and it is a distorted
view of the same four lines: passing each one through $\text{logistic}(\cdot)$
compresses it toward $0$ and $1$.

::: {.callout-note collapse="true"}
### Code
```{r cont-cont-prob-setup}
#| output: false
income_prob <- bind_rows(
  income_grid |> mutate(Model = "Additive",
                        p_hat = predict(balance_income_add,
                                        newdata = income_grid,
                                        type = "response")),
  income_grid |> mutate(Model = "Interaction",
                        p_hat = predict(balance_income_glm,
                                        newdata = income_grid,
                                        type = "response")))

# Observed default rates in $400-wide balance bins within each income
# quartile, dropped where a bin holds fewer than 50 customers.
income_binned <- Default |>
  mutate(bin = cut(balance, breaks = seq(0, 2800, by = 400),
                   include.lowest = TRUE)) |>
  group_by(income_q, bin) |>
  summarize(balance = mean(balance), rate = mean(default01), n = n(),
            .groups = "drop") |>
  filter(n >= 50)

# Four groups, so shape and linetype cannot simply duplicate a two-level
# color scale; viridis keeps the four distinguishable without color.
income_base <- ggplot(mapping = aes(x = balance, y = p_hat,
                                    color = income_q, linetype = income_q)) +
  geom_point(data = income_binned,
             aes(y = rate, size = n, shape = income_q)) +
  coord_cartesian(xlim = range(income_grid$balance), ylim = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_color_viridis_d(end = 0.8) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  scale_size_area(max_size = 4) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Income\nquartile", linetype = "Income\nquartile",
       shape = "Income\nquartile", size = "Customers\nin bin",
       title = "ISLR2::Default")

income_data <- income_base

income_additive <- income_base +
  geom_line(data = filter(income_prob, Model == "Additive"), linewidth = 0.9)

income_interaction <- income_base +
  geom_line(data = filter(income_prob, Model == "Interaction"), linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Data
```{r cont-cont-data-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
income_data
```

#### + Additive
```{r cont-cont-additive-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
income_additive
```

#### + Interaction
```{r cont-cont-interaction-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
income_interaction
```
:::

```{r cont-cont-spread}
#| include: false
income_at_2000 <- income_prob |> filter(abs(balance - 2000) ==
                                        min(abs(balance - 2000)))
income_spread <- income_at_2000 |>
  group_by(Model) |>
  summarize(spread = max(p_hat) - min(p_hat), .groups = "drop")
```

The additive tab draws four curves of identical shape, each shifted from the
next by a fixed amount of log-odds, so no pair of them ever crosses. The
interaction tab lets the balance slope grow with income, and its four curves
pull apart at high balances and close up at low ones: at a balance of \$2,000
the fitted probabilities span
`r sprintf("%.2f", income_spread$spread[income_spread$Model == "Additive"])`
under the additive model and
`r sprintf("%.2f", income_spread$spread[income_spread$Model == "Interaction"])`
under the interaction. Switching between the tabs changes the picture very
little.

The coefficients behind the interaction tab:

```{r cont-cont-fit}
summary(balance_income_glm)$coefficients
```

```{r cont-cont-extract}
#| include: false
bi_b <- coef(balance_income_glm)
bi_z <- summary(balance_income_glm)$coefficients["balance:income", "z value"]
```

The intercept $\hat\beta_0 = `r sprintf("%.2f", bi_b[1])`$ is the log-odds of
default for a customer with no balance and no income, an extrapolation well
outside the data. The estimated interaction,
$\hat\beta_3 \approx `r signif(bi_b[4], 3)`$, looks negligible beside
$\hat\beta_1 \approx `r signif(bi_b[2], 3)`$, but that comparison is not
evidence of anything: `income` is recorded in dollars running into the tens of
thousands, and re-expressing it in thousands of dollars would multiply
$\hat\beta_3$ by $1{,}000$ while leaving every fitted probability, and the
$z$-statistic $`r sprintf("%.3f", bi_z)`$, exactly where they are. Whether the
dependence matters is a question for a test.

```{r cont-cont-deviance}
anova(balance_income_add, balance_income_glm, test = "Chisq")
```

```{r cont-cont-deviance-extract}
#| include: false
bi_dev <- anova(balance_income_add, balance_income_glm, test = "Chisq")
bi_G2  <- bi_dev$Deviance[2]
bi_p   <- bi_dev[["Pr(>Chi)"]][2]
```

Dropping the product column costs $G^2 = `r sprintf("%.3f", bi_G2)`$ of
deviance on $`r bi_dev$Df[2]`$ degree of freedom,
$p = `r sprintf("%.3f", bi_p)`$, agreeing with the $z$-test and with the
figures: these data give no evidence that balance's effect on the log-odds of
default varies with income.

## Polynomials

A polynomial is a basis function like any other: the column $X^2$ joins $X$ in
the linear predictor and lets a single feature's effect curve instead of
running straight. Interacted with an indicator, it gives each group its own
quadratic rather than its own straight line. With `balance` as $X$ and the
student indicator as $D$, the two models are

$$\text{additive:} \qquad \eta = \beta_0 + \beta_1 X + \beta_2 X^2 + \beta_3 D,$$

$$\text{interaction:} \qquad \eta = \beta_0 + \beta_1 X + \beta_2 X^2 + \beta_3 D + \beta_4 XD + \beta_5 X^2 D,$$

where

- $X$ is the credit card balance and $X^2$ its square,
- $D = \mathrm{I}(\texttt{student} = \texttt{Yes})$,
- $\beta_1$ and $\beta_2$ are the reference group's linear and quadratic
coefficients, and
- $\beta_4$ and $\beta_5$ are how the indicated group's linear and quadratic
coefficients differ from the reference group's.

Setting $D$ to each of its two values splits the interaction model into a pair
of quadratics:

$$\text{reference group } (D = 0): \qquad \eta = \beta_0 + \beta_1 X + \beta_2 X^2,$$

$$\text{indicated group } (D = 1): \qquad \eta = (\beta_0 + \beta_3) + (\beta_1 + \beta_4) X + (\beta_2 + \beta_5) X^2.$$

The two quadratics share no coefficient: six parameters buy each group a
curve of its own shape. The additive model spends four instead and gives both
groups the identical shape $\beta_1 X + \beta_2 X^2$, separated by the
constant $\beta_3$ at every balance — the polynomial analogue of the two
parallel lines the continuous-categorical case produced.

In R the pair of columns is `poly(balance, 2)`, and it enters an interaction
the same way a single linear term did, `poly(balance, 2) * student`. `poly()`
returns an orthogonal basis for the same span rather than the raw $X$ and
$X^2$, which changes what the individual coefficients mean but leaves the
fitted values, and the test below, unchanged.

Four fits are drawn against the same raw data below: the additive and
interaction models of the continuous-categorical subsection, whose group
curves are straight lines on the log-odds scale, and the two quadratic models
above.

::: {.callout-note collapse="true"}
### Code
```{r poly-setup}
#| output: false
poly_additive    <- glm(default ~ poly(balance, 2) + student,
                        data = Default, family = "binomial")
poly_interaction <- glm(default ~ poly(balance, 2) * student,
                        data = Default, family = "binomial")

poly_grid <- expand_grid(
  balance = seq(0, max(Default$balance), length.out = 300),
  student = factor(c("No", "Yes"), levels = levels(Default$student)))

# The linear fits are the ones already estimated in the Interactions section,
# so the linear tabs below redraw those models rather than re-fitting them.
poly_models <- list("Additive (linear)"       = default_additive,
                    "Interaction (linear)"    = default_interaction,
                    "Additive (quadratic)"    = poly_additive,
                    "Interaction (quadratic)" = poly_interaction)

poly_pred <- bind_rows(lapply(names(poly_models), function(nm) {
  fit <- poly_models[[nm]]
  poly_grid |> mutate(Model = nm,
                      eta   = predict(fit, newdata = poly_grid),
                      p_hat = predict(fit, newdata = poly_grid,
                                      type = "response"))
}))

# The same shared-base idea used throughout this lecture: one base layer per
# scale, with each tab adding only its own pair of curves. Both bases use a
# single window covering all four fits, so switching tabs moves the curves
# and nothing else.
poly_link_base <- ggplot(mapping = aes(x = balance, y = eta,
                                       color = student, linetype = student)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  coord_cartesian(xlim = range(poly_grid$balance),
                  ylim = range(poly_pred$eta)) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Credit card balance ($)", y = expression(hat(eta)),
       color = "Student", linetype = "Student", title = "ISLR2::Default")

poly_prob_base <- ggplot(mapping = aes(x = balance, y = p_hat,
                                       color = student, linetype = student)) +
  geom_point(data = Default, aes(y = default_jit, shape = student),
             alpha = 0.08, size = 0.7) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  coord_cartesian(xlim = range(poly_grid$balance), ylim = c(-0.05, 1.05)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Student", linetype = "Student", shape = "Student",
       title = "ISLR2::Default")

poly_curves <- function(base, nm) {
  base + geom_line(data = filter(poly_pred, Model == nm), linewidth = 0.9)
}

poly_link_lin_add  <- poly_curves(poly_link_base, "Additive (linear)")
poly_link_lin_int  <- poly_curves(poly_link_base, "Interaction (linear)")
poly_link_quad_add <- poly_curves(poly_link_base, "Additive (quadratic)")
poly_link_quad_int <- poly_curves(poly_link_base, "Interaction (quadratic)")

poly_prob_data     <- poly_prob_base
poly_prob_lin_add  <- poly_curves(poly_prob_base, "Additive (linear)")
poly_prob_lin_int  <- poly_curves(poly_prob_base, "Interaction (linear)")
poly_prob_quad_add <- poly_curves(poly_prob_base, "Additive (quadratic)")
poly_prob_quad_int <- poly_curves(poly_prob_base, "Interaction (quadratic)")
```
:::

::: {.panel-tabset}
#### Additive (linear)
```{r poly-link-lin-add-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
poly_link_lin_add
```

#### Interaction (linear)
```{r poly-link-lin-int-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
poly_link_lin_int
```

#### Additive (quadratic)
```{r poly-link-quad-add-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
poly_link_quad_add
```

#### Interaction (quadratic)
```{r poly-link-quad-int-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
poly_link_quad_int
```
:::

The two linear tabs are the parallel and near-parallel lines of the
continuous-categorical subsection. The quadratic tabs add curvature, but too
little of it to see: over
the observed range of balance both still look like straight lines. What does
differ between them is the separation between the two groups — constant in
the additive tab, where one shared shape is shifted by $\hat\beta_3$, and
widening toward both ends of the balance range in the interaction tab, where
the student curve takes its own linear and quadratic coefficients.

The same four fits on the probability scale, against the raw outcomes:

::: {.panel-tabset}
#### Data
```{r poly-prob-data-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
poly_prob_data
```

#### + Additive (linear)
```{r poly-prob-lin-add-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
poly_prob_lin_add
```

#### + Interaction (linear)
```{r poly-prob-lin-int-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
poly_prob_lin_int
```

#### + Additive (quadratic)
```{r poly-prob-quad-add-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
poly_prob_quad_add
```

#### + Interaction (quadratic)
```{r poly-prob-quad-int-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
poly_prob_quad_int
```
:::

Logistic compression makes all four tabs nearly the same picture: two S-curves
of all but identical shape with the student curve shifted right. What
separated the quadratic tabs on the log-odds scale happens at balances where
the fitted probability is already pinned near $0$ or $1$, so it leaves no
visible trace here.

Whether the quadratic model needs the interaction is a comparison of the two
quadratic fits, and the interaction contributes two coefficients — one per
polynomial column — so it takes a drop-in-deviance test rather than a $z$:

```{r poly-deviance}
anova(poly_additive, poly_interaction, test = "Chisq")
```

```{r poly-deviance-extract}
#| include: false
poly_dev <- anova(poly_additive, poly_interaction, test = "Chisq")
poly_G2  <- poly_dev$Deviance[2]
poly_p   <- poly_dev[["Pr(>Chi)"]][2]
```

Forcing both groups onto one quadratic shape costs
$G^2 = `r sprintf("%.3f", poly_G2)`$ of deviance on $`r poly_dev$Df[2]`$
degrees of freedom, $p = `r sprintf("%.3f", poly_p)`$. Adding curvature does
not manufacture an interaction where the data do not support one: the verdict
is the same as it was for the straight-line fits.

## Step functions

A step function is likewise just a set of indicator columns,
$C_j(X) = \mathrm{I}(c_j \le X < c_{j+1})$ for $j = 1, \ldots, p$, built from
cut points $-\infty = c_0 < c_1 < \cdots < c_p < c_{p+1} = \infty$ exactly as
in the flexibility lecture, with $[c_0, c_1)$ the reference bin and `cut(x,
breaks)` in place of `poly(x, degree)`. Coefficients are read the same way:
$\beta_j$ is the log-odds difference between bin $j$ and the reference bin, so
$e^{\beta_j}$ is the odds ratio between those two bins — where in linear
regression $\beta_j$ was a difference between bin means.

More bins mean a more flexible fit, at the same bias-variance cost flexibility
always carries: with enough bins the step function tracks the training data
closely, at the price of a noisier estimate in any bin with few observations.

Paired with an indicator $D$, the bins offer the same choice between an
additive model and an interaction the previous two sections did:

$$\text{additive:} \qquad \eta = \beta_0 + \sum_{j=1}^{p} \beta_j C_j(X) + \gamma D,$$

$$\text{interaction:} \qquad \eta = \beta_0 + \sum_{j=1}^{p} \beta_j C_j(X) + \gamma D + \sum_{j=1}^{p} \delta_j C_j(X) D,$$

where

- $C_j(X)$ is the indicator for bin $j$, with bin $0$ the reference,
- $\gamma$ is the log-odds difference between the two groups in the reference
bin, and
- $\delta_j$ is how that difference in bin $j$ departs from $\gamma$.

The additive model spends $p + 2$ parameters and makes every bin's
group difference the same $\gamma$, so the two groups' staircases have
identical step heights and are separated by $\gamma$ everywhere. The
interaction model spends $2(p + 1)$ and gives the two groups a free
log-odds value in every bin, one parameter per bin-by-group cell — a
categorical-categorical interaction with $p + 1$ bins in place of two
categories.

::: {.callout-note collapse="true"}
### Code
```{r step-setup}
#| output: false
step_grid <- expand_grid(
  balance = seq(0, max(Default$balance), length.out = 400),
  student = factor(c("No", "Yes"), levels = levels(Default$student)))

# The same $400-wide-bin, by-student grouping used to check the additive fit
# in the previous lecture, so the observed data look the same way here as
# they did there. default01 was added to Default in the Interactions section
# above.
balance_bin_width <- 400
balance_binned <- Default |>
  mutate(bin = cut(balance, breaks = seq(0, 2800, by = balance_bin_width),
                   include.lowest = TRUE)) |>
  group_by(student, bin) |>
  summarize(balance = mean(balance), rate = mean(default01), n = n(),
            .groups = "drop") |>
  filter(n >= 100)

step_base <- ggplot(mapping = aes(x = balance)) +
  geom_point(data = balance_binned,
             aes(y = rate, color = student, shape = student, size = n)) +
  coord_cartesian(xlim = range(step_grid$balance), ylim = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  scale_size_area(max_size = 5) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Student", shape = "Student", size = "Customers\nin bin",
       title = "ISLR2::Default")

step_curve <- function(breaks) {
  fit <- glm(default ~ cut(balance, breaks = breaks, include.lowest = TRUE) *
               student,
             data = Default, family = "binomial")
  step_grid |> mutate(p_hat = predict(fit, newdata = step_grid,
                                      type = "response"))
}

step_line <- function(breaks) {
  list(geom_line(data = step_curve(breaks),
                 aes(y = p_hat, color = student, linetype = student),
                 linewidth = 0.9),
       labs(linetype = "Student"))
}

# Explicit, unequal-width breaks rather than cut(balance, breaks = k): every
# bin now has to hold both outcomes within *each* student group, not just
# overall, and equal-width bins leave the lowest bin with no student defaults
# at all once k reaches 4 (an empty cell is a preview of the separation the
# next lecture covers). Predicting from a re-fit cut() on new data is also
# fragile unless the same breaks are passed to both the fit and the grid.
step_data <- step_base
step_2bin <- step_base + step_line(c(0, 1350, 2700))
step_4bin <- step_base + step_line(c(0, 1100, 1500, 1900, 2700))
step_8bin <- step_base +
  step_line(c(0, 1100, 1300, 1500, 1700, 1900, 2100, 2300, 2700))
```
:::

Each fit below is the interaction model, `cut(balance, breaks) * student`,
giving the two groups separate staircases over the same bins as the bin count
doubles from two to eight; the observed points are the default
rate in the same \$400-wide bins the previous lecture used, split by student.

::: {.panel-tabset}
#### Data
```{r step-data-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
step_data
```

#### + 2 bins
```{r step-2bin-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
step_2bin
```

#### + 4 bins
```{r step-4bin-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
step_4bin
```

#### + 8 bins
```{r step-8bin-plot}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
step_8bin
```
:::

```{r step-extract}
#| include: false
step_top <- Default |>
  filter(balance > 2300) |>
  count(student)
```

Two bins give each group one flat stretch and one jump, barely distinguishing
low balances from high; each doubling adds detail the staircases can track,
until eight bins trace nearly the same rise the smooth logistic curve of the
earlier lectures did. Interacting the bins with `student` gives each group its
own height in every bin, so the two staircases run together at low balances,
where almost nobody defaults, and separate above roughly \$1,500, the student
staircase sitting below the non-student one over most of that range.

The price of that detail is visible in the eight-bin tab, where the staircases
stop rising monotonically, cross each other, and jump around: only
`r step_top$n[step_top$student == "No"]` non-students and
`r step_top$n[step_top$student == "Yes"]` students carry a balance above
\$2,300, so the top bin's two estimates are each built on a handful of
customers. Every added bin is another pair of coefficients to estimate, and
the data have to be spread across all of them.

## Examples

Two examples follow, asking different kinds of question of the same machinery.
The first is an inference question: is one feature's effect the same in two
groups? Its answer is a statement about a parameter, and a test delivers it.
The second is a prediction question: what probability should a decision use at
a feature combination the model is about to be asked about? Its answer is a
number the fitted model produces, and the choice between additive and
interaction changes that number.

### Student-specific balance effects

**Question.** The previous lecture fit `default ~ balance + student`
additively, which forces the per-dollar balance odds ratio to be identical for
students and non-students. Is that assumption right? The quantity of interest
is the parameter $\beta_3$ in
$\eta = \beta_0 + \beta_1 X + \beta_2 D + \beta_3 XD$, and the analysis is
finished when there is a verdict on whether it is zero — no customer's
predicted probability is being asked for.

**Data.** The same `r format(nrow(Default), big.mark = ",")` simulated credit
card customers as in the previous two lectures, with `balance` as the
quantitative feature $X$, the student indicator
$D = \mathrm{I}(\texttt{student} = \texttt{Yes})$, and $Y_i = 1$ when customer
$i$ defaults. The raw outcomes, and both fitted models on both scales, are the
figures of the continuous-categorical subsection above.

The interaction fit gives each group its own balance slope, and so its own
per-\$100 odds ratio:

::: {.callout-note collapse="true"}
### Code
```{r default-inference}
#| output: false
dcoef <- summary(default_interaction)$coefficients
z_d   <- dcoef["balance:studentYes", "z value"]
p_d   <- dcoef["balance:studentYes", "Pr(>|z|)"]
ci_d  <- confint.default(default_interaction)["balance:studentYes", ]

or_d <- tibble(
  student = c("No (reference)", "Yes"),
  slope   = c(bdi[2], bdi[2] + bdi[4]),
  or100   = exp(100 * c(bdi[2], bdi[2] + bdi[4]))
)

or_d_table <- or_d |>
  mutate(slope = sprintf("%.6f", slope),
         or100 = sprintf("%.3f", or100)) |>
  knitr::kable(col.names = c("Student", "Slope (log-odds per \\$1)",
                             "Odds ratio per \\$100"),
               align = "lcc")
```
:::

```{r default-or-table}
#| echo: false
or_d_table
```

The two per-\$100 odds ratios differ by
`r sprintf("%.2f", abs(diff(or_d$or100)))`, the difference the interaction was
added to capture, and it is $\hat\beta_3$ that carries it. That estimate has
$z = `r sprintf("%.2f", z_d)`$ with a $p$-value of $`r sprintf("%.3f", p_d)`$,
and a 95% Wald interval running from $`r sprintf("%.5f", ci_d[1])`$ to
$`r sprintf("%.5f", ci_d[2])`$, comfortably containing $0$; the
drop-in-deviance test of the same hypothesis reached the same verdict,
$G^2 = `r sprintf("%.3f", dG2)`$ on $`r ddev$Df[2]`$ degree of freedom with
$p = `r sprintf("%.3f", dp)`$.

::: {.callout-note collapse="true"}
### Beyond this course
The **Akaike information criterion**, $\text{AIC} = -2\ell(\hat\beta) + 2k$ for
a model with $k$ estimated parameters, scores a fit by its log-likelihood with
a penalty for the number of parameters it used; lower is better. It gives a
second, test-free reading of whether a term earns its place.

```{r default-aic-fit}
#| output: false
aic_d_table <- tibble(
  Model = c("`default ~ balance + student`", "`default ~ balance * student`"),
  k     = c(length(bda), length(bdi)),
  AIC   = sprintf("%.2f", c(AIC(default_additive), AIC(default_interaction)))
) |>
  knitr::kable(col.names = c("Model", "Parameters", "AIC"), align = "lcc")
```

```{r default-aic-table}
#| echo: false
aic_d_table
```

Adding the interaction makes AIC *worse* here: the extra parameter buys less
log-likelihood than it costs.
:::

**Conclusion.** Balance's effect on the odds of default does not depend on
student status in these data. The interval for $\beta_3$ contains zero, both
tests decline to reject, and the two fitted models are visually
indistinguishable on both scales, so the additive model of the previous
lecture is the one to report. Not every plausible interaction is a real one;
fitting it was still the right move, because the additive assumption was
checked rather than assumed.

### Store-specific price effects

**Question.** Citrus Hill and Minute Maid are competing orange juice brands,
sold at several different stores. A store setting next week's Citrus Hill
price wants a number: at \$1.99, what fraction of its customers will buy
Citrus Hill rather than Minute Maid? A model with one price effect shared by
every store and a model with a price effect per store both produce one, and
the analysis is finished only once one of the two has been chosen.

**Data.** `ISLR2::OJ` records `r format(nrow(OJ), big.mark = ",")` individual
orange juice purchases. These are observed purchases, not simulated ones.
Three of its `r ncol(OJ)` variables are used here:

- `Purchase`: which brand the customer bought, a factor with levels
`r paste(levels(OJ$Purchase), collapse = " and ")` (Citrus Hill and Minute
Maid). This is the response.
- `SalePriceCH`: the price actually paid for Citrus Hill, in dollars, ranging
from \$`r sprintf("%.2f", min(OJ$SalePriceCH))` to
\$`r sprintf("%.2f", max(OJ$SalePriceCH))`. This is the quantitative feature
$X$.
- `StoreID`: which of `r length(unique(OJ$StoreID))` stores the purchase
happened at, with `r paste(sort(unique(OJ$StoreID)), collapse = ", ")` as the
store identifiers. This is a categorical feature with more than two levels, so
no single $z$-test can address its interaction with price.

Buying Citrus Hill is the event of interest, so $Y_i = 1$ when purchase $i$ is
a Citrus Hill purchase. `glm()` models the probability of a factor's *second*
level, and `r levels(OJ$Purchase)[2]` is second, so the event has to be set
explicitly with `relevel()`:

```{r oj-relevel}
OJ <- OJ |> mutate(Purchase = relevel(Purchase, ref = "MM"))
levels(OJ$Purchase)
```

With `MM` as the reference level, every model below estimates
$P(\texttt{Purchase} = \texttt{CH})$.

Before fitting anything, plot each purchase against the price paid for
Citrus Hill, split by store:

::: {.callout-note collapse="true"}
### Code
```{r oj-eda}
#| output: false
OJ <- OJ |> mutate(ch01 = as.numeric(Purchase == "CH"))

oj_rates <- OJ |>
  group_by(StoreID, SalePriceCH) |>
  summarize(n = n(), rate = mean(ch01), .groups = "drop")

oj_eda_plot <- ggplot(OJ, aes(x = SalePriceCH, y = ch01)) +
  geom_jitter(height = 0.05, width = 0.008, alpha = 0.15, size = 0.8) +
  geom_point(data = oj_rates, aes(y = rate, size = n),
             shape = 1, color = "#D55E00") +
  facet_wrap(~ StoreID, labeller = label_both) +
  scale_y_continuous(breaks = c(0, 0.5, 1),
                     labels = c("MM (0)", "0.5", "CH (1)")) +
  scale_size_area(max_size = 5) +
  labs(x = "Citrus Hill sale price ($)", y = "Purchase",
       size = "Purchases\nat that price", title = "ISLR2::OJ")

store_cors <- OJ |>
  group_by(StoreID) |>
  summarize(cor = cor(SalePriceCH, ch01), .groups = "drop")
```
:::

```{r oj-eda-figure}
#| echo: false
#| fig-width: 7.5
#| fig-height: 4.5
oj_eda_plot
```

The open circles are the observed proportion of Citrus Hill purchases at each
distinct price, sized by how many purchases were made at that price. Even
before any model is fit, the raw correlation between price and purchase
already changes sign from store to store: it runs from
`r sprintf("%.2f", min(store_cors$cor))` at store
`r store_cors$StoreID[which.min(store_cors$cor)]` (higher prices, fewer Citrus
Hill purchases, as expected) to
`r sprintf("%.2f", max(store_cors$cor))` at store
`r store_cors$StoreID[which.max(store_cors$cor)]` (the opposite).

Fitting `SalePriceCH + factor(StoreID)` and `SalePriceCH * factor(StoreID)`
puts both candidate models on the probability scale a price decision uses,
with the observed proportions overlaid:

::: {.callout-note collapse="true"}
### Code
```{r oj-prob-setup}
#| output: false
oj_additive    <- glm(Purchase ~ SalePriceCH + factor(StoreID),
                      data = OJ, family = "binomial")
oj_interaction <- glm(Purchase ~ SalePriceCH * factor(StoreID),
                      data = OJ, family = "binomial")

boa <- coef(oj_additive)
boi <- coef(oj_interaction)
store_ids <- sort(unique(OJ$StoreID))

# Each store's own observed price range, not the pooled range -- stores 3 and
# 4 were never sampled below $1.79, so a pooled grid would draw their fitted
# curves across prices those stores never charged.
oj_grid <- OJ |>
  group_by(StoreID) |>
  reframe(SalePriceCH = seq(min(SalePriceCH), max(SalePriceCH), length.out = 100))

oj_prob <- bind_rows(
  oj_grid |> mutate(Model = "Additive",
                    p_hat = predict(oj_additive, newdata = oj_grid,
                                    type = "response")),
  oj_grid |> mutate(Model = "Interaction",
                    p_hat = predict(oj_interaction, newdata = oj_grid,
                                    type = "response")))

store_colors <- c("#E69F00", "#56B4E9", "#009E73", "#0072B2", "#D55E00")

# The same shared-base idea, with the observed proportions as the layer every
# tab keeps.
oj_prob_base <- ggplot(mapping = aes(x = SalePriceCH, y = p_hat,
                                     color = factor(StoreID),
                                     linetype = factor(StoreID))) +
  geom_hline(yintercept = 0.5, color = "grey60", linewidth = 0.3) +
  geom_point(data = oj_rates, aes(y = rate, size = n, shape = factor(StoreID)),
             alpha = 0.8) +
  coord_cartesian(xlim = range(oj_grid$SalePriceCH), ylim = c(0, 1)) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_size_area(max_size = 5) +
  scale_color_manual(values = store_colors) +
  labs(x = "Citrus Hill sale price ($)", y = "P(Purchase = CH)",
       color = "Store", linetype = "Store", shape = "Store",
       size = "Purchases\nat that price", title = "ISLR2::OJ")

oj_prob_data <- oj_prob_base

oj_prob_additive <- oj_prob_base +
  geom_line(data = filter(oj_prob, Model == "Additive"), linewidth = 0.9)

oj_prob_interaction <- oj_prob_base +
  geom_line(data = filter(oj_prob, Model == "Interaction"), linewidth = 0.9)
```
:::

::: {.panel-tabset}
#### Data
```{r oj-prob-data-plot}
#| echo: false
#| fig-width: 7.5
#| fig-height: 3.2
oj_prob_data
```

#### + Additive
```{r oj-prob-additive-plot}
#| echo: false
#| fig-width: 7.5
#| fig-height: 3.2
oj_prob_additive
```

#### + Interaction
```{r oj-prob-interaction-plot}
#| echo: false
#| fig-width: 7.5
#| fig-height: 3.2
oj_prob_interaction
```
:::

The data tab shows the observed proportion of Citrus Hill purchases at each
price, by store. The additive tab fits them with `r length(unique(OJ$StoreID))`
curves of identical shape that never cross, all drifting down with price. The
interaction tab lets each store's curve rise or fall at its own rate, and
several now cross within the observed price range. Which of the two a store
should predict from is the question the interaction test answers:

```{r oj-interaction-summary}
oj_dev <- anova(oj_additive, oj_interaction, test = "Chisq")
oj_dev
```

```{r oj-inference}
#| include: false
oj_G2 <- oj_dev$Deviance[2]
oj_df <- oj_dev$Df[2]
oj_p  <- oj_dev[["Pr(>Chi)"]][2]
```

With `r length(store_ids)` stores the interaction contributes
`r length(store_ids) - 1` coefficients, one per non-reference store. Dropping
all `r oj_df` of them at once would raise the deviance by
$G^2 = `r sprintf("%.2f", oj_G2)`$ on $`r oj_df`$ degrees of freedom, giving
$p = `r signif(oj_p, 3)`$: price's effect on the odds of a Citrus Hill
purchase is not the same across stores, and the store-specific model is the
one to predict from.

Reading each store's slope off that model, and converting it to an odds ratio,
makes the split concrete:

::: {.callout-note collapse="true"}
### Code
```{r oj-slope-setup}
#| output: false
store_slope <- function(s) {
  if (s == store_ids[1]) return(unname(boi["SalePriceCH"]))
  term <- paste0("SalePriceCH:factor(StoreID)", s)
  unname(boi["SalePriceCH"] + boi[term])
}
store_slopes <- sapply(store_ids, store_slope)

oj_slope_vcov <- vcov(oj_interaction)
store_slope_se <- function(s) {
  if (s == store_ids[1]) {
    return(sqrt(oj_slope_vcov["SalePriceCH", "SalePriceCH"]))
  }
  term <- paste0("SalePriceCH:factor(StoreID)", s)
  sqrt(oj_slope_vcov["SalePriceCH", "SalePriceCH"] + oj_slope_vcov[term, term] +
       2 * oj_slope_vcov["SalePriceCH", term])
}
store_ses <- sapply(store_ids, store_slope_se)
store_z   <- store_slopes / store_ses
store_p   <- 2 * pnorm(-abs(store_z))
n_decisive     <- sum(store_p < 0.05)
decisive_store <- store_ids[which.max(abs(store_z))]

oj_slope_table <- tibble(
  store = store_ids,
  slope = store_slopes,
  or10  = exp(0.10 * store_slopes)
) |>
  mutate(slope = sprintf("%.3f", slope),
         or10  = sprintf("%.3f", or10)) |>
  knitr::kable(col.names = c("Store", "Slope (log-odds per \\$1)",
                             "Odds ratio per \\$0.10"),
               align = "ccc")

n_negative <- sum(store_slopes < 0)
n_positive <- sum(store_slopes > 0)
```
:::

```{r oj-slope-table}
#| echo: false
oj_slope_table
```

The fitted slopes run downward, the direction a price increase is expected to
push demand, at `r n_negative` of the `r length(store_ids)` stores; the other
`r n_positive` run backwards, the same \$0.10 *raising* the estimated odds.
Only store `r decisive_store`'s slope is individually far enough from zero to
rule out sampling error on its own ($z = `r sprintf("%.2f", store_z[store_ids == decisive_store])`$,
$p = `r signif(store_p[store_ids == decisive_store], 3)`$); the other
`r length(store_ids) - n_decisive` point estimates are not — the
drop-in-deviance test above is what supports "not all five slopes are the
same," not any one store's estimate by itself. What these data cannot say is
why the slopes that do run backwards do
so — `SalePriceCH` records what was paid, not what else varied alongside it
across stores and weeks, so the association it carries need not be the effect
of price by itself.

Evaluating both models at \$1.99, a price all `r length(store_ids)` stores
actually charged, turns those slopes into the number the question asked for:

::: {.callout-note collapse="true"}
### Code
```{r oj-prediction}
#| output: false
oj_price <- 1.99
oj_pred_grid <- tibble(StoreID = store_ids, SalePriceCH = oj_price)

oj_pred <- oj_pred_grid |>
  mutate(additive    = predict(oj_additive, newdata = oj_pred_grid,
                               type = "response"),
         interaction = predict(oj_interaction, newdata = oj_pred_grid,
                               type = "response"))

oj_pred_gap <- max(abs(oj_pred$interaction - oj_pred$additive))

oj_pred_table <- oj_pred |>
  mutate(across(c(additive, interaction), \(x) sprintf("%.3f", x))) |>
  select(StoreID, additive, interaction) |>
  knitr::kable(col.names = c("Store", "Additive fit", "Interaction fit"),
               align = "ccc")
```
:::

```{r oj-pred-table}
#| echo: false
oj_pred_table
```

The two models disagree by as much as
`r sprintf("%.2f", oj_pred_gap)` in predicted probability at the same price
and the same store — the difference between expecting a minority of customers
to choose Citrus Hill and expecting a majority. Had the interaction test
failed to reject, the two columns would have sat close together and the choice
between the models would not have changed the number a store acts on.

A second number a price decision uses is the **break-even price**: the Citrus
Hill price at which the model puts the two brands at 50/50, computed
separately for each store. It is that store's **decision boundary** for a
$0.5$ threshold — where its fitted log-odds crosses zero, $\hat\eta = 0$,
since $\hat p = 0.5$ exactly there.

::: {.callout-note collapse="true"}
### Code
```{r oj-boundaries}
#| output: false
store_intercept <- function(s) {
  if (s == store_ids[1]) return(unname(boi["(Intercept)"]))
  term <- paste0("factor(StoreID)", s)
  unname(boi["(Intercept)"] + boi[term])
}
store_intercepts <- sapply(store_ids, store_intercept)
store_boundaries <- -store_intercepts / store_slopes

# Each store's own observed price range, not the pooled range -- store 7's
# break-even price, for instance, is inside the range OJ$SalePriceCH covers
# overall but above the highest price ever charged at store 7 specifically.
store_ranges <- OJ |>
  group_by(StoreID) |>
  summarize(lo = min(SalePriceCH), hi = max(SalePriceCH), .groups = "drop")

oj_boundary_table <- tibble(
  store = store_ids,
  price = store_boundaries
) |>
  left_join(store_ranges, by = c("store" = "StoreID")) |>
  mutate(observed = ifelse(price >= lo & price <= hi, "yes", "no"),
         price = sprintf("\\$%.2f", price)) |>
  select(store, price, observed) |>
  knitr::kable(col.names = c("Store", "Price where $\\hat p = 0.5$",
                             "In observed range?"),
               align = "ccc")

n_outside <- sum(store_boundaries < store_ranges$lo | store_boundaries > store_ranges$hi)
```
:::

```{r oj-boundary-table}
#| echo: false
oj_boundary_table
```

`r n_outside` of the `r length(store_ids)` break-even prices fall outside the
range of prices that store itself charged — a store whose own data never
approached 50/50 can only place its break-even point by extrapolating past
what was observed there, which is a property of the fitted line, not evidence
about prices that store never charged.

::: {.callout-note collapse="true"}
### Beyond this course
Giving each store its own fixed intercept and slope, as the interaction model
above does, spends one parameter per store and says nothing about a store that
was not in the sample. A **random effects** (or **mixed-effects**) logistic
regression instead treats the stores actually observed as a sample from a
population of stores, and estimates the *distribution* of intercepts and
slopes across that population rather than a separate fixed value for each one.

A **random intercept** model lets each store's own intercept vary around a
shared mean, $b_{0,\text{store}} \sim N(0, \sigma_0^2)$, while every store
keeps the same price slope — the mixed-effects analogue of the additive model
above. A **random slope** model goes further and lets the price slope itself
vary by store, $b_{1,\text{store}} \sim N(0, \sigma_1^2)$, typically allowed to
correlate with $b_{0,\text{store}}$ — the analogue of the interaction model,
but estimated with two variances and a correlation instead of one fixed
coefficient per store, which matters far more once there are dozens of stores
instead of five. In R, `lme4::glmer()` fits both:

```{r glmer-syntax}
#| eval: false
lme4::glmer(Purchase ~ SalePriceCH + (1 | StoreID),
            data = OJ, family = "binomial")          # random intercept
lme4::glmer(Purchase ~ SalePriceCH + (SalePriceCH | StoreID),
            data = OJ, family = "binomial")          # random intercept and slope
```
:::

**Conclusion.** A store predicting Citrus Hill purchases at a candidate price
should predict from the store-specific model: the drop-in-deviance test
rejects a common price slope decisively ($p = `r signif(oj_p, 3)`$), and the
two models' predictions at \$1.99 differ by up to
`r sprintf("%.2f", oj_pred_gap)`. `r n_negative` of the
`r length(store_ids)` stores show the expected pattern of higher prices
costing Citrus Hill sales and `r n_positive` run the opposite way, which these
data describe without explaining; a prediction for a price far outside a
store's own observed range, its break-even price included, is an
extrapolation of the fitted curve rather than evidence from that store.

### Two questions, two verdicts

Both examples fit the same pair of models — one feature's effect held common
across groups, then allowed to vary — and both settled the choice with a
drop-in-deviance test. What differed is what hung on the answer. For `Default`
the answer *was* the result: $\beta_3$ is the scientific quantity, and it is
indistinguishable from zero, so the additive model of the previous lecture
stands. For `OJ` the answer was a means to a number: the test rejects a common
price slope, and the consequence is that a store's predicted purchase
probability at a candidate price moves by as much as
`r sprintf("%.2f", oj_pred_gap)` depending on which model supplies it.

Nothing about either pair of features announced which answer was coming. The
additive assumption is checkable either way, and the reason to check it
differs with the question: when the interest is understanding, the assumption
*is* the question; when the interest is prediction, the assumption decides
what number comes out.

## Conclusion

Interactions, polynomials, and step functions are all choices of basis
function entering the linear predictor as their own column, estimated by the
same Bernoulli maximum likelihood as every other logistic regression, with
every coefficient still read as a change in log-odds. An interaction lets one
feature's effect depend on another's: categorical-categorical, adding an
extra log-odds for the combination of two non-reference levels;
continuous-categorical, giving each group its own log-odds slope and so its
own odds ratio; continuous-continuous, letting one feature's slope vary
continuously with the other's value. A coefficient's raw magnitude never
settles whether an interaction matters, since it depends on arbitrary units —
only a $z$-test, for one coefficient, or a drop-in-deviance test, for several
at once, does.

The diagnostic for the continuous-categorical case is a picture on the
linear-predictor scale, where both fits are straight lines: if the interaction
fit's two lines are all but parallel, the additive model already captures each
group's log-odds slope; if they visibly cross or diverge, the slope changes
from one group to the next. Each group's decision boundary is where its own
fitted line crosses zero, so an interaction moves the two boundaries
independently rather than by a common shift.

Polynomials and step functions raise no new estimation or interpretation
issues once basis functions are in view — only the question of how much
flexibility the data can support, the same bias-variance tradeoff the
flexibility lecture raised for a quantitative response. Adding columns to a
binary-response model can also break the estimation itself in a way that least
squares never breaks, a failure that goes by the name separation — the next
lecture's subject.
