# Problems in Logistic Regression

```{r packages}
#| include: false
library("tidyverse")
theme_set(theme_bw())

# Renders a number in LaTeX scientific notation, so that a value like
# 7.9e-10 appears as 7.9 x 10^{-10} in math mode rather than as "7.9e-10".
sci <- function(x, digits = 2) {
  e <- floor(log10(abs(x)))
  m <- signif(x / 10^e, digits)
  sprintf("%s \\times 10^{%d}", format(m), e)
}
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **define** complete and quasi-complete separation as conditions on the data,
stated through the existence of a coefficient vector whose linear predictor
classifies every observation correctly,
- **derive**, from the Bernoulli log-likelihood, why separation leaves $\ell$
with a supremum but no finite maximizer,
- **diagnose** an enormous coefficient paired with a still larger standard
error, alongside a `glm.fit` warning, as separation rather than as evidence
about the feature,
- **evaluate** the responses available when separation occurs — a larger
sample, a simpler model, or a fit still usable for prediction away from the
separated region — and **distinguish** separation from a warning that a single
extreme observation caused, and
- **compare** the potential problems of linear regression against their
logistic regression counterparts and **identify** which do not transfer.
:::

## A fit that should not work

Every logistic regression in the previous three lectures converged, returned
standard errors of a believable size, and produced $z$-statistics that could be
read against a normal distribution. Nothing in those outputs suggested that
maximum likelihood might do anything else.

It can. Least squares always returns an estimate as long as $X^\top X$ is
invertible, because $\hat\beta = (X^\top X)^{-1}X^\top y$ is a formula. Maximum
likelihood for logistic regression has no formula: $\hat\beta$ is defined as
the maximizer of $\ell(\beta)$, and a function need not have one.

So ask what happens at the opposite extreme from a useless feature: one that
predicts the response perfectly.

### Example: fitting a feature that predicts the outcome perfectly

**Question.** Suppose a single feature sorts the two classes without a single
mistake. What does `glm()` estimate?

**Data.** Ten observations, constructed rather than collected, because the
point is about a data-generating arrangement rather than about any real
measurement:

```{r toy-data}
x <- 1:10
y <- as.numeric(x > 5)
toy <- tibble(x = x, y = y, class = factor(y))
```

- $x_i$ takes the values $1, 2, \ldots, `r max(x)`$, and
- $y_i = 1$ exactly when $x_i > 5$, so the five smallest feature values carry
$y = 0$ and the five largest carry $y = 1$.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r toy-eda}
#| output: false
toy_eda_plot <- ggplot(toy, aes(x = x, y = y, color = class, shape = class)) +
  geom_vline(xintercept = 5.5, linetype = "dashed", color = "grey50") +
  geom_point(size = 3) +
  scale_x_continuous(breaks = x) +
  scale_y_continuous(breaks = c(0, 1)) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "x", y = "y", color = "y", shape = "y",
       title = "A perfectly predictive feature")
```
:::

```{r toy-eda-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 2.6
toy_eda_plot
```

Every $y = 0$ lies to the left of the dashed line and every $y = 1$ lies to the
right, with a gap of one unit between the last $0$ and the first $1$. No
observation contradicts the rule "predict $1$ when $x > 5.5$," and the dashed
line could be moved anywhere in $(5, 6)$ without changing that.

Now fit the model.

```{r toy-glm}
toy_glm <- glm(y ~ x, data = toy, family = "binomial")
summary(toy_glm)
```

**The two warnings are the output of the chunk above, not a defect in it.**
This lecture is about the condition that produces them, so they are left
visible here and in the second example below rather than suppressed;
recognizing the exact strings `glm.fit: algorithm did not converge` and
`glm.fit: fitted probabilities numerically 0 or 1 occurred` is part of what is
being taught.

::: {.callout-note collapse="true"}
### Code for the values quoted below
```{r toy-extract}
#| output: false
toy_coef <- summary(toy_glm)$coefficients
toy_b    <- toy_coef[, "Estimate"]
toy_se   <- toy_coef[, "Std. Error"]
toy_z    <- toy_coef[, "z value"]
toy_p    <- toy_coef[, "Pr(>|z|)"]
toy_ratio <- toy_se / abs(toy_b)
```
:::

Read the coefficient table one column at a time.

- The estimates are $\hat\beta_0 = `r sprintf("%.1f", toy_b[1])`$ and
$\hat\beta_1 = `r sprintf("%.1f", toy_b[2])`$. A slope of
$`r sprintf("%.1f", toy_b[2])`$ log-odds per unit of $x$ is an odds ratio of
$e^{`r sprintf("%.1f", toy_b[2])`}$, a number with
`r floor(toy_b[2] / log(10)) + 1` digits before the decimal point.
- The standard errors are larger still:
`r format(round(toy_se[1]), big.mark = ",")` and
`r format(round(toy_se[2]), big.mark = ",")`, each roughly
`r format(signif(mean(toy_ratio), 2), big.mark = ",")` times the coefficient it
belongs to.
- So $z = `r sprintf("%.3f", toy_z[2])`$ for the slope, with a $p$-value of
$`r sprintf("%.3f", toy_p[2])`$. **The feature that predicts the response
without error is not statistically significant.**

Two more lines of the output say the same thing in different ways. The number
of Fisher scoring iterations is `r toy_glm$iter`, which is `glm()`'s default
maximum, so the algorithm stopped because it ran out of iterations rather than
because it arrived anywhere, and `toy_glm$converged` is
`r toy_glm$converged`. The residual deviance is
$`r sci(deviance(toy_glm))`$, which is $-2\ell(\hat\beta)$: the log-likelihood
at the reported estimates is zero to within rounding, meaning the model
reproduces all `r nrow(toy)` observed responses exactly.

A model that fits the training data perfectly and reports no significant
coefficient is not a contradiction. It is the signature of the problem this
lecture is about.

## Separation

The arrangement in that example has a name, and a definition in terms of the
linear predictor rather than in terms of a picture.

Write $x_i$ for the $i$th row of the model matrix and $b$ for a candidate
coefficient vector, so $x_i b$ is the linear predictor that $b$ assigns to
observation $i$. The data exhibit **complete separation** (also called perfect
separation) if there exists a $b$ with

$$x_i b > 0 \ \text{ for every } i \text{ with } y_i = 1, \qquad x_i b < 0 \ \text{ for every } i \text{ with } y_i = 0,$$

where

- $x_i = (1, x_{i1}, \ldots, x_{ip})$ is the feature row for observation $i$,
- $b$ is any vector in $\mathbb{R}^{p+1}$, not necessarily the maximum
likelihood estimate, and
- $y_i \in \{0, 1\}$ is the observed response.

Note that this is a condition on the data, not on a fitted model: it asks
whether *some* $b$ works, and says nothing about $\hat\beta$, which under this
condition does not exist. Since the probability that $b$ implies exceeds $0.5$
exactly when $x_i b > 0$, the condition says the decision boundary of the
previous lecture, drawn using $b$, classifies every training observation
correctly, with no observation lying on the boundary itself.

The data exhibit **quasi-complete separation** if the strict inequalities can
be weakened to

$$x_i b \ge 0 \ \text{ for every } i \text{ with } y_i = 1, \qquad x_i b \le 0 \ \text{ for every } i \text{ with } y_i = 0,$$

for some $b \ne 0$, with equality holding for at least one observation. Some
observations sit exactly on the boundary; none sits on the wrong side of it.

Both conditions break maximum likelihood, and neither is visible in a
coefficient table without knowing what to look for.

::: {.callout-note collapse="true"}
### For example, one tie is enough

Complete separation needs a gap between the classes. Quasi-complete separation
does not: it needs only that no observation falls strictly on the wrong side.
Give two observations the same feature value and opposite responses, and leave
everything else separated:

```{r quasi-data}
xq <- c(1, 2, 3, 4, 5, 5, 6, 7, 8, 9)
yq <- c(0, 0, 0, 0, 0, 1, 1, 1, 1, 1)
```

With $b = (-5\beta_1, \beta_1)$ for any $\beta_1 > 0$, the two observations at
$x = 5$ get a linear predictor of exactly $0$ and every other observation gets
the sign its response requires, so this is quasi-complete separation.

```{r quasi-glm}
#| warning: false
quasi <- tibble(xq = xq, yq = yq)
quasi_glm <- glm(yq ~ xq, data = quasi, family = "binomial")
summary(quasi_glm)$coefficients
```

```{r quasi-extract}
#| output: false
quasi_coef <- summary(quasi_glm)$coefficients
```

The estimates are as unusable as in the completely separated case —
$\hat\beta_1 = `r sprintf("%.1f", quasi_coef[2, 1])`$ with a standard error of
`r format(round(quasi_coef[2, 2]), big.mark = ",")` — but this fit reports
`quasi_glm$converged` as `r quasi_glm$converged`, in
`r quasi_glm$iter` iterations. The only warning it raises is the
`fitted probabilities numerically 0 or 1` one, suppressed in this callout;
there is no convergence warning to go with it. A reported convergence is not
evidence that an estimate is trustworthy.

The two tied observations are why the residual deviance is
`r sprintf("%.2f", deviance(quasi_glm))` rather than zero: the fitted
probability at $x = 5$ is forced to $0.5$, contributing
$-2 \times 2 \log(0.5) = `r sprintf("%.2f", -4 * log(0.5))`$ no matter how large
$\beta_1$ grows.
:::

## Why maximum likelihood fails

The simple logistic regression lecture derived the Bernoulli log-likelihood

$$\ell(\beta_0, \beta_1) = \sum_{i=1}^n \left\{ y_i (\beta_0 + \beta_1 x_i) - \log\left(1 + e^{\beta_0 + \beta_1 x_i}\right) \right\},$$

and nothing beyond that is needed to see what goes wrong. With one feature,
$x_i$ is the scalar feature value again rather than the model-matrix row of the
definitions above.

Restrict attention to a one-parameter family of coefficient vectors that all
place the decision boundary at $x = 5.5$, by setting $\beta_0 = -5.5\beta_1$.
Then the linear predictor is $\eta_i = \beta_1(x_i - 5.5)$ and

$$\ell(\beta_1) = \sum_{i=1}^n \left\{ y_i \beta_1 (x_i - 5.5) - \log\left(1 + e^{\beta_1(x_i - 5.5)}\right) \right\},$$

where

- $\beta_1$ is the only free parameter, and
- $p_i = \text{logistic}(\beta_1(x_i - 5.5))$ is the probability the family
implies for observation $i$.

Differentiating with respect to $\beta_1$, by the same steps that produced the
score equations,

$$\frac{d\ell}{d\beta_1} = \sum_{i=1}^n (x_i - 5.5)\left(y_i - p_i\right).$$

Now use the separation. Every observation with $y_i = 1$ has $x_i > 5.5$, so
$x_i - 5.5 > 0$, and $y_i - p_i = 1 - p_i > 0$ because $p_i < 1$ always. Every
observation with $y_i = 0$ has $x_i < 5.5$, so $x_i - 5.5 < 0$, and
$y_i - p_i = -p_i < 0$. Both cases contribute a positive product, so

$$\frac{d\ell}{d\beta_1} > 0 \quad \text{for every finite } \beta_1.$$

The derivative is never zero, so there is no stationary point and no finite
maximizer. Meanwhile $\ell < 0$ for every finite $\beta_1$, and as
$\beta_1 \to \infty$ each $p_i$ approaches the $y_i$ it is paired with, so
every term approaches $\log 1 = 0$ and $\ell \to 0$. The log-likelihood has a
**supremum** of $0$ that it attains only in the limit.

What `glm()` reports is therefore not an estimate. It is whatever point the
iteration happened to be at when it stopped, and the reported standard errors
come from the curvature of $\ell$ at that point — curvature that is flattening
toward zero as the iteration runs, which is why they grow without bound.

Drawing the fitted curve at a few values of $\beta_1$ along this family shows
what the derivative is describing.

::: {.callout-note collapse="true"}
### Code for the figure below
```{r toy-curves}
#| output: false
toy_curve_betas <- c(0.5, 1, 2, 5)

toy_curves <- expand_grid(beta1 = toy_curve_betas,
                          xg = seq(0, 11, length.out = 400)) |>
  mutate(p = plogis(beta1 * (xg - 5.5)),
         beta1 = factor(sprintf("beta[1] == %.1f", beta1),
                        levels = sprintf("beta[1] == %.1f", toy_curve_betas)))

toy_curve_plot <- ggplot(toy_curves,
                         aes(x = xg, y = p, color = beta1, linetype = beta1)) +
  geom_line(linewidth = 0.9) +
  geom_point(data = toy, aes(x = x, y = y), inherit.aes = FALSE,
             size = 2, alpha = 0.7) +
  scale_color_manual(values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7"),
                     labels = scales::parse_format()) +
  scale_linetype_manual(values = c("solid", "dashed", "dotted", "dotdash"),
                        labels = scales::parse_format()) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  labs(x = "x", y = "p(x)", color = NULL, linetype = NULL,
       title = "Fitted curves along the family with boundary fixed at 5.5")
```
:::

```{r toy-curves-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
toy_curve_plot
```

Each increase in $\beta_1$ steepens the curve at the boundary and pushes it
closer to $0$ on the left and to $1$ on the right, so every fitted probability
moves toward the response it is paired with. Nothing stops this: the curve
approaches a step function and the likelihood approaches $1$.

Evaluating $\ell$ along the same family puts numbers on it.

::: {.callout-note collapse="true"}
### Code for the table and figure below
```{r toy-loglik}
#| output: false
toy_loglik <- function(beta1) {
  eta <- beta1 * (toy$x - 5.5)
  sum(toy$y * eta - log(1 + exp(eta)))
}

toy_ll_table <- tibble(beta1 = c(0, 0.5, 1, 2, 5, 8)) |>
  mutate(ll = map_dbl(beta1, toy_loglik),
         L  = exp(ll)) |>
  mutate(beta1 = sprintf("%.1f", beta1),
         ll    = sprintf("%.3f", ll),
         L     = sprintf("%.3f", L)) |>
  knitr::kable(col.names = c("$\\beta_1$", "$\\ell(\\beta_1)$",
                             "$L(\\beta_1)$"),
               align = "ccc")

toy_ll_curve <- tibble(beta1 = seq(0, 8, length.out = 400)) |>
  mutate(ll = map_dbl(beta1, toy_loglik))

toy_ll_plot <- ggplot(toy_ll_curve, aes(x = beta1, y = ll)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line(color = "#0072B2", linewidth = 0.9) +
  labs(x = expression(beta[1]), y = "Log-likelihood",
       title = expression(paste("Log-likelihood along the family ",
                                beta[0] == -5.5 * beta[1])))
```
:::

```{r toy-loglik-table}
#| echo: false
toy_ll_table
```

```{r toy-loglik-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3
toy_ll_plot
```

The log-likelihood rises steeply at first and then flattens against the dashed
line at $0$ without ever touching it, exactly as the derivative predicted. The
likelihood column reaches
`r sprintf("%.3f", exp(toy_loglik(8)))` at $\beta_1 = 8$ and would keep
climbing toward $1$ for larger values.

::: {.callout-note collapse="true"}
### Beyond this course
Two numerical artifacts surface when $\ell$ is evaluated by hand at large
$\beta_1$. Neither changes any conclusion above, and neither is tested here.

```{r toy-artifacts}
#| output: false
toy_loglik_prob <- function(beta1) {
  p <- plogis(beta1 * (toy$x - 5.5))
  sum(toy$y * log(p) + (1 - toy$y) * log(1 - p))
}

artifact_grid <- seq(0.1, 400, by = 0.1)
first_nan <- artifact_grid[which(is.na(map_dbl(artifact_grid,
                                               toy_loglik_prob)))[1]]
first_inf <- artifact_grid[which(is.infinite(map_dbl(artifact_grid,
                                                     toy_loglik)))[1]]
```

Evaluating $\ell$ on the probability scale as
$\sum_i \{y_i \log p_i + (1 - y_i)\log(1 - p_i)\}$ is algebraically identical
to the form used above, but in this example it returns `NaN` from
$\beta_1 = `r first_nan`$ onward. Once $p_i$ rounds to exactly $1$, the term
$(1 - y_i)\log(1 - p_i)$ becomes $0 \times \log 0 = 0 \times (-\infty)$, which
is `NaN`.

The $\eta$-scale form survives further but eventually fails too: at
$\beta_1 = `r first_inf`$ the largest $e^{\eta_i}$ overflows to `Inf`, and
$\ell$ returns $-\infty$.

Neither is a property of $\ell$, which is finite, negative, and increasing at
every one of those values. Both are finite-precision arithmetic running out of
room, which is also the mechanism behind the `fitted probabilities numerically
0 or 1` warning. The range $0 \le \beta_1 \le 8$ plotted above stays clear of
both.
:::

## Separation in real data

Constructed data makes the mechanism visible, and it also makes separation look
unmistakable. It is not always. Real data can trigger the same warning without
being separated at all, and telling the two cases apart takes one more step
after the warning appears.

### Example: screening for muscular dystrophy carriers

**Question.** Duchenne muscular dystrophy is inherited through female
carriers, who show no symptoms themselves. Can a blood test identify likely
carriers, so that a woman with a family history can be counseled about her
risk?

**Data.** `Sleuth3::ex2012` records
`r nrow(Sleuth3::ex2012)` women, each already classified by other means:

- `Group`: `Case` for a known carrier, `Control` otherwise. There are
`r sum(Sleuth3::ex2012$Group == "Case")` carriers and
`r sum(Sleuth3::ex2012$Group == "Control")` non-carriers. This is the response,
with $Y_i = 1$ for a carrier.
- `CK`: serum creatine kinase, an enzyme that leaks from damaged muscle.
This is the feature.
- `H`: serum hemopexin, a second blood measurement, recorded but not used
below.

`Case` is the first level of `Group`, and `glm()` models the probability of a
factor's *second* level, so the event has to be set explicitly:

```{r dmd-relevel}
dmd <- Sleuth3::ex2012 |>
  mutate(Group = relevel(Group, ref = "Control"),
         case01 = as.numeric(Group == "Case"))
levels(dmd$Group)
```

::: {.callout-note collapse="true"}
### Code for the figure below
```{r dmd-eda}
#| output: false
# The jitter is computed once and stored, rather than left to geom_jitter(),
# so re-rendering does not move the points.
set.seed(20260919)
dmd <- dmd |> mutate(case_jittered = case01 + runif(n(), -0.04, 0.04))

max_control_ck <- max(dmd$CK[dmd$Group == "Control"])
n_above        <- sum(dmd$CK > max_control_ck)
n_case_below   <- sum(dmd$Group == "Case" & dmd$CK <= max_control_ck)

dmd_eda_plot <- ggplot(dmd, aes(x = CK, y = case_jittered,
                                color = Group, shape = Group)) +
  geom_vline(xintercept = max_control_ck, linetype = "dashed",
             color = "grey50") +
  geom_point(size = 1.8, alpha = 0.8) +
  scale_y_continuous(breaks = c(0, 1),
                     labels = c("Control (0)", "Case (1)")) +
  scale_color_manual(values = c("#0072B2", "#D55E00")) +
  labs(x = "Serum creatine kinase (CK)", y = "Carrier status",
       title = "Sleuth3::ex2012")
```
:::

```{r dmd-eda-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3
dmd_eda_plot
```

The dashed line marks `r max_control_ck`, the largest `CK` recorded on any
non-carrier. All `r n_above` observations to its right are carriers. To its
left the two groups overlap heavily: `r n_case_below` of the
`r sum(dmd$Group == "Case")` carriers have `CK` values in the same range as the
non-carriers. Carriers have a median `CK` of
`r median(dmd$CK[dmd$Group == "Case"])` against
`r median(dmd$CK[dmd$Group == "Control"])` for non-carriers. No threshold on
`CK` classifies these women without error, so this is not separation under
either definition above.

Fit the model anyway.

```{r dmd-glm}
dmd_glm <- glm(Group ~ CK, data = dmd, family = "binomial")
summary(dmd_glm)$coefficients
```

The warning from the toy example has reappeared. The convergence warning has
not, and again this chunk's warning is intended output rather than a defect.

::: {.callout-note collapse="true"}
### Code for the values and table below
```{r dmd-extract}
#| output: false
dmd_b   <- coef(dmd_glm)
dmd_se  <- summary(dmd_glm)$coefficients[, "Std. Error"]
dmd_eta <- predict(dmd_glm)
dmd_p   <- fitted(dmd_glm)

dmd_or10    <- unname(exp(10 * dmd_b["CK"]))
dmd_ci10    <- exp(10 * confint.default(dmd_glm)["CK", ])
glm_eps     <- 10 * .Machine$double.eps
n_triggered <- sum(dmd_p > 1 - glm_eps | dmd_p < glm_eps)

# The true mathematical value of 1 - logistic(eta), computed without ever
# forming logistic(eta) itself, so it is not limited by how close to 1 a
# double can get.
true_gap <- exp(-max(dmd_eta)) / (1 + exp(-max(dmd_eta)))

dmd_tail_gaps <- (1 - dmd_p)[order(-dmd$CK)][1:5]

dmd_tail_table <- dmd |>
  mutate(eta = dmd_eta, p = dmd_p) |>
  slice_max(CK, n = 5) |>
  transmute(CK,
            Group = as.character(Group),
            eta = sprintf("%.1f", eta),
            gap = sprintf("$%s$", sci(1 - p))) |>
  knitr::kable(col.names = c("CK", "Group", "$\\hat\\eta$", "$1 - \\hat p$"),
               align = "ccrr")
```
:::

The estimated slope is $\hat\beta_1 = `r sprintf("%.4f", dmd_b["CK"])`$ with a
standard error of `r sprintf("%.4f", dmd_se["CK"])`, a ratio of
$z = `r sprintf("%.2f", dmd_b["CK"] / dmd_se["CK"])`$. The algorithm reports
`dmd_glm$converged` as `r dmd_glm$converged` after
`r dmd_glm$iter` iterations. Nothing in the coefficient table resembles the
toy example's standard error of
`r format(round(toy_se[2]), big.mark = ",")`.

What the warning is responding to sits at the far right of the plot. The last
column below is $1 - \hat p$ as `fitted()` reports it.

```{r dmd-tail-table}
#| echo: false
dmd_tail_table
```

The largest linear predictor in the fit is
$\hat\eta = `r sprintf("%.2f", max(dmd_eta))`$. Mathematically,

$$1 - \text{logistic}(\hat\eta) = \frac{e^{-\hat\eta}}{1 + e^{-\hat\eta}} = `r sci(true_gap)`$$

at that value — far below $2^{-53} = `r sci(2^-53)`$, the gap between $1$ and
the largest double-precision number below it (`.Machine$double.eps`
$= 2^{-52} = `r sci(.Machine$double.eps)`$ is the corresponding gap just
*above* $1$). No double can hold a probability that close to $1$: computed
directly, $1$ minus that gap rounds to exactly $1$. A fitted probability of
exactly $0$ or $1$ would make the deviance infinite for any observation whose
response disagreed with it, and would give that observation zero weight in
the next iteration, so `glm()` does not let it happen — it clamps the linear
predictor before inverting the link, and `fitted()` never leaves
$[`r sci(.Machine$double.eps)`,\ 1 - `r sci(.Machine$double.eps)`]$. The top
row of the table is therefore the clamped value rather than the true one:
`r round(log10(.Machine$double.eps / true_gap))` orders of magnitude larger
than the true gap just derived, and
`r round(log10(as.numeric(dmd_tail_gaps[2]) / .Machine$double.eps))` orders of
magnitude smaller than the row beneath it in the table — the column only
looks like it descends smoothly.

`glm()` warns whenever a fitted probability comes within
$`r sci(glm_eps)`$ of $0$ or $1$, and here exactly
`r n_triggered` of the `r nrow(dmd)` observations does.

So the warning fired for a reason that has nothing to do with the estimator
failing. One woman has a `CK` of `r max(dmd$CK)`, while the other
`r nrow(dmd) - 1` observations sit at
`r sort(dmd$CK, decreasing = TRUE)[2]` or below, and the fitted curve has
saturated well before it reaches her.

**Conclusion.** Creatine kinase is strongly associated with carrier status:
each additional 10 units of `CK` multiplies the estimated odds of being a
carrier by `r sprintf("%.2f", dmd_or10)` (95% CI
`r sprintf("%.2f", dmd_ci10[1])` to `r sprintf("%.2f", dmd_ci10[2])`). The
estimate is finite, its standard error is small relative to it, and the
iteration converged, so the warning here is a report about one extreme
observation rather than about the estimator. The same warning in the toy
example meant the estimate did not exist.

## Recognizing and responding to separation

The two examples give the same warning for different reasons, so the warning is
where an investigation starts rather than where it ends. What settles the
question is the coefficient table.

### Recognizing separation

- A `glm.fit: algorithm did not converge` warning. This one is close to
conclusive: when a strictly concave log-likelihood has a finite maximizer, the
algorithm locates it in a handful of iterations.
- A `glm.fit: fitted probabilities numerically 0 or 1 occurred` warning. This
one is not conclusive on its own, as the muscular dystrophy example shows, but
separation always produces it.
- A coefficient whose magnitude is absurd on the feature's own scale. An odds
ratio of $e^{`r sprintf("%.1f", toy_b[2])`}$ per unit is not a quantity any
real process produces.
- A standard error far larger than the coefficient it belongs to, so that a
feature which classifies the training data perfectly returns $z \approx 0$ and
a $p$-value near $1$. This is the diagnostic that is hardest to misread,
because it is the reverse of what a strong feature normally looks like.
- Under complete separation, a residual deviance of essentially zero, meaning
the fitted probabilities reproduce every observed response exactly. Under
quasi-complete separation the deviance settles at whatever the tied
observations on the boundary contribute, so it is small rather than zero.
- Fitted probabilities numerically indistinguishable from $0$ or $1$ in
`fitted()`, for observations on the separated side of the boundary. They are
never *exactly* $0$ or $1$: `glm()` clamps them into
$[`r sci(.Machine$double.eps)`,\ 1 - `r sci(.Machine$double.eps)`]$, so that is
what a separated observation returns.

Checking `fit$converged` and comparing each $|\hat\beta_j|$ against its
standard error takes one line and distinguishes the two situations.

### Responding to separation

Within the methods this course has covered, there are three responses.

**Collect more data.** Separation is a property of a sample, not of the
population it came from. Ten observations and one threshold feature separate
almost by construction; `r nrow(dmd)` observations with two overlapping
distributions do not, even with one `CK` value
`r round(max(dmd$CK) / median(dmd$CK))` times the sample median. The
probability that a sample happens to be separated falls as $n$ grows and as the
classes overlap more.

**Simplify the model.** If a feature or a dummy-coded level is what makes the
data separable, and that feature is not the object of the analysis, dropping it
or merging it with another level removes the separation. This is most common
with a categorical feature that has a level in which the event never occurs:
that level alone can separate the data, and combining it with a neighboring
level fixes the fit without touching the rest of the model.

**Use the fit for what it can still support.** Separation destroys the
*coefficient*, not necessarily the *classification*. The toy model's slope is
meaningless, but its decision boundary sits between $x = 5$ and $x = 6$ and
classifies every observation correctly. A model that separates on one feature
can still produce usable predicted classes away from the separated region, as
long as no probability, odds ratio, standard error, or $p$-value involving the
offending coefficient is reported.

::: {.callout-note collapse="true"}
### Beyond this course
The principled fix is to stop maximizing $\ell(\beta)$ by itself and maximize a
*penalized* version, $\ell(\beta) - \lambda \cdot \text{pen}(\beta)$, where the
penalty grows with the magnitude of $\beta$. A penalty that increases without
bound as $\|\beta\| \to \infty$ restores a finite maximizer no matter how the
data are arranged, because the penalty eventually outruns the gain in $\ell$.

**Firth's penalized likelihood** uses a penalty derived from the information
matrix and is the standard remedy for separation; it also removes most of the
small-sample bias in $\hat\beta$. The **ridge** and **lasso** penalties
forward-referenced in the multiple linear regression lecture do the same job
from a different motivation, and both apply to logistic regression unchanged.
A Bayesian analysis with a proper prior on $\beta$ achieves the same effect
again, since the prior plays the role of the penalty.

All of these are shrinkage methods, and shrinkage is ISLR2 Chapter 6 material.
Name the fix if you meet separation before then; do not try to implement it
from this callout.
:::

## Problems carried over from linear regression

The flexibility lecture listed six potential problems for a linear regression:
non-linearity, correlated errors, non-constant variance, outliers,
high-leverage points, and collinearity. Separation has no counterpart on that
list: whenever $X^\top X$ is invertible, least squares evaluates a formula and
so cannot fail this way. The six themselves transfer to logistic regression
unevenly.

### Correlated observations transfer unchanged

Correlated errors are the one that transfers with no change at all — the same
item, under the name **correlated observations**, since logistic regression has
no $\epsilon_i$ to call correlated. The Bernoulli log-likelihood is a sum over
observations precisely because they were assumed independent, so responses
correlated in time or space distort every standard error from `glm()`, for the
same reason and in the same direction as in linear regression: positively
correlated observations carry less information than their count suggests, and
the reported standard errors come out too small.

### Non-constant variance is not a separate assumption

Linear regression assumes $Var[\epsilon_i] = \sigma^2$ and needs that
assumption checked, usually by looking for a fan shape in a plot of residuals
against fitted values.

Logistic regression has nothing to check. The simple logistic regression
lecture established that

$$Var[Y_i|X_i] = p(X_i)\left[1 - p(X_i)\right]$$

is a consequence of $Y_i$ being Bernoulli, not an extra assumption laid on top
of the model. The variance is *required* to change with $X_i$, and it is
determined once $p(X_i)$ is, which is also why the model has no $\sigma^2$
parameter. A residual plot that fanned out would not be evidence against this
model.

### Non-linearity of the log-odds

What logistic regression does assume is that the *log-odds* are linear in the
features,

$$\log\left(\frac{p(x_i)}{1 - p(x_i)}\right) = x_i\beta.$$

That assumption can fail, and the previous lecture is where it was addressed:
interactions let a feature's log-odds slope depend on another feature, and
polynomial and step-function columns let it bend within a single feature. Both
enter as columns of the model matrix, and both are checked the same way — fit
the more flexible model and test the added coefficients.

### Collinearity, leverage, and outliers

These three carry over in substance. Features carrying nearly the same
information still leave the data unable to separate their contributions, so
$\text{SE}(\hat\beta_j)$ inflates and a $z$-test loses power. An observation
with an unusual feature value still has disproportionate pull on the fit, and
an observation whose response disagrees with the rest of the data still
distorts it.

What does not carry over is the arithmetic. The hat matrix
$H = X(X^\top X)^{-1}X^\top$, studentized residuals, and Cook's distance were
all defined through the least squares fit, and none of those formulas applies
to a likelihood maximized iteratively. Whether the woman with
`CK` = `r max(dmd$CK)` counts as a high-leverage point is a question the two
fits answer differently, and the least squares formula gives the wrong answer.

::: {.callout-note collapse="true"}
### Beyond this course
Each of the three has a generalized-linear-model analogue, built from the
weighted least squares problem that the iterative algorithm solves at its last
step. The weights $\hat p_i(1 - \hat p_i)$ from that final iteration produce a
weighted hat matrix, and from it a leverage, a standardized residual, and a
Cook's distance that mean what their linear-model namesakes mean. In R,
`hatvalues()`, `rstudent()`, and `cooks.distance()` all accept a `glm` object
and return exactly these.

The weighting is what settles the question above.

```{r dmd-leverage}
#| output: false
dmd_extreme <- which.max(dmd$CK)

dmd_lm  <- lm(case01 ~ CK, data = dmd)
h_lm    <- hatvalues(dmd_lm)
h_glm   <- hatvalues(dmd_glm)
cd_glm  <- cooks.distance(dmd_glm)

# The weights p(1-p) computed from eta rather than from the clamped fitted(),
# so the extreme one is its true mathematical value rather than a floored one.
w_eta       <- exp(-dmd_eta) / (1 + exp(-dmd_eta))^2
w_true      <- unname(w_eta[which.max(dmd$CK)])
w_ratio_min <- min(w_eta[-which.max(dmd$CK)]) / w_true

# Fixed notation where it reads well, scientific where it does not.
fmt_val <- function(v) ifelse(abs(v) >= 1e-3, sprintf("%.3f", v), sci(v))

dmd_leverage_table <- tibble(
  Quantity = c("Least squares leverage $h_i$",
               "Logistic leverage $h_i$",
               "Logistic Cook's distance"),
  Value    = sprintf("$%s$", fmt_val(c(h_lm[dmd_extreme], h_glm[dmd_extreme],
                                       cd_glm[dmd_extreme]))),
  Rank     = c("largest", "smallest", "smallest")
) |>
  knitr::kable(col.names = c("Quantity",
                             sprintf("Value for the `CK` = %d woman",
                                     max(dmd$CK)),
                             sprintf("Rank among the %d", nrow(dmd))),
               align = "lcc")
```

```{r dmd-leverage-table}
#| echo: false
dmd_leverage_table
```

Regressing the $0/1$ response on `CK` by least squares makes her the
highest-leverage observation in the data, since leverage there depends only on
how far a feature value sits from the mean. The logistic fit weights that
geometry by $\hat p_i(1 - \hat p_i)$ — reweighting $X^\top X$ along with it,
not just scaling her row — and that weight at
$\hat\eta = `r sprintf("%.2f", max(dmd_eta))`$ is $`r sci(w_true)`$ — smaller
than every other observation's weight by a factor of at least
`r format(signif(w_ratio_min, 2), big.mark = ",", scientific = FALSE)`. So she
ends up with the *smallest* leverage of the
`r nrow(dmd)` observations, against a maximum of
`r sprintf("%.3f", max(h_glm))`, and the smallest Cook's distance. Saturating
the curve is exactly what costs an observation its influence: the model has
already committed to a $\hat p$ indistinguishable from $1$ for her, and no
small change in $\hat\beta$ alters that, so she cannot pull the fit anywhere.
This is the other half of why the fit converged to something usable despite the
warning she triggered.

Collinearity needs no new definition: the variance inflation factor of the
flexibility lecture, $1/(1-R_j^2)$, is computed from the features alone and
never looks at the response, so that quantity is unchanged. (`car::vif()`
run on a `glm` reports a different, weighted version built from
$(X^\top\hat W X)^{-1}$, so the two do not agree numerically — the *idea*
carries over, not the specific number a package function returns.)

None of this is tested in this course, which is why the diagnostics in the
examples above were read off the fitted coefficients and the data rather than
off these quantities.
:::

## Conclusion

Maximum likelihood for logistic regression can fail in a way least squares
cannot. When some linear predictor classifies every training observation
correctly — complete separation, or quasi-complete separation if ties are
allowed on the boundary — the log-likelihood climbs toward a supremum it
attains only in the limit, so no finite $\hat\beta$ maximizes it. What `glm()`
prints in that situation is an arbitrary point along a divergent path, with
standard errors that grow as the curvature of $\ell$ flattens.

The resulting output is recognizable once its signature is familiar: a
`glm.fit` warning, a coefficient far too large for its feature's scale, a
standard error larger still, a $z$-statistic near zero, and a residual deviance
near nothing. The two examples bracket the range. Ten constructed observations
produced a slope of `r sprintf("%.1f", toy_b[2])` with a standard error of
`r format(round(toy_se[2]), big.mark = ",")`, a fit that failed to converge,
and a perfectly predictive feature reported as not significant. The sample of
`r nrow(dmd)` real women produced the same warning from a single extreme blood
measurement, with a converged fit and an entirely usable estimate.

That completes what this course covers of logistic regression before Exam 1.
