# Linear Discriminant Analysis

[Slides for this lecture](50-lda_slides.html)

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
set.seed(20261001)

# MASS is called as MASS::lda() rather than attached, because attaching it
# masks dplyr::select().
# Renders a number in LaTeX scientific notation for use inside $...$.
sci <- function(x, digits = 2) {
  e <- floor(log10(abs(x)))
  m <- signif(x / 10^e, digits)
  sprintf("%s \\times 10^{%d}", format(m), e)
}

class_colors <- c("#0072B2", "#D55E00")
model_colors <- c("LDA" = "#009E73", "Logistic regression" = "#CC79A7")
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **derive** the posterior probability $p_k(x) = P(Y = k \mid X = x)$ from the
prior probabilities $\pi_k$ and the class-conditional densities $f_k(x)$ using
Bayes' theorem, and **explain** why classifying to the class with the largest
posterior probability minimizes the expected error rate,
- **derive** the LDA discriminant function $\delta_k(x)$ for one feature under
Gaussian class-conditional densities with a shared variance $\sigma^2$, and
**show** that it is linear in $x$, so that the decision boundary is a single
point that moves toward, and when the prior is small enough past, the mean of
the class whose prior probability is smaller,
- **write** the discriminant function for $p > 1$ features in matrix form using
class mean vectors $\mu_k$ and a shared covariance matrix $\Sigma$, and
**compute** the estimates $\hat\pi_k$, $\hat\mu_k$, and $\hat\sigma^2$ or
$\hat\Sigma$ from training data,
- **fit** LDA with `MASS::lda()`, **interpret** its prior probabilities, group
means, and posterior probabilities, and **compare** its posterior
probabilities and decision boundary with those of logistic regression,
- **construct** a confusion matrix at a stated threshold (or a $K \times K$
confusion matrix for $K$ classes), **compute** the error rate, sensitivity,
and specificity from it, and **explain** how sensitivity
and specificity trade off as the threshold moves, and
- **interpret** an ROC curve and its AUC, including AUC as the probability that
a randomly chosen positive observation receives a higher score than a randomly
chosen negative one.
:::


## Generative classifiers

### Credit card default data

**Question.** Can a classifier built from how balance and income are
distributed within defaulters and within non-defaulters identify credit card
customers who will default, and how does it compare with logistic regressions
fit to the same customers and features?

These customers were analyzed in the logistic regression lectures: the
[Simple Logistic Regression](10-logistic-regression.qmd) lecture modeled
default from `balance`, the
[Multiple Logistic Regression](20-multiple-logistic-regression.qmd) lecture
from `balance` and `student`, and the
[Flexible Logistic Regression](30-flexible-logistic-regression.qmd) lecture
added interactions, including one between `balance` and `income`.

**Data.** `ISLR2::Default` is a simulated data set of
`r format(nrow(Default), big.mark = ",")` credit card customers. The response
is `default`, with `Yes` the positive class. This lecture uses the two
quantitative explanatory variables, `balance`, the customer's monthly credit
card balance in dollars, and `income`, the customer's income in dollars,
because linear discriminant analysis (LDA), the method of this lecture,
models each class's features with a Gaussian density. The
features are the explanatory variables themselves. `income` enters the models
in dollars as recorded, as in the flexible logistic regression lecture; its
coefficients are reported here per \$1,000 of income. `student` (Yes/No)
explains a feature of the income data in the Multiple-feature LDA section and
splits the customers in the Student strata section.

The table gives each default class's number and proportion of customers and
the mean, standard deviation, and number of zero values of its balances. The
first tab below
shows balance within each class, one panel per class, on the density scale;
the second shows all customers in one histogram on the count scale, with each
bar split into non-defaulters (light blue) and defaulters (dark orange).

::: {.callout-note collapse="true"}
### Code
```{r default-data}
#| output: false
Default <- Default |>
  mutate(default01   = as.numeric(default == "Yes"),
         default_jit = default01 + runif(n(), -0.03, 0.03))

dollars <- function(v) format(round(v), big.mark = ",")
# Axis labels such as $60,000, without the scales package.
dollar_labels <- function(v) {
  ifelse(is.na(v), NA,
         paste0("$", format(v, big.mark = ",", scientific = FALSE, trim = TRUE)))
}

class_summary <- Default |>
  group_by(default) |>
  summarize(n = n(), prop = n() / nrow(Default),
            mean_balance = mean(balance), sd_balance = sd(balance),
            n_zero = sum(balance == 0), .groups = "drop")
summary_of <- function(k, col) class_summary[[col]][class_summary$default == k]

class_summary_table <- class_summary |>
  transmute(default, n = format(n, big.mark = ","),
            prop = sprintf("%.4f", prop),
            mean_balance = dollars(mean_balance),
            sd_balance = dollars(sd_balance),
            n_zero = format(n_zero, big.mark = ",")) |>
  knitr::kable(col.names = c("Default", "Customers", "Proportion",
                             "Mean balance (\\$)", "SD of balance (\\$)",
                             "Zero balances"),
               align = "cccccc")

balance_hist_plot <- ggplot(Default, aes(x = balance)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 100,
                 boundary = 0, fill = "grey80", color = "grey55") +
  facet_wrap(~ default, ncol = 1, labeller = label_both) +
  scale_x_continuous(labels = dollar_labels) +
  labs(x = "Credit card balance ($)", y = "Density",
       title = "ISLR2::Default, balance within each default class")

# All customers on the count scale, each bar split by default status. The
# fills differ in lightness so the split survives without color.
balance_stack_plot <- ggplot(Default, aes(x = balance, fill = default)) +
  geom_histogram(binwidth = 100, boundary = 0, color = "grey30",
                 linewidth = 0.2) +
  scale_fill_manual(values = c(No = "#9DC3E6", Yes = "#D55E00")) +
  scale_x_continuous(labels = dollar_labels) +
  labs(x = "Credit card balance ($)", y = "Customers", fill = "Default",
       title = "ISLR2::Default, all customers by default status")

# Share of defaulters in each $100 balance bin holding at least min_bin_n
# customers.
bin_width <- 100
min_bin_n <- 10
example_bin <- 2000
bin_share <- Default |>
  mutate(bin = floor(balance / bin_width) * bin_width) |>
  group_by(bin) |>
  summarize(n = n(), share = mean(default == "Yes"), .groups = "drop") |>
  filter(n >= min_bin_n)
share_at <- function(b) bin_share$share[bin_share$bin == b]
first_majority_bin <- min(bin_share$bin[bin_share$share > 0.5])
```
:::

```{r class-summary-table}
#| echo: false
class_summary_table
```

::: {.panel-tabset}
#### Within each class
```{r default-hist-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
balance_hist_plot
```

#### All customers
```{r default-stack-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.6
balance_stack_plot
```
:::

Only `r format(summary_of("Yes", "n"), big.mark = ",")` customers, a proportion
of `r sprintf("%.4f", summary_of("Yes", "prop"))`, defaulted. Within each
class the balances form a single, roughly bell-shaped pile, and the two piles
are centered at different values: defaulters average
\$`r dollars(summary_of("Yes", "mean_balance"))` against
\$`r dollars(summary_of("No", "mean_balance"))` for non-defaulters. The
standard deviations, \$`r dollars(summary_of("No", "sd_balance"))` for
non-defaulters and \$`r dollars(summary_of("Yes", "sd_balance"))` for
defaulters, are of the same order but not equal, and the non-defaulter
histogram has a spike at zero:
`r format(summary_of("No", "n_zero"), big.mark = ",")` customers carry no
balance, and
`r if (summary_of("Yes", "n_zero") == 0) "no defaulter does" else paste(summary_of("Yes", "n_zero"), "defaulters do")`.

On the count scale, defaulters are a thin layer at the bottom of the bars at
most balances and make up most of a bar only at the largest balances.

### Bayes' theorem

With the response falling in one of $K$ classes, indexed $k = 1, \ldots, K$,
logistic regression models the conditional probability
$P(Y = k \mid X = x)$ directly. Its likelihood is the conditional
distribution of the response given the features, and the distribution of the
features themselves never enters it.

A **generative classifier** takes the other route. It models how the features
are distributed *within* each class, as in the first tab above, and how
common each class is, and then reverses the conditioning with Bayes' theorem
to obtain $P(Y = k \mid X = x)$. LDA is the first generative classifier in
this course.

Every earlier lecture conditioned on the features and treated them as fixed. A
generative classifier models their distribution, so here $X$ is a random
vector — the features of a randomly chosen observation — and $x$ is a value it
takes. With $Y$ the response, define

$$\pi_k = P(Y = k), \qquad f_k(x) = \text{density of } X \text{ given } Y = k,$$

where

- $K$ is the number of classes ($K = 2$ for a binary response),
- $\pi_k$ is the **prior probability** of class $k$, the probability that a
randomly chosen observation belongs to class $k$ before its features are
seen, with $\sum_{k=1}^K \pi_k = 1$, and
- $f_k(x)$ is the **class-conditional density** of $X$ among observations in
class $k$ (a probability mass function if $X$ is discrete).

The classification and logistic regression lectures wrote $C$ for the number
of classes; following ISLR2, this and later lectures write $K$.

Bayes' theorem, with densities in place of the probabilities of the events
$\{X = x\}$, gives

$$P(Y = k \mid X = x) = \frac{P(Y = k)\, f_k(x)}{f(x)},$$

where $f(x)$ is the marginal density of $X$. By the law of total probability,
$f(x) = \sum_{l=1}^K P(Y = l)\, f_l(x)$, so

$$p_k(x) = P(Y = k \mid X = x) = \frac{\pi_k f_k(x)}{\sum_{l=1}^K \pi_l f_l(x)} \propto \pi_k f_k(x),$$

where

- $p_k(x)$ is the **posterior probability** of class $k$ for an observation
with features $x$,
- $l$ indexes the classes in the sum, kept distinct from the $k$ being
evaluated, and
- $\propto$ means "proportional to" as a function of $k$ at a fixed $x$.

Only the numerator distinguishes one class from another: a class is probable
at $x$ when it is common ($\pi_k$ large) and when $x$ is a typical feature
value for it ($f_k(x)$ large).

For the credit card customers the class proportions estimate the priors,
$\hat\pi_{\text{No}} = `r sprintf("%.4f", summary_of("No", "prop"))`$ and
$\hat\pi_{\text{Yes}} = `r sprintf("%.4f", summary_of("Yes", "prop"))`$, and
the two panels of the first tab picture $f_{\text{No}}$ and
$f_{\text{Yes}}$ for balance. The second tab pictures the numerators: each
colored segment's height is proportional to $\pi_k f_k(x)$ over its bin, the
whole bar's to their sum $f(x)$, and the defaulters' share of the bar
estimates $p_{\text{Yes}}(x)$. Among customers with balances from
\$`r dollars(example_bin)` to \$`r dollars(example_bin + bin_width)`, for
example, a share of `r sprintf("%.3f", share_at(example_bin))` defaulted.

### Bayes classifier

A **classifier** is an assignment rule $\hat y(x)$ that maps each value of
the features to one of the classes $1, \ldots, K$. The **Bayes classifier**
assigns each observation to the class with the largest posterior probability,

$$\hat y(x) = \arg\max_k\, p_k(x),$$

where

- $\hat y(x)$ is the class assigned to an observation with features $x$, and
- $\arg\max_k p_k(x)$ is the class $k$ at which $p_k(x)$ is largest.

With $K = 2$ classes, $p_1(x) = 1 - p_2(x)$, so the Bayes classifier assigns
class $2$ whenever $p_2(x) > 0.5$.

For any classifier, the probability that the assignment at $x$ is wrong is

$$P\left(Y \ne \hat y(x) \mid X = x\right) = 1 - p_{\hat y(x)}(x).$$

This is smallest when $p_{\hat y(x)}(x)$ is largest, which is exactly the
Bayes classifier's choice. The Bayes classifier therefore minimizes the
conditional error probability at every $x$, and so it also minimizes the
overall error rate $E\left[1 - p_{\hat y(X)}(X)\right]$, which averages those
conditional error probabilities over the distribution of $X$. Its overall
error rate, the **Bayes error rate**, is the lowest achievable by any
classifier.

An estimate of this rule from the bars of the stacked histogram assigns `Yes`
where defaulters make up more than half a bar, which first happens in the bin
starting at \$`r dollars(first_majority_bin)` (among bins with at least
`r min_bin_n` customers). With no feature information the largest posterior
probability is the largest prior, so every customer would be assigned `No`, and the
`r sprintf("%.2f", 100 * summary_of("Yes", "prop"))`% who defaulted would all
be misclassified; that error rate is the one any classifier using features
has to beat.

The Bayes classifier needs the true $\pi_k$ and $f_k$, which are never known.
A generative classifier estimates both from training data and substitutes the
estimates into Bayes' theorem. Estimating $\pi_k$ takes nothing more than the
proportion of training observations in class $k$. Estimating $f_k$ requires an
assumption about its form, and LDA's assumption is that it is Gaussian.

## One-feature LDA

The one feature here is balance. The histograms below show balance within
each default class, one panel per class, on the density scale.

```{r onefeature-hist-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
balance_hist_plot
```

Each class's balances form a single, roughly bell-shaped pile; the two piles
are centered at different balances and have spreads of the same order. That
suggests a model: describe each class's balance by a Gaussian density with its
own mean, the center of its pile, and one variance shared by both classes, the
common width of the piles.

### Gaussian class-conditional densities

With a single feature ($p = 1$), LDA assumes the statistical model

$$P(Y = k) = \pi_k, \qquad X \mid Y = k \sim N\left(\mu_k, \sigma^2\right), \qquad k = 1, \ldots, K,$$

where

- $\mu_k = E[X \mid Y = k]$ is the mean of the feature in class $k$, and
- $\sigma^2 = Var[X \mid Y = k]$ is the variance of the feature, **the same in
every class**.

The class-conditional density is therefore the normal density

$$f_k(x) = \frac{1}{\sqrt{2\pi}\,\sigma} \exp\left[-\frac{(x - \mu_k)^2}{2\sigma^2}\right],$$

where $\pi$ without a subscript is the constant $3.14159\ldots$, unrelated to the
prior probabilities $\pi_k$.

The classes differ only in where their densities are centered. For the credit
card customers, $\mu_k$ is the center of class $k$'s pile of balances in the
histograms above, and $\sigma$ is the width the two piles share.

### Parameter estimation

The parameters $\pi_k$, $\mu_k$, and $\sigma^2$ are estimated from training
observations $(x_1, y_1), \ldots, (x_n, y_n)$, here with $x_i$ the scalar
feature value of observation $i$:

$$\hat\pi_k = \frac{n_k}{n}, \qquad \hat\mu_k = \frac{1}{n_k}\sum_{i:\, y_i = k} x_i, \qquad \hat\sigma^2 = \frac{1}{n - K}\sum_{k=1}^K \sum_{i:\, y_i = k}\left(x_i - \hat\mu_k\right)^2,$$

where

- $n_k$ is the number of training observations in class $k$, so
$\sum_k n_k = n$,
- $\sum_{i:\, y_i = k}$ sums over the training observations in class $k$, and
- $\hat\sigma^2$ is the **pooled variance**: squared deviations from each
observation's own class mean, summed over all classes and divided by $n - K$.

$\hat\pi_k$ and $\hat\mu_k$ are maximum likelihood estimates; $\hat\sigma^2$
is the maximum likelihood estimate rescaled by $n/(n - K)$ to make it
unbiased. Under the generative model the joint
density of observation $i$ is $P(Y = y_i)\, f_{y_i}(x_i) = \pi_{y_i} f_{y_i}(x_i)$,
so for independent observations the log-likelihood is

$$\ell(\pi_1, \ldots, \pi_K, \mu_1, \ldots, \mu_K, \sigma^2) = \sum_{k=1}^K n_k \log \pi_k \;+\; \sum_{k=1}^K \sum_{i:\, y_i = k}\left[-\frac{1}{2}\log\left(2\pi\sigma^2\right) - \frac{(x_i - \mu_k)^2}{2\sigma^2}\right].$$

The first sum involves only the priors; maximizing it subject to
$\sum_k \pi_k = 1$ (with a Lagrange multiplier) gives $\hat\pi_k = n_k/n$. In
the second, $\mu_k$ appears only in class $k$'s terms, so its maximizer is the
class-$k$ sample mean. Setting the derivative with respect to $\sigma^2$ to
zero gives the pooled sum of squares divided by $n$; dividing by $n - K$
instead, one degree of freedom per estimated mean, makes it unbiased, in the
same way that $RSS/(n-2)$ replaced $RSS/n$ in simple linear regression.
`MASS::lda()` uses $n - K$.

Logistic regression maximizes a different likelihood, the conditional one,
$\prod_{i=1}^n p(x_i)^{y_i}\left[1 - p(x_i)\right]^{1 - y_i}$ (with $y_i$
coded $0/1$ as in the logistic regression lectures), which has no closed-form
maximizer. LDA maximizes the joint likelihood of features and
responses under a stronger assumption — Gaussian features with a shared
variance — and every estimate is a sample proportion, mean, or variance.

`MASS::lda()` takes the same formula interface as `glm()`. Fitting `default`
on `balance` prints three blocks: the prior probabilities, the group means,
and the coefficients of linear discriminants.

```{r balance-lda}
balance_lda <- MASS::lda(default ~ balance, data = Default)
balance_lda
```

::: {.callout-note collapse="true"}
### Code
```{r lda-by-hand}
#| output: false
# Estimates of pi_k, mu_k, Sigma and the implied log posterior odds
# beta0 + x' beta of the second class, computed from the formulas above.
lda_by_hand <- function(X, y) {
  X <- as.matrix(X)
  classes <- levels(y)
  n <- nrow(X)
  K <- length(classes)
  pi_hat <- as.numeric(table(y)) / n
  mu_hat <- do.call(rbind, lapply(classes, function(k) {
    colMeans(X[y == k, , drop = FALSE])
  }))
  rownames(mu_hat) <- classes
  Sigma_hat <- Reduce(`+`, lapply(classes, function(k) {
    Z <- sweep(X[y == k, , drop = FALSE], 2, mu_hat[k, ])
    crossprod(Z)
  })) / (n - K)
  beta  <- solve(Sigma_hat, mu_hat[2, ] - mu_hat[1, ])
  beta0 <- log(pi_hat[2] / pi_hat[1]) -
    0.5 * (sum(mu_hat[2, ] * solve(Sigma_hat, mu_hat[2, ])) -
           sum(mu_hat[1, ] * solve(Sigma_hat, mu_hat[1, ])))
  list(pi = pi_hat, mu = mu_hat, Sigma = Sigma_hat,
       beta0 = beta0, beta = beta)
}

hand_1 <- lda_by_hand(Default["balance"], Default$default)
sigma_hat_1 <- sqrt(hand_1$Sigma[1, 1])

balance_grid <- tibble(balance = seq(0, 2800, length.out = 500))
gauss_curves <- tibble(default = factor(c("No", "Yes"))) |>
  mutate(mu = hand_1$mu[, 1]) |>
  cross_join(balance_grid) |>
  mutate(f = dnorm(balance, mu, sigma_hat_1))

gauss_plot <- balance_hist_plot +
  geom_line(data = gauss_curves, aes(y = f, color = default,
                                     linetype = default),
            linewidth = 0.9) +
  scale_color_manual(values = class_colors) +
  labs(color = "Default", linetype = "Default",
       title = expression(paste("Histograms with fitted ",
                                N(hat(mu)[k], hat(sigma)^2), " densities")))
```
:::

The figure below draws each class's fitted density
$N\left(\hat\mu_k, \hat\sigma^2\right)$ over that class's histogram from the
start of this section.

```{r balance-gauss-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
gauss_plot
```

The **prior probabilities** are $\hat\pi_{\text{No}} = `r sprintf("%.4f", hand_1$pi[1])`$
and $\hat\pi_{\text{Yes}} = `r sprintf("%.4f", hand_1$pi[2])`$, the class
proportions of the class summary table. The **group means** are
$\hat\mu_{\text{No}} = `r gsub(",", "{,}", format(round(hand_1$mu[1, 1], 2), nsmall = 2, big.mark = ","))`$
and $\hat\mu_{\text{Yes}} = `r gsub(",", "{,}", format(round(hand_1$mu[2, 1], 2), nsmall = 2, big.mark = ","))`$
dollars, the mean balances of the same table. The pooled standard deviation is
$\hat\sigma = `r sprintf("%.2f", sigma_hat_1)`$, between the two
class-specific standard deviations.

In the figure, the fitted density for non-defaulters misses the spike at zero,
and the one for defaulters, forced to share $\hat\sigma$ with the larger
class, is wider than their histogram: the two departures from LDA's
assumptions that the histograms showed.

The **coefficient of the linear discriminant**,
`r signif(balance_lda$scaling[1, 1], 4)`, equals
$1/\hat\sigma = `r signif(1 / sigma_hat_1, 4)`$: `MASS::lda()`
reports the discriminant direction rescaled so that the printed linear
combination, $\text{LD1} = \text{coefficient} \times \texttt{balance}$, has
variance $1$ within each class, and with one feature the direction itself is
just "larger balance points toward `Yes`." This is Fisher's scaling,
described in the callout below.

::: {.callout-note collapse="true"}
### Beyond this course
The "coefficients of linear discriminants" printed by `MASS::lda()` come from
R. A. Fisher's original formulation of discriminant analysis, which asks for
the linear combination of features that maximizes the ratio of
between-class to within-class variance and makes no Gaussian assumption.
With $K$ classes that problem yields up to $K - 1$ discriminant directions,
printed as `LD1`, `LD2`, ..., and normalized to have unit within-class
variance. With Gaussian classes and a shared covariance matrix, Fisher's
directions and the Bayes'-theorem discriminants of this lecture produce the
same classifier.
:::

::: {.callout-note collapse="true"}
### For example,
LDA's estimates have closed forms, and separation does not prevent them from
existing, though a zero pooled variance still would. The previous
lecture's completely separated toy data, $x = 1, \ldots, 10$ with $y = 1$
exactly when $x > 5$, left logistic regression with no finite maximum
likelihood estimate. `MASS::lda()` fits LDA to the same data:

```{r toy-lda}
toy <- tibble(x = 1:10, y = factor(as.numeric(x > 5)))
toy_lda <- MASS::lda(y ~ x, data = toy)
toy_lda$means
```

The class means are `r toy_lda$means[1]` and `r toy_lda$means[2]` and the
priors are equal, so LDA classifies by which class mean is closer: the
boundary is their midpoint, `r mean(toy_lda$means)`.
:::

### Discriminant function

Since $p_k(x)$ is proportional to $\pi_k f_k(x)$ and the logarithm is
increasing, the class with the largest posterior probability is also the class
with the largest $\log\left[\pi_k f_k(x)\right]$. Expanding the square and
collecting terms,

$$\log\left[\pi_k f_k(x)\right] = \log \pi_k - \frac{\mu_k^2}{2\sigma^2} + \frac{\mu_k}{\sigma^2}\, x - \log\left(\sqrt{2\pi}\,\sigma\right) - \frac{x^2}{2\sigma^2}.$$

The last two terms do not involve $k$ because $\sigma^2$ is shared, so they
cannot change which class is largest. Dropping them gives the **discriminant
function**

$$\delta_k(x) = \left(\log \pi_k - \frac{\mu_k^2}{2\sigma^2}\right) + \frac{\mu_k}{\sigma^2}\, x,$$

where

- $\delta_k(x)$ is the discriminant value of class $k$ at feature value $x$,
- $\log \pi_k - \mu_k^2/\left(2\sigma^2\right)$ is its intercept, the score
at $x = 0$, and
- $\mu_k/\sigma^2$ is its slope.

The dropped term $-x^2/\left(2\sigma^2\right)$ was the only one quadratic in
$x$, so each $\delta_k$ is linear in $x$, with the same intercept-plus-slope form as a regression's
linear predictor. The classifier assigns an observation to the
class with the largest $\delta_k(x)$. The slope $\mu_k/\sigma^2$ is not the
coefficient of the linear discriminant printed by `MASS::lda()`, which is
Fisher's scaling of the discriminant direction.

Had the variances differed, with $\sigma_k^2$ the class-$k$ variance, the
term $-\log\left(\sqrt{2\pi}\,\sigma_k\right)$ would differ between classes
and stay in each intercept, and the term $-x^2/\left(2\sigma_k^2\right)$ would
carry a different coefficient in each class, would not drop out, and would
make each discriminant quadratic in $x$.

The terms dropped from $\log\left[\pi_k f_k(x)\right]$ to form $\delta_k(x)$
are the same in every class, so $e^{\delta_k(x)}$ is $\pi_k f_k(x)$ multiplied
by a factor common to every class. That factor cancels from the ratio in Bayes'
theorem and the posterior probabilities can be recovered from the discriminant
values alone:

$$p_k(x) = \frac{e^{\delta_k(x)}}{\sum_{l=1}^K e^{\delta_l(x)}}.$$

::: {.callout-note collapse="true"}
### Code
```{r delta-coefficients}
#| output: false
delta_1 <- tibble(default = c("No", "Yes"),
                  intercept = log(hand_1$pi) -
                    hand_1$mu[, 1]^2 / (2 * sigma_hat_1^2),
                  slope = hand_1$mu[, 1] / sigma_hat_1^2)
delta_of <- function(k, col) delta_1[[col]][delta_1$default == k]
```
:::

Substituting $\hat\pi_k$, $\hat\mu_k$, and $\hat\sigma^2$ for the credit card
customers gives the estimated discriminants, with balance in dollars,

$$\hat\delta_{\text{No}}(x) = `r sprintf("%.3f", delta_of("No", "intercept"))` + `r sprintf("%.6f", delta_of("No", "slope"))`\, x, \qquad \hat\delta_{\text{Yes}}(x) = `r sprintf("%.3f", delta_of("Yes", "intercept"))` + `r sprintf("%.6f", delta_of("Yes", "slope"))`\, x.$$

The defaulters' line starts far lower, because $\hat\pi_{\text{Yes}}$ is
small and $\hat\mu_{\text{Yes}}$ large, and climbs
`r sprintf("%.1f", delta_of("Yes", "slope") / delta_of("No", "slope"))` times
as fast, because $\hat\mu_{\text{Yes}}$ is the larger mean.

### Decision boundary

With $K = 2$ classes, $p_1(x) = 1 - p_2(x)$, and the log posterior odds of
class $2$ are

$$\log\left(\frac{p_2(x)}{1 - p_2(x)}\right) = \delta_2(x) - \delta_1(x) = \underbrace{\log\frac{\pi_2}{\pi_1} - \frac{\mu_2^2 - \mu_1^2}{2\sigma^2}}_{\beta_0} + \underbrace{\frac{\mu_2 - \mu_1}{\sigma^2}}_{\beta_1}\, x.$$

The log-odds are linear in $x$, so LDA's posterior probability has exactly the
form of a simple logistic regression. The two methods differ in how they
estimate $\beta_0$ and $\beta_1$, not in the shape of the curve they fit.

The **decision boundary** is the set of $x$ where the two classes are equally
probable, $\delta_1(x) = \delta_2(x)$. Setting the log posterior odds to zero
and solving,

$$x^* = \frac{\mu_1 + \mu_2}{2} + \frac{\sigma^2}{\mu_2 - \mu_1}\log\frac{\pi_1}{\pi_2},$$

where

- $x^*$ is the feature value at which the classifier switches from one class
to the other.

With equal prior probabilities the logarithm is zero and the boundary is the
midpoint of the two means. Take $\mu_2 > \mu_1$. If class $2$ is the rarer
class, $\pi_2 < \pi_1$, the logarithm is positive and the boundary moves to the
right, toward $\mu_2$, and past it once
$\sigma^2 \log(\pi_1/\pi_2)/(\mu_2 - \mu_1) > (\mu_2 - \mu_1)/2$: an observation must look more like a typical class-$2$
observation before the classifier assigns it to the rarer class. Since
$\delta_1(x) = \delta_2(x)$ exactly where $\pi_1 f_1(x) = \pi_2 f_2(x)$, the
boundary is also where the two prior-weighted densities cross.

::: {.callout-note collapse="true"}
### Code
```{r balance-boundary}
#| output: false
x_mid <- mean(hand_1$mu[, 1])
x_star_lda <- -hand_1$beta0 / hand_1$beta[1]

# Prior-weighted densities pi_k f_k(x) and discriminant functions delta_k(x)
# for the fitted means and pooled variance, under a given pair of priors.
weighted_curves <- function(priors) {
  tibble(default = factor(c("No", "Yes")), mu = hand_1$mu[, 1],
         prior = priors) |>
    cross_join(balance_grid) |>
    mutate(weighted = prior * dnorm(balance, mu, sigma_hat_1),
           delta = (log(prior) - mu^2 / (2 * sigma_hat_1^2)) +
                   mu / sigma_hat_1^2 * balance)
}

weighted_plot <- function(priors, boundary, title, y_label) {
  ggplot(weighted_curves(priors),
         aes(x = balance, y = weighted, color = default, linetype = default)) +
    geom_line(linewidth = 0.9) +
    geom_vline(xintercept = boundary, linetype = "dotted") +
    scale_color_manual(values = class_colors) +
    scale_x_continuous(labels = dollar_labels) +
    labs(x = "Credit card balance ($)",
         y = y_label,
         color = "Default", linetype = "Default", title = title)
}

equal_priors_plot <- weighted_plot(
  c(0.5, 0.5), x_mid,
  expression(paste("Equal priors, ", pi[No] == pi[Yes], " = 0.5")),
  expression(pi[k] * hat(f)[k](x)))
estimated_priors_plot <- weighted_plot(
  hand_1$pi, x_star_lda,
  bquote(paste("Estimated priors, ",
               hat(pi)[Yes] == .(sprintf("%.4f", hand_1$pi[2])))),
  expression(hat(pi)[k] * hat(f)[k](x)))

delta_default_plot <- ggplot(weighted_curves(hand_1$pi),
                             aes(x = balance, y = delta,
                                 color = default, linetype = default)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = x_star_lda, linetype = "dotted") +
  scale_color_manual(values = class_colors) +
  scale_x_continuous(labels = dollar_labels) +
  labs(x = "Credit card balance ($)", y = expression(hat(delta)[k](x)),
       color = "Default", linetype = "Default",
       title = "Estimated discriminant functions, estimated priors")
```
:::

For the credit card fit, the first tab below plots the prior-weighted fitted
densities with equal priors, a counterfactual in which defaulting is as
common as not, and the second with the estimated priors. The third plots the
two estimated discriminant functions. In every tab a dotted line marks that
tab's decision boundary.

::: {.panel-tabset}
#### Equal priors
```{r balance-equal-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
equal_priors_plot
```

#### Estimated priors
```{r balance-weighted-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
estimated_priors_plot
```

#### Discriminant functions
```{r balance-delta-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
delta_default_plot
```
:::

With equal priors the two curves are the same height and cross at the
midpoint of the class means, \$`r dollars(x_mid)`. With the estimated priors
the defaulter curve shrinks to a small bump that rises above the non-defaulter
curve only past
$\hat x^* = `r gsub(",", "{,}", paste0("\\$", dollars(x_star_lda)))`$, deep in
the non-defaulters' right tail. In the boundary formula that shift is
the prior term
$\hat\sigma^2\log\left(\hat\pi_{\text{No}}/\hat\pi_{\text{Yes}}\right)/\left(\hat\mu_{\text{Yes}} - \hat\mu_{\text{No}}\right)$,
which adds \$`r dollars(x_star_lda - x_mid)` because $\hat\pi_{\text{Yes}}$ is
so small. The boundary lies beyond
$\hat\mu_{\text{Yes}} = `r gsub(",", "{,}", paste0("\\$", dollars(hand_1$mu[2, 1])))`$,
so even a customer with the defaulters' average balance is classified `No`. The discriminant functions are two straight lines crossing at the
same balance. A linear discriminant function for each class, with the
boundary where two of them are equal, is the "linear" in linear discriminant
analysis.

### Simple logistic regression comparison

Because the log posterior odds are linear in $x$, LDA's fit can be set beside
the simple logistic regression of the same customers on the log-odds scale.

::: {.callout-note collapse="true"}
### Code
```{r balance-glm}
#| output: false
balance_glm <- glm(default ~ balance, data = Default, family = "binomial")
b_glm_1     <- coef(balance_glm)
x_star_glm  <- -b_glm_1[1] / b_glm_1[2]

post_1 <- predict(balance_lda)$posterior[, "Yes"]
max_diff_1 <- max(abs(post_1 - plogis(hand_1$beta0 +
                                       hand_1$beta[1] * Default$balance)))

coef_table_1 <- tibble(
  term = c("Intercept $\\hat\\beta_0$", "`balance` $\\hat\\beta_1$"),
  lda = c(hand_1$beta0, hand_1$beta[1]),
  glm = unname(b_glm_1)
) |>
  mutate(across(c(lda, glm), \(v) sprintf("%.6f", v))) |>
  knitr::kable(col.names = c("Coefficient", "LDA", "Logistic regression"),
               align = "lcc")

posterior_curves <- balance_grid |>
  mutate(LDA = plogis(hand_1$beta0 + hand_1$beta[1] * balance),
         `Logistic regression` = predict(balance_glm,
                                         newdata = balance_grid,
                                         type = "response")) |>
  pivot_longer(-balance, names_to = "model", values_to = "p_hat")

posterior_default_plot <- ggplot(posterior_curves,
                                 aes(x = balance, y = p_hat)) +
  geom_point(data = Default, aes(y = default_jit), alpha = 0.08,
             size = 0.7) +
  geom_hline(yintercept = 0.5, color = "grey60") +
  geom_line(aes(color = model, linetype = model), linewidth = 0.9) +
  scale_color_manual(values = model_colors) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_x_continuous(labels = dollar_labels) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Model", linetype = "Model",
       title = "Posterior probability of default")
```
:::

The table gives LDA's $\hat\beta_0$ and $\hat\beta_1$, from the estimates
above, beside the simple logistic regression's coefficients. The figure plots
both models' probability of default against balance, with each customer's
jittered outcome and a grey line at $0.5$.

```{r coef-table-1}
#| echo: false
coef_table_1
```

```{r balance-posterior-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
posterior_default_plot
```

The intercept $\hat\beta_0 = `r sprintf("%.3f", hand_1$beta0)`$ is LDA's
estimated log posterior odds of default for a customer carrying no balance, a
posterior probability of
`r format(signif(plogis(hand_1$beta0), 2), scientific = FALSE)`. Each additional
\$100 of balance adds `r sprintf("%.3f", 100 * hand_1$beta[1])` to the log
posterior odds, against `r sprintf("%.3f", 100 * b_glm_1[2])` in the logistic
regression. Applying the logistic function to
$\hat\beta_0 + \hat\beta_1 x$ reproduces the posterior probabilities returned
by `predict()` to within $`r sci(max_diff_1, 1)`$.

The LDA curve is a logistic curve, slightly flatter than the logistic
regression fit, and it crosses $0.5$ a little further right, at
\$`r dollars(x_star_lda)` against \$`r dollars(x_star_glm)`. Logistic
regression has no priors or class means; its boundary is high because
defaults are rare among customers at all but the largest balances, so its
fitted probability reaches $0.5$ only there.

## Multiple-feature LDA

A second feature turns each customer into a point. The first tab below plots
income against credit card balance for every customer, with defaulters and
non-defaulters distinguished by color and shape; the second shows income
within each default class, one panel per class, on the density scale.

::: {.callout-note collapse="true"}
### Code
```{r default-scatter}
#| output: false
# Defaulters are drawn last so they are not hidden under non-defaulters.
scatter_base <- ggplot(arrange(Default, default),
                       aes(x = balance, y = income)) +
  geom_point(aes(color = default, shape = default), alpha = 0.35,
             size = 0.8) +
  scale_color_manual(values = class_colors) +
  scale_shape_manual(values = c(16, 17)) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(labels = dollar_labels) +
  scale_y_continuous(labels = dollar_labels) +
  labs(x = "Credit card balance ($)", y = "Income ($)", color = "Default",
       shape = "Default")

eda_scatter <- scatter_base +
  labs(title = "ISLR2::Default, income against balance")

income_hist_plot <- ggplot(Default, aes(x = income)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 2500,
                 boundary = 0, fill = "grey80", color = "grey55") +
  facet_wrap(~ default, ncol = 1, labeller = label_both) +
  scale_x_continuous(labels = dollar_labels) +
  labs(x = "Income ($)", y = "Density",
       title = "ISLR2::Default, income within each default class")

within_cor <- Default |>
  group_by(default) |>
  summarize(r = cor(balance, income), mean_income = mean(income),
            .groups = "drop")
within_of <- function(k, col) within_cor[[col]][within_cor$default == k]

income_by_student <- Default |>
  group_by(student) |>
  summarize(median_income = median(income), .groups = "drop")
student_income <- function(s) {
  income_by_student$median_income[income_by_student$student == s]
}
```
:::

::: {.panel-tabset}
#### Scatterplot
```{r default-scatter-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
eda_scatter
```

#### Income histograms
```{r default-income-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
income_hist_plot
```
:::

Each class forms a broad cloud, and the two clouds overlap heavily:
defaulters sit at larger balances, across nearly the whole range of income.
Within each class balance and income are weakly negatively correlated,
$r = `r sprintf("%.3f", within_of("No", "r"))`$ among non-defaulters and
$r = `r sprintf("%.3f", within_of("Yes", "r"))`$ among defaulters, so a
description of each class's cloud needs a center and a spread for each
feature and the correlation between them. The two classes' mean incomes are
close, \$`r dollars(within_of("Yes", "mean_income"))` for defaulters against
\$`r dollars(within_of("No", "mean_income"))`.

The two classes' income histograms look alike, and each is bimodal. The
lower mode is mostly students: in these data students' median income is
\$`r dollars(student_income("Yes"))`, against
\$`r dollars(student_income("No"))` for non-students. No Gaussian density has
two modes, so income strains LDA's assumptions just as balance did.

The clouds suggest the multivariate version of the one-feature model:
describe each class's features by a multivariate Gaussian density with its own
mean vector, the center of its cloud, and a covariance matrix shared by both
classes, which gives the clouds one common spread and tilt.

### Multivariate Gaussian densities

With $p > 1$ features, $X = (X_1, \ldots, X_p)^\top$ is a random column vector
whose entries are the features, and $x$ is a $p$-vector value it takes. A
value $x$ holds the same kind of entries as a row of the model matrix
$\mathbf{X}$ but without the leading $1$, and it is a column because it is the
argument of a multivariate density. LDA assumes the statistical model

$$P(Y = k) = \pi_k, \qquad X \mid Y = k \sim N_p\left(\mu_k, \Sigma\right), \qquad k = 1, \ldots, K,$$

where

- $N_p$ is the $p$-dimensional multivariate normal distribution,
- $\mu_k = E[X \mid Y = k] = (\mu_{k1}, \ldots, \mu_{kp})^\top$ is the
$p$-vector of feature means in class $k$, and
- $\Sigma = Cov[X \mid Y = k]$ is the $p \times p$ covariance matrix of the
features, **the same in every class**, with feature variances on its diagonal
and covariances between pairs of features off it; it is symmetric and
positive definite.

The class-conditional density is therefore the multivariate normal density

$$f_k(x) = \frac{1}{(2\pi)^{p/2}\,|\Sigma|^{1/2}} \exp\left[-\frac{1}{2}\left(x - \mu_k\right)^\top \Sigma^{-1} \left(x - \mu_k\right)\right],$$

where $|\Sigma|$ is the determinant of $\Sigma$.

Each class's density has contours that are nested ellipses centered at
$\mu_k$. Sharing
$\Sigma$ means every class's ellipses have the same shape and orientation and
differ only in their center. For the credit card customers, $\mu_k$ is the
center of class $k$'s cloud in the scatterplot above, and $\Sigma$ gives the
two clouds one common spread and tilt.

### Pooled covariance estimate

The estimates generalize directly. The priors are still $\hat\pi_k = n_k/n$,
and with $x_i = (x_{i1}, \ldots, x_{ip})^\top$ now the $p$-vector of feature
values for observation $i$ (again without the leading $1$),

$$\hat\mu_k = \frac{1}{n_k}\sum_{i:\, y_i = k} x_i, \qquad \hat\Sigma = \frac{1}{n - K}\sum_{k=1}^K \sum_{i:\, y_i = k}\left(x_i - \hat\mu_k\right)\left(x_i - \hat\mu_k\right)^\top,$$

where

- $\hat\mu_k$ is the vector of class-$k$ sample means, one per feature, and
- $\hat\Sigma$ is the **pooled covariance matrix**, the sum of each class's
matrix of cross-products of deviations from its own mean, divided by $n - K$;
as in one dimension, it is the maximum likelihood estimate rescaled by
$n/(n - K)$ to make it unbiased.

Adding `income` to the formula fits the $p = 2$ model to the credit card
customers:

```{r additive-lda}
additive_lda <- MASS::lda(default ~ balance + income, data = Default)
additive_lda
```

::: {.callout-note collapse="true"}
### Code
```{r additive-lda-extract}
#| output: false
X2 <- as.matrix(Default[c("balance", "income")])
hand_2 <- lda_by_hand(X2, Default$default)

scaling_ratio <- additive_lda$scaling[, 1] / hand_2$beta
sd_within <- sqrt(diag(hand_2$Sigma))
cor_within <- hand_2$Sigma[1, 2] / prod(sd_within)

post_lda_2 <- predict(additive_lda)$posterior[, "Yes"]
max_diff_2 <- max(abs(post_lda_2 -
                      plogis(hand_2$beta0 + drop(X2 %*% hand_2$beta))))
```
:::

The prior probabilities are unchanged, since they do not depend on the
features. The group means now form a vector for each class: the `balance`
column repeats the one-feature means, and the `income` column holds the class
mean incomes. The pooled covariance matrix $\hat\Sigma$ has within-class
standard deviations of \$`r dollars(sd_within[1])` for balance and
\$`r dollars(sd_within[2])` for income and a within-class correlation of
$`r sprintf("%.3f", cor_within)`$, a pooled version of the two correlations
quoted with the scatterplot.

### Matrix-form discriminant

The derivation follows the one-feature case. Because $\Sigma^{-1}$ is
symmetric, the scalar $\mu_k^\top \Sigma^{-1} x$ equals its own transpose
$x^\top \Sigma^{-1} \mu_k$, and the quadratic form expands to

$$\left(x - \mu_k\right)^\top \Sigma^{-1} \left(x - \mu_k\right) = x^\top \Sigma^{-1} x - 2\, x^\top \Sigma^{-1} \mu_k + \mu_k^\top \Sigma^{-1} \mu_k.$$

Then, collecting terms in the same order as before,

$$\log\left[\pi_k f_k(x)\right] = \log \pi_k - \frac{1}{2}\mu_k^\top \Sigma^{-1} \mu_k + x^\top \Sigma^{-1} \mu_k - \frac{p}{2}\log(2\pi) - \frac{1}{2}\log|\Sigma| - \frac{1}{2}x^\top \Sigma^{-1} x.$$

The last three terms do not involve $k$: the first is a constant, and the
other two because $\Sigma$ is shared. Dropping them gives

$$\delta_k(x) = \left(\log \pi_k - \frac{1}{2}\mu_k^\top \Sigma^{-1} \mu_k\right) + x^\top \Sigma^{-1} \mu_k,$$

where

- $\log \pi_k - \frac{1}{2}\mu_k^\top \Sigma^{-1} \mu_k$ is the intercept of
class $k$'s discriminant, and
- $\Sigma^{-1}\mu_k$ is its $p$-vector of slopes, one per feature.

The dropped term $-\frac{1}{2}x^\top \Sigma^{-1} x$ was the only one quadratic
in $x$, so the discriminant is linear in $x$ and reduces to the one-feature discriminant
when $\Sigma = \sigma^2$. As before, the classifier assigns the class with the
largest $\delta_k(x)$, and $p_k(x) = e^{\delta_k(x)} / \sum_l e^{\delta_l(x)}$.

::: {.callout-note collapse="true"}
### Code
```{r delta-coefficients-2}
#| output: false
Sigma2_inv <- solve(hand_2$Sigma)
# Each class's estimated discriminant: intercept and slopes Sigma^{-1} mu_k.
delta_row_2 <- function(k) {
  slopes <- drop(Sigma2_inv %*% hand_2$mu[k, ])
  tibble(default = k,
         intercept = log(hand_2$pi[rownames(hand_2$mu) == k]) -
           0.5 * sum(hand_2$mu[k, ] * slopes),
         balance = slopes[1], income = 1000 * slopes[2])
}
delta_table_2 <- map(c("No", "Yes"), delta_row_2) |> list_rbind()
delta_kable_2 <- delta_table_2 |>
  mutate(intercept = sprintf("%.3f", intercept),
         balance = sprintf("%.6f", balance),
         income = sprintf("%.6f", income)) |>
  knitr::kable(col.names = c("Default", "Intercept", "`balance` slope",
                             "`income` slope, per \\$1,000"),
               align = "lccc")
```
:::

For the credit card fit, the table gives each class's estimated
discriminant, $\hat\delta_k(x)$, with income slopes per \$1,000 of income.

```{r delta-table-2}
#| echo: false
delta_kable_2
```

As in one dimension, the defaulters' discriminant starts far lower and climbs
faster in balance. Its income slope is also the larger of the two, so at a
fixed balance a higher income moves the comparison slightly toward `Yes`.

### Linear decision boundary

With $K = 2$, subtracting the two discriminants gives the log posterior odds
of class $2$,

$$\log\left(\frac{p_2(x)}{1 - p_2(x)}\right) = \delta_2(x) - \delta_1(x) = \beta_0 + x^\top\beta,$$

where

- $\beta = \Sigma^{-1}(\mu_2 - \mu_1)$ is a $p$-vector of slopes, one per
feature, with no intercept entry, and
- $\beta_0 = \log(\pi_2/\pi_1) - \frac{1}{2}\left(\mu_2^\top \Sigma^{-1} \mu_2 - \mu_1^\top \Sigma^{-1} \mu_1\right)$
is the intercept.

This is again the form of a multiple logistic regression with $p$ features.
The decision boundary $\{x : \beta_0 + x^\top\beta = 0\}$ is a hyperplane in
the $p$-dimensional feature space — a line when $p = 2$.

The slopes are $\Sigma^{-1}(\mu_2 - \mu_1)$, not $\mu_2 - \mu_1$, so the
boundary is generally not perpendicular to the segment joining the two means:
$\Sigma^{-1}$ tilts it according to how the features covary. With equal
priors it still passes through the midpoint $(\mu_1 + \mu_2)/2$, since
substituting that point gives
$\frac{1}{2}\left(\mu_1 + \mu_2\right)^\top\Sigma^{-1}\left(\mu_2 - \mu_1\right) = \frac{1}{2}\left(\mu_2^\top \Sigma^{-1} \mu_2 - \mu_1^\top \Sigma^{-1} \mu_1\right)$.
The boundary is where $\pi_1 f_1(x) = \pi_2 f_2(x)$, so it passes through
every point where a contour of $\pi_1 f_1$ crosses the contour of $\pi_2 f_2$
of the same height.

::: {.callout-note collapse="true"}
### Code
```{r additive-contours}
#| output: false
mu_No  <- hand_2$mu["No", ]
mu_Yes <- hand_2$mu["Yes", ]
beta_2 <- hand_2$beta
f_peak_2 <- 1 / (2 * pi * sqrt(det(hand_2$Sigma)))
dens_2 <- function(b, i, mu) {
  d <- cbind(b - mu[1], i - mu[2])
  f_peak_2 * exp(-rowSums((d %*% Sigma2_inv) * d) / 2)
}

window_2 <- list(balance = c(0, 2700), income = c(0, 75000))
grid_2 <- expand_grid(balance = seq(0, 2700, length.out = 240),
                      income = seq(0, 75000, length.out = 220)) |>
  mutate(f_No = dens_2(balance, income, mu_No),
         f_Yes = dens_2(balance, income, mu_Yes))

# Long grid of pi_k * f_k(x); prior weights of 1 give the unweighted f_k.
weighted_grid_2 <- function(p) {
  grid_2 |>
    transmute(balance, income, No = p[1] * f_No, Yes = p[2] * f_Yes) |>
    pivot_longer(c(No, Yes), names_to = "default", values_to = "density")
}

# Fitted densities: contours where (x - mu_k)' Sigma^{-1} (x - mu_k) is 1 and
# 4. Prior-weighted
# densities: the same three heights, as fractions of the peak of f_k, in both
# prior tabs.
density_levels_2  <- f_peak_2 * exp(-c(1, 2)^2 / 2)
weighted_levels_2 <- f_peak_2 * c(0.01, 0.02, 0.03)

beta0_2 <- function(p) {
  log(p[2] / p[1]) -
    0.5 * (sum(mu_Yes * (Sigma2_inv %*% mu_Yes)) -
           sum(mu_No * (Sigma2_inv %*% mu_No)))
}
# beta0 + beta1 balance + beta2 income = 0 => income = intercept + slope balance
line_2 <- function(p) {
  c(intercept = unname(-beta0_2(p) / beta_2[2]),
    slope = unname(-beta_2[1] / beta_2[2]))
}

# Where the same-height contours of the two classes cross: where the boundary
# line, on which pi_1 f_1 = pi_2 f_2, meets the class-1 contour at that height.
# Along the line x = p0 + s u, log(pi_1 f_1(x)) is quadratic in s. Only
# crossings inside the plotted window are kept.
crossings_2 <- function(p, levels) {
  p0 <- -beta0_2(p) * beta_2 / sum(beta_2^2)
  u  <- c(beta_2[2], -beta_2[1])
  d0 <- p0 - mu_No
  a  <- 0.5 * sum(u * (Sigma2_inv %*% u))
  b  <- sum(u * (Sigma2_inv %*% d0))
  map(levels, \(L) {
    c0 <- 0.5 * sum(d0 * (Sigma2_inv %*% d0)) - log(p[1] * f_peak_2 / L)
    disc <- b^2 - 4 * a * c0
    if (disc < 0) return(NULL)
    s <- (-b + c(-1, 1) * sqrt(disc)) / (2 * a)
    tibble(balance = p0[1] + s * u[1], income = p0[2] + s * u[2])
  }) |>
    list_rbind() |>
    filter(between(balance, window_2$balance[1], window_2$balance[2]),
           between(income, window_2$income[1], window_2$income[2]))
}

means_2 <- tibble(balance = c(mu_No[1], mu_Yes[1]),
                  income = c(mu_No[2], mu_Yes[2]))

contour_plot_2 <- function(p, levels) {
  ggplot(weighted_grid_2(p), aes(x = balance, y = income)) +
    geom_point(data = Default, aes(color = default, shape = default),
               alpha = 0.12, size = 0.6) +
    geom_contour(aes(z = density, color = default, linetype = default),
                 breaks = levels, linewidth = 0.8) +
    geom_line(data = means_2, color = "grey40", linewidth = 0.5) +
    geom_point(data = means_2, shape = 4, size = 4, stroke = 1.5) +
    scale_color_manual(values = class_colors) +
    scale_shape_manual(values = c(16, 17)) +
    scale_x_continuous(labels = dollar_labels) +
    scale_y_continuous(labels = dollar_labels) +
    coord_cartesian(xlim = window_2$balance, ylim = window_2$income) +
    guides(color = guide_legend(override.aes = list(alpha = 1))) +
    labs(x = "Credit card balance ($)", y = "Income ($)", color = "Default",
         shape = "Default", linetype = "Default")
}
# The boundary uses "longdash", distinct from both class-contour linetypes.
boundary_line_2 <- function(p, ...) {
  l <- line_2(p)
  geom_abline(intercept = l["intercept"], slope = l["slope"], ...)
}
crossing_points_2 <- function(p) {
  geom_point(data = crossings_2(p, weighted_levels_2),
             aes(x = balance, y = income), inherit.aes = FALSE,
             shape = 21, size = 3, stroke = 1, fill = "white", color = "black")
}

contour_densities_2 <- contour_plot_2(c(1, 1), density_levels_2) +
  labs(title = expression(paste("Contours of ", hat(f)[k](x))))
contour_equal_2 <- contour_plot_2(c(0.5, 0.5), weighted_levels_2) +
  boundary_line_2(c(0.5, 0.5), linewidth = 1, linetype = "longdash") +
  crossing_points_2(c(0.5, 0.5)) +
  labs(title = expression(paste("Contours of ", pi[k] * hat(f)[k](x), "; ",
                                pi[No] == pi[Yes], " = 0.5")))
contour_estimated_2 <- contour_plot_2(hand_2$pi, weighted_levels_2) +
  boundary_line_2(c(0.5, 0.5), linewidth = 0.6, color = "grey50",
                  linetype = "dotted") +
  boundary_line_2(hand_2$pi, linewidth = 1, linetype = "longdash") +
  crossing_points_2(hand_2$pi) +
  labs(title = bquote(paste("Contours of ", hat(pi)[k] * hat(f)[k](x), "; ",
                            hat(pi)[Yes] == .(sprintf("%.4f", hand_2$pi[2])))))

# Read as a balance threshold at a given income, and over the whole observed
# range of income; and the change in the log posterior odds per standard
# deviation of each feature.
threshold_at <- function(b0, b1, b2, inc) -(b0 + b2 * inc) / b1
x_star_lda_range <- threshold_at(hand_2$beta0, hand_2$beta[1],
                                 hand_2$beta[2], range(Default$income))
per_sd_lda <- hand_2$beta * c(sd(Default$balance), sd(Default$income))
yes_peak_share <- hand_2$pi[2]
```
:::

For the credit card fit, each tab below draws faint customer points, the two
estimated class means (crosses) joined by a grey segment, and contours: the
first tab draws two contours of each class's fitted density $\hat f_k$, the
ellipses on which
$\left(x - \hat\mu_k\right)^\top \hat\Sigma^{-1} \left(x - \hat\mu_k\right)$
equals $1$ and $4$, and the second and third draw contours of the
prior-weighted densities at the same three heights,
$\pi_k \hat f_k$ with equal priors and $\hat\pi_k \hat f_k$ with the
estimated priors, together with that tab's decision boundary (long dashes).
Open circles mark where a non-defaulter contour crosses the defaulter
contour of the same height, and the dotted line in the third tab is the equal-prior boundary.

::: {.panel-tabset}
#### Fitted densities
```{r additive-densities-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
contour_densities_2
```

#### Equal priors
```{r additive-equal-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
contour_equal_2
```

#### Estimated priors
```{r additive-estimated-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
contour_estimated_2
```
:::

The two sets of ellipses in the first tab have the same shape and tilt,
because $\hat\Sigma$ is shared, and they tilt only slightly, as the weak
negative correlation implies. With equal priors the
boundary passes through the midpoint of the grey segment, as the algebra
requires, and through every circled crossing. With the estimated priors the
defaulter surface is scaled by $\hat\pi_{\text{Yes}} =
`r sprintf("%.4f", yes_peak_share)`$, so its contours shrink to small
ellipses around the defaulters' mean, and the boundary, still through the
circled crossings, moves parallel to the equal-prior line toward the
defaulters, because the prior enters only $\beta_0$.

The estimated boundary is nearly vertical: across the whole observed range of
income, LDA's balance threshold shifts by only
\$`r dollars(abs(diff(x_star_lda_range)))`, against a range of balance of
\$`r dollars(diff(range(Default$balance)))`.

### Multiple logistic regression comparison {#balance-and-income}

The log posterior odds $\hat\beta_0 + x^\top\hat\beta$ can again be set beside
a logistic regression with the same two features.

::: {.callout-note collapse="true"}
### Code
```{r additive-glm}
#| output: false
additive_glm <- glm(default ~ balance + income, data = Default,
                    family = "binomial")
b_glm_2 <- coef(additive_glm)
z_income_glm <- summary(additive_glm)$coefficients["income", "z value"]
post_glm_2 <- fitted(additive_glm)

# Income difference between the classes that their balance difference
# predicts through the pooled within-class regression of income on balance.
mean_diff <- hand_2$mu["Yes", ] - hand_2$mu["No", ]
income_diff_predicted <- hand_2$Sigma[1, 2] / hand_2$Sigma[1, 1] * mean_diff[1]

# Income coefficients are reported per $1,000 of income.
coef_table_2 <- tibble(
  term = c("Intercept $\\hat\\beta_0$", "`balance` $\\hat\\beta_1$",
           "`income` $\\hat\\beta_2$, per \\$1,000"),
  lda = c(hand_2$beta0, hand_2$beta[1], 1000 * hand_2$beta[2]),
  glm = c(b_glm_2[1], b_glm_2[2], 1000 * b_glm_2[3])
) |>
  mutate(across(c(lda, glm), \(v) sprintf("%.6f", v))) |>
  knitr::kable(col.names = c("Coefficient", "LDA", "Logistic regression"),
               align = "lcc")

income_levels <- quantile(Default$income, c(0.25, 0.5, 0.75))
income_level_names <- c("lower quartile", "median", "upper quartile")
x_star_lda_2 <- threshold_at(hand_2$beta0, hand_2$beta[1], hand_2$beta[2],
                             income_levels)
x_star_glm_2 <- threshold_at(b_glm_2[1], b_glm_2[2], b_glm_2[3],
                             income_levels)

boundary_lines <- tibble(
  model = c("LDA", "Logistic regression"),
  intercept = c(-hand_2$beta0 / hand_2$beta[2], -b_glm_2[1] / b_glm_2[3]),
  slope = c(-hand_2$beta[1] / hand_2$beta[2], -b_glm_2[2] / b_glm_2[3])
)

# Boundary lines take the models' colors as fixed colors, because the color
# scale already encodes default; linetype carries the legend.
boundary_layers <- function(lines) {
  list(
    geom_abline(data = filter(lines, model == "LDA"),
                aes(intercept = intercept, slope = slope, linetype = model),
                color = model_colors[["LDA"]], linewidth = 1),
    geom_abline(data = filter(lines, model == "Logistic regression"),
                aes(intercept = intercept, slope = slope, linetype = model),
                color = model_colors[["Logistic regression"]], linewidth = 1),
    scale_linetype_manual(values = c("solid", "dashed")),
    guides(linetype = guide_legend(
      override.aes = list(color = unname(model_colors))))
  )
}

boundary_plot <- scatter_base +
  boundary_layers(boundary_lines) +
  coord_cartesian(xlim = range(Default$balance),
                  ylim = range(Default$income)) +
  labs(linetype = "Boundary",
       title = "Decision boundaries in the (balance, income) plane")

# Posterior probability against balance at three fixed incomes.
income_facet <- function(v) {
  factor(v, levels = unname(income_levels),
         labels = sprintf("Income $%s (%s)", dollars(income_levels),
                          income_level_names))
}
posterior_grid_2 <- expand_grid(balance = seq(0, 2800, length.out = 400),
                                income = unname(income_levels))
posterior_curves_2 <- posterior_grid_2 |>
  mutate(LDA = predict(additive_lda, newdata = posterior_grid_2)$posterior[, "Yes"],
         `Logistic regression` = predict(additive_glm,
                                         newdata = posterior_grid_2,
                                         type = "response"),
         income_label = income_facet(income)) |>
  pivot_longer(c(LDA, `Logistic regression`), names_to = "model",
               values_to = "p_hat")

posterior_plot_2 <- ggplot(posterior_curves_2, aes(x = balance, y = p_hat)) +
  geom_hline(yintercept = 0.5, color = "grey60") +
  geom_line(aes(color = model, linetype = model), linewidth = 0.9) +
  facet_wrap(~ income_label, nrow = 1) +
  scale_color_manual(values = model_colors) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_x_continuous(labels = dollar_labels) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Model", linetype = "Model",
       title = "Posterior probability of default at three incomes")
```
:::

The table gives LDA's $\hat\beta_0$ and $\hat\beta$ beside the coefficients of
the logistic regression with the same two features. The first tab below draws
each model's decision boundary over the scatterplot of income against
balance, LDA's as a solid green line and logistic regression's as a dashed
pink one. The second plots each model's posterior probability of default
against balance at the lower quartile, median, and upper quartile of income,
with a grey line at $0.5$.

```{r coef-table-2}
#| echo: false
coef_table_2
```

::: {.panel-tabset}
#### Decision boundary
```{r additive-boundary-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
boundary_plot
```

#### Posterior
```{r additive-posterior-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 3.4
posterior_plot_2
```
:::

The intercept $\hat\beta_0 = `r sprintf("%.3f", hand_2$beta0)`$ is the log
posterior odds of default for a customer with no balance and no income, an
extrapolation, since the lowest recorded income is
\$`r dollars(min(Default$income))`. The `balance` slope plays the same role as
before. The `income` slope,
$`r sprintf("%.4f", 1000 * hand_2$beta[2])`$ per \$1,000, is positive: at a
fixed balance, LDA assigns a slightly higher posterior probability of default
to a customer with a higher income, although defaulters' mean income is the
lower of the two. Multiplying by $\hat\Sigma^{-1}$ adjusts each feature for
the other. With the pooled within-class correlation of
$`r sprintf("%.3f", cor_within)`$, defaulters' higher balances predict a mean
income \$`r dollars(-income_diff_predicted)` below non-defaulters', but it is
only \$`r dollars(-mean_diff[2])` below; the positive slope reflects that
excess. Logistic regression's `income` coefficient,
$`r sprintf("%.4f", 1000 * b_glm_2[3])`$ per \$1,000, is also positive, with
$z = `r sprintf("%.2f", z_income_glm)`$; it too compares customers with the
same balance. Applying the logistic function to
$\hat\beta_0 + x^\top\hat\beta$ again reproduces `predict()`'s posterior
probabilities, to within $`r sci(max_diff_2, 1)`$.

Per overall standard deviation, balance moves LDA's log posterior odds by
`r sprintf("%.2f", per_sd_lda[1])` and income by only
`r sprintf("%.2f", per_sd_lda[2])`. The coefficients of the linear
discriminant printed by `MASS::lda()` are proportional to the LDA column of
the table: dividing each printed coefficient by the corresponding entry of
$\hat\beta = \hat\Sigma^{-1}\left(\hat\mu_{\text{Yes}} - \hat\mu_{\text{No}}\right)$
gives `r sprintf("%.4f", scaling_ratio[1])` for both.

Read as a balance threshold, LDA's boundary sits at
\$`r dollars(x_star_lda_2[1])` for a customer at the lower quartile of
income and \$`r dollars(x_star_lda_2[3])` at the upper quartile; logistic
regression's sits at \$`r dollars(x_star_glm_2[1])` and
\$`r dollars(x_star_glm_2[3])`. In both models the threshold falls as income
rises, and logistic regression's lies to the left of LDA's. At each income
the LDA posterior curve is flatter than the logistic regression curve and
crosses $0.5$ further right.

## Classifier evaluation

An error rate counts every mistake the same way. Failing to flag a customer
who goes on to default and flagging a customer who does not are different
mistakes with different costs, and a classifier can have a small error rate
while making almost every mistake of one kind. The classification lecture
introduced the confusion matrix, which separates the mistakes; this section
develops it in full for two classes and for $K$ classes, and adds the
ROC curve and AUC, which summarize a two-class classifier at every threshold.
The running example is the LDA classifier with `balance` and `income`.

The figure shows, for each true default class, the histogram of the
`balance` and `income` LDA classifier's log posterior odds of default, on the
density scale, with a dotted line at $0$, where the classifier's estimated
probability of default, $\hat p(x)$, equals $0.5$.

::: {.callout-note collapse="true"}
### Code
```{r score-hist}
#| output: false
# Log posterior odds of default from the balance + income LDA, binned once so
# this figure and the threshold slider below share the same bars.
lda_score <- qlogis(post_lda_2)
score_breaks <- seq(floor(min(lda_score)), ceiling(max(lda_score)), by = 0.5)
class_label <- c(No = "No (negative)", Yes = "Yes (positive)")
slider_bins <- map(c("No", "Yes"), \(k) {
  h <- hist(lda_score[Default$default == k], breaks = score_breaks,
            plot = FALSE)
  tibble(class = class_label[[k]], x0 = head(h$breaks, -1),
         x1 = tail(h$breaks, -1), density = h$density)
}) |>
  list_rbind()

score_by_class <- tibble(score = lda_score, default = Default$default) |>
  group_by(default) |>
  summarize(median = median(score), below_0 = mean(score <= 0),
            above_0 = mean(score > 0), .groups = "drop")
score_of <- function(k, col) score_by_class[[col]][score_by_class$default == k]

score_hist_plot <- tibble(score = lda_score, default = Default$default) |>
  ggplot(aes(x = score)) +
  geom_histogram(aes(y = after_stat(density)), breaks = score_breaks,
                 fill = "grey80", color = "grey55") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  facet_wrap(~ default, ncol = 1, labeller = label_both) +
  labs(x = "LDA log posterior odds of default", y = "Density",
       title = "LDA (balance + income) scores within each true class")
```
:::

```{r score-hist-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
score_hist_plot
```

The defaulters' scores sit well to the right of the non-defaulters' (medians
`r sprintf("%.2f", score_of("Yes", "median"))` and
`r sprintf("%.2f", score_of("No", "median"))`), but the two histograms
overlap: `r sprintf("%.1f", 100 * score_of("Yes", "below_0"))`% of defaulters
score at or below $0$, and `r sprintf("%.1f", 100 * score_of("No", "above_0"))`%
of non-defaulters above it. Any threshold inside the overlap makes two kinds
of mistake, defaulters to its left and non-defaulters to its right.

### Confusion matrix

With two classes, call the class of interest **positive** and the other
**negative**. For a classifier that produces an estimated probability
$\hat p(x)$ of the positive class, a **threshold** $t$ turns probabilities into
classes: predict positive when $\hat p(x) > t$.

As in the classification lecture, a confusion matrix cross-tabulates the true
class against the predicted class:

::: {.callout-note collapse="true"}
### Code
```{r confusion-template}
#| output: false
confusion_template <- tibble(
  Truth = c("Positive", "Negative"),
  `Predicted positive` = c("True positive (TP)", "False positive (FP)"),
  `Predicted negative` = c("False negative (FN)", "True negative (TN)")
) |>
  knitr::kable(align = "lcc")
```
:::

```{r confusion-template-table}
#| echo: false
confusion_template
```

where

- $TP$ is the number of positive observations predicted positive,
- $FN$ is the number of positive observations predicted negative,
- $FP$ is the number of negative observations predicted positive, and
- $TN$ is the number of negative observations predicted negative.

In the score histograms above, with $t = 0.5$, $TP$ and $FN$ count the defaulters to
the right and left of the dotted line, and $FP$ and $TN$ the non-defaulters
to its right and left. The diagonal holds the correct classifications, so the
error rate is $(FP + FN)/n$.

Classifying each credit card customer as a defaulter when $\hat p(x) > 0.5$,
the LDA classifier with `balance` and `income` produces this confusion matrix:

::: {.callout-note collapse="true"}
### Code
```{r confusion-lda}
#| output: false
confusion_counts <- function(p_hat, t) {
  truth <- Default$default == "Yes"
  pred  <- p_hat > t
  c(TP = sum(truth & pred), FN = sum(truth & !pred),
    FP = sum(!truth & pred), TN = sum(!truth & !pred))
}

cm_lda_05 <- confusion_counts(post_lda_2, 0.5)
# A lower threshold, compared with 0.5 below.
t_low <- 0.2

confusion_lda_table <- tibble(
  Truth = c("Yes", "No"),
  `Predicted Yes` = c(cm_lda_05["TP"], cm_lda_05["FP"]),
  `Predicted No`  = c(cm_lda_05["FN"], cm_lda_05["TN"])
) |>
  mutate(Total = `Predicted Yes` + `Predicted No`) |>
  mutate(across(-Truth, \(v) format(v, big.mark = ","))) |>
  knitr::kable(align = "lccc")

metrics <- function(cm) {
  c(error = unname((cm["FP"] + cm["FN"]) / sum(cm)),
    sensitivity = unname(cm["TP"] / (cm["TP"] + cm["FN"])),
    specificity = unname(cm["TN"] / (cm["TN"] + cm["FP"])))
}
m_lda_05 <- metrics(cm_lda_05)
```
:::

```{r confusion-lda-table}
#| echo: false
confusion_lda_table
```

The classifier misclassifies
`r cm_lda_05["FP"] + cm_lda_05["FN"]` of
`r format(sum(cm_lda_05), big.mark = ",")` customers, an error rate of
`r sprintf("%.2f", 100 * m_lda_05["error"])`%, against the
`r sprintf("%.2f", 100 * hand_1$pi[2])`% of the classifier that predicts `No`
for everyone. The error rate is small because the
`r format(cm_lda_05["TN"] + cm_lda_05["FP"], big.mark = ",")` non-defaulters,
nearly all classified correctly, dominate it, even though
`r cm_lda_05["FN"]` of the `r cm_lda_05["FP"] + cm_lda_05["FN"]` errors are
missed defaulters.

These are training error rates: the customers being classified are the ones
the model was fit to, so they are optimistic estimates of performance on new
customers, as the classification lecture discussed. Estimating test error from
the training data alone is the subject of the resampling unit.

### Sensitivity and specificity

Dividing each row by its total gives two rates that condition on the true
class:

$$\text{Sensitivity} = \frac{TP}{TP + FN}, \qquad \text{Specificity} = \frac{TN}{TN + FP},$$

where

- **sensitivity**, also called **recall** or the **true positive rate** (TPR), is the
proportion of positive observations the classifier catches, an estimate of
$P\left(\hat y(X) = \text{positive} \mid Y = \text{positive}\right)$, and
- **specificity**, also called the **true negative rate**, is the proportion of
negative observations the classifier correctly leaves alone, an estimate of
$P\left(\hat y(X) = \text{negative} \mid Y = \text{negative}\right)$.

Their complements are error rates within each true class:
$1 - \text{sensitivity} = FN/(TP + FN)$ is the **false negative rate**, and
$1 - \text{specificity} = FP/(TN + FP)$ is the **false positive rate** (FPR).

In the credit card confusion matrix, of the
`r cm_lda_05["TP"] + cm_lda_05["FN"]` customers who defaulted the classifier
flags only `r cm_lda_05["TP"]`, a sensitivity of
`r sprintf("%.1f", 100 * m_lda_05["sensitivity"])`%. Of the
`r format(cm_lda_05["TN"] + cm_lda_05["FP"], big.mark = ",")` who did not, it
correctly leaves alone `r format(cm_lda_05["TN"], big.mark = ",")`, a
specificity of `r sprintf("%.1f", 100 * m_lda_05["specificity"])`%.

### Multiple classes

With $K$ classes the confusion matrix is $K \times K$, with true classes in
the rows and predicted classes in the columns:

::: {.callout-note collapse="true"}
### Code
```{r confusion-template-k}
#| output: false
k_labels <- c("1", "2", "$\\vdots$", "$K$")
cell <- function(k, j) {
  ifelse(k == "$\\vdots$" | j == "$\\vdots$", "$\\vdots$",
         sprintf("$n_{%s%s}$", gsub("\\$", "", k), gsub("\\$", "", j)))
}
confusion_template_k <- tibble(Truth = k_labels) |>
  mutate(`Predicted 1` = cell(Truth, "1"),
         `Predicted 2` = cell(Truth, "2"),
         `$\\cdots$`   = ifelse(Truth == "$\\vdots$", "$\\ddots$", "$\\cdots$"),
         `Predicted $K$` = cell(Truth, "$K$")) |>
  knitr::kable(align = "lcccc", escape = FALSE)
```
:::

```{r confusion-template-k-table}
#| echo: false
confusion_template_k
```

where

- $n_{kj}$ is the number of observations in true class $k$ predicted to be in
class $j$, so the diagonal entries $n_{kk}$ are the correct classifications.

The error rate is $1 - \sum_{k=1}^K n_{kk} / n$. Sensitivity and specificity
are defined one class at a time, **one versus the rest**: treat class $k$ as
positive and the other $K - 1$ classes together as negative, which collapses
the table to two-by-two. Class $k$'s rates are

$$\text{Sensitivity}_k = \frac{n_{kk}}{\sum_{j=1}^K n_{kj}}, \qquad \text{Specificity}_k = \frac{\sum_{l \ne k} \sum_{j \ne k} n_{lj}}{\sum_{l \ne k} \sum_{j=1}^K n_{lj}},$$

where

- the sensitivity's denominator is the number of observations in class $k$,
and its numerator the number of those predicted to be in class $k$, and
- the specificity's denominator is the number of observations in the other
classes, and its numerator the number of those predicted to be in some class
other than $k$, whether or not it is their own.

With $K$ classes the threshold $t = 0.5$ generalizes to assigning each
observation to the class with the largest estimated posterior probability;
with $K = 2$ the two rules coincide. Other thresholds, and the ROC curve and
AUC built from them, are two-class tools, and the rest of this section
returns to two classes.

::: {.callout-note collapse="true"}
### Beyond this course
ROC curves and AUC extend to $K > 2$ classes by averaging the AUCs of the $K$
one-versus-rest problems, or of all $K(K-1)/2$ pairs of classes. Logistic
regression as developed in this course is also a two-class method; its
multi-class extension, multinomial logistic regression, is not covered in
this course.
:::

### Classification threshold {#threshold-choice}

The threshold $t = 0.5$ is the Bayes classifier's: if $\hat p(x)$ equalled
the true $p_2(x)$, the posterior probability of the positive class, it would
minimize the expected error rate. An estimated
$\hat p(x)$ carries no such guarantee, and even for the Bayes classifier the
overall error rate need not be the quantity of interest.

Lowering $t$ enlarges the set $\left\{x : \hat p(x) > t\right\}$ of feature
values classified positive, so every
observation predicted positive at the old threshold is still predicted
positive at the new one. Consequently $TP$ and $FP$ can only grow while $FN$
and $TN$ can only shrink: sensitivity rises or stays the same, and
specificity falls or stays the same. Raising $t$ does the reverse, so no
threshold improves both.

::: {.callout-note collapse="true"}
### Beyond this course
If the two mistakes have known costs, $c_{FN}$ for a missed positive and
$c_{FP}$ for a false alarm, the expected cost of predicting positive at $x$ is
$c_{FP}\left[1 - p_2(x)\right]$ and of predicting negative is $c_{FN}\, p_2(x)$.
Predicting positive is the cheaper choice exactly when
$p_2(x) > c_{FP}/(c_{FP} + c_{FN})$, so that ratio is the cost-minimizing
threshold. Equal costs recover $t = 0.5$.
:::

A credit card company that wants to catch more of its eventual defaulters can
lower the threshold. The figure below tracks four rates of the LDA classifier
with `balance` and `income` as $t$ moves from $0$ to $0.5$ — the overall
error rate, sensitivity (the true positive rate), specificity (the true
negative rate), and the false positive rate ($1 - \text{specificity}$) — with
dotted lines at $t = `r t_low`$ and $t = 0.5$. The checkboxes choose which
rates are drawn; the false positive rate starts unchecked because it mirrors
specificity. The table below the figure gives the confusion-matrix counts and
rates at $t = 0.5$ and $t = `r t_low`$ for LDA with `balance` alone, LDA with
`balance` and `income`, and the logistic regression with `balance` and
`income`.

::: {.callout-note collapse="true"}
### Code
```{r threshold-sweep}
#| output: false
threshold_rates <- tibble(t = seq(0, 0.5, by = 0.005)) |>
  mutate(m = map(t, \(tt) metrics(confusion_counts(post_lda_2, tt))),
         `Overall error rate` = map_dbl(m, "error"),
         `Sensitivity (true positive rate)` = map_dbl(m, "sensitivity"),
         `Specificity (true negative rate)` = map_dbl(m, "specificity"),
         `False positive rate (1 - specificity)` =
           map_dbl(m, \(v) 1 - v[["specificity"]])) |>
  select(-m) |>
  pivot_longer(-t, names_to = "rate", values_to = "value")

# Training error rate at every distinct posterior probability used as t;
# the error rate only changes when t passes an observation's score.
min_error_threshold <- function(p_hat) {
  cand <- sort(unique(p_hat))
  err  <- map_dbl(cand, \(tt) metrics(confusion_counts(p_hat, tt))[["error"]])
  c(t = cand[which.min(err)], error = min(err))
}
min_err_lda <- min_error_threshold(post_lda_2)
min_err_glm <- min_error_threshold(post_glm_2)

cm_lda1_05  <- confusion_counts(post_1, 0.5)
cm_lda1_low <- confusion_counts(post_1, t_low)
cm_lda_low  <- confusion_counts(post_lda_2, t_low)
cm_glm_05   <- confusion_counts(post_glm_2, 0.5)
cm_glm_low  <- confusion_counts(post_glm_2, t_low)
m_lda1_05   <- metrics(cm_lda1_05)
m_lda1_low  <- metrics(cm_lda1_low)
m_lda_low   <- metrics(cm_lda_low)
m_glm_05    <- metrics(cm_glm_05)
m_glm_low   <- metrics(cm_glm_low)

compare_table <- tibble(
  Model = rep(c("LDA, balance", "LDA, balance + income",
                "Logistic regression, balance + income"), each = 2),
  t = rep(c(0.5, t_low), 3),
  cm = list(cm_lda1_05, cm_lda1_low, cm_lda_05, cm_lda_low,
            cm_glm_05, cm_glm_low)
) |>
  mutate(TP = map_dbl(cm, "TP"), FN = map_dbl(cm, "FN"),
         FP = map_dbl(cm, "FP"), TN = map_dbl(cm, "TN"),
         m = map(cm, metrics),
         error = map_dbl(m, "error"),
         sensitivity = map_dbl(m, "sensitivity"),
         specificity = map_dbl(m, "specificity")) |>
  select(-cm, -m) |>
  mutate(across(c(TP, FN, FP, TN), \(v) format(v, big.mark = ",")),
         across(c(error, sensitivity, specificity),
                \(v) sprintf("%.4f", v))) |>
  knitr::kable(col.names = c("Model", "Threshold", "TP", "FN", "FP", "TN",
                             "Error rate", "Sensitivity", "Specificity"),
               align = "lcccccccc")
```
:::

```{r threshold-data}
#| echo: false
ojs_define(threshold_rates = threshold_rates,
           threshold_marks = c(t_low, 0.5))
```

```{ojs}
//| label: threshold-checkboxes
//| echo: false
rateNames = ["Overall error rate", "Sensitivity (true positive rate)",
             "Specificity (true negative rate)",
             "False positive rate (1 - specificity)"]
rateStyles = ({
  "Overall error rate": {stroke: "black", dash: "none"},
  "Sensitivity (true positive rate)": {stroke: "#D55E00", dash: "8,4"},
  "Specificity (true negative rate)": {stroke: "#0072B2", dash: "2,4"},
  "False positive rate (1 - specificity)": {stroke: "#009E73", dash: "12,4,2,4"}
})
viewof shownRates = Inputs.checkbox(rateNames, {
  value: rateNames.slice(0, 3),
  format: name => html`<span style="display: inline-flex; align-items: center; gap: 0.4em">
    <svg width="36" height="10"><line x1="2" y1="5" x2="34" y2="5"
      stroke="${rateStyles[name].stroke}" stroke-width="2.5"
      stroke-dasharray="${rateStyles[name].dash}" stroke-linecap="round"/></svg>${name}</span>`
})
```

```{ojs}
//| label: threshold-plot
//| echo: false
{
  const rates = transpose(threshold_rates);
  return Plot.plot({
    width: 640, height: 340,
    marginTop: 24,
    x: {domain: [0, 0.5], label: "Threshold t"},
    y: {domain: [0, 1], label: "Rate", grid: true},
    marks: [
      Plot.ruleX(threshold_marks, {stroke: "gray", strokeDasharray: "2,3"}),
      Plot.text(threshold_marks, {x: d => d, y: 1, text: d => `t = ${d}`,
                                  dy: -12, fill: "gray"}),
      ...rateNames.filter(r => shownRates.includes(r)).map(r =>
        Plot.line(rates.filter(d => d.rate === r),
                  {x: "t", y: "value", stroke: rateStyles[r].stroke,
                   strokeDasharray: rateStyles[r].dash, strokeWidth: 2.5,
                   strokeLinecap: "round"}))
    ]
  });
}
```

```{r compare-table}
#| echo: false
compare_table
```

As $t$ falls from $0.5$, sensitivity rises steadily while specificity stays
near $1$, and the false positive rate near $0$, until $t$ is small. The
overall error rate, dominated by the
`r format(cm_lda_05["TN"] + cm_lda_05["FP"], big.mark = ",")`
non-defaulters, follows the false positive rate: nearly flat over most of the
range, then climbing sharply as $t$ approaches $0$ and almost every customer is
flagged.

The overall error rate need not be smallest at $t = 0.5$. LDA's training
error rate is lowest near $t = `r sprintf("%.3f", min_err_lda[["t"]])`$, at
`r sprintf("%.2f", 100 * min_err_lda[["error"]])`%, against
`r sprintf("%.2f", 100 * m_lda_05["error"])`% at $t = 0.5$. The logistic
regression's is lowest near
$t = `r sprintf("%.3f", min_err_glm[["t"]])`$, at
`r sprintf("%.2f", 100 * min_err_glm[["error"]])`%, against
`r sprintf("%.2f", 100 * m_glm_05["error"])`% at $t = 0.5$.

In the table, lowering the threshold to $t = `r t_low`$ raises the
two-feature LDA classifier's sensitivity from
`r sprintf("%.1f", 100 * m_lda_05["sensitivity"])`% to
`r sprintf("%.1f", 100 * m_lda_low["sensitivity"])`%, at the cost of
specificity falling from `r sprintf("%.1f", 100 * m_lda_05["specificity"])`% to
`r sprintf("%.1f", 100 * m_lda_low["specificity"])`% and the overall error rate
rising from `r sprintf("%.2f", 100 * m_lda_05["error"])`% to
`r sprintf("%.2f", 100 * m_lda_low["error"])`%. Adding `income` changes LDA
very little: with `balance` alone the error rate is
`r sprintf("%.2f", 100 * m_lda1_05["error"])`% and the sensitivity
`r sprintf("%.1f", 100 * m_lda1_05["sensitivity"])`% at $t = 0.5$, and
`r sprintf("%.2f", 100 * m_lda1_low["error"])`% and
`r sprintf("%.1f", 100 * m_lda1_low["sensitivity"])`% at $t = `r t_low`$. At
both thresholds logistic regression flags more customers as defaulters than
the two-feature LDA classifier does —
`r cm_glm_05["TP"] + cm_glm_05["FP"]` against
`r cm_lda_05["TP"] + cm_lda_05["FP"]` at $t = 0.5$ — and has the higher
sensitivity and the lower specificity of the two.

### ROC curve

The **ROC curve** (receiver operating characteristic curve) plots sensitivity
against $1 - \text{specificity}$ — the true positive rate against the false
positive rate — at every threshold. At $t = 1$ nothing is predicted positive,
giving the point $(0, 0)$; at $t = 0$ everything is, giving $(1, 1)$; and the
curve traces every tradeoff in between.

The ROC curve depends on a classifier only through how it *orders* the
observations. Any strictly increasing transformation of the score — the
log posterior odds $\hat\delta_2(x) - \hat\delta_1(x)$ or $\hat p(x)$ itself —
sorts the observations identically, produces the same sequence of confusion
matrices, and so traces the same curve. Predicting positive when
$\hat p(x) > t$ is the same as predicting positive when the log posterior
odds exceed the cutoff $c = \log\left[t/(1 - t)\right]$.

For the credit card LDA classifier, the slider below moves the threshold
$t$; its left panel uses the same bins as the score histograms above. A line above
the plots reports $t$, the cutoff $c$ on the log posterior odds, sensitivity,
and specificity. The left panel draws the histograms of the classifier's log
posterior odds within each class, on the density scale, the defaulters' with
a dashed outline and the non-defaulters' with a solid one, and a dotted
vertical line at $c$; the shaded bars to the right of $c$ are the
customers flagged, darker for defaulters (sensitivity) and lighter for
non-defaulters (the false positive rate). The right panel marks where the
classifier lands on its ROC curve, with a dashed diagonal for reference.

::: {.callout-note collapse="true"}
### Code
```{r roc-slider-data}
#| output: false
is_yes <- Default$default == "Yes"

# Empirical ROC curve: every distinct score is a cutoff; classify positive
# when score >= cutoff.
roc_curve <- function(score, positive) {
  ord <- order(score, decreasing = TRUE)
  s   <- score[ord]
  pos <- positive[ord]
  keep <- c(diff(s) != 0, TRUE)
  tibble(fpr = c(0, cumsum(!pos)[keep] / sum(!positive)),
         tpr = c(0, cumsum(pos)[keep] / sum(positive)))
}

# Area under an ROC curve (sorted by fpr) by the trapezoid rule.
auc_trapezoid <- function(roc) {
  with(roc, sum(diff(fpr) * (head(tpr, -1) + tail(tpr, -1)) / 2))
}

# AUC as the proportion of (positive, negative) pairs ordered correctly.
auc_pairs <- function(score, positive) {
  s_pos <- score[positive]
  s_neg <- score[!positive]
  mean(outer(s_pos, s_neg, ">")) + 0.5 * mean(outer(s_pos, s_neg, "=="))
}

# Everything the slider needs, indexed by t = 0.01, 0.02, ..., 0.99, is
# computed here and passed to the page, which only looks values up.
# Each histogram's outline, as a stair through the bin edges.
slider_outline <- slider_bins |>
  pivot_longer(c(x0, x1), values_to = "x") |>
  select(class, x, density)

slider_thresholds <- tibble(t = (1:99) / 100) |>
  mutate(cutoff = qlogis(t),
         m = map(t, \(tt) metrics(confusion_counts(post_lda_2, tt))),
         fpr = 1 - map_dbl(m, "specificity"),
         tpr = map_dbl(m, "sensitivity")) |>
  select(-m)

roc_lda_2 <- roc_curve(post_lda_2, is_yes)
slider_roc <- slice(roc_lda_2, unique(c(seq(1, n(), by = 10), n())))
```
:::

```{r roc-slider-define}
#| echo: false
ojs_define(slider_bins = slider_bins,
           slider_outline = slider_outline,
           slider_thresholds = slider_thresholds,
           slider_roc = slider_roc,
           slider_colors = unname(class_colors))
```

```{ojs}
//| label: roc-slider
//| echo: false
viewof t = Inputs.range([0.01, 0.99], {value: 0.5, step: 0.01, label: "Threshold t"})
```

```{ojs}
//| label: roc-slider-plots
//| echo: false
{
  const bins = transpose(slider_bins);
  const outline = transpose(slider_outline);
  const thresholds = transpose(slider_thresholds);
  const roc = transpose(slider_roc);
  const now = thresholds[Math.round(t * 100) - 1];
  const classes = ["No (negative)", "Yes (positive)"];
  const flagged = k => bins
    .filter(d => d.class === k && d.x1 > now.cutoff)
    .map(d => ({...d, x0: Math.max(d.x0, now.cutoff)}));

  const densityPlot = Plot.plot({
    width: 360, height: 300,
    x: {label: "LDA log posterior odds of default"},
    y: {label: "Density within class"},
    color: {domain: classes, range: slider_colors, legend: true},
    marks: [
      Plot.rect(flagged(classes[0]),
                {x1: "x0", x2: "x1", y1: 0, y2: "density", fill: "class",
                 fillOpacity: 0.2}),
      Plot.rect(flagged(classes[1]),
                {x1: "x0", x2: "x1", y1: 0, y2: "density", fill: "class",
                 fillOpacity: 0.45}),
      Plot.line(outline.filter(d => d.class === classes[0]),
                {x: "x", y: "density", stroke: "class", strokeWidth: 2}),
      Plot.line(outline.filter(d => d.class === classes[1]),
                {x: "x", y: "density", stroke: "class", strokeDasharray: "6,4",
                 strokeWidth: 2}),
      Plot.ruleX([now.cutoff], {strokeDasharray: "2,3"})
    ]
  });

  const rocPlot = Plot.plot({
    width: 300, height: 300,
    x: {domain: [0, 1], label: "False positive rate (1 − specificity)"},
    y: {domain: [0, 1], label: "True positive rate (sensitivity)"},
    marks: [
      Plot.line([[0, 0], [1, 1]], {stroke: "gray", strokeDasharray: "4,4"}),
      Plot.line(roc, {x: "fpr", y: "tpr", strokeWidth: 2}),
      Plot.dot([now], {x: "fpr", y: "tpr", r: 6, fill: "white", stroke: "black",
                       strokeWidth: 2})
    ]
  });

  return html`<p style="text-align: center"><b>t = ${t.toFixed(2)}</b>,
    cutoff c = ${now.cutoff.toFixed(2)}; sensitivity = ${now.tpr.toFixed(3)},
    specificity = ${(1 - now.fpr).toFixed(3)}</p>
    <div style="display: flex; flex-wrap: wrap; gap: 1rem; justify-content: center">
      ${densityPlot}${rocPlot}
    </div>`;
}
```

Moving the slider from $t = 0.9$ to $t = 0.1$ slides the cutoff left, so the
shaded share of the defaulters' histogram (sensitivity) and of the
non-defaulters' (false positive rate) both grow, and the point climbs along
the curve toward $(1, 1)$. The dashed diagonal is the
ROC curve of a classifier that ignores the features and predicts positive at
random: its true and false positive rates are equal at every threshold. The
LDA curve bows far toward the top-left corner, where sensitivity is high and
the false positive rate is low.

The ROC curves below compare LDA and logistic regression with `balance` and
`income` at every threshold. Their points at $t = 0.5$ (filled) and
$t = `r t_low`$ (open) from the threshold table are marked, a triangle for LDA
and a square for logistic regression, with a dashed diagonal for reference.

::: {.callout-note collapse="true"}
### Code
```{r roc-default}
#| output: false
roc_2 <- bind_rows(
  roc_lda_2 |> mutate(model = "LDA"),
  roc_curve(post_glm_2, is_yes) |> mutate(model = "Logistic regression")
)

threshold_labels <- c("t = 0.5", paste("t =", t_low))
roc_points <- tibble(
  model = rep(c("LDA", "Logistic regression"), each = 2),
  t = factor(rep(threshold_labels, 2), levels = threshold_labels),
  m = list(m_lda_05, m_lda_low, m_glm_05, m_glm_low)
) |>
  mutate(fpr = 1 - map_dbl(m, "specificity"),
         tpr = map_dbl(m, "sensitivity"))

roc_plot <- ggplot(roc_2, aes(x = fpr, y = tpr)) +
  geom_abline(intercept = 0, slope = 1, color = "grey60",
              linetype = "dashed") +
  geom_line(aes(color = model, linetype = model), linewidth = 0.9) +
  geom_point(data = roc_points, aes(shape = model, color = model, fill = t),
             size = 3, stroke = 1.2) +
  scale_color_manual(values = model_colors) +
  # Shape separates the models, filled versus open separates the thresholds.
  scale_shape_manual(values = c(24, 22)) +
  scale_fill_manual(values = c("black", "white")) +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +
  coord_fixed() +
  labs(x = "False positive rate (1 - specificity)",
       y = "True positive rate (sensitivity)",
       color = "Model", linetype = "Model", shape = "Model",
       fill = "Threshold", title = "ROC curves, balance + income")

# Balance at which the two models' posterior curves cross, at the lower
# quartile, median, and upper quartile of income, and the posterior
# probability there.
posterior_both <- function(b, inc) {
  nd <- tibble(balance = b, income = inc)
  c(lda = predict(additive_lda, newdata = nd)$posterior[, "Yes"],
    glm = unname(predict(additive_glm, newdata = nd, type = "response")))
}
crossover <- tibble(income = unname(income_levels)) |>
  mutate(balance = map_dbl(income, \(i) {
           uniroot(\(b) diff(posterior_both(b, i)), c(800, 2500))$root
         }),
         p = map2_dbl(balance, income, \(b, i) posterior_both(b, i)[["lda"]]))

# Is one model's flagged set contained in the other's at a threshold?
flags_within <- function(p_inner, p_outer, t) !any(p_inner > t & p_outer <= t)
t_below <- 0.1
lda_within_glm <- flags_within(post_lda_2, post_glm_2, 0.5) &&
  flags_within(post_lda_2, post_glm_2, t_low)
glm_within_lda_below <- flags_within(post_glm_2, post_lda_2, t_below)
n_flag_lda_below <- sum(post_lda_2 > t_below)
n_flag_glm_below <- sum(post_glm_2 > t_below)
```
:::

```{r roc-figure}
#| echo: false
#| fig-width: 6.5
#| fig-height: 5.5
roc_plot
```

The two ROC curves lie almost on top of one another and rise steeply near the
origin, well above the diagonal. Lowering $t$ from $0.5$ to `r t_low` moves
each model's marked point up and slightly right along its curve. At each of
these two thresholds, logistic regression's point sits further along its
nearly shared curve than LDA's.

At the lower quartile, median, and upper quartile of income, the two models'
posterior curves cross at balances of
\$`r dollars(crossover$balance[1])`, \$`r dollars(crossover$balance[2])`,
and \$`r dollars(crossover$balance[3])`, where both give posterior
probabilities of `r sprintf("%.3f", crossover$p[1])`,
`r sprintf("%.3f", crossover$p[2])`, and `r sprintf("%.3f", crossover$p[3])`.
Above those balances LDA's flatter curve lies below logistic regression's, so
a threshold above the crossing probabilities is reached by LDA only at a
higher balance.
`r if (lda_within_glm) "At both marked thresholds, every customer LDA flags is also flagged by logistic regression." else "At the marked thresholds, LDA flags fewer customers than logistic regression."`
At a low threshold the order reverses: at $t = `r t_below`$, LDA flags
`r format(n_flag_lda_below, big.mark = ",")` customers and logistic
regression `r format(n_flag_glm_below, big.mark = ",")``r if (glm_within_lda_below) ", and every customer logistic regression flags is also flagged by LDA" else ""`.

### AUC

The **AUC** is the area under the ROC curve, a single number summarizing the
curve over all thresholds at once. It has a probabilistic interpretation.
Let $S_+$ be the score of a randomly chosen positive observation and $S_-$ the
score of an independently chosen negative one, and let $g_-$ be the density of
$S_-$.
At cutoff $c$ the true positive rate is $P(S_+ > c)$ and the false positive
rate is $P(S_- > c)$, whose derivative with respect to $c$ is $-g_-(c)$. As
$c$ runs from $+\infty$ down to $-\infty$ the false positive rate runs from
$0$ to $1$, so

$$\text{AUC} = \int_0^1 \text{TPR}\; d\,\text{FPR} = \int_{-\infty}^{\infty} P(S_+ > c)\, g_-(c)\, dc = P(S_+ > S_-),$$

where

- $S_+$ and $S_-$ are the classifier's scores for a random positive and a
random negative observation, and
- the final equality conditions on $S_- = c$ and integrates over its density.

The AUC is the probability that the classifier scores a randomly chosen
positive observation above a randomly chosen negative one. A classifier that
ignores the features has $\text{AUC} = 0.5$; one that ranks every positive
above every negative has $\text{AUC} = 1$.

With training data the same probability is estimated by comparing every
positive observation with every negative one,

$$\widehat{\text{AUC}} = \frac{1}{n_+ n_-}\sum_{i:\, y_i = \text{positive}}\ \sum_{j:\, y_j = \text{negative}}\left[\mathrm{I}(s_i > s_j) + \tfrac{1}{2}\mathrm{I}(s_i = s_j)\right],$$

where

- $n_+$ and $n_-$ are the numbers of positive and negative training
observations,
- $s_i$ is the classifier's score for observation $i$, and
- tied scores count as half a correctly ordered pair.

This equals the area under the empirical ROC curve computed by the trapezoid
rule.

Under the one-feature Gaussian model the score can be taken as $x$ itself
(with class $2$ positive and $\mu_2 > \mu_1$), so
$S_+ - S_- \sim N\left(\mu_2 - \mu_1, 2\sigma^2\right)$ and

$$\text{AUC} = P(S_+ - S_- > 0) = \Phi\left(\frac{\Delta}{\sqrt{2}}\right), \qquad \Delta = \frac{\mu_2 - \mu_1}{\sigma},$$

where

- $\Phi$ is the standard normal cumulative distribution function, and
- $\Delta$ is the separation of the class means in standard deviations.

For the credit card customers, the table below gives the AUC of LDA and
logistic regression with `balance` alone and with `balance` and `income`,
computed both as the area under the empirical ROC curve and as the proportion
of (defaulter, non-defaulter) pairs ordered correctly. In the figure, the left
panel shades the area under the running example's ROC curve, labeled with
that area, with a dashed diagonal for reference. The right panel draws the
ROC curves of four classifiers fit to the same customers — LDA with `balance`
only, with `income` only, and with both, and a classifier that ignores the
features — each labeled with its AUC.

::: {.callout-note collapse="true"}
### Code
```{r auc-default}
#| output: false
scores <- list(
  "LDA, balance" = post_1,
  "Logistic regression, balance" = fitted(balance_glm),
  "LDA, balance + income" = post_lda_2,
  "Logistic regression, balance + income" = post_glm_2
)
auc_values <- tibble(model = names(scores)) |>
  mutate(trapezoid = map_dbl(scores, \(s) auc_trapezoid(roc_curve(s, is_yes))),
         pairs = map_dbl(scores, \(s) auc_pairs(s, is_yes)))
auc_of <- function(m) auc_values$trapezoid[auc_values$model == m]
auc_table <- auc_values |>
  mutate(across(c(trapezoid, pairs), \(v) sprintf("%.5f", v))) |>
  knitr::kable(col.names = c("Model", "AUC (trapezoid rule)",
                             "AUC (pairwise comparison)"),
               align = "lcc")

# AUC implied by the fitted one-feature Gaussian model for balance.
delta_hat_1 <- unname(diff(hand_1$mu[, 1]) / sigma_hat_1)
auc_gauss_1 <- pnorm(delta_hat_1 / sqrt(2))
# The same calculation with each class's own standard deviation.
auc_gauss_sep <- pnorm(diff(hand_1$mu[, 1]) /
                         sqrt(sum(class_summary$sd_balance^2)))

auc_area_plot <- ggplot(roc_lda_2, aes(x = fpr, y = tpr)) +
  geom_ribbon(aes(ymin = 0, ymax = tpr), fill = "grey70", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "grey40",
              linetype = "dashed") +
  geom_line(linewidth = 0.9) +
  annotate("text", x = 0.65, y = 0.3,
           label = sprintf("AUC = %.5f", auc_of("LDA, balance + income"))) +
  coord_fixed() +
  labs(x = "False positive rate (1 - specificity)",
       y = "True positive rate (sensitivity)",
       title = "Area under the ROC curve")

# Classifiers fit to the same customers; the one that ignores the features
# gives every customer the same score.
compare_scores <- list(
  balance = post_1,
  income  = predict(MASS::lda(default ~ income, data = Default))$posterior[, "Yes"],
  both    = post_lda_2,
  none    = rep(0.5, nrow(Default))
)
compare_auc <- map_dbl(compare_scores, \(s) auc_trapezoid(roc_curve(s, is_yes)))
compare_labels <- parse(text = c(
  sprintf('"Balance only:" ~ AUC == "%.3f"', compare_auc[["balance"]]),
  sprintf('"Income only:" ~ AUC == "%.3f"', compare_auc[["income"]]),
  sprintf('"Balance and income:" ~ AUC == "%.3f"', compare_auc[["both"]]),
  sprintf('"No features:" ~ AUC == "%.3f"', compare_auc[["none"]])
))
compare_roc <- imap(compare_scores, \(s, nm) {
  mutate(roc_curve(s, is_yes), classifier = nm)
}) |>
  list_rbind() |>
  mutate(classifier = factor(classifier, levels = names(compare_scores)))

auc_classifier_plot <- ggplot(compare_roc,
                              aes(x = fpr, y = tpr, color = classifier,
                                  linetype = classifier)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("#E69F00", "#56B4E9", "#009E73", "grey40"),
                     labels = compare_labels) +
  scale_linetype_manual(values = c("dashed", "dotdash", "solid", "dotted"),
                        labels = compare_labels) +
  coord_fixed() +
  labs(x = "False positive rate (1 - specificity)",
       y = "True positive rate (sensitivity)", color = NULL, linetype = NULL,
       title = "ROC curves, same customers") +
  theme(legend.position = "inside", legend.position.inside = c(0.97, 0.03),
        legend.justification = c(1, 0),
        legend.text = element_text(hjust = 0),
        legend.key.width = unit(1.6, "lines"))

# Where along the false positive rates each of two nearly equal curves is
# the higher one.
fpr_grid <- seq(0, 1, by = 0.001)
tpr_on_grid <- function(s) {
  r <- roc_curve(s, is_yes)
  approx(r$fpr, r$tpr, xout = fpr_grid, ties = max)$y
}
tpr_both    <- tpr_on_grid(compare_scores$both)
tpr_balance <- tpr_on_grid(compare_scores$balance)
share_both_higher    <- mean(tpr_both > tpr_balance + 1e-12)
share_balance_higher <- mean(tpr_balance > tpr_both + 1e-12)
income_dips <- any(tpr_on_grid(compare_scores$income) < fpr_grid - 1e-12)
```
:::

```{r auc-table}
#| echo: false
auc_table
```

```{r auc-figure-display}
#| echo: false
#| layout-ncol: 2
#| fig-width: 4.5
#| fig-height: 4.5
auc_area_plot
auc_classifier_plot
```

The two AUC computations agree in every row of the table. With `balance`
alone, LDA and logistic regression have identical AUCs of
`r sprintf("%.5f", auc_of("LDA, balance"))`: both posterior probabilities are
increasing functions of balance, so they order the customers identically and
trace the same ROC curve. With `balance` and `income`, the two classifiers'
posterior probabilities weight income differently relative to balance, so
their orderings differ slightly. Their AUCs,
`r sprintf("%.5f", auc_of("LDA, balance + income"))` and
`r sprintf("%.5f", auc_of("Logistic regression, balance + income"))`, differ
by $`r sci(auc_of("LDA, balance + income") - auc_of("Logistic regression, balance + income"), 2)`$.
An AUC of `r sprintf("%.3f", auc_of("LDA, balance + income"))` means that for
a randomly chosen defaulter and non-defaulter, the LDA classifier gives the
defaulter the higher posterior probability about
`r round(100 * auc_of("LDA, balance + income"))`% of the time.

The fitted one-feature Gaussian model for balance has
$\hat\Delta = `r sprintf("%.3f", delta_hat_1)`$, so it implies an AUC of
$\Phi\left(`r sprintf("%.3f", delta_hat_1)`/\sqrt{2}\right) = `r sprintf("%.3f", auc_gauss_1)`$,
below the empirical `r sprintf("%.3f", auc_of("LDA, balance"))`. That is
consistent with the defaulters' balances being less spread out than the
shared $\hat\sigma$ assumes, which separates the classes better than the
fitted model expects. With each class's own standard deviation in place of
$\hat\sigma$, $S_+ - S_-$ has variance
$\hat\sigma_{\text{No}}^2 + \hat\sigma_{\text{Yes}}^2$, where
$\hat\sigma_{\text{No}}$ and $\hat\sigma_{\text{Yes}}$ are the class sample
standard deviations in the summary table, and
$\Phi\left(`r sprintf("%.2f", diff(hand_1$mu[, 1]))`/\sqrt{`r dollars(summary_of("No", "sd_balance"))`^2 + `r dollars(summary_of("Yes", "sd_balance"))`^2}\right) = `r sprintf("%.3f", auc_gauss_sep)`$,
close to the empirical value.

In the right panel, the AUC ranks the classifiers without a threshold.
Income alone barely separates the classes: its curve stays near the
diagonal`r if (income_dips) ", even dipping below it in places," else ","` and
its AUC is `r sprintf("%.3f", compare_auc[["income"]])`. The balance-only and
two-feature curves nearly coincide and cross: the two-feature curve is the
higher one over `r round(100 * share_both_higher)`% of the range of false
positive rates and the balance-only curve over
`r round(100 * share_balance_higher)`%. When two classifiers' ROC curves
cross, each has the higher true positive rate over part of the range of false
positive rates, and the AUC averages the true positive rate over all false
positive rates, so the classifier with the larger AUC can still be the worse
one over the false positive rates that matter for a given application.

## Credit card findings

LDA with `balance` and `income` identifies defaulters about as well as the
logistic regression with the same features: although its Gaussian
assumptions fit these customers imperfectly, it orders them almost exactly as
logistic regression does, and differs mainly in flagging fewer customers at
high thresholds. Income adds little to classification beyond balance,
raising LDA's AUC by
`r sprintf("%.4f", auc_of("LDA, balance + income") - auc_of("LDA, balance"))`,
although its logistic regression coefficient is clearly nonzero
($z = `r sprintf("%.2f", z_income_glm)`$). At $t = 0.5$
either classifier catches only a minority of defaulters, so a lender that
cares about missed defaults would use a lower threshold.

## Student strata

Student status has not entered any model so far. This additional analysis
asks whether separate models for students and non-students classify better.
The table below gives each stratum's number of customers, number of
defaulters, default rate, and median income.

::: {.callout-note collapse="true"}
### Code
```{r strata-summary}
#| output: false
strata <- c(No = "No", Yes = "Yes")
in_stratum <- map(strata, \(s) filter(Default, student == s))
strata_summary <- Default |>
  group_by(student) |>
  summarize(n = n(), defaulters = sum(default == "Yes"),
            rate = mean(default == "Yes"), median_income = median(income),
            min_income = min(income), max_income = max(income),
            .groups = "drop")
strata_summary_of <- function(s, col) {
  strata_summary[[col]][strata_summary$student == s]
}
strata_summary_table <- strata_summary |>
  transmute(student, n = format(n, big.mark = ","), defaulters,
            rate = sprintf("%.2f%%", 100 * rate),
            median_income = paste0("\\$", dollars(median_income))) |>
  knitr::kable(col.names = c("Student", "Customers", "Defaulters",
                             "Default rate", "Median income"),
               align = "ccccc")
```
:::

```{r strata-summary-table}
#| echo: false
strata_summary_table
```

Students default more often than non-students,
`r sprintf("%.2f", 100 * strata_summary_of("Yes", "rate"))`% against
`r sprintf("%.2f", 100 * strata_summary_of("No", "rate"))`%, and, as noted
above, their median income is far lower.

The figure below plots income against balance separately for non-students
and students, with defaulters and non-defaulters distinguished by color and
shape, and each panel's income axis covering its own stratum's incomes.

::: {.callout-note collapse="true"}
### Code
```{r strata-data}
#| output: false
strata_data_plot <- scatter_base +
  facet_wrap(~ student, labeller = label_both, scales = "free_y") +
  labs(title = "ISLR2::Default, income against balance by student status")
```
:::

```{r strata-data-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 4
strata_data_plot
```

In both strata the defaulters sit at the larger balances. The students'
incomes occupy a lower, narrower band, from
\$`r dollars(strata_summary_of("Yes", "min_income"))` to
\$`r dollars(strata_summary_of("Yes", "max_income"))`, against
\$`r dollars(strata_summary_of("No", "min_income"))` to
\$`r dollars(strata_summary_of("No", "max_income"))` for non-students.

Below, LDA with `balance` and `income` is fit separately to the non-students
and the students, so its priors, class means, and pooled covariance can all
differ by stratum. Beside it is the logistic regression in which the
intercept and both slopes interact with student status. Each customer is
scored by the model for its own stratum. Because each stratum's priors are
its own default rates, each posterior estimates the probability of default
given balance, income, and student status, so posteriors from the two strata
are on one scale and can be compared with each other.

::: {.callout-note collapse="true"}
### Code
```{r strata}
#| output: false
strata_lda <- map(in_stratum, \(d) MASS::lda(default ~ balance + income, data = d))
strata_glm <- map(in_stratum, \(d) glm(default ~ balance + income, data = d,
                                       family = "binomial"))
interacted_glm <- glm(default ~ (balance + income) * student, data = Default,
                      family = "binomial")

# Score every customer with its own stratum's model.
by_stratum <- function(score) {
  p <- numeric(nrow(Default))
  for (s in strata) p[Default$student == s] <- score(s)
  p
}
post_lda_strat <- by_stratum(\(s) {
  predict(strata_lda[[s]], newdata = in_stratum[[s]])$posterior[, "Yes"]
})
post_glm_strat <- by_stratum(\(s) fitted(strata_glm[[s]]))
max_diff_int <- max(abs(fitted(interacted_glm) - post_glm_strat))

strata_metrics <- imap(list(
  "LDA, single fit" = post_lda_2, "LDA, by student stratum" = post_lda_strat,
  "Logistic regression, additive" = post_glm_2,
  "Logistic regression, interacted with student" = fitted(interacted_glm)
), \(p, nm) {
  m <- metrics(confusion_counts(p, 0.5))
  tibble(model = nm, error = m[["error"]],
         auc = auc_trapezoid(roc_curve(p, is_yes)))
}) |>
  list_rbind()
strata_of <- function(m, col) strata_metrics[[col]][strata_metrics$model == m]
strata_table <- strata_metrics |>
  mutate(error = sprintf("%.4f", error), auc = sprintf("%.5f", auc)) |>
  knitr::kable(col.names = c("Model", "Error rate at $t = 0.5$", "AUC"),
               align = "lcc")

# Each stratum's boundary from each model, with its balance threshold at the
# stratum's median income.
strata_hand <- map(in_stratum, \(d) lda_by_hand(d[c("balance", "income")],
                                                 d$default))
strata_lines <- imap(strata, \(s, nm) {
  h <- strata_hand[[s]]
  b <- coef(strata_glm[[s]])
  tibble(student = s, model = c("LDA", "Logistic regression"),
         b0 = c(h$beta0, b[[1]]), b1 = c(h$beta[[1]], b[[2]]),
         b2 = c(h$beta[[2]], b[[3]]))
}) |>
  list_rbind() |>
  left_join(strata_summary |>
              transmute(student = as.character(student), median_income),
            by = "student") |>
  mutate(intercept = -b0 / b2, slope = -b1 / b2,
         threshold = threshold_at(b0, b1, b2, median_income))
strata_value <- function(s, col, m = "LDA") {
  strata_lines[[col]][strata_lines$student == s & strata_lines$model == m]
}
# The LDA and logistic boundaries of a stratum cross where their balance
# thresholds are equal. If each crossing lies outside the incomes the stratum
# has, and logistic regression's threshold is lower at the median income,
# logistic regression's boundary is left of LDA's throughout.
boundary_crossing <- function(s) {
  l <- filter(strata_lines, student == s)
  (l$b0[2] / l$b1[2] - l$b0[1] / l$b1[1]) /
    (l$b2[1] / l$b1[1] - l$b2[2] / l$b1[2])
}
glm_left_throughout <- all(map_lgl(strata, \(s) {
  cr <- boundary_crossing(s)
  inc <- in_stratum[[s]]$income
  (cr < min(inc) || cr > max(inc)) &&
    strata_value(s, "threshold", "Logistic regression") <
      strata_value(s, "threshold")
}))

# How much the students' larger prior alone lowers their LDA threshold:
# the change in log prior odds, converted to balance by the balance slope.
log_prior_odds <- function(h) log(h$pi[2] / h$pi[1])
prior_shift <- (log_prior_odds(strata_hand$Yes) - log_prior_odds(strata_hand$No)) /
  strata_hand$Yes$beta[[1]]
# Class mean balances by stratum, and the midpoint of each stratum's pair.
class_mean <- function(s, k) strata_hand[[s]]$mu[k, "balance"]
midpoint_shift <- mean(strata_hand$Yes$mu[, "balance"]) -
  mean(strata_hand$No$mu[, "balance"])
# Students' minus non-students' LDA threshold over the incomes both strata
# have; the thresholds are linear in income, so the ends bound the gap.
overlap <- c(max(strata_summary$min_income), min(strata_summary$max_income))
threshold_gap <- threshold_at(strata_hand$Yes$beta0, strata_hand$Yes$beta[[1]],
                              strata_hand$Yes$beta[[2]], overlap) -
  threshold_at(strata_hand$No$beta0, strata_hand$No$beta[[1]],
               strata_hand$No$beta[[2]], overlap)

# LDA estimates K - 1 priors, K p means, and p (p + 1) / 2 covariances.
K <- nlevels(Default$default)
p_features <- length(c("balance", "income"))
n_par_lda <- (K - 1) + K * p_features + p_features * (p_features + 1) / 2

strata_plot <- scatter_base +
  boundary_layers(strata_lines) +
  # Each panel's income axis covers its own stratum's incomes.
  facet_wrap(~ student, labeller = label_both, scales = "free_y") +
  labs(linetype = "Boundary",
       title = "Boundaries fit within each student stratum")
```
:::

The table gives the training error rate at $t = 0.5$ and the AUC of these two
models and of the single LDA and additive logistic regression of the earlier
sections, all computed on the training customers. The figure repeats the
scatterplot of income against balance for each stratum, over that stratum's
own range of income, with each stratum's LDA boundary (solid green) and
interacted logistic regression boundary (dashed pink).

```{r strata-table}
#| echo: false
strata_table
```

```{r strata-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 4
strata_plot
```

Fitting the logistic regression with every coefficient interacted with
student status is the same as fitting separate logistic regressions to the
two strata: the two give the same fitted probabilities, to within
$`r sci(max_diff_int, 1)`$.

The extra flexibility barely changes the classifications. Stratifying LDA
moves its training error rate from
`r sprintf("%.2f", 100 * strata_of("LDA, single fit", "error"))`% to
`r sprintf("%.2f", 100 * strata_of("LDA, by student stratum", "error"))`% and
its AUC from `r sprintf("%.4f", strata_of("LDA, single fit", "auc"))` to
`r sprintf("%.4f", strata_of("LDA, by student stratum", "auc"))`; interacting
the logistic regression with student status moves its error rate from
`r sprintf("%.2f", 100 * strata_of("Logistic regression, additive", "error"))`%
to
`r sprintf("%.2f", 100 * strata_of("Logistic regression, interacted with student", "error"))`%
and its AUC from
`r sprintf("%.4f", strata_of("Logistic regression, additive", "auc"))` to
`r sprintf("%.4f", strata_of("Logistic regression, interacted with student", "auc"))`.
The AUC compares defaulters with non-defaulters across strata as well as
within them. These are training rates, and the stratified models estimate
twice as many parameters — `r 2 * n_par_lda` against `r n_par_lda` for LDA
(`r K - 1` prior, `r K * p_features` means, and
`r p_features * (p_features + 1) / 2` covariance entries per fit), and
`r length(coef(interacted_glm))` against `r length(coef(additive_glm))` for
logistic regression — so their training error rates tend to understate test
error, and their training AUCs to overstate test AUC, by more; whether stratifying helps on new customers is a question for
test data.

At each stratum's median income, LDA's balance threshold is
\$`r dollars(strata_value("No", "threshold"))` for non-students and
\$`r dollars(strata_value("Yes", "threshold"))` for students, and logistic
regression's is
\$`r dollars(strata_value("No", "threshold", "Logistic regression"))` and
\$`r dollars(strata_value("Yes", "threshold", "Logistic regression"))`.
`r if (glm_left_throughout) "Over the incomes each stratum actually has, logistic regression's boundary lies to the left of LDA's." else "Over part of the incomes the strata have, the two boundaries swap sides."`

The students' larger prior for `Yes` alone would lower their LDA threshold
by \$`r dollars(prior_shift)`. Their threshold is higher nonetheless because
students in both classes carry higher balances: non-defaulting students
average \$`r dollars(class_mean("Yes", "No"))` against
\$`r dollars(class_mean("No", "No"))` for non-students, and defaulting
students \$`r dollars(class_mean("Yes", "Yes"))` against
\$`r dollars(class_mean("No", "Yes"))`, which puts the midpoint of the
students' class means \$`r dollars(midpoint_shift)` higher. This is the
confounding of the
[Multiple Logistic Regression](20-multiple-logistic-regression.qmd) lecture.
Over the incomes both strata have, \$`r dollars(overlap[1])` to
\$`r dollars(overlap[2])`, the students' threshold minus the non-students'
ranges from \$`r dollars(min(threshold_gap))` to
\$`r dollars(max(threshold_gap))`, so
`r if (all(threshold_gap > 0)) "it stays higher when the strata are compared at the same income" else "the comparison depends on income"`.

## Conclusion

A generative classifier models the distribution of the features within each
class, $f_k(x)$, and the prior probability of each class, $\pi_k$, and
combines them through Bayes' theorem into the posterior probability
$p_k(x) = \pi_k f_k(x) / \sum_l \pi_l f_l(x)$. Assigning each observation to
the class with the largest posterior probability is the Bayes classifier,
which minimizes the expected error rate; a generative classifier approximates
it by estimating $\pi_k$ and $f_k$ from training data.

Linear discriminant analysis assumes Gaussian class-conditional densities with
class-specific means and a covariance matrix shared by every class. Because
the covariance is shared, the quadratic term in $x$ is the same for every
class, so it can be dropped from the discriminant functions $\delta_k(x)$,
leaving them linear in $x$ and a decision boundary that is a point, a line, or a
hyperplane. With two classes the log posterior odds are linear in the
features, the same form as logistic regression, but the coefficients come
from closed-form estimates from the joint likelihood of features and
responses rather than from the iteratively maximized conditional likelihood
of logistic regression. The class proportions and class means maximize that
joint likelihood, and the pooled variance or covariance matrix is its
maximizer rescaled to be unbiased. A smaller prior probability moves the
boundary toward that class's mean, and past it when the prior is small
enough.

A classifier's error rate hides which mistakes it makes. The confusion matrix
separates them, sensitivity and specificity measure performance within each
true class, and moving the threshold trades one against the other. With $K$
classes the confusion matrix is $K \times K$ and sensitivity and specificity
are computed one class versus the rest. The ROC curve displays the two-class
tradeoff at every threshold and depends only on how a classifier orders the
observations. Its area, the AUC, is the probability that a randomly chosen
positive observation is scored above a randomly chosen negative one, and it
ranks classifiers without fixing a threshold.

The next lecture relaxes LDA's shared covariance assumption. Quadratic
discriminant analysis gives each class its own covariance matrix, so the
quadratic term no longer cancels and the boundary curves, and naive Bayes
replaces the multivariate Gaussian with a product of one-feature densities.
That lecture then compares all five classifiers, adding KNN.
