# Linear Discriminant Analysis

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
set.seed(20261001)

# MASS is called as MASS::lda() rather than attached, because attaching it
# masks dplyr::select().
# Renders a number in LaTeX scientific notation for use inside $...$.
sci <- function(x, digits = 2) {
  e <- floor(log10(abs(x)))
  m <- signif(x / 10^e, digits)
  sprintf("%s \\times 10^{%d}", format(m), e)
}

class_colors <- c("#0072B2", "#D55E00")
model_colors <- c("LDA" = "#009E73", "Logistic regression" = "#CC79A7")
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **derive** the posterior probability $p_k(x) = P(Y = k \mid X = x)$ from the
prior probabilities $\pi_k$ and the class-conditional densities $f_k(x)$ using
Bayes' theorem, and **explain** why classifying to the class with the largest
posterior probability minimizes the expected error rate,
- **derive** the LDA discriminant function $\delta_k(x)$ for one feature under
Gaussian class-conditional densities with a shared variance $\sigma^2$, and
**show** that it is linear in $x$, so that the decision boundary is a single
point that moves toward the mean of the class whose prior probability is
smaller,
- **write** the discriminant function for $p > 1$ features in matrix form using
class mean vectors $\mu_k$ and a shared covariance matrix $\Sigma$, and
**compute** the estimates $\hat\pi_k$, $\hat\mu_k$, and $\hat\sigma^2$ or
$\hat\Sigma$ from training data,
- **fit** LDA with `MASS::lda()`, **interpret** its prior probabilities, group
means, and posterior probabilities, and **compare** its posterior
probabilities and decision boundary with those of logistic regression,
- **construct** a confusion matrix at a stated threshold (or a $K \times K$
confusion matrix for $K$ classes), **compute** the error rate, sensitivity,
and specificity from it, and **explain** how sensitivity
and specificity trade off as the threshold moves, and
- **interpret** an ROC curve and its AUC, including AUC as the probability that
a randomly chosen positive observation receives a higher score than a randomly
chosen negative one.
:::

## Generative classifiers

Logistic regression models the conditional probability $P(Y = k \mid X = x)$
of each class $k = 1, \ldots, K$ directly. Its likelihood is the conditional distribution of the response given
the features, and the distribution of the features themselves never enters it.

A **generative classifier** takes the other route. It models how the features
are distributed *within* each class, and how common each class is, and then
reverses the conditioning with Bayes' theorem to obtain
$P(Y = k \mid X = x)$. Linear discriminant analysis (LDA) is the first generative
classifier in this course.

### Bayes' theorem

Every earlier lecture conditioned on the features and treated them as fixed. A
generative classifier models their distribution, so here $X$ is a random
vector — the features of a randomly chosen observation — and $x$ is a value it
takes. Let the response $Y$ take one of $K$ classes, $k = 1, \ldots, K$, and
define

$$\pi_k = P(Y = k), \qquad f_k(x) = \text{density of } X \text{ given } Y = k,$$

where

- $K$ is the number of classes ($K = 2$ for a binary response),
- $\pi_k$ is the **prior probability** of class $k$, the probability that a
randomly chosen observation belongs to class $k$ before its features are
seen, with $\sum_{k=1}^K \pi_k = 1$, and
- $f_k(x)$ is the **class-conditional density** of $X$ among observations in
class $k$ (a probability mass function if $X$ is discrete).

The classification and logistic regression lectures wrote $C$ for the number
of classes; following ISLR2, this and later lectures write $K$.

Bayes' theorem, with densities in place of the probabilities of the events
$\{X = x\}$, gives

$$P(Y = k \mid X = x) = \frac{P(Y = k)\, f_k(x)}{f(x)},$$

where $f(x)$ is the marginal density of $X$. By the law of total probability,
$f(x) = \sum_{l=1}^K P(Y = l)\, f_l(x)$, so

$$p_k(x) = P(Y = k \mid X = x) = \frac{\pi_k f_k(x)}{\sum_{l=1}^K \pi_l f_l(x)} \propto \pi_k f_k(x),$$

where

- $p_k(x)$ is the **posterior probability** of class $k$ for an observation
with features $x$,
- $l$ indexes the classes in the sum, kept distinct from the $k$ being
evaluated, and
- $\propto$ means "proportional to" as a function of $k$ at a fixed $x$.

Only the numerator distinguishes one class from another: a class is probable
at $x$ when it is common ($\pi_k$ large) and when $x$ is a typical feature
value for it ($f_k(x)$ large).

### Bayes classifier

A **classifier** is an assignment rule $\hat y(x)$ that maps each value of
the features to one of the classes $1, \ldots, K$. The **Bayes classifier**
assigns each observation to the class with the largest posterior probability,

$$\hat y(x) = \arg\max_k\, p_k(x),$$

where

- $\hat y(x)$ is the class assigned to an observation with features $x$, and
- $\arg\max_k p_k(x)$ is the class $k$ at which $p_k(x)$ is largest.

With $K = 2$ classes, $p_1(x) = 1 - p_2(x)$, so the Bayes classifier assigns
class $2$ whenever $p_2(x) > 0.5$.

For any classifier, the probability that the assignment at $x$ is wrong is

$$P\left(Y \ne \hat y(x) \mid X = x\right) = 1 - p_{\hat y(x)}(x).$$

This is smallest when $p_{\hat y(x)}(x)$ is largest, which is exactly the
Bayes classifier's choice. The Bayes classifier therefore minimizes the
conditional error probability at every $x$, and so it also minimizes the
overall error rate $E\left[1 - p_{\hat y(X)}(X)\right]$, which averages those
conditional error probabilities over the distribution of $X$. Its overall
error rate, the **Bayes error rate**, is the lowest achievable by any
classifier.

The Bayes classifier needs the true $\pi_k$ and $f_k$, which are never known.
A generative classifier estimates both from training data and substitutes the
estimates into Bayes' theorem. Estimating $\pi_k$ takes nothing more than the
proportion of training observations in class $k$. Estimating $f_k$ requires an
assumption about its form, and LDA's assumption is that it is Gaussian.

## One-feature LDA

### Gaussian class-conditional densities

With a single feature ($p = 1$), LDA assumes the statistical model

$$P(Y = k) = \pi_k, \qquad X \mid Y = k \sim N\left(\mu_k, \sigma^2\right), \qquad k = 1, \ldots, K,$$

where

- $\mu_k = E[X \mid Y = k]$ is the mean of the feature in class $k$, and
- $\sigma^2 = Var[X \mid Y = k]$ is the variance of the feature, **the same in
every class**.

The class-conditional density is therefore the normal density

$$f_k(x) = \frac{1}{\sqrt{2\pi}\,\sigma} \exp\left[-\frac{(x - \mu_k)^2}{2\sigma^2}\right],$$

where $\pi$ without a subscript is the constant $3.14159\ldots$, unrelated to the
prior probabilities $\pi_k$.

The classes differ only in where their densities are centered. The shared
variance is what makes the classifier linear.

### Discriminant function

Since $p_k(x)$ is proportional to $\pi_k f_k(x)$ and the logarithm is
increasing, the class with the largest posterior probability is also the class
with the largest $\log\left(\pi_k f_k(x)\right)$. Expanding the square and
collecting terms,

$$\log\left(\pi_k f_k(x)\right) = \log \pi_k - \frac{\mu_k^2}{2\sigma^2} + \frac{\mu_k}{\sigma^2}\, x - \log\left(\sqrt{2\pi}\,\sigma\right) - \frac{x^2}{2\sigma^2}.$$

The last two terms do not involve $k$ because $\sigma^2$ is shared, so they
cannot change which class is largest. The last is the only term involving
$x^2$, which is why the quadratic term cancels too. Dropping them leaves the
**discriminant function**

$$\delta_k(x) = \left(\log \pi_k - \frac{\mu_k^2}{2\sigma^2}\right) + \frac{\mu_k}{\sigma^2}\, x,$$

where

- $\delta_k(x)$ is the discriminant score of class $k$ at feature value $x$,
- $\log \pi_k - \mu_k^2/\left(2\sigma^2\right)$ is its intercept, the score
at $x = 0$, and
- $\mu_k/\sigma^2$ is its slope.

Each $\delta_k$ is linear in $x$, with the same intercept-plus-slope form as a
regression's linear predictor. The classifier assigns an observation to the
class with the largest $\delta_k(x)$.

Had the variances differed, with $\sigma_k^2$ the class-$k$ variance, the
term $-\log\left(\sqrt{2\pi}\,\sigma_k\right)$ would differ between classes
and stay in each intercept, and the term $-x^2/\left(2\sigma_k^2\right)$ would
carry a different coefficient in each class, would not drop out, and would
make each discriminant quadratic in $x$.

The dropped terms are the same in every class's $\log\left(\pi_k f_k(x)\right)$,
so their exponentials are common factors of every $\pi_k f_k(x)$. They cancel
from the ratio in Bayes' theorem and the posterior probabilities can be
recovered from the discriminant scores alone:

$$p_k(x) = \frac{e^{\delta_k(x)}}{\sum_{l=1}^K e^{\delta_l(x)}}.$$

### Decision boundary

With $K = 2$ classes, $p_1(x) = 1 - p_2(x)$, and the log posterior odds of
class $2$ are

$$\log\left(\frac{p_2(x)}{1 - p_2(x)}\right) = \delta_2(x) - \delta_1(x) = \underbrace{\log\frac{\pi_2}{\pi_1} - \frac{\mu_2^2 - \mu_1^2}{2\sigma^2}}_{\beta_0} + \underbrace{\frac{\mu_2 - \mu_1}{\sigma^2}}_{\beta_1}\, x.$$

The log-odds are linear in $x$, so LDA's posterior probability has exactly the
form of a simple logistic regression. The two methods differ in how they
estimate $\beta_0$ and $\beta_1$, not in the shape of the curve they fit.

The **decision boundary** is the set of $x$ where the two classes are equally
probable, $\delta_1(x) = \delta_2(x)$. Setting the log posterior odds to zero
and solving,

$$x^* = \frac{\mu_1 + \mu_2}{2} + \frac{\sigma^2}{\mu_2 - \mu_1}\log\frac{\pi_1}{\pi_2},$$

where

- $x^*$ is the feature value at which the classifier switches from one class
to the other.

With equal prior probabilities the logarithm is zero and the boundary is the
midpoint of the two means. Take $\mu_2 > \mu_1$. If class $2$ is the rarer
class, $\pi_2 < \pi_1$, the logarithm is positive and the boundary moves to the
right, toward $\mu_2$: an observation must look more like a typical class-$2$
observation before the classifier assigns it to the rarer class.

The figure below uses two classes with $\mu_1 = -1.25$, $\mu_2 = 1.25$, and
$\sigma = 1$. The first two tabs plot $\pi_k f_k(x)$ for equal and unequal
priors, the third plots $\delta_1(x)$ and $\delta_2(x)$ for the unequal
priors, and the dotted line marks $x^*$.

::: {.callout-note collapse="true"}
### Code
```{r theory-1d}
#| output: false
mu_1 <- -1.25
mu_2 <-  1.25
sigma <- 1
pi_unequal <- c(0.8, 0.2)

boundary_1d <- function(pi1, pi2) {
  (mu_1 + mu_2) / 2 + sigma^2 / (mu_2 - mu_1) * log(pi1 / pi2)
}
# Plot title "pi_1 = a, pi_2 = b; boundary x* = c" as a plotmath expression.
# The asterisk is juxtaposed (x * "*"), not a superscript: a superscripted
# asterisk renders far above the x.
prior_title <- function(pi1, pi2, x_star) {
  bquote(paste(pi[1] == .(sprintf("%.1f", pi1)), ", ",
               pi[2] == .(sprintf("%.1f", pi2)), "; boundary ",
               x * "*" == .(sprintf("%.2f", x_star))))
}
x_star_equal   <- boundary_1d(0.5, 0.5)
x_star_unequal <- boundary_1d(pi_unequal[1], pi_unequal[2])

theory_grid <- tibble(x = seq(-4.5, 4.5, length.out = 500))

weighted_curves <- function(pi1, pi2) {
  theory_grid |>
    mutate(`k = 1` = pi1 * dnorm(x, mu_1, sigma),
           `k = 2` = pi2 * dnorm(x, mu_2, sigma)) |>
    pivot_longer(-x, names_to = "class", values_to = "density")
}

weighted_plot <- function(pi1, pi2, x_star) {
  ggplot(weighted_curves(pi1, pi2),
         aes(x = x, y = density, color = class, linetype = class)) +
    geom_line(linewidth = 0.9) +
    geom_vline(xintercept = x_star, linetype = "dotted") +
    scale_color_manual(values = class_colors) +
    coord_cartesian(ylim = c(0, 0.33)) +
    labs(x = "x", y = expression(pi[k] * f[k](x)), color = "Class",
         linetype = "Class", title = prior_title(pi1, pi2, x_star))
}

theory_equal_plot   <- weighted_plot(0.5, 0.5, x_star_equal)
theory_unequal_plot <- weighted_plot(pi_unequal[1], pi_unequal[2],
                                     x_star_unequal)

delta_lines <- theory_grid |>
  mutate(`k = 1` = (log(pi_unequal[1]) - mu_1^2 / (2 * sigma^2)) +
                   mu_1 / sigma^2 * x,
         `k = 2` = (log(pi_unequal[2]) - mu_2^2 / (2 * sigma^2)) +
                   mu_2 / sigma^2 * x) |>
  pivot_longer(-x, names_to = "class", values_to = "delta")

theory_delta_plot <- ggplot(delta_lines,
                            aes(x = x, y = delta, color = class,
                                linetype = class)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = x_star_unequal, linetype = "dotted") +
  scale_color_manual(values = class_colors) +
  labs(x = "x", y = expression(delta[k](x)), color = "Class",
       linetype = "Class",
       title = prior_title(pi_unequal[1], pi_unequal[2], x_star_unequal))
```
:::

::: {.panel-tabset}
#### Equal priors
```{r theory-equal-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
theory_equal_plot
```

#### Unequal priors
```{r theory-unequal-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
theory_unequal_plot
```

#### Discriminant functions
```{r theory-delta-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
theory_delta_plot
```
:::

Since $\delta_1(x) = \delta_2(x)$ exactly where
$\pi_1 f_1(x) = \pi_2 f_2(x)$, the boundary is where the two curves cross.
With equal priors it sits at the midpoint,
$x^* = `r sprintf("%.2f", x_star_equal)`$. Shrinking $\pi_2$ to
`r pi_unequal[2]` scales the class-$2$ curve down, so the crossing moves right
to $x^* = `r sprintf("%.2f", x_star_unequal)`$. In the third tab the two
discriminant functions are straight lines that cross at the same $x^*$. A linear discriminant function for each class, with
the boundary where two of them are equal, is the "linear" in linear
discriminant analysis.

### Parameter estimation

The parameters $\pi_k$, $\mu_k$, and $\sigma^2$ are estimated from training
observations $(x_1, y_1), \ldots, (x_n, y_n)$, here with $x_i$ the scalar
feature value of observation $i$:

$$\hat\pi_k = \frac{n_k}{n}, \qquad \hat\mu_k = \frac{1}{n_k}\sum_{i:\, y_i = k} x_i, \qquad \hat\sigma^2 = \frac{1}{n - K}\sum_{k=1}^K \sum_{i:\, y_i = k}\left(x_i - \hat\mu_k\right)^2,$$

where

- $n_k$ is the number of training observations in class $k$, so
$\sum_k n_k = n$,
- $\sum_{i:\, y_i = k}$ sums over the training observations in class $k$, and
- $\hat\sigma^2$ is the **pooled variance**: squared deviations from each
observation's own class mean, summed over all classes and divided by $n - K$.

$\hat\pi_k$ and $\hat\mu_k$ are maximum likelihood estimates; $\hat\sigma^2$
is the maximum likelihood estimate rescaled by $n/(n - K)$ to make it
unbiased. Under the generative model the joint
density of observation $i$ is $P(Y = y_i)\, f_{y_i}(x_i) = \pi_{y_i} f_{y_i}(x_i)$,
so for independent observations the log-likelihood is

$$\ell(\pi_1, \ldots, \pi_K, \mu_1, \ldots, \mu_K, \sigma^2) = \sum_{k=1}^K n_k \log \pi_k \;+\; \sum_{k=1}^K \sum_{i:\, y_i = k}\left[-\frac{1}{2}\log\left(2\pi\sigma^2\right) - \frac{(x_i - \mu_k)^2}{2\sigma^2}\right].$$

The first sum involves only the priors; maximizing it subject to
$\sum_k \pi_k = 1$ (with a Lagrange multiplier) gives $\hat\pi_k = n_k/n$. In
the second, $\mu_k$ appears only in class $k$'s terms, so its maximizer is the
class-$k$ sample mean. Setting the derivative with respect to $\sigma^2$ to
zero gives the pooled sum of squares divided by $n$; dividing by $n - K$
instead, one degree of freedom per estimated mean, makes it unbiased, in the
same way that $RSS/(n-2)$ replaced $RSS/n$ in simple linear regression.
`MASS::lda()` uses $n - K$.

Logistic regression maximizes a different likelihood, the conditional one,
$\prod_{i=1}^n p(x_i)^{y_i}\left[1 - p(x_i)\right]^{1 - y_i}$ (with $y_i$
coded $0/1$ as in the logistic regression lectures), which has no closed-form
maximizer. LDA maximizes the joint likelihood of features and
responses under a stronger assumption — Gaussian features with a shared
variance — and every estimate is a sample proportion, mean, or variance.

::: {.callout-note collapse="true"}
### For example,
LDA's estimates have closed forms, and separation does not prevent them from
existing, though a singular pooled covariance still would. The previous lecture's completely separated toy data,
$x = 1, \ldots, 10$ with $y = 1$ exactly when $x > 5$, left logistic
regression with no finite maximum likelihood estimate. `MASS::lda()`, used
again in the worked example below, fits LDA to the same data:

```{r toy-lda}
toy <- tibble(x = 1:10, y = factor(as.numeric(x > 5)))
toy_lda <- MASS::lda(y ~ x, data = toy)
toy_lda$means
```

The class means are `r toy_lda$means[1]` and `r toy_lda$means[2]`, the priors
are equal, and so the LDA boundary is the midpoint,
$x^* = `r mean(toy_lda$means)`$.
:::

## Multiple-feature LDA

### Multivariate Gaussian densities

With $p > 1$ features, $X = (X_1, \ldots, X_p)^\top$ is a random column vector
whose entries are the features, and $x$ is a $p$-vector value it takes. A
value $x$ holds the same kind of entries as a row of the model matrix
$\mathbf{X}$ but without the leading $1$, and it is a column because it is the
argument of a multivariate density. LDA assumes the statistical model

$$P(Y = k) = \pi_k, \qquad X \mid Y = k \sim N_p\left(\mu_k, \Sigma\right), \qquad k = 1, \ldots, K,$$

where

- $N_p$ is the $p$-dimensional multivariate normal distribution,
- $\mu_k = E[X \mid Y = k] = (\mu_{k1}, \ldots, \mu_{kp})^\top$ is the
$p$-vector of feature means in class $k$, and
- $\Sigma = Cov[X \mid Y = k]$ is the $p \times p$ covariance matrix of the
features, **the same in every class**, with feature variances on its diagonal
and covariances between pairs of features off it; it is symmetric and
positive definite.

The class-conditional density is therefore the multivariate normal density

$$f_k(x) = \frac{1}{(2\pi)^{p/2}\,|\Sigma|^{1/2}} \exp\left[-\frac{1}{2}\left(x - \mu_k\right)^\top \Sigma^{-1} \left(x - \mu_k\right)\right],$$

where $|\Sigma|$ is the determinant of $\Sigma$.

Each class's density has contours that are nested ellipses centered at
$\mu_k$. Sharing
$\Sigma$ means every class's ellipses have the same shape and orientation and
differ only in their center.

### Matrix-form discriminant

The derivation follows the one-feature case. Because $\Sigma^{-1}$ is
symmetric, the scalar $\mu_k^\top \Sigma^{-1} x$ equals its own transpose
$x^\top \Sigma^{-1} \mu_k$, and the quadratic form expands to

$$\left(x - \mu_k\right)^\top \Sigma^{-1} \left(x - \mu_k\right) = x^\top \Sigma^{-1} x - 2\, x^\top \Sigma^{-1} \mu_k + \mu_k^\top \Sigma^{-1} \mu_k.$$

Then, collecting terms in the same order as before,

$$\log\left(\pi_k f_k(x)\right) = \log \pi_k - \frac{1}{2}\mu_k^\top \Sigma^{-1} \mu_k + x^\top \Sigma^{-1} \mu_k - \frac{p}{2}\log(2\pi) - \frac{1}{2}\log|\Sigma| - \frac{1}{2}x^\top \Sigma^{-1} x.$$

The last three terms do not involve $k$: the first is a constant, and the
other two because $\Sigma$ is shared. The last is the only term quadratic in
$x$, which is why the quadratic term cancels too. Dropping them,

$$\delta_k(x) = \left(\log \pi_k - \frac{1}{2}\mu_k^\top \Sigma^{-1} \mu_k\right) + x^\top \Sigma^{-1} \mu_k,$$

where

- $\log \pi_k - \frac{1}{2}\mu_k^\top \Sigma^{-1} \mu_k$ is the intercept of
class $k$'s discriminant, and
- $\Sigma^{-1}\mu_k$ is its $p$-vector of slopes, one per feature.

The discriminant is linear in $x$ and reduces to the one-feature discriminant
when $\Sigma = \sigma^2$. As before, the classifier assigns the class with the
largest $\delta_k(x)$, and $p_k(x) = e^{\delta_k(x)} / \sum_l e^{\delta_l(x)}$.

### Linear decision boundary

With $K = 2$, subtracting the two discriminants gives the log posterior odds
of class $2$,

$$\log\left(\frac{p_2(x)}{1 - p_2(x)}\right) = \delta_2(x) - \delta_1(x) = \beta_0 + x^\top\beta,$$

where

- $\beta = \Sigma^{-1}(\mu_2 - \mu_1)$ is a $p$-vector of slopes, one per
feature, with no intercept entry, and
- $\beta_0 = \log(\pi_2/\pi_1) - \frac{1}{2}\left(\mu_2^\top \Sigma^{-1} \mu_2 - \mu_1^\top \Sigma^{-1} \mu_1\right)$
is the intercept.

This is again the form of a multiple logistic regression with $p$ features.
The decision boundary $\{x : \beta_0 + x^\top\beta = 0\}$ is a hyperplane in
the $p$-dimensional feature space — a line when $p = 2$.

The slopes are $\Sigma^{-1}(\mu_2 - \mu_1)$, not $\mu_2 - \mu_1$, so the
boundary is generally not perpendicular to the segment joining the two means:
$\Sigma^{-1}$ tilts it according to how the features covary. With equal
priors it still passes through the midpoint $(\mu_1 + \mu_2)/2$, since
substituting that point gives
$\frac{1}{2}\left(\mu_1 + \mu_2\right)^\top\Sigma^{-1}\left(\mu_2 - \mu_1\right) = \frac{1}{2}\left(\mu_2^\top \Sigma^{-1} \mu_2 - \mu_1^\top \Sigma^{-1} \mu_1\right)$.

The figure below uses $p = 2$, $\mu_1 = (-1, 0.25)^\top$,
$\mu_2 = (1, -0.25)^\top$, unit variances, and a correlation of $0.6$ between
the two features. Every tab marks the two class means, joined by a grey
segment. The first tab draws contours of each $f_k$. The other two
draw contours of the prior-weighted densities $\pi_k f_k(x)$, at the same
heights in both tabs, together with the decision boundary (long dashes); open
circles mark the points where a class-$1$ contour crosses the class-$2$
contour of the same height, and the dotted line in the third tab is the
equal-prior boundary.

::: {.callout-note collapse="true"}
### Code
```{r theory-2d}
#| output: false
mu2d_1  <- c(-1,  0.25)
mu2d_2  <- c( 1, -0.25)
Sigma2d <- matrix(c(1, 0.6, 0.6, 1), nrow = 2)
Sigma2d_inv <- solve(Sigma2d)

dmvnorm2 <- function(x1, x2, mu) {
  d <- cbind(x1 - mu[1], x2 - mu[2])
  q <- rowSums((d %*% Sigma2d_inv) * d)
  exp(-q / 2) / (2 * pi * sqrt(det(Sigma2d)))
}

grid2d <- expand_grid(x1 = seq(-4, 4, length.out = 240),
                      x2 = seq(-3.5, 3.5, length.out = 210)) |>
  mutate(f_1 = dmvnorm2(x1, x2, mu2d_1),
         f_2 = dmvnorm2(x1, x2, mu2d_2))

# Long grid of pi_k * f_k(x); prior weights of 1 give the unweighted f_k.
weighted_grid <- function(pi1, pi2) {
  grid2d |>
    transmute(x1, x2, `k = 1` = pi1 * f_1, `k = 2` = pi2 * f_2) |>
    pivot_longer(c(`k = 1`, `k = 2`), names_to = "class",
                 values_to = "density")
}

# Class densities: contours of f_k at Mahalanobis radii 1 and 2.
f_peak <- 1 / (2 * pi * sqrt(det(Sigma2d)))
density_levels <- f_peak * exp(-c(1, 2)^2 / 2)
# Prior-weighted densities: the same three heights in both prior tabs. The
# highest exceeds the peak of 0.2 * f_2, so that contour vanishes for class 2
# when pi_2 = 0.2.
weighted_levels <- c(0.005, 0.015, 0.045)

beta2d <- drop(Sigma2d_inv %*% (mu2d_2 - mu2d_1))
beta0_2d <- function(pi1, pi2) {
  log(pi2 / pi1) -
    0.5 * (sum(mu2d_2 * (Sigma2d_inv %*% mu2d_2)) -
           sum(mu2d_1 * (Sigma2d_inv %*% mu2d_1)))
}
# beta0 + beta1 x1 + beta2 x2 = 0  =>  x2 = -(beta0 + beta1 x1) / beta2
boundary_2d <- function(pi1, pi2) {
  c(intercept = -beta0_2d(pi1, pi2) / beta2d[2],
    slope = -beta2d[1] / beta2d[2])
}
line_equal   <- boundary_2d(0.5, 0.5)
line_unequal <- boundary_2d(pi_unequal[1], pi_unequal[2])

# Points where the class-1 and class-2 contours of the same height cross:
# where the boundary line, on which pi_1 f_1 = pi_2 f_2, meets the class-1
# contour at that height. Along the line x = p0 + s u, log(pi_1 f_1(x)) is
# quadratic in s.
contour_crossings <- function(pi1, pi2, levels) {
  p0 <- -beta0_2d(pi1, pi2) * beta2d / sum(beta2d^2)   # a point on the line
  u  <- c(beta2d[2], -beta2d[1])                        # its direction
  d0 <- p0 - mu2d_1
  a  <- 0.5 * sum(u * (Sigma2d_inv %*% u))
  b  <- sum(u * (Sigma2d_inv %*% d0))
  map(levels, \(L) {
    c0 <- 0.5 * sum(d0 * (Sigma2d_inv %*% d0)) - log(pi1 * f_peak / L)
    disc <- b^2 - 4 * a * c0
    if (disc < 0) return(NULL)
    s <- (-b + c(-1, 1) * sqrt(disc)) / (2 * a)
    tibble(x1 = p0[1] + s * u[1], x2 = p0[2] + s * u[2], level = L)
  }) |>
    list_rbind()
}
crossings_equal   <- contour_crossings(0.5, 0.5, weighted_levels)
crossings_unequal <- contour_crossings(pi_unequal[1], pi_unequal[2],
                                       weighted_levels)

# Angle, in degrees, between the boundary and the segment joining the means.
segment_2d <- mu2d_2 - mu2d_1
boundary_dir <- c(beta2d[2], -beta2d[1])
angle_2d <- acos(abs(sum(boundary_dir * segment_2d)) /
                   sqrt(sum(boundary_dir^2) * sum(segment_2d^2))) * 180 / pi

means2d <- tibble(class = c("k = 1", "k = 2"),
                  x1 = c(mu2d_1[1], mu2d_2[1]),
                  x2 = c(mu2d_1[2], mu2d_2[2]))

contour_plot <- function(long_grid, levels) {
  ggplot(long_grid, aes(x = x1, y = x2)) +
    geom_contour(aes(z = density, color = class, linetype = class),
                 breaks = levels, linewidth = 0.8) +
    geom_point(data = means2d, aes(color = class, shape = class), size = 3) +
    annotate("segment", x = mu2d_1[1], y = mu2d_1[2],
             xend = mu2d_2[1], yend = mu2d_2[2],
             color = "grey50", linewidth = 0.4) +
    scale_color_manual(values = class_colors) +
    coord_fixed(xlim = c(-4, 4), ylim = c(-3.5, 3.5)) +
    labs(x = expression(X[1]), y = expression(X[2]), color = "Class",
         linetype = "Class", shape = "Class")
}

# The boundary uses "longdash", distinct from both class-contour linetypes.
boundary_layer <- function(line, ...) {
  geom_abline(intercept = line["intercept"], slope = line["slope"], ...)
}
crossing_layer <- function(points) {
  geom_point(data = points, aes(x = x1, y = x2), inherit.aes = FALSE,
             shape = 21, size = 3, stroke = 1, fill = "white", color = "black")
}

theory_2d_densities <- contour_plot(weighted_grid(1, 1), density_levels) +
  labs(title = expression(paste("Contours of ", f[k](x),
                                ", sharing one covariance matrix")))

theory_2d_equal <- contour_plot(weighted_grid(0.5, 0.5), weighted_levels) +
  boundary_layer(line_equal, linewidth = 1, linetype = "longdash") +
  crossing_layer(crossings_equal) +
  labs(title = expression(paste("Contours of ", pi[k] * f[k](x), "; ",
                                pi[1] == 0.5, ", ", pi[2] == 0.5)))

theory_2d_unequal <- contour_plot(weighted_grid(pi_unequal[1], pi_unequal[2]),
                                  weighted_levels) +
  boundary_layer(line_equal, linewidth = 0.6, color = "grey50",
                 linetype = "dotted") +
  boundary_layer(line_unequal, linewidth = 1, linetype = "longdash") +
  crossing_layer(crossings_unequal) +
  labs(title = bquote(paste("Contours of ", pi[k] * f[k](x), "; ",
                            pi[1] == .(pi_unequal[1]), ", ",
                            pi[2] == .(pi_unequal[2]))))
```
:::

::: {.panel-tabset}
#### Class densities
```{r theory-2d-densities-figure}
#| echo: false
#| fig-width: 6
#| fig-height: 5
theory_2d_densities
```

#### Equal priors
```{r theory-2d-equal-figure}
#| echo: false
#| fig-width: 6
#| fig-height: 5
theory_2d_equal
```

#### Unequal priors
```{r theory-2d-unequal-figure}
#| echo: false
#| fig-width: 6
#| fig-height: 5
theory_2d_unequal
```
:::

The two sets of ellipses in the first tab are translates of one another. With
equal priors the boundary passes through the midpoint of the grey segment
joining the means but meets that segment at an angle of
`r round(angle_2d)` degrees rather than $90$: the positive correlation
between the features tilts it.

The boundary is the set of $x$ where $\pi_1 f_1(x) = \pi_2 f_2(x)$, so it
passes through every circled point, where the two weighted surfaces have the
same height. Shrinking $\pi_2$ to `r pi_unequal[2]` lowers the
class-$2$ surface: its contours shrink, and the one at height
`r max(weighted_levels)` disappears because
$\pi_2 f_2(x) \le `r sprintf("%.3f", pi_unequal[2] * f_peak)`$ everywhere.
The crossings, and the boundary through them, move away from $\mu_1$ toward
$\mu_2$, and the line stays parallel to the equal-prior boundary because the
prior enters only $\beta_0$.

### Pooled covariance estimate

The estimates generalize directly. The priors are still $\hat\pi_k = n_k/n$,
and with $x_i = (x_{i1}, \ldots, x_{ip})^\top$ now the $p$-vector of feature
values for observation $i$ (again without the leading $1$),

$$\hat\mu_k = \frac{1}{n_k}\sum_{i:\, y_i = k} x_i, \qquad \hat\Sigma = \frac{1}{n - K}\sum_{k=1}^K \sum_{i:\, y_i = k}\left(x_i - \hat\mu_k\right)\left(x_i - \hat\mu_k\right)^\top,$$

where

- $\hat\mu_k$ is the vector of class-$k$ sample means, one per feature, and
- $\hat\Sigma$ is the **pooled covariance matrix**, the sum of each class's
matrix of cross-products of deviations from its own mean, divided by $n - K$;
as in one dimension, it is the maximum likelihood estimate rescaled by
$n/(n - K)$ to make it unbiased.

Substituting $\hat\pi_k$, $\hat\mu_k$, and $\hat\Sigma$ into $\delta_k(x)$ gives
the estimated discriminant $\hat\delta_k(x)$, the estimated posterior
probabilities
$\hat p_k(x) = e^{\hat\delta_k(x)} / \sum_l e^{\hat\delta_l(x)}$, and the LDA
classifier, which assigns each observation to the class with the largest
$\hat\delta_k(x)$.

## Classifier evaluation

An error rate counts every mistake the same way. Failing to flag a customer
who goes on to default and flagging a customer who does not are different
mistakes with different costs, and a classifier can have a small error rate
while making almost every mistake of one kind. The classification lecture
introduced the confusion matrix, which separates the mistakes; this section
develops it in full for two classes, extends it to $K$ classes, and adds the
ROC curve and AUC, which summarize a two-class classifier at every threshold.

### Confusion matrix

With two classes, call the class of interest **positive** and the other
**negative**. For a classifier that produces an estimated probability
$\hat p(x)$ of the positive class, a **threshold** $t$ turns probabilities into
classes: predict positive when $\hat p(x) > t$.

As in the classification lecture, a confusion matrix cross-tabulates the true
class against the predicted class:

::: {.callout-note collapse="true"}
### Code
```{r confusion-template}
#| output: false
confusion_template <- tibble(
  Truth = c("Positive", "Negative"),
  `Predicted positive` = c("True positive (TP)", "False positive (FP)"),
  `Predicted negative` = c("False negative (FN)", "True negative (TN)")
) |>
  knitr::kable(align = "lcc")
```
:::

```{r confusion-template-table}
#| echo: false
confusion_template
```

where

- $TP$ is the number of positive observations predicted positive,
- $FN$ is the number of positive observations predicted negative,
- $FP$ is the number of negative observations predicted positive, and
- $TN$ is the number of negative observations predicted negative.

The diagonal holds the correct classifications, so the error rate is
$(FP + FN)/n$.

### Sensitivity and specificity

Dividing each row by its total gives two rates that condition on the true
class:

$$\text{Sensitivity} = \frac{TP}{TP + FN}, \qquad \text{Specificity} = \frac{TN}{TN + FP},$$

where

- **sensitivity**, also called **recall** or the **true positive rate**, is the
proportion of positive observations the classifier catches, an estimate of
$P\left(\hat y(X) = \text{positive} \mid Y = \text{positive}\right)$, and
- **specificity**, also called the **true negative rate**, is the proportion of
negative observations the classifier correctly leaves alone, an estimate of
$P\left(\hat y(X) = \text{negative} \mid Y = \text{negative}\right)$.

Their complements are error rates within each true class:
$1 - \text{sensitivity} = FN/(TP + FN)$ is the **false negative rate**, and
$1 - \text{specificity} = FP/(TN + FP)$ is the **false positive rate**.

### Multiple classes

With $K$ classes the confusion matrix is $K \times K$, with true classes in
the rows and predicted classes in the columns:

::: {.callout-note collapse="true"}
### Code
```{r confusion-template-k}
#| output: false
k_labels <- c("1", "2", "$\\vdots$", "$K$")
cell <- function(k, j) {
  ifelse(k == "$\\vdots$" | j == "$\\vdots$", "$\\vdots$",
         sprintf("$n_{%s%s}$", gsub("\\$", "", k), gsub("\\$", "", j)))
}
confusion_template_k <- tibble(Truth = k_labels) |>
  mutate(`Predicted 1` = cell(Truth, "1"),
         `Predicted 2` = cell(Truth, "2"),
         `$\\cdots$`   = ifelse(Truth == "$\\vdots$", "$\\ddots$", "$\\cdots$"),
         `Predicted $K$` = cell(Truth, "$K$")) |>
  knitr::kable(align = "lcccc", escape = FALSE)
```
:::

```{r confusion-template-k-table}
#| echo: false
confusion_template_k
```

where

- $n_{kj}$ is the number of observations in true class $k$ predicted to be in
class $j$, so the diagonal entries $n_{kk}$ are the correct classifications.

The error rate is $1 - \sum_{k=1}^K n_{kk} / n$. Sensitivity and specificity
are defined one class at a time, **one versus the rest**: treat class $k$ as
positive and the other $K - 1$ classes together as negative, which collapses
the table to two-by-two. Class $k$'s rates are

$$\text{Sensitivity}_k = \frac{n_{kk}}{\sum_{j=1}^K n_{kj}}, \qquad \text{Specificity}_k = \frac{\sum_{l \ne k} \sum_{j \ne k} n_{lj}}{\sum_{l \ne k} \sum_{j=1}^K n_{lj}},$$

where

- the sensitivity's denominator is the number of observations in class $k$,
and its numerator the number of those predicted to be in class $k$, and
- the specificity's denominator is the number of observations in the other
classes, and its numerator the number of those predicted to be in some class
other than $k$, whether or not it is their own.

With $K$ classes the threshold $t = 0.5$ generalizes to assigning each
observation to the class with the largest estimated posterior probability;
with $K = 2$ the two rules coincide. Other thresholds, and the ROC curve and
AUC built from them, are two-class tools, and the rest of this section
returns to two classes. Logistic regression as developed in this course is
also a two-class method; its multi-class extension, multinomial logistic
regression, is not covered in this course.

::: {.callout-note collapse="true"}
### Beyond this course
ROC curves and AUC extend to $K > 2$ classes by averaging the AUCs of the $K$
one-versus-rest problems, or of all $K(K-1)/2$ pairs of classes.
:::

### Classification threshold

The threshold $t = 0.5$ is the Bayes classifier's: if $\hat p(x)$ equalled
the true $p(x)$, it would minimize the expected error rate. An estimated
$\hat p(x)$ carries no such guarantee, and even for the Bayes classifier the
overall error rate need not be the quantity of interest.

Lowering $t$ enlarges the set $\left\{x : \hat p(x) > t\right\}$ of feature
values classified positive, so every
observation predicted positive at the old threshold is still predicted
positive at the new one. Consequently $TP$ and $FP$ can only grow while $FN$
and $TN$ can only shrink: sensitivity rises or stays the same, and
specificity falls or stays the same. Raising $t$ does the reverse. No
threshold improves both, so choosing $t$ means choosing a tradeoff between the
two kinds of mistake.

::: {.callout-note collapse="true"}
### Beyond this course
If the two mistakes have known costs, $c_{FN}$ for a missed positive and
$c_{FP}$ for a false alarm, the expected cost of predicting positive at $x$ is
$c_{FP}\left[1 - p(x)\right]$ and of predicting negative is $c_{FN}\, p(x)$.
Predicting positive is the cheaper choice exactly when
$p(x) > c_{FP}/(c_{FP} + c_{FN})$, so that ratio is the cost-minimizing
threshold. Equal costs recover $t = 0.5$.
:::

### ROC curve

The **ROC curve** (receiver operating characteristic curve) plots sensitivity
against $1 - \text{specificity}$ — the true positive rate against the false
positive rate — at every threshold. At $t = 1$ nothing is predicted positive,
giving the point $(0, 0)$; at $t = 0$ everything is, giving $(1, 1)$; and the
curve traces every tradeoff in between.

For the two-class, one-feature model of the decision-boundary figure, with
equal priors and class $2$ positive, $p(x) > t$ is equivalent to $x > c$
for a cutoff $c$ that decreases as $t$ decreases, because the posterior is
increasing in $x$. Sensitivity is then the area under $f_2$ to the right of
$c$, and $1 - \text{specificity}$ is the area under $f_1$ to the right of $c$.
The slider below moves the threshold. A line above the plots reports $t$,
the cutoff $c$, sensitivity, and specificity. The left panel draws $f_1$ and
$f_2$ with a dotted vertical rule at $c$ and shades the areas that determine
sensitivity (darker) and the false positive rate (lighter), and the right
panel marks where the classifier lands on the ROC curve, with a dashed
diagonal for reference.

::: {.callout-note collapse="true"}
### Code
```{r theory-roc}
#| output: false
# Empirical ROC curve: every distinct score is a cutoff; classify positive
# when score >= cutoff.
roc_curve <- function(score, positive) {
  ord <- order(score, decreasing = TRUE)
  s   <- score[ord]
  pos <- positive[ord]
  keep <- c(diff(s) != 0, TRUE)
  tibble(fpr = c(0, cumsum(!pos)[keep] / sum(!positive)),
         tpr = c(0, cumsum(pos)[keep] / sum(positive)))
}

# Area under an ROC curve (sorted by fpr) by the trapezoid rule.
auc_trapezoid <- function(roc) {
  with(roc, sum(diff(fpr) * (head(tpr, -1) + tail(tpr, -1)) / 2))
}

# AUC as the proportion of (positive, negative) pairs ordered correctly.
auc_pairs <- function(score, positive) {
  s_pos <- score[positive]
  s_neg <- score[!positive]
  mean(outer(s_pos, s_neg, ">")) + 0.5 * mean(outer(s_pos, s_neg, "=="))
}

# Equal-prior posterior: log-odds = beta1 * x with beta1 = (mu_2 - mu_1) / sigma^2.
beta1_theory <- (mu_2 - mu_1) / sigma^2
cutoff_for <- function(t) qlogis(t) / beta1_theory

roc_theory <- tibble(c = seq(-8, 8, length.out = 4000)) |>
  mutate(fpr = 1 - pnorm(c, mu_1, sigma),
         tpr = 1 - pnorm(c, mu_2, sigma))

auc_theory_formula <- pnorm((mu_2 - mu_1) / (sigma * sqrt(2)))
roc_theory_sorted  <- arrange(roc_theory, fpr)
auc_theory_trap    <- auc_trapezoid(roc_theory_sorted)

# Everything the slider needs, indexed by t = 0.01, 0.02, ..., 0.99, is
# computed here and passed to the page, which only looks values up.
slider_thresholds <- tibble(t = (1:99) / 100) |>
  mutate(cutoff = cutoff_for(t),
         fpr = 1 - pnorm(cutoff, mu_1, sigma),
         tpr = 1 - pnorm(cutoff, mu_2, sigma))

slider_densities <- theory_grid |>
  mutate(`k = 1 (negative)` = dnorm(x, mu_1, sigma),
         `k = 2 (positive)` = dnorm(x, mu_2, sigma)) |>
  pivot_longer(-x, names_to = "class", values_to = "density")
```
:::

```{r theory-roc-data}
#| echo: false
ojs_define(slider_thresholds = slider_thresholds,
           slider_densities  = slider_densities,
           slider_roc        = select(slice(roc_theory, seq(1, n(), by = 8)), fpr, tpr),
           slider_colors     = unname(class_colors))
```

```{ojs}
//| label: theory-roc-slider
//| echo: false
viewof t = Inputs.range([0.01, 0.99], {value: 0.5, step: 0.01, label: "Threshold t"})
```

```{ojs}
//| label: theory-roc-slider-plots
//| echo: false
{
  const thresholds = transpose(slider_thresholds);
  const densities = transpose(slider_densities);
  const roc = transpose(slider_roc);
  const now = thresholds[Math.round(t * 100) - 1];
  const classes = ["k = 1 (negative)", "k = 2 (positive)"];

  const densityPlot = Plot.plot({
    width: 340, height: 300,
    x: {label: "x"},
    y: {label: "Density fₖ(x)"},
    color: {domain: classes, range: slider_colors, legend: true},
    marks: [
      Plot.areaY(densities.filter(d => d.x > now.cutoff && d.class === classes[0]),
                 {x: "x", y1: 0, y2: "density", fill: "class", fillOpacity: 0.15}),
      Plot.areaY(densities.filter(d => d.x > now.cutoff && d.class === classes[1]),
                 {x: "x", y1: 0, y2: "density", fill: "class", fillOpacity: 0.45}),
      Plot.line(densities.filter(d => d.class === classes[0]),
                {x: "x", y: "density", stroke: "class", strokeWidth: 2}),
      Plot.line(densities.filter(d => d.class === classes[1]),
                {x: "x", y: "density", stroke: "class", strokeDasharray: "6,4",
                 strokeWidth: 2}),
      Plot.ruleX([now.cutoff], {strokeDasharray: "2,3"})
    ]
  });

  const rocPlot = Plot.plot({
    width: 300, height: 300,
    x: {domain: [0, 1], label: "False positive rate (1 − specificity)"},
    y: {domain: [0, 1], label: "True positive rate (sensitivity)"},
    marks: [
      Plot.line([[0, 0], [1, 1]], {stroke: "gray", strokeDasharray: "4,4"}),
      Plot.line(roc, {x: "fpr", y: "tpr", strokeWidth: 2}),
      Plot.dot([now], {x: "fpr", y: "tpr", r: 6, fill: "white", stroke: "black",
                       strokeWidth: 2})
    ]
  });

  return html`<p style="text-align: center"><b>t = ${t.toFixed(2)}</b>,
    cutoff c = ${now.cutoff.toFixed(2)}; sensitivity = ${now.tpr.toFixed(2)},
    specificity = ${(1 - now.fpr).toFixed(2)}</p>
    <div style="display: flex; flex-wrap: wrap; gap: 1rem; justify-content: center">
      ${densityPlot}${rocPlot}
    </div>`;
}
```

Moving the slider from $t = 0.9$ to $t = 0.1$ slides the cutoff left, so the
shaded area under $f_2$ (sensitivity) and the shaded area under $f_1$ (false
positive rate) both grow, and the point climbs along the curve toward $(1, 1)$. The
dashed diagonal is the ROC curve of a classifier that ignores $x$ and
predicts positive at random: its true and false positive rates are equal at
every threshold. A useful classifier's curve bows toward the top-left corner,
where sensitivity is high and the false positive rate is low.

The ROC curve depends on a classifier only through how it *orders* the
observations. Any strictly increasing transformation of the score — $x$,
$\hat\delta_2(x) - \hat\delta_1(x)$, or $\hat p(x)$ — sorts the observations
identically, produces the same sequence of confusion matrices, and so traces
the same curve.

### AUC

The **AUC** is the area under the ROC curve, a single number summarizing the
curve over all thresholds at once. It has a probabilistic interpretation.
Let $S_+$ be the score of a randomly chosen positive observation and $S_-$ the
score of an independently chosen negative one, and let $g_-$ be the density of
$S_-$.
At cutoff $c$ the true positive rate is $P(S_+ > c)$ and the false positive
rate is $P(S_- > c)$, whose derivative with respect to $c$ is $-g_-(c)$. As
$c$ runs from $+\infty$ down to $-\infty$ the false positive rate runs from
$0$ to $1$, so

$$\text{AUC} = \int_0^1 \text{TPR}\; d\,\text{FPR} = \int_{-\infty}^{\infty} P(S_+ > c)\, g_-(c)\, dc = P(S_+ > S_-),$$

where

- $S_+$ and $S_-$ are the classifier's scores for a random positive and a
random negative observation, and
- the final equality conditions on $S_- = c$ and integrates over its density.

The AUC is the probability that the classifier scores a randomly chosen
positive observation above a randomly chosen negative one. A classifier that
ignores the features has $\text{AUC} = 0.5$; one that ranks every positive
above every negative has $\text{AUC} = 1$.

In the one-feature Gaussian model the score can be taken as $x$ itself, so
$S_+ - S_- \sim N\left(\mu_2 - \mu_1, 2\sigma^2\right)$ and

$$\text{AUC} = P(S_+ - S_- > 0) = \Phi\left(\frac{\Delta}{\sqrt{2}}\right), \qquad \Delta = \frac{\mu_2 - \mu_1}{\sigma},$$

where

- $\Phi$ is the standard normal cumulative distribution function, and
- $\Delta$ is the separation of the class means in standard deviations.

For the curve plotted above, $\Delta = `r (mu_2 - mu_1) / sigma`$ and
$\text{AUC} = \Phi\left(`r (mu_2 - mu_1) / sigma`/\sqrt{2}\right) = `r sprintf("%.5f", auc_theory_formula)`$.

With training data the same probability is estimated by comparing every
positive observation with every negative one,

$$\widehat{\text{AUC}} = \frac{1}{n_+ n_-}\sum_{i:\, y_i = \text{positive}}\ \sum_{j:\, y_j = \text{negative}}\left[\mathrm{I}(s_i > s_j) + \tfrac{1}{2}\mathrm{I}(s_i = s_j)\right],$$

where

- $n_+$ and $n_-$ are the numbers of positive and negative training
observations,
- $s_i$ is the classifier's score for observation $i$, and
- tied scores count as half a correctly ordered pair.

This equals the area under the empirical ROC curve computed by the trapezoid
rule.

The left panel below shades the area under the ROC curve of the previous
subsection, labeled with that area, and adds a dashed diagonal for reference.
The right panel uses one simulated training set from the
two-feature model of the multiple-feature section, with class $2$ positive,
and draws the empirical ROC curves of three LDA classifiers fit to it — one
using $X_1$ only, one using $X_2$ only, and one using both features — and of
a classifier that ignores the features, each labeled with its AUC.

::: {.callout-note collapse="true"}
### Code
```{r auc-figure}
#| output: false
auc_area_plot <- ggplot(roc_theory_sorted, aes(x = fpr, y = tpr)) +
  geom_area(fill = "grey70", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "grey40",
              linetype = "dashed") +
  geom_line(linewidth = 0.9) +
  annotate("text", x = 0.65, y = 0.3,
           label = sprintf("AUC = %.5f", auc_theory_trap)) +
  coord_fixed() +
  labs(x = "False positive rate (1 - specificity)",
       y = "True positive rate (sensitivity)",
       title = "Area under the ROC curve")

# One training set from the two-feature model, n_sim observations per class.
n_sim <- 1000
set.seed(20261001)
sim_2d <- bind_rows(
  as_tibble(MASS::mvrnorm(n_sim, mu2d_1, Sigma2d), .name_repair = ~ c("X1", "X2")) |>
    mutate(class = "1"),
  as_tibble(MASS::mvrnorm(n_sim, mu2d_2, Sigma2d), .name_repair = ~ c("X1", "X2")) |>
    mutate(class = "2")
) |>
  mutate(class = factor(class))
sim_positive <- sim_2d$class == "2"

# Score of each classifier: its posterior probability of class 2. The
# classifier that ignores the features gives every observation the same score.
sim_score <- function(model_formula) {
  fit <- MASS::lda(model_formula, data = sim_2d)
  predict(fit, newdata = sim_2d)$posterior[, "2"]
}
sim_scores <- list(
  x1   = sim_score(class ~ X1),
  x2   = sim_score(class ~ X2),
  both = sim_score(class ~ X1 + X2),
  none = rep(0.5, nrow(sim_2d))
)
sim_auc <- map_dbl(sim_scores, \(s) auc_trapezoid(roc_curve(s, sim_positive)))

# Legend entries as plotmath, each with its AUC.
sim_labels <- parse(text = c(
  sprintf('X[1] ~ "only:" ~ AUC == "%.3f"', sim_auc[["x1"]]),
  sprintf('X[2] ~ "only:" ~ AUC == "%.3f"', sim_auc[["x2"]]),
  sprintf('X[1] ~ "and" ~ X[2] * ":" ~ AUC == "%.3f"', sim_auc[["both"]]),
  sprintf('"No features:" ~ AUC == "%.3f"', sim_auc[["none"]])
))
sim_roc <- imap(sim_scores, \(s, nm) mutate(roc_curve(s, sim_positive),
                                            classifier = nm)) |>
  list_rbind() |>
  mutate(classifier = factor(classifier, levels = names(sim_scores)))

auc_classifier_plot <- ggplot(sim_roc, aes(x = fpr, y = tpr,
                                           color = classifier,
                                           linetype = classifier)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("#E69F00", "#56B4E9", "#009E73", "grey40"),
                     labels = sim_labels) +
  scale_linetype_manual(values = c("dashed", "dotdash", "solid", "dotted"),
                        labels = sim_labels) +
  coord_fixed() +
  labs(x = "False positive rate (1 - specificity)",
       y = "True positive rate (sensitivity)", color = NULL, linetype = NULL,
       title = "ROC curves, same data") +
  theme(legend.position = "inside", legend.position.inside = c(0.97, 0.03),
        legend.justification = c(1, 0),
        legend.text = element_text(hjust = 0),
        legend.key.width = unit(1.6, "lines"))
```
:::

```{r auc-figure-display}
#| echo: false
#| layout-ncol: 2
#| fig-width: 4.5
#| fig-height: 4.5
auc_area_plot
auc_classifier_plot
```

The left panel shades the area under the slider's ROC curve,
`r sprintf("%.5f", auc_theory_trap)` by the trapezoid rule, in agreement with
the formula. In the right panel, because the AUC needs no threshold, it ranks
the classifiers directly: the curve that uses both features lies on or above
both single-feature curves and has the largest AUC. Alone, $X_2$ separates the classes only weakly, since its class means differ
by `r abs(diff(c(mu2d_1[2], mu2d_2[2])))` standard deviations, but combined
with the correlated $X_1$ it adds information that $X_1$ lacks. The
$X_1$-only AUC is close to the population value from the one-feature formula,
$\Phi\left(`r abs(diff(c(mu2d_1[1], mu2d_2[1])))`/\sqrt{2}\right) = `r sprintf("%.3f", pnorm(abs(diff(c(mu2d_1[1], mu2d_2[1]))) / sqrt(2)))`$.

The three LDA curves do not cross one another, so the AUC's ranking of them
holds at every false positive rate. When two classifiers' ROC curves do cross, each has the higher true
positive rate over part of the range of false positive rates, and the AUC
averages the true positive rate over all false positive rates, so the
classifier with the larger AUC can still be the worse one over the false
positive rates that matter for a given application.

## Credit card default classification

**Question.** Can LDA identify credit card customers who will default from
their balance and income, and how does it compare with logistic regressions
fit to the same customers and features?

These customers were analyzed in the logistic regression lectures: the
[Simple Logistic Regression](10-logistic-regression.qmd) lecture modeled
default from `balance`, the
[Multiple Logistic Regression](20-multiple-logistic-regression.qmd) lecture
from `balance` and `student`, and the
[Flexible Logistic Regression](30-flexible-logistic-regression.qmd) lecture
added interactions, including one between `balance` and `income`.

**Data.** `ISLR2::Default` is a simulated data set of
`r format(nrow(Default), big.mark = ",")` credit card customers. The response is `default`, with `Yes` the positive class. The
explanatory variables are `balance`, the customer's monthly credit card
balance in dollars, and `income`, the customer's income in dollars, the two
quantitative explanatory variables; they are used because LDA models each
class's features with a Gaussian density. The features are the explanatory
variables themselves. `income` enters the models in dollars as recorded, as
in the flexible logistic regression lecture; its coefficients are reported
here per \$1,000 of income. `student` (Yes/No) is used in the last subsection
only to split the customers.

### Class summaries

The quantities LDA estimates — class proportions, class means, and
within-class spreads — can all be computed before fitting anything. The table
summarizes each class.

::: {.callout-note collapse="true"}
### Code
```{r default-setup}
#| output: false
Default <- Default |>
  mutate(default01   = as.numeric(default == "Yes"),
         default_jit = default01 + runif(n(), -0.03, 0.03))

class_summary <- Default |>
  group_by(default) |>
  summarize(n = n(),
            prop = n() / nrow(Default),
            mean_balance = mean(balance),
            sd_balance = sd(balance),
            mean_income = mean(income),
            sd_income = sd(income),
            n_zero = sum(balance == 0),
            cor_balance_income = cor(balance, income),
            .groups = "drop")

dollars <- function(v) format(round(v), big.mark = ",")
# Axis labels such as $60,000, without the scales package.
dollar_labels <- function(v) {
  ifelse(is.na(v), NA,
         paste0("$", format(v, big.mark = ",", scientific = FALSE, trim = TRUE)))
}

class_summary_table <- class_summary |>
  transmute(default,
            n = format(n, big.mark = ","),
            prop = sprintf("%.4f", prop),
            mean_balance = dollars(mean_balance),
            sd_balance = dollars(sd_balance),
            mean_income = dollars(mean_income),
            sd_income = dollars(sd_income)) |>
  knitr::kable(col.names = c("Default", "Customers", "Proportion",
                             "Mean balance (\\$)", "SD of balance (\\$)",
                             "Mean income (\\$)", "SD of income (\\$)"),
               align = "ccccccc")

summary_of <- function(k, col) class_summary[[col]][class_summary$default == k]
```
:::

```{r class-summary-table}
#| echo: false
class_summary_table
```

Only `r format(summary_of("Yes", "n"), big.mark = ",")` customers, a proportion
of `r sprintf("%.4f", summary_of("Yes", "prop"))`, defaulted. Defaulters carry
an average balance of \$`r dollars(summary_of("Yes", "mean_balance"))`
against \$`r dollars(summary_of("No", "mean_balance"))` for non-defaulters,
and the standard deviations of balance differ between classes,
\$`r dollars(summary_of("No", "sd_balance"))` against
\$`r dollars(summary_of("Yes", "sd_balance"))`, where LDA assumes one shared
variance. The two classes' incomes are much closer: defaulters average
\$`r dollars(summary_of("Yes", "mean_income"))` against
\$`r dollars(summary_of("No", "mean_income"))`, with similar standard
deviations.

### Balance alone

Before any model is fit, the histograms below show balance within each
default class, one panel per class, on the density scale.

::: {.callout-note collapse="true"}
### Code
```{r default-hist}
#| output: false
balance_hist_plot <- ggplot(Default, aes(x = balance)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 100,
                 boundary = 0, fill = "grey80", color = "grey55") +
  facet_wrap(~ default, ncol = 1, labeller = label_both) +
  scale_x_continuous(labels = dollar_labels) +
  labs(x = "Credit card balance ($)", y = "Density",
       title = "ISLR2::Default, balance within each default class")
```
:::

```{r default-hist-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
balance_hist_plot
```

The defaulters' histogram sits well to the right of the non-defaulters'. The
non-defaulter histogram also has a spike at zero —
`r format(summary_of("No", "n_zero"), big.mark = ",")` customers carry no
balance, and `r if (summary_of("Yes", "n_zero") == 0) "no defaulter does" else paste(summary_of("Yes", "n_zero"), "defaulters do")`
— which no Gaussian density reproduces.

`MASS::lda()` takes the same formula interface as `glm()`. Fitting `default`
on `balance`:

```{r balance-lda}
balance_lda <- MASS::lda(default ~ balance, data = Default)
balance_lda
```

::: {.callout-note collapse="true"}
### Code
```{r lda-by-hand}
#| output: false
# Estimates of pi_k, mu_k, Sigma and the implied log posterior odds
# beta0 + x' beta of the second class, computed from the formulas above.
lda_by_hand <- function(X, y) {
  X <- as.matrix(X)
  classes <- levels(y)
  n <- nrow(X)
  K <- length(classes)
  pi_hat <- as.numeric(table(y)) / n
  mu_hat <- do.call(rbind, lapply(classes, function(k) {
    colMeans(X[y == k, , drop = FALSE])
  }))
  rownames(mu_hat) <- classes
  Sigma_hat <- Reduce(`+`, lapply(classes, function(k) {
    Z <- sweep(X[y == k, , drop = FALSE], 2, mu_hat[k, ])
    crossprod(Z)
  })) / (n - K)
  beta  <- solve(Sigma_hat, mu_hat[2, ] - mu_hat[1, ])
  beta0 <- log(pi_hat[2] / pi_hat[1]) -
    0.5 * (sum(mu_hat[2, ] * solve(Sigma_hat, mu_hat[2, ])) -
           sum(mu_hat[1, ] * solve(Sigma_hat, mu_hat[1, ])))
  list(pi = pi_hat, mu = mu_hat, Sigma = Sigma_hat,
       beta0 = beta0, beta = beta)
}

hand_1 <- lda_by_hand(Default["balance"], Default$default)
sigma_hat_1 <- sqrt(hand_1$Sigma[1, 1])
x_star_lda  <- -hand_1$beta0 / hand_1$beta[1]

balance_glm <- glm(default ~ balance, data = Default, family = "binomial")
b_glm_1     <- coef(balance_glm)
x_star_glm  <- -b_glm_1[1] / b_glm_1[2]

post_1 <- predict(balance_lda)$posterior[, "Yes"]
max_diff_1 <- max(abs(post_1 - plogis(hand_1$beta0 +
                                       hand_1$beta[1] * Default$balance)))

coef_table_1 <- tibble(
  term = c("Intercept $\\hat\\beta_0$", "`balance` $\\hat\\beta_1$"),
  lda = c(hand_1$beta0, hand_1$beta[1]),
  glm = unname(b_glm_1)
) |>
  mutate(across(c(lda, glm), \(v) sprintf("%.6f", v))) |>
  knitr::kable(col.names = c("Coefficient", "LDA", "Logistic regression"),
               align = "lcc")
```
:::

Substituting the estimates into the log posterior odds
$\beta_0 + \beta_1 x$ derived above gives LDA's coefficients, shown beside
those of the simple logistic regression fit to the same customers:

```{r coef-table-1}
#| echo: false
coef_table_1
```

The first tab below shows the fitted Gaussian class-conditional densities
over each class's histogram. The second shows the same densities weighted by
the estimated priors, and the third the two estimated discriminant functions,
each with a dotted line at LDA's boundary. The fourth shows the
posterior probability of default from LDA and from logistic regression, with
each customer's jittered outcome and a grey line at $0.5$.

::: {.callout-note collapse="true"}
### Code
```{r balance-tabset}
#| output: false
balance_grid <- tibble(balance = seq(0, 2800, length.out = 500))

gauss_curves <- tibble(default = factor(c("No", "Yes"))) |>
  mutate(mu = hand_1$mu[, 1], prior = hand_1$pi) |>
  cross_join(balance_grid) |>
  mutate(f = dnorm(balance, mu, sigma_hat_1),
         weighted = prior * f,
         delta = (log(prior) - mu^2 / (2 * sigma_hat_1^2)) +
                 mu / sigma_hat_1^2 * balance)

gauss_plot <- balance_hist_plot +
  geom_line(data = gauss_curves, aes(y = f, color = default,
                                     linetype = default),
            linewidth = 0.9) +
  scale_color_manual(values = class_colors) +
  labs(color = "Default", linetype = "Default",
       title = expression(paste("Histograms with fitted ",
                                N(hat(mu)[k], hat(sigma)^2), " densities")))

weighted_default_plot <- ggplot(gauss_curves,
                                aes(x = balance, y = weighted,
                                    color = default, linetype = default)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = x_star_lda, linetype = "dotted") +
  scale_color_manual(values = class_colors) +
  scale_x_continuous(labels = dollar_labels) +
  labs(x = "Credit card balance ($)",
       y = expression(hat(pi)[k] * hat(f)[k](x)),
       color = "Default", linetype = "Default",
       title = "Prior-weighted densities; dotted line at the LDA boundary")

delta_default_plot <- ggplot(gauss_curves,
                             aes(x = balance, y = delta,
                                 color = default, linetype = default)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = x_star_lda, linetype = "dotted") +
  scale_color_manual(values = class_colors) +
  scale_x_continuous(labels = dollar_labels) +
  labs(x = "Credit card balance ($)", y = expression(hat(delta)[k](x)),
       color = "Default", linetype = "Default",
       title = "Estimated discriminant functions")

posterior_curves <- balance_grid |>
  mutate(LDA = plogis(hand_1$beta0 + hand_1$beta[1] * balance),
         `Logistic regression` = predict(balance_glm,
                                         newdata = balance_grid,
                                         type = "response")) |>
  pivot_longer(-balance, names_to = "model", values_to = "p_hat")

posterior_default_plot <- ggplot(posterior_curves,
                                 aes(x = balance, y = p_hat)) +
  geom_point(data = Default, aes(y = default_jit), alpha = 0.08,
             size = 0.7) +
  geom_hline(yintercept = 0.5, color = "grey60") +
  geom_line(aes(color = model, linetype = model), linewidth = 0.9) +
  scale_color_manual(values = model_colors) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_x_continuous(labels = dollar_labels) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Model", linetype = "Model",
       title = "Posterior probability of default")
```
:::

::: {.panel-tabset}
#### Gaussian densities
```{r balance-gauss-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
gauss_plot
```

#### Prior-weighted densities
```{r balance-weighted-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
weighted_default_plot
```

#### Discriminant functions
```{r balance-delta-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
delta_default_plot
```

#### Posterior
```{r balance-posterior-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
posterior_default_plot
```
:::

The **prior probabilities** are $\hat\pi_{\text{No}} = `r sprintf("%.4f", hand_1$pi[1])`$
and $\hat\pi_{\text{Yes}} = `r sprintf("%.4f", hand_1$pi[2])`$, the class
proportions in the training data. They are the posterior probabilities LDA
would assign with no feature information. A classifier that used only them
would predict `No` for every customer and misclassify the
`r sprintf("%.2f", 100 * hand_1$pi[2])`% who defaulted; that error rate is the
one any classifier using features has to beat.

The **group means** are $\hat\mu_{\text{No}} = `r gsub(",", "{,}", format(round(hand_1$mu[1, 1], 2), nsmall = 2, big.mark = ","))`$
and $\hat\mu_{\text{Yes}} = `r gsub(",", "{,}", format(round(hand_1$mu[2, 1], 2), nsmall = 2, big.mark = ","))`$
dollars, the mean balances of the class summary table. The pooled
standard deviation is $\hat\sigma = `r sprintf("%.2f", sigma_hat_1)`$, between
the two class-specific standard deviations. In the Gaussian densities tab, the
fitted density for non-defaulters misses the spike at zero, and the one for
defaulters, forced to share $\hat\sigma$ with the larger class, is wider than
their histogram: the two departures the histograms and class summaries
flagged.

The **coefficient of the linear discriminant**,
`r signif(balance_lda$scaling[1, 1], 4)`, equals
$1/\hat\sigma = `r signif(1 / sigma_hat_1, 4)`$: `MASS::lda()`
reports the discriminant direction rescaled so that the discriminant has
variance $1$ within each class, and with one feature the direction itself is
just "larger balance points toward `Yes`."

The coefficient table puts LDA on the same log-odds scale as logistic
regression. The intercept $\hat\beta_0 = `r sprintf("%.3f", hand_1$beta0)`$ is
LDA's estimated log posterior odds of default for a customer carrying no
balance, a posterior probability of
`r format(signif(plogis(hand_1$beta0), 2), scientific = FALSE)`. Each additional
\$100 of balance adds `r sprintf("%.3f", 100 * hand_1$beta[1])` to the log
posterior odds, against `r sprintf("%.3f", 100 * b_glm_1[2])` in the logistic
regression. Applying the logistic function to
$\hat\beta_0 + \hat\beta_1 x$ reproduces the posterior probabilities returned
by `predict()` to within $`r sci(max_diff_1, 1)`$.

The decision boundary is
$\hat x^* = -\hat\beta_0/\hat\beta_1 = \$`r gsub(",", "{,}", format(round(x_star_lda), big.mark = ","))`$
for LDA and \$`r format(round(x_star_glm), big.mark = ",")` for logistic
regression. Both lie far above the midpoint of the class means,
\$`r format(round(mean(hand_1$mu[, 1])), big.mark = ",")`, for different
reasons. In LDA's boundary formula the shift comes from the prior
term
$\hat\sigma^2\log\left(\hat\pi_{\text{No}}/\hat\pi_{\text{Yes}}\right)/\left(\hat\mu_{\text{Yes}} - \hat\mu_{\text{No}}\right)$,
which adds \$`r format(round(x_star_lda - mean(hand_1$mu[, 1])), big.mark = ",")`
because $\hat\pi_{\text{Yes}}$ is so small. The prior-weighted
densities tab shows the effect: weighting by the priors shrinks the defaulter
curve to a small bump that rises above the non-defaulter curve only past the
dotted line at $\hat x^*$, deep in the right tail where both are small. The
discriminant functions tab shows two straight lines crossing at that same
balance. Logistic regression has no priors or class means; its boundary is
high because defaults are rare among customers at all but the largest
balances, so its fitted probability reaches $0.5$ only there.

In the posterior tab, the LDA curve is a logistic curve, slightly flatter than
the logistic regression fit and crossing $0.5$ a little further right, as its
smaller slope and higher boundary imply.

::: {.callout-note collapse="true"}
### Beyond this course
The "coefficients of linear discriminants" printed by `MASS::lda()` come from
R. A. Fisher's original formulation of discriminant analysis, which asks for
the linear combination of features that maximizes the ratio of
between-class to within-class variance and makes no Gaussian assumption.
With $K$ classes that problem yields up to $K - 1$ discriminant directions,
printed as `LD1`, `LD2`, ..., and normalized to have unit within-class
variance. With Gaussian classes and a shared covariance matrix, Fisher's
directions and the Bayes'-theorem discriminants of this lecture produce the
same classifier.
:::

### Balance and income

Before `income` enters the model, the first tab below plots income against
balance for every customer, with defaulters and non-defaulters distinguished
by color and shape, and the second shows income within each default class,
one panel per class, on the density scale.

::: {.callout-note collapse="true"}
### Code
```{r default-income-eda}
#| output: false
income_by_student <- Default |>
  group_by(student) |>
  summarize(median_income = median(income), .groups = "drop")
student_income <- function(s) {
  income_by_student$median_income[income_by_student$student == s]
}

# Defaulters are drawn last so they are not hidden under non-defaulters.
scatter_base <- ggplot(arrange(Default, default),
                       aes(x = balance, y = income)) +
  geom_point(aes(color = default, shape = default), alpha = 0.35,
             size = 0.8) +
  scale_color_manual(values = class_colors) +
  scale_shape_manual(values = c(16, 17)) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(labels = dollar_labels) +
  scale_y_continuous(labels = dollar_labels) +
  labs(x = "Credit card balance ($)", y = "Income ($)", color = "Default",
       shape = "Default")

eda_scatter <- scatter_base +
  labs(title = "ISLR2::Default, income against balance")

income_hist_plot <- ggplot(Default, aes(x = income)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 2500,
                 boundary = 0, fill = "grey80", color = "grey55") +
  facet_wrap(~ default, ncol = 1, labeller = label_both) +
  scale_x_continuous(labels = dollar_labels) +
  labs(x = "Income ($)", y = "Density",
       title = "ISLR2::Default, income within each default class")
```
:::

::: {.panel-tabset}
#### Scatterplot
```{r default-scatter-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
eda_scatter
```

#### Income histograms
```{r default-income-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
income_hist_plot
```
:::

In the scatterplot, defaulters are concentrated at high balances across
nearly the whole range of income, and at any given balance the two classes
overlap in income. Within each class balance and income are weakly negatively
correlated: $r = `r sprintf("%.3f", summary_of("No", "cor_balance_income"))`$
among non-defaulters and
$r = `r sprintf("%.3f", summary_of("Yes", "cor_balance_income"))`$ among
defaulters.

The two classes' income histograms look alike, and each is bimodal. The
lower mode is mostly students: in these data students' median income is
\$`r dollars(student_income("Yes"))`, against
\$`r dollars(student_income("No"))` for non-students. No Gaussian density has
two modes, so income strains LDA's assumptions just as balance did.

Adding `income` to the formula gives the $p = 2$ model:

```{r additive-lda}
additive_lda <- MASS::lda(default ~ balance + income, data = Default)
additive_lda
```

::: {.callout-note collapse="true"}
### Code
```{r additive-lda-extract}
#| output: false
X2 <- as.matrix(Default[c("balance", "income")])
hand_2 <- lda_by_hand(X2, Default$default)

scaling_ratio <- additive_lda$scaling[, 1] / hand_2$beta
cor_within <- hand_2$Sigma[1, 2] / sqrt(hand_2$Sigma[1, 1] * hand_2$Sigma[2, 2])
# Income difference between the classes that their balance difference
# predicts through the pooled within-class regression of income on balance.
mean_diff <- hand_2$mu["Yes", ] - hand_2$mu["No", ]
income_diff_predicted <- hand_2$Sigma[1, 2] / hand_2$Sigma[1, 1] * mean_diff[1]

additive_glm <- glm(default ~ balance + income, data = Default,
                    family = "binomial")
b_glm_2 <- coef(additive_glm)
z_income_glm <- summary(additive_glm)$coefficients["income", "z value"]

post_lda_2 <- predict(additive_lda)$posterior[, "Yes"]
post_glm_2 <- fitted(additive_glm)
max_diff_2 <- max(abs(post_lda_2 -
                      plogis(hand_2$beta0 + drop(X2 %*% hand_2$beta))))

# Income coefficients are reported per $1,000 of income.
coef_table_2 <- tibble(
  term = c("Intercept $\\hat\\beta_0$", "`balance` $\\hat\\beta_1$",
           "`income` $\\hat\\beta_2$, per \\$1,000"),
  lda = c(hand_2$beta0, hand_2$beta[1], 1000 * hand_2$beta[2]),
  glm = c(b_glm_2[1], b_glm_2[2], 1000 * b_glm_2[3])
) |>
  mutate(across(c(lda, glm), \(v) sprintf("%.6f", v))) |>
  knitr::kable(col.names = c("Coefficient", "LDA", "Logistic regression"),
               align = "lcc")

# Decision boundary in the (balance, income) plane:
# beta0 + beta1 * balance + beta2 * income = 0.
income_levels <- quantile(Default$income, c(0.25, 0.5, 0.75))
income_level_names <- c("lower quartile", "median", "upper quartile")
threshold_at <- function(b0, b1, b2, inc) -(b0 + b2 * inc) / b1
x_star_lda_2 <- threshold_at(hand_2$beta0, hand_2$beta[1], hand_2$beta[2],
                             income_levels)
x_star_glm_2 <- threshold_at(b_glm_2[1], b_glm_2[2], b_glm_2[3],
                             income_levels)
# LDA's balance threshold across the whole observed range of income, and the
# change in LDA's log posterior odds per standard deviation of each feature.
x_star_lda_range <- threshold_at(hand_2$beta0, hand_2$beta[1],
                                 hand_2$beta[2], range(Default$income))
per_sd_lda <- hand_2$beta * c(sd(Default$balance), sd(Default$income))

boundary_lines <- tibble(
  model = c("LDA", "Logistic regression"),
  intercept = c(-hand_2$beta0 / hand_2$beta[2], -b_glm_2[1] / b_glm_2[3]),
  slope = c(-hand_2$beta[1] / hand_2$beta[2], -b_glm_2[2] / b_glm_2[3])
)

# Boundary lines take the section's model colors as fixed colors, because
# the color scale already encodes default; linetype carries the legend.
boundary_layers <- function(lines) {
  list(
    geom_abline(data = filter(lines, model == "LDA"),
                aes(intercept = intercept, slope = slope, linetype = model),
                color = model_colors[["LDA"]], linewidth = 1),
    geom_abline(data = filter(lines, model == "Logistic regression"),
                aes(intercept = intercept, slope = slope, linetype = model),
                color = model_colors[["Logistic regression"]], linewidth = 1),
    scale_linetype_manual(values = c("solid", "dashed")),
    guides(linetype = guide_legend(
      override.aes = list(color = unname(model_colors))))
  )
}

boundary_plot <- scatter_base +
  boundary_layers(boundary_lines) +
  coord_cartesian(xlim = range(Default$balance),
                  ylim = range(Default$income)) +
  labs(linetype = "Boundary",
       title = "Decision boundaries in the (balance, income) plane")

# Posterior probability against balance at three fixed incomes.
income_facet <- function(v) {
  factor(v, levels = unname(income_levels),
         labels = sprintf("Income $%s (%s)", dollars(income_levels),
                          income_level_names))
}
posterior_grid_2 <- expand_grid(balance = seq(0, 2800, length.out = 400),
                                income = unname(income_levels))
posterior_curves_2 <- posterior_grid_2 |>
  mutate(LDA = predict(additive_lda, newdata = posterior_grid_2)$posterior[, "Yes"],
         `Logistic regression` = predict(additive_glm,
                                         newdata = posterior_grid_2,
                                         type = "response"),
         income_label = income_facet(income)) |>
  pivot_longer(c(LDA, `Logistic regression`), names_to = "model",
               values_to = "p_hat")

posterior_plot_2 <- ggplot(posterior_curves_2, aes(x = balance, y = p_hat)) +
  geom_hline(yintercept = 0.5, color = "grey60") +
  geom_line(aes(color = model, linetype = model), linewidth = 0.9) +
  facet_wrap(~ income_label, nrow = 1) +
  scale_color_manual(values = model_colors) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  scale_x_continuous(labels = dollar_labels) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Model", linetype = "Model",
       title = "Posterior probability of default at three incomes")
```
:::

Substituting the estimates into $\beta_0$ and $\beta$ gives LDA's
coefficients, shown beside those of the logistic regression with the same two
features:

```{r coef-table-2}
#| echo: false
coef_table_2
```

The first tab below draws each model's decision boundary over the
scatterplot of income against balance, LDA's as a solid green line and
logistic regression's as a dashed pink one. The second plots each model's
posterior probability of default against balance at the lower quartile,
median, and upper quartile of income, with a grey line at $0.5$.

::: {.panel-tabset}
#### Decision boundary
```{r additive-boundary-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
boundary_plot
```

#### Posterior
```{r additive-posterior-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 3.4
posterior_plot_2
```
:::

The prior probabilities are unchanged, since they do not depend on the
features. The group means now form a vector for each class: the `balance`
column repeats the one-feature means, and the `income` column holds the class
mean incomes of the class summary table. The coefficients of the linear
discriminant are proportional to
$\hat\beta = \hat\Sigma^{-1}\left(\hat\mu_{\text{Yes}} - \hat\mu_{\text{No}}\right)$: dividing
each printed coefficient by the corresponding entry of $\hat\beta$ computed
from the formulas gives `r sprintf("%.4f", scaling_ratio[1])` for both.

In the coefficient table, the intercept
$\hat\beta_0 = `r sprintf("%.3f", hand_2$beta0)`$ is the log posterior odds of
default for a customer with no balance and no income, an extrapolation, since
the lowest recorded income is \$`r dollars(min(Default$income))`. The
`balance` slope plays the same role as before. The `income` slope,
$`r sprintf("%.4f", 1000 * hand_2$beta[2])`$ per \$1,000, is positive: at a
fixed balance, LDA assigns a slightly higher posterior probability of default
to a customer with a higher income, although defaulters' mean income is the
lower of the two. Multiplying by $\hat\Sigma^{-1}$ adjusts each feature for
the other. With a pooled within-class correlation of
$`r sprintf("%.3f", cor_within)`$, defaulters' higher balances predict a mean
income \$`r dollars(-income_diff_predicted)` below non-defaulters', but it is
only \$`r dollars(-mean_diff[2])` below; the positive slope reflects that
excess. Logistic regression's `income` coefficient,
$`r sprintf("%.4f", 1000 * b_glm_2[3])`$ per \$1,000, is also positive, with
$z = `r sprintf("%.2f", z_income_glm)`$; it too compares customers with the
same balance. Applying the logistic
function to $\hat\beta_0 + x^\top\hat\beta$ again reproduces `predict()`'s
posterior probabilities, to within $`r sci(max_diff_2, 1)`$.

The boundary is a straight line in the $(\texttt{balance}, \texttt{income})$
plane, and a nearly vertical one: across the whole observed range of income,
LDA's balance threshold shifts by only
\$`r dollars(abs(diff(x_star_lda_range)))`, against a range of balance of
\$`r dollars(diff(range(Default$balance)))`. Per standard deviation, balance
moves LDA's log posterior odds by `r sprintf("%.2f", per_sd_lda[1])` and
income by only `r sprintf("%.2f", per_sd_lda[2])`. Read as a balance
threshold, LDA's boundary sits at
\$`r dollars(x_star_lda_2[1])` for a customer at the lower quartile of
income and \$`r dollars(x_star_lda_2[3])` at the upper quartile; logistic
regression's sits at \$`r dollars(x_star_glm_2[1])` and
\$`r dollars(x_star_glm_2[3])`. In both models the threshold falls as income
rises, and logistic regression's lies to the left of LDA's. The posterior
tab shows the same comparison on the probability scale: at each income the
LDA curve is flatter than the logistic regression curve and crosses $0.5$
further right.

Whether adding `income` improves the classifications is the question of the
next three subsections.

### Confusion matrices

Classifying each customer as a defaulter when $\hat p(x) > 0.5$, the LDA
classifier with `balance` and `income` produces this confusion matrix:

::: {.callout-note collapse="true"}
### Code
```{r confusion-lda}
#| output: false
confusion_counts <- function(p_hat, t) {
  truth <- Default$default == "Yes"
  pred  <- p_hat > t
  c(TP = sum(truth & pred), FN = sum(truth & !pred),
    FP = sum(!truth & pred), TN = sum(!truth & !pred))
}

cm_lda_05 <- confusion_counts(post_lda_2, 0.5)
# A lower threshold, compared with 0.5 in the next subsection.
t_low <- 0.2

confusion_lda_table <- tibble(
  Truth = c("Yes", "No"),
  `Predicted Yes` = c(cm_lda_05["TP"], cm_lda_05["FP"]),
  `Predicted No`  = c(cm_lda_05["FN"], cm_lda_05["TN"])
) |>
  mutate(Total = `Predicted Yes` + `Predicted No`) |>
  mutate(across(-Truth, \(v) format(v, big.mark = ","))) |>
  knitr::kable(align = "lccc")

metrics <- function(cm) {
  c(error = unname((cm["FP"] + cm["FN"]) / sum(cm)),
    sensitivity = unname(cm["TP"] / (cm["TP"] + cm["FN"])),
    specificity = unname(cm["TN"] / (cm["TN"] + cm["FP"])))
}
m_lda_05 <- metrics(cm_lda_05)
```
:::

```{r confusion-lda-table}
#| echo: false
confusion_lda_table
```

The classifier misclassifies
`r cm_lda_05["FP"] + cm_lda_05["FN"]` of
`r format(sum(cm_lda_05), big.mark = ",")` customers, an error rate of
`r sprintf("%.2f", 100 * m_lda_05["error"])`%, against the
`r sprintf("%.2f", 100 * hand_1$pi[2])`% of the classifier that predicts `No`
for everyone. The rows tell a different story. Of the
`r cm_lda_05["TP"] + cm_lda_05["FN"]` customers who defaulted, it flags only
`r cm_lda_05["TP"]`, a sensitivity of
`r sprintf("%.1f", 100 * m_lda_05["sensitivity"])`%. Of the
`r format(cm_lda_05["TN"] + cm_lda_05["FP"], big.mark = ",")` who did not, it
correctly leaves alone `r format(cm_lda_05["TN"], big.mark = ",")`, a
specificity of `r sprintf("%.1f", 100 * m_lda_05["specificity"])`%. The error
rate is small because the
`r format(cm_lda_05["TN"] + cm_lda_05["FP"], big.mark = ",")` non-defaulters,
nearly all classified correctly, dominate it, even though
`r cm_lda_05["FN"]` of the `r cm_lda_05["FP"] + cm_lda_05["FN"]` errors are
missed defaulters.

These are training error rates: the customers being classified are the ones
the model was fit to, so they are optimistic estimates of performance on new
customers, as the classification lecture discussed. Estimating test error from
the training data alone is the subject of the resampling unit.

### Threshold choice

A credit card company that wants to catch more of its eventual defaulters can
lower the threshold. The figure below tracks four rates of the LDA classifier
with `balance` and `income` as $t$ moves from $0$ to $0.5$ — the overall
error rate, sensitivity (the true positive rate), specificity (the true
negative rate), and the false positive rate ($1 - \text{specificity}$) — with
dotted lines at $t = `r t_low`$ and $t = 0.5$. The checkboxes choose which
rates are drawn; the false positive rate starts unchecked because it mirrors
specificity.

::: {.callout-note collapse="true"}
### Code
```{r threshold-sweep}
#| output: false
threshold_rates <- tibble(t = seq(0, 0.5, by = 0.005)) |>
  mutate(m = map(t, \(tt) metrics(confusion_counts(post_lda_2, tt))),
         `Overall error rate` = map_dbl(m, "error"),
         `Sensitivity (true positive rate)` = map_dbl(m, "sensitivity"),
         `Specificity (true negative rate)` = map_dbl(m, "specificity"),
         `False positive rate (1 - specificity)` =
           map_dbl(m, \(v) 1 - v[["specificity"]])) |>
  select(-m) |>
  pivot_longer(-t, names_to = "rate", values_to = "value")

# Training error rate at every distinct posterior probability used as t;
# the error rate only changes when t passes an observation's score.
min_error_threshold <- function(p_hat) {
  cand <- sort(unique(p_hat))
  err  <- map_dbl(cand, \(tt) metrics(confusion_counts(p_hat, tt))[["error"]])
  c(t = cand[which.min(err)], error = min(err))
}
min_err_lda <- min_error_threshold(post_lda_2)
min_err_glm <- min_error_threshold(post_glm_2)

cm_lda1_05  <- confusion_counts(post_1, 0.5)
cm_lda1_low <- confusion_counts(post_1, t_low)
cm_lda_low  <- confusion_counts(post_lda_2, t_low)
cm_glm_05   <- confusion_counts(post_glm_2, 0.5)
cm_glm_low  <- confusion_counts(post_glm_2, t_low)
m_lda1_05   <- metrics(cm_lda1_05)
m_lda1_low  <- metrics(cm_lda1_low)
m_lda_low   <- metrics(cm_lda_low)
m_glm_05    <- metrics(cm_glm_05)
m_glm_low   <- metrics(cm_glm_low)

compare_table <- tibble(
  Model = rep(c("LDA, balance", "LDA, balance + income",
                "Logistic regression, balance + income"), each = 2),
  t = rep(c(0.5, t_low), 3),
  cm = list(cm_lda1_05, cm_lda1_low, cm_lda_05, cm_lda_low,
            cm_glm_05, cm_glm_low)
) |>
  mutate(TP = map_dbl(cm, "TP"), FN = map_dbl(cm, "FN"),
         FP = map_dbl(cm, "FP"), TN = map_dbl(cm, "TN"),
         m = map(cm, metrics),
         error = map_dbl(m, "error"),
         sensitivity = map_dbl(m, "sensitivity"),
         specificity = map_dbl(m, "specificity")) |>
  select(-cm, -m) |>
  mutate(across(c(TP, FN, FP, TN), \(v) format(v, big.mark = ",")),
         across(c(error, sensitivity, specificity),
                \(v) sprintf("%.4f", v))) |>
  knitr::kable(col.names = c("Model", "Threshold", "TP", "FN", "FP", "TN",
                             "Error rate", "Sensitivity", "Specificity"),
               align = "lcccccccc")
```
:::

```{r threshold-data}
#| echo: false
ojs_define(threshold_rates = threshold_rates,
           threshold_marks = c(t_low, 0.5))
```

```{ojs}
//| label: threshold-checkboxes
//| echo: false
rateNames = ["Overall error rate", "Sensitivity (true positive rate)",
             "Specificity (true negative rate)",
             "False positive rate (1 - specificity)"]
rateStyles = ({
  "Overall error rate": {stroke: "black", dash: "none"},
  "Sensitivity (true positive rate)": {stroke: "#D55E00", dash: "8,4"},
  "Specificity (true negative rate)": {stroke: "#0072B2", dash: "2,4"},
  "False positive rate (1 - specificity)": {stroke: "#009E73", dash: "12,4,2,4"}
})
viewof shownRates = Inputs.checkbox(rateNames, {
  value: rateNames.slice(0, 3),
  format: name => html`<span style="display: inline-flex; align-items: center; gap: 0.4em">
    <svg width="36" height="10"><line x1="2" y1="5" x2="34" y2="5"
      stroke="${rateStyles[name].stroke}" stroke-width="2.5"
      stroke-dasharray="${rateStyles[name].dash}" stroke-linecap="round"/></svg>${name}</span>`
})
```

```{ojs}
//| label: threshold-plot
//| echo: false
{
  const rates = transpose(threshold_rates);
  return Plot.plot({
    width: 640, height: 340,
    marginTop: 24,
    x: {domain: [0, 0.5], label: "Threshold t"},
    y: {domain: [0, 1], label: "Rate", grid: true},
    marks: [
      Plot.ruleX(threshold_marks, {stroke: "gray", strokeDasharray: "2,3"}),
      Plot.text(threshold_marks, {x: d => d, y: 1, text: d => `t = ${d}`,
                                  dy: -12, fill: "gray"}),
      ...rateNames.filter(r => shownRates.includes(r)).map(r =>
        Plot.line(rates.filter(d => d.rate === r),
                  {x: "t", y: "value", stroke: rateStyles[r].stroke,
                   strokeDasharray: rateStyles[r].dash, strokeWidth: 2.5,
                   strokeLinecap: "round"}))
    ]
  });
}
```

The confusion-matrix counts and rates at $t = 0.5$ and $t = `r t_low`$ for LDA with
`balance` alone, LDA with `balance` and `income`, and the logistic regression
with `balance` and `income`:

```{r compare-table}
#| echo: false
compare_table
```

As $t$ falls from $0.5$, sensitivity rises steadily while specificity stays
near $1$, and the false positive rate near $0$, until $t$ is small. The
overall error rate, dominated by the
`r format(cm_lda_05["TN"] + cm_lda_05["FP"], big.mark = ",")`
non-defaulters, follows the false positive rate: nearly flat over most of the
range, then climbing sharply as $t$ approaches $0$ and almost every customer is
flagged.

The overall error rate need not be smallest at $t = 0.5$. LDA's training
error rate is lowest near $t = `r sprintf("%.3f", min_err_lda[["t"]])`$, at
`r sprintf("%.2f", 100 * min_err_lda[["error"]])`%, against
`r sprintf("%.2f", 100 * m_lda_05["error"])`% at $t = 0.5$. The logistic
regression's is lowest near
$t = `r sprintf("%.3f", min_err_glm[["t"]])`$, at
`r sprintf("%.2f", 100 * min_err_glm[["error"]])`%, against
`r sprintf("%.2f", 100 * m_glm_05["error"])`% at $t = 0.5$.

In the table, lowering the threshold to $t = `r t_low`$ raises the
two-feature LDA classifier's sensitivity from
`r sprintf("%.1f", 100 * m_lda_05["sensitivity"])`% to
`r sprintf("%.1f", 100 * m_lda_low["sensitivity"])`%, at the cost of
specificity falling from `r sprintf("%.1f", 100 * m_lda_05["specificity"])`% to
`r sprintf("%.1f", 100 * m_lda_low["specificity"])`% and the overall error rate
rising from `r sprintf("%.2f", 100 * m_lda_05["error"])`% to
`r sprintf("%.2f", 100 * m_lda_low["error"])`%. Adding `income` changes LDA
very little: with `balance` alone the error rate is
`r sprintf("%.2f", 100 * m_lda1_05["error"])`% and the sensitivity
`r sprintf("%.1f", 100 * m_lda1_05["sensitivity"])`% at $t = 0.5$, and
`r sprintf("%.2f", 100 * m_lda1_low["error"])`% and
`r sprintf("%.1f", 100 * m_lda1_low["sensitivity"])`% at $t = `r t_low`$. At
both thresholds logistic regression flags more customers as defaulters than
the two-feature LDA classifier does —
`r cm_glm_05["TP"] + cm_glm_05["FP"]` against
`r cm_lda_05["TP"] + cm_lda_05["FP"]` at $t = 0.5$ — and has the higher
sensitivity and the lower specificity of the two.

### ROC curves and AUC

A comparison at one or two thresholds depends on which thresholds were
chosen. The ROC curves below compare LDA and logistic regression with
`balance` and `income` at every threshold, with their points at $t = 0.5$ and
$t = `r t_low`$ from the previous table marked and a dashed diagonal for
reference.

::: {.callout-note collapse="true"}
### Code
```{r roc-default}
#| output: false
is_yes <- Default$default == "Yes"
scores <- list(
  "LDA, balance" = post_1,
  "Logistic regression, balance" = fitted(balance_glm),
  "LDA, balance + income" = post_lda_2,
  "Logistic regression, balance + income" = post_glm_2
)

auc_values <- tibble(model = names(scores)) |>
  mutate(trapezoid = map_dbl(scores, \(s) auc_trapezoid(roc_curve(s, is_yes))),
         pairs = map_dbl(scores, \(s) auc_pairs(s, is_yes)))

auc_of <- function(m) auc_values$trapezoid[auc_values$model == m]

auc_table <- auc_values |>
  mutate(across(c(trapezoid, pairs), \(v) sprintf("%.5f", v))) |>
  knitr::kable(col.names = c("Model", "AUC (trapezoid rule)",
                             "AUC (pairwise comparison)"),
               align = "lcc")

roc_2 <- bind_rows(
  roc_curve(post_lda_2, is_yes) |> mutate(model = "LDA"),
  roc_curve(post_glm_2, is_yes) |> mutate(model = "Logistic regression")
)

threshold_labels <- c("t = 0.5", paste("t =", t_low))
roc_points <- tibble(
  model = rep(c("LDA", "Logistic regression"), each = 2),
  t = factor(rep(threshold_labels, 2), levels = threshold_labels),
  m = list(m_lda_05, m_lda_low, m_glm_05, m_glm_low)
) |>
  mutate(fpr = 1 - map_dbl(m, "specificity"),
         tpr = map_dbl(m, "sensitivity"))

roc_plot <- ggplot(roc_2, aes(x = fpr, y = tpr)) +
  geom_abline(intercept = 0, slope = 1, color = "grey60",
              linetype = "dashed") +
  geom_line(aes(color = model, linetype = model), linewidth = 0.9) +
  geom_point(data = roc_points, aes(shape = model, color = model, fill = t),
             size = 3, stroke = 1.2) +
  scale_color_manual(values = model_colors) +
  # Shape separates the models, filled versus open separates the thresholds.
  scale_shape_manual(values = c(24, 22)) +
  scale_fill_manual(values = c("black", "white")) +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +
  coord_fixed() +
  labs(x = "False positive rate (1 - specificity)",
       y = "True positive rate (sensitivity)",
       color = "Model", linetype = "Model", shape = "Model",
       fill = "Threshold", title = "ROC curves, balance + income")

# Balance at which the two-feature models' posterior curves cross, at the
# lower quartile, median, and upper quartile of income, and the posterior
# probability there.
posterior_both <- function(b, inc) {
  nd <- tibble(balance = b, income = inc)
  c(lda = predict(additive_lda, newdata = nd)$posterior[, "Yes"],
    glm = unname(predict(additive_glm, newdata = nd, type = "response")))
}
crossover <- tibble(income = unname(income_levels)) |>
  mutate(balance = map_dbl(income, \(i) {
           uniroot(\(b) diff(posterior_both(b, i)), c(800, 2500))$root
         }),
         p = map2_dbl(balance, income, \(b, i) posterior_both(b, i)[["lda"]]))

# A low threshold, and checks of which model's flagged set contains the
# other's: at t_below every customer logistic regression flags is flagged by
# LDA; at t_low and 0.5 every customer LDA flags is flagged by logistic
# regression.
t_below <- 0.1
flags_within <- function(p_inner, p_outer, t) !any(p_inner > t & p_outer <= t)
stopifnot(flags_within(post_glm_2, post_lda_2, t_below),
          flags_within(post_lda_2, post_glm_2, t_low),
          flags_within(post_lda_2, post_glm_2, 0.5))
n_flag_lda_below <- sum(post_lda_2 > t_below)
n_flag_glm_below <- sum(post_glm_2 > t_below)
```
:::

```{r roc-figure}
#| echo: false
#| fig-width: 6.5
#| fig-height: 5.5
roc_plot
```

The AUC of each classifier, for both the `balance` and the
`balance + income` models, computed both as the area under the empirical ROC
curve and as the proportion of (defaulter, non-defaulter) pairs ordered
correctly:

```{r auc-table}
#| echo: false
auc_table
```

The two ROC curves lie almost on top of one another and rise steeply near the
origin, well above the diagonal. Lowering $t$ from $0.5$ to `r t_low` moves
each model's marked point up and slightly right along its curve. At each of
these two thresholds, logistic regression's point sits further along its
nearly shared curve than LDA's.

At the lower quartile, median, and upper quartile of income, the two models'
posterior curves cross at balances of
\$`r dollars(crossover$balance[1])`, \$`r dollars(crossover$balance[2])`,
and \$`r dollars(crossover$balance[3])`, where both give posterior
probabilities of `r sprintf("%.3f", crossover$p[1])`,
`r sprintf("%.3f", crossover$p[2])`, and `r sprintf("%.3f", crossover$p[3])`.
Above those balances LDA's flatter curve lies below logistic regression's, so
a threshold above the crossing probabilities is reached by LDA only at a
higher balance. At both marked thresholds, every customer LDA flags is also
flagged by logistic regression. At a low threshold the order reverses: at
$t = `r t_below`$, LDA flags `r format(n_flag_lda_below, big.mark = ",")`
customers and logistic regression
`r format(n_flag_glm_below, big.mark = ",")`, and every customer logistic
regression flags is also flagged by LDA.

The two AUC computations agree in every row of the table. With `balance`
alone, LDA and logistic regression have identical AUCs of
`r sprintf("%.5f", auc_of("LDA, balance"))`: both posterior probabilities are
increasing functions of balance, so they order the customers identically and
trace the same ROC curve. With `balance` and `income`, the two classifiers'
posterior probabilities weight income differently relative to balance, so
their orderings differ slightly. Their AUCs,
`r sprintf("%.5f", auc_of("LDA, balance + income"))` and
`r sprintf("%.5f", auc_of("Logistic regression, balance + income"))`, differ
by $`r sci(auc_of("LDA, balance + income") - auc_of("Logistic regression, balance + income"), 2)`$,
and adding `income` raises LDA's AUC by only
`r sprintf("%.4f", auc_of("LDA, balance + income") - auc_of("LDA, balance"))`.
An AUC of `r sprintf("%.3f", auc_of("LDA, balance + income"))` means that for
a randomly chosen defaulter and non-defaulter, the LDA classifier gives the
defaulter the higher posterior probability about
`r round(100 * auc_of("LDA, balance + income"))`% of the time.

### Student strata

Student status has not entered these models. The table below summarizes the
two student strata.

::: {.callout-note collapse="true"}
### Code
```{r strata-summary}
#| output: false
strata <- c(No = "No", Yes = "Yes")
in_stratum <- map(strata, \(s) filter(Default, student == s))
strata_summary <- Default |>
  group_by(student) |>
  summarize(n = n(), defaulters = sum(default == "Yes"),
            rate = mean(default == "Yes"), median_income = median(income),
            min_income = min(income), max_income = max(income),
            .groups = "drop")
strata_summary_of <- function(s, col) {
  strata_summary[[col]][strata_summary$student == s]
}
strata_summary_table <- strata_summary |>
  transmute(student, n = format(n, big.mark = ","), defaulters,
            rate = sprintf("%.2f%%", 100 * rate),
            median_income = paste0("\\$", dollars(median_income))) |>
  knitr::kable(col.names = c("Student", "Customers", "Defaulters",
                             "Default rate", "Median income"),
               align = "ccccc")
```
:::

```{r strata-summary-table}
#| echo: false
strata_summary_table
```

Students default more often than non-students,
`r sprintf("%.2f", 100 * strata_summary_of("Yes", "rate"))`% against
`r sprintf("%.2f", 100 * strata_summary_of("No", "rate"))`%, and, as noted
above, their median income is far lower.

Below, LDA with `balance` and `income` is fit separately to the non-students
and the students, so its priors, class means, and pooled covariance can all
differ by stratum. Beside it is the logistic regression in which the
intercept and both slopes interact with student status. Each customer is
scored by the model for its own stratum. Because each stratum's priors are
its own default rates, each posterior estimates the probability of default
given balance, income, and student status, so posteriors from the two strata
are on one scale and can be compared with each other.

::: {.callout-note collapse="true"}
### Code
```{r strata}
#| output: false
strata_lda <- map(in_stratum, \(d) MASS::lda(default ~ balance + income, data = d))
strata_glm <- map(in_stratum, \(d) glm(default ~ balance + income, data = d,
                                       family = "binomial"))
interacted_glm <- glm(default ~ (balance + income) * student, data = Default,
                      family = "binomial")

# Score every customer with its own stratum's model.
by_stratum <- function(score) {
  p <- numeric(nrow(Default))
  for (s in strata) p[Default$student == s] <- score(s)
  p
}
post_lda_strat <- by_stratum(\(s) {
  predict(strata_lda[[s]], newdata = in_stratum[[s]])$posterior[, "Yes"]
})
post_glm_strat <- by_stratum(\(s) fitted(strata_glm[[s]]))
max_diff_int <- max(abs(fitted(interacted_glm) - post_glm_strat))

strata_metrics <- imap(list(
  "LDA, single fit" = post_lda_2, "LDA, by student stratum" = post_lda_strat,
  "Logistic regression, additive" = post_glm_2,
  "Logistic regression, interacted with student" = fitted(interacted_glm)
), \(p, nm) {
  m <- metrics(confusion_counts(p, 0.5))
  tibble(model = nm, error = m[["error"]],
         auc = auc_trapezoid(roc_curve(p, is_yes)))
}) |>
  list_rbind()
strata_of <- function(m, col) strata_metrics[[col]][strata_metrics$model == m]
strata_table <- strata_metrics |>
  mutate(error = sprintf("%.4f", error), auc = sprintf("%.5f", auc)) |>
  knitr::kable(col.names = c("Model", "Error rate at $t = 0.5$", "AUC"),
               align = "lcc")

# Each stratum's boundary from each model, with its balance threshold at the
# stratum's median income.
strata_hand <- map(in_stratum, \(d) lda_by_hand(d[c("balance", "income")],
                                                 d$default))
strata_lines <- imap(strata, \(s, nm) {
  h <- strata_hand[[s]]
  b <- coef(strata_glm[[s]])
  tibble(student = s, model = c("LDA", "Logistic regression"),
         b0 = c(h$beta0, b[[1]]), b1 = c(h$beta[[1]], b[[2]]),
         b2 = c(h$beta[[2]], b[[3]]))
}) |>
  list_rbind() |>
  left_join(strata_summary |>
              transmute(student = as.character(student), median_income),
            by = "student") |>
  mutate(intercept = -b0 / b2, slope = -b1 / b2,
         threshold = threshold_at(b0, b1, b2, median_income))
strata_value <- function(s, col, m = "LDA") {
  strata_lines[[col]][strata_lines$student == s & strata_lines$model == m]
}
# The LDA and logistic boundaries of a stratum cross where their balance
# thresholds are equal; check that each crossing lies outside the incomes the
# stratum has, so logistic regression's boundary is left of LDA's throughout.
boundary_crossing <- function(s) {
  l <- filter(strata_lines, student == s)
  (l$b0[2] / l$b1[2] - l$b0[1] / l$b1[1]) /
    (l$b2[1] / l$b1[1] - l$b2[2] / l$b1[2])
}
for (s in strata) {
  cr <- boundary_crossing(s)
  inc <- in_stratum[[s]]$income
  stopifnot(cr < min(inc) || cr > max(inc),
            strata_value(s, "threshold", "Logistic regression") <
              strata_value(s, "threshold"))
}

# How much the students' larger prior alone lowers their LDA threshold:
# the change in log prior odds, converted to balance by the balance slope.
log_prior_odds <- function(h) log(h$pi[2] / h$pi[1])
prior_shift <- (log_prior_odds(strata_hand$Yes) - log_prior_odds(strata_hand$No)) /
  strata_hand$Yes$beta[[1]]
# Class mean balances by stratum, and the midpoint of each stratum's pair.
class_mean <- function(s, k) strata_hand[[s]]$mu[k, "balance"]
midpoint_shift <- mean(strata_hand$Yes$mu[, "balance"]) -
  mean(strata_hand$No$mu[, "balance"])
# Students' minus non-students' LDA threshold over the incomes both strata
# have; the thresholds are linear in income, so the ends bound the gap.
overlap <- c(max(strata_summary$min_income), min(strata_summary$max_income))
threshold_gap <- threshold_at(strata_hand$Yes$beta0, strata_hand$Yes$beta[[1]],
                              strata_hand$Yes$beta[[2]], overlap) -
  threshold_at(strata_hand$No$beta0, strata_hand$No$beta[[1]],
               strata_hand$No$beta[[2]], overlap)
stopifnot(all(threshold_gap > 0))

# LDA estimates K - 1 priors, K p means, and p (p + 1) / 2 covariances.
K <- nlevels(Default$default)
p_features <- length(c("balance", "income"))
n_par_lda <- (K - 1) + K * p_features + p_features * (p_features + 1) / 2

strata_plot <- scatter_base +
  boundary_layers(strata_lines) +
  # Each panel's income axis covers its own stratum's incomes.
  facet_wrap(~ student, labeller = label_both, scales = "free_y") +
  labs(linetype = "Boundary",
       title = "Boundaries fit within each student stratum")
```
:::

The table gives the training error rate at $t = 0.5$ and the AUC of these two
models and of the single LDA and additive logistic regression of the previous
subsections, all computed on the training customers. The figure repeats the
scatterplot of income against balance for each stratum, over that stratum's
own range of income, with each stratum's LDA boundary (solid green) and
interacted logistic regression boundary (dashed pink).

```{r strata-table}
#| echo: false
strata_table
```

```{r strata-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 4
strata_plot
```

Fitting the logistic regression with every coefficient interacted with
student status is the same as fitting separate logistic regressions to the
two strata: the two give the same fitted probabilities, to within
$`r sci(max_diff_int, 1)`$.

The extra flexibility barely changes the classifications. Stratifying LDA
moves its training error rate from
`r sprintf("%.2f", 100 * strata_of("LDA, single fit", "error"))`% to
`r sprintf("%.2f", 100 * strata_of("LDA, by student stratum", "error"))`% and
its AUC from `r sprintf("%.4f", strata_of("LDA, single fit", "auc"))` to
`r sprintf("%.4f", strata_of("LDA, by student stratum", "auc"))`; interacting
the logistic regression with student status moves its error rate from
`r sprintf("%.2f", 100 * strata_of("Logistic regression, additive", "error"))`%
to
`r sprintf("%.2f", 100 * strata_of("Logistic regression, interacted with student", "error"))`%
and its AUC from
`r sprintf("%.4f", strata_of("Logistic regression, additive", "auc"))` to
`r sprintf("%.4f", strata_of("Logistic regression, interacted with student", "auc"))`.
The AUC compares defaulters with non-defaulters across strata as well as
within them, which is meaningful because the two strata's posteriors share
that scale. These are training rates, and the stratified models estimate twice as many
parameters — `r 2 * n_par_lda` against `r n_par_lda` for LDA (one prior,
four means, and three covariance entries per fit), and
`r length(coef(interacted_glm))` against `r length(coef(additive_glm))` for
logistic regression — so their training rates tend to understate their test
error rates by more; whether stratifying helps on new customers is a question for
test data.

At each stratum's median income, LDA's balance threshold is
\$`r dollars(strata_value("No", "threshold"))` for non-students and
\$`r dollars(strata_value("Yes", "threshold"))` for students, and logistic
regression's is
\$`r dollars(strata_value("No", "threshold", "Logistic regression"))` and
\$`r dollars(strata_value("Yes", "threshold", "Logistic regression"))`. Over
the incomes each stratum actually has, logistic regression's boundary lies to
the left of LDA's.

The students' larger prior for `Yes` alone would lower their LDA threshold
by \$`r dollars(prior_shift)`. Their threshold is higher nonetheless because
students in both classes carry higher balances: non-defaulting students
average \$`r dollars(class_mean("Yes", "No"))` against
\$`r dollars(class_mean("No", "No"))` for non-students, and defaulting
students \$`r dollars(class_mean("Yes", "Yes"))` against
\$`r dollars(class_mean("No", "Yes"))`, which puts the midpoint of the
students' class means \$`r dollars(midpoint_shift)` higher. This is the
confounding of the
[Multiple Logistic Regression](20-multiple-logistic-regression.qmd) lecture.
The students' threshold stays higher when the strata are compared at the same
income: over the incomes both strata have,
\$`r dollars(overlap[1])` to \$`r dollars(overlap[2])`, it exceeds the
non-students' by between \$`r dollars(min(threshold_gap))` and
\$`r dollars(max(threshold_gap))`.

**Conclusion.** LDA's Gaussian assumptions fit these data imperfectly —
balance has a spike at zero and a smaller spread among defaulters, and income
is bimodal — yet LDA with `balance` and `income` orders the training
customers almost exactly as the logistic regression with the same features
does. Income adds little to classification beyond balance — it raises LDA's
AUC by
`r sprintf("%.4f", auc_of("LDA, balance + income") - auc_of("LDA, balance"))`,
moves its training error rate at $t = 0.5$ from
`r sprintf("%.2f", 100 * m_lda1_05["error"])`% to
`r sprintf("%.2f", 100 * m_lda_05["error"])`%, and at $t = `r t_low`$ raises it
from `r sprintf("%.2f", 100 * m_lda1_low["error"])`% to
`r sprintf("%.2f", 100 * m_lda_low["error"])`% and lowers sensitivity from
`r sprintf("%.1f", 100 * m_lda1_low["sensitivity"])`% to
`r sprintf("%.1f", 100 * m_lda_low["sensitivity"])`% — although it is a real
predictor in the logistic regression. Fitting separate models
for students and non-students changes the training error rates and AUCs only
slightly. At $t = 0.5$ and
$t = `r t_low`$ LDA flags fewer customers than logistic regression, because
its flatter posterior curve lies below logistic regression's over the
balances where those thresholds are crossed. At $t = 0.5$ it catches
`r sprintf("%.1f", 100 * m_lda_05["sensitivity"])`% of the defaulters, and
lowering the threshold to `r t_low` raises that to
`r sprintf("%.1f", 100 * m_lda_low["sensitivity"])`% at a cost of
`r sprintf("%.1f", 100 * (m_lda_05["specificity"] - m_lda_low["specificity"]))`
percentage points of specificity.

## Conclusion

A generative classifier models the distribution of the features within each
class, $f_k(x)$, and the prior probability of each class, $\pi_k$, and
combines them through Bayes' theorem into the posterior probability
$p_k(x) = \pi_k f_k(x) / \sum_l \pi_l f_l(x)$. Assigning each observation to
the class with the largest posterior probability is the Bayes classifier,
which minimizes the expected error rate; a generative classifier approximates
it by estimating $\pi_k$ and $f_k$ from training data.

Linear discriminant analysis assumes Gaussian class-conditional densities with
class-specific means and a covariance matrix shared by every class. Because
the covariance is shared, the quadratic term in $x$ is the same for every
class and drops out, leaving discriminant functions $\delta_k(x)$ that are
linear in $x$ and a decision boundary that is a point, a line, or a
hyperplane. With two classes the log posterior odds are linear in the
features, the same form as logistic regression, but the coefficients come
from closed-form estimates from the joint likelihood of features and
responses rather than from the iteratively maximized conditional likelihood
of logistic regression. The class proportions and class means maximize that
joint likelihood, and the pooled variance or covariance matrix is its
maximizer rescaled to be unbiased. A smaller prior probability moves the
boundary toward that class's mean.

A classifier's error rate hides which mistakes it makes. The confusion matrix
separates them, sensitivity and specificity measure performance within each
true class, and moving the threshold trades one against the other. With $K$
classes the confusion matrix is $K \times K$ and sensitivity and specificity
are computed one class versus the rest. The ROC curve displays the two-class
tradeoff at every threshold and depends only on how a classifier orders the
observations. Its area, the AUC, is the probability that a randomly chosen
positive observation is scored above a randomly chosen negative one, and it
ranks classifiers without fixing a threshold.

The next lecture relaxes LDA's shared covariance assumption. Quadratic
discriminant analysis gives each class its own covariance matrix, so the
quadratic term no longer cancels and the boundary curves, and naive Bayes
replaces the multivariate Gaussian with a product of one-feature densities.
That lecture then compares all five classifiers, adding KNN.
