# Linear Discriminant Analysis

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
set.seed(20261001)

# MASS is called as MASS::lda() rather than attached, because attaching it
# masks dplyr::select().
# Renders a number in LaTeX scientific notation for use inside $...$.
sci <- function(x, digits = 2) {
  e <- floor(log10(abs(x)))
  m <- signif(x / 10^e, digits)
  sprintf("%s \\times 10^{%d}", format(m), e)
}

class_colors <- c("#0072B2", "#D55E00")
model_colors <- c("LDA" = "#009E73", "Logistic regression" = "#CC79A7")
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **derive** the posterior probability $p_k(x) = P(Y = k \mid X = x)$ from the
prior probabilities $\pi_k$ and the class-conditional densities $f_k(x)$ using
Bayes' theorem, and **explain** why classifying to the class with the largest
posterior probability minimizes the expected error rate,
- **derive** the LDA discriminant function $\delta_k(x)$ for one feature under
Gaussian class-conditional densities with a shared variance $\sigma^2$, and
**show** that the resulting decision boundary is linear in $x$ and moves toward
the mean of the class whose prior probability is smaller,
- **write** the discriminant function for $p > 1$ features in matrix form using
class mean vectors $\mu_k$ and a shared covariance matrix $\Sigma$, and
**compute** the estimates $\hat\pi_k$, $\hat\mu_k$, and $\hat\sigma^2$ or
$\hat\Sigma$ from training data,
- **fit** LDA with `MASS::lda()`, **interpret** its prior probabilities, group
means, and posterior probabilities, and **compare** its posterior
probabilities and decision boundary with those of logistic regression,
- **construct** a confusion matrix at a stated threshold, **compute** the error
rate, sensitivity, and specificity from it, and **explain** how sensitivity
and specificity trade off as the threshold moves, and
- **interpret** an ROC curve and its AUC, including AUC as the probability that
a randomly chosen positive observation receives a higher score than a randomly
chosen negative one.
:::

## Generative classifiers

Logistic regression models the conditional probability $P(Y = k \mid X = x)$
directly. Its likelihood is the conditional distribution of the response given
the features, and the distribution of the features themselves never enters it.

A **generative classifier** takes the other route. It models how the features
are distributed *within* each class, and how common each class is, and then
reverses the conditioning with Bayes' theorem to obtain
$P(Y = k \mid X = x)$. Linear discriminant analysis (LDA) is the first generative
classifier in this course.

### Bayes' theorem

Every earlier lecture conditioned on the features and treated them as fixed. A
generative classifier models their distribution, so here $X$ is a random
vector — the features of a randomly chosen observation — and $x$ is a value it
takes. Let the response $Y$ take one of $K$ classes, $k = 1, \ldots, K$, and
define

$$\pi_k = P(Y = k), \qquad f_k(x) = \text{density of } X \text{ given } Y = k,$$

where

- $K$ is the number of classes ($K = 2$ for a binary response),
- $\pi_k$ is the **prior probability** of class $k$, the probability that a
randomly chosen observation belongs to class $k$ before its features are
seen, with $\sum_{k=1}^K \pi_k = 1$, and
- $f_k(x)$ is the **class-conditional density** of $X$ among observations in
class $k$ (a probability mass function if $X$ is discrete).

The classification and logistic regression lectures wrote $C$ for the number
of classes; following ISLR2, this and later lectures write $K$.

Bayes' theorem, with densities in place of the probabilities of the events
$\{X = x\}$, gives

$$P(Y = k \mid X = x) = \frac{P(Y = k)\, f_k(x)}{f(x)},$$

where $f(x)$ is the marginal density of $X$. By the law of total probability,
$f(x) = \sum_{l=1}^K P(Y = l)\, f_l(x)$, so

$$p_k(x) = P(Y = k \mid X = x) = \frac{\pi_k f_k(x)}{\sum_{l=1}^K \pi_l f_l(x)},$$

where

- $p_k(x)$ is the **posterior probability** of class $k$ for an observation
with features $x$, and
- $l$ indexes the classes in the sum, kept distinct from the $k$ being
evaluated.

The denominator is the same for every $k$. The posterior is therefore
proportional to $\pi_k f_k(x)$: a class is probable at $x$ when it is common
($\pi_k$ large) and when $x$ is a typical feature value for it ($f_k(x)$
large).

### Bayes classifier

Suppose an observation with features $x$ is assigned to class $\hat y(x)$. The
probability that the assignment is wrong is

$$P\left(Y \ne \hat y(x) \mid X = x\right) = 1 - p_{\hat y(x)}(x).$$

This is smallest when $\hat y(x)$ is the class with the largest posterior
probability. Choosing that class at every $x$ minimizes the conditional error
probability at every $x$, and so it also minimizes the overall error rate
$E\left[1 - p_{\hat y(X)}(X)\right]$, which averages those conditional error
probabilities over the distribution of $X$.

The rule "assign each observation to $\arg\max_k p_k(x)$" is the **Bayes
classifier**, and its error rate, the **Bayes error rate**, is the lowest
achievable by any classifier. With $K = 2$ classes it reduces to assigning
class $2$ whenever $p_2(x) > 0.5$.

The Bayes classifier needs the true $\pi_k$ and $f_k$, which are never known.
A generative classifier estimates both from training data and substitutes the
estimates into Bayes' theorem. Estimating $\pi_k$ takes nothing more than the
proportion of training observations in class $k$. Estimating $f_k$ requires an
assumption about its form, and LDA's assumption is that it is Gaussian.

## One-feature LDA

### Gaussian class-conditional densities

With a single feature ($p = 1$), LDA assumes the statistical model

$$P(Y = k) = \pi_k, \qquad X \mid Y = k \sim N\left(\mu_k, \sigma^2\right), \qquad k = 1, \ldots, K,$$

where

- $\mu_k = E[X \mid Y = k]$ is the mean of the feature in class $k$, and
- $\sigma^2 = Var[X \mid Y = k]$ is the variance of the feature, **the same in
every class**.

The class-conditional density is therefore the normal density

$$f_k(x) = \frac{1}{\sqrt{2\pi}\,\sigma} \exp\left[-\frac{(x - \mu_k)^2}{2\sigma^2}\right],$$

where $\pi$ without a subscript is the constant $3.14159\ldots$, unrelated to the
prior probabilities $\pi_k$.

The classes differ only in where their densities are centered. The shared
variance is what makes the classifier linear.

### Discriminant function

Since $p_k(x)$ is proportional to $\pi_k f_k(x)$ and the logarithm is
increasing, the class with the largest posterior probability is also the class
with the largest $\log\left(\pi_k f_k(x)\right)$. Expanding the square,

$$\log\left(\pi_k f_k(x)\right) = \log \pi_k - \log\left(\sqrt{2\pi}\,\sigma\right) - \frac{x^2}{2\sigma^2} + x\frac{\mu_k}{\sigma^2} - \frac{\mu_k^2}{2\sigma^2}.$$

The second and third terms do not involve $k$ — the third only because
$\sigma^2$ is shared — so they cannot change which class is largest. Dropping
them leaves the **discriminant function**

$$\delta_k(x) = x\frac{\mu_k}{\sigma^2} - \frac{\mu_k^2}{2\sigma^2} + \log \pi_k,$$

where

- $\delta_k(x)$ is the discriminant score of class $k$ at feature value $x$.

The classifier assigns an observation to the class with the largest
$\delta_k(x)$. Each $\delta_k$ is a linear function of $x$, with slope
$\mu_k/\sigma^2$ and intercept $\log \pi_k - \mu_k^2/\left(2\sigma^2\right)$.

The dropped terms are added to every class's $\log\left(\pi_k f_k(x)\right)$,
so their exponentials are common factors of every $\pi_k f_k(x)$. They cancel
from the ratio in Bayes' theorem and the posterior probabilities can be
recovered from the discriminant scores alone:

$$p_k(x) = \frac{e^{\delta_k(x)}}{\sum_{l=1}^K e^{\delta_l(x)}}.$$

### Decision boundary

With $K = 2$ classes, $p_1(x) = 1 - p_2(x)$, and the log posterior odds of
class $2$ are

$$\log\left(\frac{p_2(x)}{1 - p_2(x)}\right) = \delta_2(x) - \delta_1(x) = \underbrace{\log\frac{\pi_2}{\pi_1} - \frac{\mu_2^2 - \mu_1^2}{2\sigma^2}}_{\beta_0} + \underbrace{\frac{\mu_2 - \mu_1}{\sigma^2}}_{\beta_1}\, x.$$

The log-odds are linear in $x$, so LDA's posterior probability has exactly the
form of a simple logistic regression. The two methods differ in how they
estimate $\beta_0$ and $\beta_1$, not in the shape of the curve they fit.

The **decision boundary** is the set of $x$ where the two classes are equally
probable, $\delta_1(x) = \delta_2(x)$. Setting the log posterior odds to zero
and solving,

$$x^* = \frac{\mu_1 + \mu_2}{2} + \frac{\sigma^2}{\mu_2 - \mu_1}\log\frac{\pi_1}{\pi_2},$$

where

- $x^*$ is the feature value at which the classifier switches from one class
to the other.

With equal prior probabilities the logarithm is zero and the boundary is the
midpoint of the two means. Take $\mu_2 > \mu_1$. If class $2$ is the rarer
class, $\pi_2 < \pi_1$, the logarithm is positive and the boundary moves to the
right, toward $\mu_2$: an observation must look more like a typical class-$2$
observation before the classifier assigns it to the rarer class.

The figure below uses two classes with $\mu_1 = -1.25$, $\mu_2 = 1.25$, and
$\sigma = 1$. Each curve is $\pi_k f_k(x)$, so the boundary is where the two
curves cross.

::: {.callout-note collapse="true"}
### Code
```{r theory-1d}
#| output: false
mu_1 <- -1.25
mu_2 <-  1.25
sigma <- 1
pi_unequal <- c(0.8, 0.2)

boundary_1d <- function(pi1, pi2) {
  (mu_1 + mu_2) / 2 + sigma^2 / (mu_2 - mu_1) * log(pi1 / pi2)
}
# Plot title "pi_1 = a, pi_2 = b; boundary x* = c" as a plotmath expression.
prior_title <- function(pi1, pi2, x_star) {
  bquote(paste(pi[1] == .(sprintf("%.1f", pi1)), ", ",
               pi[2] == .(sprintf("%.1f", pi2)), "; boundary ",
               x^"*" == .(sprintf("%.2f", x_star))))
}
x_star_equal   <- boundary_1d(0.5, 0.5)
x_star_unequal <- boundary_1d(pi_unequal[1], pi_unequal[2])

theory_grid <- tibble(x = seq(-4.5, 4.5, length.out = 500))

weighted_curves <- function(pi1, pi2) {
  theory_grid |>
    mutate(`k = 1` = pi1 * dnorm(x, mu_1, sigma),
           `k = 2` = pi2 * dnorm(x, mu_2, sigma)) |>
    pivot_longer(-x, names_to = "class", values_to = "density")
}

weighted_plot <- function(pi1, pi2, x_star) {
  ggplot(weighted_curves(pi1, pi2),
         aes(x = x, y = density, color = class, linetype = class)) +
    geom_line(linewidth = 0.9) +
    geom_vline(xintercept = x_star, linetype = "dotted") +
    scale_color_manual(values = class_colors) +
    coord_cartesian(ylim = c(0, 0.33)) +
    labs(x = "x", y = expression(pi[k] * f[k](x)), color = "Class",
         linetype = "Class", title = prior_title(pi1, pi2, x_star))
}

theory_equal_plot   <- weighted_plot(0.5, 0.5, x_star_equal)
theory_unequal_plot <- weighted_plot(pi_unequal[1], pi_unequal[2],
                                     x_star_unequal)

delta_lines <- theory_grid |>
  mutate(`k = 1` = x * mu_1 / sigma^2 - mu_1^2 / (2 * sigma^2) +
                   log(pi_unequal[1]),
         `k = 2` = x * mu_2 / sigma^2 - mu_2^2 / (2 * sigma^2) +
                   log(pi_unequal[2])) |>
  pivot_longer(-x, names_to = "class", values_to = "delta")

theory_delta_plot <- ggplot(delta_lines,
                            aes(x = x, y = delta, color = class,
                                linetype = class)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = x_star_unequal, linetype = "dotted") +
  scale_color_manual(values = class_colors) +
  labs(x = "x", y = expression(delta[k](x)), color = "Class",
       linetype = "Class",
       title = prior_title(pi_unequal[1], pi_unequal[2], x_star_unequal))
```
:::

::: {.panel-tabset}
#### Equal priors
```{r theory-equal-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
theory_equal_plot
```

#### Unequal priors
```{r theory-unequal-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
theory_unequal_plot
```

#### Discriminant functions
```{r theory-delta-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
theory_delta_plot
```
:::

With equal priors the boundary sits at the midpoint,
$x^* = `r sprintf("%.2f", x_star_equal)`$. Shrinking $\pi_2$ to
`r pi_unequal[2]` scales the class-$2$ curve down, so the crossing moves right
to $x^* = `r sprintf("%.2f", x_star_unequal)`$. The third tab plots
$\delta_1(x)$ and $\delta_2(x)$ for the unequal priors: two straight lines that
cross at the same $x^*$. A linear discriminant function for each class, with
the boundary where two of them are equal, is the "linear" in linear
discriminant analysis.

Had the variances differed, the term $-x^2/\left(2\sigma_k^2\right)$, where $\sigma_k^2$
is the class-$k$ variance, would carry a
different coefficient in each class, would not drop out, and would make each
discriminant quadratic in $x$.

### Parameter estimation

The parameters $\pi_k$, $\mu_k$, and $\sigma^2$ are estimated from training
observations $(x_1, y_1), \ldots, (x_n, y_n)$, here with $x_i$ the scalar
feature value of observation $i$:

$$\hat\pi_k = \frac{n_k}{n}, \qquad \hat\mu_k = \frac{1}{n_k}\sum_{i:\, y_i = k} x_i, \qquad \hat\sigma^2 = \frac{1}{n - K}\sum_{k=1}^K \sum_{i:\, y_i = k}\left(x_i - \hat\mu_k\right)^2,$$

where

- $n_k$ is the number of training observations in class $k$, so
$\sum_k n_k = n$,
- $\sum_{i:\, y_i = k}$ sums over the training observations in class $k$, and
- $\hat\sigma^2$ is the **pooled variance**: squared deviations from each
observation's own class mean, summed over all classes and divided by $n - K$.

These are maximum likelihood estimates. Under the generative model the joint
density of observation $i$ is $P(Y = y_i)\, f_{y_i}(x_i) = \pi_{y_i} f_{y_i}(x_i)$,
so for independent observations the log-likelihood is

$$\ell(\pi_1, \ldots, \pi_K, \mu_1, \ldots, \mu_K, \sigma^2) = \sum_{k=1}^K n_k \log \pi_k \;+\; \sum_{k=1}^K \sum_{i:\, y_i = k}\left[-\frac{1}{2}\log\left(2\pi\sigma^2\right) - \frac{(x_i - \mu_k)^2}{2\sigma^2}\right].$$

The first sum involves only the priors; maximizing it subject to
$\sum_k \pi_k = 1$ (with a Lagrange multiplier) gives $\hat\pi_k = n_k/n$. In
the second, $\mu_k$ appears only in class $k$'s terms, so its maximizer is the
class-$k$ sample mean. Setting the derivative with respect to $\sigma^2$ to
zero gives the pooled sum of squares divided by $n$; dividing by $n - K$
instead, one degree of freedom per estimated mean, makes it unbiased, in the
same way that $RSS/(n-2)$ replaced $RSS/n$ in simple linear regression.
`MASS::lda()` uses $n - K$.

Logistic regression maximizes a different likelihood, the conditional one,
$\prod_{i=1}^n p(x_i)^{y_i}\left[1 - p(x_i)\right]^{1 - y_i}$ (with $y_i$
coded $0/1$ as in the logistic regression lectures), which has no closed-form
maximizer. LDA maximizes the joint likelihood of features and
responses under a stronger assumption — Gaussian features with a shared
variance — and every estimate is a sample proportion, mean, or variance. Both
methods end with log-odds linear in $x$; their estimates of $\beta_0$ and
$\beta_1$ differ because they maximize different likelihoods.

::: {.callout-note collapse="true"}
### For example,
LDA's estimates have closed forms, and separation does not prevent them from
existing, though a singular pooled covariance still would. The previous lecture's completely separated toy data,
$x = 1, \ldots, 10$ with $y = 1$ exactly when $x > 5$, left logistic
regression with no finite maximum likelihood estimate. `MASS::lda()`, used
again in the worked example below, fits LDA to the same data:

```{r toy-lda}
toy <- tibble(x = 1:10, y = factor(as.numeric(x > 5)))
toy_lda <- MASS::lda(y ~ x, data = toy)
toy_lda$means
```

The class means are `r toy_lda$means[1]` and `r toy_lda$means[2]`, the priors
are equal, and so the LDA boundary is the midpoint,
$x^* = `r mean(toy_lda$means)`$.
:::

## Multiple-feature LDA

### Multivariate Gaussian densities

With $p > 1$ features, $X = (X_1, \ldots, X_p)^\top$ is a random column vector
whose entries are the features, and $x$ is a $p$-vector value it takes. A
value $x$ holds the same kind of entries as a row of the model matrix
$\mathbf{X}$ but without the leading $1$, and it is a column because it is the
argument of a multivariate density. LDA assumes the statistical model

$$P(Y = k) = \pi_k, \qquad X \mid Y = k \sim N_p\left(\mu_k, \Sigma\right), \qquad k = 1, \ldots, K,$$

where

- $N_p$ is the $p$-dimensional multivariate normal distribution,
- $\mu_k = E[X \mid Y = k] = (\mu_{k1}, \ldots, \mu_{kp})^\top$ is the
$p$-vector of feature means in class $k$, and
- $\Sigma = Cov[X \mid Y = k]$ is the $p \times p$ covariance matrix of the
features, **the same in every class**, with feature variances on its diagonal
and covariances between pairs of features off it; it is symmetric and
positive definite.

The class-conditional density is therefore the multivariate normal density

$$f_k(x) = \frac{1}{(2\pi)^{p/2}\,|\Sigma|^{1/2}} \exp\left[-\frac{1}{2}\left(x - \mu_k\right)^\top \Sigma^{-1} \left(x - \mu_k\right)\right],$$

where $|\Sigma|$ is the determinant of $\Sigma$.

Each class's density has contours that are nested ellipses centered at
$\mu_k$. Sharing
$\Sigma$ means every class's ellipses have the same shape and orientation and
differ only in their center.

### Matrix-form discriminant

The derivation follows the one-feature case. Because $\Sigma^{-1}$ is
symmetric, the scalar $\mu_k^\top \Sigma^{-1} x$ equals its own transpose
$x^\top \Sigma^{-1} \mu_k$, and the quadratic form expands to

$$\left(x - \mu_k\right)^\top \Sigma^{-1} \left(x - \mu_k\right) = x^\top \Sigma^{-1} x - 2\, x^\top \Sigma^{-1} \mu_k + \mu_k^\top \Sigma^{-1} \mu_k.$$

Then

$$\log\left(\pi_k f_k(x)\right) = \log \pi_k - \frac{p}{2}\log(2\pi) - \frac{1}{2}\log|\Sigma| - \frac{1}{2}x^\top \Sigma^{-1} x + x^\top \Sigma^{-1} \mu_k - \frac{1}{2}\mu_k^\top \Sigma^{-1} \mu_k.$$

The second, third, and fourth terms do not involve $k$; the fourth only
because $\Sigma$ is shared. Dropping them,

$$\delta_k(x) = x^\top \Sigma^{-1} \mu_k - \frac{1}{2}\mu_k^\top \Sigma^{-1} \mu_k + \log \pi_k,$$

which is linear in $x$ and reduces to the one-feature discriminant when
$\Sigma = \sigma^2$. As before, the classifier assigns the class with the
largest $\delta_k(x)$, and $p_k(x) = e^{\delta_k(x)} / \sum_l e^{\delta_l(x)}$.

### Linear decision boundary

With $K = 2$, subtracting the two discriminants gives the log posterior odds
of class $2$,

$$\log\left(\frac{p_2(x)}{1 - p_2(x)}\right) = \delta_2(x) - \delta_1(x) = \beta_0 + x^\top\beta,$$

where

- $\beta = \Sigma^{-1}(\mu_2 - \mu_1)$ is a $p$-vector of slopes, one per
feature, with no intercept entry, and
- $\beta_0 = \log(\pi_2/\pi_1) - \frac{1}{2}\left(\mu_2^\top \Sigma^{-1} \mu_2 - \mu_1^\top \Sigma^{-1} \mu_1\right)$
is the intercept.

This is again the form of a multiple logistic regression with $p$ features.
The decision boundary $\{x : \beta_0 + x^\top\beta = 0\}$ is a hyperplane in
the $p$-dimensional feature space — a line when $p = 2$.

The slopes are $\Sigma^{-1}(\mu_2 - \mu_1)$, not $\mu_2 - \mu_1$, so the
boundary is generally not perpendicular to the segment joining the two means:
$\Sigma^{-1}$ tilts it according to how the features covary. With equal
priors it still passes through the midpoint $(\mu_1 + \mu_2)/2$, since
substituting that point gives
$\frac{1}{2}\left(\mu_1 + \mu_2\right)^\top\Sigma^{-1}\left(\mu_2 - \mu_1\right) = \frac{1}{2}\left(\mu_2^\top \Sigma^{-1} \mu_2 - \mu_1^\top \Sigma^{-1} \mu_1\right)$.

The figure below uses $p = 2$, $\mu_1 = (-1, 0)^\top$, $\mu_2 = (1, 0)^\top$,
unit variances, and a correlation of $0.6$ between the two features. The
ellipses are contours of each $f_k$.

::: {.callout-note collapse="true"}
### Code
```{r theory-2d}
#| output: false
mu2d_1  <- c(-1, 0)
mu2d_2  <- c( 1, 0)
Sigma2d <- matrix(c(1, 0.6, 0.6, 1), nrow = 2)
Sigma2d_inv <- solve(Sigma2d)

dmvnorm2 <- function(x1, x2, mu) {
  d <- cbind(x1 - mu[1], x2 - mu[2])
  q <- rowSums((d %*% Sigma2d_inv) * d)
  exp(-q / 2) / (2 * pi * sqrt(det(Sigma2d)))
}

grid2d <- expand_grid(x1 = seq(-4, 4, length.out = 200),
                      x2 = seq(-3.5, 3.5, length.out = 200)) |>
  mutate(`k = 1` = dmvnorm2(x1, x2, mu2d_1),
         `k = 2` = dmvnorm2(x1, x2, mu2d_2)) |>
  pivot_longer(c(`k = 1`, `k = 2`), names_to = "class",
               values_to = "density")

# Contours at Mahalanobis radii 1 and 2.
contour_levels <- exp(-c(1, 2)^2 / 2) / (2 * pi * sqrt(det(Sigma2d)))

boundary_2d <- function(pi1, pi2) {
  beta  <- Sigma2d_inv %*% (mu2d_2 - mu2d_1)
  beta0 <- log(pi2 / pi1) -
    0.5 * (t(mu2d_2) %*% Sigma2d_inv %*% mu2d_2 -
           t(mu2d_1) %*% Sigma2d_inv %*% mu2d_1)
  # beta0 + beta1 x1 + beta2 x2 = 0  =>  x2 = -(beta0 + beta1 x1) / beta2
  c(intercept = -beta0[1] / beta[2], slope = -beta[1] / beta[2])
}
line_equal   <- boundary_2d(0.5, 0.5)
line_unequal <- boundary_2d(pi_unequal[1], pi_unequal[2])

means2d <- tibble(class = c("k = 1", "k = 2"),
                  x1 = c(mu2d_1[1], mu2d_2[1]),
                  x2 = c(mu2d_1[2], mu2d_2[2]))

theory_2d_base <- ggplot(grid2d, aes(x = x1, y = x2)) +
  geom_contour(aes(z = density, color = class, linetype = class),
               breaks = contour_levels, linewidth = 0.8) +
  geom_point(data = means2d, aes(color = class, shape = class), size = 3) +
  annotate("segment", x = mu2d_1[1], y = mu2d_1[2],
           xend = mu2d_2[1], yend = mu2d_2[2],
           color = "grey50", linewidth = 0.4) +
  scale_color_manual(values = class_colors) +
  coord_fixed(xlim = c(-4, 4), ylim = c(-3.5, 3.5)) +
  labs(x = expression(X[1]), y = expression(X[2]), color = "Class",
       linetype = "Class", shape = "Class")

theory_2d_densities <- theory_2d_base +
  labs(title = "Class densities sharing one covariance matrix")

theory_2d_equal <- theory_2d_base +
  geom_abline(intercept = line_equal["intercept"],
              slope = line_equal["slope"], linewidth = 0.9) +
  labs(title = expression(paste("Boundary with ", pi[1] == 0.5, ", ",
                                pi[2] == 0.5)))

theory_2d_unequal <- theory_2d_base +
  geom_abline(intercept = line_equal["intercept"],
              slope = line_equal["slope"], linewidth = 0.5,
              color = "grey60", linetype = "dotted") +
  geom_abline(intercept = line_unequal["intercept"],
              slope = line_unequal["slope"], linewidth = 0.9) +
  labs(title = bquote(paste("Boundary with ", pi[1] == .(pi_unequal[1]), ", ",
                            pi[2] == .(pi_unequal[2]),
                            " (equal-prior boundary dotted)")))
```
:::

::: {.panel-tabset}
#### Class densities
```{r theory-2d-densities-figure}
#| echo: false
#| fig-width: 6
#| fig-height: 5
theory_2d_densities
```

#### Equal priors
```{r theory-2d-equal-figure}
#| echo: false
#| fig-width: 6
#| fig-height: 5
theory_2d_equal
```

#### Unequal priors
```{r theory-2d-unequal-figure}
#| echo: false
#| fig-width: 6
#| fig-height: 5
theory_2d_unequal
```
:::

The two sets of ellipses are translates of one another. With equal priors the
boundary passes through the midpoint of the grey segment joining the means,
but it is not perpendicular to that segment: the positive correlation between
the features tilts it. Shrinking $\pi_2$ to `r pi_unequal[2]` shifts the line
parallel to itself, away from $\mu_1$ and toward $\mu_2$, because the prior
enters only $\beta_0$.

### Pooled covariance estimate

The estimates generalize directly. The priors are still $\hat\pi_k = n_k/n$,
and with $x_i = (x_{i1}, \ldots, x_{ip})^\top$ now the $p$-vector of feature
values for observation $i$ (again without the leading $1$),

$$\hat\mu_k = \frac{1}{n_k}\sum_{i:\, y_i = k} x_i, \qquad \hat\Sigma = \frac{1}{n - K}\sum_{k=1}^K \sum_{i:\, y_i = k}\left(x_i - \hat\mu_k\right)\left(x_i - \hat\mu_k\right)^\top,$$

where

- $\hat\mu_k$ is the vector of class-$k$ sample means, one per feature, and
- $\hat\Sigma$ is the **pooled covariance matrix**, the sum of each class's
matrix of cross-products of deviations from its own mean, divided by $n - K$.

Substituting $\hat\pi_k$, $\hat\mu_k$, and $\hat\Sigma$ into $\delta_k(x)$ gives
the estimated discriminant $\hat\delta_k(x)$, the estimated posterior
probabilities
$\hat p_k(x) = e^{\hat\delta_k(x)} / \sum_l e^{\hat\delta_l(x)}$, and the LDA
classifier, which assigns each observation to the class with the largest
$\hat\delta_k(x)$.

## Classifier evaluation

The classification lecture introduced the confusion matrix and the
sensitivity, specificity, and precision read from it, alongside the error
rate. This section develops the two-class case in full: the threshold that
produces a confusion matrix, how sensitivity and specificity trade off as the
threshold moves, and the ROC curve and AUC that summarize every threshold at
once.

An error rate counts every mistake the same way. Failing to flag a customer who goes on to default and flagging a
customer who does not are different mistakes with different costs, and a
classifier can have a small error rate while making almost every mistake of
one kind.

### Confusion matrix

With two classes, call the class of interest **positive** and the other
**negative**. For a classifier that produces an estimated probability
$\hat p(x)$ of the positive class, a **threshold** $t$ turns probabilities into
classes: predict positive when $\hat p(x) > t$. The Bayes classifier uses
$t = 0.5$.

As in the classification lecture, a confusion matrix cross-tabulates the true
class against the predicted class:

::: {.callout-note collapse="true"}
### Code
```{r confusion-template}
#| output: false
confusion_template <- tibble(
  Truth = c("Positive", "Negative"),
  `Predicted positive` = c("True positive (TP)", "False positive (FP)"),
  `Predicted negative` = c("False negative (FN)", "True negative (TN)")
) |>
  knitr::kable(align = "lcc")
```
:::

```{r confusion-template-table}
#| echo: false
confusion_template
```

where

- $TP$ is the number of positive observations predicted positive,
- $FN$ is the number of positive observations predicted negative,
- $FP$ is the number of negative observations predicted positive, and
- $TN$ is the number of negative observations predicted negative.

The diagonal holds the correct classifications, so the error rate is
$(FP + FN)/n$.

### Sensitivity and specificity

Dividing each row by its total gives two rates that condition on the true
class:

$$\text{Sensitivity} = \frac{TP}{TP + FN}, \qquad \text{Specificity} = \frac{TN}{TN + FP},$$

where

- **sensitivity**, also called **recall** or the **true positive rate**, is the
proportion of positive observations the classifier catches, an estimate of
$P\left(\hat y = \text{positive} \mid Y = \text{positive}\right)$, and
- **specificity**, also called the **true negative rate**, is the proportion of
negative observations the classifier correctly leaves alone, an estimate of
$P\left(\hat y = \text{negative} \mid Y = \text{negative}\right)$.

Their complements are error rates within each true class:
$1 - \text{sensitivity} = FN/(TP + FN)$ is the **false negative rate**, and
$1 - \text{specificity} = FP/(TN + FP)$ is the **false positive rate**.

### Classification threshold

The threshold $t = 0.5$ is the Bayes classifier's: if $\hat p(x)$ equalled
the true $p(x)$, it would minimize the expected error rate. An estimated
$\hat p(x)$ carries no such guarantee, and even for the Bayes classifier the
overall error rate need not be the quantity of interest.

Lowering $t$ enlarges the set $\left\{x : \hat p(x) > t\right\}$ of feature
values classified positive, so every
observation predicted positive at the old threshold is still predicted
positive at the new one. Consequently $TP$ and $FP$ can only grow while $FN$
and $TN$ can only shrink: sensitivity rises or stays the same, and
specificity falls or stays the same. Raising $t$ does the reverse. No
threshold improves both, so choosing $t$ means choosing a tradeoff between the
two kinds of mistake.

::: {.callout-note collapse="true"}
### Beyond this course
If the two mistakes have known costs, $c_{FN}$ for a missed positive and
$c_{FP}$ for a false alarm, the expected cost of predicting positive at $x$ is
$c_{FP}\left[1 - p(x)\right]$ and of predicting negative is $c_{FN}\, p(x)$.
Predicting positive is the cheaper choice exactly when
$p(x) > c_{FP}/(c_{FP} + c_{FN})$, so that ratio is the cost-minimizing
threshold. Equal costs recover $t = 0.5$.
:::

### ROC curve

The **ROC curve** (receiver operating characteristic curve) plots sensitivity
against $1 - \text{specificity}$ — the true positive rate against the false
positive rate — at every threshold. At $t = 1$ nothing is predicted positive,
giving the point $(0, 0)$; at $t = 0$ everything is, giving $(1, 1)$; and the
curve traces every tradeoff in between.

For the two-class, one-feature model of the decision-boundary figure, with
equal priors and class $2$ positive, $p(x) > t$ is equivalent to $x > c$
for a cutoff $c$ that decreases as $t$ decreases, because the posterior is
increasing in $x$. Sensitivity is then the area under $f_2$ to the right of
$c$, and $1 - \text{specificity}$ is the area under $f_1$ to the right of $c$.
The slider below moves the threshold. The left panel shades the areas that
determine sensitivity and the false positive rate, and the right panel marks
where the classifier lands on the ROC curve.

::: {.callout-note collapse="true"}
### Code
```{r theory-roc}
#| output: false
# Equal-prior posterior: log-odds = beta1 * x with beta1 = (mu_2 - mu_1) / sigma^2.
beta1_theory <- (mu_2 - mu_1) / sigma^2
cutoff_for <- function(t) qlogis(t) / beta1_theory

roc_theory <- tibble(c = seq(-8, 8, length.out = 4000)) |>
  mutate(fpr = 1 - pnorm(c, mu_1, sigma),
         tpr = 1 - pnorm(c, mu_2, sigma))

auc_theory_formula <- pnorm((mu_2 - mu_1) / (sigma * sqrt(2)))
roc_theory_sorted  <- arrange(roc_theory, fpr)
auc_theory_trap    <- with(roc_theory_sorted,
                           sum(diff(fpr) * (head(tpr, -1) + tail(tpr, -1)) / 2))

# Everything the slider needs, indexed by t = 0.01, 0.02, ..., 0.99, is
# computed here and passed to the page, which only looks values up.
slider_thresholds <- tibble(t = (1:99) / 100) |>
  mutate(cutoff = cutoff_for(t),
         fpr = 1 - pnorm(cutoff, mu_1, sigma),
         tpr = 1 - pnorm(cutoff, mu_2, sigma))

slider_densities <- theory_grid |>
  mutate(`k = 1 (negative)` = dnorm(x, mu_1, sigma),
         `k = 2 (positive)` = dnorm(x, mu_2, sigma)) |>
  pivot_longer(-x, names_to = "class", values_to = "density")
```
:::

```{r theory-roc-data}
#| echo: false
ojs_define(slider_thresholds = slider_thresholds,
           slider_densities  = slider_densities,
           slider_roc        = select(slice(roc_theory, seq(1, n(), by = 8)), fpr, tpr),
           slider_colors     = unname(class_colors))
```

```{ojs}
//| label: theory-roc-slider
//| echo: false
viewof t = Inputs.range([0.01, 0.99], {value: 0.5, step: 0.01, label: "Threshold t"})
```

```{ojs}
//| label: theory-roc-slider-plots
//| echo: false
{
  const thresholds = transpose(slider_thresholds);
  const densities = transpose(slider_densities);
  const roc = transpose(slider_roc);
  const now = thresholds[Math.round(t * 100) - 1];
  const classes = ["k = 1 (negative)", "k = 2 (positive)"];

  const densityPlot = Plot.plot({
    width: 340, height: 300,
    x: {label: "x"},
    y: {label: "Density fₖ(x)"},
    color: {domain: classes, range: slider_colors, legend: true},
    marks: [
      Plot.areaY(densities.filter(d => d.x > now.cutoff),
                 {x: "x", y1: 0, y2: "density", fill: "class", fillOpacity: 0.3}),
      Plot.line(densities.filter(d => d.class === classes[0]),
                {x: "x", y: "density", stroke: "class", strokeWidth: 2}),
      Plot.line(densities.filter(d => d.class === classes[1]),
                {x: "x", y: "density", stroke: "class", strokeDasharray: "6,4",
                 strokeWidth: 2}),
      Plot.ruleX([now.cutoff], {strokeDasharray: "2,3"})
    ]
  });

  const rocPlot = Plot.plot({
    width: 300, height: 300,
    x: {domain: [0, 1], label: "False positive rate (1 − specificity)"},
    y: {domain: [0, 1], label: "True positive rate (sensitivity)"},
    marks: [
      Plot.line([[0, 0], [1, 1]], {stroke: "gray", strokeDasharray: "4,4"}),
      Plot.line(roc, {x: "fpr", y: "tpr", strokeWidth: 2}),
      Plot.dot([now], {x: "fpr", y: "tpr", r: 6, fill: "white", stroke: "black",
                       strokeWidth: 2})
    ]
  });

  return html`<p style="text-align: center"><b>t = ${t.toFixed(2)}</b>,
    cutoff c = ${now.cutoff.toFixed(2)}; sensitivity = ${now.tpr.toFixed(2)},
    specificity = ${(1 - now.fpr).toFixed(2)}</p>
    <div style="display: flex; flex-wrap: wrap; gap: 1rem; justify-content: center">
      ${densityPlot}${rocPlot}
    </div>`;
}
```

Moving the slider from $t = 0.9$ to $t = 0.1$ slides the cutoff left, so the
shaded area under $f_2$ (sensitivity) and the shaded area under $f_1$ (false
positive rate) both grow, and the point climbs along the curve toward $(1, 1)$. The
dashed diagonal is the ROC curve of a classifier that ignores $x$ and
predicts positive at random: its true and false positive rates are equal at
every threshold. A useful classifier's curve bows toward the top-left corner,
where sensitivity is high and the false positive rate is low.

The ROC curve depends on a classifier only through how it *orders* the
observations. Any strictly increasing transformation of the score — $x$,
$\hat\delta_2(x) - \hat\delta_1(x)$, or $\hat p(x)$ — sorts the observations
identically, produces the same sequence of confusion matrices, and so traces
the same curve.

### AUC

The **AUC** is the area under the ROC curve, a single number summarizing the
curve over all thresholds at once. It has a probabilistic interpretation.
Let $S_+$ be the score of a randomly chosen positive observation and $S_-$ the
score of an independently chosen negative one, and let $g_-$ be the density of
$S_-$.
At cutoff $c$ the true positive rate is $P(S_+ > c)$ and the false positive
rate is $P(S_- > c)$, whose derivative with respect to $c$ is $-g_-(c)$. As
$c$ runs from $+\infty$ down to $-\infty$ the false positive rate runs from
$0$ to $1$, so

$$\text{AUC} = \int_0^1 \text{TPR}\; d\,\text{FPR} = \int_{-\infty}^{\infty} P(S_+ > c)\, g_-(c)\, dc = P(S_+ > S_-),$$

where

- $S_+$ and $S_-$ are the classifier's scores for a random positive and a
random negative observation, and
- the final equality conditions on $S_- = c$ and integrates over its density.

The AUC is the probability that the classifier scores a randomly chosen
positive observation above a randomly chosen negative one. A classifier that
ignores the features has $\text{AUC} = 0.5$; one that ranks every positive
above every negative has $\text{AUC} = 1$.

In the one-feature Gaussian model the score can be taken as $x$ itself, so
$S_+ - S_- \sim N\left(\mu_2 - \mu_1, 2\sigma^2\right)$ and

$$\text{AUC} = P(S_+ - S_- > 0) = \Phi\left(\frac{\Delta}{\sqrt{2}}\right), \qquad \Delta = \frac{\mu_2 - \mu_1}{\sigma},$$

where

- $\Phi$ is the standard normal cumulative distribution function, and
- $\Delta$ is the separation of the class means in standard deviations.

For the curve plotted above, $\Delta = `r mu_2 - mu_1`$ and
$\text{AUC} = \Phi\left(`r mu_2 - mu_1`/\sqrt{2}\right) = `r sprintf("%.5f", auc_theory_formula)`$;
numerically integrating the plotted curve gives
`r sprintf("%.5f", auc_theory_trap)`.

The same model gives the ROC curve in closed form. A cutoff $c$ has
$\text{FPR} = 1 - \Phi\left((c - \mu_1)/\sigma\right)$, so
$(c - \mu_1)/\sigma = -\Phi^{-1}(\text{FPR})$, and
$\text{TPR} = \Phi\left((\mu_2 - c)/\sigma\right)$ becomes

$$\text{TPR} = \Phi\left(\Delta + \Phi^{-1}(\text{FPR})\right).$$

The left panel below shades the area under the ROC curve of the previous
section, the $\Delta = `r mu_2 - mu_1`$ curve in the right panel. The right
panel draws the ROC curves for several values of $\Delta$, each labeled with its
AUC. The curve for $\Delta = 0$ is the diagonal.

::: {.callout-note collapse="true"}
### Code
```{r auc-figure}
#| output: false
auc_delta <- function(delta) pnorm(delta / sqrt(2))
delta_theory <- (mu_2 - mu_1) / sigma
deltas <- c(0, 1, delta_theory, 4)

auc_area_plot <- ggplot(roc_theory_sorted, aes(x = fpr, y = tpr)) +
  geom_area(fill = "grey70", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "grey40",
              linetype = "dashed") +
  geom_line(linewidth = 0.9) +
  annotate("text", x = 0.65, y = 0.3,
           label = sprintf("AUC = %.5f", auc_theory_trap)) +
  coord_fixed() +
  labs(x = "False positive rate (1 - specificity)",
       y = "True positive rate (sensitivity)",
       title = "Area under the ROC curve")

# Legend entries as plotmath: Delta = d, AUC = a.
curve_labels <- parse(text = sprintf('Delta == "%.1f" ~~ AUC == "%.3f"', deltas,
                                     auc_delta(deltas)))
auc_curves <- expand_grid(separation = deltas, fpr = seq(0, 1, length.out = 300)) |>
  mutate(tpr = pnorm(separation + qnorm(fpr)),
         curve = factor(separation, levels = deltas))

auc_curves_plot <- ggplot(auc_curves, aes(x = fpr, y = tpr, color = curve,
                                          linetype = curve)) +
  geom_line(linewidth = 0.9) +
  scale_color_viridis_d(end = 0.9, labels = curve_labels) +
  scale_linetype_discrete(labels = curve_labels) +
  coord_fixed() +
  labs(x = "False positive rate (1 - specificity)",
       y = "True positive rate (sensitivity)", color = NULL, linetype = NULL,
       title = "ROC curves by separation") +
  theme(legend.position = "inside", legend.position.inside = c(0.97, 0.03),
        legend.justification = c(1, 0))
```
:::

```{r auc-figure-display}
#| echo: false
#| layout-ncol: 2
#| fig-width: 4.5
#| fig-height: 4.5
auc_area_plot
auc_curves_plot
```

The curves in the right panel are nested: for every false positive rate, a
larger $\Delta$ gives a larger true positive rate, so one curve lies above
another exactly when its AUC is larger. The AUC therefore orders these
classifiers the same way a comparison of true positive rates at any single
false positive rate would. It grows with the separation of the class means
measured in standard deviations, from $0.5$ at $\Delta = 0$ toward $1$, and
does not depend on the priors or the threshold. When two classifiers' ROC
curves cross instead, each has the higher true positive rate over part of the
range of false positive rates, and the AUC, which averages the true positive
rate over all false positive rates, summarizes that disagreement in one number.

With training data the same probability is estimated by comparing every
positive observation with every negative one,

$$\widehat{\text{AUC}} = \frac{1}{n_+ n_-}\sum_{i:\, y_i = \text{positive}}\ \sum_{j:\, y_j = \text{negative}}\left[\mathrm{I}(s_i > s_j) + \tfrac{1}{2}\mathrm{I}(s_i = s_j)\right],$$

where

- $n_+$ and $n_-$ are the numbers of positive and negative training
observations,
- $s_i$ is the classifier's score for observation $i$, and
- tied scores count as half a correctly ordered pair.

This equals the area under the empirical ROC curve computed by the trapezoid
rule.

## Credit card default classification

**Question.** Can LDA identify credit card customers who will default from
their balance and student status, and how does it compare with the logistic
regressions fit to the same customers in the previous lectures?

**Data.** `ISLR2::Default` records `r format(nrow(Default), big.mark = ",")`
customers. The response is `default`, with `Yes` the positive class. The
explanatory variables are `balance`, the customer's monthly credit card
balance in dollars, and `student`, a two-level factor. The features are
`balance` itself and the indicator
$D = \mathrm{I}(\texttt{student} = \texttt{Yes})$.

### Class summaries

The quantities LDA estimates — class proportions, class means, and a common
spread — can all be computed before fitting anything.

::: {.callout-note collapse="true"}
### Code
```{r default-setup}
#| output: false
Default <- Default |>
  mutate(default01   = as.numeric(default == "Yes"),
         student01   = as.numeric(student == "Yes"),
         default_jit = default01 + runif(n(), -0.03, 0.03),
         student_jit = student01 + runif(n(), -0.2, 0.2))

class_summary <- Default |>
  group_by(default) |>
  summarize(n = n(),
            prop = n() / nrow(Default),
            mean_balance = mean(balance),
            sd_balance = sd(balance),
            prop_student = mean(student01),
            n_zero = sum(balance == 0),
            .groups = "drop")

class_summary_table <- class_summary |>
  transmute(default,
            n = format(n, big.mark = ","),
            prop = sprintf("%.4f", prop),
            mean_balance = format(round(mean_balance), big.mark = ","),
            sd_balance = format(round(sd_balance), big.mark = ","),
            prop_student = sprintf("%.3f", prop_student)) |>
  knitr::kable(col.names = c("Default", "Customers", "Proportion",
                             "Mean balance (\\$)", "SD of balance (\\$)",
                             "Proportion students"),
               align = "cccccc")

summary_of <- function(k, col) class_summary[[col]][class_summary$default == k]

eda_plot <- ggplot(Default, aes(x = balance)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 100,
                 boundary = 0, fill = "grey70", color = "grey40") +
  facet_grid(default ~ student,
             labeller = labeller(default = label_both, student = label_both)) +
  labs(x = "Credit card balance ($)", y = "Density",
       title = "ISLR2::Default, balance within each default-by-student group")
```
:::

```{r class-summary-table}
#| echo: false
class_summary_table
```

```{r default-eda-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
eda_plot
```

Only `r format(summary_of("Yes", "n"), big.mark = ",")` customers, a proportion
of `r sprintf("%.4f", summary_of("Yes", "prop"))`, defaulted. Defaulters carry
an average balance of
\$`r format(round(summary_of("Yes", "mean_balance")), big.mark = ",")`
against \$`r format(round(summary_of("No", "mean_balance")), big.mark = ",")`
for non-defaulters, and the histograms in the bottom row sit well to the right
of those in the top row whether or not the customer is a student.

Two aspects of the data strain LDA's assumptions. The standard deviations of
balance differ between classes,
\$`r format(round(summary_of("No", "sd_balance")), big.mark = ",")` against
\$`r format(round(summary_of("Yes", "sd_balance")), big.mark = ",")`, where
LDA assumes one shared variance. And the non-defaulter histograms have a spike
at zero — `r format(summary_of("No", "n_zero"), big.mark = ",")` customers
carry no balance, none of whom defaulted — which no Gaussian density
reproduces.

Students make up a larger share of defaulters
(`r sprintf("%.3f", summary_of("Yes", "prop_student"))`) than of
non-defaulters (`r sprintf("%.3f", summary_of("No", "prop_student"))`).

### Balance alone

`MASS::lda()` takes the same formula interface as `glm()`:

```{r balance-lda}
balance_lda <- MASS::lda(default ~ balance, data = Default)
balance_lda
```

::: {.callout-note collapse="true"}
### Code
```{r lda-by-hand}
#| output: false
# Estimates of pi_k, mu_k, Sigma and the implied log posterior odds
# beta0 + x' beta of the second class, computed from the formulas above.
lda_by_hand <- function(X, y) {
  X <- as.matrix(X)
  classes <- levels(y)
  n <- nrow(X)
  K <- length(classes)
  pi_hat <- as.numeric(table(y)) / n
  mu_hat <- do.call(rbind, lapply(classes, function(k) {
    colMeans(X[y == k, , drop = FALSE])
  }))
  rownames(mu_hat) <- classes
  Sigma_hat <- Reduce(`+`, lapply(classes, function(k) {
    Z <- sweep(X[y == k, , drop = FALSE], 2, mu_hat[k, ])
    crossprod(Z)
  })) / (n - K)
  beta  <- solve(Sigma_hat, mu_hat[2, ] - mu_hat[1, ])
  beta0 <- log(pi_hat[2] / pi_hat[1]) -
    0.5 * (sum(mu_hat[2, ] * solve(Sigma_hat, mu_hat[2, ])) -
           sum(mu_hat[1, ] * solve(Sigma_hat, mu_hat[1, ])))
  list(pi = pi_hat, mu = mu_hat, Sigma = Sigma_hat,
       beta0 = beta0, beta = beta)
}

hand_1 <- lda_by_hand(Default["balance"], Default$default)
sigma_hat_1 <- sqrt(hand_1$Sigma[1, 1])
x_star_lda  <- -hand_1$beta0 / hand_1$beta[1]

balance_glm <- glm(default ~ balance, data = Default, family = "binomial")
b_glm_1     <- coef(balance_glm)
x_star_glm  <- -b_glm_1[1] / b_glm_1[2]

post_1 <- predict(balance_lda)$posterior[, "Yes"]
max_diff_1 <- max(abs(post_1 - plogis(hand_1$beta0 +
                                       hand_1$beta[1] * Default$balance)))

coef_table_1 <- tibble(
  term = c("Intercept $\\hat\\beta_0$", "`balance` $\\hat\\beta_1$"),
  lda = c(hand_1$beta0, hand_1$beta[1]),
  glm = unname(b_glm_1)
) |>
  mutate(across(c(lda, glm), \(v) sprintf("%.6f", v))) |>
  knitr::kable(col.names = c("Coefficient", "LDA", "Logistic regression"),
               align = "lcc")
```
:::

The output lists three things, and the prior probabilities come first.

The **prior probabilities** are $\hat\pi_{\text{No}} = `r sprintf("%.4f", hand_1$pi[1])`$
and $\hat\pi_{\text{Yes}} = `r sprintf("%.4f", hand_1$pi[2])`$, the class
proportions in the training data. They are the posterior probabilities LDA
would assign with no feature information. A classifier that used only them
would predict `No` for every customer and misclassify the
`r sprintf("%.2f", 100 * hand_1$pi[2])`% who defaulted; that error rate is the
baseline any classifier using features has to beat.

The **group means** are $\hat\mu_{\text{No}} = `r format(round(hand_1$mu[1, 1], 2), nsmall = 2)`$
and $\hat\mu_{\text{Yes}} = `r format(round(hand_1$mu[2, 1], 2), nsmall = 2)`$
dollars, the mean balances already seen in the class summaries. The pooled
standard deviation is $\hat\sigma = `r sprintf("%.2f", sigma_hat_1)`$, between
the two class-specific standard deviations.

The **coefficient of the linear discriminant**,
`r signif(balance_lda$scaling[1, 1], 4)`, equals
$1/\hat\sigma = `r signif(1 / sigma_hat_1, 4)`$: `MASS::lda()`
reports the discriminant direction rescaled so that the discriminant has
variance $1$ within each class, and with one feature the direction itself is
just "larger balance points toward `Yes`."

The log posterior odds derived above are more directly interpretable.
Substituting the estimates into $\beta_0$ and $\beta_1$, and placing the
coefficients of the simple logistic regression from the earlier lecture
beside them:

```{r coef-table-1}
#| echo: false
coef_table_1
```

The intercept $\hat\beta_0 = `r sprintf("%.3f", hand_1$beta0)`$ is LDA's
estimated log posterior odds of default for a customer carrying no balance, a
posterior probability of
`r format(signif(plogis(hand_1$beta0), 2), scientific = FALSE)`. Each additional
\$100 of balance adds `r sprintf("%.3f", 100 * hand_1$beta[1])` to the log
posterior odds, against `r sprintf("%.3f", 100 * b_glm_1[2])` in the logistic
regression. Applying the logistic function to
$\hat\beta_0 + \hat\beta_1 x$ reproduces the posterior probabilities returned
by `predict()` to within $`r sci(max_diff_1, 1)`$.

The decision boundary is
$\hat x^* = -\hat\beta_0/\hat\beta_1 = `r format(round(x_star_lda), big.mark = ",")`$
dollars for LDA and
`r format(round(x_star_glm), big.mark = ",")` dollars for logistic regression.
Both lie far above the midpoint of the class means,
`r format(round(mean(hand_1$mu[, 1])), big.mark = ",")` dollars, for
different reasons. In LDA's boundary formula the shift comes from the prior
term
$\hat\sigma^2\log\left(\hat\pi_{\text{No}}/\hat\pi_{\text{Yes}}\right)/\left(\hat\mu_{\text{Yes}} - \hat\mu_{\text{No}}\right)$,
which adds `r format(round(x_star_lda - mean(hand_1$mu[, 1])), big.mark = ",")`
dollars because $\hat\pi_{\text{Yes}}$ is so small. Logistic regression has
no priors or class means; its boundary is high because defaults are rare
among customers at all but the largest balances, so its fitted probability
reaches $0.5$ only there.

The tabs below build the fitted classifier from its pieces: the Gaussian
class-conditional densities, the same densities weighted by the priors, the
two discriminant functions, and the resulting posterior probability.

::: {.callout-note collapse="true"}
### Code
```{r balance-tabset}
#| output: false
balance_grid <- tibble(balance = seq(0, 2800, length.out = 500))

gauss_curves <- tibble(default = factor(c("No", "Yes"))) |>
  mutate(mu = hand_1$mu[, 1], prior = hand_1$pi) |>
  cross_join(balance_grid) |>
  mutate(f = dnorm(balance, mu, sigma_hat_1),
         weighted = prior * f,
         delta = balance * mu / sigma_hat_1^2 - mu^2 / (2 * sigma_hat_1^2) +
                 log(prior))

gauss_plot <- ggplot(Default, aes(x = balance)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 100, boundary = 0,
                 fill = "grey80", color = "grey55") +
  geom_line(data = gauss_curves, aes(y = f, color = default,
                                     linetype = default),
            linewidth = 0.9) +
  facet_wrap(~ default, ncol = 1, labeller = label_both) +
  scale_color_manual(values = class_colors) +
  labs(x = "Credit card balance ($)", y = "Density",
       color = "Default", linetype = "Default",
       title = expression(paste("Histograms with fitted ",
                                N(hat(mu)[k], hat(sigma)^2), " densities")))

weighted_default_plot <- ggplot(gauss_curves,
                                aes(x = balance, y = weighted,
                                    color = default, linetype = default)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = x_star_lda, linetype = "dotted") +
  scale_color_manual(values = class_colors) +
  labs(x = "Credit card balance ($)",
       y = expression(hat(pi)[k] * hat(f)[k](x)),
       color = "Default", linetype = "Default",
       title = "Prior-weighted densities; dotted line at the LDA boundary")

delta_default_plot <- ggplot(gauss_curves,
                             aes(x = balance, y = delta,
                                 color = default, linetype = default)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = x_star_lda, linetype = "dotted") +
  scale_color_manual(values = class_colors) +
  labs(x = "Credit card balance ($)", y = expression(hat(delta)[k](x)),
       color = "Default", linetype = "Default",
       title = "Estimated discriminant functions")

posterior_curves <- balance_grid |>
  mutate(LDA = plogis(hand_1$beta0 + hand_1$beta[1] * balance),
         `Logistic regression` = predict(balance_glm,
                                         newdata = balance_grid,
                                         type = "response")) |>
  pivot_longer(-balance, names_to = "model", values_to = "p_hat")

posterior_default_plot <- ggplot(posterior_curves,
                                 aes(x = balance, y = p_hat)) +
  geom_point(data = Default, aes(y = default_jit), alpha = 0.08,
             size = 0.7) +
  geom_hline(yintercept = 0.5, color = "grey60") +
  geom_line(aes(color = model, linetype = model), linewidth = 0.9) +
  scale_color_manual(values = model_colors) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Model", linetype = "Model",
       title = "Posterior probability of default")
```
:::

::: {.panel-tabset}
#### Gaussian densities
```{r balance-gauss-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
gauss_plot
```

#### Prior-weighted densities
```{r balance-weighted-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
weighted_default_plot
```

#### Discriminant functions
```{r balance-delta-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
delta_default_plot
```

#### Posterior
```{r balance-posterior-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
posterior_default_plot
```
:::

The fitted Gaussian for non-defaulters misses the spike at zero, and the one
for defaulters, forced to share $\hat\sigma$ with the larger class, is wider
than their histogram: the two departures the class summaries flagged. Weighting by the priors shrinks the defaulter curve to a
small bump; it rises above the non-defaulter curve only past the dotted line,
deep in the right tail where both are small. The discriminant functions are
two straight lines crossing at that same balance. On the probability scale the
LDA curve is a logistic curve, slightly flatter than the logistic regression
fit and crossing $0.5$ a little further right, as the coefficient table
implied.

::: {.callout-note collapse="true"}
### Beyond this course
The "coefficients of linear discriminants" printed by `MASS::lda()` come from
R. A. Fisher's original formulation of discriminant analysis, which asks for
the linear combination of features that maximizes the ratio of
between-class to within-class variance and makes no Gaussian assumption.
With $K$ classes that problem yields up to $K - 1$ discriminant directions,
printed as `LD1`, `LD2`, ..., and normalized to have unit within-class
variance. With Gaussian classes and a shared covariance matrix, Fisher's
directions and the Bayes'-theorem discriminants of this lecture produce the
same classifier.
:::

### Balance and student

Adding `student` to the formula gives the $p = 2$ model:

```{r additive-lda}
additive_lda <- MASS::lda(default ~ balance + student, data = Default)
additive_lda
```

::: {.callout-note collapse="true"}
### Code
```{r additive-lda-extract}
#| output: false
X2 <- model.matrix(~ balance + student, data = Default)[, -1]
hand_2 <- lda_by_hand(X2, Default$default)

scaling_ratio <- additive_lda$scaling[, 1] / hand_2$beta
cor_within <- hand_2$Sigma[1, 2] / sqrt(hand_2$Sigma[1, 1] * hand_2$Sigma[2, 2])

additive_glm <- glm(default ~ balance + student, data = Default,
                    family = "binomial")
b_glm_2 <- coef(additive_glm)

post_lda_2 <- predict(additive_lda)$posterior[, "Yes"]
post_glm_2 <- fitted(additive_glm)
max_diff_2 <- max(abs(post_lda_2 -
                      plogis(hand_2$beta0 + drop(X2 %*% hand_2$beta))))

coef_table_2 <- tibble(
  term = c("Intercept $\\hat\\beta_0$", "`balance` $\\hat\\beta_1$",
           "`studentYes` $\\hat\\beta_2$"),
  lda = c(hand_2$beta0, hand_2$beta),
  glm = unname(b_glm_2)
) |>
  mutate(across(c(lda, glm), \(v) sprintf("%.6f", v))) |>
  knitr::kable(col.names = c("Coefficient", "LDA", "Logistic regression"),
               align = "lcc")

# Decision boundary in the (balance, D) plane:
# beta0 + beta1 * balance + beta2 * D = 0.
threshold_at <- function(b0, b1, b2, d) -(b0 + b2 * d) / b1
x_star_lda_2 <- sapply(0:1, \(d) threshold_at(hand_2$beta0, hand_2$beta[1],
                                              hand_2$beta[2], d))
x_star_glm_2 <- sapply(0:1, \(d) threshold_at(b_glm_2[1], b_glm_2[2],
                                              b_glm_2[3], d))

boundary_lines <- tibble(
  model = c("LDA", "Logistic regression"),
  intercept = c(-hand_2$beta0 / hand_2$beta[2], -b_glm_2[1] / b_glm_2[3]),
  slope = c(-hand_2$beta[1] / hand_2$beta[2], -b_glm_2[2] / b_glm_2[3])
)

boundary_plot <- ggplot(Default, aes(x = balance, y = student_jit)) +
  geom_point(aes(color = default, shape = default), alpha = 0.25,
             size = 0.9) +
  geom_abline(data = boundary_lines,
              aes(intercept = intercept, slope = slope, linetype = model),
              linewidth = 0.9) +
  scale_color_manual(values = class_colors) +
  scale_shape_manual(values = c(16, 17)) +
  scale_y_continuous(breaks = c(0, 1), labels = c("No (0)", "Yes (1)")) +
  coord_cartesian(ylim = c(-0.35, 1.35)) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  labs(x = "Credit card balance ($)", y = "Student indicator D",
       color = "Default", shape = "Default", linetype = "Boundary",
       title = "Decision boundaries in the (balance, D) plane")

posterior_grid_2 <- expand_grid(
  balance = seq(0, 2800, length.out = 400),
  student = factor(c("No", "Yes"), levels = levels(Default$student))
)
posterior_curves_2 <- posterior_grid_2 |>
  mutate(LDA = predict(additive_lda, newdata = posterior_grid_2)$posterior[, "Yes"],
         `Logistic regression` = predict(additive_glm,
                                         newdata = posterior_grid_2,
                                         type = "response")) |>
  pivot_longer(c(LDA, `Logistic regression`), names_to = "model",
               values_to = "p_hat")

posterior_plot_2 <- ggplot(posterior_curves_2, aes(x = balance, y = p_hat)) +
  geom_point(data = Default, aes(y = default_jit), alpha = 0.08, size = 0.7) +
  geom_hline(yintercept = 0.5, color = "grey60") +
  geom_line(aes(color = model, linetype = model), linewidth = 0.9) +
  facet_wrap(~ student, labeller = label_both) +
  scale_color_manual(values = model_colors) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Model", linetype = "Model",
       title = "Posterior probability of default by student status")
```
:::

The prior probabilities are unchanged, since they do not depend on the
features. The group means now form a vector for each class: the `balance`
column repeats the one-feature means, and the `studentYes` column is the
class mean of the indicator $D$, which is the proportion of students in that
class. The coefficients of the linear discriminant are proportional to
$\hat\beta = \hat\Sigma^{-1}\left(\hat\mu_{\text{Yes}} - \hat\mu_{\text{No}}\right)$: dividing
each printed coefficient by the corresponding entry of $\hat\beta$ computed
from the formulas gives `r sprintf("%.4f", scaling_ratio[1])` for both.

Substituting the estimates into $\beta_0$ and $\beta$, beside the additive
logistic regression of the multiple logistic regression lecture:

```{r coef-table-2}
#| echo: false
coef_table_2
```

The intercept $\hat\beta_0 = `r sprintf("%.3f", hand_2$beta0)`$ is the log
posterior odds of default for a non-student with no balance. The `balance`
slope plays the same role as before. The `studentYes` slope,
$`r sprintf("%.3f", hand_2$beta[2])`$, is negative: at a fixed balance, LDA
assigns a student a lower posterior probability of default than a
non-student. That is the opposite direction from the raw class means, where
students made up a larger share of defaulters. Multiplying by
$\hat\Sigma^{-1}$ accounts for the within-class correlation of
`r sprintf("%.3f", cor_within)` between balance and $D$ — students carry higher
balances — in the same way that holding balance fixed reversed the sign of the
student coefficient in logistic regression. Applying the logistic function to
$\hat\beta_0 + x^\top\hat\beta$ again reproduces `predict()`'s posterior
probabilities, to within $`r sci(max_diff_2, 1)`$.

$D$ takes only the values $0$ and $1$, so the class-conditional distribution
of $(\texttt{balance}, D)$ cannot be bivariate Gaussian. `MASS::lda()` fits
the model regardless, because the estimates are sample means and covariances,
which can be computed for any numeric column. Whether the resulting classifier is
useful is a question about its classifications, which the next subsection
answers.

::: {.panel-tabset}
#### Decision boundary
```{r additive-boundary-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.6
boundary_plot
```

#### Posterior
```{r additive-posterior-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
posterior_plot_2
```
:::

The boundary is a straight line in the $(\texttt{balance}, D)$ plane, though
customers occupy only its two horizontal strips. Where the LDA line crosses
each strip gives a balance threshold for each group:
\$`r format(round(x_star_lda_2[1]), big.mark = ",")` for non-students and
\$`r format(round(x_star_lda_2[2]), big.mark = ",")` for students. The
logistic regression boundary gives
\$`r format(round(x_star_glm_2[1]), big.mark = ",")` and
\$`r format(round(x_star_glm_2[2]), big.mark = ",")`. In both models students
need a higher balance than non-students before being classified as
defaulters. The posterior tab shows the same comparison on the probability
scale: in each panel the LDA curve is flatter than the logistic regression
curve and crosses $0.5$ further right.

### Confusion matrices

Classifying each customer as a defaulter when $\hat p(x) > 0.5$, the LDA
classifier with `balance` and `student` produces this confusion matrix:

::: {.callout-note collapse="true"}
### Code
```{r confusion-lda}
#| output: false
confusion_counts <- function(p_hat, t) {
  truth <- Default$default == "Yes"
  pred  <- p_hat > t
  c(TP = sum(truth & pred), FN = sum(truth & !pred),
    FP = sum(!truth & pred), TN = sum(!truth & !pred))
}

cm_lda_05 <- confusion_counts(post_lda_2, 0.5)

confusion_lda_table <- tibble(
  Truth = c("Yes", "No"),
  `Predicted Yes` = c(cm_lda_05["TP"], cm_lda_05["FP"]),
  `Predicted No`  = c(cm_lda_05["FN"], cm_lda_05["TN"])
) |>
  mutate(Total = `Predicted Yes` + `Predicted No`) |>
  mutate(across(-Truth, \(v) format(v, big.mark = ","))) |>
  knitr::kable(align = "lccc")

metrics <- function(cm) {
  c(error = unname((cm["FP"] + cm["FN"]) / sum(cm)),
    sensitivity = unname(cm["TP"] / (cm["TP"] + cm["FN"])),
    specificity = unname(cm["TN"] / (cm["TN"] + cm["FP"])))
}
m_lda_05 <- metrics(cm_lda_05)
```
:::

```{r confusion-lda-table}
#| echo: false
confusion_lda_table
```

The classifier misclassifies
`r cm_lda_05["FP"] + cm_lda_05["FN"]` of
`r format(sum(cm_lda_05), big.mark = ",")` customers, an error rate of
`r sprintf("%.2f", 100 * m_lda_05["error"])`%, against the
`r sprintf("%.2f", 100 * hand_1$pi[2])`% of the classifier that predicts `No`
for everyone. The rows tell a different story. Of the
`r cm_lda_05["TP"] + cm_lda_05["FN"]` customers who defaulted, it flags only
`r cm_lda_05["TP"]`, a sensitivity of
`r sprintf("%.1f", 100 * m_lda_05["sensitivity"])`%. Of the
`r format(cm_lda_05["TN"] + cm_lda_05["FP"], big.mark = ",")` who did not, it
correctly leaves alone `r format(cm_lda_05["TN"], big.mark = ",")`, a
specificity of `r sprintf("%.1f", 100 * m_lda_05["specificity"])`%. The small
error rate comes almost entirely from the large negative class.

These are training error rates: the customers being classified are the ones
the model was fit to, so they are optimistic estimates of performance on new
customers, as the classification lecture discussed. Estimating test error from
the training data alone is the subject of the resampling unit.

### Threshold choice

A credit card company that wants to catch more of its eventual defaulters can
lower the threshold. The figure below tracks the overall error rate, the false
negative rate ($1 - \text{sensitivity}$), and the false positive rate
($1 - \text{specificity}$) of the LDA classifier as $t$ moves from $0$ to
$0.5$.

::: {.callout-note collapse="true"}
### Code
```{r threshold-sweep}
#| output: false
t_low <- 0.2

threshold_rates <- tibble(t = seq(0, 0.5, by = 0.005)) |>
  mutate(m = map(t, \(tt) metrics(confusion_counts(post_lda_2, tt))),
         `Overall error rate` = map_dbl(m, "error"),
         `False negative rate` = map_dbl(m, \(v) 1 - v[["sensitivity"]]),
         `False positive rate` = map_dbl(m, \(v) 1 - v[["specificity"]])) |>
  select(-m) |>
  pivot_longer(-t, names_to = "rate", values_to = "value") |>
  mutate(rate = factor(rate, levels = c("Overall error rate",
                                        "False negative rate",
                                        "False positive rate")))

threshold_plot <- ggplot(threshold_rates,
                         aes(x = t, y = value, color = rate, linetype = rate)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = c(t_low, 0.5), color = "grey60",
             linetype = "dotted") +
  scale_color_manual(values = c("black", "#D55E00", "#0072B2")) +
  labs(x = "Threshold t", y = "Rate", color = NULL, linetype = NULL,
       title = "LDA (balance + student) error rates by threshold")

# Training error rate at every distinct posterior probability used as t;
# the error rate only changes when t passes an observation's score.
min_error_threshold <- function(p_hat) {
  cand <- sort(unique(p_hat))
  err  <- map_dbl(cand, \(tt) metrics(confusion_counts(p_hat, tt))[["error"]])
  c(t = cand[which.min(err)], error = min(err))
}
min_err_lda <- min_error_threshold(post_lda_2)
min_err_glm <- min_error_threshold(post_glm_2)

cm_lda_low <- confusion_counts(post_lda_2, t_low)
cm_glm_05  <- confusion_counts(post_glm_2, 0.5)
cm_glm_low <- confusion_counts(post_glm_2, t_low)
m_lda_low  <- metrics(cm_lda_low)
m_glm_05   <- metrics(cm_glm_05)
m_glm_low  <- metrics(cm_glm_low)

compare_table <- tibble(
  Model = c("LDA", "LDA", "Logistic regression", "Logistic regression"),
  t = c(0.5, t_low, 0.5, t_low),
  cm = list(cm_lda_05, cm_lda_low, cm_glm_05, cm_glm_low)
) |>
  mutate(TP = map_dbl(cm, "TP"), FN = map_dbl(cm, "FN"),
         FP = map_dbl(cm, "FP"), TN = map_dbl(cm, "TN"),
         m = map(cm, metrics),
         error = map_dbl(m, "error"),
         sensitivity = map_dbl(m, "sensitivity"),
         specificity = map_dbl(m, "specificity")) |>
  select(-cm, -m) |>
  mutate(across(c(TP, FN, FP, TN), \(v) format(v, big.mark = ",")),
         across(c(error, sensitivity, specificity),
                \(v) sprintf("%.4f", v))) |>
  knitr::kable(col.names = c("Model", "Threshold", "TP", "FN", "FP", "TN",
                             "Error rate", "Sensitivity", "Specificity"),
               align = "lcccccccc")
```
:::

```{r threshold-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
threshold_plot
```

As $t$ falls from $0.5$, the false negative rate falls steadily while the
false positive rate stays near zero until $t$ is small. The overall error rate,
dominated by the `r format(cm_lda_05["TN"] + cm_lda_05["FP"], big.mark = ",")`
non-defaulters, follows the false positive rate: nearly flat over most of the
range, then climbing sharply as $t$ approaches $0$ and almost every customer is
flagged.

The overall error rate is not smallest at $t = 0.5$. LDA's training error rate
is lowest near $t = `r sprintf("%.2f", min_err_lda[["t"]])`$, at
`r sprintf("%.2f", 100 * min_err_lda[["error"]])`%, against
`r sprintf("%.2f", 100 * m_lda_05["error"])`% at $t = 0.5$. The additive
logistic regression's is lowest near
$t = `r sprintf("%.2f", min_err_glm[["t"]])`$, at
`r sprintf("%.2f", 100 * min_err_glm[["error"]])`%, against
`r sprintf("%.2f", 100 * m_glm_05["error"])`% at $t = 0.5$. Both
$\hat p(x)$ are estimates, so the Bayes classifier's threshold carries no
guarantee for them.

The confusion-matrix counts at $t = 0.5$ and $t = `r t_low`$, for LDA and for
the additive logistic regression:

```{r compare-table}
#| echo: false
compare_table
```

Lowering the threshold to $t = `r t_low`$ raises LDA's sensitivity from
`r sprintf("%.1f", 100 * m_lda_05["sensitivity"])`% to
`r sprintf("%.1f", 100 * m_lda_low["sensitivity"])`%, at the cost of
specificity falling from `r sprintf("%.1f", 100 * m_lda_05["specificity"])`% to
`r sprintf("%.1f", 100 * m_lda_low["specificity"])`% and the overall error rate
rising from `r sprintf("%.2f", 100 * m_lda_05["error"])`% to
`r sprintf("%.2f", 100 * m_lda_low["error"])`%. At both thresholds logistic
regression flags more customers as defaulters than LDA does —
`r cm_glm_05["TP"] + cm_glm_05["FP"]` against
`r cm_lda_05["TP"] + cm_lda_05["FP"]` at $t = 0.5$ — and so has the higher
sensitivity and the lower specificity of the two.

### ROC curves and AUC

A comparison at one or two thresholds depends on which thresholds were
chosen. The ROC curves compare the two classifiers at every threshold.

::: {.callout-note collapse="true"}
### Code
```{r roc-default}
#| output: false
# Every distinct score is a cutoff; classify positive when score >= cutoff.
roc_curve <- function(score, positive) {
  ord <- order(score, decreasing = TRUE)
  s   <- score[ord]
  pos <- positive[ord]
  keep <- c(diff(s) != 0, TRUE)
  tibble(fpr = c(0, cumsum(!pos)[keep] / sum(!positive)),
         tpr = c(0, cumsum(pos)[keep] / sum(positive)))
}

auc_trapezoid <- function(roc) {
  with(roc, sum(diff(fpr) * (head(tpr, -1) + tail(tpr, -1)) / 2))
}

auc_pairs <- function(score, positive) {
  s_pos <- score[positive]
  s_neg <- score[!positive]
  mean(outer(s_pos, s_neg, ">")) + 0.5 * mean(outer(s_pos, s_neg, "=="))
}

is_yes <- Default$default == "Yes"
scores <- list(
  "LDA, balance" = post_1,
  "Logistic regression, balance" = fitted(balance_glm),
  "LDA, balance + student" = post_lda_2,
  "Logistic regression, balance + student" = post_glm_2
)

auc_values <- tibble(model = names(scores)) |>
  mutate(trapezoid = map_dbl(scores, \(s) auc_trapezoid(roc_curve(s, is_yes))),
         pairs = map_dbl(scores, \(s) auc_pairs(s, is_yes)))

auc_of <- function(m) auc_values$trapezoid[auc_values$model == m]

auc_table <- auc_values |>
  mutate(across(c(trapezoid, pairs), \(v) sprintf("%.5f", v))) |>
  knitr::kable(col.names = c("Model", "AUC (trapezoid rule)",
                             "AUC (pairwise comparison)"),
               align = "lcc")

roc_2 <- bind_rows(
  roc_curve(post_lda_2, is_yes) |> mutate(model = "LDA"),
  roc_curve(post_glm_2, is_yes) |> mutate(model = "Logistic regression")
)

threshold_labels <- c("t = 0.5", paste("t =", t_low))
roc_points <- tibble(
  model = rep(c("LDA", "Logistic regression"), each = 2),
  t = factor(rep(threshold_labels, 2), levels = threshold_labels),
  m = list(m_lda_05, m_lda_low, m_glm_05, m_glm_low)
) |>
  mutate(fpr = 1 - map_dbl(m, "specificity"),
         tpr = map_dbl(m, "sensitivity"))

roc_plot <- ggplot(roc_2, aes(x = fpr, y = tpr)) +
  geom_abline(intercept = 0, slope = 1, color = "grey60",
              linetype = "dashed") +
  geom_line(aes(color = model, linetype = model), linewidth = 0.9) +
  geom_point(data = roc_points, aes(shape = t, color = model), size = 3,
             stroke = 1.2) +
  scale_color_manual(values = model_colors) +
  scale_shape_manual(values = c(24, 21)) +
  coord_fixed() +
  labs(x = "False positive rate (1 - specificity)",
       y = "True positive rate (sensitivity)",
       color = "Model", linetype = "Model", shape = "Threshold",
       title = "ROC curves, balance + student")

# Balance at which the additive models' posterior curves cross, by student
# status, and the posterior probability there.
posterior_both <- function(b, s) {
  nd <- tibble(balance = b,
               student = factor(s, levels = levels(Default$student)))
  c(lda = predict(additive_lda, newdata = nd)$posterior[, "Yes"],
    glm = unname(predict(additive_glm, newdata = nd, type = "response")))
}
crossover <- tibble(student = c("No", "Yes")) |>
  mutate(balance = map_dbl(student, \(s) {
           uniroot(\(b) diff(posterior_both(b, s)), c(800, 2500))$root
         }),
         p = map2_dbl(balance, student, \(b, s) posterior_both(b, s)[["lda"]]))
cross_of <- function(s, col) crossover[[col]][crossover$student == s]

# A threshold below both crossing probabilities.
t_below <- 0.1
n_flag_lda_below <- sum(post_lda_2 > t_below)
n_flag_glm_below <- sum(post_glm_2 > t_below)
```
:::

```{r roc-figure}
#| echo: false
#| fig-width: 6.5
#| fig-height: 5.5
roc_plot
```

The two curves lie almost on top of one another and rise steeply near the
origin, well above the diagonal. The marked points are the four classifiers of
the previous table. Lowering $t$ from $0.5$ to `r t_low` moves each model up
and slightly right along its curve. At each of these two thresholds, logistic
regression's point sits further along its nearly shared curve than LDA's.

The two models' posterior curves cross at a balance of
\$`r format(round(cross_of("No", "balance")), big.mark = ",")` for
non-students and \$`r format(round(cross_of("Yes", "balance")), big.mark = ",")`
for students, where both models give posterior probabilities of
`r sprintf("%.3f", cross_of("No", "p"))` and
`r sprintf("%.3f", cross_of("Yes", "p"))`, respectively. Above those balances LDA's flatter
curve lies below logistic regression's, so a threshold above both crossing
probabilities is reached by LDA only at a higher balance, and LDA flags fewer
customers; both marked thresholds are of this kind. Below both crossing
probabilities the order reverses: at $t = `r t_below`$, LDA flags
`r format(n_flag_lda_below, big.mark = ",")` customers and logistic regression
`r format(n_flag_glm_below, big.mark = ",")`. The differences between the two models'
confusion matrices come from where each model's probabilities place a given
threshold on the curve, not from a difference in how well the two models order
the customers.

The AUC of each classifier, computed both as the area under the empirical ROC
curve and as the proportion of (defaulter, non-defaulter) pairs ordered
correctly:

```{r auc-table}
#| echo: false
auc_table
```

The two computations agree in every row. With `balance` alone, LDA and
logistic regression have identical AUCs of
`r sprintf("%.5f", auc_of("LDA, balance"))`: both posterior probabilities are
increasing functions of balance, so they order the customers identically and
trace the same ROC curve. With `balance` and `student`, the two classifiers'
posterior probabilities weight $D$ differently relative to balance, so their
orderings differ slightly. Their AUCs,
`r sprintf("%.5f", auc_of("LDA, balance + student"))` and
`r sprintf("%.5f", auc_of("Logistic regression, balance + student"))`, differ
by $`r sci(auc_of("LDA, balance + student") - auc_of("Logistic regression, balance + student"), 2)`$.
An AUC of
`r sprintf("%.3f", auc_of("LDA, balance + student"))` means that for a randomly
chosen defaulter and non-defaulter, the LDA classifier gives the defaulter
the higher posterior probability about
`r round(100 * auc_of("LDA, balance + student"))`% of the time.

**Conclusion.** LDA's Gaussian assumptions describe these data poorly — the
student indicator is not Gaussian and the two classes' balance variances
differ — yet LDA with `balance` and `student` orders the training customers
almost exactly as the additive logistic regression does. Its decision
boundary likewise assigns students a higher balance threshold. At $t = 0.5$
and $t = `r t_low`$ it flags fewer customers than logistic regression, because
its flatter posterior curve lies below logistic regression's over the balances
where those thresholds are crossed. At $t = 0.5$ it
catches `r sprintf("%.1f", 100 * m_lda_05["sensitivity"])`% of the
defaulters, and lowering the threshold to `r t_low` raises that to
`r sprintf("%.1f", 100 * m_lda_low["sensitivity"])`% at a cost of
`r sprintf("%.1f", 100 * (m_lda_05["specificity"] - m_lda_low["specificity"]))`
percentage points of specificity.

## Conclusion

A generative classifier models the distribution of the features within each
class, $f_k(x)$, and the prior probability of each class, $\pi_k$, and
combines them through Bayes' theorem into the posterior probability
$p_k(x) = \pi_k f_k(x) / \sum_l \pi_l f_l(x)$. Assigning each observation to
the class with the largest posterior probability is the Bayes classifier,
which minimizes the expected error rate; a generative classifier approximates
it by estimating $\pi_k$ and $f_k$ from training data.

Linear discriminant analysis assumes Gaussian class-conditional densities with
class-specific means and a covariance matrix shared by every class. Because
the covariance is shared, the quadratic term in $x$ is the same for every
class and drops out, leaving discriminant functions $\delta_k(x)$ that are
linear in $x$ and a decision boundary that is a point, a line, or a
hyperplane. With two classes the log posterior odds are linear in the
features — the same form as logistic regression — but the coefficients come
from closed-form estimates (class proportions, class means, and a pooled
variance or covariance matrix) that maximize the joint likelihood of features
and responses, rather than from the iteratively maximized conditional
likelihood of logistic regression. A smaller prior probability moves the
boundary toward that class's mean.

A classifier's error rate hides which mistakes it makes. The confusion matrix
separates them, sensitivity and specificity measure performance within each
true class, and moving the threshold trades one against the other. The ROC
curve displays that tradeoff at every threshold, depends only on how a
classifier orders the observations, and its area, the AUC, is the probability
that a randomly chosen positive observation is scored above a randomly chosen
negative one.

The next lecture relaxes LDA's shared covariance assumption. Quadratic
discriminant analysis gives each class its own covariance matrix, so the
quadratic term no longer cancels and the boundary curves, and naive Bayes
replaces the multivariate Gaussian with a product of one-feature densities.
Together with LDA and logistic regression, these are the classifiers ISLR2
Chapter 4 compares; that lecture also adds KNN to the comparison.
