# Classifier Evaluation

[Slides for this lecture](55-classifier-evaluation_slides.html)

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
# The same seed as 50-lda.qmd, the lecture these sections were moved from. No
# result below depends on a random draw (predict() for a MASS fit breaks ties
# between classes at random, but only its posterior probabilities are used),
# so the seed only keeps renders reproducible.
set.seed(20261001)

# MASS is called as MASS::lda() rather than attached, because attaching it
# masks dplyr::select().
# Renders a number in LaTeX scientific notation for use inside $...$.
sci <- function(x, digits = 2) {
  e <- floor(log10(abs(x)))
  m <- signif(x / 10^e, digits)
  sprintf("%s \\times 10^{%d}", format(m), e)
}

class_colors <- c("#0072B2", "#D55E00")
model_colors <- c("LDA" = "#009E73", "Logistic regression" = "#CC79A7")
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **construct** a confusion matrix at a stated threshold (or a $K \times K$
confusion matrix for $K$ classes), **compute** the error rate, sensitivity,
and specificity from it, and **explain** how sensitivity
and specificity trade off as the threshold moves,
- **interpret** an ROC curve and its AUC, the area under the curve, for which
larger values (at most 1) generally indicate a better classifier, **compute**
the AUC from the points of an ROC curve by the trapezoid rule, and **explain**
why the ROC curve and the AUC depend on a classifier only through how it
orders the observations, and
- **compare** classifiers fit to the same observations, including models fit
separately within the strata of a categorical variable, by error rate,
sensitivity, specificity, and AUC, and **explain** why training rates
overstate performance on new observations.
:::


## Credit card classifiers

### Credit card default data

**Question.** How well do the LDA and logistic regression classifiers of the
[Linear Discriminant Analysis](50-lda.qmd) lecture identify the credit card
customers who will default, and which kinds of mistakes do they make?

**Data.** `ISLR2::Default` is a simulated data set of
`r format(nrow(Default), big.mark = ",")` credit card customers. The response
is `default`, with `Yes` the positive class. The features are `balance`, the
customer's monthly credit card balance in dollars, and `income`, the
customer's income in dollars; `student` (Yes/No) splits the customers in the
Student strata section.

The table gives each default class's number and proportion of customers and
the mean, standard deviation, and number of zero values of its balances. The
figure plots income against balance for every customer, with defaulters and
non-defaulters distinguished by color and shape.

::: {.callout-note collapse="true"}
### Code
```{r default-data}
#| output: false
dollars <- function(v) format(round(v), big.mark = ",")
# Axis labels such as $60,000, without the scales package.
dollar_labels <- function(v) {
  ifelse(is.na(v), NA,
         paste0("$", format(v, big.mark = ",", scientific = FALSE, trim = TRUE)))
}

class_summary <- Default |>
  group_by(default) |>
  summarize(n = n(), prop = n() / nrow(Default),
            mean_balance = mean(balance), sd_balance = sd(balance),
            n_zero = sum(balance == 0), .groups = "drop")
summary_of <- function(k, col) class_summary[[col]][class_summary$default == k]

class_summary_table <- class_summary |>
  transmute(default, n = format(n, big.mark = ","),
            prop = sprintf("%.4f", prop),
            mean_balance = dollars(mean_balance),
            sd_balance = dollars(sd_balance),
            n_zero = format(n_zero, big.mark = ",")) |>
  knitr::kable(col.names = c("Default", "Customers", "Proportion",
                             "Mean balance (\\$)", "SD of balance (\\$)",
                             "Zero balances"),
               align = "cccccc")

# Defaulters are drawn last so they are not hidden under non-defaulters.
scatter_base <- ggplot(arrange(Default, default),
                       aes(x = balance, y = income)) +
  geom_point(aes(color = default, shape = default), alpha = 0.35,
             size = 0.8) +
  scale_color_manual(values = class_colors) +
  scale_shape_manual(values = c(16, 17)) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  scale_x_continuous(labels = dollar_labels) +
  scale_y_continuous(labels = dollar_labels) +
  labs(x = "Credit card balance ($)", y = "Income ($)", color = "Default",
       shape = "Default")

eda_scatter <- scatter_base +
  labs(title = "ISLR2::Default, income against balance")
```
:::

```{r class-summary-table}
#| echo: false
class_summary_table
```

```{r default-scatter-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
eda_scatter
```

Only `r format(summary_of("Yes", "n"), big.mark = ",")` customers, a proportion
of `r sprintf("%.4f", summary_of("Yes", "prop"))`, defaulted. Defaulters sit
at larger balances, across nearly the whole range of income, and the two
classes' clouds overlap heavily.

### Fitted classifiers

The LDA lecture fit four classifiers to all of these customers: LDA and
logistic regression with `balance` alone, and LDA and logistic regression
with `balance` and `income`. Each gives every customer a posterior
probability of default, $\hat p(x)$, and this lecture evaluates those
probabilities as scores. The figure draws the two-feature classifiers'
decision boundaries, where $\hat p(x) = 0.5$, over the scatterplot, LDA's as
a solid green line and logistic regression's as a dashed pink one.

::: {.callout-note collapse="true"}
### Code
```{r fitted-classifiers}
#| output: false
# Estimates of pi_k, mu_k, Sigma and the implied log posterior odds
# beta0 + x' beta of the second class, as in the LDA lecture.
lda_by_hand <- function(X, y) {
  X <- as.matrix(X)
  classes <- levels(y)
  n <- nrow(X)
  K <- length(classes)
  pi_hat <- as.numeric(table(y)) / n
  mu_hat <- do.call(rbind, lapply(classes, function(k) {
    colMeans(X[y == k, , drop = FALSE])
  }))
  rownames(mu_hat) <- classes
  Sigma_hat <- Reduce(`+`, lapply(classes, function(k) {
    Z <- sweep(X[y == k, , drop = FALSE], 2, mu_hat[k, ])
    crossprod(Z)
  })) / (n - K)
  beta  <- solve(Sigma_hat, mu_hat[2, ] - mu_hat[1, ])
  beta0 <- log(pi_hat[2] / pi_hat[1]) -
    0.5 * (sum(mu_hat[2, ] * solve(Sigma_hat, mu_hat[2, ])) -
           sum(mu_hat[1, ] * solve(Sigma_hat, mu_hat[1, ])))
  list(pi = pi_hat, mu = mu_hat, Sigma = Sigma_hat,
       beta0 = beta0, beta = beta)
}

# Balance alone.
hand_1 <- lda_by_hand(Default["balance"], Default$default)
balance_lda <- MASS::lda(default ~ balance, data = Default)
balance_glm <- glm(default ~ balance, data = Default, family = "binomial")
post_1 <- predict(balance_lda)$posterior[, "Yes"]

# Balance and income.
additive_lda <- MASS::lda(default ~ balance + income, data = Default)
hand_2 <- lda_by_hand(Default[c("balance", "income")], Default$default)
post_lda_2 <- predict(additive_lda)$posterior[, "Yes"]
additive_glm <- glm(default ~ balance + income, data = Default,
                    family = "binomial")
b_glm_2 <- coef(additive_glm)
z_income_glm <- summary(additive_glm)$coefficients["income", "z value"]
post_glm_2 <- fitted(additive_glm)

income_levels <- quantile(Default$income, c(0.25, 0.5, 0.75))
# A boundary read as a balance threshold at a given income.
threshold_at <- function(b0, b1, b2, inc) -(b0 + b2 * inc) / b1

boundary_lines <- tibble(
  model = c("LDA", "Logistic regression"),
  intercept = c(-hand_2$beta0 / hand_2$beta[2], -b_glm_2[1] / b_glm_2[3]),
  slope = c(-hand_2$beta[1] / hand_2$beta[2], -b_glm_2[2] / b_glm_2[3])
)

# Boundary lines take the models' colors as fixed colors, because the color
# scale already encodes default; linetype carries the legend.
boundary_layers <- function(lines) {
  list(
    geom_abline(data = filter(lines, model == "LDA"),
                aes(intercept = intercept, slope = slope, linetype = model),
                color = model_colors[["LDA"]], linewidth = 1),
    geom_abline(data = filter(lines, model == "Logistic regression"),
                aes(intercept = intercept, slope = slope, linetype = model),
                color = model_colors[["Logistic regression"]], linewidth = 1),
    scale_linetype_manual(values = c("solid", "dashed")),
    guides(linetype = guide_legend(
      override.aes = list(color = unname(model_colors))))
  )
}

boundary_plot <- scatter_base +
  boundary_layers(boundary_lines) +
  coord_cartesian(xlim = range(Default$balance),
                  ylim = range(Default$income)) +
  labs(linetype = "Boundary",
       title = "Decision boundaries in the (balance, income) plane")
```
:::

```{r fitted-boundary-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
boundary_plot
```

The two boundaries are nearly vertical and close together, with logistic
regression's slightly to the left of LDA's. Customers to the right of a
boundary are classified as defaulters, and many defaulters lie to its left.

## Classifier evaluation

An error rate counts every mistake the same way. Failing to flag a customer
who goes on to default and flagging a customer who does not are different
mistakes with different costs, and a classifier can have a small error rate
while making almost every mistake of one kind. The classification lecture
introduced the confusion matrix, which separates the mistakes; this section
develops it in full for two classes and for $K$ classes, and adds the
ROC curve and AUC, which summarize a two-class classifier at every threshold.
The running example is the LDA classifier with `balance` and `income`.

The figure shows, for each true default class, the histogram of the
`balance` and `income` LDA classifier's log posterior odds of default, on the
density scale, with a dotted line at $0$, where the classifier's estimated
probability of default, $\hat p(x)$, equals $0.5$.

::: {.callout-note collapse="true"}
### Code
```{r score-hist}
#| output: false
# Log posterior odds of default from the balance + income LDA, binned once so
# this figure and the threshold slider below share the same bars.
lda_score <- qlogis(post_lda_2)
score_breaks <- seq(floor(min(lda_score)), ceiling(max(lda_score)), by = 0.5)
class_label <- c(No = "No (negative)", Yes = "Yes (positive)")
slider_bins <- map(c("No", "Yes"), \(k) {
  h <- hist(lda_score[Default$default == k], breaks = score_breaks,
            plot = FALSE)
  tibble(class = class_label[[k]], x0 = head(h$breaks, -1),
         x1 = tail(h$breaks, -1), density = h$density)
}) |>
  list_rbind()

score_by_class <- tibble(score = lda_score, default = Default$default) |>
  group_by(default) |>
  summarize(median = median(score), below_0 = mean(score <= 0),
            above_0 = mean(score > 0), .groups = "drop")
score_of <- function(k, col) score_by_class[[col]][score_by_class$default == k]

score_hist_plot <- tibble(score = lda_score, default = Default$default) |>
  ggplot(aes(x = score)) +
  geom_histogram(aes(y = after_stat(density)), breaks = score_breaks,
                 fill = "grey80", color = "grey55") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  facet_wrap(~ default, ncol = 1, labeller = label_both) +
  labs(x = "LDA log posterior odds of default", y = "Density",
       title = "LDA (balance + income) scores within each true class")
```
:::

```{r score-hist-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
score_hist_plot
```

The defaulters' scores sit well to the right of the non-defaulters' (medians
`r sprintf("%.2f", score_of("Yes", "median"))` and
`r sprintf("%.2f", score_of("No", "median"))`), but the two histograms
overlap: `r sprintf("%.1f", 100 * score_of("Yes", "below_0"))`% of defaulters
score at or below $0$, and `r sprintf("%.1f", 100 * score_of("No", "above_0"))`%
of non-defaulters above it. Any threshold inside the overlap makes two kinds
of mistake, defaulters to its left and non-defaulters to its right.

### Confusion matrix

With two classes, call the class of interest **positive** and the other
**negative**. For a classifier that produces an estimated probability
$\hat p(x)$ of the positive class, a **threshold** $t$ turns probabilities into
classes: predict positive when $\hat p(x) > t$.

As in the classification lecture, a confusion matrix cross-tabulates the true
class against the predicted class:

::: {.callout-note collapse="true"}
### Code
```{r confusion-template}
#| output: false
confusion_template <- tibble(
  Truth = c("Positive", "Negative"),
  `Predicted positive` = c("True positive (TP)", "False positive (FP)"),
  `Predicted negative` = c("False negative (FN)", "True negative (TN)")
) |>
  knitr::kable(align = "lcc")
```
:::

```{r confusion-template-table}
#| echo: false
confusion_template
```

where

- $TP$ is the number of positive observations predicted positive,
- $FN$ is the number of positive observations predicted negative,
- $FP$ is the number of negative observations predicted positive, and
- $TN$ is the number of negative observations predicted negative.

In the score histograms above, with $t = 0.5$, $TP$ and $FN$ count the defaulters to
the right and left of the dotted line, and $FP$ and $TN$ the non-defaulters
to its right and left. The diagonal holds the correct classifications, so the
error rate is $(FP + FN)/n$.

Classifying each credit card customer as a defaulter when $\hat p(x) > 0.5$,
the LDA classifier with `balance` and `income` produces this confusion matrix:

::: {.callout-note collapse="true"}
### Code
```{r confusion-lda}
#| output: false
confusion_counts <- function(p_hat, t) {
  truth <- Default$default == "Yes"
  pred  <- p_hat > t
  c(TP = sum(truth & pred), FN = sum(truth & !pred),
    FP = sum(!truth & pred), TN = sum(!truth & !pred))
}

cm_lda_05 <- confusion_counts(post_lda_2, 0.5)
# A lower threshold, compared with 0.5 below.
t_low <- 0.2

confusion_lda_table <- tibble(
  Truth = c("Yes", "No"),
  `Predicted Yes` = c(cm_lda_05["TP"], cm_lda_05["FP"]),
  `Predicted No`  = c(cm_lda_05["FN"], cm_lda_05["TN"])
) |>
  mutate(Total = `Predicted Yes` + `Predicted No`) |>
  mutate(across(-Truth, \(v) format(v, big.mark = ","))) |>
  knitr::kable(align = "lccc")

metrics <- function(cm) {
  c(error = unname((cm["FP"] + cm["FN"]) / sum(cm)),
    sensitivity = unname(cm["TP"] / (cm["TP"] + cm["FN"])),
    specificity = unname(cm["TN"] / (cm["TN"] + cm["FP"])))
}
m_lda_05 <- metrics(cm_lda_05)
```
:::

```{r confusion-lda-table}
#| echo: false
confusion_lda_table
```

The classifier misclassifies
`r cm_lda_05["FP"] + cm_lda_05["FN"]` of
`r format(sum(cm_lda_05), big.mark = ",")` customers, an error rate of
`r sprintf("%.2f", 100 * m_lda_05["error"])`%, against the
`r sprintf("%.2f", 100 * hand_1$pi[2])`% of the classifier that predicts `No`
for everyone. The error rate is small because the
`r format(cm_lda_05["TN"] + cm_lda_05["FP"], big.mark = ",")` non-defaulters,
nearly all classified correctly, dominate it, even though
`r cm_lda_05["FN"]` of the `r cm_lda_05["FP"] + cm_lda_05["FN"]` errors are
missed defaulters.

These are training error rates: the customers being classified are the ones
the model was fit to, so they are optimistic estimates of performance on new
customers, as the classification lecture discussed. Estimating test error from
the training data alone is the subject of the resampling unit.

### Sensitivity and specificity

Dividing each row by its total gives two rates that condition on the true
class:

$$\text{Sensitivity} = \frac{TP}{TP + FN}, \qquad \text{Specificity} = \frac{TN}{TN + FP},$$

where

- **sensitivity**, also called **recall** or the **true positive rate** (TPR), is the
proportion of positive observations the classifier catches, an estimate of
$P\left(\hat y(X) = \text{positive} \mid Y = \text{positive}\right)$, and
- **specificity**, also called the **true negative rate**, is the proportion of
negative observations the classifier correctly leaves alone, an estimate of
$P\left(\hat y(X) = \text{negative} \mid Y = \text{negative}\right)$.

Their complements are error rates within each true class:
$1 - \text{sensitivity} = FN/(TP + FN)$ is the **false negative rate**, and
$1 - \text{specificity} = FP/(TN + FP)$ is the **false positive rate** (FPR).

In the credit card confusion matrix, of the
`r cm_lda_05["TP"] + cm_lda_05["FN"]` customers who defaulted the classifier
flags only `r cm_lda_05["TP"]`, a sensitivity of
`r sprintf("%.1f", 100 * m_lda_05["sensitivity"])`%. Of the
`r format(cm_lda_05["TN"] + cm_lda_05["FP"], big.mark = ",")` who did not, it
correctly leaves alone `r format(cm_lda_05["TN"], big.mark = ",")`, a
specificity of `r sprintf("%.1f", 100 * m_lda_05["specificity"])`%.

### Multiple classes

With $K$ classes the confusion matrix is $K \times K$, with true classes in
the rows and predicted classes in the columns:

::: {.callout-note collapse="true"}
### Code
```{r confusion-template-k}
#| output: false
k_labels <- c("1", "2", "$\\vdots$", "$K$")
cell <- function(k, j) {
  ifelse(k == "$\\vdots$" | j == "$\\vdots$", "$\\vdots$",
         sprintf("$n_{%s%s}$", gsub("\\$", "", k), gsub("\\$", "", j)))
}
confusion_template_k <- tibble(Truth = k_labels) |>
  mutate(`Predicted 1` = cell(Truth, "1"),
         `Predicted 2` = cell(Truth, "2"),
         `$\\cdots$`   = ifelse(Truth == "$\\vdots$", "$\\ddots$", "$\\cdots$"),
         `Predicted $K$` = cell(Truth, "$K$")) |>
  knitr::kable(align = "lcccc", escape = FALSE)
```
:::

```{r confusion-template-k-table}
#| echo: false
confusion_template_k
```

where

- $n_{kj}$ is the number of observations in true class $k$ predicted to be in
class $j$, so the diagonal entries $n_{kk}$ are the correct classifications.

The error rate is $1 - \sum_{k=1}^K n_{kk} / n$. Sensitivity and specificity
are defined one class at a time, **one versus the rest**: treat class $k$ as
positive and the other $K - 1$ classes together as negative, which collapses
the table to two-by-two. Class $k$'s rates are

$$\text{Sensitivity}_k = \frac{n_{kk}}{\sum_{j=1}^K n_{kj}}, \qquad \text{Specificity}_k = \frac{\sum_{l \ne k} \sum_{j \ne k} n_{lj}}{\sum_{l \ne k} \sum_{j=1}^K n_{lj}},$$

where

- the sensitivity's denominator is the number of observations in class $k$,
and its numerator the number of those predicted to be in class $k$, and
- the specificity's denominator is the number of observations in the other
classes, and its numerator the number of those predicted to be in some class
other than $k$, whether or not it is their own.

With $K$ classes the threshold $t = 0.5$ generalizes to assigning each
observation to the class with the largest estimated posterior probability;
with $K = 2$ the two rules coincide. Other thresholds, and the ROC curve and
AUC built from them, are two-class tools, and the rest of this section
returns to two classes.

::: {.callout-note collapse="true"}
### Beyond this course
ROC curves and AUC extend to $K > 2$ classes by averaging the AUCs of the $K$
one-versus-rest problems, or of all $K(K-1)/2$ pairs of classes. Logistic
regression as developed in this course is also a two-class method; its
multi-class extension, multinomial logistic regression, is not covered in
this course.
:::

### Classification threshold {#threshold-choice}

The threshold $t = 0.5$ is the Bayes classifier's: if $\hat p(x)$ equalled
the true $p_2(x)$, the posterior probability of the positive class, it would
minimize the expected error rate. An estimated
$\hat p(x)$ carries no such guarantee, and even for the Bayes classifier the
overall error rate need not be the quantity of interest.

Lowering $t$ enlarges the set $\left\{x : \hat p(x) > t\right\}$ of feature
values classified positive, so every
observation predicted positive at the old threshold is still predicted
positive at the new one. Consequently $TP$ and $FP$ can only grow while $FN$
and $TN$ can only shrink: sensitivity rises or stays the same, and
specificity falls or stays the same. Raising $t$ does the reverse, so no
threshold improves both.

::: {.callout-note collapse="true"}
### Beyond this course
If the two mistakes have known costs, $c_{FN}$ for a missed positive and
$c_{FP}$ for a false alarm, the expected cost of predicting positive at $x$ is
$c_{FP}\left[1 - p_2(x)\right]$ and of predicting negative is $c_{FN}\, p_2(x)$.
Predicting positive is the cheaper choice exactly when
$p_2(x) > c_{FP}/(c_{FP} + c_{FN})$, so that ratio is the cost-minimizing
threshold. Equal costs recover $t = 0.5$.
:::

A credit card company that wants to catch more of its eventual defaulters can
lower the threshold. The figure below tracks four rates of the LDA classifier
with `balance` and `income` as $t$ moves from $0$ to $0.5$ — the overall
error rate, sensitivity (the true positive rate), specificity (the true
negative rate), and the false positive rate ($1 - \text{specificity}$) — with
dotted lines at $t = `r t_low`$ and $t = 0.5$. The checkboxes choose which
rates are drawn; the false positive rate starts unchecked because it mirrors
specificity. The table below the figure gives the confusion-matrix counts and
rates at $t = 0.5$ and $t = `r t_low`$ for LDA with `balance` alone, LDA with
`balance` and `income`, and the logistic regression with `balance` and
`income`.

::: {.callout-note collapse="true"}
### Code
```{r threshold-sweep}
#| output: false
threshold_rates <- tibble(t = seq(0, 0.5, by = 0.005)) |>
  mutate(m = map(t, \(tt) metrics(confusion_counts(post_lda_2, tt))),
         `Overall error rate` = map_dbl(m, "error"),
         `Sensitivity (true positive rate)` = map_dbl(m, "sensitivity"),
         `Specificity (true negative rate)` = map_dbl(m, "specificity"),
         `False positive rate (1 - specificity)` =
           map_dbl(m, \(v) 1 - v[["specificity"]])) |>
  select(-m) |>
  pivot_longer(-t, names_to = "rate", values_to = "value")

# Training error rate at every distinct posterior probability used as t;
# the error rate only changes when t passes an observation's score.
min_error_threshold <- function(p_hat) {
  cand <- sort(unique(p_hat))
  err  <- map_dbl(cand, \(tt) metrics(confusion_counts(p_hat, tt))[["error"]])
  c(t = cand[which.min(err)], error = min(err))
}
min_err_lda <- min_error_threshold(post_lda_2)
min_err_glm <- min_error_threshold(post_glm_2)

cm_lda1_05  <- confusion_counts(post_1, 0.5)
cm_lda1_low <- confusion_counts(post_1, t_low)
cm_lda_low  <- confusion_counts(post_lda_2, t_low)
cm_glm_05   <- confusion_counts(post_glm_2, 0.5)
cm_glm_low  <- confusion_counts(post_glm_2, t_low)
m_lda1_05   <- metrics(cm_lda1_05)
m_lda1_low  <- metrics(cm_lda1_low)
m_lda_low   <- metrics(cm_lda_low)
m_glm_05    <- metrics(cm_glm_05)
m_glm_low   <- metrics(cm_glm_low)

compare_table <- tibble(
  Model = rep(c("LDA, balance", "LDA, balance + income",
                "Logistic regression, balance + income"), each = 2),
  t = rep(c(0.5, t_low), 3),
  cm = list(cm_lda1_05, cm_lda1_low, cm_lda_05, cm_lda_low,
            cm_glm_05, cm_glm_low)
) |>
  mutate(TP = map_dbl(cm, "TP"), FN = map_dbl(cm, "FN"),
         FP = map_dbl(cm, "FP"), TN = map_dbl(cm, "TN"),
         m = map(cm, metrics),
         error = map_dbl(m, "error"),
         sensitivity = map_dbl(m, "sensitivity"),
         specificity = map_dbl(m, "specificity")) |>
  select(-cm, -m) |>
  mutate(across(c(TP, FN, FP, TN), \(v) format(v, big.mark = ",")),
         across(c(error, sensitivity, specificity),
                \(v) sprintf("%.4f", v))) |>
  knitr::kable(col.names = c("Model", "Threshold", "TP", "FN", "FP", "TN",
                             "Error rate", "Sensitivity", "Specificity"),
               align = "lcccccccc")
```
:::

```{r threshold-data}
#| echo: false
ojs_define(threshold_rates = threshold_rates,
           threshold_marks = c(t_low, 0.5))
```

```{ojs}
//| label: threshold-checkboxes
//| echo: false
rateNames = ["Overall error rate", "Sensitivity (true positive rate)",
             "Specificity (true negative rate)",
             "False positive rate (1 - specificity)"]
rateStyles = ({
  "Overall error rate": {stroke: "black", dash: "none"},
  "Sensitivity (true positive rate)": {stroke: "#D55E00", dash: "8,4"},
  "Specificity (true negative rate)": {stroke: "#0072B2", dash: "2,4"},
  "False positive rate (1 - specificity)": {stroke: "#009E73", dash: "12,4,2,4"}
})
viewof shownRates = Inputs.checkbox(rateNames, {
  value: rateNames.slice(0, 3),
  format: name => html`<span style="display: inline-flex; align-items: center; gap: 0.4em">
    <svg width="36" height="10"><line x1="2" y1="5" x2="34" y2="5"
      stroke="${rateStyles[name].stroke}" stroke-width="2.5"
      stroke-dasharray="${rateStyles[name].dash}" stroke-linecap="round"/></svg>${name}</span>`
})
```

```{ojs}
//| label: threshold-plot
//| echo: false
{
  const rates = transpose(threshold_rates);
  return Plot.plot({
    width: 640, height: 340,
    marginTop: 24,
    x: {domain: [0, 0.5], label: "Threshold t"},
    y: {domain: [0, 1], label: "Rate", grid: true},
    marks: [
      Plot.ruleX(threshold_marks, {stroke: "gray", strokeDasharray: "2,3"}),
      Plot.text(threshold_marks, {x: d => d, y: 1, text: d => `t = ${d}`,
                                  dy: -12, fill: "gray"}),
      ...rateNames.filter(r => shownRates.includes(r)).map(r =>
        Plot.line(rates.filter(d => d.rate === r),
                  {x: "t", y: "value", stroke: rateStyles[r].stroke,
                   strokeDasharray: rateStyles[r].dash, strokeWidth: 2.5,
                   strokeLinecap: "round"}))
    ]
  });
}
```

```{r compare-table}
#| echo: false
compare_table
```

As $t$ falls from $0.5$, sensitivity rises steadily while specificity stays
near $1$, and the false positive rate near $0$, until $t$ is small. The
overall error rate, dominated by the
`r format(cm_lda_05["TN"] + cm_lda_05["FP"], big.mark = ",")`
non-defaulters, follows the false positive rate: nearly flat over most of the
range, then climbing sharply as $t$ approaches $0$ and almost every customer is
flagged.

The overall error rate need not be smallest at $t = 0.5$. LDA's training
error rate is lowest near $t = `r sprintf("%.3f", min_err_lda[["t"]])`$, at
`r sprintf("%.2f", 100 * min_err_lda[["error"]])`%, against
`r sprintf("%.2f", 100 * m_lda_05["error"])`% at $t = 0.5$. The logistic
regression's is lowest near
$t = `r sprintf("%.3f", min_err_glm[["t"]])`$, at
`r sprintf("%.2f", 100 * min_err_glm[["error"]])`%, against
`r sprintf("%.2f", 100 * m_glm_05["error"])`% at $t = 0.5$.

In the table, lowering the threshold to $t = `r t_low`$ raises the
two-feature LDA classifier's sensitivity from
`r sprintf("%.1f", 100 * m_lda_05["sensitivity"])`% to
`r sprintf("%.1f", 100 * m_lda_low["sensitivity"])`%, at the cost of
specificity falling from `r sprintf("%.1f", 100 * m_lda_05["specificity"])`% to
`r sprintf("%.1f", 100 * m_lda_low["specificity"])`% and the overall error rate
rising from `r sprintf("%.2f", 100 * m_lda_05["error"])`% to
`r sprintf("%.2f", 100 * m_lda_low["error"])`%. Adding `income` changes LDA
very little: with `balance` alone the error rate is
`r sprintf("%.2f", 100 * m_lda1_05["error"])`% and the sensitivity
`r sprintf("%.1f", 100 * m_lda1_05["sensitivity"])`% at $t = 0.5$, and
`r sprintf("%.2f", 100 * m_lda1_low["error"])`% and
`r sprintf("%.1f", 100 * m_lda1_low["sensitivity"])`% at $t = `r t_low`$. At
both thresholds logistic regression flags more customers as defaulters than
the two-feature LDA classifier does —
`r cm_glm_05["TP"] + cm_glm_05["FP"]` against
`r cm_lda_05["TP"] + cm_lda_05["FP"]` at $t = 0.5$ — and has the higher
sensitivity and the lower specificity of the two.

### ROC curve

The **ROC curve** (receiver operating characteristic curve) plots sensitivity
against $1 - \text{specificity}$ — the true positive rate against the false
positive rate — at every threshold. At $t = 1$ nothing is predicted positive,
giving the point $(0, 0)$; at $t = 0$ everything is, giving $(1, 1)$; and the
curve traces every tradeoff in between.

The ROC curve depends on a classifier only through how it *orders* the
observations. Any strictly increasing transformation of the score — the
log posterior odds $\hat\delta_2(x) - \hat\delta_1(x)$ or $\hat p(x)$ itself —
sorts the observations identically, produces the same sequence of confusion
matrices, and so traces the same curve. Predicting positive when
$\hat p(x) > t$ is the same as predicting positive when the log posterior
odds exceed the cutoff $c = \log\left[t/(1 - t)\right]$.

For the credit card LDA classifier, the slider below moves the threshold
$t$; its left panel uses the same bins as the score histograms above. A line above
the plots reports $t$, the cutoff $c$ on the log posterior odds, sensitivity,
and specificity. The left panel draws the histograms of the classifier's log
posterior odds within each class, on the density scale, the defaulters' with
a dashed outline and the non-defaulters' with a solid one, and a dotted
vertical line at $c$; the shaded bars to the right of $c$ are the
customers flagged, darker for defaulters (sensitivity) and lighter for
non-defaulters (the false positive rate). The right panel marks where the
classifier lands on its ROC curve, with a dashed diagonal for reference.

::: {.callout-note collapse="true"}
### Code
```{r roc-slider-data}
#| output: false
is_yes <- Default$default == "Yes"

# Empirical ROC curve: every distinct score is a cutoff; classify positive
# when score >= cutoff.
roc_curve <- function(score, positive) {
  ord <- order(score, decreasing = TRUE)
  s   <- score[ord]
  pos <- positive[ord]
  keep <- c(diff(s) != 0, TRUE)
  tibble(fpr = c(0, cumsum(!pos)[keep] / sum(!positive)),
         tpr = c(0, cumsum(pos)[keep] / sum(positive)))
}

# Area under an ROC curve (sorted by fpr) by the trapezoid rule.
auc_trapezoid <- function(roc) {
  with(roc, sum(diff(fpr) * (head(tpr, -1) + tail(tpr, -1)) / 2))
}

# Everything the slider needs, indexed by t = 0.01, 0.02, ..., 0.99, is
# computed here and passed to the page, which only looks values up.
# Each histogram's outline, as a stair through the bin edges.
slider_outline <- slider_bins |>
  pivot_longer(c(x0, x1), values_to = "x") |>
  select(class, x, density)

slider_thresholds <- tibble(t = (1:99) / 100) |>
  mutate(cutoff = qlogis(t),
         m = map(t, \(tt) metrics(confusion_counts(post_lda_2, tt))),
         fpr = 1 - map_dbl(m, "specificity"),
         tpr = map_dbl(m, "sensitivity")) |>
  select(-m)

roc_lda_2 <- roc_curve(post_lda_2, is_yes)
slider_roc <- slice(roc_lda_2, unique(c(seq(1, n(), by = 10), n())))
```
:::

```{r roc-slider-define}
#| echo: false
ojs_define(slider_bins = slider_bins,
           slider_outline = slider_outline,
           slider_thresholds = slider_thresholds,
           slider_roc = slider_roc,
           slider_colors = unname(class_colors))
```

```{ojs}
//| label: roc-slider
//| echo: false
viewof t = Inputs.range([0.01, 0.99], {value: 0.5, step: 0.01, label: "Threshold t"})
```

```{ojs}
//| label: roc-slider-plots
//| echo: false
{
  const bins = transpose(slider_bins);
  const outline = transpose(slider_outline);
  const thresholds = transpose(slider_thresholds);
  const roc = transpose(slider_roc);
  const now = thresholds[Math.round(t * 100) - 1];
  const classes = ["No (negative)", "Yes (positive)"];
  const flagged = k => bins
    .filter(d => d.class === k && d.x1 > now.cutoff)
    .map(d => ({...d, x0: Math.max(d.x0, now.cutoff)}));

  const densityPlot = Plot.plot({
    width: 360, height: 300,
    x: {label: "LDA log posterior odds of default"},
    y: {label: "Density within class"},
    color: {domain: classes, range: slider_colors, legend: true},
    marks: [
      Plot.rect(flagged(classes[0]),
                {x1: "x0", x2: "x1", y1: 0, y2: "density", fill: "class",
                 fillOpacity: 0.2}),
      Plot.rect(flagged(classes[1]),
                {x1: "x0", x2: "x1", y1: 0, y2: "density", fill: "class",
                 fillOpacity: 0.45}),
      Plot.line(outline.filter(d => d.class === classes[0]),
                {x: "x", y: "density", stroke: "class", strokeWidth: 2}),
      Plot.line(outline.filter(d => d.class === classes[1]),
                {x: "x", y: "density", stroke: "class", strokeDasharray: "6,4",
                 strokeWidth: 2}),
      Plot.ruleX([now.cutoff], {strokeDasharray: "2,3"})
    ]
  });

  const rocPlot = Plot.plot({
    width: 300, height: 300,
    x: {domain: [0, 1], label: "False positive rate (1 − specificity)"},
    y: {domain: [0, 1], label: "True positive rate (sensitivity)"},
    marks: [
      Plot.line([[0, 0], [1, 1]], {stroke: "gray", strokeDasharray: "4,4"}),
      Plot.line(roc, {x: "fpr", y: "tpr", strokeWidth: 2}),
      Plot.dot([now], {x: "fpr", y: "tpr", r: 6, fill: "white", stroke: "black",
                       strokeWidth: 2})
    ]
  });

  return html`<p style="text-align: center"><b>t = ${t.toFixed(2)}</b>,
    cutoff c = ${now.cutoff.toFixed(2)}; sensitivity = ${now.tpr.toFixed(3)},
    specificity = ${(1 - now.fpr).toFixed(3)}</p>
    <div style="display: flex; flex-wrap: wrap; gap: 1rem; justify-content: center">
      ${densityPlot}${rocPlot}
    </div>`;
}
```

Moving the slider from $t = 0.9$ to $t = 0.1$ slides the cutoff left, so the
shaded share of the defaulters' histogram (sensitivity) and of the
non-defaulters' (false positive rate) both grow, and the point climbs along
the curve toward $(1, 1)$. The dashed diagonal is the
ROC curve of a classifier that ignores the features and predicts positive at
random: its true and false positive rates are equal at every threshold. The
LDA curve bows far toward the top-left corner, where sensitivity is high and
the false positive rate is low.

The ROC curves below compare LDA and logistic regression with `balance` and
`income` at every threshold. Their points at $t = 0.5$ (filled) and
$t = `r t_low`$ (open) from the threshold table are marked, a triangle for LDA
and a square for logistic regression, with a dashed diagonal for reference.

::: {.callout-note collapse="true"}
### Code
```{r roc-default}
#| output: false
roc_2 <- bind_rows(
  roc_lda_2 |> mutate(model = "LDA"),
  roc_curve(post_glm_2, is_yes) |> mutate(model = "Logistic regression")
)

threshold_labels <- c("t = 0.5", paste("t =", t_low))
roc_points <- tibble(
  model = rep(c("LDA", "Logistic regression"), each = 2),
  t = factor(rep(threshold_labels, 2), levels = threshold_labels),
  m = list(m_lda_05, m_lda_low, m_glm_05, m_glm_low)
) |>
  mutate(fpr = 1 - map_dbl(m, "specificity"),
         tpr = map_dbl(m, "sensitivity"))

roc_plot <- ggplot(roc_2, aes(x = fpr, y = tpr)) +
  geom_abline(intercept = 0, slope = 1, color = "grey60",
              linetype = "dashed") +
  geom_line(aes(color = model, linetype = model), linewidth = 0.9) +
  geom_point(data = roc_points, aes(shape = model, color = model, fill = t),
             size = 3, stroke = 1.2) +
  scale_color_manual(values = model_colors) +
  # Shape separates the models, filled versus open separates the thresholds.
  scale_shape_manual(values = c(24, 22)) +
  scale_fill_manual(values = c("black", "white")) +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +
  coord_fixed() +
  labs(x = "False positive rate (1 - specificity)",
       y = "True positive rate (sensitivity)",
       color = "Model", linetype = "Model", shape = "Model",
       fill = "Threshold", title = "ROC curves, balance + income")

# Balance at which the two models' posterior curves cross, at the lower
# quartile, median, and upper quartile of income, and the posterior
# probability there.
posterior_both <- function(b, inc) {
  nd <- tibble(balance = b, income = inc)
  c(lda = predict(additive_lda, newdata = nd)$posterior[, "Yes"],
    glm = unname(predict(additive_glm, newdata = nd, type = "response")))
}
crossover <- tibble(income = unname(income_levels)) |>
  mutate(balance = map_dbl(income, \(i) {
           uniroot(\(b) diff(posterior_both(b, i)), c(800, 2500))$root
         }),
         p = map2_dbl(balance, income, \(b, i) posterior_both(b, i)[["lda"]]))

# Is one model's flagged set contained in the other's at a threshold?
flags_within <- function(p_inner, p_outer, t) !any(p_inner > t & p_outer <= t)
t_below <- 0.1
lda_within_glm <- flags_within(post_lda_2, post_glm_2, 0.5) &&
  flags_within(post_lda_2, post_glm_2, t_low)
glm_within_lda_below <- flags_within(post_glm_2, post_lda_2, t_below)
n_flag_lda_below <- sum(post_lda_2 > t_below)
n_flag_glm_below <- sum(post_glm_2 > t_below)
```
:::

```{r roc-figure}
#| echo: false
#| fig-width: 6.5
#| fig-height: 5.5
roc_plot
```

The two ROC curves lie almost on top of one another and rise steeply near the
origin, well above the diagonal. Lowering $t$ from $0.5$ to `r t_low` moves
each model's marked point up and slightly right along its curve. At each of
these two thresholds, logistic regression's point sits further along its
nearly shared curve than LDA's.

At the lower quartile, median, and upper quartile of income, the two models'
posterior curves cross at balances of
\$`r dollars(crossover$balance[1])`, \$`r dollars(crossover$balance[2])`,
and \$`r dollars(crossover$balance[3])`, where both give posterior
probabilities of `r sprintf("%.3f", crossover$p[1])`,
`r sprintf("%.3f", crossover$p[2])`, and `r sprintf("%.3f", crossover$p[3])`.
Above those balances LDA's flatter curve lies below logistic regression's, so
a threshold above the crossing probabilities is reached by LDA only at a
higher balance.
`r if (lda_within_glm) "At both marked thresholds, every customer LDA flags is also flagged by logistic regression." else "At the marked thresholds, LDA flags fewer customers than logistic regression."`
At a low threshold the order reverses: at $t = `r t_below`$, LDA flags
`r format(n_flag_lda_below, big.mark = ",")` customers and logistic
regression `r format(n_flag_glm_below, big.mark = ",")``r if (glm_within_lda_below) ", and every customer logistic regression flags is also flagged by LDA" else ""`.

### AUC

The **AUC** is the area under the ROC curve, a single number summarizing the
curve over all thresholds at once. The AUC is at most 1 and, generally, a
larger AUC is better; the caveat below is that a classifier with the larger AUC
can still be the worse one over the false positive rates that matter for an
application. A classifier that ignores the features has an ROC curve along the
diagonal and an AUC of 0.5; a classifier whose curve reaches the top-left
corner, with sensitivity 1 at a false positive rate of 0, has an AUC of 1.

The ROC curve of a sample is a set of points
$\left(\text{FPR}_m, \text{TPR}_m\right)$, one for each threshold at which
the confusion matrix changes. Joining consecutive points by straight lines
and adding up the trapezoids beneath them gives the AUC,

$$\text{AUC} = \sum_{m=1}^{M}\left(\text{FPR}_m - \text{FPR}_{m-1}\right)\frac{\text{TPR}_m + \text{TPR}_{m-1}}{2},$$

where

- $\left(\text{FPR}_m, \text{TPR}_m\right)$ is the false and true positive rate
at the $m$th point, ordered by increasing false positive rate, and
- $m = 0$ is the point $(0, 0)$ and $m = M$ is the point $(1, 1)$.

For the credit card customers, the table below gives the AUC of LDA and
logistic regression with `balance` alone and with `balance` and `income`,
computed this way. In the figure, the left panel shades the area under the
running example's ROC curve, labeled with that area, with a dashed diagonal for
reference. The right panel draws the
ROC curves of four classifiers fit to the same customers — LDA with `balance`
only, with `income` only, and with both, and a classifier that ignores the
features — each labeled with its AUC.

::: {.callout-note collapse="true"}
### Code
```{r auc-default}
#| output: false
scores <- list(
  "LDA, balance" = post_1,
  "Logistic regression, balance" = fitted(balance_glm),
  "LDA, balance + income" = post_lda_2,
  "Logistic regression, balance + income" = post_glm_2
)
auc_values <- tibble(model = names(scores)) |>
  mutate(auc = map_dbl(scores, \(s) auc_trapezoid(roc_curve(s, is_yes))))
auc_of <- function(m) auc_values$auc[auc_values$model == m]
auc_table <- auc_values |>
  mutate(auc = sprintf("%.5f", auc)) |>
  knitr::kable(col.names = c("Model", "AUC"), align = "lc")

auc_area_plot <- ggplot(roc_lda_2, aes(x = fpr, y = tpr)) +
  geom_ribbon(aes(ymin = 0, ymax = tpr), fill = "grey70", alpha = 0.6) +
  geom_abline(intercept = 0, slope = 1, color = "grey40",
              linetype = "dashed") +
  geom_line(linewidth = 0.9) +
  annotate("text", x = 0.65, y = 0.3,
           label = sprintf("AUC = %.5f", auc_of("LDA, balance + income"))) +
  coord_fixed() +
  labs(x = "False positive rate (1 - specificity)",
       y = "True positive rate (sensitivity)",
       title = "Area under the ROC curve")

# Classifiers fit to the same customers; the one that ignores the features
# gives every customer the same score.
compare_scores <- list(
  balance = post_1,
  income  = predict(MASS::lda(default ~ income, data = Default))$posterior[, "Yes"],
  both    = post_lda_2,
  none    = rep(0.5, nrow(Default))
)
compare_auc <- map_dbl(compare_scores, \(s) auc_trapezoid(roc_curve(s, is_yes)))
compare_labels <- parse(text = c(
  sprintf('"Balance only:" ~ AUC == "%.3f"', compare_auc[["balance"]]),
  sprintf('"Income only:" ~ AUC == "%.3f"', compare_auc[["income"]]),
  sprintf('"Balance and income:" ~ AUC == "%.3f"', compare_auc[["both"]]),
  sprintf('"No features:" ~ AUC == "%.3f"', compare_auc[["none"]])
))
compare_roc <- imap(compare_scores, \(s, nm) {
  mutate(roc_curve(s, is_yes), classifier = nm)
}) |>
  list_rbind() |>
  mutate(classifier = factor(classifier, levels = names(compare_scores)))

auc_classifier_plot <- ggplot(compare_roc,
                              aes(x = fpr, y = tpr, color = classifier,
                                  linetype = classifier)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c("#E69F00", "#56B4E9", "#009E73", "grey40"),
                     labels = compare_labels) +
  scale_linetype_manual(values = c("dashed", "dotdash", "solid", "dotted"),
                        labels = compare_labels) +
  coord_fixed() +
  labs(x = "False positive rate (1 - specificity)",
       y = "True positive rate (sensitivity)", color = NULL, linetype = NULL,
       title = "ROC curves, same customers") +
  theme(legend.position = "inside", legend.position.inside = c(0.97, 0.03),
        legend.justification = c(1, 0),
        legend.text = element_text(hjust = 0),
        legend.key.width = unit(1.6, "lines"))

# Where along the false positive rates each of two nearly equal curves is
# the higher one.
fpr_grid <- seq(0, 1, by = 0.001)
tpr_on_grid <- function(s) {
  r <- roc_curve(s, is_yes)
  approx(r$fpr, r$tpr, xout = fpr_grid, ties = max)$y
}
tpr_both    <- tpr_on_grid(compare_scores$both)
tpr_balance <- tpr_on_grid(compare_scores$balance)
share_both_higher    <- mean(tpr_both > tpr_balance + 1e-12)
share_balance_higher <- mean(tpr_balance > tpr_both + 1e-12)
income_dips <- any(tpr_on_grid(compare_scores$income) < fpr_grid - 1e-12)
```
:::

```{r auc-table}
#| echo: false
auc_table
```

```{r auc-figure-display}
#| echo: false
#| layout-ncol: 2
#| fig-width: 4.5
#| fig-height: 4.5
auc_area_plot
auc_classifier_plot
```

With `balance` alone, LDA and logistic regression have identical AUCs of
`r sprintf("%.5f", auc_of("LDA, balance"))`: both posterior probabilities are
increasing functions of balance, so they order the customers identically and
trace the same ROC curve. With `balance` and `income`, the two classifiers'
posterior probabilities weight income differently relative to balance, so
their orderings differ slightly. Their AUCs,
`r sprintf("%.5f", auc_of("LDA, balance + income"))` and
`r sprintf("%.5f", auc_of("Logistic regression, balance + income"))`, differ
by $`r sci(auc_of("LDA, balance + income") - auc_of("Logistic regression, balance + income"), 2)`$.
An AUC of `r sprintf("%.3f", auc_of("LDA, balance + income"))`, close to 1,
reflects an ROC curve that bows far toward the top-left corner.

In the right panel, the AUC ranks the classifiers without a threshold.
Income alone barely separates the classes: its curve stays near the
diagonal`r if (income_dips) ", even dipping below it in places," else ","` and
its AUC is `r sprintf("%.3f", compare_auc[["income"]])`. The balance-only and
two-feature curves nearly coincide and cross: the two-feature curve is the
higher one over `r round(100 * share_both_higher)`% of the range of false
positive rates and the balance-only curve over
`r round(100 * share_balance_higher)`%. When two classifiers' ROC curves
cross, each has the higher true positive rate over part of the range of false
positive rates, and the AUC averages the true positive rate over all false
positive rates, so the classifier with the larger AUC can still be the worse
one over the false positive rates that matter for a given application.

## Credit card findings

LDA with `balance` and `income` identifies defaulters about as well as the
logistic regression with the same features: although its Gaussian
assumptions fit these customers imperfectly, it orders them almost exactly as
logistic regression does, and differs mainly in flagging fewer customers at
high thresholds. Income adds little to classification beyond balance,
raising LDA's AUC by
`r sprintf("%.4f", auc_of("LDA, balance + income") - auc_of("LDA, balance"))`,
although its logistic regression coefficient is clearly nonzero
($z = `r sprintf("%.2f", z_income_glm)`$). At $t = 0.5$
either classifier catches only a minority of defaulters, so a lender that
cares about missed defaults would use a lower threshold.

## Student strata

Student status has not entered any model so far. This additional analysis
asks whether separate models for students and non-students classify better.
The table below gives each stratum's number of customers, number of
defaulters, default rate, and median income.

::: {.callout-note collapse="true"}
### Code
```{r strata-summary}
#| output: false
strata <- c(No = "No", Yes = "Yes")
in_stratum <- map(strata, \(s) filter(Default, student == s))
strata_summary <- Default |>
  group_by(student) |>
  summarize(n = n(), defaulters = sum(default == "Yes"),
            rate = mean(default == "Yes"), median_income = median(income),
            min_income = min(income), max_income = max(income),
            .groups = "drop")
strata_summary_of <- function(s, col) {
  strata_summary[[col]][strata_summary$student == s]
}
strata_summary_table <- strata_summary |>
  transmute(student, n = format(n, big.mark = ","), defaulters,
            rate = sprintf("%.2f%%", 100 * rate),
            median_income = paste0("\\$", dollars(median_income))) |>
  knitr::kable(col.names = c("Student", "Customers", "Defaulters",
                             "Default rate", "Median income"),
               align = "ccccc")
```
:::

```{r strata-summary-table}
#| echo: false
strata_summary_table
```

Students default more often than non-students,
`r sprintf("%.2f", 100 * strata_summary_of("Yes", "rate"))`% against
`r sprintf("%.2f", 100 * strata_summary_of("No", "rate"))`%, and, as the
LDA lecture noted, their median income is far lower.

The figure below plots income against balance separately for non-students
and students, with defaulters and non-defaulters distinguished by color and
shape, and each panel's income axis covering its own stratum's incomes.

::: {.callout-note collapse="true"}
### Code
```{r strata-data}
#| output: false
strata_data_plot <- scatter_base +
  facet_wrap(~ student, labeller = label_both, scales = "free_y") +
  labs(title = "ISLR2::Default, income against balance by student status")
```
:::

```{r strata-data-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 4
strata_data_plot
```

In both strata the defaulters sit at the larger balances. The students'
incomes occupy a lower, narrower band, from
\$`r dollars(strata_summary_of("Yes", "min_income"))` to
\$`r dollars(strata_summary_of("Yes", "max_income"))`, against
\$`r dollars(strata_summary_of("No", "min_income"))` to
\$`r dollars(strata_summary_of("No", "max_income"))` for non-students.

Below, LDA with `balance` and `income` is fit separately to the non-students
and the students, so its priors, class means, and pooled covariance can all
differ by stratum. Beside it is the logistic regression in which the
intercept and both slopes interact with student status. Each customer is
scored by the model for its own stratum. Because each stratum's priors are
its own default rates, each posterior estimates the probability of default
given balance, income, and student status, so posteriors from the two strata
are on one scale and can be compared with each other.

::: {.callout-note collapse="true"}
### Code
```{r strata}
#| output: false
strata_lda <- map(in_stratum, \(d) MASS::lda(default ~ balance + income, data = d))
strata_glm <- map(in_stratum, \(d) glm(default ~ balance + income, data = d,
                                       family = "binomial"))
interacted_glm <- glm(default ~ (balance + income) * student, data = Default,
                      family = "binomial")

# Score every customer with its own stratum's model.
by_stratum <- function(score) {
  p <- numeric(nrow(Default))
  for (s in strata) p[Default$student == s] <- score(s)
  p
}
post_lda_strat <- by_stratum(\(s) {
  predict(strata_lda[[s]], newdata = in_stratum[[s]])$posterior[, "Yes"]
})
post_glm_strat <- by_stratum(\(s) fitted(strata_glm[[s]]))
max_diff_int <- max(abs(fitted(interacted_glm) - post_glm_strat))

strata_metrics <- imap(list(
  "LDA, single fit" = post_lda_2, "LDA, by student stratum" = post_lda_strat,
  "Logistic regression, additive" = post_glm_2,
  "Logistic regression, interacted with student" = fitted(interacted_glm)
), \(p, nm) {
  m <- metrics(confusion_counts(p, 0.5))
  tibble(model = nm, error = m[["error"]],
         auc = auc_trapezoid(roc_curve(p, is_yes)))
}) |>
  list_rbind()
strata_of <- function(m, col) strata_metrics[[col]][strata_metrics$model == m]
strata_table <- strata_metrics |>
  mutate(error = sprintf("%.4f", error), auc = sprintf("%.5f", auc)) |>
  knitr::kable(col.names = c("Model", "Error rate at $t = 0.5$", "AUC"),
               align = "lcc")

# Each stratum's boundary from each model, with its balance threshold at the
# stratum's median income.
strata_hand <- map(in_stratum, \(d) lda_by_hand(d[c("balance", "income")],
                                                 d$default))
strata_lines <- imap(strata, \(s, nm) {
  h <- strata_hand[[s]]
  b <- coef(strata_glm[[s]])
  tibble(student = s, model = c("LDA", "Logistic regression"),
         b0 = c(h$beta0, b[[1]]), b1 = c(h$beta[[1]], b[[2]]),
         b2 = c(h$beta[[2]], b[[3]]))
}) |>
  list_rbind() |>
  left_join(strata_summary |>
              transmute(student = as.character(student), median_income),
            by = "student") |>
  mutate(intercept = -b0 / b2, slope = -b1 / b2,
         threshold = threshold_at(b0, b1, b2, median_income))
strata_value <- function(s, col, m = "LDA") {
  strata_lines[[col]][strata_lines$student == s & strata_lines$model == m]
}
# The LDA and logistic boundaries of a stratum cross where their balance
# thresholds are equal. If each crossing lies outside the incomes the stratum
# has, and logistic regression's threshold is lower at the median income,
# logistic regression's boundary is left of LDA's throughout.
boundary_crossing <- function(s) {
  l <- filter(strata_lines, student == s)
  (l$b0[2] / l$b1[2] - l$b0[1] / l$b1[1]) /
    (l$b2[1] / l$b1[1] - l$b2[2] / l$b1[2])
}
glm_left_throughout <- all(map_lgl(strata, \(s) {
  cr <- boundary_crossing(s)
  inc <- in_stratum[[s]]$income
  (cr < min(inc) || cr > max(inc)) &&
    strata_value(s, "threshold", "Logistic regression") <
      strata_value(s, "threshold")
}))

# How much the students' larger prior alone lowers their LDA threshold:
# the change in log prior odds, converted to balance by the balance slope.
log_prior_odds <- function(h) log(h$pi[2] / h$pi[1])
prior_shift <- (log_prior_odds(strata_hand$Yes) - log_prior_odds(strata_hand$No)) /
  strata_hand$Yes$beta[[1]]
# Class mean balances by stratum, and the midpoint of each stratum's pair.
class_mean <- function(s, k) strata_hand[[s]]$mu[k, "balance"]
midpoint_shift <- mean(strata_hand$Yes$mu[, "balance"]) -
  mean(strata_hand$No$mu[, "balance"])
# Students' minus non-students' LDA threshold over the incomes both strata
# have; the thresholds are linear in income, so the ends bound the gap.
overlap <- c(max(strata_summary$min_income), min(strata_summary$max_income))
threshold_gap <- threshold_at(strata_hand$Yes$beta0, strata_hand$Yes$beta[[1]],
                              strata_hand$Yes$beta[[2]], overlap) -
  threshold_at(strata_hand$No$beta0, strata_hand$No$beta[[1]],
               strata_hand$No$beta[[2]], overlap)

# LDA estimates K - 1 priors, K p means, and p (p + 1) / 2 covariances.
K <- nlevels(Default$default)
p_features <- length(c("balance", "income"))
n_par_lda <- (K - 1) + K * p_features + p_features * (p_features + 1) / 2

strata_plot <- scatter_base +
  boundary_layers(strata_lines) +
  # Each panel's income axis covers its own stratum's incomes.
  facet_wrap(~ student, labeller = label_both, scales = "free_y") +
  labs(linetype = "Boundary",
       title = "Boundaries fit within each student stratum")
```
:::

The table gives the training error rate at $t = 0.5$ and the AUC of these two
models and of the single LDA and additive logistic regression of the earlier
sections, all computed on the training customers. The figure repeats the
scatterplot of income against balance for each stratum, over that stratum's
own range of income, with each stratum's LDA boundary (solid green) and
interacted logistic regression boundary (dashed pink).

```{r strata-table}
#| echo: false
strata_table
```

```{r strata-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 4
strata_plot
```

Fitting the logistic regression with every coefficient interacted with
student status is the same as fitting separate logistic regressions to the
two strata: the two give the same fitted probabilities, to within
$`r sci(max_diff_int, 1)`$.

The extra flexibility barely changes the classifications. Stratifying LDA
moves its training error rate from
`r sprintf("%.2f", 100 * strata_of("LDA, single fit", "error"))`% to
`r sprintf("%.2f", 100 * strata_of("LDA, by student stratum", "error"))`% and
its AUC from `r sprintf("%.4f", strata_of("LDA, single fit", "auc"))` to
`r sprintf("%.4f", strata_of("LDA, by student stratum", "auc"))`; interacting
the logistic regression with student status moves its error rate from
`r sprintf("%.2f", 100 * strata_of("Logistic regression, additive", "error"))`%
to
`r sprintf("%.2f", 100 * strata_of("Logistic regression, interacted with student", "error"))`%
and its AUC from
`r sprintf("%.4f", strata_of("Logistic regression, additive", "auc"))` to
`r sprintf("%.4f", strata_of("Logistic regression, interacted with student", "auc"))`.
The pooled ROC curve sweeps one threshold over the customers of both strata
together, so it compares students' scores with non-students' as well as
within each stratum. These are training rates, and the stratified models estimate
twice as many parameters — `r 2 * n_par_lda` against `r n_par_lda` for LDA
(`r K - 1` prior, `r K * p_features` means, and
`r p_features * (p_features + 1) / 2` covariance entries per fit), and
`r length(coef(interacted_glm))` against `r length(coef(additive_glm))` for
logistic regression — so their training error rates tend to understate test
error, and their training AUCs to overstate test AUC, by more; whether stratifying helps on new customers is a question for
test data.

At each stratum's median income, LDA's balance threshold is
\$`r dollars(strata_value("No", "threshold"))` for non-students and
\$`r dollars(strata_value("Yes", "threshold"))` for students, and logistic
regression's is
\$`r dollars(strata_value("No", "threshold", "Logistic regression"))` and
\$`r dollars(strata_value("Yes", "threshold", "Logistic regression"))`.
`r if (glm_left_throughout) "Over the incomes each stratum actually has, logistic regression's boundary lies to the left of LDA's." else "Over part of the incomes the strata have, the two boundaries swap sides."`

The students' larger prior for `Yes` alone would lower their LDA threshold
by \$`r dollars(prior_shift)`. Their threshold is higher nonetheless because
students in both classes carry higher balances: non-defaulting students
average \$`r dollars(class_mean("Yes", "No"))` against
\$`r dollars(class_mean("No", "No"))` for non-students, and defaulting
students \$`r dollars(class_mean("Yes", "Yes"))` against
\$`r dollars(class_mean("No", "Yes"))`, which puts the midpoint of the
students' class means \$`r dollars(midpoint_shift)` higher. This is the
confounding of the
[Multiple Logistic Regression](20-multiple-logistic-regression.qmd) lecture.
Over the incomes both strata have, \$`r dollars(overlap[1])` to
\$`r dollars(overlap[2])`, the students' threshold minus the non-students'
ranges from \$`r dollars(min(threshold_gap))` to
\$`r dollars(max(threshold_gap))`, so
`r if (all(threshold_gap > 0)) "it stays higher when the strata are compared at the same income" else "the comparison depends on income"`.

## Conclusion

A classifier's error rate hides which mistakes it makes. The confusion matrix
separates them, sensitivity and specificity measure performance within each
true class, and moving the threshold trades one against the other. With $K$
classes the confusion matrix is $K \times K$ and sensitivity and specificity
are computed one class versus the rest. The ROC curve displays the two-class
tradeoff at every threshold and depends only on how a classifier orders the
observations. Its area, the AUC, is at most 1 with larger values generally
better, is computed from the points of the ROC curve by the trapezoid rule, and
ranks classifiers without fixing a threshold, although a classifier with the
larger AUC can be the worse one over the false positive rates that matter for
an application.

Fitting a model separately within each stratum of a categorical variable lets
every parameter differ by stratum; for logistic regression it is the same as
interacting every coefficient with that variable. Stratified
models estimate more parameters, so their training error rates and training
AUCs are more optimistic, and whether stratifying helps is a question for
test data. When each stratum's model uses that stratum's own class
proportions as priors, the posteriors from different strata share one scale,
so a single error rate and AUC can be computed over all observations.

The next lecture returns to generative classifiers. Quadratic discriminant
analysis relaxes LDA's shared covariance matrix, giving each class its own,
and naive Bayes replaces the multivariate Gaussian with a product of
one-feature densities. That lecture then adds K-nearest neighbors and
compares the classifiers on a held-out test set.
