# Quadratic Discriminant Analysis and Naive Bayes

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
set.seed(20261001)

# MASS, e1071, and class are called with :: rather than attached, because
# attaching MASS masks dplyr::select().
class_colors <- c("#0072B2", "#D55E00")
model_colors <- c("Logistic regression" = "#CC79A7", "LDA" = "#009E73",
                  "QDA" = "#E69F00", "Naive Bayes" = "#56B4E9",
                  "KNN" = "black")
model_linetypes <- c("Logistic regression" = "solid", "LDA" = "dashed",
                     "QDA" = "dotted", "Naive Bayes" = "dotdash",
                     "KNN" = "longdash")

# Density of N_2(mu, Sigma) at the rows (x1, x2).
dmvnorm2 <- function(x1, x2, mu, Sigma) {
  d <- cbind(x1 - mu[1], x2 - mu[2])
  q <- rowSums((d %*% solve(Sigma)) * d)
  exp(-q / 2) / (2 * pi * sqrt(det(Sigma)))
}

# Smallest odd integer greater than sqrt(n): a fixed rule of thumb for the
# number of neighbors that does not look at any test data. Odd, so a
# two-class vote cannot tie.
odd_above_sqrt <- function(n) {
  k <- floor(sqrt(n))
  if (k %% 2 == 0) k + 1 else k + 2
}
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **derive** the QDA discriminant function $\delta_k(x)$ from Gaussian
class-conditional densities with class-specific covariance matrices
$\Sigma_k$, and **show** that the resulting decision boundary is quadratic in
$x$, with up to two boundary points when $p = 1$,
- **count** the mean and covariance parameters that LDA, QDA, and Gaussian
naive Bayes estimate, and **use** the counts to **explain** through the
bias-variance tradeoff when each should have the lowest test error rate,
- **state** the naive Bayes assumption that the features are conditionally
independent given the class, **write** $f_k(x)$ as a product of one-feature
densities of possibly different types, and **show** that the log posterior
odds are a sum of one-feature terms,
- **identify** Gaussian naive Bayes as LDA or QDA with a diagonal covariance
matrix,
- **compare** logistic regression, LDA, QDA, naive Bayes, and KNN by the shape
of their decision boundaries, their assumptions, and how they are affected by
$n$ and $p$, and **evaluate** which should perform best in a described
setting, and
- **fit** QDA with `MASS::qda()` and naive Bayes with `e1071::naiveBayes()`,
**interpret** their estimates, and **compare** classifiers by test error
rate, sensitivity, specificity, and AUC on a held-out test set.
:::

## Generative classifier recap

The previous lecture built a classifier from the prior probabilities
$\pi_k = P(Y = k)$ and the class-conditional densities $f_k(x)$ of the
features through Bayes' theorem,

$$p_k(x) = P(Y = k \mid X = x) = \frac{\pi_k f_k(x)}{\sum_{l=1}^K \pi_l f_l(x)},$$

and assigned each observation to the class with the largest posterior
probability. Every generative classifier in this course estimates $\pi_k$ by
the class proportion $n_k/n$. They differ only in the model assumed for $f_k(x)$.

LDA assumed $X \mid Y = k \sim N_p\left(\mu_k, \Sigma\right)$, with one
covariance matrix $\Sigma$ shared by every class. That is what made the
quadratic term in $x$ cancel and the boundary linear. In the credit card data
it is already doubtful: the standard deviation of balance was smaller among
defaulters than among non-defaulters.

This lecture changes the model for $f_k(x)$ in two directions. Quadratic
discriminant analysis gives each class its own covariance matrix. Naive Bayes
drops the multivariate Gaussian and treats the features as independent within
each class.

## Quadratic discriminant analysis

### Class-specific covariance matrices

Quadratic discriminant analysis (QDA) assumes the statistical model

$$P(Y = k) = \pi_k, \qquad X \mid Y = k \sim N_p\left(\mu_k, \Sigma_k\right), \qquad k = 1, \ldots, K,$$

where

- $\mu_k = E[X \mid Y = k]$ is the $p$-vector of feature means in class $k$,
as in LDA, and
- $\Sigma_k = Cov[X \mid Y = k]$ is the $p \times p$ covariance matrix of the
features **in class $k$**, symmetric and positive definite, and free to differ
from class to class.

The class-conditional density is therefore

$$f_k(x) = \frac{1}{(2\pi)^{p/2}\left|\Sigma_k\right|^{1/2}} \exp\left(-\frac{1}{2}\left(x - \mu_k\right)^\top \Sigma_k^{-1} \left(x - \mu_k\right)\right),$$

where $\left|\Sigma_k\right|$ is the determinant of $\Sigma_k$. Each class's
contours are still ellipses centered at $\mu_k$, but now each class's
ellipses have their own size, shape, and orientation.

### Quadratic discriminant

As in LDA, the class with the largest posterior probability is the class with
the largest $\log\left(\pi_k f_k(x)\right)$,

$$\log\left(\pi_k f_k(x)\right) = \log \pi_k - \frac{p}{2}\log(2\pi) - \frac{1}{2}\log\left|\Sigma_k\right| - \frac{1}{2}\left(x - \mu_k\right)^\top \Sigma_k^{-1}\left(x - \mu_k\right).$$

Only the term $-\frac{p}{2}\log(2\pi)$ is common to every class. Dropping it
gives the **quadratic discriminant function**

$$\delta_k(x) = -\frac{1}{2}x^\top \Sigma_k^{-1} x + x^\top \Sigma_k^{-1}\mu_k - \frac{1}{2}\mu_k^\top \Sigma_k^{-1}\mu_k - \frac{1}{2}\log\left|\Sigma_k\right| + \log \pi_k,$$

after expanding the quadratic form as in the previous lecture. Two terms that
cancelled in LDA survive here:

- $-\frac{1}{2}x^\top \Sigma_k^{-1} x$ is quadratic in $x$ and carries a
different matrix in each class, and
- $-\frac{1}{2}\log\left|\Sigma_k\right|$ penalizes a class whose density is
spread over a larger volume.

When every $\Sigma_k$ equals a common $\Sigma$, both terms are the same in
every class, drop out, and $\delta_k(x)$ reduces to the LDA discriminant.
Because only $-\frac{p}{2}\log(2\pi)$, common to every class, was dropped,
the posterior probabilities are recovered as before, $p_k(x) = e^{\delta_k(x)} / \sum_l e^{\delta_l(x)}$.

### Quadratic decision boundary

With $K = 2$ classes the log posterior odds of class $2$ are

$$\log\left(\frac{p_2(x)}{1 - p_2(x)}\right) = \delta_2(x) - \delta_1(x) = -\frac{1}{2}x^\top \left(\Sigma_2^{-1} - \Sigma_1^{-1}\right) x + x^\top\left(\Sigma_2^{-1}\mu_2 - \Sigma_1^{-1}\mu_1\right) + \beta_0,$$

where

- $\beta_0 = \log(\pi_2/\pi_1) - \frac{1}{2}\left(\mu_2^\top \Sigma_2^{-1}\mu_2 - \mu_1^\top \Sigma_1^{-1}\mu_1\right) - \frac{1}{2}\log\left(\left|\Sigma_2\right| / \left|\Sigma_1\right|\right)$
is the intercept, collecting the terms that do not involve $x$.

The log posterior odds are a quadratic function of the features: an
intercept, $p$ linear terms $x_j$, where $x_j$ is the $j$th entry of $x$, and
$p(p+1)/2$ squares $x_j^2$ and cross-products $x_j x_l$, with coefficients
determined by the $\mu_k$ and $\Sigma_k$. The decision boundary, where the log
posterior odds are zero, is a conic section when $p = 2$ instead of a line.

With one feature, writing $\sigma_k^2$ for the class-$k$ variance, the log
posterior odds are

$$\log\left(\frac{p_2(x)}{1 - p_2(x)}\right) = \frac{1}{2}\left(\frac{1}{\sigma_1^2} - \frac{1}{\sigma_2^2}\right)x^2 + \left(\frac{\mu_2}{\sigma_2^2} - \frac{\mu_1}{\sigma_1^2}\right)x + \frac{\mu_1^2}{2\sigma_1^2} - \frac{\mu_2^2}{2\sigma_2^2} + \log\frac{\sigma_1}{\sigma_2} + \log\frac{\pi_2}{\pi_1}.$$

A quadratic in $x$ has up to two roots, so the boundary is up to two points.
If $\sigma_2 > \sigma_1$, the coefficient of $x^2$ is positive, the log
posterior odds grow without bound in both directions, and the class with the
larger variance is assigned in both tails.

The figure below uses $\mu_1 = 0$, $\mu_2 = 1$, and equal priors. Each curve is
$\pi_k f_k(x)$, so the classifier switches classes wherever the curves cross.

::: {.callout-note collapse="true"}
### Code
```{r theory-1d}
#| output: false
mu_1 <- 0
mu_2 <- 1
sigma_shared <- 1
sigma_1 <- 0.7
sigma_2 <- 1.5

# Roots of the one-feature QDA log posterior odds (equal priors).
qda_roots_1d <- function(m1, s1, m2, s2) {
  a  <- (1 / s1^2 - 1 / s2^2) / 2
  b  <- m2 / s2^2 - m1 / s1^2
  c0 <- m1^2 / (2 * s1^2) - m2^2 / (2 * s2^2) + log(s1 / s2)
  sort(Re(polyroot(c(c0, b, a))))
}
roots_1d <- qda_roots_1d(mu_1, sigma_1, mu_2, sigma_2)
x_star_shared <- (mu_1 + mu_2) / 2

grid_1d <- tibble(x = seq(-4, 5, length.out = 600))

curves_1d <- function(s1, s2) {
  grid_1d |>
    mutate(`k = 1` = 0.5 * dnorm(x, mu_1, s1),
           `k = 2` = 0.5 * dnorm(x, mu_2, s2))
}

weighted_1d_plot <- function(s1, s2, x_star, title) {
  curves_1d(s1, s2) |>
    select(x, `k = 1`, `k = 2`) |>
    pivot_longer(-x, names_to = "class", values_to = "density") |>
    ggplot(aes(x = x, y = density, color = class, linetype = class)) +
    geom_line(linewidth = 0.9) +
    geom_vline(xintercept = x_star, linetype = "dotted") +
    scale_color_manual(values = class_colors) +
    labs(x = "x", y = expression(pi[k] * f[k](x)), color = "Class",
         linetype = "Class", title = title)
}

shared_1d_plot <- weighted_1d_plot(
  sigma_shared, sigma_shared, x_star_shared,
  bquote(paste("Shared variance: ", sigma[1] == sigma[2],
               " = ", .(sigma_shared))))

unequal_1d_plot <- weighted_1d_plot(
  sigma_1, sigma_2, roots_1d,
  bquote(paste("Class-specific variances: ", sigma[1] == .(sigma_1), ", ",
               sigma[2] == .(sigma_2))))
```
:::

::: {.panel-tabset}
#### Shared variance
```{r theory-1d-shared-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
shared_1d_plot
```

#### Class-specific variances
```{r theory-1d-unequal-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.2
unequal_1d_plot
```
:::

With a shared variance the curves cross once, at the midpoint
$x^* = `r x_star_shared`$: the LDA picture. With $\sigma_2 > \sigma_1$ the wider
class-$2$ curve rises above the narrow class-$1$ curve on both sides, and the
classifier assigns class $1$ only between
$x = `r sprintf("%.2f", roots_1d[1])`$ and
$x = `r sprintf("%.2f", roots_1d[2])`$.

In two dimensions the boundary is a curve. The figure below gives class $1$
mean $(0, 0)^\top$ and positively correlated features, and class $2$ mean
$(1.5, 0.5)^\top$, larger variances, and negatively correlated features, with
equal priors. The ellipses are contours of each $f_k$.

::: {.callout-note collapse="true"}
### Code
```{r theory-2d}
#| output: false
mu_a    <- c(0, 0)
mu_b    <- c(1.5, 0.5)
Sigma_a <- matrix(c(1, 0.5, 0.5, 1), nrow = 2)
Sigma_b <- matrix(c(3, -0.6, -0.6, 2), nrow = 2)

# Two-class Gaussian log posterior odds of class 2, equal priors.
log_odds_gauss <- function(x1, x2, mu1, S1, mu2, S2) {
  log(dmvnorm2(x1, x2, mu2, S2)) - log(dmvnorm2(x1, x2, mu1, S1))
}

grid_2d <- expand_grid(x1 = seq(-3.5, 5.5, length.out = 220),
                       x2 = seq(-3.5, 4, length.out = 200)) |>
  mutate(`k = 1` = dmvnorm2(x1, x2, mu_a, Sigma_a),
         `k = 2` = dmvnorm2(x1, x2, mu_b, Sigma_b),
         qda = log_odds_gauss(x1, x2, mu_a, Sigma_a, mu_b, Sigma_b))

grid_2d_long <- grid_2d |>
  pivot_longer(c(`k = 1`, `k = 2`), names_to = "class",
               values_to = "density")

# Contours at Mahalanobis radii 1 and 2 for each class.
contour_at <- function(S) exp(-c(1, 2)^2 / 2) / (2 * pi * sqrt(det(S)))

means_2d <- tibble(class = c("k = 1", "k = 2"),
                   x1 = c(mu_a[1], mu_b[1]), x2 = c(mu_a[2], mu_b[2]))

density_layers <- function(long_grid) {
  list(
    geom_contour(data = filter(long_grid, class == "k = 1"),
                 aes(z = density, color = class, linetype = class),
                 breaks = contour_at(Sigma_a), linewidth = 0.7),
    geom_contour(data = filter(long_grid, class == "k = 2"),
                 aes(z = density, color = class, linetype = class),
                 breaks = contour_at(Sigma_b), linewidth = 0.7),
    geom_point(data = means_2d, aes(color = class, shape = class), size = 3),
    scale_color_manual(values = class_colors),
    coord_fixed(xlim = c(-3.5, 5.5), ylim = c(-3.5, 4), expand = FALSE),
    labs(x = expression(X[1]), y = expression(X[2]), color = "Class",
         linetype = "Class", shape = "Class")
  )
}

theory_2d_base <- ggplot(mapping = aes(x = x1, y = x2)) +
  density_layers(grid_2d_long)

theory_2d_densities <- theory_2d_base +
  labs(title = "Class densities with class-specific covariance matrices")

# The boundary uses "longdash", distinct from both class-contour linetypes.
theory_2d_qda <- theory_2d_base +
  geom_contour(data = grid_2d, aes(z = qda), breaks = 0, color = "black",
               linewidth = 1.1, linetype = "longdash") +
  labs(title = "QDA boundary (black, long dashes)")
```
:::

::: {.panel-tabset}
#### Class densities
```{r theory-2d-densities-figure}
#| echo: false
#| fig-width: 6
#| fig-height: 5
theory_2d_densities
```

#### QDA boundary
```{r theory-2d-qda-figure}
#| echo: false
#| fig-width: 6
#| fig-height: 5
theory_2d_qda
```
:::

The class-$2$ ellipses are larger in every direction, so the QDA boundary is
an ellipse around the class-$1$ mean, long enough to run off the plot at the
lower left: class $1$ is assigned inside it and class $2$ everywhere outside
it.

### Class-specific covariance estimates

The priors and means are estimated as in LDA, $\hat\pi_k = n_k/n$ and
$\hat\mu_k = \frac{1}{n_k}\sum_{i:\, y_i = k} x_i$. Each covariance matrix is
estimated from its own class alone,

$$\hat\Sigma_k = \frac{1}{n_k - 1}\sum_{i:\, y_i = k}\left(x_i - \hat\mu_k\right)\left(x_i - \hat\mu_k\right)^\top,$$

where

- $x_i$ is the $p$-vector of feature values of observation $i$, as in the
previous lecture, and
- $\hat\Sigma_k$ is the **class-$k$ sample covariance matrix**, whose divisor
$n_k - 1$ accounts for the one estimated mean.

`MASS::qda()` uses this divisor (maximum likelihood would use $n_k$), and
$\hat\Sigma_k$, whose rank is at most $n_k - 1$, is invertible only if
$n_k > p$.

### Bias-variance tradeoff

QDA's flexibility costs parameters. Not counting the $K - 1$ free priors,
which every generative classifier estimates,

$$\text{LDA: } Kp + \frac{p(p+1)}{2}, \qquad \text{QDA: } Kp + K\,\frac{p(p+1)}{2},$$

where

- $Kp$ counts the entries of the $K$ mean vectors, and
- $p(p+1)/2$ counts the distinct entries of one symmetric $p \times p$
covariance matrix: $p$ variances and $p(p-1)/2$ covariances.

::: {.callout-note collapse="true"}
### Code
```{r parameter-counts}
#| output: false
count_lda <- function(K, p) K * p + p * (p + 1) / 2
count_qda <- function(K, p) K * p + K * p * (p + 1) / 2

parameter_table <- tibble(p = c(1, 2, 5, 10, 50)) |>
  mutate(LDA = count_lda(2, p), QDA = count_qda(2, p),
         across(c(LDA, QDA), \(v) format(v, big.mark = ","))) |>
  knitr::kable(col.names = c("Features $p$", "LDA", "QDA"), align = "ccc")
```
:::

With $K = 2$ classes:

```{r parameter-table}
#| echo: false
parameter_table
```

QDA's count grows like $Kp^2/2$, against LDA's $p^2/2$. When the
$\Sigma_k$ really differ, LDA's shared $\Sigma$ makes its boundary the wrong
shape however much data it sees: bias. QDA's boundary has the right shape, but
each of its many covariance estimates comes from one class's observations and
varies from one training set to the next: variance.

The figure below draws repeated training sets from the two classes of the
previous figure, fits LDA and QDA to each, and overlays the fitted boundaries
(thin grey) on the Bayes decision boundary (thick dashed), which is the QDA
boundary computed from the true parameters. The first tab uses small training
sets and the second large ones.

::: {.callout-note collapse="true"}
### Code
```{r boundary-variability}
#| output: false
rmvnorm2 <- function(n, mu, Sigma) {
  sweep(matrix(rnorm(n * 2), n) %*% chol(Sigma), 2, mu, "+")
}

draw_quadratic <- function(n_k) {
  X <- rbind(rmvnorm2(n_k, mu_a, Sigma_a), rmvnorm2(n_k, mu_b, Sigma_b))
  tibble(y = factor(rep(c("k = 1", "k = 2"), each = n_k)),
         x1 = X[, 1], x2 = X[, 2])
}

grid_fit <- expand_grid(x1 = seq(-3.5, 5.5, length.out = 120),
                        x2 = seq(-3.5, 4, length.out = 110))

n_k_small <- 25
n_k_large <- 250
n_boundary_reps <- 20

fitted_boundaries <- expand_grid(n_k = c(n_k_small, n_k_large),
                                 rep = seq_len(n_boundary_reps)) |>
  mutate(data = map(n_k, draw_quadratic),
         LDA = map(data, \(d) {
           predict(MASS::lda(y ~ x1 + x2, data = d),
                   newdata = grid_fit)$posterior[, "k = 2"]
         }),
         QDA = map(data, \(d) {
           predict(MASS::qda(y ~ x1 + x2, data = d),
                   newdata = grid_fit)$posterior[, "k = 2"]
         })) |>
  select(-data) |>
  pivot_longer(c(LDA, QDA), names_to = "method", values_to = "p_hat") |>
  mutate(grid = list(grid_fit)) |>
  unnest(c(grid, p_hat))

variability_plot <- function(nk) {
  ggplot(filter(fitted_boundaries, n_k == nk), aes(x = x1, y = x2)) +
    geom_contour(aes(z = p_hat - 0.5, group = rep), breaks = 0,
                 color = "grey55", linewidth = 0.35) +
    geom_contour(data = grid_2d, aes(z = qda), breaks = 0, color = "black",
                 linewidth = 1.1, linetype = "dashed") +
    facet_wrap(~ method) +
    coord_fixed(xlim = c(-3.5, 5.5), ylim = c(-3.5, 4), expand = FALSE) +
    labs(x = expression(X[1]), y = expression(X[2]),
         title = bquote(paste(.(n_boundary_reps), " fitted boundaries, ",
                              n[k] == .(nk),
                              " per class; Bayes boundary dashed")))
}
variability_small <- variability_plot(n_k_small)
variability_large <- variability_plot(n_k_large)
```
:::

::: {.panel-tabset}
#### Small training sets
```{r variability-small-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 4
variability_small
```

#### Large training sets
```{r variability-large-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 4
variability_large
```
:::

With `r n_k_small` observations per class the LDA lines vary in angle, and
none follows the curvature of the Bayes boundary.
The QDA curves follow the shape of the Bayes boundary but scatter widely
around it, some closing into small ellipses and others opening out of the
plot. With `r n_k_large` per class the LDA lines tighten around a single line
that still misses the curvature, while the QDA curves collapse onto the Bayes
boundary: more data reduces QDA's variance but not LDA's bias.

LDA should therefore win when the training set is small relative to the
number of covariance parameters, or when the class covariance matrices are
close to equal so that its bias is small. QDA should win when the covariance
matrices clearly differ and the training set is large enough to estimate
them. The simulation study later in this lecture checks the covariance half
of this at a fixed training-set size.

::: {.callout-note collapse="true"}
### Beyond this course
Regularized discriminant analysis compromises between the two by shrinking
each class covariance toward the pooled one,
$\hat\Sigma_k(\alpha) = \alpha\hat\Sigma_k + (1 - \alpha)\hat\Sigma$, where
$\alpha = 0$ gives LDA and $\alpha = 1$ gives QDA. The value of $\alpha$ is a
flexibility dial chosen from the data, as the next unit will choose the
number of neighbors for KNN.
:::

## Naive Bayes

### Conditional independence

Naive Bayes keeps the generative framework but replaces the multivariate model
for $f_k$ with an assumption about how the features relate to one another:

$$P(Y = k) = \pi_k, \qquad X_1, \ldots, X_p \text{ are independent given } Y = k, \qquad X_j \mid Y = k \sim f_{kj},$$

where

- $X_j$ is the $j$th feature, the $j$th entry of the random vector $X$, and
- $f_{kj}$ is the one-feature density (or probability mass function) of $X_j$
among observations in class $k$.

Independence turns the joint density into a product of its marginals, so the
class-conditional density is

$$f_k(x) = f_{k1}(x_1) \times f_{k2}(x_2) \times \cdots \times f_{kp}(x_p) = \prod_{j=1}^p f_{kj}(x_j).$$

The posterior probability is Bayes' theorem with this product in place of
$f_k(x)$.

The assumption is conditional on the class. Features that are related only
because both differ between classes, such as two symptoms that are each more
common among patients with a disease, can be independent within each class
while being correlated overall.

### Feature densities

Each $f_{kj}$ describes a single feature in a single class, so each can be
modeled and estimated separately, and features of different types can use
densities of different types.

- A quantitative feature is usually modeled as Gaussian,
$X_j \mid Y = k \sim N\left(\mu_{kj}, \sigma_{kj}^2\right)$, with $\mu_{kj}$
and $\sigma_{kj}^2$ estimated by the sample mean and variance of feature $j$
among the class-$k$ training observations.
- A binary feature is Bernoulli,
$f_{kj}(x_j) = \theta_{kj}^{x_j}\left(1 - \theta_{kj}\right)^{1 - x_j}$ for
$x_j \in \{0, 1\}$, where $\theta_{kj} = P(X_j = 1 \mid Y = k)$ is estimated
by the proportion of class-$k$ training observations with $x_j = 1$.
- A categorical feature with more than two levels gets one probability per
level in each class, estimated by the class-$k$ proportions at each level.

### Additive log posterior odds

With $K = 2$, the logarithm of the product is a sum, and the log posterior
odds of class $2$ are

$$\log\left(\frac{p_2(x)}{1 - p_2(x)}\right) = \log\frac{\pi_2}{\pi_1} + \sum_{j=1}^p \log\frac{f_{2j}(x_j)}{f_{1j}(x_j)} = \log\frac{\pi_2}{\pi_1} + \sum_{j=1}^p g_j(x_j),$$

where

- $g_j(x_j) = \log\left(f_{2j}(x_j)/f_{1j}(x_j)\right)$ is the contribution of
feature $j$, a function of $x_j$ alone.

Each feature shifts the log posterior odds by an amount that does not depend
on the values of the other features: naive Bayes is an additive model with no
interactions, the same structural restriction as the additive logistic
regressions of the multiple logistic regression lecture. The form of each
$g_j$ follows from the type of $f_{kj}$:

- Gaussian with class-specific variances: $g_j$ is quadratic in $x_j$, with
coefficient $\frac{1}{2}\left(1/\sigma_{1j}^2 - 1/\sigma_{2j}^2\right)$ on
$x_j^2$.
- Gaussian with a variance shared by the two classes: the $x_j^2$ terms
cancel and $g_j$ is linear in $x_j$.
- Bernoulli: $g_j(x_j) = x_j\log\left(\theta_{2j}/\theta_{1j}\right) + \left(1 - x_j\right)\log\left[\left(1 - \theta_{2j}\right)/\left(1 - \theta_{1j}\right)\right]$,
linear in $x_j$, so the feature shifts the log posterior odds by
$\log\left\{\left[\theta_{2j}/\left(1 - \theta_{2j}\right)\right] / \left[\theta_{1j}/\left(1 - \theta_{1j}\right)\right]\right\}$
when $x_j$ changes from $0$ to $1$.

### Diagonal covariance

When every $f_{kj}$ is Gaussian, the product of $p$ independent normal
densities is a multivariate normal density with a diagonal covariance matrix:

$$\prod_{j=1}^p \frac{1}{\sqrt{2\pi}\,\sigma_{kj}}\exp\left(-\frac{\left(x_j - \mu_{kj}\right)^2}{2\sigma_{kj}^2}\right) = \frac{1}{(2\pi)^{p/2}\left|\Lambda_k\right|^{1/2}}\exp\left(-\frac{1}{2}\left(x - \mu_k\right)^\top \Lambda_k^{-1}\left(x - \mu_k\right)\right),$$

where

- $\Lambda_k = \text{diag}\left(\sigma_{k1}^2, \ldots, \sigma_{kp}^2\right)$ is the
diagonal matrix of class-$k$ feature variances, whose determinant is
$\prod_j \sigma_{kj}^2$ and whose inverse is diagonal with entries
$1/\sigma_{kj}^2$.

Gaussian naive Bayes with class-specific variances is therefore QDA with
every $\Sigma_k$ restricted to be diagonal, and Gaussian naive Bayes with one
variance per feature shared across classes is LDA with $\Sigma$ restricted to
be diagonal.

The figure below uses the two classes of the QDA figures. Naive Bayes keeps
each class's mean and feature variances but discards the correlations; the
ellipses drawn are contours of the naive Bayes densities.

::: {.callout-note collapse="true"}
### Code
```{r theory-nb}
#| output: false
Lambda_a <- diag(diag(Sigma_a))   # diagonal covariance: variances only
Lambda_b <- diag(diag(Sigma_b))

grid_nb <- grid_2d |>
  select(x1, x2, qda) |>
  mutate(nb = log_odds_gauss(x1, x2, mu_a, Lambda_a, mu_b, Lambda_b),
         `k = 1` = dmvnorm2(x1, x2, mu_a, Lambda_a),
         `k = 2` = dmvnorm2(x1, x2, mu_b, Lambda_b))

grid_nb_long <- grid_nb |>
  pivot_longer(c(`k = 1`, `k = 2`), names_to = "class",
               values_to = "density")

nb_density_plot <- ggplot(mapping = aes(x = x1, y = x2)) +
  geom_contour(data = filter(grid_nb_long, class == "k = 1"),
               aes(z = density, color = class, linetype = class),
               breaks = contour_at(Lambda_a), linewidth = 0.7) +
  geom_contour(data = filter(grid_nb_long, class == "k = 2"),
               aes(z = density, color = class, linetype = class),
               breaks = contour_at(Lambda_b), linewidth = 0.7) +
  geom_point(data = means_2d, aes(color = class, shape = class), size = 3) +
  scale_color_manual(values = class_colors) +
  coord_fixed(xlim = c(-3.5, 5.5), ylim = c(-3.5, 4), expand = FALSE) +
  labs(x = expression(X[1]), y = expression(X[2]), color = "Class",
       linetype = "Class", shape = "Class")

# Boundary linetypes "longdash" and "twodash" differ from the contour linetypes.
nb_boundary_plot <- nb_density_plot +
  geom_contour(data = grid_nb, aes(z = nb), breaks = 0, color = "grey40",
               linewidth = 1.1, linetype = "twodash") +
  geom_contour(data = grid_nb, aes(z = qda), breaks = 0, color = "black",
               linewidth = 1.1, linetype = "longdash") +
  labs(title = "Bayes boundary (black, long dashes) and\nnaive Bayes boundary (grey, dash-dot)")
```
:::

```{r theory-nb-boundary-figure}
#| echo: false
#| fig-width: 6
#| fig-height: 5
nb_boundary_plot
```

Unlike the class densities of the QDA figure, the naive Bayes ellipses are
aligned with the feature axes. The naive Bayes boundary is still an ellipse,
because class $2$'s variance is the larger in both features, and since its
quadratic has no $x_1 x_2$ term it is aligned with the axes too. Ignoring the
correlations replaces the long, tilted Bayes ellipse with a rounder one: naive Bayes assigns class $1$ to
part of the upper left where the Bayes classifier assigns class $2$, and
class $2$ to the lower-left extension of the Bayes ellipse, where the Bayes
classifier assigns class $1$.

### Parameter count

Gaussian naive Bayes with class-specific variances estimates $Kp$ means and
$Kp$ variances, $2Kp$ parameters, against QDA's
$Kp + Kp(p+1)/2$. With $K = 2$ and $p = 50$ that is
`r format(2 * 2 * 50, big.mark = ",")` parameters against
`r format(count_qda(2, 50), big.mark = ",")`. The count grows linearly in
$p$, and each parameter is estimated from one feature in one class, using all
$n_k$ observations of that class.

For $p$ binary features the saving is larger still: a joint probability mass
function has $2^p$ cells per class ($2^p - 1$ free probabilities, or
`r format(2^20 - 1, big.mark = ",")` when $p = 20$), most of them never
observed in a training set, against naive Bayes's $p$ Bernoulli
probabilities, which sidesteps the curse of dimensionality of the flexibility
lecture.

The price is bias whenever the features are dependent within a class, which
in practice is almost always. A classification depends only on which side of
the threshold the estimated posterior probability falls, so a biased estimate
of $f_k(x)$ still gives the Bayes classifier's answer wherever the posterior
probability it produces falls on the same side of the threshold as the true
one. Whether it does is a question for data, taken up in the
next two sections.

::: {.callout-note collapse="true"}
### Beyond this course
A quantitative feature need not be modeled as Gaussian. Each $f_{kj}$ can
instead be estimated nonparametrically, by a histogram or a kernel density
estimate of feature $j$ within class $k$, since each is a one-dimensional
density. The `naivebayes` package offers this option.
:::

## Classifier comparison

### KNN classification

Five classifiers are now available: logistic regression, LDA, QDA, naive
Bayes, and K-nearest neighbors (KNN). From here through the worked example,
$K$ is the number of neighbors rather than the number of classes: every comparison has two classes, and the number of
classes is written as $2$.

The flexibility lecture introduced KNN for a quantitative response. For a
qualitative response the neighbors vote:

$$\hat p_k(x_0) = \frac{1}{K}\sum_{i \in \mathcal{N}_0}\mathrm{I}(y_i = k),$$

where

- $K$ is the **number of neighbors**,
- $\mathcal{N}_0$ is the set of the $K$ training observations closest to $x_0$
in Euclidean distance after every feature, in the training data and in $x_0$
alike, is standardized by its training mean and standard deviation, since
that distance depends on each feature's scale (as in the flexibility
lecture), and
- $\hat p_k(x_0)$ is the proportion of those neighbors in class $k$, and the
classifier assigns the class with the largest proportion.

### Boundary shapes

With two classes, the four model-based classifiers differ in the form of the
log posterior odds they can represent:

$$\begin{aligned}
\text{Logistic regression, LDA: } & \beta_0 + \sum_{j=1}^p \beta_j x_j, \\
\text{QDA: } & \beta_0 + \sum_{j=1}^p \beta_j x_j + \sum_{j \le l} \gamma_{jl} x_j x_l, \\
\text{Gaussian naive Bayes: } & \beta_0 + \sum_{j=1}^p \left(\beta_j x_j + \gamma_{jj} x_j^2\right),
\end{aligned}$$

where

- $\beta_0$ is an intercept and $\beta_j$ the coefficient of $x_j$, with
values that differ from method to method, and
- $\gamma_{jl}$ is the coefficient of the product $x_j x_l$, a square when
$l = j$.

A naive Bayes classifier with Bernoulli or categorical features adds terms
linear in those features' indicators. KNN has no formula for its log
posterior odds. The table summarizes all five:

::: {.callout-note collapse="true"}
### Code
```{r comparison-table}
#| output: false
comparison_table <- tibble(
  Method = c("Logistic regression", "LDA", "QDA", "Gaussian naive Bayes",
             "KNN"),
  Boundary = c("Linear", "Linear", "Quadratic", "Quadratic, no cross-products",
               "Any shape"),
  Assumption = c("Linear log-odds",
                 "Gaussian, shared covariance",
                 "Gaussian, class covariances",
                 "Features independent within each class",
                 "Smooth class probabilities"),
  Estimation = c("Iterative", "Closed form", "Closed form", "Closed form",
                 "None (stores data)")
) |>
  knitr::kable(align = "llll")
```
:::

```{r comparison-table-display}
#| echo: false
comparison_table
```

Reading down the table:

- **Logistic regression and LDA** share a linear form and differ in
estimation. LDA adds the Gaussian, shared-covariance assumption; when it
holds, both estimate the same log posterior odds, and when it fails, logistic
regression, which never models the features' distribution, is unaffected by
that failure provided the log-odds remain linear.
- **QDA** adds every square and cross-product. A logistic regression given
those same squared and product features, as in the flexible logistic
regression lecture, has the same quadratic form, fit by the conditional
likelihood instead.
- **Naive Bayes** is the most restricted of the quadratic forms, with no
cross-products, and in exchange has fewer parameters than QDA.
- **KNN** is nonparametric: it assumes no form at all, and its flexibility is
set by $1/K$. It can follow any boundary given enough data, but it needs $n$
large relative to $p$, suffers most from the curse of dimensionality, and has
no coefficients to interpret.

Increasing $n$ favors the more flexible methods, QDA and KNN, whose lower
bias becomes affordable as their variance shrinks. Increasing $p$ favors the
more restricted ones: naive Bayes, whose parameter count grows linearly in
$p$, and LDA, whose count grows like $p^2/2$ rather than like QDA's $p^2$ for
two classes. No method is best in every setting.

### Simulated scenarios

A small simulation puts these predictions to a test. Each of four scenarios
has two classes with equal priors, a data-generating process chosen so that a
different method's assumptions hold:

- **Linear**: two Gaussian features, a shared covariance matrix with
correlation $0.5$, and different means, so the Bayes boundary is a line.
- **Quadratic**: the two classes of the QDA figures, with class-specific
covariance matrices.
- **Independent, $p = 10$**: ten Gaussian features, independent within each
class; the classes differ in the means of the first five features and the
variances of the last five.
- **Non-linear**: two features uniform on a square, class $2$ above the curve
$x_2 = \sin(2x_1)$ and class $1$ below it, and then $10\%$ of the class labels
flipped at random, so neither class is Gaussian and the Bayes boundary is the
sine curve.

::: {.callout-note collapse="true"}
### Code
```{r scenario-study}
#| output: false
scenario_names <- c("Linear", "Quadratic", "Independent, p = 10",
                    "Non-linear")
n_reps <- 100       # training sets per scenario
n_train_k <- 50     # training observations per class
n_test_k <- 2000    # test observations per class
k_sim <- odd_above_sqrt(2 * n_train_k)
knn_sim_name <- paste0("KNN, K = ", k_sim)

rmvnorm_p <- function(n, mu, Sigma) {
  sweep(matrix(rnorm(n * length(mu)), n) %*% chol(Sigma), 2, mu, "+")
}

# One data set of 2 * n_k observations, n_k per class (in expectation for the
# non-linear scenario, whose class sizes are random).
draw_scenario <- function(scenario, n_k) {
  labels <- c("Class 1", "Class 2")
  if (scenario == "Non-linear") {
    X <- matrix(runif(4 * n_k, -2, 2), ncol = 2)
    y <- as.numeric(X[, 2] > sin(2 * X[, 1]))
    flip <- runif(2 * n_k) < 0.1
    y[flip] <- 1 - y[flip]
    y <- factor(labels[y + 1], levels = labels)
  } else {
    if (scenario == "Linear") {
      S <- matrix(c(1, 0.5, 0.5, 1), nrow = 2)
      X <- rbind(rmvnorm_p(n_k, c(0, 0), S), rmvnorm_p(n_k, c(1.5, 0.5), S))
    }
    if (scenario == "Quadratic") {
      X <- rbind(rmvnorm_p(n_k, mu_a, Sigma_a), rmvnorm_p(n_k, mu_b, Sigma_b))
    }
    if (scenario == "Independent, p = 10") {
      X <- rbind(rmvnorm_p(n_k, rep(0, 10), diag(10)),
                 rmvnorm_p(n_k, rep(c(0.5, 0), each = 5),
                           diag(rep(c(1, 2), each = 5))))
    }
    y <- factor(rep(labels, each = n_k), levels = labels)
  }
  colnames(X) <- paste0("x", seq_len(ncol(X)))
  bind_cols(tibble(y = y), as_tibble(X))
}

# Bayes classifier's error rate on a test set, from the true model.
bayes_error <- function(scenario, test) {
  X <- as.matrix(select(test, -y))
  truth <- test$y == "Class 2"
  if (scenario == "Non-linear") {
    pred <- X[, 2] > sin(2 * X[, 1])
  } else {
    if (scenario == "Linear") {
      S <- matrix(c(1, 0.5, 0.5, 1), nrow = 2)
      lo <- log_odds_gauss(X[, 1], X[, 2], c(0, 0), S, c(1.5, 0.5), S)
    }
    if (scenario == "Quadratic") {
      lo <- log_odds_gauss(X[, 1], X[, 2], mu_a, Sigma_a, mu_b, Sigma_b)
    }
    if (scenario == "Independent, p = 10") {
      lo <- rowSums(dnorm(X, rep(c(0.5, 0), each = 5)[col(X)],
                          sqrt(rep(c(1, 2), each = 5))[col(X)], log = TRUE) -
                    dnorm(X, 0, 1, log = TRUE))
    }
    pred <- lo > 0
  }
  mean(pred != truth)
}

# Test error rate of each method fit to one training set.
fit_methods <- function(train, test) {
  truth <- test$y
  X_train <- as.matrix(select(train, -y))
  X_test  <- as.matrix(select(test, -y))
  # KNN uses features standardized by the training means and SDs.
  centers <- colMeans(X_train)
  scales  <- apply(X_train, 2, sd)
  Z_train <- scale(X_train, centers, scales)
  Z_test  <- scale(X_test, centers, scales)
  k_sqrt  <- odd_above_sqrt(nrow(train))
  glm_fit <- glm(y ~ ., data = train, family = "binomial")
  errors <- c(`Logistic regression` = mean(
      (predict(glm_fit, newdata = test, type = "response") > 0.5) !=
        (truth == "Class 2")),
    LDA = mean(predict(MASS::lda(y ~ ., data = train), test)$class != truth),
    QDA = mean(predict(MASS::qda(y ~ ., data = train), test)$class != truth),
    `Naive Bayes` = mean(
      predict(e1071::naiveBayes(y ~ ., data = train), test) != truth),
    `KNN, K = 1` = mean(class::knn(Z_train, Z_test, train$y, k = 1) != truth),
    knn_sqrt = mean(class::knn(Z_train, Z_test, train$y, k = k_sqrt) != truth))
  names(errors)[names(errors) == "knn_sqrt"] <- paste0("KNN, K = ", k_sqrt)
  errors
}

scenario_results <- map(scenario_names, \(s) {
  test <- draw_scenario(s, n_test_k)
  errors <- map(seq_len(n_reps),
                \(r) fit_methods(draw_scenario(s, n_train_k), test))
  list(errors = bind_rows(errors) |> mutate(scenario = s, rep = row_number()),
       bayes = bayes_error(s, test))
})

method_levels <- c("Logistic regression", "LDA", "QDA", "Naive Bayes",
                   "KNN, K = 1", knn_sim_name)

scenario_errors <- map(scenario_results, "errors") |>
  bind_rows() |>
  pivot_longer(all_of(method_levels), names_to = "method",
               values_to = "error") |>
  mutate(method = factor(method, levels = method_levels),
         scenario = factor(scenario, levels = scenario_names))

scenario_bayes <- tibble(
  scenario = factor(scenario_names, levels = scenario_names),
  bayes = map_dbl(scenario_results, "bayes"))

scenario_means <- scenario_errors |>
  group_by(scenario, method) |>
  summarize(mean_error = mean(error), .groups = "drop")

mean_error_of <- function(s, m) {
  scenario_means$mean_error[scenario_means$scenario == s &
                              scenario_means$method == m]
}
best_method <- function(s) {
  d <- filter(scenario_means, scenario == s)
  as.character(d$method[which.min(d$mean_error)])
}
pct <- function(v) sprintf("%.1f%%", 100 * v)

# Prose labels for the winning method of each scenario.
prose_label <- c("Logistic regression" = "logistic regression", "LDA" = "LDA",
                 "QDA" = "QDA", "Naive Bayes" = "naive Bayes",
                 "KNN, K = 1" = "KNN with $K = 1$")
prose_label[knn_sim_name] <- paste0("KNN with $K = ", k_sim, "$")
winner <- function(s) unname(prose_label[best_method(s)])

pp <- function(v) sprintf("%.3f", 100 * v)   # percentage points

# One training set per scenario to show what the data look like; the
# independent scenario is shown in features 1 (mean shift) and 6 (variance).
scenario_examples <- map(scenario_names, \(s) {
  draw_scenario(s, n_train_k) |>
    transmute(y, first = x1, second = if (s == "Independent, p = 10") x6 else x2,
              scenario = s)
}) |>
  bind_rows() |>
  mutate(scenario = factor(scenario, levels = scenario_names))

scenario_data_plot <- ggplot(scenario_examples,
                             aes(x = first, y = second, color = y, shape = y)) +
  geom_point(size = 1.6, alpha = 0.8) +
  facet_wrap(~ scenario, scales = "free") +
  scale_color_manual(values = class_colors) +
  labs(x = "First plotted feature", y = "Second plotted feature",
       color = "Class", shape = "Class",
       title = "One training set per scenario (Independent: features 1 and 6)")

scenario_error_plot <- ggplot(scenario_errors, aes(x = method, y = error)) +
  geom_boxplot(outlier.size = 0.6, fill = "grey90") +
  geom_hline(data = scenario_bayes, aes(yintercept = bayes),
             linetype = "dashed") +
  facet_wrap(~ scenario, scales = "free_y") +
  labs(x = NULL, y = "Test error rate",
       title = paste("Test error rates over", n_reps,
                     "training sets; Bayes error rate dashed")) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```
:::

In each scenario, `r n_reps` training sets of `r 2 * n_train_k` observations
were drawn, `r n_train_k` per class (in expectation in the non-linear
scenario). Each method was fit to every training set and its error rate
computed on one large test set of `r format(2 * n_test_k, big.mark = ",")`
observations drawn from the same scenario. KNN was fit on standardized
features with $K = 1$ and with
$K = `r k_sim`$, the smallest odd integer above $\sqrt{n}$ for the
$n = `r 2 * n_train_k`$ training observations: odd so that a two-class vote
cannot tie, and near $\sqrt{n}$ as a rule of thumb rather than an optimal
choice. Both are fixed in advance, since choosing $K$ from the data requires
the resampling methods of the next unit.

::: {.panel-tabset}
#### Scenarios
```{r scenario-data-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 6.5
scenario_data_plot
```

#### Test error rates
```{r scenario-error-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 6.5
scenario_error_plot
```
:::

Each box summarizes the test error rates of one method over the
`r n_reps` training sets, and the dashed line is the Bayes error rate, the
floor no classifier can beat on average.

- **Linear.** Logistic regression and LDA
`r if (pct(mean_error_of("Linear", "Logistic regression")) == pct(mean_error_of("Linear", "LDA"))) paste("both average", pct(mean_error_of("Linear", "LDA"))) else paste("average", pct(mean_error_of("Linear", "Logistic regression")), "and", pct(mean_error_of("Linear", "LDA")))`,
against a Bayes error rate of
`r pct(scenario_bayes$bayes[1])`. QDA, whose extra covariance parameters are
unnecessary here, averages `r pct(mean_error_of("Linear", "QDA"))`: with only
two features its variance penalty is small. Naive Bayes, which ignores the
correlation, averages `r pct(mean_error_of("Linear", "Naive Bayes"))`.
- **Quadratic.** QDA's average of `r pct(mean_error_of("Quadratic", "QDA"))`
beats logistic regression's
`r pct(mean_error_of("Quadratic", "Logistic regression"))` and LDA's
`r pct(mean_error_of("Quadratic", "LDA"))`: the linear methods' bias outweighs
QDA's variance. Naive Bayes, curved but missing the cross-product, falls between
at `r pct(mean_error_of("Quadratic", "Naive Bayes"))`.
- **Independent, $p = 10$.** Naive Bayes, whose assumption holds exactly,
averages `r pct(mean_error_of("Independent, p = 10", "Naive Bayes"))` with
`r 2 * 2 * 10` parameters. QDA's model also contains the truth, but it
estimates `r count_qda(2, 10)` parameters from `r 2 * n_train_k` observations and averages
`r pct(mean_error_of("Independent, p = 10", "QDA"))`. KNN, with ten
dimensions to search for neighbors, averages
`r pct(mean_error_of("Independent, p = 10", knn_sim_name))` with
$K = `r k_sim`$.
- **Non-linear.** No model-based method can represent a sine curve, and all
four average between
`r pct(min(sapply(method_levels[1:4], \(m) mean_error_of("Non-linear", m))))`
and
`r pct(max(sapply(method_levels[1:4], \(m) mean_error_of("Non-linear", m))))`.
KNN with $K = `r k_sim`$ averages
`r pct(mean_error_of("Non-linear", knn_sim_name))`. With $K = 1$ its variance
erases that advantage: it averages
`r pct(mean_error_of("Non-linear", "KNN, K = 1"))`, about the same as the
model-based methods.

In the quadratic, independent, and non-linear scenarios the best method is the
one whose assumptions are the most restrictive ones that still hold:
`r winner("Quadratic")` in the quadratic, `r winner("Independent, p = 10")` in
the independent, and `r winner("Non-linear")` in the non-linear scenario. In
the linear scenario both LDA and logistic regression are correct models, and
their averages differ by only
`r pp(abs(mean_error_of("Linear", "Logistic regression") - mean_error_of("Linear", "LDA")))`
percentage points, with
`r if (best_method("Linear") == "LDA") "LDA" else "logistic regression"` ahead:
with two features, LDA's extra assumption buys essentially nothing. None wins
in every scenario.

## Credit card default classification

**Question.** The previous lecture found LDA's shared-covariance assumption
strained in the credit card data, and balance and student status were
correlated within each class. Do QDA and naive Bayes, which change LDA's model in different directions,
classify new customers more accurately than LDA and logistic regression, and
how does KNN compare?

**Data.** `ISLR2::Default` records `r format(nrow(Default), big.mark = ",")`
customers. The response is `default`, with `Yes` the positive class. The
explanatory variables are `balance`, the monthly credit card balance in
dollars, and `student`. The features are `balance` and the indicator
$D = \mathrm{I}(\texttt{student} = \texttt{Yes})$.

### Training and test sets

The previous lecture's error rates were training error rates, optimistic
because each customer helped fit the model that classified it. Here, as in the
classification lecture's bias-variance script, the customers are split at
random into two halves. Every model is fit to the training half, and its
error rate, sensitivity, specificity, and AUC are computed on the test half,
whose customers the model never saw.

::: {.callout-note collapse="true"}
### Code
```{r default-split}
#| output: false
Default <- Default |>
  mutate(D = as.numeric(student == "Yes"))

# Its own seed, so the split does not depend on the simulation above.
set.seed(20261001)
train_id <- sample(nrow(Default), nrow(Default) / 2)
train <- Default[train_id, ] |>
  mutate(default_jit = as.numeric(default == "Yes") + runif(n(), -0.03, 0.03))
test  <- Default[-train_id, ]

n_defaults <- c(train = sum(train$default == "Yes"),
                test  = sum(test$default == "Yes"))
```
:::

Each half holds `r format(nrow(train), big.mark = ",")` customers, with
`r n_defaults["train"]` defaulters in the training set and
`r n_defaults["test"]` in the test set. A single split is one random draw: a
different split would give somewhat different test error rates, and averaging
over many splits is a topic of the resampling unit.

### Class summaries

QDA allows a different covariance matrix in each class, and naive Bayes
assumes balance and $D$ are independent within each class. The training data
show, before anything is fit, whether either matters.

::: {.callout-note collapse="true"}
### Code
```{r default-summaries}
#| output: false
class_summary <- train |>
  group_by(default) |>
  summarize(sd_balance = sd(balance),
            mean_balance_student = mean(balance[D == 1]),
            mean_balance_nonstudent = mean(balance[D == 0]),
            cor_balance_D = cor(balance, D),
            .groups = "drop")

summary_of <- function(k, col) class_summary[[col]][class_summary$default == k]

eda_plot <- ggplot(train, aes(x = balance, color = student,
                              linetype = student)) +
  geom_density(linewidth = 0.9, bounds = c(0, Inf), key_glyph = "path") +
  facet_wrap(~ default, ncol = 1, labeller = label_both, scales = "free_y") +
  scale_color_manual(values = class_colors) +
  labs(x = "Credit card balance ($)", y = "Density", color = "Student",
       linetype = "Student",
       title = "Training set: balance by student status within each class") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```
:::

```{r default-eda-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
eda_plot
```

The standard deviation of balance is
\$`r format(round(summary_of("No", "sd_balance")), big.mark = ",")` among
non-defaulters and
\$`r format(round(summary_of("Yes", "sd_balance")), big.mark = ",")` among
defaulters, a difference QDA can represent and LDA cannot.

Within each class, the student curve sits to the right of the non-student
curve. Among defaulters, students carry an average balance of
\$`r format(round(summary_of("Yes", "mean_balance_student")), big.mark = ",")`
against
\$`r format(round(summary_of("Yes", "mean_balance_nonstudent")), big.mark = ",")`
for non-students, and the within-class correlation between balance and $D$ is
`r sprintf("%.3f", summary_of("No", "cor_balance_D"))` among non-defaulters
and `r sprintf("%.3f", summary_of("Yes", "cor_balance_D"))` among defaulters.
Balance and student status are not independent within either class, so the
naive Bayes assumption fails.

### QDA fit

`MASS::qda()` takes the same formula as `MASS::lda()`:

```{r default-qda}
default_qda <- MASS::qda(default ~ balance + student, data = train)
default_qda
```

::: {.callout-note collapse="true"}
### Code
```{r default-qda-extract}
#| output: false
# QDA balance thresholds for each student group: the posterior crosses 0.5
# once inside the observed range and once far above it.
qda_posterior_at <- function(b, s) {
  nd <- tibble(balance = b, student = factor(s, levels = c("No", "Yes")))
  predict(default_qda, newdata = nd)$posterior[, "Yes"]
}
qda_thresholds <- tibble(student = c("No", "Yes")) |>
  mutate(lower = map_dbl(student, \(s) {
           uniroot(\(b) qda_posterior_at(b, s) - 0.5, c(1000, 2600))$root
         }),
         upper = map_dbl(student, \(s) {
           uniroot(\(b) qda_posterior_at(b, s) - 0.5, c(2700, 10000))$root
         }))
threshold_of <- function(s, col) qda_thresholds[[col]][qda_thresholds$student == s]
dollars <- function(v) paste0("\\$", format(round(v), big.mark = ","))
```
:::

The output reports the prior probabilities and the class means of balance and
of $D$, the same priors and means LDA reports. The covariance matrices are
not printed; they are the two class-specific sample covariance matrices of
$(\texttt{balance}, D)$.

Because non-defaulters have the larger balance variance given student status,
QDA's posterior probability of default eventually falls as balance grows: for
non-students it crosses $0.5$ at `r dollars(threshold_of("No", "lower"))` and
again at `r dollars(threshold_of("No", "upper"))` (for students,
`r dollars(threshold_of("Yes", "lower"))` and
`r dollars(threshold_of("Yes", "upper"))`), the second crossing far beyond the
largest training balance of `r dollars(max(train$balance))`.

### Naive Bayes fit

`e1071::naiveBayes()` models a numeric explanatory variable as Gaussian and a
factor with a probability for each level, so `balance` gets a Gaussian density
and `student` a Bernoulli one in each class:

```{r default-nb}
default_nb <- e1071::naiveBayes(default ~ balance + student, data = train)
default_nb
```

The **A-priori probabilities** are $\hat\pi_k$, the same as QDA's. Under
**Conditional probabilities**, the `balance` table gives each class's mean
(first column) and standard deviation (second column), the Gaussian
$\hat\mu_{kj}$ and $\hat\sigma_{kj}$. The `student` table gives
$\hat\theta_{kj}$, the proportion of students and non-students within each
class.

### Posterior probabilities

Logistic regression and LDA are refit to the same training half for
comparison. The table below gives each model's difference in log posterior
odds of default between a student and a non-student with the same balance.

::: {.callout-note collapse="true"}
### Code
```{r default-compare-fits}
#| output: false
default_lda <- MASS::lda(default ~ balance + student, data = train)
default_glm <- glm(default ~ balance + student, data = train,
                   family = "binomial")

posterior_all <- function(newdata) {
  tibble(`Logistic regression` = unname(predict(default_glm, newdata = newdata,
                                                type = "response")),
         LDA = predict(default_lda, newdata = newdata)$posterior[, "Yes"],
         QDA = predict(default_qda, newdata = newdata)$posterior[, "Yes"],
         `Naive Bayes` = predict(default_nb, newdata = newdata,
                                 type = "raw")[, "Yes"])
}

shift_balances <- c(1000, 1500, 2000)
shift_grid <- expand_grid(balance = shift_balances,
                          student = factor(c("No", "Yes")))
student_shift <- bind_cols(shift_grid, posterior_all(shift_grid)) |>
  pivot_longer(-c(balance, student), names_to = "model",
               values_to = "p_hat") |>
  mutate(log_odds = qlogis(p_hat)) |>
  select(-p_hat) |>
  pivot_wider(names_from = student, values_from = log_odds) |>
  mutate(shift = Yes - No)

shift_of <- function(m, b) {
  student_shift$shift[student_shift$model == m & student_shift$balance == b]
}

student_shift_table <- student_shift |>
  select(model, balance, shift) |>
  mutate(balance = paste0("Balance \\$", format(balance, big.mark = ",")),
         shift = sprintf("%.3f", shift)) |>
  pivot_wider(names_from = balance, values_from = shift) |>
  mutate(model = factor(model, levels = names(model_colors))) |>
  arrange(model) |>
  knitr::kable(col.names = c("Model", paste0("Balance \\$",
                                             format(shift_balances,
                                                    big.mark = ","))),
               align = "lccc")

posterior_grid <- expand_grid(balance = seq(0, 4500, length.out = 451),
                              student = factor(c("No", "Yes")))
posterior_curves <- bind_cols(posterior_grid, posterior_all(posterior_grid)) |>
  pivot_longer(-c(balance, student), names_to = "model",
               values_to = "p_hat") |>
  mutate(model = factor(model, levels = names(model_colors)))

posterior_plot <- ggplot(posterior_curves, aes(x = balance, y = p_hat)) +
  annotate("rect", xmin = max(train$balance), xmax = Inf, ymin = -Inf,
           ymax = Inf, fill = "grey85") +
  geom_point(data = train, aes(y = default_jit), alpha = 0.08, size = 0.7) +
  geom_hline(yintercept = 0.5, color = "grey60") +
  geom_line(aes(color = model, linetype = model), linewidth = 0.9) +
  facet_wrap(~ student, labeller = label_both) +
  scale_color_manual(values = model_colors) +
  scale_linetype_manual(values = model_linetypes) +
  scale_y_continuous(breaks = c(0, 0.5, 1)) +
  labs(x = "Credit card balance ($)", y = "P(default = Yes)",
       color = "Model", linetype = "Model",
       title = "Posterior probability of default; shading beyond the largest training balance")
```
:::

```{r student-shift-table}
#| echo: false
student_shift_table
```

```{r default-posterior-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 3.8
posterior_plot
```

Logistic regression, LDA, and QDA all give a student lower log posterior
odds of default than a non-student with the same balance: logistic
regression by `r sprintf("%.3f", -shift_of("Logistic regression", shift_balances[1]))`
and LDA by `r sprintf("%.3f", -shift_of("LDA", shift_balances[1]))` at every
balance, and QDA by an amount that depends on balance, from
`r sprintf("%.3f", -shift_of("QDA", shift_balances[1]))` at
`r dollars(shift_balances[1])` to
`r sprintf("%.3f", -shift_of("QDA", shift_balances[3]))` at
`r dollars(shift_balances[3])`, because QDA's
quadratic includes a product of balance and $D$. Naive Bayes reverses the
sign, adding `r sprintf("%.3f", shift_of("Naive Bayes", shift_balances[1]))` at every
balance. Its student term uses only the within-class proportions of students,
which are higher among defaulters, and cannot see that students are more often defaulters because they carry
higher balances, which the balance term already counts. This is the
confounding of the multiple logistic regression lecture, which conditional
independence assumes away.

On the probability scale, the four curves cross $0.5$ at similar balances,
close to the lower QDA thresholds above, with the naive Bayes curve shifted
toward lower balances for students and toward higher balances for
non-students, the two directions of its reversed student term. Above
that they separate. Logistic regression and LDA continue toward $1$, while
QDA and naive Bayes, whose balance terms are both quadratic because the two
classes' balance variances differ, level off below $1$ and then turn down in
the shaded region beyond the data.

### Test performance

::: {.callout-note collapse="true"}
### Code
```{r default-test}
#| output: false
k_default <- odd_above_sqrt(nrow(train))

# KNN on standardized features, with the training means and SDs used for both
# sets. Returns the vote proportion for "Yes" at each target row.
standardize <- function(d) {
  cbind((d$balance - mean(train$balance)) / sd(train$balance),
        (d$D - mean(train$D)) / sd(train$D))
}
knn_yes <- function(target, K) {
  fit <- class::knn(standardize(train), standardize(target), train$default,
                    k = K, prob = TRUE)
  ifelse(fit == "Yes", attr(fit, "prob"), 1 - attr(fit, "prob"))
}

knn_default_name <- paste0("KNN, K = ", k_default)
scores_train <- posterior_all(train) |>
  mutate(`KNN, K = 1` = knn_yes(train, 1),
         "{knn_default_name}" := knn_yes(train, k_default))
scores_test <- posterior_all(test) |>
  mutate(`KNN, K = 1` = knn_yes(test, 1),
         "{knn_default_name}" := knn_yes(test, k_default))

roc_curve <- function(score, positive) {
  ord  <- order(score, decreasing = TRUE)
  s    <- score[ord]
  pos  <- positive[ord]
  keep <- c(diff(s) != 0, TRUE)
  tibble(fpr = c(0, cumsum(!pos)[keep] / sum(!positive)),
         tpr = c(0, cumsum(pos)[keep] / sum(positive)))
}
auc_pairs <- function(score, positive) {
  s_pos <- score[positive]
  s_neg <- score[!positive]
  mean(outer(s_pos, s_neg, ">")) + 0.5 * mean(outer(s_pos, s_neg, "=="))
}

yes_train <- train$default == "Yes"
yes_test  <- test$default == "Yes"

performance <- tibble(model = names(scores_test)) |>
  mutate(train_error = map_dbl(model, \(m) {
           mean((scores_train[[m]] > 0.5) != yes_train)
         }),
         test_error = map_dbl(model, \(m) {
           mean((scores_test[[m]] > 0.5) != yes_test)
         }),
         sensitivity = map_dbl(model, \(m) {
           mean(scores_test[[m]][yes_test] > 0.5)
         }),
         specificity = map_dbl(model, \(m) {
           mean(scores_test[[m]][!yes_test] <= 0.5)
         }),
         auc = map_dbl(model, \(m) auc_pairs(scores_test[[m]], yes_test)))

perf_of <- function(m, col) performance[[col]][performance$model == m]

performance_table <- performance |>
  mutate(across(c(train_error, test_error, sensitivity, specificity),
                \(v) sprintf("%.2f%%", 100 * v)),
         auc = sprintf("%.4f", auc)) |>
  knitr::kable(col.names = c("Model", "Training error", "Test error",
                             "Test sensitivity", "Test specificity",
                             "Test AUC"),
               align = "lccccc")

roc_models <- c("Logistic regression", "LDA", "QDA", "Naive Bayes",
                knn_default_name)
roc_test <- map(roc_models, \(m) {
  roc_curve(scores_test[[m]], yes_test) |> mutate(model = m)
}) |>
  bind_rows() |>
  mutate(model = factor(model, levels = roc_models))

roc_colors    <- setNames(unname(model_colors), roc_models)
roc_linetypes <- setNames(unname(model_linetypes), roc_models)

roc_plot <- ggplot(roc_test, aes(x = fpr, y = tpr, color = model,
                                 linetype = model)) +
  geom_abline(intercept = 0, slope = 1, color = "grey60",
              linetype = "dashed") +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = roc_colors) +
  scale_linetype_manual(values = roc_linetypes) +
  coord_fixed() +
  labs(x = "False positive rate (1 - specificity)",
       y = "True positive rate (sensitivity)", color = "Model",
       linetype = "Model", title = "Test-set ROC curves, balance + student")
```
:::

KNN uses balance and $D$, each standardized by its training mean and standard
deviation (the binary $D$ like the quantitative balance), with $K = 1$ and
$K = `r k_default`$, the smallest odd integer above
$\sqrt{`r format(nrow(train), big.mark = "{,}")`}$. Every classifier is
evaluated at threshold $0.5$ (for KNN, a majority vote), with the training
error rate computed on the customers each model was fit to:

```{r performance-table}
#| echo: false
performance_table
```

```{r roc-figure}
#| echo: false
#| fig-width: 6.5
#| fig-height: 5.5
roc_plot
```

The training error rates are optimistic. KNN with $K = 1$ makes a training
error rate of `r sprintf("%.2f%%", 100 * perf_of("KNN, K = 1", "train_error"))`,
since each training customer is its own nearest neighbor, but a test error
rate of `r sprintf("%.2f%%", 100 * perf_of("KNN, K = 1", "test_error"))`,
`r if (perf_of("KNN, K = 1", "test_error") == max(performance$test_error)) "the worst in the table" else "far above its training error rate"`.

The four model-based classifiers have test error rates between
`r sprintf("%.2f%%", 100 * min(performance$test_error[1:4]))` and
`r sprintf("%.2f%%", 100 * max(performance$test_error[1:4]))`, against the
`r sprintf("%.2f%%", 100 * mean(yes_test))` of predicting `No` for every
test customer. They differ more in sensitivity, from
`r sprintf("%.1f%%", 100 * min(performance$sensitivity[1:4]))` to
`r sprintf("%.1f%%", 100 * max(performance$sensitivity[1:4]))`, which with
`r n_defaults["test"]` test defaulters is a difference of
`r round(n_defaults["test"] * (max(performance$sensitivity[1:4]) - min(performance$sensitivity[1:4])))`
customers. Their ROC curves are nearly indistinguishable, and their test AUCs
range only from `r sprintf("%.4f", min(performance$auc[1:4]))` to
`r sprintf("%.4f", max(performance$auc[1:4]))`. Naive Bayes, despite the
failed independence assumption and the wrong sign on student status, orders
the test customers almost as well as the others, with AUC
`r sprintf("%.4f", perf_of("Naive Bayes", "auc"))`: balance dominates every
model's ordering, and the student term moves a customer's log posterior
odds far less than the range of balance does. KNN with
$K = `r k_default`$ has the lowest AUC of the five curves,
`r sprintf("%.4f", perf_of(knn_default_name, "auc"))`.

**Conclusion.** Relaxing LDA's shared covariance (QDA) or replacing its joint
Gaussian with independent features (naive Bayes) makes no practical
difference to classifying credit card default: on the test set the four
model-based error rates lie within
`r sprintf("%.2f", 100 * (max(performance$test_error[1:4]) - min(performance$test_error[1:4])))`
percentage points of each other, and their AUCs within
`r sprintf("%.3f", max(performance$auc[1:4]) - min(performance$auc[1:4]))`.
With two features and a boundary close
to a single balance threshold, the linear methods are not biased enough for
the extra flexibility to pay. The curvature of QDA and naive Bayes matters
only at the largest balances, where they level off below $1$, and naive Bayes
misstates the effect of student status while still
ordering customers almost as well as the other three, because balance carries
most of the information. KNN with $K = `r k_default`$ has the lowest AUC of
the five curves, and with $K = 1$ its
`r sprintf("%.2f%%", 100 * perf_of("KNN, K = 1", "train_error"))` training
error rate is no guide to its test error rate.

## Conclusion

QDA is the generative classifier with Gaussian class-conditional densities and
a covariance matrix $\Sigma_k$ for each class. The quadratic term
$-\frac{1}{2}x^\top\Sigma_k^{-1}x$ and the term
$-\frac{1}{2}\log\left|\Sigma_k\right|$ no longer cancel between classes, so
the discriminant functions and the log posterior odds are quadratic in $x$,
the decision boundary is a quadratic curve, and with one feature the class
with the larger variance is assigned in both tails. Each $\Sigma_k$ is
estimated by its class's sample covariance matrix, which requires $n_k > p$.

The choice between LDA and QDA is a bias-variance tradeoff. With $K$ classes,
QDA estimates $K p(p+1)/2$ covariance parameters against LDA's $p(p+1)/2$, so it removes
LDA's bias when the class covariances differ at the cost of variance that
grows with $p$ and shrinks with $n$.

Naive Bayes assumes the features are independent within each class, so
$f_k(x) = \prod_j f_{kj}(x_j)$, each one-feature density estimated on its own
and of whatever type suits the feature. The log posterior odds are a sum of
one-feature terms, with no interactions. Gaussian naive Bayes is LDA or QDA
with a diagonal covariance matrix. With class-specific variances it estimates
$2Kp$ parameters, sidestepping the curse of dimensionality that a joint
density faces. Its bias, which comes from ignoring within-class dependence,
need not change its classifications: it agrees with the Bayes classifier
wherever its estimated posterior probability falls on the same side of the
threshold as the true one.

Logistic regression and LDA produce linear boundaries, QDA and naive Bayes
quadratic ones, and KNN boundaries of any shape. Which classifier has the
lowest test error rate depends on how closely its assumptions match the
data-generating process and on the sizes of $n$ and $p$; no method is best
in every setting. Comparing classifiers requires test data the models were
not fit to, since training error rates reward flexibility.

The next lecture steps back from classification to a framework that contains
linear and logistic regression as special cases. Generalized linear models
pair a distribution for the response with a link function for its mean, and
Poisson regression, the generalized linear model for counts, is the first new
member of the family.
