# Quadratic Discriminant Analysis and Naive Bayes

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
set.seed(20261001)

# MASS, e1071, and class are called with :: rather than attached, because
# attaching MASS masks dplyr::select().
class_colors <- c("#0072B2", "#D55E00")
model_colors <- c("Logistic regression" = "#CC79A7", "LDA" = "#009E73",
                  "QDA" = "#E69F00", "Naive Bayes" = "#56B4E9",
                  "KNN" = "black")
model_linetypes <- c("Logistic regression" = "dashed", "LDA" = "solid",
                     "QDA" = "dotted", "Naive Bayes" = "dotdash",
                     "KNN" = "longdash")

# Density of N_2(mu, Sigma) at the rows (x1, x2), and the two-class Gaussian
# log posterior odds of class 2 with equal priors.
dmvnorm2 <- function(x1, x2, mu, Sigma) {
  d <- cbind(x1 - mu[1], x2 - mu[2])
  q <- rowSums((d %*% solve(Sigma)) * d)
  exp(-q / 2) / (2 * pi * sqrt(det(Sigma)))
}
log_odds_gauss <- function(x1, x2, mu1, S1, mu2, S2) {
  log(dmvnorm2(x1, x2, mu2, S2)) - log(dmvnorm2(x1, x2, mu1, S1))
}

# Renders a number in LaTeX scientific notation for use inside $...$.
sci <- function(x, digits = 1) {
  if (x == 0) return("0")
  e <- floor(log10(abs(x)))
  m <- signif(x / 10^e, digits)
  if (abs(m) >= 10) {
    m <- m / 10
    e <- e + 1
  }
  sprintf("%s \\times 10^{%d}", format(m), e)
}

# Dollar amounts for prose, with the "$" escaped for markdown: "\$1,234".
dollar_text <- function(v) paste0("\\$", format(round(v), big.mark = ","))
# Axis labels such as $60,000, without the scales package.
dollar_labels <- function(v) {
  ifelse(is.na(v), NA,
         paste0("$", format(v, big.mark = ",", scientific = FALSE, trim = TRUE)))
}
pct2 <- function(v) sprintf("%.2f%%", 100 * v)

# Smallest odd integer greater than sqrt(n): a fixed rule of thumb for the
# number of neighbors that does not look at any test data. Odd, so a
# two-class vote cannot tie.
odd_above_sqrt <- function(n) {
  k <- floor(sqrt(n))
  if (k %% 2 == 0) k + 1 else k + 2
}
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **derive** the QDA discriminant function $\delta_k(x)$ from Gaussian
class-conditional densities with class-specific covariance matrices
$\Sigma_k$, and **show** that the resulting decision boundary is quadratic in
$x$, with up to two boundary points when $p = 1$,
- **count** the mean and covariance parameters that LDA, QDA, and Gaussian
naive Bayes estimate, and **use** the counts to **explain** through the
bias-variance tradeoff when each should have the lowest test error rate,
- **state** the naive Bayes assumption that the features are conditionally
independent given the class, **write** $f_k(x)$ as a product of one-feature
densities of possibly different types, and **show** that the log posterior
odds are a sum of one-feature terms,
- **identify** Gaussian naive Bayes as LDA or QDA with a diagonal covariance
matrix,
- **compare** logistic regression, LDA, QDA, naive Bayes, and KNN by the shape
of their decision boundaries, their assumptions, and how they are affected by
$n$ and $p$, and **evaluate** which should perform best in a described
setting, and
- **fit** QDA with `MASS::qda()` and naive Bayes with `e1071::naiveBayes()`,
**interpret** their estimates, and **compare** classifiers by test error
rate, sensitivity, specificity, and AUC on a held-out test set.
:::

## Generative classifiers

### Credit card default data

**Question.** Do QDA and naive Bayes, which change LDA's model of the features
in different directions, classify credit card customers more accurately than
LDA and logistic regression, and how does KNN compare?

These customers were analyzed in the logistic regression lectures and, with
`balance` and `income` as the features, in the previous lecture, which modeled
each class's features by a Gaussian density with one covariance matrix
shared by both classes.

**Data.** `ISLR2::Default` is a simulated data set of
`r format(nrow(Default), big.mark = ",")` credit card customers. The response
is `default`, with `Yes` the positive class. The explanatory variables are
`balance`, the customer's monthly credit card balance in dollars, and
`income`, the customer's income in dollars; the features are the explanatory
variables themselves. `student` (Yes/No) is used only in the Student strata
section, to split the customers.

The previous lecture's error rates were training error rates, optimistic on
average because each customer helped fit the model that classified it. Here,
as in the classification lecture's bias-variance script, the customers are
first split at random into halves: every model in this lecture is fit to the
training half and evaluated on the test half, whose customers it never saw.

::: {.callout-note collapse="true"}
### Code
```{r default-split}
#| output: false
# Its own seed, so the split does not depend on anything drawn before it.
set.seed(20261001)
train_id <- sample(nrow(Default), nrow(Default) / 2)
train <- Default[train_id, ]
test  <- Default[-train_id, ]

n_defaults <- c(train = sum(train$default == "Yes"),
                test  = sum(test$default == "Yes"))
```
:::

Each half holds `r format(nrow(train), big.mark = ",")` customers, with
`r n_defaults["train"]` defaulters in the training half and
`r n_defaults["test"]` in the test half. A different split would give somewhat
different test error rates; averaging over many splits is a topic of the
resampling unit.

The figure below plots income against balance for the training customers,
defaulters and non-defaulters distinguished by color and shape, with each
class's 95% normal-theory ellipse (solid for non-defaulters, dashed for
defaulters), the contour of a Gaussian density with that class's sample mean
and covariance matrix.

::: {.callout-note collapse="true"}
### Code
```{r default-summaries}
#| output: false
class_summary <- train |>
  group_by(default) |>
  summarize(sd_balance = sd(balance), sd_income = sd(income),
            cor_balance_income = cor(balance, income), .groups = "drop")
summary_of <- function(k, col) class_summary[[col]][class_summary$default == k]
student_income <- function(s) median(train$income[train$student == s])

# Defaulters are drawn last so they are not hidden under non-defaulters.
scatter_points <- list(
  geom_point(data = arrange(train, default),
             aes(x = balance, y = income, color = default, shape = default),
             alpha = 0.35, size = 0.8),
  scale_color_manual(values = class_colors),
  scale_shape_manual(values = c(16, 17)),
  scale_x_continuous(labels = dollar_labels),
  scale_y_continuous(labels = dollar_labels),
  labs(x = "Credit card balance ($)", y = "Income ($)", color = "Default",
       shape = "Default")
)
eda_scatter <- ggplot() +
  scatter_points +
  stat_ellipse(data = train, aes(x = balance, y = income, color = default,
                                 linetype = default),
               type = "norm", level = 0.95, linewidth = 0.9) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  labs(linetype = "Default",
       title = "Training set, with each class's 95% normal-theory ellipse")
```
:::

```{r default-eda-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
eda_scatter
```

The defaulters' ellipse is centered at higher balances and is narrower in the
balance direction: the standard deviation of balance is
`r dollar_text(summary_of("No", "sd_balance"))` among non-defaulters and
`r dollar_text(summary_of("Yes", "sd_balance"))` among defaulters. Income is
about equally spread in the two classes, with standard deviations of
`r dollar_text(summary_of("No", "sd_income"))` and
`r dollar_text(summary_of("Yes", "sd_income"))`.

Within each class balance and income are weakly correlated, with
$r = `r sprintf("%.3f", summary_of("No", "cor_balance_income"))`$ among
non-defaulters and
$r = `r sprintf("%.3f", summary_of("Yes", "cor_balance_income"))`$ among
defaulters, so both ellipses are only slightly tilted. Among non-defaulters
the points are densest in two horizontal bands, the bimodal income of the
previous lecture's histograms. The lower band is mostly students: in the
training half students' median income is
`r dollar_text(student_income("Yes"))`, against
`r dollar_text(student_income("No"))` for non-students. No Gaussian density
has two modes.

### Bayes' theorem and LDA

The previous lecture built a classifier from the prior probabilities
$\pi_k = P(Y = k)$ and the class-conditional densities $f_k(x)$ of the
features through Bayes' theorem,

$$p_k(x) = P(Y = k \mid X = x) = \frac{\pi_k f_k(x)}{\sum_{l=1}^K \pi_l f_l(x)},$$

and assigned each observation to the class with the largest posterior
probability. Every generative classifier in this course estimates $\pi_k$ by
the class proportion $n_k/n$. They differ only in the model assumed for
$f_k(x)$.

LDA assumed $X \mid Y = k \sim N_p\left(\mu_k, \Sigma\right)$, with one
covariance matrix $\Sigma$ shared by every class. That is what made the
quadratic term in $x$ cancel and the boundary linear. It would give the two
ellipses of the scatterplot the same shape, but the defaulters' is the
narrower in balance.

This lecture changes the model for $f_k(x)$ in two directions. Quadratic
discriminant analysis gives each class its own covariance matrix. Naive Bayes
drops the multivariate Gaussian and treats the features as independent within
each class.

## Quadratic discriminant analysis

The scatterplot's two ellipses differ in width and slightly in tilt. A model
that lets each class's cloud have its own spread and tilt describes them
directly.

### Class-specific covariance matrices

Quadratic discriminant analysis (QDA) assumes the statistical model

$$P(Y = k) = \pi_k, \qquad X \mid Y = k \sim N_p\left(\mu_k, \Sigma_k\right), \qquad k = 1, \ldots, K,$$

where

- $\mu_k = E[X \mid Y = k]$ is the $p$-vector of feature means in class $k$,
as in LDA, and
- $\Sigma_k = Cov[X \mid Y = k]$ is the $p \times p$ covariance matrix of the
features **in class $k$**, symmetric and positive definite, and free to differ
from class to class.

The class-conditional density is therefore

$$f_k(x) = \frac{1}{(2\pi)^{p/2}\left|\Sigma_k\right|^{1/2}} \exp\left[-\frac{1}{2}\left(x - \mu_k\right)^\top \Sigma_k^{-1} \left(x - \mu_k\right)\right],$$

where $\left|\Sigma_k\right|$ is the determinant of $\Sigma_k$. Each class's
contours are still ellipses centered at $\mu_k$, but now each class's
ellipses have their own size, shape, and orientation, as the two ellipses of
the scatterplot do.

### Class-specific covariance estimates

The priors and means are estimated as in LDA, $\hat\pi_k = n_k/n$ and
$\hat\mu_k = \frac{1}{n_k}\sum_{i:\, y_i = k} x_i$. Each covariance matrix is
estimated from its own class alone,

$$\hat\Sigma_k = \frac{1}{n_k - 1}\sum_{i:\, y_i = k}\left(x_i - \hat\mu_k\right)\left(x_i - \hat\mu_k\right)^\top,$$

where

- $x_i$ is the $p$-vector of feature values of observation $i$, as in the
previous lecture, and
- $\hat\Sigma_k$ is the **class-$k$ sample covariance matrix**, whose divisor
$n_k - 1$ accounts for the one estimated mean.

`MASS::qda()` uses this divisor (maximum likelihood would use $n_k$), and
$\hat\Sigma_k$, whose rank is at most $n_k - 1$, is invertible only if
$n_k > p$. It takes the same formula as `MASS::lda()`; fitting `default` on
`balance` and `income` in the training half:

```{r default-qda}
default_qda <- MASS::qda(default ~ balance + income, data = train)
default_qda
```

::: {.callout-note collapse="true"}
### Code
```{r qda-estimates}
#| output: false
qda_par <- map(c(No = "No", Yes = "Yes"), \(k) {
  X <- as.matrix(train[train$default == k, c("balance", "income")])
  list(prior = mean(train$default == k), mu = colMeans(X), Sigma = cov(X))
})
# A 2 x 2 matrix as LaTeX, entries rounded with LaTeX-safe thousands commas.
matrix_tex <- function(S) {
  e <- format(round(S), big.mark = "{,}", trim = TRUE)
  sprintf("\\begin{pmatrix} %s & %s \\\\ %s & %s \\end{pmatrix}",
          e[1, 1], e[1, 2], e[2, 1], e[2, 2])
}
```
:::

The output reports the prior probabilities and the class means of balance and
income, the same priors and means `MASS::lda()` reports for these training
customers; the priors differ slightly from the previous lecture's
all-customer fit, $\hat\pi_{\text{Yes}} = `r sprintf("%.4f", mean(Default$default == "Yes"))`$.
The covariance matrices are not printed. They are each class's sample
covariance matrix of (`balance`, `income`), in squared dollars,

$$\hat\Sigma_{\text{No}} = `r matrix_tex(qda_par$No$Sigma)`, \qquad \hat\Sigma_{\text{Yes}} = `r matrix_tex(qda_par$Yes$Sigma)`,$$

whose square-rooted diagonals and correlations are the standard deviations
and correlations quoted with the scatterplot, and whose 95% contours are its
ellipses.

### Quadratic discriminant

As in LDA, the class with the largest posterior probability is the class with
the largest $\log\left(\pi_k f_k(x)\right)$,

$$\log\left(\pi_k f_k(x)\right) = \log \pi_k - \frac{p}{2}\log(2\pi) - \frac{1}{2}\log\left|\Sigma_k\right| - \frac{1}{2}\left(x - \mu_k\right)^\top \Sigma_k^{-1}\left(x - \mu_k\right).$$

Only the term $-\frac{p}{2}\log(2\pi)$ is common to every class. Dropping it
gives the **quadratic discriminant function**

$$\delta_k(x) = -\frac{1}{2}x^\top \Sigma_k^{-1} x + x^\top \Sigma_k^{-1}\mu_k - \frac{1}{2}\mu_k^\top \Sigma_k^{-1}\mu_k - \frac{1}{2}\log\left|\Sigma_k\right| + \log \pi_k,$$

after expanding the quadratic form as in the previous lecture. Two terms that
cancelled in LDA survive here:

- $-\frac{1}{2}x^\top \Sigma_k^{-1} x$ is quadratic in $x$ and carries a
different matrix in each class, and
- $-\frac{1}{2}\log\left|\Sigma_k\right|$ penalizes a class whose density is
spread over a larger volume.

When every $\Sigma_k$ equals a common $\Sigma$, both terms are the same in
every class, drop out, and $\delta_k(x)$ reduces to the LDA discriminant.
Because only $-\frac{p}{2}\log(2\pi)$, common to every class, was dropped,
the posterior probabilities are recovered as before,
$p_k(x) = e^{\delta_k(x)} / \sum_l e^{\delta_l(x)}$.

With $K = 2$ classes the log posterior odds of class $2$ are

$$\log\left(\frac{p_2(x)}{1 - p_2(x)}\right) = \delta_2(x) - \delta_1(x) = \beta_0 + x^\top\left(\Sigma_2^{-1}\mu_2 - \Sigma_1^{-1}\mu_1\right) - \frac{1}{2}x^\top \left(\Sigma_2^{-1} - \Sigma_1^{-1}\right) x,$$

where

- $\beta_0 = \log(\pi_2/\pi_1) - \frac{1}{2}\left(\mu_2^\top \Sigma_2^{-1}\mu_2 - \mu_1^\top \Sigma_1^{-1}\mu_1\right) - \frac{1}{2}\log\left(\left|\Sigma_2\right| / \left|\Sigma_1\right|\right)$
is the intercept, collecting the terms that do not involve $x$.

The log posterior odds are a quadratic function of the features: an
intercept, $p$ linear terms $x_j$, where $x_j$ is the $j$th entry of $x$, and
$p(p+1)/2$ squares $x_j^2$ and cross-products $x_j x_l$, with coefficients
determined by the $\mu_k$ and $\Sigma_k$.

::: {.callout-note collapse="true"}
### Code
```{r qda-discriminant}
#| output: false
S_no_inv  <- solve(qda_par$No$Sigma)
S_yes_inv <- solve(qda_par$Yes$Sigma)
m_no  <- qda_par$No$mu
m_yes <- qda_par$Yes$mu
A_qda <- -0.5 * (S_yes_inv - S_no_inv)
b_qda <- drop(S_yes_inv %*% m_yes - S_no_inv %*% m_no)
c_qda <- log(qda_par$Yes$prior / qda_par$No$prior) -
  0.5 * (sum(m_yes * (S_yes_inv %*% m_yes)) - sum(m_no * (S_no_inv %*% m_no))) -
  0.5 * log(det(qda_par$Yes$Sigma) / det(qda_par$No$Sigma))
balance_sq_coef <- A_qda[1, 1]

# Log posterior odds of Yes from the coefficients, checked against predict().
X_train <- as.matrix(train[c("balance", "income")])
qda_log_odds <- c_qda + drop(X_train %*% b_qda) + rowSums((X_train %*% A_qda) * X_train)
max_diff_qda <- max(abs(plogis(qda_log_odds) -
                        predict(default_qda)$posterior[, "Yes"]))

# Income in thousands of dollars, so the coefficients have readable sizes.
qda_coef_table <- tibble(
  term = c("Intercept $\\hat\\beta_0$", "`balance`", "`income`, per \\$1,000",
           "`balance`$^2$", "`balance` $\\times$ `income`, per \\$1,000",
           "`income`$^2$, per (\\$1,000)$^2$"),
  value = c(c_qda, b_qda[1], 1000 * b_qda[2], A_qda[1, 1],
            2 * 1000 * A_qda[1, 2], 1e6 * A_qda[2, 2])
) |>
  mutate(value = map_chr(value, \(v) paste0("$", sci(v, 3), "$"))) |>
  knitr::kable(col.names = c("Term", "Coefficient"), align = "lc")
```
:::

For the credit card fit, the table gives the coefficients of QDA's estimated
log posterior odds of default, with income terms per \$1,000 of income.

```{r qda-discriminant-table}
#| echo: false
qda_coef_table
```

The intercept is the log posterior odds of default for a customer with no
balance and no income, an extrapolation, since the lowest recorded income is
`r dollar_text(min(Default$income))`. With squared and cross-product terms, no
single coefficient gives a feature's effect: as in the polynomial logistic
regressions of the flexible logistic regression lecture, the effect of
balance depends on the balance and income at which it is evaluated.
Applying the logistic function to these log posterior odds reproduces
`predict()`'s posterior probabilities to within $`r sci(max_diff_qda)`$.

### Quadratic decision boundary

The decision boundary, where the log posterior odds are zero, is a conic
section when $p = 2$ instead of a line. With one feature, writing
$\sigma_k^2$ for the class-$k$ variance, the log posterior odds are

$$\log\left(\frac{p_2(x)}{1 - p_2(x)}\right) = \frac{1}{2}\left(\frac{1}{\sigma_1^2} - \frac{1}{\sigma_2^2}\right)x^2 + \left(\frac{\mu_2}{\sigma_2^2} - \frac{\mu_1}{\sigma_1^2}\right)x + \frac{\mu_1^2}{2\sigma_1^2} - \frac{\mu_2^2}{2\sigma_2^2} + \log\frac{\sigma_1}{\sigma_2} + \log\frac{\pi_2}{\pi_1}.$$

A quadratic in $x$ has up to two roots, so the boundary is up to two points.
If $\sigma_1 > \sigma_2$, the coefficient of $x^2$ is negative, the log
posterior odds fall without bound in both directions, and class $1$, the
class with the larger variance, is assigned in both tails. For the credit
card customers class $1$ is `No`, whose balances have the larger standard
deviation.

LDA (`MASS::lda()`) is refit to the same training half for contrast. The
figure below draws the QDA boundary (orange, dotted) and the LDA boundary
(green, solid) over the training customers, non-defaulters as grey circles
and defaulters as black triangles, with balances beyond the largest training
balance shaded.

::: {.callout-note collapse="true"}
### Code
```{r qda-boundary}
#| output: false
default_lda <- MASS::lda(default ~ balance + income, data = train)

# Each model's posterior probability of default; a model's object is looked
# up only when that model is requested, so later models can be added.
posterior_of <- function(model, d) {
  switch(model,
         "Logistic regression" = unname(predict(default_glm, newdata = d,
                                                type = "response")),
         LDA = predict(default_lda, newdata = d)$posterior[, "Yes"],
         QDA = predict(default_qda, newdata = d)$posterior[, "Yes"],
         "Naive Bayes" = predict(default_nb, newdata = d, type = "raw")[, "Yes"])
}

boundary_grid <- expand_grid(
  balance = seq(0, 4500, length.out = 181),
  income = seq(min(Default$income), max(Default$income), length.out = 121))

# Training customers under the 0.5 contour of each requested model, with
# balances beyond the largest training balance shaded.
boundary_figure <- function(models, title) {
  curves <- map(models, \(m) {
    mutate(boundary_grid, model = m, p_hat = posterior_of(m, boundary_grid))
  }) |>
    list_rbind() |>
    mutate(model = factor(model, levels = names(model_colors)))
  # Points are filled grey or black by default status, so that color is free
  # to identify the models' boundaries.
  ggplot(mapping = aes(x = balance, y = income)) +
    annotate("rect", xmin = max(train$balance), xmax = Inf, ymin = -Inf,
             ymax = Inf, fill = "grey92") +
    geom_point(data = arrange(train, default),
               aes(shape = default, fill = default), color = "grey40",
               stroke = 0.2, size = 1.1, alpha = 0.6) +
    geom_contour(data = curves, aes(z = p_hat, color = model, linetype = model),
                 breaks = 0.5, linewidth = 0.9) +
    scale_shape_manual(values = c(21, 24)) +
    scale_fill_manual(values = c("grey70", "black")) +
    scale_color_manual(values = model_colors) +
    scale_linetype_manual(values = model_linetypes) +
    scale_x_continuous(labels = dollar_labels) +
    scale_y_continuous(labels = dollar_labels) +
    labs(x = "Credit card balance ($)", y = "Income ($)", shape = "Default",
         fill = "Default", color = "Model", linetype = "Model", title = title)
}
qda_boundary_plot <- boundary_figure(
  c("LDA", "QDA"), "QDA and LDA boundaries; shading beyond the largest training balance")
```
:::

```{r qda-boundary-figure}
#| echo: false
#| fig-width: 7.5
#| fig-height: 4.5
qda_boundary_plot
```

Within the training balances the QDA boundary runs close to LDA's nearly
vertical line, so income barely moves either model's balance threshold there.
Beyond them QDA's boundary has a second branch, where its posterior
probability of default falls back below $0.5$.

The second branch is easiest to see with one feature. Fitting QDA to balance
alone, the first tab below plots the prior-weighted fitted densities
$\hat\pi_k \hat f_k(x)$ with the class-specific standard deviations, and the
second the two estimated discriminant functions $\hat\delta_k(x)$. Dotted
lines mark the boundary points, and balances beyond the largest training
balance are shaded.

::: {.callout-note collapse="true"}
### Code
```{r qda-one-feature}
#| output: false
one_feature <- train |>
  group_by(default) |>
  summarize(prior = n() / nrow(train), mu = mean(balance), sd = sd(balance),
            .groups = "drop")
of_class <- function(k, col) one_feature[[col]][one_feature$default == k]

# Boundary points: roots of the one-feature log posterior odds of Yes.
m1 <- of_class("No", "mu");  s1 <- of_class("No", "sd")
m2 <- of_class("Yes", "mu"); s2 <- of_class("Yes", "sd")
log_prior_odds <- log(of_class("Yes", "prior") / of_class("No", "prior"))
roots_qda_1 <- sort(Re(polyroot(c(
  m1^2 / (2 * s1^2) - m2^2 / (2 * s2^2) + log(s1 / s2) + log_prior_odds,
  m2 / s2^2 - m1 / s1^2,
  (1 / s1^2 - 1 / s2^2) / 2))))

balance_grid_1 <- tibble(balance = seq(0, 4500, length.out = 600))
one_feature_curves <- one_feature |>
  cross_join(balance_grid_1) |>
  mutate(weighted = prior * dnorm(balance, mu, sd),
         delta = log(prior) - log(sd) - (balance - mu)^2 / (2 * sd^2))
one_feature_plot <- function(y, y_label, title) {
  ggplot(one_feature_curves,
         aes(x = balance, y = .data[[y]], color = default, linetype = default)) +
    annotate("rect", xmin = max(train$balance), xmax = Inf, ymin = -Inf,
             ymax = Inf, fill = "grey92") +
    geom_line(linewidth = 0.9) +
    geom_vline(xintercept = roots_qda_1, linetype = "dotted") +
    scale_color_manual(values = class_colors) +
    scale_x_continuous(labels = dollar_labels) +
    labs(x = "Credit card balance ($)", y = y_label, color = "Default",
         linetype = "Default", title = title)
}
qda_weighted_1 <- one_feature_plot(
  "weighted", expression(hat(pi)[k] * hat(f)[k](x)),
  expression(paste("QDA prior-weighted densities, class-specific ", hat(sigma)[k])))
# Vertical range from the curves' maximum down past the second crossing.
delta_at <- function(b) {
  log(of_class("No", "prior")) - log(s1) - (b - m1)^2 / (2 * s1^2)
}
delta_range <- c(delta_at(roots_qda_1[2]) - 8, max(one_feature_curves$delta) + 1)
qda_delta_1 <- one_feature_plot("delta", expression(hat(delta)[k](x)),
                                "QDA discriminant functions") +
  coord_cartesian(ylim = delta_range)
```
:::

::: {.panel-tabset}
#### Prior-weighted densities
```{r qda-one-qda-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
qda_weighted_1
```

#### Discriminant functions
```{r qda-one-delta-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
qda_delta_1
```
:::

The weighted densities cross at `r dollar_text(roots_qda_1[1])` and again at
`r dollar_text(roots_qda_1[2])`, where both are too small to see; the
discriminant functions, two downward parabolas, visibly cross a second
time. Past that point the wider non-defaulter
parabola is the higher one again, so `No` is assigned in both tails.

::: {.callout-note collapse="true"}
### Code
```{r qda-crossings}
#| output: false
# Balances at which a model's two-feature posterior crosses 0.5 at the
# training income quartiles.
income_levels <- quantile(train$income, c(0.25, 0.5, 0.75))
crossings <- function(model) {
  gap <- function(b, inc) posterior_of(model, tibble(balance = b, income = inc)) - 0.5
  map(income_levels, \(inc) {
    b <- seq(0, 10000, by = 10)
    j <- which(diff(sign(gap(b, inc))) != 0)
    map_dbl(j, \(i) uniroot(gap, b[c(i, i + 1)], inc = inc)$root)
  })
}
qda_cross <- crossings("QDA")
cross_at <- function(cr, i) map_dbl(cr, i)
# The prose describes two crossings per income quartile, the second beyond
# the training data, and a negative balance^2 coefficient.
stopifnot(balance_sq_coef < 0, all(lengths(qda_cross) == 2),
          all(cross_at(qda_cross, 2) > max(train$balance)))
```
:::

With both features the same happens at every income. The coefficient of
$\texttt{balance}^2$ in the table above is negative, because balance varies
less among defaulters at a given income, so QDA's posterior probability of
default eventually falls as balance grows. At the lower quartile, median, and
upper quartile of training income, `r dollar_text(income_levels[1])`,
`r dollar_text(income_levels[2])`, and `r dollar_text(income_levels[3])`,
QDA's posterior crosses $0.5$ at balances of
`r dollar_text(cross_at(qda_cross, 1)[1])`,
`r dollar_text(cross_at(qda_cross, 1)[2])`, and
`r dollar_text(cross_at(qda_cross, 1)[3])` and again at
`r dollar_text(cross_at(qda_cross, 2)[1])`,
`r dollar_text(cross_at(qda_cross, 2)[2])`, and
`r dollar_text(cross_at(qda_cross, 2)[3])`, beyond the largest training
balance, `r dollar_text(max(train$balance))`. The second branch is an
extrapolation of the Gaussian tails`r if (min(cross_at(qda_cross, 2)) > max(Default$balance)) ": no customer in either half has a balance that large" else ""`.

### Bias-variance tradeoff

QDA's flexibility costs parameters. Not counting the $K - 1$ free priors,
which every generative classifier estimates (the previous lecture's count of
8 for LDA with balance and income included its one free prior),

$$\text{LDA: } Kp + \frac{p(p+1)}{2}, \qquad \text{QDA: } Kp + K\,\frac{p(p+1)}{2},$$

where

- $Kp$ counts the entries of the $K$ mean vectors, and
- $p(p+1)/2$ counts the distinct entries of one symmetric $p \times p$
covariance matrix: $p$ variances and $p(p-1)/2$ covariances.

::: {.callout-note collapse="true"}
### Code
```{r parameter-counts}
#| output: false
count_lda <- function(K, p) K * p + p * (p + 1) / 2
count_qda <- function(K, p) K * p + K * p * (p + 1) / 2

parameter_table <- tibble(p = c(1, 2, 5, 10, 50)) |>
  mutate(LDA = count_lda(2, p), QDA = count_qda(2, p),
         across(c(LDA, QDA), \(v) format(v, big.mark = ","))) |>
  knitr::kable(col.names = c("Features $p$", "LDA", "QDA"), align = "ccc")
```
:::

With $K = 2$ classes:

```{r parameter-table}
#| echo: false
parameter_table
```

QDA's count grows like $Kp^2/2$, against LDA's $p^2/2$. When the
$\Sigma_k$ really differ, LDA's shared $\Sigma$ makes its boundary the wrong
shape however much data it sees: bias. QDA's boundary has the right shape, but
each of its many covariance estimates comes from one class's observations and
varies from one training set to the next: variance. For the credit card
customers, QDA estimates the defaulters' `r 2 * (2 + 1) / 2`
covariance entries from the `r n_defaults["train"]` training defaulters
alone, while LDA's pooled covariance uses all
`r format(nrow(train), big.mark = ",")` training customers.

The figure below shows the tradeoff over repeated training sets, which the
credit card data cannot supply, so it uses simulated data: two classes with
known Gaussian densities, class 1 with mean $(0, 0)^\top$ and positively
correlated features and class 2 with mean $(1.5, 0.5)^\top$, larger
variances, and negatively correlated features, with equal priors. Each panel
overlays the boundaries of LDA (left) or QDA (right) fit to each of repeated
training sets (thin grey) on the Bayes decision boundary (thick black, long
dashes), the QDA boundary computed from the true parameters. The first tab
uses small training sets and the second large ones.

::: {.callout-note collapse="true"}
### Code
```{r boundary-variability}
#| output: false
mu_a    <- c(0, 0)
mu_b    <- c(1.5, 0.5)
Sigma_a <- matrix(c(1, 0.5, 0.5, 1), nrow = 2)
Sigma_b <- matrix(c(3, -0.6, -0.6, 2), nrow = 2)

rmvnorm2 <- function(n, mu, Sigma) {
  sweep(matrix(rnorm(n * 2), n) %*% chol(Sigma), 2, mu, "+")
}
draw_quadratic <- function(n_k) {
  X <- rbind(rmvnorm2(n_k, mu_a, Sigma_a), rmvnorm2(n_k, mu_b, Sigma_b))
  tibble(y = factor(rep(c("k = 1", "k = 2"), each = n_k)),
         x1 = X[, 1], x2 = X[, 2])
}

grid_fit <- expand_grid(x1 = seq(-3.5, 5.5, length.out = 120),
                        x2 = seq(-3.5, 4, length.out = 110))
bayes_grid <- expand_grid(x1 = seq(-3.5, 5.5, length.out = 220),
                          x2 = seq(-3.5, 4, length.out = 200)) |>
  mutate(qda = log_odds_gauss(x1, x2, mu_a, Sigma_a, mu_b, Sigma_b))

n_k_small <- 25
n_k_large <- 250
n_boundary_reps <- 20

fitted_boundaries <- expand_grid(n_k = c(n_k_small, n_k_large),
                                 rep = seq_len(n_boundary_reps)) |>
  mutate(data = map(n_k, draw_quadratic),
         LDA = map(data, \(d) {
           predict(MASS::lda(y ~ x1 + x2, data = d),
                   newdata = grid_fit)$posterior[, "k = 2"]
         }),
         QDA = map(data, \(d) {
           predict(MASS::qda(y ~ x1 + x2, data = d),
                   newdata = grid_fit)$posterior[, "k = 2"]
         })) |>
  select(-data) |>
  pivot_longer(c(LDA, QDA), names_to = "method", values_to = "p_hat") |>
  mutate(grid = list(grid_fit)) |>
  unnest(c(grid, p_hat))

variability_plot <- function(nk) {
  ggplot(filter(fitted_boundaries, n_k == nk), aes(x = x1, y = x2)) +
    geom_contour(aes(z = p_hat - 0.5, group = rep), breaks = 0,
                 color = "grey55", linewidth = 0.35) +
    geom_contour(data = bayes_grid, aes(z = qda), breaks = 0, color = "black",
                 linewidth = 1.1, linetype = "longdash") +
    facet_wrap(~ method) +
    coord_fixed(xlim = c(-3.5, 5.5), ylim = c(-3.5, 4), expand = FALSE) +
    labs(x = expression(X[1]), y = expression(X[2]),
         title = bquote(paste(.(n_boundary_reps), " fitted boundaries, ",
                              n[k] == .(nk),
                              " per class; Bayes boundary long-dashed")))
}
variability_small <- variability_plot(n_k_small)
variability_large <- variability_plot(n_k_large)
```
:::

::: {.panel-tabset}
#### Small training sets
```{r variability-small-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 4
variability_small
```

#### Large training sets
```{r variability-large-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 4
variability_large
```
:::

With `r n_k_small` observations per class the LDA lines vary in angle, and
none follows the curvature of the Bayes boundary. The QDA curves follow the
shape of the Bayes boundary but scatter widely around it. With
`r n_k_large` per class the LDA lines tighten around a single line that still
misses the curvature, while the QDA curves collapse onto the Bayes boundary:
more data reduces QDA's variance but not LDA's bias.

LDA should therefore win when the training set is small relative to the
number of covariance parameters, or when the class covariance matrices are
close to equal so that its bias is small. QDA should win when the covariance
matrices clearly differ and the training set is large enough to estimate
them.

::: {.callout-note collapse="true"}
### Beyond this course
Regularized discriminant analysis compromises between the two by shrinking
each class covariance toward the pooled one,
$\hat\Sigma_k(\alpha) = \alpha\hat\Sigma_k + (1 - \alpha)\hat\Sigma$, where
$\alpha = 0$ gives LDA and $\alpha = 1$ gives QDA. The value of $\alpha$ is a
flexibility dial chosen from the data, as the next unit will choose the
number of neighbors for KNN.
:::

## Naive Bayes

Both ellipses in the scatterplot are only slightly tilted: within each class,
balance and income are nearly uncorrelated. A model that treats the features
as independent within each class would describe each cloud by a spread for
each feature alone.

### Conditional independence

Naive Bayes keeps the generative framework but replaces the multivariate model
for $f_k$ with an assumption about how the features relate to one another:

$$P(Y = k) = \pi_k, \qquad X_1, \ldots, X_p \text{ are independent given } Y = k, \qquad X_j \mid Y = k \sim f_{kj},$$

where

- $X_j$ is the $j$th feature, the $j$th entry of the random vector $X$, and
- $f_{kj}$ is the one-feature density (or probability mass function) of $X_j$
among observations in class $k$.

Independence turns the joint density into a product of its marginals, so the
class-conditional density is

$$f_k(x) = f_{k1}(x_1) \times f_{k2}(x_2) \times \cdots \times f_{kp}(x_p) = \prod_{j=1}^p f_{kj}(x_j).$$

The posterior probability is Bayes' theorem with this product in place of
$f_k(x)$. The assumption is conditional on the class. Features that are
related only because both differ between classes, such as two symptoms that
are each more common among patients with a disease, can be independent within
each class while being correlated overall.

Each $f_{kj}$ describes a single feature in a single class, so each can be
modeled and estimated separately, and features of different types can use
densities of different types.

- A quantitative feature is usually modeled as Gaussian,
$X_j \mid Y = k \sim N\left(\mu_{kj}, \sigma_{kj}^2\right)$, with $\mu_{kj}$
and $\sigma_{kj}^2$ estimated by the sample mean and variance of feature $j$
among the class-$k$ training observations.
- A binary feature is Bernoulli,
$f_{kj}(x_j) = \theta_{kj}^{x_j}\left(1 - \theta_{kj}\right)^{1 - x_j}$ for
$x_j \in \{0, 1\}$, where $\theta_{kj} = P(X_j = 1 \mid Y = k)$ is estimated
by the proportion of class-$k$ training observations with $x_j = 1$.
- A categorical feature with more than two levels gets one probability per
level in each class, estimated by the class-$k$ proportions at each level.

### Naive Bayes fit

`e1071::naiveBayes()` takes the same formula and models each numeric
explanatory variable with a Gaussian density in each class. Fitting it to the
training half:

```{r default-nb}
default_nb <- e1071::naiveBayes(default ~ balance + income, data = train)
default_nb
```

The **A-priori probabilities** are $\hat\pi_k$, the same as QDA's. Under
**Conditional probabilities**, the `balance` and `income` tables give each
class's mean (first column) and standard deviation (second column), the
Gaussian $\hat\mu_{kj}$ and $\hat\sigma_{kj}$.

### Additive log posterior odds

With $K = 2$, the logarithm of the product is a sum, and the log posterior
odds of class $2$ are

$$\log\left(\frac{p_2(x)}{1 - p_2(x)}\right) = \log\frac{\pi_2}{\pi_1} + \sum_{j=1}^p \log\frac{f_{2j}(x_j)}{f_{1j}(x_j)} = \log\frac{\pi_2}{\pi_1} + \sum_{j=1}^p g_j(x_j),$$

where

- $g_j(x_j) = \log\left(f_{2j}(x_j)/f_{1j}(x_j)\right)$ is the contribution of
feature $j$, a function of $x_j$ alone.

Each feature shifts the log posterior odds by an amount that does not depend
on the values of the other features: naive Bayes is an additive model with no
interactions, the same structural restriction as the additive logistic
regressions of the multiple logistic regression lecture. The form of each
$g_j$ follows from the type of $f_{kj}$:

- Gaussian with class-specific variances: $g_j$ is quadratic in $x_j$, with
coefficient $\frac{1}{2}\left(1/\sigma_{1j}^2 - 1/\sigma_{2j}^2\right)$ on
$x_j^2$.
- Gaussian with a variance shared by the two classes: the $x_j^2$ terms
cancel and $g_j$ is linear in $x_j$.
- Bernoulli: $g_j(x_j) = x_j\log\left(\theta_{2j}/\theta_{1j}\right) + \left(1 - x_j\right)\log\left[\left(1 - \theta_{2j}\right)/\left(1 - \theta_{1j}\right)\right]$,
linear in $x_j$, so the feature shifts the log posterior odds by
$\log\left\{\left[\theta_{2j}/\left(1 - \theta_{2j}\right)\right] / \left[\theta_{1j}/\left(1 - \theta_{1j}\right)\right]\right\}$
when $x_j$ changes from $0$ to $1$.

For the credit card fit both terms are Gaussian with class-specific variances,
so the balance term and the income term are each quadratic, and there is no
balance-by-income term.

### Diagonal covariance

When every $f_{kj}$ is Gaussian, the product of $p$ independent normal
densities is a multivariate normal density with a diagonal covariance matrix:

$$\prod_{j=1}^p \frac{1}{\sqrt{2\pi}\,\sigma_{kj}}\exp\left[-\frac{\left(x_j - \mu_{kj}\right)^2}{2\sigma_{kj}^2}\right] = \frac{1}{(2\pi)^{p/2}\left|\Lambda_k\right|^{1/2}}\exp\left[-\frac{1}{2}\left(x - \mu_k\right)^\top \Lambda_k^{-1}\left(x - \mu_k\right)\right],$$

where

- $\Lambda_k = \text{diag}\left(\sigma_{k1}^2, \ldots, \sigma_{kp}^2\right)$ is the
diagonal matrix of class-$k$ feature variances, whose determinant is
$\prod_j \sigma_{kj}^2$ and whose inverse is diagonal with entries
$1/\sigma_{kj}^2$.

Gaussian naive Bayes with class-specific variances is therefore QDA with
every $\Sigma_k$ restricted to be diagonal, and Gaussian naive Bayes with one
variance per feature shared across classes is LDA with $\Sigma$ restricted to
be diagonal.

::: {.callout-note collapse="true"}
### Code
```{r nb-diagonal}
#| output: false
nb_sd <- function(k) c(default_nb$tables$balance[k, 2], default_nb$tables$income[k, 2])
max_diff_sd <- max(map_dbl(c("No", "Yes"), \(k) {
  max(abs(nb_sd(k) - sqrt(diag(qda_par[[k]]$Sigma))))
}))
```
:::

For the credit card customers, the naive Bayes means are QDA's and its
standard deviations are the square roots of the diagonals of QDA's
$\hat\Sigma_k$`r if (max_diff_sd < 1e-6) ", up to rounding error" else paste0(", to within $", sci(max_diff_sd), "$ dollars")`: $\hat\Lambda_k$ is
$\hat\Sigma_k$ with the within-class correlation, $`r sprintf("%.3f", summary_of("No", "cor_balance_income"))`$
or $`r sprintf("%.3f", summary_of("Yes", "cor_balance_income"))`$, set to zero.

### Parameter count

Gaussian naive Bayes with class-specific variances estimates $Kp$ means and
$Kp$ variances, $2Kp$ parameters, against QDA's
$Kp + Kp(p+1)/2$. With $K = 2$ and $p = 50$ that is
`r format(2 * 2 * 50, big.mark = ",")` parameters against
`r format(count_qda(2, 50), big.mark = ",")`. The count grows linearly in
$p$, and each parameter is estimated from one feature in one class, using all
$n_k$ observations of that class.

For $p$ binary features the saving is larger still: a joint probability mass
function has $2^p$ cells per class ($2^p - 1$ free probabilities, or
`r format(2^20 - 1, big.mark = ",")` when $p = 20$), most of them never
observed in a training set, against naive Bayes's $p$ Bernoulli
probabilities, which sidesteps the curse of dimensionality of the flexibility
lecture.

The price is bias whenever the features are dependent within a class, which
in practice is almost always. A classification depends only on which side of
the threshold the estimated posterior probability falls, so a biased estimate
of $f_k(x)$ still gives the Bayes classifier's answer wherever the posterior
probability it produces falls on the same side of the threshold as the true
one.

::: {.callout-note collapse="true"}
### Beyond this course
A quantitative feature need not be modeled as Gaussian. Each $f_{kj}$ can
instead be estimated nonparametrically, by a histogram or a kernel density
estimate of feature $j$ within class $k$, since each is a one-dimensional
density. The `naivebayes` package offers this option.
:::

### Naive Bayes boundary

The figure below draws the naive Bayes boundary (blue, dot-dash) beside the
QDA boundary (orange, dotted) over the training customers, non-defaulters as
grey circles and defaulters as black triangles, with balances beyond the
largest training balance shaded.

::: {.callout-note collapse="true"}
### Code
```{r nb-boundary}
#| output: false
nb_boundary_plot <- boundary_figure(
  c("QDA", "Naive Bayes"),
  "Naive Bayes and QDA boundaries; shading beyond the largest training balance")
nb_cross <- crossings("Naive Bayes")
stopifnot(all(lengths(nb_cross) == 2),
          all(cross_at(nb_cross, 2) > max(train$balance)))

# Naive Bayes's income term at the median income, per $1,000, against LDA's
# income slope (pooled-covariance coefficient), per $1,000.
nb_income <- default_nb$tables$income
g_income <- function(x) {
  dnorm(x, nb_income["Yes", 1], nb_income["Yes", 2], log = TRUE) -
    dnorm(x, nb_income["No", 1], nb_income["No", 2], log = TRUE)
}
nb_income_slope <- g_income(median(train$income) + 500) -
  g_income(median(train$income) - 500)
income_gap <- nb_income["Yes", 1] - nb_income["No", 1]
pooled <- Reduce(`+`, map(c("No", "Yes"), \(k) {
  crossprod(scale(X_train[train$default == k, ], scale = FALSE))
})) / (nrow(train) - 2)
mean_diff <- c(diff(default_nb$tables$balance[, 1]), income_gap)
lda_income_slope <- 1000 * solve(pooled, mean_diff)[2]
raises <- function(v) if (v > 0) "raises" else "lowers"
# Two significant digits without rounding a small value to 0.0000.
small_num <- function(v) format(signif(v, 2), scientific = FALSE)
```
:::

```{r nb-boundary-figure}
#| echo: false
#| fig-width: 7.5
#| fig-height: 4.5
nb_boundary_plot
```

Within the training balances the two boundaries run close together, as
nearly uncorrelated features would lead one to expect. Naive Bayes's balance
term is quadratic for the same reason as QDA's, so its boundary also has a
second branch beyond the training balances, crossing $0.5$ at
`r dollar_text(cross_at(nb_cross, 2)[1])`,
`r dollar_text(cross_at(nb_cross, 2)[2])`, and
`r dollar_text(cross_at(nb_cross, 2)[3])` at the three income quartiles.

Naive Bayes drops the within-class correlation, so its income term compares
the two classes' income distributions without adjusting for balance:
defaulters' mean training income is `r dollar_text(abs(income_gap))`
`r if (income_gap > 0) "higher" else "lower"` than non-defaulters', and at the
median income each additional \$1,000 `r raises(nb_income_slope)` naive
Bayes's log posterior odds by `r small_num(abs(nb_income_slope))`. The term is
slightly curved, because the two classes' income standard deviations differ.
LDA's income slope, adjusted for balance through the pooled within-class
covariance as the [previous lecture](50-lda.qmd#balance-and-income)
explained, `r raises(lda_income_slope)` them by
`r small_num(abs(lda_income_slope))` per \$1,000,
`r if (sign(nb_income_slope) == sign(lda_income_slope)) "the same direction" else "the opposite direction"`.

## Classifier comparison

### KNN classification

Five classifiers are now available: logistic regression, LDA, QDA, naive
Bayes, and K-nearest neighbors (KNN). From here to the end of this section,
$K$ is the number of neighbors rather than the number of classes: every comparison has two classes, and the number of
classes is written as $2$.

The flexibility lecture introduced KNN for a quantitative response. For a
qualitative response the neighbors vote:

$$\hat p_k(x_0) = \frac{1}{K}\sum_{i \in \mathcal{N}_0}\mathrm{I}(y_i = k),$$

where

- $K$ is the **number of neighbors**,
- $\mathcal{N}_0$ is the set of the $K$ training observations closest to $x_0$
in Euclidean distance after every feature, in the training data and in $x_0$
alike, is standardized by its training mean and standard deviation, since
that distance depends on each feature's scale (as in the flexibility
lecture), and
- $\hat p_k(x_0)$ is the proportion of those neighbors in class $k$, and the
classifier assigns the class with the largest proportion.

### Boundary shapes

With two classes, the four model-based classifiers differ in the form of the
log posterior odds they can represent:

$$\begin{aligned}
\text{Logistic regression, LDA: } & \beta_0 + \sum_{j=1}^p \beta_j x_j, \\
\text{QDA: } & \beta_0 + \sum_{j=1}^p \beta_j x_j + \sum_{j \le l} \gamma_{jl} x_j x_l, \\
\text{Gaussian naive Bayes: } & \beta_0 + \sum_{j=1}^p \left(\beta_j x_j + \gamma_{jj} x_j^2\right),
\end{aligned}$$

where

- $\beta_0$ is an intercept and $\beta_j$ the coefficient of $x_j$, with
values that differ from method to method, and
- $\gamma_{jl}$ is the coefficient of the product $x_j x_l$, a square when
$l = j$.

A naive Bayes classifier with Bernoulli or categorical features adds terms
linear in those features' indicators. KNN has no formula for its log
posterior odds, and its boundary can take any shape. The table summarizes all
five:

::: {.callout-note collapse="true"}
### Code
```{r comparison-table}
#| output: false
comparison_table <- tibble(
  Method = c("Logistic regression", "LDA", "QDA", "Gaussian naive Bayes",
             "KNN"),
  Assumption = c("Linear log-odds",
                 "Gaussian, shared covariance",
                 "Gaussian, class covariances",
                 "Features independent within each class",
                 "Smooth class probabilities"),
  Estimation = c("Iterative", "Closed form", "Closed form", "Closed form",
                 "None (stores data)")
) |>
  knitr::kable(align = "lll")
```
:::

```{r comparison-table-display}
#| echo: false
comparison_table
```

Reading down the table:

- **Logistic regression and LDA** share a linear form and differ in
estimation. LDA adds the Gaussian, shared-covariance assumption; when it
holds, both estimate the same log posterior odds, and when it fails, logistic
regression, which never models the features' distribution, is unaffected by
that failure provided the log-odds remain linear.
- **QDA** adds every square and cross-product. A logistic regression given
those same squared and product features, as in the flexible logistic
regression lecture, has the same quadratic form, fit by the conditional
likelihood instead.
- **Naive Bayes** is the most restricted of the quadratic forms, with no
cross-products, and in exchange has fewer parameters than QDA.
- **KNN** is nonparametric: it assumes no form at all, and its flexibility is
set by $1/K$. It can follow any boundary given enough data, but it needs $n$
large relative to $p$, suffers most from the curse of dimensionality, and has
no coefficients to interpret.

Increasing $n$ favors the more flexible methods, QDA and KNN, whose lower
bias becomes affordable as their variance shrinks. Increasing $p$ favors the
more restricted ones: naive Bayes, whose parameter count grows linearly in
$p$, and LDA, whose count grows like $p^2/2$ rather than like QDA's $p^2$ for
two classes. No method is best in every setting.

### Simulated scenarios

A small simulation puts these predictions to a test. Each of four scenarios
has two classes with equal priors, a data-generating process chosen so that a
different method's assumptions hold:

- **Linear**: two Gaussian features, a shared covariance matrix with
correlation $0.5$, and different means, so the Bayes boundary is a line.
- **Quadratic**: the two classes of the variability figure, with
class-specific covariance matrices.
- **Independent, $p = 10$**: ten Gaussian features, independent within each
class; the classes differ in the means of the first five features and the
variances of the last five.
- **Non-linear**: two features uniform on a square, class $2$ above the curve
$x_2 = \sin(2x_1)$ and class $1$ below it, and then $10\%$ of the class labels
flipped at random, so neither class is Gaussian and the Bayes boundary is the
sine curve.

::: {.callout-note collapse="true"}
### Code
```{r scenario-study}
#| output: false
scenario_names <- c("Linear", "Quadratic", "Independent, p = 10",
                    "Non-linear")
n_reps <- 100       # training sets per scenario
n_train_k <- 50     # training observations per class
n_test_k <- 2000    # test observations per class
k_sim <- odd_above_sqrt(2 * n_train_k)
knn_sim_name <- paste0("KNN, K = ", k_sim)

rmvnorm_p <- function(n, mu, Sigma) {
  sweep(matrix(rnorm(n * length(mu)), n) %*% chol(Sigma), 2, mu, "+")
}

# One data set of 2 * n_k observations, n_k per class (in expectation for the
# non-linear scenario, whose class sizes are random).
draw_scenario <- function(scenario, n_k) {
  labels <- c("Class 1", "Class 2")
  if (scenario == "Non-linear") {
    X <- matrix(runif(4 * n_k, -2, 2), ncol = 2)
    y <- as.numeric(X[, 2] > sin(2 * X[, 1]))
    flip <- runif(2 * n_k) < 0.1
    y[flip] <- 1 - y[flip]
    y <- factor(labels[y + 1], levels = labels)
  } else {
    if (scenario == "Linear") {
      S <- matrix(c(1, 0.5, 0.5, 1), nrow = 2)
      X <- rbind(rmvnorm_p(n_k, c(0, 0), S), rmvnorm_p(n_k, c(1.5, 0.5), S))
    }
    if (scenario == "Quadratic") {
      X <- rbind(rmvnorm_p(n_k, mu_a, Sigma_a), rmvnorm_p(n_k, mu_b, Sigma_b))
    }
    if (scenario == "Independent, p = 10") {
      X <- rbind(rmvnorm_p(n_k, rep(0, 10), diag(10)),
                 rmvnorm_p(n_k, rep(c(0.5, 0), each = 5),
                           diag(rep(c(1, 2), each = 5))))
    }
    y <- factor(rep(labels, each = n_k), levels = labels)
  }
  colnames(X) <- paste0("x", seq_len(ncol(X)))
  bind_cols(tibble(y = y), as_tibble(X))
}

# Bayes classifier's error rate on a test set, from the true model.
bayes_error <- function(scenario, test) {
  X <- as.matrix(select(test, -y))
  truth <- test$y == "Class 2"
  if (scenario == "Non-linear") {
    pred <- X[, 2] > sin(2 * X[, 1])
  } else {
    if (scenario == "Linear") {
      S <- matrix(c(1, 0.5, 0.5, 1), nrow = 2)
      lo <- log_odds_gauss(X[, 1], X[, 2], c(0, 0), S, c(1.5, 0.5), S)
    }
    if (scenario == "Quadratic") {
      lo <- log_odds_gauss(X[, 1], X[, 2], mu_a, Sigma_a, mu_b, Sigma_b)
    }
    if (scenario == "Independent, p = 10") {
      lo <- rowSums(dnorm(X, rep(c(0.5, 0), each = 5)[col(X)],
                          sqrt(rep(c(1, 2), each = 5))[col(X)], log = TRUE) -
                    dnorm(X, 0, 1, log = TRUE))
    }
    pred <- lo > 0
  }
  mean(pred != truth)
}

# Test error rate of each method fit to one training set.
fit_methods <- function(train, test) {
  truth <- test$y
  X_train <- as.matrix(select(train, -y))
  X_test  <- as.matrix(select(test, -y))
  # KNN uses features standardized by the training means and SDs.
  centers <- colMeans(X_train)
  scales  <- apply(X_train, 2, sd)
  Z_train <- scale(X_train, centers, scales)
  Z_test  <- scale(X_test, centers, scales)
  k_sqrt  <- odd_above_sqrt(nrow(train))
  glm_fit <- glm(y ~ ., data = train, family = "binomial")
  errors <- c(`Logistic regression` = mean(
      (predict(glm_fit, newdata = test, type = "response") > 0.5) !=
        (truth == "Class 2")),
    LDA = mean(predict(MASS::lda(y ~ ., data = train), test)$class != truth),
    QDA = mean(predict(MASS::qda(y ~ ., data = train), test)$class != truth),
    `Naive Bayes` = mean(
      predict(e1071::naiveBayes(y ~ ., data = train), test) != truth),
    `KNN, K = 1` = mean(class::knn(Z_train, Z_test, train$y, k = 1) != truth),
    knn_sqrt = mean(class::knn(Z_train, Z_test, train$y, k = k_sqrt) != truth))
  names(errors)[names(errors) == "knn_sqrt"] <- paste0("KNN, K = ", k_sqrt)
  errors
}

scenario_results <- map(scenario_names, \(s) {
  test <- draw_scenario(s, n_test_k)
  errors <- map(seq_len(n_reps),
                \(r) fit_methods(draw_scenario(s, n_train_k), test))
  list(errors = bind_rows(errors) |> mutate(scenario = s, rep = row_number()),
       bayes = bayes_error(s, test))
})

method_levels <- c("Logistic regression", "LDA", "QDA", "Naive Bayes",
                   "KNN, K = 1", knn_sim_name)

scenario_errors <- map(scenario_results, "errors") |>
  bind_rows() |>
  pivot_longer(all_of(method_levels), names_to = "method",
               values_to = "error") |>
  mutate(method = factor(method, levels = method_levels),
         scenario = factor(scenario, levels = scenario_names))

scenario_bayes <- tibble(
  scenario = factor(scenario_names, levels = scenario_names),
  bayes = map_dbl(scenario_results, "bayes"))

scenario_means <- scenario_errors |>
  group_by(scenario, method) |>
  summarize(mean_error = mean(error), .groups = "drop")

mean_error_of <- function(s, m) {
  scenario_means$mean_error[scenario_means$scenario == s &
                              scenario_means$method == m]
}
best_method <- function(s) {
  d <- filter(scenario_means, scenario == s)
  as.character(d$method[which.min(d$mean_error)])
}
pct <- function(v) sprintf("%.1f%%", 100 * v)

# Prose labels for the winning method of each scenario.
prose_label <- c("Logistic regression" = "logistic regression", "LDA" = "LDA",
                 "QDA" = "QDA", "Naive Bayes" = "naive Bayes",
                 "KNN, K = 1" = "KNN with $K = 1$")
prose_label[knn_sim_name] <- paste0("KNN with $K = ", k_sim, "$")
winner <- function(s) unname(prose_label[best_method(s)])

pp <- function(v) sprintf("%.3f", 100 * v)   # percentage points

# One training set per scenario to show what the data look like; the
# independent scenario is shown in features 1 (mean shift) and 6 (variance).
scenario_examples <- map(scenario_names, \(s) {
  draw_scenario(s, n_train_k) |>
    transmute(y, first = x1, second = if (s == "Independent, p = 10") x6 else x2,
              scenario = s)
}) |>
  bind_rows() |>
  mutate(scenario = factor(scenario, levels = scenario_names))

scenario_data_plot <- ggplot(scenario_examples,
                             aes(x = first, y = second, color = y, shape = y)) +
  geom_point(size = 1.6, alpha = 0.8) +
  facet_wrap(~ scenario, scales = "free") +
  scale_color_manual(values = class_colors) +
  labs(x = "First plotted feature", y = "Second plotted feature",
       color = "Class", shape = "Class",
       title = "One training set per scenario (Independent: features 1 and 6)")

scenario_error_plot <- ggplot(scenario_errors, aes(x = method, y = error)) +
  geom_boxplot(outlier.size = 0.6, fill = "grey90") +
  geom_hline(data = scenario_bayes, aes(yintercept = bayes),
             linetype = "dashed") +
  facet_wrap(~ scenario, scales = "free_y") +
  labs(x = NULL, y = "Test error rate",
       title = paste("Test error rates over", n_reps,
                     "training sets; Bayes error rate dashed")) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```
:::

In each scenario, `r n_reps` training sets of `r 2 * n_train_k` observations
were drawn, `r n_train_k` per class (in expectation in the non-linear
scenario). Each method was fit to every training set and its error rate
computed on one large test set of `r format(2 * n_test_k, big.mark = ",")`
observations drawn from the same scenario. KNN was fit on standardized
features with $K = 1$ and with
$K = `r k_sim`$, the smallest odd integer above $\sqrt{n}$ for the
$n = `r 2 * n_train_k`$ training observations: odd so that a two-class vote
cannot tie, and near $\sqrt{n}$ as a rule of thumb rather than an optimal
choice. Both are fixed in advance, since choosing $K$ from the data requires
the resampling methods of the next unit.

The first tab below shows one training set from each scenario, classes
distinguished by color and shape, with features 1 and 6 for the independent
scenario. In the second, each box summarizes the test error rates of one
method over the `r n_reps` training sets, and the dashed line is the Bayes
error rate, the floor no classifier can beat on average.

::: {.panel-tabset}
#### Scenarios
```{r scenario-data-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 6.5
scenario_data_plot
```

#### Test error rates
```{r scenario-error-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 6.5
scenario_error_plot
```
:::

- **Linear.** Logistic regression and LDA
`r if (pct(mean_error_of("Linear", "Logistic regression")) == pct(mean_error_of("Linear", "LDA"))) paste("both average", pct(mean_error_of("Linear", "LDA"))) else paste("average", pct(mean_error_of("Linear", "Logistic regression")), "and", pct(mean_error_of("Linear", "LDA")))`,
against a Bayes error rate of
`r pct(scenario_bayes$bayes[1])`. QDA, whose extra covariance parameters are
unnecessary here, averages `r pct(mean_error_of("Linear", "QDA"))`: with only
two features its variance penalty is small. Naive Bayes, which ignores the
correlation, averages `r pct(mean_error_of("Linear", "Naive Bayes"))`.
- **Quadratic.** QDA's average of `r pct(mean_error_of("Quadratic", "QDA"))`
beats logistic regression's
`r pct(mean_error_of("Quadratic", "Logistic regression"))` and LDA's
`r pct(mean_error_of("Quadratic", "LDA"))`: the linear methods' bias outweighs
QDA's variance. Naive Bayes, curved but missing the cross-product, falls
between them, at `r pct(mean_error_of("Quadratic", "Naive Bayes"))`.
- **Independent, $p = 10$.** Naive Bayes, whose assumption holds exactly,
averages `r pct(mean_error_of("Independent, p = 10", "Naive Bayes"))` with
`r 2 * 2 * 10` parameters. QDA's model also contains the truth, but it
estimates `r count_qda(2, 10)` parameters from `r 2 * n_train_k` observations and averages
`r pct(mean_error_of("Independent, p = 10", "QDA"))`. KNN, with ten
dimensions to search for neighbors, averages
`r pct(mean_error_of("Independent, p = 10", knn_sim_name))` with
$K = `r k_sim`$.
- **Non-linear.** No model-based method can represent a sine curve, and all
four average
`r nl <- pct(range(sapply(method_levels[1:4], \(m) mean_error_of("Non-linear", m)))); if (nl[1] == nl[2]) paste("about", nl[1]) else paste("between", nl[1], "and", nl[2])`.
KNN with $K = `r k_sim`$ averages
`r pct(mean_error_of("Non-linear", knn_sim_name))`. With $K = 1$ its variance
erases that advantage: it averages
`r pct(mean_error_of("Non-linear", "KNN, K = 1"))`, about the same as the
model-based methods.

In the quadratic, independent, and non-linear scenarios the best method is the
one whose assumptions are the most restrictive ones that still hold:
`r winner("Quadratic")` in the quadratic, `r winner("Independent, p = 10")` in
the independent, and `r winner("Non-linear")` in the non-linear scenario. In
the linear scenario both LDA and logistic regression are correct models, and
their averages differ by only
`r pp(abs(mean_error_of("Linear", "Logistic regression") - mean_error_of("Linear", "LDA")))`
percentage points, with
`r if (best_method("Linear") == "LDA") "LDA" else "logistic regression"` ahead:
with two features, LDA's extra assumption buys essentially nothing. None wins
in every scenario.

### Test performance

Returning to the credit card customers, logistic regression with `balance`
and `income` and KNN are fit to the training half beside LDA, QDA, and naive
Bayes, and every classifier is scored on both halves.

::: {.callout-note collapse="true"}
### Code
```{r default-test}
#| output: false
default_glm <- glm(default ~ balance + income, data = train,
                   family = "binomial")
k_default <- odd_above_sqrt(nrow(train))

# KNN vote proportion for "Yes", on features standardized by training means/SDs.
standardize <- function(d) {
  cbind((d$balance - mean(train$balance)) / sd(train$balance),
        (d$income - mean(train$income)) / sd(train$income))
}
knn_yes <- function(target, K) {
  fit <- class::knn(standardize(train), standardize(target), train$default,
                    k = K, prob = TRUE)
  ifelse(fit == "Yes", attr(fit, "prob"), 1 - attr(fit, "prob"))
}

knn_default_name <- paste0("KNN, K = ", k_default)
model_based_names <- c("Logistic regression", "LDA", "QDA", "Naive Bayes")
all_scores <- function(d) {
  map(model_based_names, \(m) posterior_of(m, d)) |>
    set_names(model_based_names) |>
    as_tibble() |>
    mutate(`KNN, K = 1` = knn_yes(d, 1), "{knn_default_name}" := knn_yes(d, k_default))
}
scores_train <- all_scores(train)
scores_test  <- all_scores(test)

# AUC as the proportion of (defaulter, non-defaulter) pairs ordered correctly.
auc_pairs <- function(s, pos) {
  mean(outer(s[pos], s[!pos], ">")) + 0.5 * mean(outer(s[pos], s[!pos], "=="))
}
yes_train <- train$default == "Yes"
yes_test  <- test$default == "Yes"

rates <- function(p_train, p_test) {
  tibble(train_error = mean((p_train > 0.5) != yes_train),
         test_error  = mean((p_test > 0.5) != yes_test),
         sensitivity = mean(p_test[yes_test] > 0.5),
         specificity = mean(p_test[!yes_test] <= 0.5),
         auc         = auc_pairs(p_test, yes_test),
         flagged     = sum(p_test > 0.5))
}
performance <- map(names(scores_test), \(m) rates(scores_train[[m]], scores_test[[m]])) |>
  set_names(names(scores_test)) |>
  list_rbind(names_to = "model")
perf_of <- function(m, col) performance[[col]][performance$model == m]
model_based <- performance |> slice(1:4)
n_train_above <- sum(model_based$train_error > model_based$test_error)
defaulter_counts <- sprintf("%d defaulters (%s) and the test half %d (%s)",
                            n_defaults[["train"]], pct2(mean(yes_train)),
                            n_defaults[["test"]], pct2(mean(yes_test)))
optimism_text <- if (n_train_above == 0) {
  paste0("No model-based classifier has a training error rate above its test ",
         "error rate, as optimism predicts; the training half drew ",
         defaulter_counts, ".")
} else if (n_train_above < 4) {
  paste0("Training error rates are optimistic on average, but on a single ",
         "split chance can outweigh that: ",
         c("one of the four model-based classifiers has",
           "two of the four model-based classifiers have",
           "three of the four model-based classifiers have")[n_train_above],
         " a training error rate above the test error rate.")
} else {
  paste0("Training error rates are optimistic on average, but on a single ",
         "split chance can outweigh that: all four model-based classifiers ",
         "have a training error rate above their test error rate",
         if (n_defaults[["train"]] > n_defaults[["test"]]) {
           paste0(", perhaps because the training half drew ",
                  defaulter_counts, ".")
         } else ".")
}

performance_table <- performance |>
  select(-flagged) |>
  mutate(across(train_error:specificity, pct2), auc = sprintf("%.4f", auc)) |>
  knitr::kable(col.names = c("Model", "Training error", "Test error",
                             "Test sensitivity", "Test specificity", "Test AUC"),
               align = "lccccc")

# Naive Bayes against QDA, from the rates as displayed in the table.
nb_error_diff <- round(perf_of("Naive Bayes", "test_error") - perf_of("QDA", "test_error"), 4)
nb_auc_diff   <- round(perf_of("Naive Bayes", "auc"), 4) - round(perf_of("QDA", "auc"), 4)
nb_error_text <- if (nb_error_diff == 0) "the same as QDA's" else
  paste(sprintf("%.2f", 100 * abs(nb_error_diff)), "percentage points",
        if (nb_error_diff > 0) "higher" else "lower", "than QDA's")
nb_auc_text <- if (nb_auc_diff == 0) "the same as QDA's" else
  paste(sprintf("%.4f", abs(nb_auc_diff)), if (nb_auc_diff > 0) "higher" else "lower")
nb_cost <- if (nb_error_diff <= 0 && nb_auc_diff >= 0) "does not hurt naive Bayes" else
  "costs naive Bayes little"

# Model-based classifiers with the highest sensitivity and the most customers
# flagged, named in prose (ties listed together).
model_label <- c(`Logistic regression` = "logistic regression", LDA = "LDA",
                 QDA = "QDA", `Naive Bayes` = "naive Bayes")
and_list <- function(x) {
  if (length(x) == 1) x else paste(paste(x[-length(x)], collapse = ", "), "and", x[length(x)])
}
top_models <- function(col) {
  v <- round(model_based[[col]], 6)
  unname(model_label[model_based$model[v == max(v)]])
}
sens_top <- top_models("sensitivity")
flag_top <- top_models("flagged")
top_text <- paste0(
  if (length(sens_top) == 1) paste0("The highest sensitivity is ", sens_top, "'s") else
    paste("The highest sensitivity is shared by", and_list(sens_top)),
  if (identical(sens_top, flag_top) && length(flag_top) == 1) {
    paste0(", which flags the most test customers, ", max(model_based$flagged), ".")
  } else {
    paste0("; ", and_list(flag_top), if (length(flag_top) == 1) " flags" else " flag",
           " the most test customers, ", max(model_based$flagged), ".")
  })

# KNN's place among the classifiers, stated only when it holds.
knn1_worst <- perf_of("KNN, K = 1", "test_error") == max(performance$test_error)
knn_auc_lowest <- perf_of(knn_default_name, "auc") < min(model_based$auc)
```
:::

KNN uses balance and income, each standardized by its training mean and
standard deviation, with $K = 1$ and $K = `r k_default`$, the smallest odd
integer above $\sqrt{`r format(nrow(train), big.mark = "{,}")`}$. The table
gives each classifier's training error rate, and its test error rate,
sensitivity, and specificity at threshold $0.5$ (for KNN, a majority vote),
and its test AUC; the training error rate is computed on the customers each
model was fit to.

```{r performance-table}
#| echo: false
performance_table
```

`r optimism_text` KNN with $K = 1$ shows the optimism plainly: its training
error rate is `r pct2(perf_of("KNN, K = 1", "train_error"))`, since each
training customer is its own nearest neighbor, against a test error rate of
`r pct2(perf_of("KNN, K = 1", "test_error"))``r if (knn1_worst) ", the worst in the table" else ""`.

The four model-based classifiers have test error rates
`r te <- pct2(range(model_based$test_error)); if (te[1] == te[2]) paste("of", te[1]) else paste("between", te[1], "and", te[2])`,
against the
`r pct2(mean(yes_test))` of predicting `No` for every test customer. They
differ more in sensitivity, from
`r sprintf("%.1f%%", 100 * min(model_based$sensitivity))` to
`r sprintf("%.1f%%", 100 * max(model_based$sensitivity))`, a difference of
`r round(n_defaults["test"] * diff(range(model_based$sensitivity)))` of the
`r n_defaults["test"]` test defaulters. `r top_text` Their test AUCs range
only from `r sprintf("%.4f", min(model_based$auc))` to
`r sprintf("%.4f", max(model_based$auc))`. With the within-class correlation
this small, dropping it `r nb_cost`: its test error rate is
`r nb_error_text`, and its AUC is `r nb_auc_text`. KNN with
$K = `r k_default`$ has `r if (knn_auc_lowest) "a lower test AUC than any of them" else "a test AUC"`,
`r sprintf("%.4f", perf_of(knn_default_name, "auc"))`.

## Student strata

Student status has not entered any model so far. In this additional
analysis, QDA with `balance` and `income` is fit separately to the training set's
`r format(sum(train$student == "No"), big.mark = ",")` non-students, of whom
`r sprintf("%.2f", 100 * mean(train$default[train$student == "No"] == "Yes"))`%
defaulted, and `r format(sum(train$student == "Yes"), big.mark = ",")`
students, of whom
`r sprintf("%.2f", 100 * mean(train$default[train$student == "Yes"] == "Yes"))`%
defaulted, beside the logistic regression whose intercept and slopes all
interact with student status. Each stratum's QDA uses that stratum's class
proportions as priors, so every customer's score is its posterior probability
given balance, income, and student status, on one common scale across strata,
which is what lets a single error rate and AUC be computed over all customers.

::: {.callout-note collapse="true"}
### Code
```{r strata}
#| output: false
strata <- c(No = "No", Yes = "Yes")
train_by <- map(strata, \(s) filter(train, student == s))
strata_qda <- map(train_by, \(d) MASS::qda(default ~ balance + income, data = d))
strata_glm <- map(train_by, \(d) glm(default ~ balance + income, data = d,
                                     family = "binomial"))
interacted_glm <- glm(default ~ (balance + income) * student, data = train,
                      family = "binomial")

# Score every customer with its own stratum's model.
by_stratum <- function(d, score) {
  p <- numeric(nrow(d))
  for (s in strata) p[d$student == s] <- score(s, d[d$student == s, ])
  p
}
qda_strat <- function(d) by_stratum(d, \(s, ds) predict(strata_qda[[s]], ds)$posterior[, "Yes"])
max_diff_int <- max(abs(fitted(interacted_glm) - by_stratum(train, \(s, ds) {
  predict(strata_glm[[s]], ds, type = "response")
})))
glm_int <- function(d) predict(interacted_glm, newdata = d, type = "response")

# Rates rounded as displayed, so changes computed from them match the table.
strata_rates <- list(
  "QDA, single fit" = rates(scores_train$QDA, scores_test$QDA),
  "QDA, by student stratum" = rates(qda_strat(train), qda_strat(test)),
  "Logistic regression, additive" = rates(scores_train$`Logistic regression`,
                                          scores_test$`Logistic regression`),
  "Logistic regression, interacted with student" =
    rates(glm_int(train), glm_int(test))
) |>
  list_rbind(names_to = "model") |>
  mutate(across(train_error:sensitivity, \(v) round(v, 4)), auc = round(auc, 4))
strata_table <- strata_rates |>
  select(-specificity, -flagged) |>
  mutate(across(train_error:sensitivity, pct2), auc = sprintf("%.4f", auc)) |>
  knitr::kable(col.names = c("Model", "Training error", "Test error",
                             "Test sensitivity", "Test AUC"), align = "lcccc")
# Largest change in test error rate and test AUC from stratifying either
# model: rows 2 and 4 are the stratified versions of rows 1 and 3.
max_error_change <- max(abs(diff(strata_rates$test_error)[c(1, 3)]))
max_auc_change   <- max(abs(diff(strata_rates$auc)[c(1, 3)]))
# Mean and covariance parameters of a two-class QDA, priors not counted, as in
# the parameter-count table.
n_par_qda <- count_qda(2, 2)
```
:::

The table gives each model's training and test error rates and test
sensitivity at $t = 0.5$, and its test AUC.

```{r strata-table}
#| echo: false
strata_table
```

Stratifying changes no model's test error rate by more than
`r sprintf("%.2f", 100 * max_error_change)` percentage points or its test AUC
by more than `r sprintf("%.4f", max_auc_change)`. (The interacted logistic
regression reproduces separate per-stratum logistic regressions, to within
$`r sci(max_diff_int)`$ on the training set.) Each stratified model estimates
twice as many parameters, `r 2 * n_par_qda` means and covariance entries against
`r n_par_qda` for QDA (priors not counted, as in the parameter-count table)
and `r length(coef(interacted_glm))` coefficients against
`r length(coef(default_glm))` for logistic regression, which tends to make its
training error rate more optimistic; the test rates are the fair comparison.

## Conclusion

QDA is the generative classifier with Gaussian class-conditional densities and
a covariance matrix $\Sigma_k$ for each class. The quadratic term
$-\frac{1}{2}x^\top\Sigma_k^{-1}x$ and the term
$-\frac{1}{2}\log\left|\Sigma_k\right|$ no longer cancel between classes, so
the discriminant functions and the log posterior odds are quadratic in $x$,
the decision boundary is a quadratic curve or surface, and with one feature the class
with the larger variance is assigned in both tails. Each $\Sigma_k$ is
estimated by its class's sample covariance matrix, which requires $n_k > p$.

The choice between LDA and QDA is a bias-variance tradeoff. With $K$ classes,
QDA estimates $K p(p+1)/2$ covariance parameters against LDA's $p(p+1)/2$, so it removes
LDA's bias when the class covariances differ at the cost of variance that
grows with $p$ and shrinks with $n$.

Naive Bayes assumes the features are independent within each class, so
$f_k(x) = \prod_j f_{kj}(x_j)$, each one-feature density estimated on its own
and of whatever type suits the feature. The log posterior odds are a sum of
one-feature terms, with no interactions. Gaussian naive Bayes is LDA or QDA
with a diagonal covariance matrix. With class-specific variances it estimates
$2Kp$ parameters, sidestepping the curse of dimensionality that a joint
density faces. Its bias, which comes from ignoring within-class dependence,
need not change its classifications: it agrees with the Bayes classifier
wherever its estimated posterior probability falls on the same side of the
threshold as the true one.

Logistic regression and LDA produce linear boundaries, QDA and Gaussian naive
Bayes quadratic ones, and KNN boundaries of any shape. Which classifier has the
lowest test error rate depends on how closely its assumptions match the
data-generating process and on the sizes of $n$ and $p$; no method is best
in every setting. Comparing classifiers requires test data the models were
not fit to, since training error rates reward flexibility.

The next lecture steps back from classification to a framework that contains
linear and logistic regression as special cases. Generalized linear models
pair a distribution for the response with a link function for its mean, and
Poisson regression, the generalized linear model for counts, is the first new
member of the family.
