# Quadratic Discriminant Analysis and Naive Bayes

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
set.seed(20261001)

# MASS, e1071, and class are called with :: rather than attached, because
# attaching MASS masks dplyr::select().
class_colors <- c("#0072B2", "#D55E00")
model_colors <- c("Logistic regression" = "#CC79A7", "LDA" = "#009E73",
                  "QDA" = "#E69F00", "Naive Bayes" = "#56B4E9",
                  "KNN" = "black")
model_linetypes <- c("Logistic regression" = "dashed", "LDA" = "solid",
                     "QDA" = "dotted", "Naive Bayes" = "dotdash",
                     "KNN" = "twodash")

# Density of N_2(mu, Sigma) at the rows (x1, x2), and the two-class Gaussian
# log posterior odds of class 2 with equal priors.
dmvnorm2 <- function(x1, x2, mu, Sigma) {
  d <- cbind(x1 - mu[1], x2 - mu[2])
  q <- rowSums((d %*% solve(Sigma)) * d)
  exp(-q / 2) / (2 * pi * sqrt(det(Sigma)))
}
log_odds_gauss <- function(x1, x2, mu1, S1, mu2, S2) {
  log(dmvnorm2(x1, x2, mu2, S2)) - log(dmvnorm2(x1, x2, mu1, S1))
}

# A number to `digits` significant digits, keeping trailing zeros ("0.010").
sig_text <- function(v, digits = 2) {
  sub("\\.$", "", formatC(signif(v, digits), digits = digits, format = "fg",
                           flag = "#"))
}
# Renders a number in LaTeX scientific notation for use inside $...$.
sci <- function(x, digits = 1) {
  if (x == 0) return("0")
  e <- floor(log10(abs(x)))
  m <- signif(x / 10^e, digits)
  if (abs(m) >= 10) {
    m <- m / 10
    e <- e + 1
  }
  sprintf("%s \\times 10^{%d}", sig_text(m, digits), e)
}
# Fixed notation for magnitudes from 0.01 to 1000, scientific otherwise.
num_tex <- function(v, digits = 3) {
  if (abs(v) >= 0.01 && abs(v) < 1000) sig_text(v, digits) else sci(v, digits)
}

# Dollar amounts for prose, with the "$" escaped for markdown: "\$1,234".
dollar_text <- function(v) paste0("\\$", format(round(v), big.mark = ","))
# Axis labels such as $60,000, without the scales package.
dollar_labels <- function(v) {
  ifelse(is.na(v), NA,
         paste0("$", format(v, big.mark = ",", scientific = FALSE, trim = TRUE)))
}
pct2 <- function(v) sprintf("%.2f%%", 100 * v)
# "a", "a and b", "a, b, and c".
and_list <- function(x) {
  if (length(x) == 1) x else paste(paste(x[-length(x)], collapse = ", "), "and", x[length(x)])
}

# Smallest odd integer greater than sqrt(n): a fixed rule of thumb for the
# number of neighbors that does not look at any test data. Odd, so a
# two-class vote cannot tie.
odd_above_sqrt <- function(n) {
  k <- floor(sqrt(n))
  if (k %% 2 == 0) k + 1 else k + 2
}
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **derive** the QDA discriminant function $\delta_k(x)$ from Gaussian
class-conditional densities with class-specific covariance matrices
$\Sigma_k$, and **show** that the resulting decision boundary is quadratic in
$x$, with up to two boundary points when $p = 1$,
- **count** the mean and covariance parameters that LDA, QDA, and Gaussian
naive Bayes estimate, and **use** the counts to **explain** through the
bias-variance tradeoff when each should have the lowest test error rate,
- **state** the naive Bayes assumption that the features are conditionally
independent given the class, **write** $f_k(x)$ as a product of one-feature
densities of possibly different types, and **show** that the log posterior
odds are a sum of one-feature terms,
- **identify** Gaussian naive Bayes as LDA or QDA with a diagonal covariance
matrix,
- **compare** logistic regression, LDA, QDA, naive Bayes, and KNN by the shape
of their decision boundaries, their assumptions, and how they are affected by
$n$ and $p$, and **evaluate** which should perform best in a described
setting, and
- **fit** QDA with `MASS::qda()` and naive Bayes with `e1071::naiveBayes()`,
**interpret** their estimates, and **compare** classifiers by test error
rate, sensitivity, specificity, and AUC on a held-out test set.
:::

## Generative classifiers

### Credit card default data

**Question.** Do quadratic discriminant analysis (QDA) and naive Bayes, which
change LDA's model of the features in different directions, classify credit card customers more accurately than
LDA and logistic regression, and how does K-nearest neighbors (KNN) compare?

These customers were analyzed in the logistic regression lectures and, with
`balance` and `income` as the features, in the previous lecture, which modeled
each class's features by a Gaussian density with one covariance matrix
shared by both classes.

**Data.** `ISLR2::Default` is a simulated data set of
`r format(nrow(Default), big.mark = ",")` credit card customers. The response
is `default`, with `Yes` the positive class. The explanatory variables are
`balance`, the customer's monthly credit card balance in dollars, and
`income`, the customer's income in dollars; the features are the explanatory
variables themselves. `student` (Yes/No) enters only twice: as a third,
binary feature in the naive Bayes section's mixed feature types, and to split
the customers in the Student strata section.

The previous lecture's error rates were training error rates, optimistic on
average because each customer helped fit the model that classified it. Here,
as in the classification lecture's bias-variance script, the customers are
first split at random into halves: every model in this lecture is fit to the
training half and evaluated on the test half, whose customers it never saw.

::: {.callout-note collapse="true"}
### Code
```{r default-split}
#| output: false
# Its own seed, so the split does not depend on anything drawn before it.
set.seed(20261001)
train_id <- sample(nrow(Default), nrow(Default) / 2)
train <- Default[train_id, ]
test  <- Default[-train_id, ]

n_defaults <- c(train = sum(train$default == "Yes"),
                test  = sum(test$default == "Yes"))
```
:::

Each half holds `r format(nrow(train), big.mark = ",")` customers, with
`r n_defaults["train"]` defaulters in the training half and
`r n_defaults["test"]` in the test half. A different split would give somewhat
different test error rates; averaging over many splits is a topic of the
resampling unit.

The figure below plots income against balance for the training customers,
defaulters and non-defaulters distinguished by color and shape, with each
class's approximately 95% normal-theory ellipse (solid for non-defaulters,
dashed for defaulters), a contour of a Gaussian density with that class's
sample mean and covariance matrix.

::: {.callout-note collapse="true"}
### Code
```{r default-summaries}
#| output: false
class_summary <- train |>
  group_by(default) |>
  summarize(sd_balance = sd(balance), sd_income = sd(income),
            cor_balance_income = cor(balance, income), .groups = "drop")
summary_of <- function(k, col) class_summary[[col]][class_summary$default == k]
student_income <- function(s) median(train$income[train$student == s])

# Defaulters are drawn last so they are not hidden under non-defaulters.
scatter_points <- list(
  geom_point(data = arrange(train, default),
             aes(x = balance, y = income, color = default, shape = default),
             alpha = 0.35, size = 0.8),
  scale_color_manual(values = class_colors),
  scale_shape_manual(values = c(16, 17)),
  scale_x_continuous(labels = dollar_labels),
  scale_y_continuous(labels = dollar_labels),
  labs(x = "Credit card balance ($)", y = "Income ($)", color = "Default",
       shape = "Default")
)
eda_scatter <- ggplot() +
  scatter_points +
  stat_ellipse(data = train, aes(x = balance, y = income, color = default,
                                 linetype = default),
               type = "norm", level = 0.95, linewidth = 0.9) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  labs(linetype = "Default",
       title = "Training set, with each class's approximately 95% normal-theory ellipse")
```
:::

```{r default-eda-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
eda_scatter
```

The defaulters' ellipse is centered at higher balances and is narrower in the
balance direction: the standard deviation of balance is
`r dollar_text(summary_of("No", "sd_balance"))` among non-defaulters and
`r dollar_text(summary_of("Yes", "sd_balance"))` among defaulters. Income is
about equally spread in the two classes, with standard deviations of
`r dollar_text(summary_of("No", "sd_income"))` and
`r dollar_text(summary_of("Yes", "sd_income"))`.

Within each class balance and income are weakly correlated, with
$r = `r sprintf("%.3f", summary_of("No", "cor_balance_income"))`$ among
non-defaulters and
$r = `r sprintf("%.3f", summary_of("Yes", "cor_balance_income"))`$ among
defaulters, so both ellipses are only slightly tilted. Among non-defaulters
the points are densest in two horizontal bands, the bimodal income of the
previous lecture's histograms. The lower band is mostly students: in the
training half students' median income is
`r dollar_text(student_income("Yes"))`, against
`r dollar_text(student_income("No"))` for non-students. No Gaussian density
has two modes.

### Bayes' theorem and LDA

With the response in one of $K$ classes, $k = 1, \ldots, K$, the previous
lecture built a classifier from the prior probabilities $\pi_k = P(Y = k)$
and the class-conditional densities $f_k(x)$ of the features through Bayes'
theorem,

$$p_k(x) = P(Y = k \mid X = x) = \frac{\pi_k f_k(x)}{\sum_{l=1}^K \pi_l f_l(x)},$$

and assigned each observation to the class with the largest posterior
probability. Every generative classifier in this course estimates $\pi_k$ by
the class proportion $n_k/n$, where $n_k$ is the number of training
observations in class $k$ and $n$ the number of training observations. They
differ only in the model assumed for $f_k(x)$.

LDA assumed $X \mid Y = k \sim N_p\left(\mu_k, \Sigma\right)$, with one
covariance matrix $\Sigma$ shared by every class. That is what made the
quadratic term in $x$ cancel and the boundary linear. It would give the two
ellipses of the scatterplot the same shape, but the defaulters' is the
narrower in balance.

This lecture changes the model for $f_k(x)$ in two directions. Quadratic
discriminant analysis gives each class its own covariance matrix. Naive Bayes
drops the multivariate Gaussian and treats the features as independent within
each class.

## Quadratic discriminant analysis

The figure below repeats the training scatterplot of income against balance,
points colored and shaped by default status, with each class's approximately
95% normal-theory ellipse (solid for non-defaulters, dashed for defaulters).

```{r qda-data-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
eda_scatter
```

The two ellipses differ in width: the defaulters' is narrower in balance,
with standard deviation `r dollar_text(summary_of("Yes", "sd_balance"))`
against `r dollar_text(summary_of("No", "sd_balance"))`. One covariance matrix
shared by both classes cannot describe both clouds; QDA gives each class its
own.

### Class-specific covariance matrices

Quadratic discriminant analysis (QDA) assumes the statistical model

$$P(Y = k) = \pi_k, \qquad X \mid Y = k \sim N_p\left(\mu_k, \Sigma_k\right), \qquad k = 1, \ldots, K,$$

where

- $\mu_k = E[X \mid Y = k]$ is the $p$-vector of feature means in class $k$,
as in LDA, and
- $\Sigma_k = Cov[X \mid Y = k]$ is the $p \times p$ covariance matrix of the
features **in class $k$**, symmetric and positive definite, and free to differ
from class to class.

The class-conditional density is therefore

$$f_k(x) = \frac{1}{(2\pi)^{p/2}\left|\Sigma_k\right|^{1/2}} \exp\left[-\frac{1}{2}\left(x - \mu_k\right)^\top \Sigma_k^{-1} \left(x - \mu_k\right)\right],$$

where $\left|\Sigma_k\right|$ is the determinant of $\Sigma_k$. Each class's
contours are still ellipses centered at $\mu_k$, but now each class's
ellipses have their own size, shape, and orientation, as the two ellipses of
the scatterplot do.

### Class-specific covariance estimates

The priors and means are estimated as in LDA, $\hat\pi_k = n_k/n$ and
$\hat\mu_k = \frac{1}{n_k}\sum_{i:\, y_i = k} x_i$. Each covariance matrix is
estimated from its own class alone,

$$\hat\Sigma_k = \frac{1}{n_k - 1}\sum_{i:\, y_i = k}\left(x_i - \hat\mu_k\right)\left(x_i - \hat\mu_k\right)^\top,$$

where

- $x_i$ is the $p$-vector of feature values of observation $i$, as in the
previous lecture, and
- $\hat\Sigma_k$ is the **class-$k$ sample covariance matrix**, whose divisor
$n_k - 1$ accounts for the one estimated mean.

`MASS::qda()` uses this divisor (maximum likelihood would use $n_k$), and
$\hat\Sigma_k$, whose rank is at most $n_k - 1$, is invertible only if
$n_k > p$. It takes the same formula as `MASS::lda()`; fitting `default` on
`balance` and `income` in the training half:

```{r default-qda}
default_qda <- MASS::qda(default ~ balance + income, data = train)
default_qda
```

::: {.callout-note collapse="true"}
### Code
```{r qda-estimates}
#| output: false
qda_par <- map(c(No = "No", Yes = "Yes"), \(k) {
  X <- as.matrix(train[train$default == k, c("balance", "income")])
  list(prior = mean(train$default == k), mu = colMeans(X), Sigma = cov(X))
})
# A 2 x 2 matrix as LaTeX, entries rounded with LaTeX-safe thousands commas.
matrix_tex <- function(S) {
  e <- format(round(S), big.mark = "{,}", trim = TRUE)
  sprintf("\\begin{pmatrix} %s & %s \\\\ %s & %s \\end{pmatrix}",
          e[1, 1], e[1, 2], e[2, 1], e[2, 2])
}
```
:::

The output reports the prior probabilities and the class means of balance and
income, the same priors and means `MASS::lda()` reports for these training
customers; the priors differ slightly from the previous lecture's
all-customer fit, $\hat\pi_{\text{Yes}} = `r sprintf("%.4f", mean(Default$default == "Yes"))`$.
The covariance matrices are not printed. They are each class's sample
covariance matrix of (`balance`, `income`), in squared dollars,

$$\hat\Sigma_{\text{No}} = `r matrix_tex(qda_par$No$Sigma)`, \qquad \hat\Sigma_{\text{Yes}} = `r matrix_tex(qda_par$Yes$Sigma)`,$$

whose square-rooted diagonals and correlations are the standard deviations
and correlations quoted with the scatterplot, and whose contours are its
ellipses.

### Quadratic discriminant

As in LDA, the class with the largest posterior probability is the class with
the largest $\log\left[\pi_k f_k(x)\right]$,

$$\log\left[\pi_k f_k(x)\right] = \log \pi_k - \frac{p}{2}\log(2\pi) - \frac{1}{2}\log\left|\Sigma_k\right| - \frac{1}{2}\left(x - \mu_k\right)^\top \Sigma_k^{-1}\left(x - \mu_k\right).$$

Only the term $-\frac{p}{2}\log(2\pi)$ is common to every class. Dropping it
gives the **quadratic discriminant function**

$$\delta_k(x) = -\frac{1}{2}x^\top \Sigma_k^{-1} x + x^\top \Sigma_k^{-1}\mu_k - \frac{1}{2}\mu_k^\top \Sigma_k^{-1}\mu_k - \frac{1}{2}\log\left|\Sigma_k\right| + \log \pi_k,$$

after expanding the quadratic form as in the previous lecture. Two terms that
LDA dropped survive here:

- $-\frac{1}{2}x^\top \Sigma_k^{-1} x$ is quadratic in $x$ and carries a
different matrix in each class, and
- $-\frac{1}{2}\log\left|\Sigma_k\right|$ penalizes a class whose density is
spread over a larger volume.

When every $\Sigma_k$ equals a common $\Sigma$, both terms are the same in
every class, drop out, and $\delta_k(x)$ reduces to the LDA discriminant.

With $K = 2$ classes the log posterior odds of class $2$ are

$$\log\left(\frac{p_2(x)}{1 - p_2(x)}\right) = \delta_2(x) - \delta_1(x) = \beta_0 + x^\top\left(\Sigma_2^{-1}\mu_2 - \Sigma_1^{-1}\mu_1\right) - \frac{1}{2}x^\top \left(\Sigma_2^{-1} - \Sigma_1^{-1}\right) x,$$

where

- $\beta_0 = \log(\pi_2/\pi_1) - \frac{1}{2}\left(\mu_2^\top \Sigma_2^{-1}\mu_2 - \mu_1^\top \Sigma_1^{-1}\mu_1\right) - \frac{1}{2}\log\left(\left|\Sigma_2\right| / \left|\Sigma_1\right|\right)$
is the intercept, collecting the terms that do not involve $x$.

The log posterior odds are a quadratic function of the features: an
intercept, $p$ linear terms $x_j$, where $x_j$ is the $j$th entry of $x$, and
$p(p+1)/2$ squares $x_j^2$ and cross-products $x_j x_l$, with coefficients
determined by the $\mu_k$ and $\Sigma_k$.

::: {.callout-note collapse="true"}
### Code
```{r qda-discriminant}
#| output: false
S_no_inv  <- solve(qda_par$No$Sigma)
S_yes_inv <- solve(qda_par$Yes$Sigma)
m_no  <- qda_par$No$mu
m_yes <- qda_par$Yes$mu
A_qda <- -0.5 * (S_yes_inv - S_no_inv)
b_qda <- drop(S_yes_inv %*% m_yes - S_no_inv %*% m_no)
c_qda <- log(qda_par$Yes$prior / qda_par$No$prior) -
  0.5 * (sum(m_yes * (S_yes_inv %*% m_yes)) - sum(m_no * (S_no_inv %*% m_no))) -
  0.5 * log(det(qda_par$Yes$Sigma) / det(qda_par$No$Sigma))
balance_sq_coef <- A_qda[1, 1]

# Log posterior odds of Yes from the coefficients, checked against predict().
X_train <- as.matrix(train[c("balance", "income")])
qda_log_odds <- c_qda + drop(X_train %*% b_qda) + rowSums((X_train %*% A_qda) * X_train)
max_diff_qda <- max(abs(plogis(qda_log_odds) -
                        predict(default_qda)$posterior[, "Yes"]))

# Income in thousands of dollars, so the coefficients have readable sizes.
qda_coef_table <- tibble(
  term = c("Intercept $\\hat\\beta_0$", "$\\texttt{balance}$",
           "$\\texttt{income}$, per \\$1,000", "$\\texttt{balance}^2$",
           "$\\texttt{balance} \\times \\texttt{income}$, per \\$1,000",
           "$\\texttt{income}^2$, per (\\$1,000)$^2$"),
  value = c(c_qda, b_qda[1], 1000 * b_qda[2], A_qda[1, 1],
            2 * 1000 * A_qda[1, 2], 1e6 * A_qda[2, 2])
) |>
  mutate(value = map_chr(value, \(v) paste0("$", num_tex(v), "$"))) |>
  knitr::kable(col.names = c("Term", "Coefficient"), align = "lc")
```
:::

For the credit card fit, the table gives the coefficients of QDA's estimated
log posterior odds of default, with income terms per \$1,000 of income.

```{r qda-discriminant-table}
#| echo: false
qda_coef_table
```

The intercept is the log posterior odds of default for a customer with no
balance and no income, an extrapolation, since the lowest recorded income is
`r dollar_text(min(Default$income))`. With squared and cross-product terms, no
single coefficient gives a feature's effect: as in the polynomial logistic
regressions of the flexible logistic regression lecture, the effect of
balance depends on the balance and income at which it is evaluated.
Applying the logistic function to these log posterior odds reproduces
`predict()`'s posterior probabilities to within $`r sci(max_diff_qda)`$.

### Quadratic decision boundary

The decision boundary, where the log posterior odds are zero, is a conic
section when $p = 2$ instead of a line. With one feature, writing
$\sigma_k^2$ for the class-$k$ variance, the log posterior odds are

$$\log\left(\frac{p_2(x)}{1 - p_2(x)}\right) = \frac{1}{2}\left(\frac{1}{\sigma_1^2} - \frac{1}{\sigma_2^2}\right)x^2 + \left(\frac{\mu_2}{\sigma_2^2} - \frac{\mu_1}{\sigma_1^2}\right)x + \frac{\mu_1^2}{2\sigma_1^2} - \frac{\mu_2^2}{2\sigma_2^2} + \log\frac{\sigma_1}{\sigma_2} + \log\frac{\pi_2}{\pi_1}.$$

A quadratic in $x$ has up to two roots, so the boundary is up to two points.
If $\sigma_1 > \sigma_2$, the coefficient of $x^2$ is negative, the log
posterior odds fall without bound in both directions, and class $1$, the
class with the larger variance, is assigned in both tails. For the credit
card customers class $1$ is `No`, whose balances have the larger standard
deviation.

LDA (`MASS::lda()`) is refit to the same training half for contrast. The
figure below draws the QDA boundary (orange, dotted) and the LDA boundary
(green, solid) over the training customers, non-defaulters as grey circles
and defaulters as black triangles, with balances beyond the largest training
balance shaded.

::: {.callout-note collapse="true"}
### Code
```{r qda-boundary}
#| output: false
default_lda <- MASS::lda(default ~ balance + income, data = train)

# Each model's posterior probability of default; a model's object is looked
# up only when that model is requested, so later models can be added.
posterior_of <- function(model, d) {
  switch(model,
         "Logistic regression" = unname(predict(default_glm, newdata = d,
                                                type = "response")),
         LDA = predict(default_lda, newdata = d)$posterior[, "Yes"],
         QDA = predict(default_qda, newdata = d)$posterior[, "Yes"],
         "Naive Bayes" = predict(default_nb, newdata = d, type = "raw")[, "Yes"],
         KNN = knn_yes(d, k_default))
}

# Balances at which a model's posterior crosses 0.5 at the training income
# quartiles.
income_levels <- quantile(train$income, c(0.25, 0.5, 0.75))
crossings <- function(model) {
  gap <- function(b, inc) posterior_of(model, tibble(balance = b, income = inc)) - 0.5
  map(income_levels, \(inc) {
    # Searching past about $10,000 would fail: e1071's naive Bayes posteriors
    # underflow between about $15,000 and $18,000 of balance, which would trip
    # the exactly-two-crossings check below.
    b <- seq(0, 10000, by = 10)
    j <- which(diff(sign(gap(b, inc))) != 0)
    map_dbl(j, \(i) uniroot(gap, b[c(i, i + 1)], inc = inc)$root)
  })
}
qda_cross <- crossings("QDA")
cross_at <- function(cr, i) map_dbl(cr, i)
# The prose describes two crossings per income quartile, the second beyond
# the training data, and a negative balance^2 coefficient.
stopifnot(balance_sq_coef < 0, all(lengths(qda_cross) == 2),
          all(cross_at(qda_cross, 2) > max(train$balance)))

# Training customers under the 0.5 contour of each requested model, with
# balances beyond the largest training balance shaded. The grid reaches past
# the largest second crossing, so the second branches are always drawn.
boundary_figure <- function(models, title, second = cross_at(qda_cross, 2)) {
  grid <- expand_grid(
    balance = seq(0, max(4500, 1.15 * max(second)), length.out = 181),
    income = seq(min(Default$income), max(Default$income), length.out = 121))
  curves <- map(models, \(m) {
    mutate(grid, model = m, p_hat = posterior_of(m, grid))
  }) |>
    list_rbind() |>
    mutate(model = factor(model, levels = names(model_colors)))
  # Points are filled grey or black by default status, so that color is free
  # to identify the models' boundaries.
  ggplot(mapping = aes(x = balance, y = income)) +
    annotate("rect", xmin = max(train$balance), xmax = Inf, ymin = -Inf,
             ymax = Inf, fill = "grey92") +
    geom_point(data = arrange(train, default),
               aes(shape = default, fill = default), color = "grey40",
               stroke = 0.2, size = 1.1, alpha = 0.6) +
    geom_contour(data = curves, aes(z = p_hat, color = model, linetype = model),
                 breaks = 0.5, linewidth = 0.9) +
    scale_shape_manual(values = c(21, 24)) +
    scale_fill_manual(values = c("grey70", "black")) +
    scale_color_manual(values = model_colors) +
    scale_linetype_manual(values = model_linetypes) +
    scale_x_continuous(labels = dollar_labels) +
    scale_y_continuous(labels = dollar_labels) +
    guides(shape = guide_legend(order = 1), fill = guide_legend(order = 1),
           color = guide_legend(order = 2), linetype = guide_legend(order = 2)) +
    labs(x = "Credit card balance ($)", y = "Income ($)", shape = "Default",
         fill = "Default", color = "Model", linetype = "Model", title = title)
}
qda_boundary_plot <- boundary_figure(
  c("LDA", "QDA"), "QDA and LDA boundaries; shading beyond the largest training balance")
```
:::

```{r qda-boundary-figure}
#| echo: false
#| fig-width: 7.5
#| fig-height: 4.5
qda_boundary_plot
```

Within the training balances the QDA boundary runs close to LDA's nearly
vertical line, so income barely moves either model's balance threshold there.
Beyond them QDA's boundary has a second branch, where its posterior
probability of default falls back below $0.5$.

The second branch is easiest to see with one feature. Fitting QDA to balance
alone, the first tab below plots the prior-weighted fitted densities
$\hat\pi_k \hat f_k(x)$ with the class-specific standard deviations, over
shaded histograms of each class's training balances on the same scale, outlined
in each class's line type, and
the second plots the two estimated discriminant functions
$\hat\delta_k(x)$. Dotted lines mark the boundary points, and balances
beyond the largest training balance are shaded grey.

::: {.callout-note collapse="true"}
### Code
```{r qda-one-feature}
#| output: false
one_feature <- train |>
  group_by(default) |>
  summarize(prior = n() / nrow(train), mu = mean(balance), sd = sd(balance),
            .groups = "drop")
of_class <- function(k, col) one_feature[[col]][one_feature$default == k]

# Boundary points: roots of the one-feature log posterior odds of Yes.
m1 <- of_class("No", "mu")
s1 <- of_class("No", "sd")
m2 <- of_class("Yes", "mu")
s2 <- of_class("Yes", "sd")
log_prior_odds <- log(of_class("Yes", "prior") / of_class("No", "prior"))
roots_qda_1 <- sort(Re(polyroot(c(
  m1^2 / (2 * s1^2) - m2^2 / (2 * s2^2) + log(s1 / s2) + log_prior_odds,
  m2 / s2^2 - m1 / s1^2,
  (1 / s1^2 - 1 / s2^2) / 2))))

balance_grid_1 <- tibble(balance = seq(0, max(4500, 1.15 * roots_qda_1[2]),
                                        length.out = 600))
one_feature_curves <- one_feature |>
  cross_join(balance_grid_1) |>
  mutate(weighted = prior * dnorm(balance, mu, sd),
         delta = log(prior) - log(sd) - (balance - mu)^2 / (2 * sd^2))
one_feature_plot <- function(y, y_label, title) {
  ggplot(one_feature_curves,
         aes(x = balance, y = .data[[y]], color = default, linetype = default)) +
    annotate("rect", xmin = max(train$balance), xmax = Inf, ymin = -Inf,
             ymax = Inf, fill = "grey92") +
    geom_line(linewidth = 0.9) +
    geom_vline(xintercept = roots_qda_1, linetype = "dotted") +
    scale_color_manual(values = class_colors) +
    scale_x_continuous(labels = dollar_labels) +
    labs(x = "Credit card balance ($)", y = y_label, color = "Default",
         linetype = "Default", title = title)
}
# Histograms of training balances on the same scale as pi_k f_k: a bar's
# count divided by n times the bin width.
bin_width_1 <- 100
qda_weighted_1 <- one_feature_plot(
  "weighted", expression(hat(pi)[k] * hat(f)[k](x)),
  expression(paste("QDA prior-weighted densities, class-specific ", hat(sigma)[k]))) +
  geom_histogram(data = train,
                 aes(x = balance, y = after_stat(count) / (nrow(train) * bin_width_1),
                     fill = default, color = default, linetype = default),
                 binwidth = bin_width_1, boundary = 0, position = "identity",
                 alpha = 0.3, linewidth = 0.3, inherit.aes = FALSE) +
  scale_fill_manual(values = class_colors) +
  labs(fill = "Default")
# Vertical range from the curves' maximum down past the second crossing.
delta_at <- function(b) {
  log(of_class("No", "prior")) - log(s1) - (b - m1)^2 / (2 * s1^2)
}
delta_range <- c(delta_at(roots_qda_1[2]) - 8, max(one_feature_curves$delta) + 1)
qda_delta_1 <- one_feature_plot("delta", expression(hat(delta)[k](x)),
                                "QDA discriminant functions") +
  coord_cartesian(ylim = delta_range)
```
:::

::: {.panel-tabset}
#### Prior-weighted densities
```{r qda-one-qda-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
qda_weighted_1
```

#### Discriminant functions
```{r qda-one-delta-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
qda_delta_1
```
:::

The fitted densities follow the two histograms, apart from the non-defaulters'
spike at zero balance. They cross at `r dollar_text(roots_qda_1[1])` and
again at `r dollar_text(roots_qda_1[2])`, where both are too small to see; the
discriminant functions, two downward parabolas, visibly cross a second time.
Past that point the wider non-defaulter parabola is the higher one again, so
`No` is assigned in both tails. With the previous lecture's one shared
standard deviation, the two discriminant functions were straight lines that
cross once.

With both features the same happens at every income. The coefficient of
$\texttt{balance}^2$ in the table above is negative, because balance varies
less among defaulters at a given income, so QDA's posterior probability of
default eventually falls as balance grows. At the lower quartile, median, and
upper quartile of training income, `r dollar_text(income_levels[1])`,
`r dollar_text(income_levels[2])`, and `r dollar_text(income_levels[3])`,
QDA's posterior crosses $0.5$ at balances of
`r dollar_text(cross_at(qda_cross, 1)[1])`,
`r dollar_text(cross_at(qda_cross, 1)[2])`, and
`r dollar_text(cross_at(qda_cross, 1)[3])` and again at
`r dollar_text(cross_at(qda_cross, 2)[1])`,
`r dollar_text(cross_at(qda_cross, 2)[2])`, and
`r dollar_text(cross_at(qda_cross, 2)[3])`, beyond the largest training
balance, `r dollar_text(max(train$balance))`. The second branch is an
extrapolation of the Gaussian tails`r if (min(cross_at(qda_cross, 2)) > max(Default$balance)) ": no customer in either half has a balance that large" else ""`.

### Bias-variance tradeoff

QDA's flexibility costs parameters. Not counting the $K - 1$ free priors,
which every generative classifier estimates,

$$\text{LDA: } Kp + \frac{p(p+1)}{2}, \qquad \text{QDA: } Kp + K\,\frac{p(p+1)}{2},$$

where

- $Kp$ counts the entries of the $K$ mean vectors, and
- $p(p+1)/2$ counts the distinct entries of one symmetric $p \times p$
covariance matrix: $p$ variances and $p(p-1)/2$ covariances.

::: {.callout-note collapse="true"}
### Code
```{r parameter-counts}
#| output: false
count_lda <- function(K, p) K * p + p * (p + 1) / 2
count_qda <- function(K, p) K * p + K * p * (p + 1) / 2

parameter_table <- tibble(p = c(1, 2, 5, 10, 50)) |>
  mutate(LDA = count_lda(2, p), QDA = count_qda(2, p),
         across(c(LDA, QDA), \(v) format(v, big.mark = ","))) |>
  knitr::kable(col.names = c("Features $p$", "LDA", "QDA"), align = "ccc")
```
:::

With $K = 2$ classes:

```{r parameter-table}
#| echo: false
parameter_table
```

(The previous lecture's count of `r 1 + count_lda(2, 2)` for LDA with balance
and income included its one free prior.) QDA's count grows like $Kp^2/2$,
against LDA's $p^2/2$. When the
$\Sigma_k$ really differ, LDA's shared $\Sigma$ makes its boundary the wrong
shape however much data it sees: bias. QDA's boundary has the right shape, but
each of its many covariance estimates comes from one class's observations and
varies from one training set to the next: variance. For the credit card
customers, QDA estimates the defaulters' `r 2 * (2 + 1) / 2`
covariance entries from the `r n_defaults["train"]` training defaulters
alone, while LDA's pooled covariance uses all
`r format(nrow(train), big.mark = ",")` training customers.

::: {.callout-note collapse="true"}
### Code
```{r boundary-variability}
#| output: false
mu_a    <- c(0, 0)
mu_b    <- c(1.5, 0.5)
Sigma_a <- matrix(c(1, 0.5, 0.5, 1), nrow = 2)
Sigma_b <- matrix(c(3, -0.6, -0.6, 2), nrow = 2)

rmvnorm2 <- function(n, mu, Sigma) {
  sweep(matrix(rnorm(n * 2), n) %*% chol(Sigma), 2, mu, "+")
}
draw_quadratic <- function(n_k) {
  X <- rbind(rmvnorm2(n_k, mu_a, Sigma_a), rmvnorm2(n_k, mu_b, Sigma_b))
  tibble(y = factor(rep(c("k = 1", "k = 2"), each = n_k)),
         x1 = X[, 1], x2 = X[, 2])
}

grid_fit <- expand_grid(x1 = seq(-3.5, 5.5, length.out = 120),
                        x2 = seq(-3.5, 4, length.out = 110))
bayes_grid <- expand_grid(x1 = seq(-3.5, 5.5, length.out = 220),
                          x2 = seq(-3.5, 4, length.out = 200)) |>
  mutate(qda = log_odds_gauss(x1, x2, mu_a, Sigma_a, mu_b, Sigma_b))

n_k_small <- 25
n_k_large <- 250
n_boundary_reps <- 20

fitted_boundaries <- expand_grid(n_k = c(n_k_small, n_k_large),
                                 rep = seq_len(n_boundary_reps)) |>
  mutate(data = map(n_k, draw_quadratic),
         LDA = map(data, \(d) {
           predict(MASS::lda(y ~ x1 + x2, data = d),
                   newdata = grid_fit)$posterior[, "k = 2"]
         }),
         QDA = map(data, \(d) {
           predict(MASS::qda(y ~ x1 + x2, data = d),
                   newdata = grid_fit)$posterior[, "k = 2"]
         })) |>
  select(-data) |>
  pivot_longer(c(LDA, QDA), names_to = "method", values_to = "p_hat") |>
  mutate(grid = list(grid_fit)) |>
  unnest(c(grid, p_hat))

variability_plot <- function(nk) {
  ggplot(filter(fitted_boundaries, n_k == nk), aes(x = x1, y = x2)) +
    geom_contour(aes(z = p_hat - 0.5, group = rep), breaks = 0,
                 color = "grey55", linewidth = 0.35) +
    geom_contour(data = bayes_grid, aes(z = qda), breaks = 0, color = "black",
                 linewidth = 1.1, linetype = "longdash") +
    facet_wrap(~ method) +
    coord_fixed(xlim = c(-3.5, 5.5), ylim = c(-3.5, 4), expand = FALSE) +
    labs(x = expression(X[1]), y = expression(X[2]),
         title = bquote(paste(.(n_boundary_reps), " fitted boundaries, ",
                              n[k] == .(nk),
                              " per class; Bayes boundary long-dashed")))
}
variability_small <- variability_plot(n_k_small)
variability_large <- variability_plot(n_k_large)
cor_of <- function(S) S[1, 2] / sqrt(S[1, 1] * S[2, 2])
```
:::

The figure below compares fits from repeated training sets against a known
Bayes decision boundary, which the credit card data cannot supply, so it uses
simulated data: two classes with known Gaussian densities and equal priors,
class 1 with mean $(`r mu_a[1]`, `r mu_a[2]`)^\top$, unit variances, and
correlation `r sprintf("%.2f", cor_of(Sigma_a))`, and class 2 with mean
$(`r mu_b[1]`, `r mu_b[2]`)^\top$, variances `r Sigma_b[1, 1]` and
`r Sigma_b[2, 2]`, and correlation `r sprintf("%.2f", cor_of(Sigma_b))`. Each
panel overlays the boundaries of LDA (left) or QDA (right) fit to each of
`r n_boundary_reps` training sets (thin grey) on the Bayes decision boundary
(thick black, long dashes), the QDA boundary computed from the true
parameters. The first tab uses `r n_k_small` observations per class and the
second `r n_k_large`.

::: {.panel-tabset}
#### Small training sets
```{r variability-small-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 4
variability_small
```

#### Large training sets
```{r variability-large-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 4
variability_large
```
:::

With the small training sets the LDA lines vary in angle and none follows the
curvature of the Bayes boundary, while the QDA curves follow its shape but
scatter widely. With the large ones the LDA lines tighten around a single line
that still misses the curvature, while the QDA curves collapse onto the Bayes
boundary: more data reduces QDA's variance but not LDA's bias. LDA should
therefore win when the training set is small relative to the number of
covariance parameters or the class covariance matrices are close to equal,
and QDA when they clearly differ and the training set is large enough to
estimate them.

::: {.callout-note collapse="true"}
### Beyond this course
Regularized discriminant analysis compromises between the two by shrinking
each class covariance toward the pooled one,
$\hat\Sigma_k(\alpha) = \alpha\hat\Sigma_k + (1 - \alpha)\hat\Sigma$, where
$\alpha = 0$ gives LDA and $\alpha = 1$ gives QDA. The value of $\alpha$ is a
flexibility dial chosen from the data, as the next unit will choose the
number of neighbors for KNN.
:::

## Naive Bayes

The figure below shows the training scatterplot once more, points colored and
shaped by default status, with each class's approximately 95% normal-theory
ellipse (solid for non-defaulters, dashed for defaulters).

```{r nb-data-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 4.5
eda_scatter
```

The ellipses tilt only slightly: the within-class correlations are
$`r sprintf("%.3f", summary_of("No", "cor_balance_income"))`$ and
$`r sprintf("%.3f", summary_of("Yes", "cor_balance_income"))`$,
`r if (max(abs(class_summary$cor_balance_income)) < 0.3) "both small" else "not large"`
in magnitude. A model that treats balance and income as independent within
each class ignores only that tilt.

### Conditional independence

Naive Bayes keeps the generative framework but replaces the multivariate model
for $f_k$ with an assumption about how the features relate to one another:

$$P(Y = k) = \pi_k, \qquad X_1, \ldots, X_p \text{ are independent given } Y = k, \qquad X_j \mid Y = k \sim f_{kj},$$

where

- $X_j$ is the $j$th feature, the $j$th entry of the random vector $X$, and
- $f_{kj}$ is the one-feature density (or probability mass function) of $X_j$
among observations in class $k$.

Independence turns the joint density into a product of its marginals, so the
class-conditional density is

$$f_k(x) = f_{k1}(x_1) \times f_{k2}(x_2) \times \cdots \times f_{kp}(x_p) = \prod_{j=1}^p f_{kj}(x_j).$$

The posterior probability is Bayes' theorem with this product in place of
$f_k(x)$. The assumption is conditional on the class: features that are
related only because both differ between classes can be independent within
each class while being correlated overall.

Each $f_{kj}$ describes a single feature in a single class, so each can be
modeled and estimated separately, and features of different types can use
densities of different types.

- A quantitative feature is usually modeled as Gaussian,
$X_j \mid Y = k \sim N\left(\mu_{kj}, \sigma_{kj}^2\right)$, with $\mu_{kj}$
and $\sigma_{kj}^2$ estimated by the sample mean and variance of feature $j$
among the class-$k$ training observations.
- A binary feature is Bernoulli,
$f_{kj}(x_j) = \theta_{kj}^{x_j}\left(1 - \theta_{kj}\right)^{1 - x_j}$ for
$x_j \in \{0, 1\}$, where $\theta_{kj} = P(X_j = 1 \mid Y = k)$ is estimated
by the proportion of class-$k$ training observations with $x_j = 1$.
- A categorical feature with more than two levels gets one probability per
level in each class, estimated by the class-$k$ proportions at each level.

### Naive Bayes fit

`e1071::naiveBayes()` takes the same formula and models each numeric
explanatory variable with a Gaussian density in each class. Fitting it to the
training half:

```{r default-nb}
default_nb <- e1071::naiveBayes(default ~ balance + income, data = train)
default_nb
```

The header "Naive Bayes Classifier for Discrete Predictors" is `e1071`'s fixed
label, printed although both features here are continuous. The
**A-priori probabilities** are $\hat\pi_k$, the same as QDA's. Under
**Conditional probabilities**, the `balance` and `income` tables give each
class's mean (first column) and standard deviation (second column), the
Gaussian $\hat\mu_{kj}$ and $\hat\sigma_{kj}$.

### Additive log posterior odds

With $K = 2$, the logarithm of the product is a sum, and the log posterior
odds of class $2$ are

$$\log\left(\frac{p_2(x)}{1 - p_2(x)}\right) = \log\frac{\pi_2}{\pi_1} + \sum_{j=1}^p \log\frac{f_{2j}(x_j)}{f_{1j}(x_j)} = \log\frac{\pi_2}{\pi_1} + \sum_{j=1}^p g_j(x_j),$$

where

- $g_j(x_j) = \log\left(f_{2j}(x_j)/f_{1j}(x_j)\right)$ is the contribution of
feature $j$, a function of $x_j$ alone.

Each feature shifts the log posterior odds by an amount that does not depend
on the values of the other features: naive Bayes is an additive model with no
interactions, the same structural restriction as the additive logistic
regressions of the multiple logistic regression lecture. The form of each
$g_j$ follows from the type of $f_{kj}$:

- Gaussian with class-specific variances: $g_j$ is quadratic in $x_j$, with
coefficient $\frac{1}{2}\left(1/\sigma_{1j}^2 - 1/\sigma_{2j}^2\right)$ on
$x_j^2$.
- Gaussian with a variance shared by the two classes: the $x_j^2$ terms
cancel and $g_j$ is linear in $x_j$.
- Bernoulli: $g_j(x_j) = x_j\log\left(\theta_{2j}/\theta_{1j}\right) + \left(1 - x_j\right)\log\left[\left(1 - \theta_{2j}\right)/\left(1 - \theta_{1j}\right)\right]$,
linear in $x_j$, so the feature shifts the log posterior odds by
$\log\left\{\left[\theta_{2j}/\left(1 - \theta_{2j}\right)\right] / \left[\theta_{1j}/\left(1 - \theta_{1j}\right)\right]\right\}$
when $x_j$ changes from $0$ to $1$.

For the credit card fit both terms are Gaussian with class-specific variances,
so the balance term and the income term are each quadratic, and there is no
balance-by-income term.

### Mixed feature types

A binary feature enters naive Bayes through its own probability table. Adding
`student` as a third feature, `naiveBayes()` keeps the Gaussian densities for
balance and income and estimates, for each class, the proportion of students
and non-students; its `student` table is printed below.

```{r nb-mixed}
mixed_nb <- e1071::naiveBayes(default ~ balance + income + student, data = train)
mixed_nb$tables$student
```

The second column holds $\hat\theta_k$ (the feature index of $\theta_{kj}$
dropped, since `student` is the only binary feature), the proportion of
students among the training customers of class $k$:
`r sprintf("%.3f", mixed_nb$tables$student["No", "Yes"])` among
non-defaulters and `r sprintf("%.3f", mixed_nb$tables$student["Yes", "Yes"])`
among defaulters. By the Bernoulli term above, being a student
`r if (qlogis(mixed_nb$tables$student["Yes", "Yes"]) > qlogis(mixed_nb$tables$student["No", "Yes"])) "raises" else "lowers"`
the log posterior odds of default by
`r sprintf("%.3f", abs(qlogis(mixed_nb$tables$student["Yes", "Yes"]) - qlogis(mixed_nb$tables$student["No", "Yes"])))`
at every balance and income. No Gaussian assumption is made for `student`,
but conditional independence still is: the term ignores that students have
much lower incomes than non-students.

### Diagonal covariance

When every $f_{kj}$ is Gaussian, the product of $p$ independent normal
densities is a multivariate normal density with a diagonal covariance matrix:

$$\prod_{j=1}^p \frac{1}{\sqrt{2\pi}\,\sigma_{kj}}\exp\left[-\frac{\left(x_j - \mu_{kj}\right)^2}{2\sigma_{kj}^2}\right] = \frac{1}{(2\pi)^{p/2}\left|\Lambda_k\right|^{1/2}}\exp\left[-\frac{1}{2}\left(x - \mu_k\right)^\top \Lambda_k^{-1}\left(x - \mu_k\right)\right],$$

where

- $\Lambda_k = \text{diag}\left(\sigma_{k1}^2, \ldots, \sigma_{kp}^2\right)$ is the
diagonal matrix of class-$k$ feature variances, whose determinant is
$\prod_j \sigma_{kj}^2$ and whose inverse is diagonal with entries
$1/\sigma_{kj}^2$.

Gaussian naive Bayes with class-specific variances is therefore QDA with
every $\Sigma_k$ restricted to be diagonal, and Gaussian naive Bayes with one
variance per feature shared across classes is LDA with $\Sigma$ restricted to
be diagonal.

::: {.callout-note collapse="true"}
### Code
```{r nb-diagonal}
#| output: false
nb_sd <- function(k) c(default_nb$tables$balance[k, 2], default_nb$tables$income[k, 2])
max_diff_sd <- max(map_dbl(c("No", "Yes"), \(k) {
  max(abs(nb_sd(k) - sqrt(diag(qda_par[[k]]$Sigma))))
}))
```
:::

For the credit card customers, the naive Bayes means are QDA's and its
standard deviations are `r if (max_diff_sd == 0) "exactly " else ""`the square
roots of the diagonals of QDA's
$\hat\Sigma_k$`r if (max_diff_sd == 0) "" else if (max_diff_sd < 1e-6) ", up to rounding error" else paste0(", to within $", sci(max_diff_sd), "$ dollars")`: $\hat\Lambda_k$ is
$\hat\Sigma_k$ with the within-class correlation, $`r sprintf("%.3f", summary_of("No", "cor_balance_income"))`$
or $`r sprintf("%.3f", summary_of("Yes", "cor_balance_income"))`$, set to zero.

### Parameter count

Gaussian naive Bayes with class-specific variances estimates $Kp$ means and
$Kp$ variances, $2Kp$ parameters, against QDA's
$Kp + Kp(p+1)/2$. With $K = 2$ and $p = 50$ that is
`r format(2 * 2 * 50, big.mark = ",")` parameters against
`r format(count_qda(2, 50), big.mark = ",")`. The count grows linearly in
$p$, and each parameter is estimated from one feature in one class, using all
$n_k$ observations of that class. For balance and income, naive Bayes
estimates `r 2 * 2 * 2` means and variances against QDA's
`r count_qda(2, 2)` means and covariance entries.

For $p$ binary features the saving is larger still: a joint probability mass
function has $2^p$ cells per class ($2^p - 1$ free probabilities, or
`r format(2^20 - 1, big.mark = ",")` when $p = 20$), most of them never
observed in a training set, against naive Bayes's $p$ Bernoulli
probabilities, which sidesteps the curse of dimensionality of the flexibility
lecture.

The price is bias whenever the features are dependent within a class, which
in practice is almost always. Only the side of the threshold matters, so a
biased estimate of $f_k(x)$ can still produce the Bayes classifier's
classifications.

::: {.callout-note collapse="true"}
### Beyond this course
A quantitative feature need not be modeled as Gaussian. Each $f_{kj}$ can
instead be estimated nonparametrically, by a histogram or a kernel density
estimate of feature $j$ within class $k$, since each is a one-dimensional
density. The `naivebayes` package offers this option.
:::

### Naive Bayes boundary

The figure below draws the naive Bayes boundary (blue, dot-dash) beside the
QDA boundary (orange, dotted) over the training customers, non-defaulters as
grey circles and defaulters as black triangles, with balances beyond the
largest training balance shaded.

::: {.callout-note collapse="true"}
### Code
```{r nb-boundary}
#| output: false
nb_cross <- crossings("Naive Bayes")
stopifnot(all(lengths(nb_cross) == 2),
          all(cross_at(nb_cross, 2) > max(train$balance)))
second_all <- c(cross_at(qda_cross, 2), cross_at(nb_cross, 2))
nb_boundary_plot <- boundary_figure(
  c("QDA", "Naive Bayes"),
  "Naive Bayes and QDA boundaries; shading beyond the largest training balance",
  second_all)
# How each model's balance threshold moves from the lower to the upper income
# quartile, and the largest gap between the two at the three quartiles.
first_qda <- cross_at(qda_cross, 1)
first_nb  <- cross_at(nb_cross, 1)
tilt <- function(v) if (v[3] < v[1]) "falls" else "rises"
nb_tilt <- if (tilt(first_nb) == tilt(first_qda)) paste("also", tilt(first_nb)) else tilt(first_nb)
max_first_gap <- max(abs(first_qda - first_nb))

# Naive Bayes's income term at the median income, per $1,000, against LDA's
# income slope (pooled-covariance coefficient), per $1,000.
nb_income <- default_nb$tables$income
g_income <- function(x) {
  dnorm(x, nb_income["Yes", 1], nb_income["Yes", 2], log = TRUE) -
    dnorm(x, nb_income["No", 1], nb_income["No", 2], log = TRUE)
}
nb_income_slope <- g_income(median(train$income) + 500) -
  g_income(median(train$income) - 500)
income_gap <- nb_income["Yes", 1] - nb_income["No", 1]
pooled <- Reduce(`+`, map(c("No", "Yes"), \(k) {
  crossprod(scale(X_train[train$default == k, ], scale = FALSE))
})) / (nrow(train) - 2)
mean_diff <- c(diff(default_nb$tables$balance[, 1]), income_gap)
lda_income_slope <- 1000 * solve(pooled, mean_diff)[2]
raises <- function(v) if (v > 0) "raises" else "lowers"
# Two significant digits without rounding a small value to 0.0000.
small_num <- function(v) sig_text(v, 2)
```
:::

```{r nb-boundary-figure}
#| echo: false
#| fig-width: 7.5
#| fig-height: 4.5
nb_boundary_plot
```

Within the training balances the two boundaries stay within
`r dollar_text(max_first_gap)` of each other at the three income quartiles.
As income rises from the lower to the upper quartile, QDA's balance threshold
`r tilt(first_qda)` from `r dollar_text(first_qda[1])` to
`r dollar_text(first_qda[3])` and naive Bayes's `r nb_tilt` from
`r dollar_text(first_nb[1])` to `r dollar_text(first_nb[3])`. Naive Bayes's balance term
is quadratic for the same reason as QDA's, so its boundary also has a second
branch beyond the training balances, crossing $0.5$ at
`r dollar_text(cross_at(nb_cross, 2)[1])`,
`r dollar_text(cross_at(nb_cross, 2)[2])`, and
`r dollar_text(cross_at(nb_cross, 2)[3])` at the three income quartiles.

Naive Bayes drops the within-class correlation, so its income term compares
the two classes' income distributions without adjusting for balance:
defaulters' mean training income is `r dollar_text(abs(income_gap))`
`r if (income_gap > 0) "higher" else "lower"` than non-defaulters', and at the
median income each additional \$1,000 `r raises(nb_income_slope)` naive
Bayes's log posterior odds by `r small_num(abs(nb_income_slope))`. The term is
slightly curved, because the two classes' income standard deviations differ.
LDA's income slope, adjusted for balance through the pooled within-class
covariance as the [previous lecture](50-lda.qmd#balance-and-income)
explained, `r raises(lda_income_slope)` them by
`r small_num(abs(lda_income_slope))` per \$1,000,
`r if (sign(nb_income_slope) == sign(lda_income_slope)) "the same direction" else "the opposite direction"`.

## Classifier comparison

Five classifiers are now available: logistic regression, LDA, QDA, naive
Bayes, and KNN. From here to the end of Classifier comparison, $K$ is the
number of neighbors rather than the number of classes: every comparison has
two classes, and the number of classes is written as $2$.

### KNN classification

The flexibility lecture introduced KNN for a quantitative response. For a
qualitative response the neighbors vote:

$$\hat p_{\text{Yes}}(x_0) = \frac{1}{K}\sum_{i \in \mathcal{N}_0}\mathrm{I}(y_i = \text{Yes}),$$

where

- $x_0$ is the feature vector, here (`balance`, `income`), of the customer
being classified,
- $K$ is the **number of neighbors**,
- $\mathcal{N}_0$ is the set of the $K$ training observations closest to $x_0$
in Euclidean distance after every feature, in the training data and in $x_0$
alike, is standardized by its training mean and standard deviation, since
that distance depends on each feature's scale (as in the flexibility
lecture), and
- $\hat p_{\text{Yes}}(x_0)$ is the proportion of those neighbors who
defaulted; the classifier assigns `Yes` when it exceeds $0.5$, a majority
vote.

This lecture fixes $K$ in advance at the smallest odd integer above
$\sqrt{n}$: odd, so that a two-class vote cannot tie, and near $\sqrt{n}$ as
a rule of thumb rather than an optimal choice, since choosing $K$ from the
data requires the resampling methods of the next unit. For the credit card
training half, $n = `r format(nrow(train), big.mark = "{,}")`$ gives
$K = `r odd_above_sqrt(nrow(train))`$.

### Fitted boundaries

::: {.callout-note collapse="true"}
### Code
```{r comparison-boundaries}
#| output: false
default_glm <- glm(default ~ balance + income, data = train,
                   family = "binomial")
k_default <- odd_above_sqrt(nrow(train))

# KNN vote proportion for "Yes", on features standardized by training means/SDs.
standardize <- function(d) {
  cbind((d$balance - mean(train$balance)) / sd(train$balance),
        (d$income - mean(train$income)) / sd(train$income))
}
knn_yes <- function(target, K) {
  fit <- class::knn(standardize(train), standardize(target), train$default,
                    k = K, prob = TRUE)
  ifelse(fit == "Yes", attr(fit, "prob"), 1 - attr(fit, "prob"))
}
all_boundaries_plot <- boundary_figure(
  c("Logistic regression", "LDA", "QDA", "Naive Bayes", "KNN"),
  "All five classifiers' boundaries; shading beyond the largest training balance",
  second_all)

# Incomes (in $1,000 steps) at which KNN flags no balance up to the largest
# training balance, as runs of consecutive grid incomes.
knn_check <- expand_grid(
  income = seq(1000 * floor(min(train$income) / 1000),
               1000 * ceiling(max(train$income) / 1000), by = 1000),
  balance = seq(0, max(train$balance), length.out = 200))
knn_check$yes <- knn_yes(knn_check, k_default) > 0.5
no_flag <- knn_check |>
  group_by(income) |>
  summarize(none = !any(yes), .groups = "drop")
runs <- rle(no_flag$none)
run_end <- cumsum(runs$lengths)
run_start <- run_end - runs$lengths + 1
# Run ends are clipped to the range of training incomes.
no_flag_runs <- tibble(from = pmax(no_flag$income[run_start[runs$values]],
                                   min(train$income)),
                       to = pmin(no_flag$income[run_end[runs$values]],
                                 max(train$income))) |>
  filter(from <= to)
range_text <- function(a, b) {
  if (a == b) paste("near", dollar_text(a)) else
    paste("from", dollar_text(a), "to", dollar_text(b))
}
no_flag_text <- if (nrow(no_flag_runs) == 0) "nowhere" else
  and_list(map2_chr(no_flag_runs$from, no_flag_runs$to, range_text))
# Training defaulters with incomes in those ranges (within $500 of a grid
# income), none of whom KNN flags.
in_no_flag <- map_lgl(train$income, \(v) {
  any(no_flag$none & abs(no_flag$income - v) <= 500)
})
n_def_unflagged <- sum(train$default == "Yes" & in_no_flag)
where_text <- if (nrow(no_flag_runs) == 1) "that range" else "those ranges"
defaulters_text <- if (n_def_unflagged == 0) {
  paste0(", although no training defaulter has an income in ", where_text)
} else if (n_def_unflagged == 1) {
  paste0(", including the one training defaulter with an income in ", where_text)
} else {
  paste0(", including the ", n_def_unflagged,
         " training defaulters with incomes in ", where_text)
}
```
:::

With $K = `r k_default`$ neighbors for KNN, the figure below shows the
training data with every classifier's fitted 0.5 boundary: non-defaulters as
grey circles and defaulters as black triangles, with balances beyond the
largest training balance shaded, and logistic regression pink dashed, LDA
green solid, QDA orange dotted, naive Bayes blue dot-dash, and KNN black
long-and-short dashes.

```{r comparison-boundaries-figure}
#| echo: false
#| fig-width: 7.5
#| fig-height: 4.5
all_boundaries_plot
```

At middle incomes, near the balances where most training defaulters sit, the
five boundaries lie close together. The logistic regression and LDA
boundaries are straight lines. The QDA and naive Bayes boundaries curve, and
each has a second branch beyond the training balances, an extrapolation of
their Gaussian tails. The KNN boundary is jagged, following the vote of the
nearest training customers rather than any fixed form. Beyond the training
balances its course is also an extrapolation, a vote among the same edge
customers. On this split, `r if (nrow(no_flag_runs) == 0) "at no income in the training range does its threshold lie beyond every training balance" else paste0("at incomes ", no_flag_text, " its threshold lies beyond every training balance, so it flags no customer there", defaulters_text)`.

### Boundary shapes

With two classes, the four model-based classifiers differ in the form of the
log posterior odds they can represent:

$$\begin{aligned}
\text{Logistic regression, LDA: } & \beta_0 + \sum_{j=1}^p \beta_j x_j, \\
\text{QDA: } & \beta_0 + \sum_{j=1}^p \beta_j x_j + \sum_{j \le l} \gamma_{jl} x_j x_l, \\
\text{Gaussian naive Bayes: } & \beta_0 + \sum_{j=1}^p \left(\beta_j x_j + \gamma_{jj} x_j^2\right),
\end{aligned}$$

where

- $\beta_0$ is an intercept and $\beta_j$ the coefficient of $x_j$, with
values that differ from method to method, and
- $\gamma_{jl}$ is the coefficient of the product $x_j x_l$, a square when
$l = j$.

A naive Bayes classifier with Bernoulli or categorical features adds terms
linear in those features' indicators. KNN has no formula for its log
posterior odds, and its boundary can take any shape. The table summarizes all
five:

::: {.callout-note collapse="true"}
### Code
```{r comparison-table}
#| output: false
comparison_table <- tibble(
  Method = c("Logistic regression", "LDA", "QDA", "Gaussian naive Bayes",
             "KNN"),
  Assumption = c("Linear log-odds",
                 "Gaussian, shared covariance",
                 "Gaussian, class covariances",
                 "Features independent within each class",
                 "Smooth class probabilities"),
  Estimation = c("Iterative", "Closed form", "Closed form", "Closed form",
                 "None (stores data)")
) |>
  knitr::kable(align = "lll")
```
:::

```{r comparison-table-display}
#| echo: false
comparison_table
```

Logistic regression and LDA share a linear form and differ in estimation:
when LDA's Gaussian, shared-covariance assumption fails, logistic regression,
which never models the features' distribution, is unaffected provided the
log-odds remain linear. A logistic regression given QDA's squares and
cross-products, as in the flexible logistic regression lecture, has QDA's
quadratic form, fit by the conditional likelihood instead. Naive Bayes is the
most restricted quadratic form and has fewer parameters than QDA. KNN is
nonparametric, with flexibility set by $1/K$: it can follow any boundary
given enough data, but it needs $n$ large relative to $p$, suffers most from
the curse of dimensionality, and has no coefficients to interpret.

Increasing $n$ favors the more flexible methods, QDA and KNN, whose lower
bias becomes affordable as their variance shrinks. Increasing $p$ favors the
more restricted ones: naive Bayes, whose parameter count grows linearly in
$p$, and LDA, whose count grows like $p^2/2$ rather than like QDA's $p^2$ for
two classes.

### Simulated scenarios

A small simulation puts these predictions to a test, in four scenarios of two
classes with equal priors, each a data-generating process chosen so that a
different method's assumptions hold.

::: {.callout-note collapse="true"}
### Code
```{r scenario-study}
#| output: false
p_indep <- 10                     # independent: number of features
indep_name <- paste0("Independent, p = ", p_indep)
scenario_names <- c("Linear", "Quadratic", indep_name, "Non-linear")
n_reps <- 100       # training sets per scenario
n_train_k <- 50     # training observations per class
n_test_k <- 2000    # test observations per class
k_sim <- odd_above_sqrt(2 * n_train_k)
knn_sim_name <- paste0("KNN, K = ", k_sim)
rho_linear <- 0.5                 # linear: correlation of the shared covariance
mu_linear <- c(1.5, 0.5)          # linear: class 2 mean (class 1 at the origin)
shift_indep <- 0.5                # independent: class 2 mean, first half
var_indep <- 2                    # independent: class 2 variance, second half
flip_rate <- 0.1                  # non-linear: share of labels flipped
S_linear <- matrix(c(1, rho_linear, rho_linear, 1), nrow = 2)
mu_indep <- rep(c(shift_indep, 0), each = p_indep / 2)
var_indep_2 <- rep(c(1, var_indep), each = p_indep / 2)

rmvnorm_p <- function(n, mu, Sigma) {
  sweep(matrix(rnorm(n * length(mu)), n) %*% chol(Sigma), 2, mu, "+")
}

# One data set of 2 * n_k observations, n_k per class (in expectation for the
# non-linear scenario, whose class sizes are random).
draw_scenario <- function(scenario, n_k) {
  labels <- c("Class 1", "Class 2")
  if (scenario == "Non-linear") {
    X <- matrix(runif(4 * n_k, -2, 2), ncol = 2)
    y <- as.numeric(X[, 2] > sin(2 * X[, 1]))
    flip <- runif(2 * n_k) < flip_rate
    y[flip] <- 1 - y[flip]
    y <- factor(labels[y + 1], levels = labels)
  } else {
    if (scenario == "Linear") {
      X <- rbind(rmvnorm_p(n_k, c(0, 0), S_linear),
                 rmvnorm_p(n_k, mu_linear, S_linear))
    }
    if (scenario == "Quadratic") {
      X <- rbind(rmvnorm_p(n_k, mu_a, Sigma_a), rmvnorm_p(n_k, mu_b, Sigma_b))
    }
    if (scenario == indep_name) {
      X <- rbind(rmvnorm_p(n_k, rep(0, p_indep), diag(p_indep)),
                 rmvnorm_p(n_k, mu_indep, diag(var_indep_2)))
    }
    y <- factor(rep(labels, each = n_k), levels = labels)
  }
  colnames(X) <- paste0("x", seq_len(ncol(X)))
  bind_cols(tibble(y = y), as_tibble(X))
}

# Bayes classifier's error rate on a test set, from the true model.
bayes_error <- function(scenario, test) {
  X <- as.matrix(select(test, -y))
  pred <- switch(scenario,
    "Non-linear" = X[, 2] > sin(2 * X[, 1]),
    Linear = log_odds_gauss(X[, 1], X[, 2], c(0, 0), S_linear, mu_linear,
                            S_linear) > 0,
    Quadratic = log_odds_gauss(X[, 1], X[, 2], mu_a, Sigma_a, mu_b, Sigma_b) > 0,
    rowSums(dnorm(X, mu_indep[col(X)], sqrt(var_indep_2)[col(X)], log = TRUE) -
              dnorm(X, log = TRUE)) > 0)
  mean(pred != (test$y == "Class 2"))
}

# Test error rate of each method fit to one training set; KNN uses features
# standardized by the training means and SDs.
fit_methods <- function(train, test) {
  truth <- test$y
  X_train <- as.matrix(select(train, -y))
  Z_train <- scale(X_train)
  Z_test  <- scale(as.matrix(select(test, -y)), attr(Z_train, "scaled:center"),
                   attr(Z_train, "scaled:scale"))
  knn_error <- function(K) mean(class::knn(Z_train, Z_test, train$y, k = K) != truth)
  p_glm <- predict(glm(y ~ ., data = train, family = "binomial"), test,
                   type = "response")
  c(`Logistic regression` = mean((p_glm > 0.5) != (truth == "Class 2")),
    LDA = mean(predict(MASS::lda(y ~ ., data = train), test)$class != truth),
    QDA = mean(predict(MASS::qda(y ~ ., data = train), test)$class != truth),
    `Naive Bayes` = mean(predict(e1071::naiveBayes(y ~ ., data = train), test) != truth),
    `KNN, K = 1` = knn_error(1)) |>
    c(setNames(knn_error(odd_above_sqrt(nrow(train))), knn_sim_name))
}

scenario_results <- map(scenario_names, \(s) {
  test <- draw_scenario(s, n_test_k)
  errors <- map(seq_len(n_reps),
                \(r) fit_methods(draw_scenario(s, n_train_k), test))
  list(errors = bind_rows(errors) |> mutate(scenario = s, rep = row_number()),
       bayes = bayes_error(s, test))
})

method_levels <- c("Logistic regression", "LDA", "QDA", "Naive Bayes",
                   "KNN, K = 1", knn_sim_name)

scenario_errors <- map(scenario_results, "errors") |>
  bind_rows() |>
  pivot_longer(all_of(method_levels), names_to = "method",
               values_to = "error") |>
  mutate(method = factor(method, levels = method_levels),
         scenario = factor(scenario, levels = scenario_names))

scenario_bayes <- tibble(scenario = factor(scenario_names, levels = scenario_names),
                         bayes = map_dbl(scenario_results, "bayes"))
scenario_means <- scenario_errors |>
  group_by(scenario, method) |>
  summarize(mean_error = mean(error), .groups = "drop")
mean_error_of <- function(s, m) {
  with(scenario_means, mean_error[scenario == s & method == m])
}
best_method <- function(s) {
  with(filter(scenario_means, scenario == s), as.character(method[which.min(mean_error)]))
}
pct <- function(v) sprintf("%.1f%%", 100 * v)

# Prose labels for the winning method of each scenario.
prose_label <- setNames(c("logistic regression", "LDA", "QDA", "naive Bayes",
                          "KNN with $K = 1$", paste0("KNN with $K = ", k_sim, "$")),
                        c("Logistic regression", "LDA", "QDA", "Naive Bayes",
                          "KNN, K = 1", knn_sim_name))
winner <- function(s) unname(prose_label[best_method(s)])

pp <- function(v) sprintf("%.3f", 100 * v)   # percentage points

# One training set per scenario to show what the data look like; the
# independent scenario is shown in features 1 (mean shift) and 6 (variance).
scenario_examples <- map(scenario_names, \(s) {
  draw_scenario(s, n_train_k) |>
    transmute(y, first = x1,
              second = if (s == indep_name) .data[[paste0("x", p_indep / 2 + 1)]] else x2,
              scenario = s)
}) |>
  bind_rows() |>
  mutate(scenario = factor(scenario, levels = scenario_names))

scenario_data_plot <- ggplot(scenario_examples,
                             aes(x = first, y = second, color = y, shape = y)) +
  geom_point(size = 1.6, alpha = 0.8) +
  facet_wrap(~ scenario, scales = "free") +
  scale_color_manual(values = class_colors) +
  labs(x = "First plotted feature", y = "Second plotted feature",
       color = "Class", shape = "Class",
       title = paste0("One training set per scenario (Independent: features 1 and ",
                      p_indep / 2 + 1, ")"))

scenario_error_plot <- ggplot(scenario_errors, aes(x = method, y = error)) +
  geom_boxplot(outlier.size = 0.6, fill = "grey90") +
  geom_hline(data = scenario_bayes, aes(yintercept = bayes),
             linetype = "dashed") +
  facet_wrap(~ scenario, scales = "free_y") +
  labs(x = NULL, y = "Test error rate",
       title = paste("Test error rates over", n_reps,
                     "training sets; Bayes error rate dashed")) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```
:::

- **Linear**: two Gaussian features, a shared covariance matrix with
correlation $`r rho_linear`$, and class means $(0, 0)^\top$ and
$(`r mu_linear[1]`, `r mu_linear[2]`)^\top$, so the Bayes boundary is a line.
- **Quadratic**: the two classes of the variability figure, with
class-specific covariance matrices.
- **Independent, $p = `r p_indep`$**: `r p_indep` Gaussian features,
independent within each class; class 2's mean is `r shift_indep` in the first
`r p_indep / 2` features and its variance `r var_indep` in the last
`r p_indep / 2`, against means 0 and variances 1 in class 1.
- **Non-linear**: two features uniform on a square, class $2$ above the curve
$x_2 = \sin(2x_1)$ and class $1$ below it, and then
`r 100 * flip_rate`% of the class labels flipped at random, so neither class
is Gaussian and the Bayes boundary is the sine curve.

In each scenario, `r n_reps` training sets of `r 2 * n_train_k` observations
were drawn, `r n_train_k` per class (in expectation in the non-linear
scenario). Each method was fit to every training set and its error rate
computed on one large test set of `r format(2 * n_test_k, big.mark = ",")`
observations drawn from the same scenario. KNN was fit on standardized
features with $K = 1$ and with $K = `r k_sim`$, the KNN rule above for the
$n = `r 2 * n_train_k`$ training observations.

The first tab below shows one training set from each scenario, classes
distinguished by color and shape, with features 1 and `r p_indep / 2 + 1` for the independent
scenario. In the second, each box summarizes the test error rates of one
method over the `r n_reps` training sets, and the dashed line is the Bayes
error rate, the floor no classifier can beat on average.

::: {.panel-tabset}
#### Scenarios
```{r scenario-data-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 6.5
scenario_data_plot
```

#### Test error rates
```{r scenario-error-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 6.5
scenario_error_plot
```
:::

- **Linear.** Logistic regression and LDA
`r if (pct(mean_error_of("Linear", "Logistic regression")) == pct(mean_error_of("Linear", "LDA"))) paste("both average", pct(mean_error_of("Linear", "LDA"))) else paste("average", pct(mean_error_of("Linear", "Logistic regression")), "and", pct(mean_error_of("Linear", "LDA")))`,
against a Bayes error rate of
`r pct(scenario_bayes$bayes[1])`. QDA, whose extra covariance parameters are
unnecessary here, averages `r pct(mean_error_of("Linear", "QDA"))`: with only
two features its variance penalty is small. Naive Bayes, which ignores the
correlation, averages `r pct(mean_error_of("Linear", "Naive Bayes"))`.
- **Quadratic.** QDA's average of `r pct(mean_error_of("Quadratic", "QDA"))`
beats logistic regression's
`r pct(mean_error_of("Quadratic", "Logistic regression"))` and LDA's
`r pct(mean_error_of("Quadratic", "LDA"))`: the linear methods' bias outweighs
QDA's variance. Naive Bayes, curved but missing the cross-product, falls
between them, at `r pct(mean_error_of("Quadratic", "Naive Bayes"))`.
- **Independent, $p = `r p_indep`$.** Naive Bayes, whose assumption holds exactly,
averages `r pct(mean_error_of(indep_name, "Naive Bayes"))` with
`r 2 * 2 * p_indep` parameters. QDA's model also contains the truth, but it
estimates `r count_qda(2, p_indep)` parameters from `r 2 * n_train_k`
observations and averages `r pct(mean_error_of(indep_name, "QDA"))`. KNN, with
`r p_indep` dimensions to search for neighbors, averages
`r pct(mean_error_of(indep_name, knn_sim_name))` with
$K = `r k_sim`$.
- **Non-linear.** No model-based method can represent a sine curve, and all
four average
`r nl <- pct(range(sapply(method_levels[1:4], \(m) mean_error_of("Non-linear", m)))); if (nl[1] == nl[2]) paste("about", nl[1]) else paste("between", nl[1], "and", nl[2])`.
KNN with $K = `r k_sim`$ averages
`r pct(mean_error_of("Non-linear", knn_sim_name))`. With $K = 1$ its variance
erases that advantage: it averages
`r pct(mean_error_of("Non-linear", "KNN, K = 1"))`, about the same as the
model-based methods.

In the quadratic, independent, and non-linear scenarios the best method is the
one whose assumptions are the most restrictive ones that still hold:
`r winner("Quadratic")` in the quadratic, `r winner(indep_name)` in
the independent, and `r winner("Non-linear")` in the non-linear scenario. In
the linear scenario both LDA and logistic regression are correct models, and
their averages differ by only
`r pp(abs(mean_error_of("Linear", "Logistic regression") - mean_error_of("Linear", "LDA")))`
percentage points, with
`r if (best_method("Linear") == "LDA") "LDA" else "logistic regression"` ahead:
with two features, LDA's extra assumption buys essentially nothing.

### Test performance

Returning to the credit card customers, every classifier of the five-boundary
figure, with KNN also at $K = 1$, is scored on both halves.

::: {.callout-note collapse="true"}
### Code
```{r default-test}
#| output: false
knn_default_name <- paste0("KNN, K = ", k_default)
model_based_names <- c("Logistic regression", "LDA", "QDA", "Naive Bayes")
all_scores <- function(d) {
  map(model_based_names, \(m) posterior_of(m, d)) |>
    set_names(model_based_names) |>
    as_tibble() |>
    mutate(`KNN, K = 1` = knn_yes(d, 1), "{knn_default_name}" := knn_yes(d, k_default))
}
scores_train <- all_scores(train)
scores_test  <- all_scores(test)

# AUC as the proportion of (defaulter, non-defaulter) pairs ordered correctly.
auc_pairs <- function(s, pos) {
  mean(outer(s[pos], s[!pos], ">")) + 0.5 * mean(outer(s[pos], s[!pos], "=="))
}
yes_train <- train$default == "Yes"
yes_test  <- test$default == "Yes"

rates <- function(p_train, p_test) {
  tibble(train_error = mean((p_train > 0.5) != yes_train),
         test_error  = mean((p_test > 0.5) != yes_test),
         sensitivity = mean(p_test[yes_test] > 0.5),
         specificity = mean(p_test[!yes_test] <= 0.5),
         auc         = auc_pairs(p_test, yes_test),
         flagged     = sum(p_test > 0.5))
}
performance <- map(names(scores_test), \(m) rates(scores_train[[m]], scores_test[[m]])) |>
  set_names(names(scores_test)) |>
  list_rbind(names_to = "model")
perf_of <- function(m, col) performance[[col]][performance$model == m]
model_based <- performance |> slice(1:4)
# Optimism sentence, with a separate wording for each count of model-based
# classifiers whose training error rate is above their test error rate.
n_train_above <- sum(model_based$train_error > model_based$test_error)
chance <- "Training error rates are optimistic on average, but on a single split chance can outweigh that: "
optimism_text <- if (n_train_above == 0) {
  "No model-based classifier has a training error rate above its test error rate, as optimism predicts."
} else if (n_train_above < 4) {
  paste0(chance, c("one of the four model-based classifiers has",
                   "two of the four model-based classifiers have",
                   "three of the four model-based classifiers have")[n_train_above],
         " a training error rate above the test error rate.")
} else {
  paste0(chance, "all four model-based classifiers have a training error rate above ",
         "their test error rate",
         if (n_defaults[["train"]] > n_defaults[["test"]]) sprintf(
           ", perhaps because the training half drew %d defaulters (%s) and the test half %d (%s).",
           n_defaults[["train"]], pct2(mean(yes_train)), n_defaults[["test"]],
           pct2(mean(yes_test))) else ".")
}

performance_table <- performance |>
  select(-flagged) |>
  mutate(across(train_error:specificity, pct2), auc = sprintf("%.4f", auc)) |>
  knitr::kable(col.names = c("Model", "Training error", "Test error",
                             "Test sensitivity", "Test specificity", "Test AUC"),
               align = "lccccc")

# Naive Bayes against QDA, from the rates as displayed in the table.
nb_error_diff <- round(perf_of("Naive Bayes", "test_error") - perf_of("QDA", "test_error"), 4)
nb_auc_diff   <- round(perf_of("Naive Bayes", "auc"), 4) - round(perf_of("QDA", "auc"), 4)
higher_lower <- function(d) if (d > 0) "higher" else "lower"
nb_error_text <- if (nb_error_diff == 0) "the same as QDA's" else
  paste(sprintf("%.2f percentage points", 100 * abs(nb_error_diff)),
        higher_lower(nb_error_diff), "than QDA's")
nb_auc_text <- if (nb_auc_diff == 0) "the same as QDA's" else
  paste(sprintf("%.4f", abs(nb_auc_diff)), higher_lower(nb_auc_diff))
nb_cost <- if (nb_error_diff <= 0 && nb_auc_diff >= 0) "does not hurt naive Bayes" else
  "costs naive Bayes little"

# Model-based classifiers with the highest sensitivity and the most customers
# flagged, named in prose (ties listed together).
model_label <- c(`Logistic regression` = "logistic regression", LDA = "LDA",
                 QDA = "QDA", `Naive Bayes` = "naive Bayes")
top_models <- function(col) {
  v <- round(model_based[[col]], 6)
  unname(model_label[model_based$model[v == max(v)]])
}
sens_top <- top_models("sensitivity")
flag_top <- top_models("flagged")
# "is X's" for one model, "is shared by X and Y" for a tie.
owner <- function(x) if (length(x) == 1) paste0("is ", x, "'s") else paste("is shared by", and_list(x))
top_text <- paste0("The highest sensitivity ", owner(sens_top),
  if (identical(sens_top, flag_top) && length(flag_top) == 1) ", which flags" else
    paste0("; ", and_list(flag_top), if (length(flag_top) == 1) " flags" else " flag"),
  " the most test customers, ", max(model_based$flagged), ".")

# KNN's place among the classifiers, stated only when it holds.
knn1_worst <- perf_of("KNN, K = 1", "test_error") == max(performance$test_error)
knn_auc_lowest <- perf_of(knn_default_name, "auc") < min(model_based$auc)

# The opening question, answered from the table's rounded test error rates.
all_label <- c(model_label, setNames(c("KNN with $K = 1$",
                                       paste0("KNN with $K = ", k_default, "$")),
                                     c("KNN, K = 1", knn_default_name)))
err4 <- setNames(round(performance$test_error, 4), performance$model)
lowest_text <- owner(unname(all_label[names(err4)[err4 == min(err4)]]))
new_vs_old <- min(err4[c("LDA", "Logistic regression")]) - min(err4[c("QDA", "Naive Bayes")])
new_vs_old_text <- if (new_vs_old == 0) "the same as" else
  paste(sprintf("%.2f percentage points", 100 * abs(new_vs_old)),
        if (new_vs_old > 0) "lower than" else "higher than")
knn_best <- min(err4[c("KNN, K = 1", knn_default_name)])
knn_vs_text <- if (knn_best > max(err4[model_based$model])) {
  "both KNN classifiers have higher test error rates than every model-based classifier"
} else if (knn_best < min(err4[model_based$model])) {
  "the better KNN classifier has a lower test error rate than every model-based classifier"
} else {
  "the better KNN classifier's test error rate falls within the model-based classifiers' range"
}
```
:::

KNN uses $K = 1$ and $K = `r k_default`$. The table
gives each classifier's training error rate and its test error rate,
sensitivity, and specificity at threshold $0.5$ (for KNN, a majority vote),
and test AUC.

```{r performance-table}
#| echo: false
performance_table
```

`r optimism_text` KNN with $K = 1$ shows the optimism plainly: its training
error rate is `r pct2(perf_of("KNN, K = 1", "train_error"))`, since each
training customer is its own nearest neighbor, against a test error rate of
`r pct2(perf_of("KNN, K = 1", "test_error"))``r if (knn1_worst) ", the worst in the table" else ""`.

The four model-based classifiers have test error rates
`r te <- pct2(range(model_based$test_error)); if (te[1] == te[2]) paste("of", te[1]) else paste("between", te[1], "and", te[2])`,
against the
`r pct2(mean(yes_test))` of predicting `No` for every test customer. They
differ more in sensitivity, from `r pct2(min(model_based$sensitivity))` to
`r pct2(max(model_based$sensitivity))`, a difference of
`r round(n_defaults["test"] * diff(range(model_based$sensitivity)))` of the
`r n_defaults["test"]` test defaulters. `r top_text` Their test AUCs range
only from `r sprintf("%.4f", min(model_based$auc))` to
`r sprintf("%.4f", max(model_based$auc))`. With the within-class correlation
this small, dropping it `r nb_cost`: its test error rate is
`r nb_error_text`, and its AUC is `r nb_auc_text`. KNN with
$K = `r k_default`$ has `r if (knn_auc_lowest) "a lower test AUC than any of them" else "a test AUC"`,
`r sprintf("%.4f", perf_of(knn_default_name, "auc"))`, and a test sensitivity
of `r pct2(perf_of(knn_default_name, "sensitivity"))`,
`r if (perf_of(knn_default_name, "sensitivity") < min(model_based$sensitivity)) "below every model-based classifier's" else "within or above the model-based classifiers' range"``r if (nrow(no_flag_runs) > 0) ", consistent with the incomes at which its boundary flags no one" else ""`.

To answer the opening question on this split: the lowest test error rate
`r lowest_text`; the better of QDA and naive Bayes has a test error rate
`r new_vs_old_text` the better of LDA and logistic regression; and
`r knn_vs_text`.

## Student strata

Apart from the mixed naive Bayes fit, student status has not entered the
comparison. The table below
summarizes the two student strata of the training half, and the figure plots
income against balance in each stratum, points colored and shaped by default
status, with each panel's income axis covering its own stratum's incomes.

::: {.callout-note collapse="true"}
### Code
```{r strata-summary}
#| output: false
strata_summary <- train |>
  group_by(student) |>
  summarize(n = n(), defaulters = sum(default == "Yes"),
            rate = mean(default == "Yes"), median_income = median(income),
            .groups = "drop")
strata_summary_of <- function(s, col) strata_summary[[col]][strata_summary$student == s]
strata_summary_table <- strata_summary |>
  transmute(student, n = format(n, big.mark = ","), defaulters,
            rate = pct2(rate), median_income = dollar_text(median_income)) |>
  knitr::kable(col.names = c("Student", "Customers", "Defaulters",
                             "Default rate", "Median income"), align = "ccccc")
strata_data_plot <- ggplot() +
  scatter_points +
  facet_wrap(~ student, labeller = label_both, scales = "free_y") +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2))) +
  labs(title = "Training set, income against balance by student status")
```
:::

```{r strata-summary-table}
#| echo: false
strata_summary_table
```

```{r strata-data-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 4
strata_data_plot
```

Students default `r if (strata_summary_of("Yes", "rate") > strata_summary_of("No", "rate")) "more" else "less"` often,
`r pct2(strata_summary_of("Yes", "rate"))` against
`r pct2(strata_summary_of("No", "rate"))`, and their incomes occupy a lower
band; in both strata the defaulters sit at the larger balances. These
numbers differ from the previous lecture's strata table because that lecture
used all customers and this one the training half.

In this additional analysis, QDA with `balance` and `income` is fit
separately to each stratum, beside the logistic regression whose intercept
and slopes all interact with student status. Each stratum's QDA uses that stratum's class
proportions as priors, so every customer's score is its posterior probability
given balance, income, and student status, on one common scale across strata,
which is what lets a single error rate and AUC be computed over all customers.

::: {.callout-note collapse="true"}
### Code
```{r strata}
#| output: false
strata <- c(No = "No", Yes = "Yes")
train_by <- map(strata, \(s) filter(train, student == s))
strata_qda <- map(train_by, \(d) MASS::qda(default ~ balance + income, data = d))
strata_glm <- map(train_by, \(d) glm(default ~ balance + income, data = d,
                                     family = "binomial"))
interacted_glm <- glm(default ~ (balance + income) * student, data = train,
                      family = "binomial")

# Score every customer with its own stratum's model.
by_stratum <- function(d, score) {
  p <- numeric(nrow(d))
  for (s in strata) p[d$student == s] <- score(s, d[d$student == s, ])
  p
}
qda_strat <- function(d) by_stratum(d, \(s, ds) predict(strata_qda[[s]], ds)$posterior[, "Yes"])
max_diff_int <- max(abs(fitted(interacted_glm) - by_stratum(train, \(s, ds) {
  predict(strata_glm[[s]], ds, type = "response")
})))
glm_int <- function(d) predict(interacted_glm, newdata = d, type = "response")

# Rates rounded as displayed, so changes computed from them match the table.
strata_rates <- list(
  "QDA, single fit" = rates(scores_train$QDA, scores_test$QDA),
  "QDA, by student stratum" = rates(qda_strat(train), qda_strat(test)),
  "Logistic regression, additive" = rates(scores_train$`Logistic regression`,
                                          scores_test$`Logistic regression`),
  "Logistic regression, interacted with student" =
    rates(glm_int(train), glm_int(test))
) |>
  list_rbind(names_to = "model") |>
  mutate(across(train_error:sensitivity, \(v) round(v, 4)), auc = round(auc, 4))
strata_table <- strata_rates |>
  select(-specificity, -flagged) |>
  mutate(across(train_error:sensitivity, pct2), auc = sprintf("%.4f", auc)) |>
  knitr::kable(col.names = c("Model", "Training error", "Test error",
                             "Test sensitivity", "Test AUC"), align = "lcccc")
# Largest change in test error rate and test AUC from stratifying either
# model: rows 2 and 4 are the stratified versions of rows 1 and 3.
max_error_change <- max(abs(diff(strata_rates$test_error)[c(1, 3)]))
max_auc_change   <- max(abs(diff(strata_rates$auc)[c(1, 3)]))
# Mean and covariance parameters of a two-class QDA, priors not counted, as in
# the parameter-count table.
n_par_qda <- count_qda(2, 2)
```
:::

The table gives each model's training and test error rates and test
sensitivity at threshold $0.5$, and its test AUC.

```{r strata-table}
#| echo: false
strata_table
```

Stratifying changes no model's test error rate by more than
`r sprintf("%.2f", 100 * max_error_change)` percentage points or its test AUC
by more than `r sprintf("%.4f", max_auc_change)`. (The interacted logistic
regression reproduces separate per-stratum logistic regressions, to within
$`r sci(max_diff_int)`$ on the training set.) Each stratified model estimates
twice as many parameters, `r 2 * n_par_qda` means and covariance entries against
`r n_par_qda` for QDA (priors not counted, as in the parameter-count table)
and `r length(coef(interacted_glm))` coefficients against
`r length(coef(default_glm))` for logistic regression, which tends to make its
training error rate more optimistic; the test rates are the fair
comparison.

## Conclusion

QDA is the generative classifier with Gaussian class-conditional densities and
a covariance matrix $\Sigma_k$ for each class. The quadratic term
$-\frac{1}{2}x^\top\Sigma_k^{-1}x$ and the term
$-\frac{1}{2}\log\left|\Sigma_k\right|$ no longer cancel between classes, so
the discriminant functions and the log posterior odds are quadratic in $x$,
the decision boundary is a quadratic curve or surface, and with one feature the class
with the larger variance is assigned in both tails. Each $\Sigma_k$ is
estimated by its class's sample covariance matrix, which requires $n_k > p$.

The choice between LDA and QDA is a bias-variance tradeoff. With $K$ classes,
QDA estimates $K p(p+1)/2$ covariance parameters against LDA's $p(p+1)/2$, so it removes
LDA's bias when the class covariances differ at the cost of variance that
grows with $p$ and shrinks with $n$.

Naive Bayes assumes the features are independent within each class, so
$f_k(x) = \prod_j f_{kj}(x_j)$, each one-feature density estimated on its own
and of whatever type suits the feature. The log posterior odds are a sum of
one-feature terms, with no interactions. Gaussian naive Bayes is LDA or QDA
with a diagonal covariance matrix. With class-specific variances it estimates
$2Kp$ parameters, sidestepping the curse of dimensionality that a joint
density faces. Its bias, which comes from ignoring within-class dependence,
need not change its classifications, since only the side of the threshold
matters.

Logistic regression and LDA produce linear boundaries, QDA and Gaussian naive
Bayes quadratic ones, and KNN boundaries of any shape. Which classifier has the
lowest test error rate depends on how closely its assumptions match the
data-generating process and on the sizes of $n$ and $p$; no method is best
in every setting. Comparing classifiers requires test data the models were
not fit to, since training error rates reward flexibility.

The next lecture steps back from classification to a framework that contains
linear and logistic regression as special cases. Generalized linear models
pair a distribution for the response with a link function for its mean, and
Poisson regression, the generalized linear model for counts, is the first new
member of the family.
