# Generalized Linear Models

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
set.seed(20261001)

working_colors  <- c("No" = "#009E73", "Yes" = "#CC79A7")
working_labels  <- c("No" = "Non-working day", "Yes" = "Working day")
model_colors    <- c("Linear regression" = "#D55E00",
                     "Poisson regression" = "#0072B2")
model_linetypes <- c("Linear regression" = "dashed",
                     "Poisson regression" = "solid")

# Renders a number in LaTeX scientific notation for use inside $...$.
sci <- function(x, digits = 2) {
  e <- floor(log10(abs(x)))
  m <- signif(x / 10^e, digits)
  sprintf("%s \\times 10^{%d}", format(m), e)
}

# Rounds and adds thousands separators for inline numbers.
fmt <- function(x, digits = 0) {
  format(round(x, digits), big.mark = ",", nsmall = digits, scientific = FALSE)
}
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **write** a generalized linear model as its three components — a
distribution for $Y_i$ given $x_i$ with mean $\mu_i$, a linear predictor
$\eta_i = x_i\beta$, and a link function $g$ with $g(\mu_i) = \eta_i$ — and
**identify** linear, logistic, and Poisson regression as special cases by their
distribution, link, and variance,
- **explain**, using fitted values and residuals, why linear regression is a
poor model for a count response,
- **write** the Poisson regression log-likelihood, **derive** its score
equations, and **explain** why its maximizer has no closed form,
- **interpret** $e^{\beta_0}$ as a mean count and $e^{\beta_j}$ as a rate
ratio, the factor multiplying the mean count per one-unit increase in feature
$j$ with the other features held fixed, and **construct** Wald tests and
confidence intervals for $\beta_j$ and $e^{\beta_j}$,
- **conduct** a drop-in-deviance test of a group of coefficients in a Poisson
regression, and
- **assess** the Poisson assumption $Var[Y_i \mid x_i] = \mu_i$ with Pearson
residuals, and **explain** what overdispersion does to Wald standard errors
and tests.
:::

## Count response

The regression lectures modeled a quantitative response whose mean could be
any real number, and the logistic regression lectures a binary response whose
mean is a probability. A **count** — the number of events in a fixed period
or region — is neither: it is a nonnegative integer, its mean must be
positive, and its spread typically grows with its mean.

### Bike rentals

**Question.** How does the number of bikes rented in an hour depend on the
time of day, whether the day is a working day, the month, the temperature,
and the weather?

**Data.** `ISLR2::Bikeshare` records `r fmt(nrow(Bikeshare))` hours of the
Capital Bikeshare system in Washington, D.C., spread over the
`r max(Bikeshare$day)` days of one year
(`r 24 * max(Bikeshare$day) - nrow(Bikeshare)` hours of the year have no
record). The response is `bikers`, the number of bikes rented in the hour.
The explanatory variables used here are

- `hr`, the hour of the day, a factor with levels `0` through `23`,
- `workingday`, $1$ for a day that is neither a weekend nor a holiday and $0$
otherwise,
- `mnth`, the month, a factor,
- `temp`, the temperature rescaled to $[0, 1]$ from the range
$-8$ to $39$ degrees Celsius, and
- `weathersit`, the weather, a factor with levels
`r paste0("\x60", levels(Bikeshare$weathersit), "\x60", collapse = ", ")`.

`bikers` is the sum of two other columns, `casual` and `registered`. Those are
pieces of the response rather than explanatory variables, and no model below
uses them.

### Hourly counts

::: {.callout-note collapse="true"}
### Code
```{r bike-eda}
#| output: false
bike <- Bikeshare |>
  mutate(working  = factor(workingday, levels = c(0, 1),
                           labels = c("No", "Yes")),
         weather  = fct_collapse(weathersit,
                                 `rain/snow` = c("light rain/snow",
                                                 "heavy rain/snow")),
         temp_c   = 47 * temp - 8,
         hour     = as.numeric(as.character(hr)),
         hour_jit = hour + runif(n(), -0.3, 0.3))

hour_groups <- bike |>
  group_by(working, hour) |>
  summarize(n = n(), mean = mean(bikers), variance = var(bikers),
            .groups = "drop")

mv_slope <- unname(coef(lm(log(variance) ~ log(mean), data = hour_groups))[2])

count_hist_plot <- ggplot(bike, aes(x = bikers)) +
  geom_histogram(binwidth = 10, boundary = 0, fill = "grey70",
                 color = "grey40") +
  labs(x = "Bikes rented in the hour", y = "Hours",
       title = "ISLR2::Bikeshare, hourly counts")

count_hour_plot <- ggplot(bike, aes(x = hour_jit, y = bikers)) +
  geom_point(alpha = 0.08, size = 0.5) +
  geom_line(data = hour_groups, aes(x = hour, y = mean), linewidth = 0.9) +
  geom_point(data = hour_groups, aes(x = hour, y = mean), size = 1.8) +
  facet_wrap(~ working, labeller = labeller(working = working_labels)) +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  labs(x = "Hour of the day", y = "Bikes rented in the hour",
       title = "Hourly counts (jittered) and hourly means")

mean_variance_plot <- ggplot(hour_groups,
                             aes(x = mean, y = variance, color = working,
                                 shape = working)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
              color = "grey40") +
  geom_point(size = 2.3) +
  annotate("text", x = 40, y = 12, label = "variance = mean", hjust = 0,
           color = "grey30") +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = working_colors, labels = working_labels) +
  scale_shape_manual(values = c(16, 17), labels = working_labels) +
  labs(x = "Mean count in the group (log scale)",
       y = "Variance of the count in the group (log scale)",
       color = NULL, shape = NULL,
       title = "One point per hour-by-working-day group")
```
:::

::: {.panel-tabset}
#### Distribution
```{r count-hist-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
count_hist_plot
```

#### By hour
```{r count-hour-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 3.6
count_hour_plot
```

#### Mean and variance
```{r mean-variance-figure}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.5
mean_variance_plot
```
:::

The counts run from `r min(bike$bikers)` to `r max(bike$bikers)` with median
`r median(bike$bikers)` and mean `r fmt(mean(bike$bikers), 1)`: every value is
a positive integer, and the distribution is strongly right-skewed.

On working days the hourly means peak twice, at hour
`r with(filter(hour_groups, working == "Yes", hour < 12), hour[which.max(mean)])`
and hour
`r with(filter(hour_groups, working == "Yes", hour >= 12), hour[which.max(mean)])`,
the commute. On non-working days they rise to a single midday plateau. The two
hourly profiles differ in shape, not just in level. Across hours, the points
spread out more where the mean is higher.

The third tab makes that precise. Each point is the mean and variance of the
counts in one of the `r nrow(hour_groups)` combinations of hour and
working-day status. The variance climbs steadily with the mean; on these
log-log axes the points lie near a line of slope
`r sprintf("%.2f", mv_slope)`. All `r sum(hour_groups$variance > hour_groups$mean)`
points lie above the dashed line on which the variance equals the mean.
Month, temperature, and weather vary within each group, so part of each
group's variance is variation those explanatory variables could explain; a
check that accounts for them comes after a model is fit.

### Linear regression fails

A linear regression fit by least squares uses these features:

- `r nlevels(bike$hr) - 1` dummy features for `hr`, with hour `0` the reference,
- the indicator $\mathrm{I}(\texttt{workingday} = 1)$,
- `r nlevels(bike$hr) - 1` products of each hour dummy with the working-day
indicator, because the two hourly profiles differ in shape,
- `r nlevels(bike$mnth) - 1` dummy features for `mnth`, with January the
reference,
- the temperature in degrees Celsius, the feature $h(\texttt{temp}) = 47\,\texttt{temp} - 8$, and
- `r nlevels(bike$weather) - 1` dummy features for the weather, with `clear`
the reference, after merging `heavy rain/snow` into `light rain/snow` as
`rain/snow`: `heavy rain/snow` occurs in
`r sum(Bikeshare$weathersit == "heavy rain/snow")` hour, and its dummy's
coefficient would be estimated from that hour alone.

```{r bike-lm}
bike_formula <- bikers ~ hr * working + mnth + temp_c + weather
bike_lm <- lm(bike_formula, data = bike)
```

::: {.callout-note collapse="true"}
### Code
```{r bike-lm-residuals}
#| output: false
bike <- bike |>
  mutate(fit_lm = fitted(bike_lm), resid_lm = residuals(bike_lm))

n_negative <- sum(bike$fit_lm < 0)
sd_low  <- sd(bike$resid_lm[bike$fit_lm < 50])
sd_high <- sd(bike$resid_lm[bike$fit_lm > 300])
b_lm    <- coef(bike_lm)

lm_residual_plot <- ggplot(bike, aes(x = fit_lm, y = resid_lm)) +
  geom_point(alpha = 0.1, size = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#D55E00") +
  geom_hline(yintercept = 0, color = "grey40") +
  labs(x = "Fitted value from least squares",
       y = "Residual", title = "Linear regression residuals")
```
:::

```{r lm-residual-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.6
lm_residual_plot
```

Three failures appear.

1. **Negative fitted means.** Every point left of the dashed line has a
negative fitted value: `r fmt(n_negative)` of the `r fmt(nrow(bike))` hours,
`r sprintf("%.1f", 100 * n_negative / nrow(bike))`%, with the smallest at
`r fmt(min(bike$fit_lm), 1)` bikers. A mean count cannot be negative.
2. **Variance growing with the mean.** The residuals fan out to the right. Their
standard deviation is `r fmt(sd_low, 1)` among hours with fitted values below
$50$ and `r fmt(sd_high, 1)` among hours with fitted values above $300$, while
every $t$-test and interval from `lm()` assumes one common variance.
3. **Additive effects.** Every additive effect — month, temperature, and
weather — subtracts the same number of riders from every hour, and those
constant subtractions are what drive the night-time fitted values below zero:
`r fmt(sum(bike$fit_lm < 0 & bike$weather == "clear"))` of the
`r fmt(n_negative)` negative fitted values are in clear weather. The
`rain/snow` coefficient, `r fmt(b_lm["weatherrain/snow"], 1)`, removes as many
riders from an early-morning hour averaging a handful as from a rush hour
averaging hundreds. Rain plausibly removes a *proportion* of the riders who
would otherwise ride, which is an effect that multiplies the mean.

A count needs a model whose mean is positive for every value of the
features, whose effects multiply the mean, and whose variance grows with the
mean.

## Generalized linear models

The problems in logistic regression lecture wrote each model as one
distributional statement: $Y_i \mid x_i \stackrel{ind}{\sim} N(x_i\beta, \sigma^2)$
for linear regression, and
$Y_i \mid x_i \stackrel{ind}{\sim} \text{Bernoulli}\left[p(x_i)\right]$ with
$\log\left[\frac{p(x_i)}{1 - p(x_i)}\right] = x_i\beta$ for logistic
regression. Both statements have the same three parts, and generalized linear
models keep those parts while letting each one vary.

### Three components

A **generalized linear model** (GLM) assumes the statistical model

$$Y_i \mid x_i \stackrel{ind}{\sim} \text{a distribution with mean } \mu_i, \qquad g(\mu_i) = \eta_i, \qquad \eta_i = x_i\beta,$$

where

- $\mu_i = E[Y_i \mid x_i]$ is the mean of the response of observation $i$
(unrelated to the class means $\mu_k$ of the discriminant analysis lectures),
- $x_i = (1, x_{i1}, \ldots, x_{ip})$ is the $i$th row of the model matrix
$\mathbf{X}$ and $\beta = (\beta_0, \ldots, \beta_p)^\top$ is the vector of
coefficients,
- $\eta_i$ is the **linear predictor**, as in the logistic regression
lectures, and
- $g$ is the **link function**, a monotone function; the three links in this
lecture also carry the possible values of the mean onto the whole real line.

The three parts have names:

1. The **random component** is the distribution of $Y_i$ given $x_i$:
normal, Bernoulli, Poisson, or another family chosen to suit the response.
2. The **linear predictor** $\eta_i = x_i\beta$ is where the features enter.
Dummy variables, polynomials, step functions, and interactions all enter it
exactly as they entered a linear or logistic regression.
3. The **link function** connects the two. Inverting it,
$\mu_i = g^{-1}(\eta_i) = g^{-1}(x_i\beta)$.

The link is what keeps the mean in range. A linear predictor can take any real
value, and for the three links in this lecture $g^{-1}$ carries the real line
onto the values the mean is allowed to take — $(0, 1)$ for a probability, $(0, \infty)$ for a count — so every
$\beta$ gives a valid mean.

ISLR2 writes $\eta$ for the link function itself. These notes keep $\eta$ for
the linear predictor, as in the logistic regression lectures, and write $g$
for the link.

### Special cases

::: {.callout-note collapse="true"}
### Code
```{r glm-table}
#| output: false
glm_table <- tibble(
  Model = c("Linear regression", "Logistic regression", "Poisson regression"),
  Response = c("Quantitative", "Binary ($0/1$)", "Count"),
  Distribution = c("$N(\\mu_i, \\sigma^2)$", "$\\text{Bernoulli}(\\mu_i)$",
                   "$\\text{Poisson}(\\mu_i)$"),
  Link = c("identity, $\\mu_i$", "logit, $\\log\\left[\\mu_i/(1 - \\mu_i)\\right]$",
           "log, $\\log \\mu_i$"),
  Mean = c("$\\eta_i$", "$e^{\\eta_i}/\\left(1 + e^{\\eta_i}\\right)$",
           "$e^{\\eta_i}$"),
  Variance = c("$\\sigma^2$", "$\\mu_i(1 - \\mu_i)$", "$\\mu_i$"),
  family = c("`gaussian`", "`binomial`", "`poisson`")
) |>
  knitr::kable(col.names = c("Model", "Response", "Distribution of $Y_i$",
                             "Link $g(\\mu_i)$", "Mean $\\mu_i$",
                             "Variance", "`glm()` family"),
               align = "lllllll")
```
:::

```{r glm-table-display}
#| echo: false
#| tbl-colwidths: [17, 12, 15, 22, 16, 10, 8]
glm_table
```

In logistic regression the mean $\mu_i = P(Y_i = 1 \mid x_i)$ is the
probability the logistic regression lectures wrote $p(x_i)$.

Each distribution fixes how the variance depends on the mean,

$$Var[Y_i \mid x_i] = \phi\, V(\mu_i),$$

where

- $V$ is the **variance function**: $V(\mu) = 1$ for the normal,
$V(\mu) = \mu(1 - \mu)$ for the Bernoulli, and $V(\mu) = \mu$ for the Poisson,
and
- $\phi$ is the **dispersion parameter**: $\phi = \sigma^2$ for the normal,
and $\phi = 1$ for the Bernoulli and the Poisson, which have no variance
parameter separate from the mean.

Linear regression's constant variance is the special case $V(\mu) = 1$. The
two other models require the variance to change with the mean, as the bike
counts did.

A GLM transforms the **mean**, not the response. The regression lectures'
remedy for a skewed response was to model $\log Y_i$, which describes
$E[\log Y_i \mid x_i]$ — under normal errors the logarithm of the median of
$Y_i$ — and fails outright for a count of zero (`bikers` has no zeros, its
minimum is `r min(bike$bikers)`, so $\log(\texttt{bikers})$ is defined here,
but a count response in general can be zero). A log link models
$\log E[Y_i \mid x_i]$, the logarithm of the mean itself, and is defined for
any count because it never takes the logarithm of $y_i$.

`glm()` fits every model in the table, with the `family` argument choosing the
distribution and, by default, the link shown:

```{r glm-families}
#| eval: false
glm(y ~ x, family = gaussian)  # linear regression: same estimates as lm()
glm(y ~ x, family = binomial)  # logistic regression
glm(y ~ x, family = poisson)   # Poisson regression
```

::: {.callout-note collapse="true"}
### Beyond this course
The normal, Bernoulli, and Poisson distributions all belong to the
**exponential family**, and each table row uses its family's **canonical
link**, the link under which the score equations take the form
$\sum_i x_{ij}(y_i - \mu_i) = 0$ derived below. The same three-part template
covers other distributions, other links for the same distribution, and
responses with more than two classes:

- **Gamma regression**, for a positive, continuous, right-skewed response such
as an insurance claim amount, has $V(\mu) = \mu^2$ and usually a log link:
`glm(y ~ x, family = Gamma(link = "log"))`.
- **Probit regression** keeps the Bernoulli distribution and replaces the logit
link with the inverse of the standard normal cumulative distribution
function: `family = binomial(link = "probit")`.
- **Multinomial logistic regression**, for an unordered response with $K > 2$
classes (ISLR2 Section 4.3.5), extends the Bernoulli to a categorical
distribution with $K - 1$ linear predictors, one for each class other than a
reference class; it is fit with `nnet::multinom()`.
- **Ordinal (proportional-odds) regression**, for ordered classes, uses one
linear predictor together with $K - 1$ cutpoints; it is fit with
`MASS::polr()`.

None of these is taught or tested here.
:::

## Poisson regression

### Log-linear model

**Poisson regression** is the GLM with a Poisson random component and a log
link:

$$Y_i \mid x_i \stackrel{ind}{\sim} \text{Poisson}(\mu_i), \qquad \log \mu_i = x_i\beta = \beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip}.$$

Inverting the link,

$$\mu_i = e^{x_i\beta} = e^{\beta_0} \times e^{\beta_1 x_{i1}} \times \cdots \times e^{\beta_p x_{ip}},$$

which is positive for every $\beta$ and every row $x_i$. The features add on
the log scale and multiply on the scale of the mean.

### Poisson distribution

Under this model a count $Y_i$ has the **Poisson distribution** with mean
$\mu_i$, whose probability mass function is

$$P(Y_i = y \mid x_i) = \frac{\mu_i^{y} e^{-\mu_i}}{y!}, \qquad y = 0, 1, 2, \ldots,$$

where

- $\mu_i > 0$ is the mean, and
- $y! = 1 \times 2 \times \cdots \times y$, with $0! = 1$.

ISLR2 writes $\lambda$ for the Poisson mean; these notes write $\mu_i$, the
GLM symbol for a mean.

The mean and the variance are both $\mu_i$. Dropping the $y = 0$ term and
cancelling $y$ against $y!$,

$$E[Y_i \mid x_i] = \sum_{y=1}^\infty y\, \frac{\mu_i^{y} e^{-\mu_i}}{y!} = \mu_i \sum_{y=1}^\infty \frac{\mu_i^{y-1} e^{-\mu_i}}{(y-1)!} = \mu_i,$$

because the last sum adds the probability mass function over every count. The
same step applied twice gives $E[Y_i(Y_i - 1) \mid x_i] = \mu_i^2$, so

$$Var[Y_i \mid x_i] = E[Y_i(Y_i - 1) \mid x_i] + E[Y_i \mid x_i] - \mu_i^2 = \mu_i^2 + \mu_i - \mu_i^2 = \mu_i.$$

The slider below moves $\mu$. The black lollipops are the probability mass
function, the dashed line marks the mean, and the shaded band runs one
standard deviation, $\sqrt{\mu}$, either side of it. The horizontal axis
widens with $\mu$, so read the spread from its labels.

```{r poisson-slider-data}
#| echo: false
poisson_pmf <- expand_grid(mu = seq(1, 40, by = 0.5), y = 0:70) |>
  mutate(prob = dpois(y, mu)) |>
  filter(prob > 5e-4)
ojs_define(poisson_pmf = poisson_pmf)
```

```{ojs}
//| label: poisson-slider
//| echo: false
viewof mu = Inputs.range([1, 40], {value: 1, step: 0.5, label: "Mean μ"})
```

```{ojs}
//| label: poisson-slider-plot
//| echo: false
{
  const pmf = transpose(poisson_pmf).filter(d => Math.abs(d.mu - mu) < 1e-9);
  const sd = Math.sqrt(mu);
  const xmax = Math.max(10, Math.ceil(mu + 4 * sd));
  const plot = Plot.plot({
    width: 640, height: 300,
    x: {domain: [-0.5, xmax + 0.5], label: "Count y"},
    y: {domain: [0, 0.4], label: "P(Y = y)"},
    marks: [
      Plot.rect([{x1: mu - sd, x2: mu + sd}],
                {x1: "x1", x2: "x2", y1: 0, y2: 0.4,
                 fill: "#0072B2", fillOpacity: 0.12}),
      Plot.ruleX(pmf.filter(d => d.y <= xmax),
                 {x: "y", y1: 0, y2: "prob", strokeWidth: 2}),
      Plot.dot(pmf.filter(d => d.y <= xmax),
               {x: "y", y: "prob", r: 2.5, fill: "black"}),
      Plot.ruleX([mu], {stroke: "#0072B2", strokeDasharray: "6,4",
                        strokeWidth: 2})
    ]
  });
  return html`<p style="text-align: center"><b>mean = variance = ${mu.toFixed(1)}</b>,
    standard deviation = ${sd.toFixed(2)}, P(Y = 0) = ${Math.exp(-mu).toFixed(3)}</p>
    <div style="display: flex; justify-content: center">${plot}</div>`;
}
```

At a mean near $1$ most of the probability sits at $0$, $1$, and $2$, and the
distribution is strongly right-skewed. As $\mu$ grows the distribution moves
right, becomes nearly symmetric, and widens: its standard deviation $\sqrt\mu$
grows with its mean, the pattern the hourly bike counts showed.

### Coefficient interpretation

Write $\mu(x)$ for the mean count at a row of feature values
$x = (1, x_1, \ldots, x_p)$. Setting every feature to zero gives
$\log \mu = \beta_0$, so $e^{\beta_0}$ is the **mean count when every feature
is zero** — for dummy variables, at the reference levels.

Increasing feature $j$ by one unit from $x_j$ while holding every other
feature fixed, every term of the linear predictor except the $j$th cancels:

$$\log \mu(x_j + 1) - \log \mu(x_j) = \beta_j, \qquad \frac{\mu(x_j + 1)}{\mu(x_j)} = e^{\beta_j},$$

where

- $\mu(x_j)$ and $\mu(x_j + 1)$ are the mean counts at two feature rows that
differ only in feature $j$, by one unit.

So $\beta_j$ is the change in the log mean count per one-unit increase in
$x_j$, and $e^{\beta_j}$, the **rate ratio**, is the factor by which the mean
count is multiplied, holding the other features fixed. This parallels the odds
ratio of logistic regression with the mean count in place of the odds:
$e^{\beta_j} = 1$ ($\beta_j = 0$) means no association, and the mean changes
by $100\left(e^{\beta_j} - 1\right)$ percent per unit. A change of $c$ units
multiplies the mean by $e^{c\beta_j}$. For a dummy variable, $e^{\beta_j}$ is
the mean at the indicated level divided by the mean at the reference level.
With an interaction $x_j x_l$ in the model, the rate ratio for $x_j$ depends
on $x_l$, as an odds ratio did in the flexible logistic regression lecture.

### Maximum likelihood estimation

For independent observations the log-likelihood is the sum of the logarithms
of the probability mass functions,

$$\ell(\beta) = \sum_{i=1}^n \left[y_i \log \mu_i - \mu_i - \log(y_i!)\right] = \sum_{i=1}^n \left[y_i\, x_i\beta - e^{x_i\beta} - \log(y_i!)\right],$$

where

- $y_i$ is the observed count for observation $i$, and
- $\log(y_i!)$ does not involve $\beta$ and so does not affect the maximizer.

Differentiating with respect to $\beta_j$, using
$\partial e^{x_i\beta} / \partial \beta_j = x_{ij} e^{x_i\beta}$, and setting
the result to zero gives the **score equations**

$$\sum_{i=1}^n x_{ij}\left(y_i - \mu_i\right) = 0, \qquad j = 0, 1, \ldots, p,$$

with $x_{i0} = 1$. They have exactly the form of the logistic regression score
equations with $\mu_i$ in place of $p(x_i)$: the residuals $y_i - \mu_i$ sum
to zero and are orthogonal to every feature. Because $\mu_i = e^{x_i\beta}$ is
nonlinear in $\beta$, the equations have no closed-form solution.

`glm()` solves them with the iterative algorithm it used for logistic
regression, **iteratively reweighted least squares**: each iteration solves a
weighted least squares problem whose weights are recomputed from the current
estimate.

Two consequences of the score equations are used below. The intercept's
equation makes the fitted means sum to the observed counts,
$\sum_i \hat\mu_i = \sum_i y_i$. More generally, the equation for a dummy
variable makes the fitted means average to the observed mean among the
observations with that dummy equal to $1$; the normal equations of least
squares do the same.

### Inference

Inference repeats the logistic regression lectures. In large samples
$\hat\beta_j$ is approximately normal, so

$$z = \frac{\hat\beta_j - \beta_j^0}{\text{SE}(\hat\beta_j)}$$

is approximately standard normal under $H_0: \beta_j = \beta_j^0$, where
$\text{SE}(\hat\beta_j)$ comes from the curvature of $\ell$ at $\hat\beta$. The
Wald interval $\hat\beta_j \pm z_{1-\alpha/2}\,\text{SE}(\hat\beta_j)$ is
returned by `confint.default()`, and exponentiating its endpoints gives an
interval for the rate ratio $e^{\beta_j}$.

A group of coefficients — every product of an hour dummy with the working-day
indicator, say — is tested with the drop-in-deviance test. For a count
response `glm()` reports the residual deviance

$$D = 2\left[\ell_{\text{sat}} - \ell(\hat\beta)\right],$$

where

- $\ell_{\text{sat}}$ is the log-likelihood of the **saturated** model, which
gives each observation its own mean $\mu_i = y_i$, the largest log-likelihood
any mean model can reach.

Here $D$ is the deviance, as in the multiple logistic regression lecture, not
an indicator variable.

$\ell_{\text{sat}}$ does not involve $\beta$, so for a reduced model nested in
a full one, $G^2 = D_r - D_f = -2\left[\ell(\hat\beta_r) - \ell(\hat\beta_f)\right]$,
the drop-in-deviance statistic of the multiple logistic regression lecture.
(For a binary response $\ell_{\text{sat}} = 0$, which is why that lecture
could write $D = -2\ell(\hat\beta)$.) Under $H_0$ that the $k$ dropped
coefficients are zero, $G^2$ is approximately $\chi^2_k$, and
`anova(reduced, full, test = "Chisq")` performs the test.

### Overdispersion

The Poisson model has no variance parameter: once $\mu_i$ is specified,
$Var[Y_i \mid x_i] = \mu_i$ is determined, and every standard error and
deviance test above is computed under that variance. **Overdispersion** is
the failure $Var[Y_i \mid x_i] > \mu_i$.

The **Pearson residual** $(y_i - \hat\mu_i)/\sqrt{\hat\mu_i}$ standardizes a
residual by the Poisson standard deviation. If $Var[Y_i \mid x_i] = \mu_i$,
then $E\left[(Y_i - \mu_i)^2/\mu_i \mid x_i\right] = 1$, so the average
squared Pearson residual,

$$\hat\phi = \frac{1}{n - (p+1)}\sum_{i=1}^n \frac{\left(y_i - \hat\mu_i\right)^2}{\hat\mu_i},$$

should be close to $1$, where

- $\hat\phi$ estimates the dispersion parameter $\phi$, which the Poisson
model fixes at $1$, and
- the divisor $n - (p+1)$ accounts for the $p + 1$ estimated coefficients, as
in $\hat\sigma^2 = RSS/[n - (p+1)]$.

A value well above $1$ is evidence of overdispersion. A plot shows the same
thing: group the observations by fitted mean, and compare each group's average
squared residual $(y_i - \hat\mu_i)^2$ with its average $\hat\mu_i$.

The score equations involve $y_i$ only through $y_i - \mu_i$, never through
its variance, so when the mean model is right $\hat\beta$ remains a sensible
estimate under overdispersion, and so do the fitted means and rate ratios.
The standard errors do not. Computed as if the variance were $\mu_i$, they are
too small when it is larger, so $z$-statistics and $G^2$ are too large,
$p$-values too small, and confidence intervals too narrow.

Overdispersion has the same sources as in the problems in logistic regression
lecture: explanatory variables missing from the model that vary among
observations with the same $x_i$, and events within one count that are not
independent — riders who rent together. The remedy within this course is the same: add features that explain the extra
variation.

::: {.callout-note collapse="true"}
### Beyond this course
Two standard remedies change the variance model rather than the mean model.
**Quasi-Poisson** regression keeps $\log \mu_i = x_i\beta$ and the estimates
$\hat\beta$ but assumes $Var[Y_i \mid x_i] = \phi\,\mu_i$, estimating $\phi$ by
$\hat\phi$ and multiplying every standard error by $\sqrt{\hat\phi}$
(`family = quasipoisson` in `glm()`); deviance tests become $F$-tests.
**Negative binomial** regression replaces the Poisson with a distribution whose
variance is $\mu_i + \mu_i^2/\theta$ for an additional parameter $\theta$,
fit by maximum likelihood with `MASS::glm.nb()`. Neither is taught or tested
here.
:::

## Hourly bike rentals

The question, data, and exploratory analysis are those of the count response
section above, and the Poisson regression uses the features of its linear
regression.

### Fitted means

```{r bike-glm}
bike_glm <- glm(bike_formula, data = bike, family = poisson)
```

The model has `r length(coef(bike_glm))` coefficients, with the same features
as `bike_lm`, and `glm()` reports convergence after `r bike_glm$iter`
iterations.

The figure below splits the hours by weather (rows) and working-day status
(columns). Each point is the observed mean count for one hour of the day in
that panel; each line is the average, over the same hours, of a model's fitted
means. The first tab shows the data alone, and the other two add one model
each.

::: {.callout-note collapse="true"}
### Code
```{r fitted-means}
#| output: false
bike <- bike |>
  mutate(fit_glm = fitted(bike_glm))

cell_means <- bike |>
  group_by(weather, working, hour) |>
  summarize(n = n(), observed = mean(bikers),
            `Linear regression` = mean(fit_lm),
            `Poisson regression` = mean(fit_glm), .groups = "drop")

cell_long <- cell_means |>
  pivot_longer(c(`Linear regression`, `Poisson regression`),
               names_to = "model", values_to = "fitted")

rain_night <- filter(cell_means, weather == "rain/snow", hour <= 6)

fitted_base <- ggplot(cell_means, aes(x = hour, y = observed)) +
  geom_hline(yintercept = 0, color = "grey50", linetype = "dotted") +
  geom_point(size = 1.3, shape = 1) +
  facet_grid(weather ~ working, scales = "free_y",
             labeller = labeller(working = working_labels)) +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  labs(x = "Hour of the day", y = "Mean bikes rented in the hour",
       color = NULL, linetype = NULL) +
  theme(legend.position = "bottom")

fitted_observed_plot <- fitted_base +
  labs(title = "Observed hourly means")

model_layer <- function(m) {
  list(geom_line(data = filter(cell_long, model == m),
                 aes(y = fitted, color = model, linetype = model),
                 linewidth = 0.8),
       scale_color_manual(values = model_colors),
       scale_linetype_manual(values = model_linetypes))
}
fitted_lm_plot <- fitted_base + model_layer("Linear regression") +
  labs(title = "Linear regression fitted means")
fitted_glm_plot <- fitted_base + model_layer("Poisson regression") +
  labs(title = "Poisson regression fitted means")
```
:::

::: {.panel-tabset}
#### Observed
```{r fitted-observed-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 6.5
fitted_observed_plot
```

#### Linear regression
```{r fitted-lm-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 6.5
fitted_lm_plot
```

#### Poisson regression
```{r fitted-glm-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 6.5
fitted_glm_plot
```
:::

In both models the hour dummies, the working-day indicator, and their
products give each combination of hour and working-day status its own mean,
so by their score and normal equations both reproduce the observed mean of
each combination exactly once the weather rows are pooled. The models
differ in how they split each combination across the weather rows. In the
`rain/snow` row, linear regression subtracts its constant weather coefficient
from every hour, and its lines fall below zero overnight, as low as
`r fmt(min(rain_night$"Linear regression"), 1)` bikers, where the observed
means never fall below `r fmt(min(rain_night$observed), 1)`. The Poisson lines
scale the hourly profile down instead and follow the points much more
closely, overnight included.

### Rate ratios

::: {.callout-note collapse="true"}
### Code
```{r rate-ratios}
#| output: false
b    <- coef(bike_glm)
se   <- summary(bike_glm)$coefficients[, "Std. Error"]
z    <- summary(bike_glm)$coefficients[, "z value"]
ci   <- confint.default(bike_glm)
shown <- c("(Intercept)", "workingYes", "temp_c", "weathercloudy/misty",
           "weatherrain/snow")

rate_ratio_table <- tibble(
  Coefficient = c("Intercept", "Working day (at hour 0)",
                  "Temperature (per degree C)", "Cloudy/misty vs. clear",
                  "Rain/snow vs. clear"),
  Estimate = sprintf("%.4f", b[shown]),
  SE = sprintf("%.4f", se[shown]),
  z = sprintf("%.1f", z[shown]),
  exp_b = sprintf("%.4f", exp(b[shown])),
  lower = sprintf("%.4f", exp(ci[shown, 1])),
  upper = sprintf("%.4f", exp(ci[shown, 2]))
) |>
  knitr::kable(col.names = c("Coefficient", "$\\hat\\beta_j$", "SE", "$z$",
                             "$e^{\\hat\\beta_j}$", "95% CI lower",
                             "95% CI upper"),
               align = "lcccccc")

rr_working_8 <- exp(b["workingYes"] + b["hr8:workingYes"])
rr_temp_5    <- exp(5 * b["temp_c"])
ci_temp_5    <- exp(5 * ci["temp_c", ])
rr_rain      <- exp(b["weatherrain/snow"])
```
:::

The intercept and the coefficients for working day, temperature, and
weather, each with its exponential and a 95% Wald interval for that
exponential:

```{r rate-ratio-table}
#| echo: false
rate_ratio_table
```

The intercept comes first. $e^{\hat\beta_0} = `r sprintf("%.1f", exp(b[1]))`$
is the estimated mean count at hour $0$ (midnight to 1 a.m.) on a non-working
day in January, in clear weather at $0$ degrees Celsius — a realistic winter
night, unlike the linear regression's intercept of
`r fmt(b_lm[1], 1)` bikers for the same conditions.

Because hour and working day interact, the working-day coefficient compares
the two kinds of day at the reference hour only: at midnight a working day's
estimated mean is $e^{\hat\beta_{\texttt{workingYes}}} = `r sprintf("%.2f", exp(b["workingYes"]))`$
times a non-working day's. At 8 a.m. the corresponding ratio adds the
`hr8:workingYes` coefficient to the exponent,
$e^{`r sprintf("%.3f", b["workingYes"])` `r sprintf("%+.3f", b["hr8:workingYes"])`} = `r sprintf("%.2f", rr_working_8)`$.

Holding hour, working-day status, month, and weather fixed, each additional
degree Celsius multiplies the estimated mean count by
`r sprintf("%.4f", exp(b["temp_c"]))`, so $5$ degrees multiply it by
$e^{5\hat\beta_{\texttt{temp}\_\texttt{c}}} = `r sprintf("%.3f", rr_temp_5)`$, an increase of
`r sprintf("%.1f", 100 * (rr_temp_5 - 1))`%, with 95% interval
$(`r sprintf("%.3f", ci_temp_5[1])`, `r sprintf("%.3f", ci_temp_5[2])`)$.

Rain or snow multiplies the estimated mean by
`r sprintf("%.3f", rr_rain)` relative to clear weather — a
`r sprintf("%.0f", 100 * (1 - rr_rain))`% reduction at every hour, month, and
temperature — and cloudy or misty weather multiplies it by
`r sprintf("%.3f", exp(b["weathercloudy/misty"]))`. Every rate-ratio interval
in the table excludes $1$.

### Interaction test

The hour-by-working-day products are `r sum(grepl(":", names(b)))`
coefficients, so no single $z$-test asks whether working and non-working days
share one hourly profile up to a constant factor. The drop-in-deviance test
compares the full model with the additive one:

```{r interaction-test}
bike_additive <- glm(bikers ~ hr + working + mnth + temp_c + weather,
                     data = bike, family = poisson)
interaction_test <- anova(bike_additive, bike_glm, test = "Chisq")
interaction_test
```

```{r interaction-test-values}
#| include: false
G2 <- interaction_test$Deviance[2]
k  <- interaction_test$Df[2]
log10_p <- pchisq(G2, df = k, lower.tail = FALSE, log.p = TRUE) / log(10)
```

The deviance drops by $G^2 = `r format(round(G2), big.mark = "{,}")`$ on $k = `r k`$ degrees of freedom.
A $\chi^2_{`r k`}$ variable has mean `r k`, and the upper tail probability
beyond $G^2$ is about $10^{`r format(round(log10_p), big.mark = "{,}")`}$,
far below the smallest value `anova()` prints,
$`r sci(.Machine$double.eps)`$. The
data are inconsistent with one hourly profile shared by both kinds of day.

### Variance check

::: {.callout-note collapse="true"}
### Code
```{r variance-check}
#| output: false
pearson_phi <- function(fit) {
  sum(residuals(fit, type = "pearson")^2) / df.residual(fit)
}
phi_full     <- pearson_phi(bike_glm)
phi_additive <- pearson_phi(bike_additive)

variance_bins <- bike |>
  mutate(bin = ntile(fit_glm, 20)) |>
  group_by(bin) |>
  summarize(mean_fit = mean(fit_glm),
            mean_sq_resid = mean((bikers - fit_glm)^2), .groups = "drop")

variance_check_plot <- ggplot(variance_bins,
                              aes(x = mean_fit, y = mean_sq_resid)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
              color = "grey40") +
  geom_point(size = 2.3) +
  annotate("text", x = 60, y = 22, label = "Poisson: variance = mean",
           hjust = 0, color = "grey30") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = expression(paste("Average ", hat(mu)[i], " in the bin (log scale)")),
       y = expression(paste("Average ", (y[i] - hat(mu)[i])^2,
                            " in the bin (log scale)")),
       title = "Poisson regression, hours binned into 20 groups by fitted mean")

rain_se    <- summary(bike_glm)$coefficients["weatherrain/snow", "Std. Error"]
rain_ci_q  <- exp(b["weatherrain/snow"] +
                  c(-1, 1) * qnorm(0.975) * sqrt(phi_full) * rain_se)
```
:::

```{r variance-check-figure}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.5
variance_check_plot
```

`r if (all(variance_bins$mean_sq_resid > variance_bins$mean_fit)) paste("All", nrow(variance_bins), "bins have") else paste(sum(variance_bins$mean_sq_resid > variance_bins$mean_fit), "of the", nrow(variance_bins), "bins have")`
an average squared residual above the dashed line. The average squared Pearson residual is
$\hat\phi = `r sprintf("%.1f", phi_full)`$: the hourly counts vary about
`r round(phi_full)` times as much around their fitted means as a Poisson
distribution allows. The additive model's $\hat\phi$ is
`r sprintf("%.1f", phi_additive)`, so the hour-by-working-day features
explained part of the excess variation, but $\hat\phi$ remains far above
$1$. The remaining excess plausibly comes from explanatory variables not in
the model, such as conditions shared by the hours of one day.

If the mean model is right, the fitted means and rate ratios above stand; the
standard errors,
$z$-statistics, intervals, and $G^2$, all computed as if $\phi = 1$,
overstate their precision.

::: {.callout-note collapse="true"}
### Beyond this course
A quasi-Poisson fit keeps every estimate above and multiplies every standard
error by $\sqrt{\hat\phi} = `r sprintf("%.2f", sqrt(phi_full))`$. The 95%
interval for the `rain/snow` rate ratio widens from
$(`r sprintf("%.3f", exp(ci["weatherrain/snow", 1]))`, `r sprintf("%.3f", exp(ci["weatherrain/snow", 2]))`)$
to $(`r sprintf("%.3f", rain_ci_q[1])`, `r sprintf("%.3f", rain_ci_q[2])`)$,
still far from $1$.
:::

**Conclusion.** The mean hourly count is modeled well by a Poisson regression
in which hour and working-day status interact and month, temperature, and
weather multiply the mean. Its fitted means are positive everywhere, and its
multiplicative weather effect follows the rainy-hour means where linear
regression's constant subtraction predicts negative counts overnight. Rain or
snow multiplies the estimated mean count by
`r sprintf("%.2f", rr_rain)`, and each $5$ degrees Celsius by
`r sprintf("%.2f", rr_temp_5)`; working and non-working days have different
hourly profiles. The counts are overdispersed, with $\hat\phi$ near
`r round(phi_full)`, so these estimates are usable if the mean model is right,
while the reported standard errors and intervals are too small.

## Conclusion

A generalized linear model has three components: a distribution for the
response given the features, with mean $\mu_i$; a linear predictor
$\eta_i = x_i\beta$, into which every feature-engineering device of the
earlier lectures enters unchanged; and a link function with
$g(\mu_i) = \eta_i$, whose inverse keeps the mean within the values the
response allows. Linear regression is the normal distribution with the
identity link, logistic regression the Bernoulli with the logit link, and
Poisson regression the Poisson with the log link. Each distribution also fixes
its variance function: constant, $\mu(1-\mu)$, and $\mu$.

Poisson regression models a count. Its mean $e^{x_i\beta}$ is positive for
every $\beta$, so $e^{\beta_0}$ is a mean count at the reference values of the
features and $e^{\beta_j}$ a rate ratio, the factor multiplying the mean count
per unit of feature $j$ with the other features held fixed. Its coefficients
are maximum likelihood estimates; the score equations
$\sum_i x_{ij}(y_i - \mu_i) = 0$ have the logistic regression form, have no
closed-form solution, and are solved by iteratively reweighted least squares.
Inference uses the same Wald $z$-tests and intervals and the same
drop-in-deviance test as logistic regression.

All of that inference assumes $Var[Y_i \mid x_i] = \mu_i$. The average
squared Pearson residual estimates the ratio of the actual variance to the
mean; a value well above $1$ signals overdispersion, under which the estimates
remain usable if the mean model is right, but standard errors are too small,
tests too liberal, and intervals too narrow.

The next lecture begins the resampling unit, ISLR2 Chapter 5. Every error
rate and fit measure so far has been computed on the data the model was fit
to, or on a single held-out split; cross-validation estimates test error by
repeatedly holding out part of the data, and uses those estimates to choose
among models and flexibility settings.
