# Generalized Linear Models

```{r packages}
#| include: false
library("tidyverse")
library("ISLR2")
theme_set(theme_bw())
set.seed(20261001)

working_colors  <- c("No" = "#009E73", "Yes" = "#CC79A7")
working_labels  <- c("No" = "Non-working day", "Yes" = "Working day")
model_colors    <- c("Linear regression" = "#D55E00",
                     "Poisson regression" = "#0072B2")
model_linetypes <- c("Linear regression" = "dashed",
                     "Poisson regression" = "solid")

# Renders a number in LaTeX scientific notation for use inside $...$.
sci <- function(x, digits = 2) {
  e <- floor(log10(abs(x)))
  m <- signif(x / 10^e, digits)
  sprintf("%s \\times 10^{%d}", format(m), e)
}

# Rounds and adds thousands separators for inline numbers.
fmt <- function(x, digits = 0) {
  format(round(x, digits), big.mark = ",", nsmall = digits, scientific = FALSE)
}
```

::: {.callout-tip}
### Learning objectives
After this lecture you should be able to

- **write** a generalized linear model as its three components — a
distribution for $Y_i$ given $x_i$ with mean $\mu_i$, a linear predictor
$\eta_i = x_i\beta$, and a link function $g$ with $g(\mu_i) = \eta_i$ — and
**identify** linear, logistic, and Poisson regression as special cases by their
distribution, link, and variance,
- **explain**, using fitted values and residuals, why linear regression is a
poor model for a count response,
- **write** the Poisson regression log-likelihood, **derive** its score
equations, and **explain** why, in general, its maximizer has no closed form,
- **interpret** $e^{\beta_0}$ as a mean count and $e^{\beta_j}$ as a rate
ratio, the factor multiplying the mean count per one-unit increase in feature
$j$ with the other features held fixed, and **construct** Wald tests and
confidence intervals for $\beta_j$ and $e^{\beta_j}$,
- **conduct** a drop-in-deviance test of a group of coefficients in a Poisson
regression, and
- **assess** the Poisson assumption $Var[Y_i \mid x_i] = \mu_i$ with Pearson
residuals, and **explain** what overdispersion does to Wald standard errors
and tests.
:::

## Count response

The regression lectures modeled a quantitative response whose mean could be
any real number, and the logistic regression lectures a binary response whose
mean is a probability. A **count** — the number of events in a fixed period
or region — is neither: it is a nonnegative integer, its mean must be
positive, and its spread typically grows with its mean.

### Bike rentals

**Question.** How does the number of bikes rented in an hour depend on the
time of day, whether the day is a working day, the month, the temperature,
and the weather?

**Data.** `ISLR2::Bikeshare` records `r fmt(nrow(Bikeshare))` hours of the
Capital Bikeshare system in Washington, D.C., spread over the
`r max(Bikeshare$day)` days of one year
(`r 24 * max(Bikeshare$day) - nrow(Bikeshare)` hours of the year have no
record). The response is `bikers`, the number of bikes rented in the hour.
The explanatory variables used here are

- `hr`, the hour of the day, a factor with levels `0` through `23`,
- `workingday`, $1$ for a day that is neither a weekend nor a holiday and $0$
otherwise,
- `mnth`, the month, a factor,
- `temp`, the temperature rescaled to $[0, 1]$ from the range
$-8$ to $39$ degrees Celsius, and
- `weathersit`, the weather, a factor with levels
`r paste0("\x60", levels(Bikeshare$weathersit), "\x60", collapse = ", ")`.

`bikers` is the sum of `casual` and `registered`; those are pieces of the
response, not explanatory variables, and no model below uses them.

### Hourly counts

::: {.callout-note collapse="true"}
### Code
```{r bike-eda}
#| output: false
bike <- Bikeshare |>
  mutate(working  = factor(workingday, levels = c(0, 1),
                           labels = c("No", "Yes")),
         weather  = fct_collapse(weathersit,
                                 `rain/snow` = c("light rain/snow",
                                                 "heavy rain/snow")),
         temp_c   = 47 * temp - 8,
         hour     = as.numeric(as.character(hr)),
         hour_jit = hour + runif(n(), -0.3, 0.3))

hour_groups <- bike |>
  group_by(working, hour) |>
  summarize(n = n(), mean = mean(bikers), variance = var(bikers),
            .groups = "drop")

mv_slope <- unname(coef(lm(log(variance) ~ log(mean), data = hour_groups))[2])
busiest  <- slice_max(hour_groups, mean, n = 1)
n_heavy  <- sum(Bikeshare$weathersit == "heavy rain/snow")

# Poisson probability mass functions for the slider below.
poisson_pmf <- expand_grid(mu = seq(1, 40, by = 0.5), y = 0:70) |>
  mutate(prob = dpois(y, mu)) |>
  filter(prob > 5e-4)

count_hist_plot <- ggplot(bike, aes(x = bikers)) +
  geom_histogram(binwidth = 10, boundary = 0, fill = "grey70",
                 color = "grey40") +
  labs(x = "Bikes rented in the hour", y = "Hours",
       title = "ISLR2::Bikeshare, hourly counts")

count_hour_plot <- ggplot(bike, aes(x = hour_jit, y = bikers)) +
  geom_point(alpha = 0.08, size = 0.5) +
  geom_line(data = hour_groups, aes(x = hour, y = mean), linewidth = 0.9) +
  geom_point(data = hour_groups, aes(x = hour, y = mean), size = 1.8) +
  facet_wrap(~ working, labeller = labeller(working = working_labels)) +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  labs(x = "Hour of the day", y = "Bikes rented in the hour",
       title = "Hourly counts (jittered) and hourly means")

mean_variance_plot <- ggplot(hour_groups,
                             aes(x = mean, y = variance, color = working,
                                 shape = working)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
              color = "grey40") +
  geom_point(size = 2.3) +
  annotate("text", x = 100, y = 55, label = "variance = mean", angle = 24,
           color = "grey30") +
  scale_x_log10(breaks = c(5, 10, 30, 100, 300)) +
  scale_y_log10() +
  scale_color_manual(values = working_colors, labels = working_labels) +
  scale_shape_manual(values = c(16, 17), labels = working_labels) +
  labs(x = "Mean count in the group (log scale)",
       y = "Variance of the count in the group (log scale)",
       color = NULL, shape = NULL,
       title = "One point per hour-by-working-day group")
```
:::

The first tab is a histogram of the hourly counts. The second plots each
hour's count, jittered horizontally, against the hour of the day, with the
hourly means joined by a line, separately for non-working and working days.
The third places each of the `r nrow(hour_groups)` hour-by-working-day groups
at its mean and variance on log scales, with color and shape marking working
and non-working days and a dashed line where the variance equals the mean.

::: {.panel-tabset}
#### Distribution
```{r count-hist-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.4
count_hist_plot
```

#### Hourly profile
```{r count-hour-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 3.6
count_hour_plot
```

#### Mean and variance
```{r mean-variance-figure}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.5
mean_variance_plot
```
:::

The counts run from `r min(bike$bikers)` to `r max(bike$bikers)` with median
`r median(bike$bikers)` and mean `r fmt(mean(bike$bikers), 1)`: every value is
a positive integer, and the distribution is strongly right-skewed.

On working days the hourly means peak twice, at hour
`r with(filter(hour_groups, working == "Yes", hour < 12), hour[which.max(mean)])`
and hour
`r with(filter(hour_groups, working == "Yes", hour >= 12), hour[which.max(mean)])`,
the commute. On non-working days they rise to a single midday plateau. The two
hourly profiles differ in shape, not just in level. Across hours, the points
spread out more where the mean is higher.

In the third tab the variance climbs steadily with the mean; on these log-log
axes the points lie near a line of slope `r sprintf("%.2f", mv_slope)`.
`r if (all(hour_groups$variance > hour_groups$mean)) paste("All", nrow(hour_groups), "points lie") else paste(sum(hour_groups$variance > hour_groups$mean), "of the", nrow(hour_groups), "points lie")`
above the dashed line. Month, temperature, and weather vary within each
group, so part of each group's variance is variation those explanatory
variables could explain.

### Poisson distribution

The dashed line is the variance of the standard model for a count. A count
$Y$ with a **Poisson distribution** with mean $\mu$ has probability mass
function

$$P(Y = y) = \frac{\mu^{y} e^{-\mu}}{y!}, \qquad y = 0, 1, 2, \ldots,$$

where

- $\mu > 0$ is the mean, and
- $y! = 1 \times 2 \times \cdots \times y$, with $0! = 1$.

ISLR2 writes $\lambda$ for the Poisson mean; these notes write $\mu$, the
symbol the generalized linear models below use for a mean.

The mean and the variance are both $\mu$. Dropping the $y = 0$ term and
cancelling $y$ against $y!$,

$$E[Y] = \sum_{y=1}^\infty y\, \frac{\mu^{y} e^{-\mu}}{y!} = \mu \sum_{y=1}^\infty \frac{\mu^{y-1} e^{-\mu}}{(y-1)!} = \mu,$$

because the last sum adds the probability mass function over every count. The
same step applied twice gives $E[Y(Y - 1)] = \mu^2$, so

$$Var[Y] = E[Y(Y - 1)] + E[Y] - \mu^2 = \mu^2 + \mu - \mu^2 = \mu.$$

```{r poisson-slider-define}
#| echo: false
ojs_define(poisson_pmf = poisson_pmf)
```

The slider below moves $\mu$. The black lollipops are the probability mass
function, the dashed line marks the mean, the shaded band runs one standard
deviation, $\sqrt{\mu}$, either side of it; both axes rescale as $\mu$ moves,
so read the spread and the heights from the axis labels.

```{ojs}
//| label: poisson-slider
//| echo: false
viewof mu = Inputs.range([1, 40], {value: 1, step: 0.5, label: "Mean μ"})
```

```{ojs}
//| label: poisson-slider-plot
//| echo: false
{
  const pmf = transpose(poisson_pmf).filter(d => Math.abs(d.mu - mu) < 1e-9);
  const sd = Math.sqrt(mu);
  const xmax = Math.max(10, Math.ceil(mu + 4 * sd));
  const ymax = 1.15 * Math.max(...pmf.map(d => d.prob));
  const plot = Plot.plot({
    width: 640, height: 300,
    x: {domain: [-0.5, xmax + 0.5], label: "Count y"},
    y: {domain: [0, ymax], label: "P(Y = y)"},
    marks: [
      Plot.rect([{x1: mu - sd, x2: mu + sd}],
                {x1: "x1", x2: "x2", y1: 0, y2: ymax,
                 fill: "#0072B2", fillOpacity: 0.12}),
      Plot.ruleX(pmf.filter(d => d.y <= xmax),
                 {x: "y", y1: 0, y2: "prob", strokeWidth: 2}),
      Plot.dot(pmf.filter(d => d.y <= xmax),
               {x: "y", y: "prob", r: 2.5, fill: "black"}),
      Plot.ruleX([mu], {stroke: "#0072B2", strokeDasharray: "6,4",
                        strokeWidth: 2})
    ]
  });
  return html`<p style="text-align: center"><b>mean = variance = ${mu.toFixed(1)}</b>,
    standard deviation = ${sd.toFixed(2)}, P(Y = 0) = ${Math.exp(-mu).toFixed(3)}</p>
    <div style="display: flex; justify-content: center">${plot}</div>`;
}
```

At a mean near $1$ most of the probability sits at $0$, $1$, and $2$, and the
distribution is strongly right-skewed. As $\mu$ grows the distribution moves
right, becomes nearly symmetric, and widens: its standard deviation $\sqrt\mu$
grows with its mean. The slider stops at $40$. The busiest bike group, with
mean `r fmt(busiest$mean)`, would have Poisson standard deviation
`r fmt(sqrt(busiest$mean), 1)`; its observed standard deviation is
`r fmt(sqrt(busiest$variance), 1)` (variance `r fmt(busiest$variance)`), the
gap the mean–variance plot showed for every group.

### Linear regression fails

A linear regression fit by least squares uses these features:

- `r nlevels(bike$hr) - 1` dummy features for `hr`, with hour `0` the reference,
- the indicator $\mathrm{I}(\texttt{workingday} = 1)$,
- `r nlevels(bike$hr) - 1` products of each hour dummy with the working-day
indicator, because the two hourly profiles differ in shape,
- `r nlevels(bike$mnth) - 1` dummy features for `mnth`, with January the
reference,
- the temperature in degrees Celsius, the feature $h(\texttt{temp}) = 47\,\texttt{temp} - 8$, and
- `r nlevels(bike$weather) - 1` dummy features for the weather, with `clear`
the reference, after merging `heavy rain/snow` into `light rain/snow` as
`rain/snow`, since `heavy rain/snow` occurs in
`r paste(n_heavy, if (n_heavy == 1) "hour" else "hours")`.

```{r bike-lm}
bike_formula <- bikers ~ hr * working + mnth + temp_c + weather
bike_lm <- lm(bike_formula, data = bike)
```

::: {.callout-note collapse="true"}
### Code
```{r bike-lm-residuals}
#| output: false
bike <- bike |>
  mutate(fit_lm = fitted(bike_lm), resid_lm = residuals(bike_lm))

n_negative <- sum(bike$fit_lm < 0)
fit_low  <- 50
fit_high <- 300
sd_low  <- sd(bike$resid_lm[bike$fit_lm < fit_low])
sd_high <- sd(bike$resid_lm[bike$fit_lm > fit_high])
b_lm    <- coef(bike_lm)

lm_residual_plot <- ggplot(bike, aes(x = fit_lm, y = resid_lm)) +
  geom_point(alpha = 0.1, size = 0.6) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#D55E00") +
  geom_hline(yintercept = 0, color = "grey40") +
  labs(x = "Fitted value from least squares",
       y = "Residual", title = "Linear regression residuals")
```
:::

The figure plots each hour's residual against its least squares fitted value;
the dashed vertical line marks a fitted value of zero and the solid horizontal
line a residual of zero.

```{r lm-residual-figure}
#| echo: false
#| fig-width: 7
#| fig-height: 3.6
lm_residual_plot
```

Three failures appear.

1. **Negative fitted means.** Every point left of the dashed line has a
negative fitted value: `r fmt(n_negative)` of the `r fmt(nrow(bike))` hours,
`r sprintf("%.1f", 100 * n_negative / nrow(bike))`%, with the smallest at
`r fmt(min(bike$fit_lm), 1)` bikers. A mean count cannot be negative.
2. **Variance growing with the mean.** The residuals fan out to the right. Their
standard deviation is `r fmt(sd_low, 1)` among hours with fitted values below
$`r fit_low`$ and `r fmt(sd_high, 1)` among hours with fitted values above
$`r fit_high`$, while every $t$-test and interval from `lm()` assumes one common
variance.
3. **Additive effects.** Every additive effect — month, temperature, and
weather — shifts every hour by the same number of riders. Within each
combination of hour and working-day status, an hour whose shift is smaller
than average — a winter month, a cold hour, rain — sits the same number of
riders below the combination's mean at 3 a.m. as at 5 p.m., and at night that
is enough to push the fit below zero:
`r fmt(sum(bike$fit_lm < 0 & bike$weather == "clear"))` of the
`r fmt(n_negative)` negative fitted values are in clear weather, and the
median temperature of all `r fmt(n_negative)` is
`r sprintf("%.1f", median(bike$temp_c[bike$fit_lm < 0]))` degrees Celsius,
against `r sprintf("%.1f", median(bike$temp_c))` for all hours.

## Generalized linear models

The figure repeats the mean–variance plot of the hourly counts: each of the
`r nrow(hour_groups)` hour-by-working-day groups at its mean and variance on log
scales, with color and shape marking working and non-working days and a dashed
line where the variance equals the mean.

```{r glm-mean-variance-figure}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.5
mean_variance_plot
```

The variance grows with the mean, and much faster than the constant variance
the linear regression assumed. A generalized linear model lets the
distribution, and with it the way the variance depends on the mean, change
with the response. The
[Problems in Logistic Regression](40-problems-in-logistic-regression.qmd)
lecture wrote each model as one distributional statement:
$Y_i \mid x_i \stackrel{ind}{\sim} N(x_i\beta, \sigma^2)$ for linear
regression, and
$Y_i \mid x_i \stackrel{ind}{\sim} \text{Bernoulli}\left[p(x_i)\right]$ with
$\log\left[\frac{p(x_i)}{1 - p(x_i)}\right] = x_i\beta$ for logistic
regression. Both statements have the same three parts — a distribution, a
linear predictor, and a link — and generalized linear models keep those parts
while letting each one vary.

### Three components

A **generalized linear model** (GLM) assumes the statistical model

$$Y_i \mid x_i \stackrel{ind}{\sim} \text{a distribution with mean } \mu_i, \qquad g(\mu_i) = \eta_i, \qquad \eta_i = x_i\beta,$$

where

- $\mu_i = E[Y_i \mid x_i]$ is the mean of the response of observation $i$
(unrelated to the class means $\mu_k$ of the discriminant analysis lectures),
- $x_i = (1, x_{i1}, \ldots, x_{ip})$ is the $i$th row of the model matrix
$\mathbf{X}$ and $\beta = (\beta_0, \ldots, \beta_p)^\top$ is the vector of
coefficients,
- $\eta_i$ is the **linear predictor**, as in the problems lecture, and
- $g$ is the **link function**, a strictly monotone function (ISLR2 writes
$\eta$ for the link itself; these notes write $g$).

The distribution of $Y_i$ given $x_i$ is the **random component**. The linear
predictor is where the features enter: dummy variables, polynomials, step
functions, and interactions enter it exactly as they entered a linear or
logistic regression. The link connects the two; inverting it,
$\mu_i = g^{-1}(\eta_i) = g^{-1}(x_i\beta)$.

The link is what keeps the mean in range. A linear predictor can take any real
value, and for the identity, logit, and log links used in this lecture $g^{-1}$ carries the real line
onto the values the mean is allowed to take — $(0, 1)$ for a probability,
$(0, \infty)$ for a count — so every $\beta$ gives a valid mean.

For the bike counts, $Y_i$ is `bikers` in hour $i$, and $x_i$ is the row of
`r length(coef(bike_lm))` entries — a leading $1$ and the features of the
linear regression above — so $\eta_i$ is the same linear predictor in the
linear regression above and in the Poisson regression introduced below. The
linear regression is the GLM with a
normal random component and the identity link, $g(\mu_i) = \mu_i$, which does
nothing to keep $x_i\hat\beta$ positive: that is how `r fmt(n_negative)` of its
fitted means fell below zero.

### Special cases

::: {.callout-note collapse="true"}
### Code
```{r glm-table}
#| output: false
glm_table <- tibble(
  Model = c("Linear regression", "Logistic regression", "Poisson regression"),
  Response = c("Quantitative", "Binary ($0/1$)", "Count"),
  Distribution = c("$N(\\mu_i, \\sigma^2)$", "$\\text{Bernoulli}(\\mu_i)$",
                   "$\\text{Poisson}(\\mu_i)$"),
  Link = c("identity, $\\mu_i$", "logit, $\\log\\left[\\mu_i/(1 - \\mu_i)\\right]$",
           "log, $\\log \\mu_i$"),
  Mean = c("$\\eta_i$", "$e^{\\eta_i}/\\left(1 + e^{\\eta_i}\\right)$",
           "$e^{\\eta_i}$"),
  Variance = c("$\\sigma^2$", "$\\mu_i(1 - \\mu_i)$", "$\\mu_i$"),
  family = c("`gaussian`", "`binomial`", "`poisson`")
) |>
  knitr::kable(col.names = c("Model", "Response", "Distribution of $Y_i$",
                             "Link $g(\\mu_i)$", "Mean $\\mu_i$",
                             "Variance", "`glm()` family"),
               align = "lllllll")
```
:::

For each model the table gives the response type, the distribution of $Y_i$,
the link, the mean as a function of $\eta_i$, the variance, and the `glm()`
family that fits it (`family = gaussian` gives the same estimates as `lm()`).

```{r glm-table-display}
#| echo: false
#| tbl-colwidths: [14, 12, 15, 20, 16, 12, 11]
glm_table
```

In logistic regression the mean $\mu_i = P(Y_i = 1 \mid x_i)$ is the
probability the [Simple Logistic Regression](10-logistic-regression.qmd)
lecture and its successors wrote $p(x_i)$; those lectures,
with the credit card default data, are this table's logistic regression row.

Each distribution fixes how the variance depends on the mean,

$$Var[Y_i \mid x_i] = \phi\, V(\mu_i),$$

where

- $V$ is the **variance function**: $V(\mu) = 1$ for the normal,
$V(\mu) = \mu(1 - \mu)$ for the Bernoulli, and $V(\mu) = \mu$ for the Poisson,
and
- $\phi$ is the **dispersion parameter**: $\phi = \sigma^2$ for the normal,
and $\phi = 1$ for the Bernoulli and the Poisson, which have no variance
parameter separate from the mean.

Linear regression's constant variance is the special case $V(\mu) = 1$, the
assumption the fanning residuals of the bike-count linear regression
contradicted. The two other models require the variance to change with the
mean, as the bike counts' variance did in the mean–variance plot.

A GLM transforms the **mean**, not the response. The regression lectures'
remedy for a variance that grows with the mean was to model $\log Y_i$, which
describes $E[\log Y_i \mid x_i]$ and fails for a count of zero (`bikers` has
none, but counts in general can). A log link models $\log E[Y_i \mid x_i]$
and never takes the logarithm of $y_i$.

::: {.callout-note collapse="true"}
### Beyond this course
The three distributions belong to the **exponential family**, and each row
uses its **canonical link**, under which the score equations take the form
$\sum_i x_{ij}(y_i - \mu_i) = 0$ derived below. The same template covers
other distributions, other links, and more than two classes:

- **Gamma regression**, for a positive, right-skewed response such as a claim
amount, has $V(\mu) = \mu^2$: `family = Gamma(link = "log")`.
- **Probit regression** keeps the Bernoulli and uses the inverse standard
normal cumulative distribution function as link: `binomial(link = "probit")`.
- **Multinomial logistic regression**, for an unordered response with $K > 2$
classes (ISLR2 Section 4.3.5), extends the Bernoulli to a categorical
distribution with $K - 1$ linear predictors, one for each class other than a
reference class; it is fit with `nnet::multinom()`.
- **Ordinal (proportional-odds) regression**, for ordered classes, uses one
linear predictor together with $K - 1$ cutpoints; it is fit with
`MASS::polr()`.

None of these is taught or tested here.
:::

## Poisson regression

::: {.callout-note collapse="true"}
### Code
```{r observed-means}
#| output: false
cell_means <- bike |>
  group_by(weather, working, hour) |>
  summarize(n = n(), observed = mean(bikers), .groups = "drop")

observed_base <- ggplot(cell_means, aes(x = hour, y = observed)) +
  geom_hline(yintercept = 0, color = "grey50", linetype = "dotted") +
  geom_point(size = 1.3, shape = 1) +
  facet_grid(weather ~ working, scales = "free_y",
             labeller = labeller(working = working_labels)) +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  labs(x = "Hour of the day", y = "Mean bikes rented in the hour",
       color = NULL, linetype = NULL) +
  theme(legend.position = "none")

observed_plot <- observed_base + labs(title = "Observed hourly means")

# Rain/snow vs. clear on working days: ratio and difference of hourly means.
rain_vs_clear <- cell_means |>
  filter(working == "Yes", weather %in% c("clear", "rain/snow")) |>
  pivot_wider(id_cols = hour, names_from = weather, values_from = observed) |>
  mutate(ratio = `rain/snow` / clear, diff = clear - `rain/snow`)
at_3  <- unlist(filter(rain_vs_clear, hour == 3))
at_17 <- unlist(filter(rain_vs_clear, hour == 17))
```
:::

The figure splits the hours by weather (rows) and working-day status
(columns); each point is the observed mean count for one hour of the day in
that panel, each row has its own vertical scale, and the dotted line marks
zero.

```{r observed-means-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 6.5
observed_plot
```

Each weather row repeats the hourly profile of the clear row at a lower
level. On working days the `rain/snow` mean is `r sprintf("%.2f", at_3["ratio"])`
times the clear mean at 3 a.m. and `r sprintf("%.2f", at_17["ratio"])` times
at 5 p.m., while the differences are `r fmt(at_3["diff"], 1)` and
`r fmt(at_17["diff"], 1)` riders: rain scales the profile, which is a
multiplicative effect, rather than lowering every hour by the same number.

### Log-linear model

**Poisson regression** is the GLM with a Poisson random component and a log
link:

$$Y_i \mid x_i \stackrel{ind}{\sim} \text{Poisson}(\mu_i), \qquad \log \mu_i = x_i\beta = \beta_0 + \beta_1 x_{i1} + \cdots + \beta_p x_{ip}.$$

Inverting the link,

$$\mu_i = e^{x_i\beta} = e^{\beta_0} \times e^{\beta_1 x_{i1}} \times \cdots \times e^{\beta_p x_{ip}},$$

which is positive for every $\beta$ and every row $x_i$. The features add on
the log scale and multiply on the scale of the mean. A weather dummy therefore
multiplies every hour's mean by the same factor $e^{\beta_j}$, the scaling of
the `rain/snow` row in the figure above.

### Maximum likelihood estimation

For independent observations the log-likelihood is the sum of the logarithms
of the probability mass functions,

$$\ell(\beta) = \sum_{i=1}^n \left[y_i \log \mu_i - \mu_i - \log(y_i!)\right] = \sum_{i=1}^n \left[y_i\, x_i\beta - e^{x_i\beta} - \log(y_i!)\right],$$

where

- $y_i$ is the observed count for observation $i$, and
- $\log(y_i!)$ does not involve $\beta$ and so does not affect the maximizer.

Differentiating with respect to $\beta_j$, using
$\partial e^{x_i\beta} / \partial \beta_j = x_{ij} e^{x_i\beta}$, and setting
the result to zero gives the **score equations**

$$\sum_{i=1}^n x_{ij}\left(y_i - \mu_i\right) = 0, \qquad j = 0, 1, \ldots, p,$$

with $x_{i0} = 1$. They have exactly the form of the logistic regression score
equations with $\mu_i$ in place of $p(x_i)$: at the solution the residuals
$y_i - \hat\mu_i$, where $\hat\mu_i = e^{x_i\hat\beta}$ is the fitted mean,
sum to zero and are orthogonal to every feature. Because
$\mu_i = e^{x_i\beta}$ is nonlinear in $\beta$, the equations in general have
no closed-form solution: with only the hour-by-working-day dummies the
solution is the cell means (which the score check below shows the fit
reproduces), but with a quantitative feature such as temperature it has none.

`glm()` solves them with the iterative algorithm it used for logistic
regression, **iteratively reweighted least squares**: each iteration solves a
weighted least squares problem whose weights are recomputed from the current
estimate. Fitting the bike counts with the features of the linear regression:

```{r bike-glm}
bike_glm <- glm(bike_formula, data = bike, family = poisson)
```

::: {.callout-note collapse="true"}
### Code
```{r score-check}
#| output: false
bike <- bike |>
  mutate(fit_glm = fitted(bike_glm))

combination_gap <- bike |>
  group_by(working, hour) |>
  summarize(gap_glm = abs(mean(bikers) - mean(fit_glm)),
            gap_lm  = abs(mean(bikers) - mean(fit_lm)), .groups = "drop")
max_gap <- max(combination_gap$gap_glm, combination_gap$gap_lm)
```
:::

The model has `r length(coef(bike_glm))` coefficients, as the linear
regression does, and `glm()` reports convergence after `r bike_glm$iter`
iterations. The intercept's score equation makes the fitted means sum to the
observed counts,
$\sum_i \hat\mu_i = `r format(round(sum(bike$fit_glm)), big.mark = "{,}")`$ and
$\sum_i y_i = `r format(sum(bike$bikers), big.mark = "{,}")`$.
A dummy variable's equation makes the fitted means average to the observed
mean where that dummy is $1$, as the least squares normal equations also do,
so both models reproduce the observed mean of each of the
`r nrow(combination_gap)` combinations of hour and working-day status (largest
gap $`r if (max_gap > 0) sci(max_gap) else "0"`$ bikers).

### Fitted means

::: {.callout-note collapse="true"}
### Code
```{r fitted-means}
#| output: false
cell_fits <- bike |>
  group_by(weather, working, hour) |>
  summarize(`Linear regression` = mean(fit_lm),
            `Poisson regression` = mean(fit_glm), .groups = "drop") |>
  right_join(cell_means, by = c("weather", "working", "hour"))

cell_long <- cell_fits |>
  pivot_longer(c(`Linear regression`, `Poisson regression`),
               names_to = "model", values_to = "fitted")

rain_night <- filter(cell_fits, weather == "rain/snow", hour <= 6)

# Every tab draws the fitted means invisibly so both tabs share y axes.
shared_axes <- geom_blank(data = cell_long, aes(y = fitted))

model_layer <- function(m) {
  list(geom_line(data = filter(cell_long, model == m),
                 aes(y = fitted, color = model, linetype = model),
                 linewidth = 0.8),
       scale_color_manual(values = model_colors),
       scale_linetype_manual(values = model_linetypes))
}
fitted_lm_plot <- observed_base + shared_axes +
  model_layer("Linear regression") +
  labs(title = "Linear regression fitted means")
fitted_glm_plot <- observed_base + shared_axes +
  model_layer("Poisson regression") +
  labs(title = "Poisson regression fitted means")
```
:::

Each tab redraws the observed hourly means of the figure above and adds a
line through one model's fitted means, averaged within each point's group;
the two tabs share their vertical axes, and each tab's title names the model.

::: {.panel-tabset}
#### Linear regression
```{r fitted-lm-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 6.5
fitted_lm_plot
```

#### Poisson regression
```{r fitted-glm-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 6.5
fitted_glm_plot
```
:::

Pooled over weather, the two models agree, since both reproduce each
hour-by-working-day mean; they differ only in how they split each combination
across the weather rows. In the `rain/snow` row, linear regression subtracts
its constant weather coefficient from every hour, and its lines fall below
zero overnight and in the early morning, as low as `r fmt(min(rain_night$"Linear regression"), 1)`
bikers, where the observed means never fall below
`r fmt(min(rain_night$observed), 1)`. The Poisson lines scale the hourly
profile down instead and follow the points much more closely, overnight
included.

### Coefficient interpretation

Write $\mu(x)$ for the mean count at a row of feature values
$x = (1, x_1, \ldots, x_p)$. At $x = (1, 0, \ldots, 0)$,
$\log \mu(x) = \beta_0$, so $e^{\beta_0}$ is the **mean count when every
feature is zero** — for dummy variables, at the reference levels.

Increasing feature $j$ by one unit from $x_j$ while holding every other
feature fixed, every term of the linear predictor except the $j$th cancels:

$$\log \mu(x_j + 1) - \log \mu(x_j) = \beta_j, \qquad \frac{\mu(x_j + 1)}{\mu(x_j)} = e^{\beta_j},$$

where

- $\mu(x_j)$ and $\mu(x_j + 1)$ are the mean counts at two feature rows that
differ only in feature $j$, by one unit.

So $\beta_j$ is the change in the log mean count per one-unit increase in
$x_j$, and $e^{\beta_j}$, the **rate ratio**, is the factor multiplying the
mean count, holding the other features fixed — the odds ratio of logistic
regression with the mean count in place of the odds. $e^{\beta_j} = 1$ means
no association; $c$ units multiply the mean by $e^{c\beta_j}$; for a dummy
variable, $e^{\beta_j}$ is the mean at the indicated level divided by the mean
at the reference level; and with an interaction $x_j x_l$ the rate ratio for
$x_j$ depends on $x_l$.

::: {.callout-note collapse="true"}
### Code
```{r rate-ratios}
#| output: false
b     <- coef(bike_glm)
shown <- c("(Intercept)", "workingYes", "temp_c", "weathercloudy/misty",
           "weatherrain/snow")
shown_labels <- c("Intercept", "Working day (at hour 0)",
                  "Temperature (per degree C)", "Cloudy/misty vs. clear",
                  "Rain/snow vs. clear")

rate_ratio_table <- tibble(
  Coefficient = shown_labels,
  Estimate = sprintf("%.4f", b[shown]),
  exp_b = sprintf("%.4f", exp(b[shown]))
) |>
  knitr::kable(col.names = c("Coefficient", "Estimate",
                             "$e^{\\hat\\beta_j}$"),
               align = "lcc")

z  <- summary(bike_glm)$coefficients[, "z value"]
ci <- confint.default(bike_glm)
rr_ci <- function(term, c = 1) {
  sprintf("(%.3f, %.3f)", exp(c * ci[term, 1]), exp(c * ci[term, 2]))
}

rr_working_8 <- exp(b["workingYes"] + b["hr8:workingYes"])
rr_temp_5    <- exp(5 * b["temp_c"])
rr_rain      <- exp(b["weatherrain/snow"])
```
:::

For the bike counts, the intercept and the coefficients for working day,
temperature, and weather, with their exponentials:

```{r rate-ratio-table}
#| echo: false
#| tbl-colwidths: [50, 25, 25]
rate_ratio_table
```

$e^{\hat\beta_0} = `r sprintf("%.1f", exp(b[1]))`$ is the estimated mean count
at hour $0$ (midnight to 1 a.m.) on a non-working day in January, in clear
weather at $0$ degrees Celsius — a realistic winter night, unlike the linear
regression's intercept of `r fmt(b_lm[1], 1)` bikers for the same conditions.

Because hour and working day interact, the working-day coefficient compares
the two kinds of day at the reference hour only: at midnight a working day's
estimated mean is $e^{\hat\beta_{\texttt{workingYes}}} = `r sprintf("%.2f", exp(b["workingYes"]))`$
times a non-working day's. At 8 a.m. the corresponding ratio adds the
`hr8:workingYes` coefficient to the exponent,
$e^{`r sprintf("%.3f", b["workingYes"])` `r sprintf("%+.3f", b["hr8:workingYes"])`} = `r sprintf("%.2f", rr_working_8)`$.

Holding hour, working-day status, month, and weather fixed, each additional
degree Celsius multiplies the estimated mean count by
`r sprintf("%.4f", exp(b["temp_c"]))`, so $5$ degrees multiply it by
$e^{5\hat\beta_{\texttt{temp}\_\texttt{c}}} = `r sprintf("%.3f", rr_temp_5)`$, an increase of
`r sprintf("%.1f", 100 * (rr_temp_5 - 1))`%.

Rain or snow multiplies the estimated mean by
`r sprintf("%.3f", rr_rain)` relative to clear weather — a
`r sprintf("%.0f", 100 * (1 - rr_rain))`% reduction at every hour, month, and
temperature, the scaling seen in the `rain/snow` row of the fitted-means
figure — and cloudy or misty weather multiplies it by
`r sprintf("%.3f", exp(b["weathercloudy/misty"]))`.

### Inference

Inference repeats the logistic regression lectures. In large samples
$\hat\beta_j$ is approximately normal, so

$$z = \frac{\hat\beta_j - \beta_j^0}{\text{SE}\left(\hat\beta_j\right)}$$

is approximately standard normal under $H_0: \beta_j = \beta_j^0$, where
$\text{SE}\left(\hat\beta_j\right)$ comes from the curvature of $\ell$ at $\hat\beta$. The
Wald interval $\hat\beta_j \pm z_{1-\alpha/2}\,\text{SE}\left(\hat\beta_j\right)$ is
returned by `confint.default()`, and exponentiating its endpoints gives an
interval for the rate ratio $e^{\beta_j}$.

For the four rate ratios of the bike-count table, $|z|$ ranges from
`r sprintf("%.1f", min(abs(z[shown[-1]])))` to
`r sprintf("%.1f", max(abs(z[shown[-1]])))`, and the 95% intervals are
$`r rr_ci("workingYes")`$ for working day at hour $0$,
$`r rr_ci("temp_c", 5)`$ per $5$ degrees Celsius,
$`r rr_ci("weathercloudy/misty")`$ for cloudy or misty weather, and
$`r rr_ci("weatherrain/snow")`$ for rain or snow: every one excludes $1$.

A group of coefficients — every product of an hour dummy with the working-day
indicator, say — is tested with the drop-in-deviance test. For a count
response `glm()` reports the residual deviance

$$D = 2\left[\ell_{\text{sat}} - \ell\left(\hat\beta\right)\right],$$

where

- $D$ is the deviance, as in the
[Multiple Logistic Regression](20-multiple-logistic-regression.qmd) lecture, not an
indicator variable, and
- $\ell_{\text{sat}}$ is the log-likelihood of the **saturated** model, which
gives each observation its own mean $\mu_i = y_i$, the largest log-likelihood
any mean model can reach.

$\ell_{\text{sat}}$ does not involve $\beta$, so for a reduced model nested in
a full one, $G^2 = D_r - D_f = -2\left[\ell\left(\hat\beta_r\right) - \ell\left(\hat\beta_f\right)\right]$,
the drop-in-deviance statistic of the multiple logistic regression lecture.
(For a binary response $\ell_{\text{sat}} = 0$, which is why that lecture
could write $D = -2\ell\left(\hat\beta\right)$.) Under $H_0$ that the $k$ dropped
coefficients are zero, $G^2$ is approximately $\chi^2_k$, and
`anova(reduced, full, test = "Chisq")` performs the test.

The bike-count model's `r sum(grepl(":", names(b)))` hour-by-working-day
products are such a group: no single $z$-test asks whether working and
non-working days share one hourly profile up to a constant factor. The test
compares the full model with the additive one:

```{r interaction-test}
bike_additive <- glm(bikers ~ hr + working + mnth + temp_c + weather,
                     data = bike, family = poisson)
interaction_test <- anova(bike_additive, bike_glm, test = "Chisq")
interaction_test
G2 <- interaction_test$Deviance[2]
k  <- interaction_test$Df[2]
```

The deviance drops by $G^2 = `r format(round(G2), big.mark = "{,}")`$ on
$k = `r k`$ degrees of freedom, against a $\chi^2_{`r k`}$ mean of `r k`. The
data are inconsistent with one hourly profile, up to a constant factor, shared
by both kinds of day.

### Overdispersion

The Poisson model has no variance parameter: $Var[Y_i \mid x_i] = \mu_i$, and
every standard error and deviance test above is computed under that variance.
**Overdispersion** is the failure $Var[Y_i \mid x_i] > \mu_i$.

The **Pearson residual** $\left(y_i - \hat\mu_i\right)/\sqrt{\hat\mu_i}$ standardizes a
residual by the Poisson standard deviation. If $Var[Y_i \mid x_i] = \mu_i$,
then $E\left[(Y_i - \mu_i)^2/\mu_i \mid x_i\right] = 1$, so the average
squared Pearson residual,

$$\hat\phi = \frac{1}{n - (p+1)}\sum_{i=1}^n \frac{\left(y_i - \hat\mu_i\right)^2}{\hat\mu_i},$$

should be close to $1$, and a value well above $1$ is evidence of
overdispersion, where

- $\hat\phi$ estimates the dispersion parameter $\phi$, which the Poisson
model fixes at $1$, and
- the divisor $n - (p+1)$ accounts for the $p + 1$ estimated coefficients, as
in $\hat\sigma^2 = RSS/[n - (p+1)]$.

::: {.callout-note collapse="true"}
### Code
```{r variance-check}
#| output: false
pearson_phi <- function(fit) {
  sum(residuals(fit, type = "pearson")^2) / df.residual(fit)
}
phi_full     <- pearson_phi(bike_glm)
phi_additive <- pearson_phi(bike_additive)

n_bins <- 20
variance_bins <- bike |>
  mutate(bin = ntile(fit_glm, n_bins)) |>
  group_by(bin) |>
  summarize(mean_fit = mean(fit_glm),
            mean_sq_resid = mean((bikers - fit_glm)^2), .groups = "drop")

variance_check_plot <- ggplot(variance_bins,
                              aes(x = mean_fit, y = mean_sq_resid)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
              color = "grey40") +
  geom_point(size = 2.3) +
  annotate("text", x = 100, y = 55, label = "Poisson: variance = mean",
           angle = 24, color = "grey30") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = expression(paste("Average ", hat(mu)[i], " in the bin (log scale)")),
       y = expression(paste("Average ", (y[i] - hat(mu)[i])^2,
                            " in the bin (log scale)")),
       title = paste("Poisson regression, hours binned into", n_bins,
                     "groups by fitted mean"))

rain_se    <- summary(bike_glm)$coefficients["weatherrain/snow", "Std. Error"]
rain_ci_q  <- exp(b["weatherrain/snow"] +
                  c(-1, 1) * qnorm(0.975) * sqrt(phi_full) * rain_se)
```
:::

For the bike-count fit, the figure below places each of `r n_bins` groups of
hours, binned by fitted mean, at its average fitted mean and its average
squared residual, on log scales; the dashed line is variance $=$ mean.

```{r variance-check-figure}
#| echo: false
#| fig-width: 6.5
#| fig-height: 4.5
variance_check_plot
```

`r if (all(variance_bins$mean_sq_resid > variance_bins$mean_fit)) paste("All", nrow(variance_bins), "bins have") else paste(sum(variance_bins$mean_sq_resid > variance_bins$mean_fit), "of the", nrow(variance_bins), "bins have")`
an average squared residual above the dashed line. The average squared Pearson
residual is $\hat\phi = `r sprintf("%.1f", phi_full)`$: the hourly counts have
about `r round(phi_full)` times the variance around their fitted means that a
Poisson distribution allows. The additive model's $\hat\phi$ is
`r sprintf("%.1f", phi_additive)`, so the hour-by-working-day features
explained part of the excess variation, but $\hat\phi$ remains far above
$1$.

The score equations involve $y_i$ only through $y_i - \mu_i$, never through
its variance, so if the mean model is right $\hat\beta$, the fitted means, and
the rate ratios remain sensible estimates under overdispersion. The standard
errors, computed as if the variance were $\mu_i$, are too small, so
$z$-statistics are too large and confidence intervals too narrow, and $G^2$ is
too large relative to its $\chi^2_k$ reference. If
$Var[Y_i \mid x_i] = \phi\,\mu_i$, $G^2$ is approximately $\phi\,\chi^2_k$, so
$G^2/\hat\phi$ is approximately $\chi^2_k$. For the bike counts
$G^2/\hat\phi = `r format(round(G2 / phi_full), big.mark = "{,}")`$, still
`r fmt(G2 / phi_full / k)` times its $`r k`$ degrees of freedom: the
interaction test's conclusion survives.

Overdispersion has two sources, as in the problems in logistic regression
lecture: explanatory variables missing from the model that vary among
observations with the same $x_i$ (for the bike counts, plausibly conditions
shared by the hours of one day), and events within one count that are not
independent, such as riders who rent together. Adding features addresses only
the first; the second falls under the remedies of the callout below.

::: {.callout-note collapse="true"}
### Beyond this course
Two standard remedies change the variance model rather than the mean model.
**Quasi-Poisson** regression keeps $\log \mu_i = x_i\beta$ and the estimates
$\hat\beta$ but assumes $Var[Y_i \mid x_i] = \phi\,\mu_i$, estimating $\phi$ by
$\hat\phi$ and multiplying every standard error by $\sqrt{\hat\phi}$
(`family = quasipoisson` in `glm()`); deviance tests become $F$-tests. For the
bike counts the factor is
$\sqrt{\hat\phi} = `r sprintf("%.2f", sqrt(phi_full))`$, and the 95% interval
for the `rain/snow` rate ratio widens from
$(`r sprintf("%.3f", exp(ci["weatherrain/snow", 1]))`, `r sprintf("%.3f", exp(ci["weatherrain/snow", 2]))`)$
to $(`r sprintf("%.3f", rain_ci_q[1])`, `r sprintf("%.3f", rain_ci_q[2])`)$,
still far from $1$. **Negative binomial** regression replaces the Poisson with
a distribution whose variance is $\mu_i + \mu_i^2/\theta$ for an additional
parameter $\theta$, fit by maximum likelihood with `MASS::glm.nb()`. Neither is
taught or tested here.
:::

## Conclusion

A generalized linear model has three components: a distribution for the
response given the features, with mean $\mu_i$; a linear predictor
$\eta_i = x_i\beta$, into which every feature-engineering device enters
unchanged; and a link function with $g(\mu_i) = \eta_i$, whose inverse keeps
the mean in range. Linear, logistic, and Poisson regression pair the normal,
Bernoulli, and Poisson distributions with the identity, logit, and log links,
and variance functions $1$, $\mu(1-\mu)$, and $\mu$.

Poisson regression models a count. Its mean $e^{x_i\beta}$ is positive for
every $\beta$; $e^{\beta_0}$ is the mean count when every feature is zero,
and $e^{\beta_j}$ a rate ratio, the factor multiplying the mean count
per unit of feature $j$ with the other features held fixed. Its coefficients
are maximum likelihood estimates; the score equations
$\sum_i x_{ij}(y_i - \mu_i) = 0$ have the logistic regression form, have no
closed-form solution, and are solved by iteratively reweighted least squares.
Inference uses the same Wald $z$-tests and intervals and the same
drop-in-deviance test as logistic regression.

All of that inference assumes $Var[Y_i \mid x_i] = \mu_i$. The average
squared Pearson residual estimates the ratio of the actual variance to the
mean; a value well above $1$ signals overdispersion, under which the estimates
remain usable if the mean model is right, but standard errors are too small,
tests too liberal, and intervals too narrow.

The next lecture begins the resampling unit, ISLR2 Chapter 5:
cross-validation estimates test error by repeatedly holding out part of the
data, and uses those estimates to choose among models and flexibility
settings.
